### 2012-12-31 edupar-13/conclusion.tex
more setup

\section{Conclusion}

we did stuff

\section*{Acknowledgment}

thanks for reading

### 2013-01-01 edupar-13/intro.tex
write intro

\section{Introduction}

Managing parallel computation, whether on thousand-node clusters or on one single-core machine, is often taught by resorting to powerful abstractions that simplify the hardest problems.
Programming patterns such as divide-and-conquer and fork/join parallelism enable developers to reason about parallel algorithms at a high, abstract level.
With the advent of massively multicore hardware and distributed systems, however, there is also a need for reasoning about parallelism at lower levels: those of operating system kernels and concurrent systems applications, which are built on (or even below) the abstractions of threads and processes.
Low-level systems programmers must often reason about parallel code without the convenience of high-level simplifications, and hence are often susceptible to notoriously difficult concurrency errors.

In this paper, we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively challenging kernel project, in which they implement threads and processes in code that will boot on real hardware.
Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in userspace on top of a Pebbles reference kernel in a two-week project, and subsequently, in a 6-week project, write their own kernel implementation.
Finally, in another two-week project, students implement a small extension to their Pebbles kernel. This project changes every semester, and in the past has included extending the kernel to be able to hypervise other kernels and to be able to run in SMP systems.

As we ask our students to write more and more difficult concurrent code in the same semester, we want to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time during the projects.
However, we also want to avoid compromising the complexity and difficulty of the project's concurrency challenges. % how to say... "because it's the most valuable thing we do"
Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads into execution sequences that expose concurrency bugs.
We present our experience sharing a prototype of Landslide with volunteering students who used it to find bugs in their own code, and discuss the possibility for integrating it as a main component of the 15-410 curriculum.

% should we do a "The contributions of this paper are as follows:"?

The rest of this paper is organized as follows.
Section~\ref{sec:projects} gives an overview of 15-410's five programming projects.
Section~\ref{sec:pebbles} discusses the Pebbles specification in detail.
Section~\ref{sec:landslide} describes Landslide's testing mechanisms.
Section~\ref{sec:future} concludes, discussing strategies for making Landslide more accessible to struggling students, and how Landslide might be applied to related operating systems projects at other universities.

### 2013-01-02 edupar-13/410.tex
write a ton more

\section{The 15-410 Curriculum}
\label{sec:projects}

\section{Pebbles}
\label{sec:pebbles}

### 2013-01-02 edupar-13/eval.tex
write a ton more

\section{Evaluation}
\label{sec:eval}

We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.

We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
We configured Landslide to search for five complicated already-known race conditions.
In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
Using decision points only on calls to \x{mutex_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.

In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two races and two deterministic errors.
These bugs manifested as infinite loops, a kernel panic, and a use-after-free.

% TODO: decide whether to put table 7.2 from the thesis here. i think probably not?

### 2013-01-02 edupar-13/landslide.tex
write a ton more

\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{grey}{RGB}{127,127,127}
\definecolor{darkcyan}{RGB}{0,127,127}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{violet}{RGB}{127,0,127}
\definecolor{brickred}{RGB}{127,0,0}
\definecolor{brown}{RGB}{127,63,0}
\definecolor{red}{RGB}{127,0,0}

\section{Landslide}
\label{sec:landslide}

Landslide is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
It follows in the footsteps of related tools such as CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter}, but is the first tool we know of to apply systematic exploration in a kernel environment.
In this section we give an overview of Landslide's design and interface, and point out the annotations students need to provide to make Landslide work with their own code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scope}

\paragraph{Testing environment}
Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels. Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and once per memory access of the simulated execution.
Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.

\paragraph{Limitations}
Landslide assumes that timer interrupts are the only nondeterministic events that influence a kernel's concurrent execution. While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
Landslide's model is also restricted to uniprocessor execution.
Supporting multiprocessor environments, device driver testing, and production kernels is left to future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Testing Mechanism}

In Figure~\ref{fig:threadfork} we show a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section. This code is a simplified example of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID. However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run and invoke \x{exit}, causing \x{child} to become a dangling pointer.

\begin{figure}[t]
\small
\begin{lstlisting}[numbers=left]
int thread_fork() {
	thread_t *child = construct_new_thread();
	add_to_runqueue(child);
	// at this point child may run and exit#\label{line:hax}#
	return child->tid;#\label{line:owned}#
}
\end{lstlisting}
\caption{Example race condition. If a timer interrupt occurs at line~\ref{line:hax}, the child thread can run, exit, and free its state, causing line~\ref{line:owned}'s access to be a use-after-free.}
\label{fig:threadfork}
\end{figure}

Systematic exploration, Landslide's mechanism for testing different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings in a system. These interleavings are defined around {\em decision points}, which intuitively indicate points during execution at which a preemption could cause different behaviour to arise.
Searching with few decision points results in coarser-grained interleavings, shorter test execution, and less likelihood of finding bugs; searching with more results in the opposite.

In our example bug, the necessary decision point for finding the bug is at line~\ref{line:hax}. With correct annotations (discussed in the next section and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point.
Landslide also automatically identifies a minimal set of decision points at voluntary reschedules between threads (e.g. \texttt{yield}). Together these decision points define an execution tree which exposes the bug, depicted in Figure~\ref{fig:tree}.

Of course, additional decision points are necessary to expose more subtle race conditions. We provide an interface for refining the state space, described in the next section.

\begin{figure}[t]
\includegraphics[width=0.48\textwidth]{threadfork/threadfork.pdf}
\caption{The state space of possible thread interleavings can be viewed as an {\em execution tree}.
Landslide uses timer interrupts to force different threads to run as it explores this tree.
If a kernel has concurrency errors, they will show up in some branches, but not others.
This tree shows the bug from Figure~\ref{fig:threadfork}, with decision points from the parent thread shaded.
}
\label{fig:tree}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Instrumentation}

Landslide's interface is comprised of two types of instrumentation which students must provide: {\em required annotations} and {\em configuring decision points}.

\paragraph{Required annotations}
Users annotate their kernels to inform Landslide of certain important concurrency events during execution. We provide a set of annotation functions, called \x{tell_landslide}, for this purpose. The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
Figure~\ref{fig:annotation} shows the code from Figure~\ref{fig:threadfork} with \x{tell_landslide} annotations.

There is also a configuration file, \x{config.landslide}, in which the user must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which userspace test program Landslide should invoke.
Finally, there are two short (nominally two-line) functions used within Landslide itself that the user must implement. These are predicates on the kernel's scheduler state, and express potentially nontrivial conditions: whether the current thread is runnable but not on the runqueue, and whether preemption is disabled while interrupts are on.

\newcommand\telllandslide[1]{\bfseries \color{violet}{#1}}
\begin{figure}[t]
\small
\begin{lstlisting}
void add_to_runqueue(thread_t *child) {
	#\telllandslide{tell\_landslide\_thread\_runnable(child->tid);}#
	// ...
}
int thread_fork() {
	thread_t *child = construct_new_thread();
	#\telllandslide{tell\_landslide\_forking(child->tid);}#
	add_to_runqueue(child);
	return child->tid;
}
\end{lstlisting}
\caption{Landslide requires the programmer to annotate important concurrency events in their kernel. \texttt{tell\_landslide\_thread\_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \texttt{tell\_landslide\_forking()} announces a new thread's existence.}
\label{fig:annotation}
\end{figure}

\paragraph{Configuring decision points}

% TODO

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Identifying Bugs}

Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.

Landslide can optionally also check heuristically for infinite loops and livelock. This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
We have never found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.

When Landslide identifies a bug, it outputs a {\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.
This trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that we caused to preempt it. With this trace, the student can better understand the concurrent execution that led to the bug.

In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it. However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that a race exists.

\newcommand\bug[1]{\hilight{red}{#1}}
\newcommand\decision[1]{\hilight{olivegreen}{#1}}
\newcommand\stacktrace[1]{\hilight{darkcyan}{#1}}
\begin{figure}[t]
\begin{lstlisting}
#\bfseries \bug{USE~AFTER~FREE:~read~from~0x15a8f0~at~IP~0x104209}#
#\bug{Block~0x15a8f0~was~allocated~by~thread~3~at~(...)}#
                   #\bug{and~freed~by~thread~4~at~(...)}#
Decision trace follows:
#\decision{1:  switched from thread 3 -> thread 4 at:}#
        0x105a10 in #\stacktrace{context\_switch}#,
        0x1041f4 in #\stacktrace{thread\_fork}#,
        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
#\decision{2:  switched from thread 4 -> thread 3 at:}#
        0x105a10 in #\stacktrace{context\_switch}#,
        0x104681 in #\stacktrace{yield}#,
        0x104570 in #\stacktrace{exit}#,
        0x103708 in #\stacktrace{exit\_wrapper}#
#\decision{Current thread 3 at:}#
        0x104209 in #\stacktrace{thread\_fork}#,
        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
Total decision points 24, total backtracks 5
\end{lstlisting}
\caption{Landslide outputs this trace for Figure~\ref{fig:threadfork}'s bug.}
\label{fig:trace}
\end{figure}

### 2013-01-02 edupar-13/landslide.tex
write about configuring

@@
 Landslide's model is also restricted to uniprocessor execution.
-Supporting multiprocessor environments, device driver testing, and production kernels is left to future work.
+Supporting multiprocessor environments and device driver testing is left to future work.
+% this used to say ", and production kernels"
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Testing Mechanism}
+\subsection{Example}
+% TODO: if necessary, this can be not its own section
 
-In Figure~\ref{fig:threadfork} we show a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section.
-This code is a simplified example of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID.
+Figure~\ref{fig:threadfork} shows a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section.
+This code is a simplified implementation of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID.
 However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run and invoke \x{exit}, causing \x{child} to become a dangling pointer.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Testing Mechanism}
+
 Systematic exploration, Landslide's mechanism for testing different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings in a system.
@@
 Landslide also automatically identifies a minimal set of decision points at voluntary reschedules between threads (e.g.
-\texttt{yield}).
+\x{yield}).
 Together these decision points define an execution tree which exposes the bug, depicted in Figure~\ref{fig:tree}.
 
-Of course, additional decision points are necessary to expose more subtle race conditions. We provide an interface for refining the state space, described in the next section.
+Of course, additional decision points are necessary to expose more subtle race conditions. We provide an interface for refining the state space, described in Section~\ref{sec:decision}.
+In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction (DPOR) algorithm~\cite{dpor} to identify and prune redundant thread interleavings.
+While the specifics of the algorithm are beyond the educational scope of this work, we also provide an interface for helping Landslide further reduce the state space using domain knowledge about the kernel's implementation.
+We discuss this in Section~\ref{sec:focusing}.
 
@@
 
-\paragraph{Required annotations}
+\subsubsection{Required annotations}
 Users annotate their kernels to inform Landslide of certain important concurrency events during execution.
@@
 \caption{Landslide requires the programmer to annotate important concurrency events in their kernel.
-\texttt{tell\_landslide\_thread\_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \texttt{tell\_landslide\_forking()} announces a new thread's existence.}
+\x{tell_landslide_thread_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \x{tell_landslide_forking()} announces a new thread's existence.}
 \label{fig:annotation}
@@
 
-\paragraph{Configuring decision points}
+\subsubsection{Configuring decision points}
+\label{sec:decision}
 
-% TODO
+Using only the decision points automatically identified on voluntary reschedules will result in coarse-grained interleavings likely to overlook bugs.
+We provide an extra annotation
+for the user to add more decision points for a finer-grained search, called
+\x{tell_landslide_decide()}.
+We recommend using it in concurrency primitives, such as at the start of \x{mutex_lock()} and at the end of \x{mutex_unlock()}.
+
+\subsubsection{Focusing the search space}
+\label{sec:focusing}
+
+The strategy we described above may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated system calls.
+We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively limit them.
+First, the student writes ``\x{DECISION_INFO_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
+Then, they write ``\x{within_function} \x{foo}'' to whitelist function \x{foo}, and ``\x{without_function} \x{bar}'' to blacklist function \x{bar}.
+For example, if testing thread death and reaping, the student should write \x{within_function exit} and \x{within_function} \x{wait}, and also
+(assuming they don't care about the virtual memory operations associated with \x{exit})
+\x{without_function} \x{destroy_address_space}.
+
+Additionally, the student may configure Landslide's state space reduction to ignore certain memory locations.
+DPOR's analysis requires a memory independence relation between thread transitions; the more transitions are independent from each other, the more reduction can be achieved.
+Unfortunately, in the kernel environment, every thread switch will execute through a common path that modifies shared scheduler data structures.
+Unchecked, this would cause every transition to conflict with each other and result in no reduction in execution time from DPOR.
+Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested; for example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
+
+We provide an interface with which the user can configure Landslide to ignore such unrelated conflicts to more efficiently test components of the kernel they care about.
+In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
 

### 2013-01-03 edupar-13/landslide.tex
Move 'identifying bugs' section and add two important paragraphs

@@
 
+We also note that, in contrast with stress testing, Landslide works best with very small test cases.
+While a more conventional stress test achieves coverage by exercising many system calls (in parallel and/or in sequence), Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
+Hence, if Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
+When working with students, we provided a suite of four test programs, each no longer than 10 lines of code.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Landslide ``explores'' decision trees by controlling the kernel's execution in two ways.
+First, at each decision point, it can force an arbitary thread to run in place of the current one by injecting timer interrupts to force the kernel to context switch.
+Second, whenever the kernel finishes executing a particular interleaving of the test case, it reverts the system to an earlier state to explore additional interleavings using Simics's backtracking mechanism.
+
 In our example bug, the necessary decision point for finding the bug is at line~\ref{line:hax}.
-With correct annotations (discussed in the next section and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point.
+With correct annotations (discussed in Section~\ref{sec:instrument} and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point.
 Landslide also automatically identifies a minimal set of decision points at voluntary reschedules between threads (e.g.
@@
 
-Of course, additional decision points are necessary to expose more subtle race conditions. We provide an interface for refining the state space, described in Section~\ref{sec:decision}.
+Of course, additional decision points are necessary to expose more subtle race conditions.
+We provide an interface for adding new decision points to refine the state space, described in Section~\ref{sec:decision}.
 In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction (DPOR) algorithm~\cite{dpor} to identify and prune redundant thread interleavings.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Identifying Bugs}
+
+Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.
+
+Landslide can optionally also check heuristically for infinite loops and livelock.
+This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
+We have never found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
+
+When Landslide identifies a bug, it outputs a {\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.
+This trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that we caused to preempt it.
+With this trace, the student can better understand the concurrent execution that led to the bug.
+
+In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.
+However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that a race exists.
+
+\newcommand\bug[1]{\hilight{red}{#1}}
+\newcommand\decision[1]{\bfseries \hilight{olivegreen}{#1}}
+\newcommand\stacktrace[1]{\hilight{darkcyan}{#1}}
+\begin{figure}[t]
+\begin{lstlisting}
+#\bfseries \bug{USE~AFTER~FREE:~read~from~0x15a8f0~at~IP~0x104209}#
+#\bug{Block~0x15a8f0~was~allocated~by~thread~3~at~(...)}#
+                   #\bug{and~freed~by~thread~4~at~(...)}#
+Decision trace follows:
+#\decision{1:  switched from thread 3 -> thread 4 at:}#
+        0x105a10 in #\stacktrace{context\_switch}#,
+        0x1041f4 in #\stacktrace{thread\_fork}#,
+        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
+#\decision{2:  switched from thread 4 -> thread 3 at:}#
+        0x105a10 in #\stacktrace{context\_switch}#,
+        0x104681 in #\stacktrace{yield}#,
+        0x104570 in #\stacktrace{exit}#,
+        0x103708 in #\stacktrace{exit\_wrapper}#
+#\decision{Current thread 3 at:}#
+        0x104209 in #\stacktrace{thread\_fork}#,
+        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
+Total decision points 24, total backtracks 5
+\end{lstlisting}
+\caption{Landslide outputs this trace for Figure~\ref{fig:threadfork}'s bug.}
+\label{fig:trace}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Instrumentation}
+\label{sec:instrument}
 
@@
 \caption{Landslide requires the programmer to annotate important concurrency events in their kernel.
-\x{tell_landslide_thread_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \x{tell_landslide_forking()} announces a new thread's existence.}
+\texttt{tell\_landslide\_thread\_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \texttt{tell\_landslide\_forking()} announces a new thread's existence.}
 \label{fig:annotation}
@@
 In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Identifying Bugs}
-
-Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.
-
-Landslide can optionally also check heuristically for infinite loops and livelock.
-This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
-We have never found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
-
-When Landslide identifies a bug, it outputs a {\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.
-This trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that we caused to preempt it.
-With this trace, the student can better understand the concurrent execution that led to the bug.
-
-In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.
-However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that a race exists.
-
-\newcommand\bug[1]{\hilight{red}{#1}}
-\newcommand\decision[1]{\hilight{olivegreen}{#1}}
-\newcommand\stacktrace[1]{\hilight{darkcyan}{#1}}
-\begin{figure}[t]
-\begin{lstlisting}
-#\bfseries \bug{USE~AFTER~FREE:~read~from~0x15a8f0~at~IP~0x104209}#
-#\bug{Block~0x15a8f0~was~allocated~by~thread~3~at~(...)}#
-                   #\bug{and~freed~by~thread~4~at~(...)}#
-Decision trace follows:
-#\decision{1:  switched from thread 3 -> thread 4 at:}#
-        0x105a10 in #\stacktrace{context\_switch}#,
-        0x1041f4 in #\stacktrace{thread\_fork}#,
-        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
-#\decision{2:  switched from thread 4 -> thread 3 at:}#
-        0x105a10 in #\stacktrace{context\_switch}#,
-        0x104681 in #\stacktrace{yield}#,
-        0x104570 in #\stacktrace{exit}#,
-        0x103708 in #\stacktrace{exit\_wrapper}#
-#\decision{Current thread 3 at:}#
-        0x104209 in #\stacktrace{thread\_fork}#,
-        0x10362b in #\stacktrace{thread\_fork\_wrapper}#
-Total decision points 24, total backtracks 5
-\end{lstlisting}
-\caption{Landslide outputs this trace for Figure~\ref{fig:threadfork}'s bug.}
-\label{fig:trace}
-\end{figure}

### 2013-01-05 edupar-13/conclusion.tex
write a ton more
conclusion.tex: Acknowledgements: selected 410 staff

@@
-\section{Conclusion}
+\section{Future Work and Conclusion}
+\label{sec:future}
 
@@
 
-\section*{Acknowledgment}
+\section*{Acknowledgments}
 
-thanks for reading
+Many people have contributed to the development of
+15-410 over the past decade.
+Stephen Muckle, currently at Qualcomm, led
+the switch to x86 projects running in Simics.
+Nathaniel Wesley Filardo,
+currently a Ph.D.\ candidate at Johns Hopkins,
+increased the sanity of the build infrastructure.
+The current reference kernel is named ``pathos''
+and was written by Michael J. Sullivan,
+currently a CMU CS Ph.D.\ candidate,
+and Elly Fong-Jones, currently at Google.
+Over the past decade the course has benefitted
+greatly from free educational Simics licenses,
+donated initially by Virtutech and recently by
+Wind River.

### 2013-01-05 edupar-13/410.tex
410.tex: only 75% done with first pass, but something that can be reacted to

@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{The 15-410 Curriculum}
-\label{sec:projects}
+\label{sec:curriculum}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+The work described takes place in the context of a
+Computer Science class called
+15-410, Operating System Design and Implementation.
+It is designed for juniors and seniors in Computer Science
+and Electrical and Computer Engineering.
+For CS students it is one of five courses which fulfill
+the ``Software Systems'' requirement (at present, the others
+are Compiler Design,
+Parallel Computer Architecture and Programming,
+Distributed Systems,
+and Computer Networks.
+The prerequisite course is 15-213,
+Introduction to Computer Systems,
+a course which has been adopted at other
+universities in the U.S.\ and worldwide~\cite{sigcse01:CSaPP}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% XXX How do we say politely "Some M.S. students
+% manage to pass it too"?
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In this section we will discuss the course learning
+objectives,
+outline the programming projects,
+discuss our use of Wind River Simics as the
+the basic simulation/debugging environment,
+and close by mentioning current limitations in the scope
+of the course material.
+
+The course learning objectives vary widely.
+%
+Students study high-level concepts
+such as protection (least privilege, access control lists vs.
+capabilities)
+and
+file-system internals,
+and log-based storage.
+%
+Emphasis is placed on acquiring information from ``primary sources,''
+including both manufacturer-provided hardware documentation
+and a non-textbook technical-literature reading assignment.
+%
+Students begin with a ``blank slate'' rather than a
+kernel-source template or an existing operating system,
+so they must synthesize design requirements from multiple sources
+and are provided with an opportunity to 
+decide on module boundaries and inter-module conventions
+themselves.
+%
+Due to the foundational nature of kernel code,
+assignment design and grading encourage students to
+think about corner cases including resource exhaustion
+instead of being satisfied by ``the right basic idea''
+implementations that handle only auspicious situations.
+%
+Finally, most relevant to this work,
+students gain substantial experience in
+analyzing and writing lock-based multi-threaded code and
+thread-synchronization objects, including detecting and documenting race conditions and deadlock,
+including
+the relationship between interrupts and other forms of concurrency.
+
+%=== Projects ===%
+\subsection{Projects}
+
+In the course of a semester, students work on five
+programming assignments; the first two are individual,
+and the remaining three are the products of two-person
+teams.
+
+The first project is a ``stack crawler'':  invoked by
+a client program, it displays the program's stack
+symbolically, rendering saved program-counter values
+as function names and printing function parameters
+in accordance with their types.
+This project meets a variety of goals:
+it enables students to review key process-model and
+language-runtime concepts from the
+prerequisite course;
+it introduces students to our expectations about
+design, analysis, and making choices;
+finally,
+because C pointers are unsafe, it requires students
+to consider robustness.
+%
+% and we READ THEIR CODE
+This project is built using a standard C tool chain
+and debugged as students see fit (generally using
+a combination of \texttt{printf()} and \texttt{gdb}).
+
+The second project is a simple game, such as Hangman,
+which runs standalone (without an underlying
+operating system).
+We provide students with selected C~library routines
+(\texttt{strcmp()}, \texttt{memmove()}, \texttt{printf()});
+they write a device-driver library consisting of
+console output (to a memory-mapped text-mode video display),
+keyboard input (interrupt driven),
+and a countdown timer.
+The game, which varies by semester, provides students
+with the ability to test their drivers.
+%
+% and the need to write enough code to get into trouble
+%
+This project and the remaining ones are written in
+C with a bit of x86-32 assembly code,
+compiled and linked into an ELF executable,
+stored into a 1.44-megabyte 3.5-inch floppy-disk image,
+and booted via GRUB.
+The floppy-disk image can be copied to a real floppy disk
+or embedded into an ``El Torito'' bootable compact disc image,
+in which case it can be booted on standard PC hardware.
+However, students most often boot the floppy image
+using the Simics system emulator (described below).
+
+The third project is implementing a 1:1 thread library for
+user-space programs.
+The library specification is essentially a stripped-down
+version of POSIX Pthreads.
+The primary goal of the project is providing students
+with a laboratory for
+experimenting with concurrency and atomicity.
+Students begin by designing mutexes using any
+x86-32 atomic instructions they choose
+(there is substantial variation across groups).
+They then write other thread-synchronization
+primitives (condition variables, semaphores,
+and reader/writer locks), infrastructure
+components (stack allocation/recycling and
+a thread registry),
+and low-level code to launch and shut down
+threads.
+Another goal is encouraging students to carefully
+consider how ingredients are assembled into
+abstractions.
+One example of this is a semantic gap
+between the library-level \texttt{thr\_create()}
+call, analogous to \texttt{pthread\_create()},
+and the \texttt{thread\_fork} system call provided
+by the underlying kernel,
+which takes no parameters,
+makes no changes to the user's address space,
+and cannot meaningfully
+be invoked from C~code.
+Because these abstractions are so different,
+students have the opportunity to
+understand each as an independent entity while
+deciding how to bridge them.
+A similar semantic gap exists between
+\texttt{cond\_wait()} and the system call
+which suspends execution of a thread.
+
+Student library code is linked against small
+test programs provided by the course staff,
+producing one statically linked ELF executable
+per test program.
+The test programs are bundled into a RAM-disk
+image and linked against a kernel written by
+the course staff
+and provided in binary form.
+%
+In addition to providing a reliable execution
+substrate,
+this ``reference kernel'' schedules
+the execution of user-space threads created by
+student code according to a variety of
+interleaving patterns.
+In order to aid debugging of concurrency problems,
+students can manually select among these
+interleaving patterns.
+The behavior of the reference kernel is specified
+via a twelve-page document.
+% the handout for the thread library project itself
+% is nineteen pages and includes not only a specification
+% of the thread-library primitives but also
+% a suggested implementation plan, grading guidelines,
+% optional design challenges, etc.
+
+For the fourth project, two-student teams
+implement a kernel which complies to the
+same specification as the reference kernel
+they previously relied on.
+They implement some approach to
+synchronizing and blocking threads while
+they are in kernel space,
+a simple round-robin scheduler,
+basic virtual memory,
+a program loader,
+code to handle various x86 exceptions,
+and code for setting up and tearing down
+threads and processes.
+Students re-use the basic device drivers
+they implemented in the second project.
+These ingredients are combined to yield
+twenty-four system calls
+(life-cycle, thread management, memory management,
+console I/O, and miscellaneous).
+For most students in the class, this is the
+largest and most complicated software artifact they
+have produced.
+Because the test suite and the grading criteria
+emphasize robustness and preemptibility of
+kernel code,
+there are many cross-cutting concerns.
+Because the students are responsible for ensuring
+the run-time invariants underlying all compiler-generated
+code in the system (kernel and user-space),
+they gain experience with debugging at both the
+``algorithm level'' and the register/bit-field level.
+
+Students who complete the kernel project on time
+then work on a smaller kernel-extension project
+with various content depending on the semester.
+Past projects have included
+writing a sound card driver,
+hibernation (suspend to disk),
+kernel profiling,
+and writing an in-kernel debugger.
+Two recent, more-aggressive, projects have been
+adding paravirtualization so that their kernels
+can host guest kernels and
+addding multi-processor support to their single-processor kernels.
+Meanwhile, students who were unable to complete
+the kernel project on time receive additional time
+to finish up and improve their code quality---this
+requirement is based on our belief that important
+learning experiences are obtained by solving the
+hard problems,
+which turn up at the end of the implementation.
+
+%=== Simics ===%
+\subsection{Simics}
+
+As mentioned above, the main execution and debugging
+platform used in the class is
+Simics~\cite{Magnusson:2002:SFS:619072.621909},
+originally developed by the Swedish Institute of
+Computer Science and
+currently a product of Wind River.
+Unlike some emulators,
+which focus on fast execution of correct code,
+Simics
+provides very faithful bit-level support
+not only for code that behaves correctly but
+also for code that is ``incorrect within reason.''
+Occasionally student code manages to invoke an unsupported
+function of some hardware device or even to cause the
+simulator to crash, but this is quite rare, and in almost
+every case when the ``bad event'' is characterized it turns
+out that what the software requested from the hardware was
+a mistake.
+Meanwhile, inside its large envelope, Simics supports
+\textit{very} detailed models of various x86 processors.
+Unlike hardware virtualization environments,
+Simics contains substantial debugger support:
+single-stepping,
+printing of source-level symbolic
+expressions,
+stack tracing,
+display of TLB entries,
+and even summaries of x86 hardware-defined descriptor tables.
+Unlike some symbolic debuggers,
+Simics is scriptable in Python,
+and it is possible to interpose user-written measurement
+frameworks to evaluate proposed hardware
+features~\cite{UW-GEMS,SimFlex,FeS2}.
+Performance is very reasonable (Simics can simulate certain
+cases, such as a CPU halted awaiting a clock interrupt,
+much faster than real time).
+
+% ability to quickly write small debugging-support snippets
+% in a few lines of Python (e.g., report the function name
+% every time a RET instruction is executed)
+
+%=== Limitations ===%
+\subsection{Limitations}
+
+At present the course accepts the following limitations
+or core assumptions.
+
+\begin{description}
+
+\item[x86] Students write code that runs on one specific
+hardware platform.
+We target x86 PCs because machines of this type are
+available from many manufacturers at many price points.
+This means that it is easy for students to obtain access
+to a machine (virtual or real) which will let them boot
+up their code, type commands, and observe the results.
+Also, unlike many System-on-a-Chip platforms,
+detailed hardware descriptions are readily available.
+% rather than being proprietary or "documented"
+% by labyrinthine Linux source code
+Finally, the platform is fairly stable---with only a
+little care, it is possible to write a compliant
+kernel which runs on anything newer than a 486DX cpu,
+i.e., almost any PC hardware which is likely to be in
+working condition.
+
+\item[32-bit] The course still targets the 32-bit
+subset of the x86 platform, for three reasons.
+First, while the small number of registers is
+frustrating for compiler writers,
+from the point of view of OS students,
+the existence of fewer registers means less work
+when it is necessary to verify that each register's
+value is correct.
+Likewise, the old-fashioned focus on the stack
+makes it easier to track function parameter values.
+Second, implementing virtual memory on x86-64 as
+opposed to x86-32 requires more code and more
+debugging time without imparting greater insight.
+Finally, since it is impossible for an x86 processor
+to be in 64-bit mode with virtual memory off,
+students would need to be coached through the
+32-bit--to--64-bit transition or would need
+to be given a template virtual-memory implementation.
+All in all, at present we are satisifed with
+the 32-bit choice.
+
+\item[memory model] At present we allow students
+to reason about concurrency based on a
+sequentially-consistent view of
+memory, which is consistent with
+existing x86 hardware~\cite{SewellSOZNM:x86tso-cacm10}
+but simplistic compared to most other
+architectures.
+Lecture material educates students about the
+general topic of out-of-order memory systems
+and fencing,
+but at present the hands-on experience does not
+reinforce that material.
+
+\end{description}
+
+%=== Outcomes ===%
+\subsection{Outcomes}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Pebbles}
 \label{sec:pebbles}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% primeval Unix model: fork()/exec()/wait()
+% Mach ``task/thread'' terminology/structure (as opposed to rfork()/clone())
+% keyboard input, screen output
+% thread creation
+% thread block/unblock
+% new_pages()/remove_pages()
+% swexn
+
+% swexn - what an auto-growing stack really means

### 2013-01-05 edupar-13/410.tex
React to Dave's work

@@
 
+\subsubsection{Stack Crawler}
 The first project is a ``stack crawler'':  invoked by
@@
 a combination of \texttt{printf()} and \texttt{gdb}).
+%% TODO I've been using "\x{printf()}" for code, an alias for \lstinline I
+%% adopted from the Rust paper template. We should be consistent because
+%% the font sizes are different (and maybe make \x{}'s font bigger too,
+%% what do you think?). -- bblum
 
+\subsubsection{Device Drivers and Game}
 The second project is a simple game, such as Hangman,
@@
 
+\subsubsection{User-space Thread Library}
 The third project is implementing a 1:1 thread library for
@@
 version of POSIX Pthreads.
+%% TODO Should be "PThreads"? -- bblum
 The primary goal of the project is providing students
@@
 abstractions.
+%% Good rhetoric (below). -- bblum
 One example of this is a semantic gap
@@
 The behavior of the reference kernel is specified
-via a twelve-page document.
+via a twelve-page document~\cite{kspec}.
 % the handout for the thread library project itself
@@
 
+\subsubsection{Pebbles Kernel}
 For the fourth project, two-student teams
@@
 they implemented in the second project.
+%% TODO Space permitting, maybe have a table of system calls?
+%% The 410 section is otherwise lacking in visual aids. -- bblum
 These ingredients are combined to yield
@@
 console I/O, and miscellaneous).
+%% TODO Justify by mentioning average lines of code. -- bblum
 For most students in the class, this is the
@@
 have produced.
+%% TODO Huh? I don't get how the sentence below is supposed to flow.
+%% Seems like it wants to elaborate on the concerns, but the next sentence
+%% is different. -- bblum
 Because the test suite and the grading criteria
@@
 
+\subsubsection{Pebbles Extension}
 Students who complete the kernel project on time
@@
 and writing an in-kernel debugger.
-Two recent, more-aggressive, projects have been
+Two recent, more aggressive projects have been
 adding paravirtualization so that their kernels
@@
 platform used in the class is
-Simics~\cite{Magnusson:2002:SFS:619072.621909},
+Simics~\cite{simics},
 originally developed by the Swedish Institute of
@@
 frameworks to evaluate proposed hardware
+%% TODO Fix citations. But also, maybe this mention isn't relevant and
+%% could be left out (if we're in space trouble?) -- bblum
 features~\cite{UW-GEMS,SimFlex,FeS2}.
@@
 much faster than real time).
+%% TODO: One huge advantage of simics over QEMU that we should mention is
+%% that QEMU only issues timer interrupts at basic block boundaries. -- bblum
 
@@
 
-\begin{description}
-
-\item[x86] Students write code that runs on one specific
+\subsubsection{x86}
+Students write code that runs on one specific
 hardware platform.
@@
 
-\item[32-bit] The course still targets the 32-bit
+\subsubsection{32-Bit} The course still targets the 32-bit
 subset of the x86 platform, for three reasons.
+%% TODO I'm not sure I believe this is an advantage. Instead of having
+%% to verify if each register is correct, students now have to verify
+%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
 First, while the small number of registers is
@@
 
-\item[memory model] At present we allow students
+\subsubsection{Memory Model} At present we allow students
 to reason about concurrency based on a
@@
 reinforce that material.
-
-\end{description}
 

### 2013-01-05 edupar-13/landslide.tex
React to Dave's work

@@
 
+% TODO: This paragraph can be removed if we're hurting for space
 We also note that, in contrast with stress testing, Landslide works best with very small test cases.
 While a more conventional stress test achieves coverage by exercising many system calls (in parallel and/or in sequence), Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
-Hence, if Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
-When working with students, we provided a suite of four test programs, each no longer than 10 lines of code.
+If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
+Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code.
 

### 2013-01-05 edupar-13/410.tex
410.tex: Added Pebbles kspec summary.

@@
 \subsubsection{x86}
+
 Students write code that runs on one specific
@@
 
-\subsubsection{32-Bit} The course still targets the 32-bit
+\subsubsection{32-Bit}
+
+The course still targets the 32-bit
 subset of the x86 platform, for three reasons.
@@
 
-\subsubsection{Memory Model} At present we allow students
+\subsubsection{Memory Model}
+
+At present we allow students
 to reason about concurrency based on a
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Pebbles}
+\section{The Pebbles kernel specification}
 \label{sec:pebbles}
@@
 
-% primeval Unix model: fork()/exec()/wait()
-% Mach ``task/thread'' terminology/structure (as opposed to rfork()/clone())
-% keyboard input, screen output
-% thread creation
-% thread block/unblock
-% new_pages()/remove_pages()
-% swexn
+The kernel that students implement is loosely based on the
+Unix model: \texttt{fork()}, \texttt{exec()}, and \texttt{wait()}.
+We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
+distinction between ``tasks,''
+which are resource containers,
+and threads,
+each of which executes within a single task.
+This partitioning is a different approach from that of
+Plan~9's \texttt{rfork()} or Linux's \texttt{clone()},
+which allow a continuum of sharing between
+identical threads and independent processes,
+at the cost of implementation complexity which we wish to avoid.
+A task's address space can be edited through two system
+calls that add and delete contiguous regions of memory.
+Students are expected to implement zero-fill-on-demand,
+but copy-on-write and demand-loading of executables are
+optional.
+Threads can print text to and read lines of text from
+the simple console that students implemented at
+the start of the semester.
 
-% swexn - what an auto-growing stack really means
+Within a task, threads are created using the
+\texttt{thread\_fork} system call, which closely
+mirrors the specification of \texttt{fork()}:
+the new thread begins with an exact copy of the
+register values of the old thread, except that
+the return-value register, \texttt{\%eax}, contains
+zero;
+in the old thread, the return value is the thread i.d.\
+of the newly created thread.
+This minimal formulation defers all policy decisions
+(e.g., stack size, location, and contents) to the
+user-space thread library.
+A thread may suspend its execution via the
+\texttt{deschedule()} system call,
+and a blocked thread can be reactivated via
+\texttt{make\_runnable()}.
+
+Each thread has the ability to register a handler
+for hardware exceptions (illegal instruction,
+division by zero, etc.)
+via the \texttt{swexn()} system call,
+which specifies the addresses of handler code
+and a handler stack.
+When a thread encounters an exception,
+the kernel reflects the exception to user space
+by
+de-registering the handler,
+pushing the thread's register state onto the stack,
+and invoking the handler code.
+The handler has the ability to re-register itself
+(or register some other handler),
+to run arbitrary code,
+and/or to atomically adopt a new set of register
+values.
+The \texttt{swexn()} facility is used to enable
+the thread library to detect when a managed
+thread has crashed
+and, more importantly,
+to enable user-space
+implementation of automatically-growing stack
+regions.
+Removing the auto-stack policy and mechanism from
+its traditional position in Unix-like kernels
+was done because it simplfies the kernel and
+``front-loads'' the policy question to an earlier
+part of the course,
+before students are facing the substantial pressure
+of the kernel project.

### 2013-01-06 edupar-13/410.tex
410.tex: Added a hopefully-visible box where "Outcomes" text should go.
Incorporate nwf's feedback

@@
 
+%% TODO: Nwf says "vary widely" does not read as intended. -- bblum
 The course learning objectives vary widely.
@@
 assignment design and grading encourage students to
-think about corner cases including resource exhaustion
+think about corner cases, including resource exhaustion,
 instead of being satisfied by ``the right basic idea''
@@
 and debugged as students see fit (generally using
-a combination of \texttt{printf()} and \texttt{gdb}).
+a combination of \x{printf()} and \x{gdb}).
 %% TODO I've been using "\x{printf()}" for code, an alias for \lstinline I
@@
 We provide students with selected C~library routines
-(\texttt{strcmp()}, \texttt{memmove()}, \texttt{printf()});
+(\x{strcmp()}, \x{memmove()}, \x{printf()});
 they write a device-driver library consisting of
@@
 version of POSIX Pthreads.
-%% TODO Should be "PThreads"? -- bblum
 The primary goal of the project is providing students
@@
 One example of this is a semantic gap
-between the library-level \texttt{thr\_create()}
-call, analogous to \texttt{pthread\_create()},
-and the \texttt{thread\_fork} system call provided
+between the library-level \x{thr_create()}
+call, analogous to \x{pthread_create()},
+and the \x{thread_fork} system call provided
 by the underlying kernel,
@@
 deciding how to bridge them.
-A similar semantic gap exists between
-\texttt{cond\_wait()} and the system call
+A similar deliberate semantic gap exists between
+\x{cond_wait()} and the system call
 which suspends execution of a thread.
@@
 and provided in binary form.
-%
+
+The behavior of the reference kernel is specified
+via a twelve-page document~\cite{kspec}.
+% the handout for the thread library project itself
+% is nineteen pages and includes not only a specification
+% of the thread-library primitives but also
+% a suggested implementation plan, grading guidelines,
+% optional design challenges, etc.
 In addition to providing a reliable execution
@@
 In order to aid debugging of concurrency problems,
-students can manually select among these
-interleaving patterns.
-The behavior of the reference kernel is specified
-via a twelve-page document~\cite{kspec}.
-% the handout for the thread library project itself
-% is nineteen pages and includes not only a specification
-% of the thread-library primitives but also
-% a suggested implementation plan, grading guidelines,
-% optional design challenges, etc.
+% students can manually select among these
+% interleaving patterns.
+the kernel offers a variety of (specification-conformant, yet deliberately undocumented) interleaving policies;
+the \x{misbehave()} system call is provided to select the current global policy.
 
@@
 
+\framebox{Some text should go here!}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Pebbles kernel specification}
+\section{The Pebbles Kernel Specification}
 \label{sec:pebbles}
@@
 The kernel that students implement is loosely based on the
-Unix model: \texttt{fork()}, \texttt{exec()}, and \texttt{wait()}.
+Unix model: \x{fork()}, \x{exec()}, and \x{wait()}.
 We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
@@
 This partitioning is a different approach from that of
-Plan~9's \texttt{rfork()} or Linux's \texttt{clone()},
+Plan~9's \x{rfork()} or Linux's \x{clone()},
 which allow a continuum of sharing between
@@
 Within a task, threads are created using the
-\texttt{thread\_fork} system call, which closely
-mirrors the specification of \texttt{fork()}:
+\x{thread_fork} system call, which closely
+mirrors the specification of \x{fork()}:
 the new thread begins with an exact copy of the
 register values of the old thread, except that
-the return-value register, \texttt{\%eax}, contains
+the return-value register, \x{\%eax}, contains
 zero;
@@
 A thread may suspend its execution via the
-\texttt{deschedule()} system call,
+\x{deschedule()} system call,
 and a blocked thread can be reactivated via
-\texttt{make\_runnable()}.
+\x{make_runnable()}.
 
@@
 division by zero, etc.)
-via the \texttt{swexn()} system call,
+via the \x{swexn()} system call,
 which specifies the addresses of handler code
@@
 values.
-The \texttt{swexn()} facility is used to enable
+The \x{swexn()} facility is used to enable
 the thread library to detect when a managed

### 2013-01-06 edupar-13/conclusion.tex
Incorporate nwf's feedback

@@
 currently a Ph.D.\ candidate at Johns Hopkins,
-increased the sanity of the build infrastructure.
+increased the sanity of the build infrastructure,
+designed the \x{swexn} system call, and helped critique this paper.
 The current reference kernel is named ``pathos''

### 2013-01-06 edupar-13/410.tex
Add syscall table

@@
 These ingredients are combined to yield
-twenty-four system calls
+twenty-five system calls
 (life-cycle, thread management, memory management,
-console I/O, and miscellaneous).
+console I/O, and miscellaneous),
+which are summarized in Figure~\ref{fig:syscalls}.
 %% TODO Justify by mentioning average lines of code. -- bblum
@@
 ``algorithm level'' and the register/bit-field level.
+
+\begin{figure}
+	\center \footnotesize
+	\begin{tabular}{|l|p{0.28\textwidth}|}
+		\hline
+		\bf System call name & \bf Summary \\
+		\hline
+		\multicolumn{2}{c}{\em Lifecycle management} \\
+		\hline
+		\x{fork} & Duplicates the invoking task, including all memory regions. \\
+		\x{thread_fork} & Creates a new thread in the current task.\\
+		\x{exec} & Replaces the program currently running in the invoking task with a new one specified. \\
+		\x{set_status} & Records the exit status of the current task. \\
+		\x{vanish} & Terminates execution of the calling thread. \\
+		\x{wait} & Blocks execution until another task terminates, and collects its exit status.\\
+		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
+		\hline
+		\multicolumn{2}{c}{\em Thread management} \\
+		\hline
+		\x{gettid} & Returns the ID of the invoking thread. \\
+		\x{yield} & Defers execution to a specified thread. \\
+		\x{deschedule} & Blocks execution of the invoking thread. \\
+		\x{make_runnable} & Wakes up another \x{deschedule}d thread. \\
+		\x{get_ticks} & Gets the number of timer ticks since bootup. \\
+		\x{sleep} & Blocks a thread for a given number of ticks. \\
+		\x{swexn} & Registers a given user-space function as a software exception handler.\\
+		\hline
+		\multicolumn{2}{c}{\em Memory management} \\
+		\hline
+		\x{new_pages} & Allocates a specified region of memory. \\
+		\x{remove_pages} & Deallocates same. \\
+		\hline
+		\multicolumn{2}{c}{\em Console I/O} \\
+		\hline
+		\x{getchar}* & Reads one character from keyboard input. \\
+		\x{readline} & Reads the next line from keyboard input. \\
+		\x{print} & Prints a given memory buffer to the console. \\
+		\x{set_term_color} & Sets the color for future console output. \\
+		\x{set_cursor_pos} & Sets the console cursor location. \\
+		\x{get_cursor_pos} & Retrieves the console cursor location. \\
+		\hline
+		\multicolumn{2}{c}{\em Miscellaneous} \\
+		\hline
+		\x{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system''. \\
+		\x{halt} & Ceases execution of the operating system. \\
+		\x{misbehave}* & Selects among several behaviors for thread interleaving in certain scenarios. \\
+		\hline
+	\end{tabular}
+	\caption{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
+	\label{fig:syscalls}
+\end{figure}
 

### 2013-01-06 edupar-13/410.tex
Talk about vanish/vanish troubles

@@
 which are summarized in Figure~\ref{fig:syscalls}.
+
 %% TODO Justify by mentioning average lines of code. -- bblum
@@
 there are many cross-cutting concerns.
+%
+%% TODO: Dave, is it OK to say the following sentence? -- bblum
+Widely regarded as the most difficult concurrency problem in the project
+is that of coordinating a parent and a child task that ``simultaneously''
+invoke \x{vanish()}: the parent task must reassign ownership of the child to
+\x{init}, the root process, while the child task must check its parent's state
+for any \x{wait()}ing threads which need to be woken up.
+Many solutions, due to design constraints imposed by other parts of the kernel,
+are prone to data races or deadlocks without careful thought.
+%
 Because the students are responsible for ensuring

### 2013-01-06 edupar-13/intro.tex
write a ton more
Talk about vanish/vanish troubles

@@
 Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in userspace on top of a Pebbles reference kernel in a two-week project, and subsequently, in a 6-week project, write their own kernel implementation.
+During these projects students face very demanding synchronization problems, including how to implement locking primitives and how to coordinate a parent and a child process that are exiting simultaneously.
 Finally, in another two-week project, students implement a small extension to their Pebbles kernel.
@@
 The rest of this paper is organized as follows.
-Section~\ref{sec:projects} gives an overview of 15-410's five programming projects.
+Section~\ref{sec:curriculum} gives an overview of 15-410's five programming projects.
 Section~\ref{sec:pebbles} discusses the Pebbles specification in detail.
 Section~\ref{sec:landslide} describes Landslide's testing mechanisms.
+Section~\ref{sec:eval} describes how we worked with students to evaluate Landslide.
 Section~\ref{sec:future} concludes, discussing strategies for making Landslide more accessible to struggling students, and how Landslide might be applied to related operating systems projects at other universities.

### 2013-01-06 edupar-13/landslide.tex
Demote eval to subsection + fix a citation

@@
 In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Evaluation}
+
+We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
+
+We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
+We configured Landslide to search for five complicated already-known race conditions.
+In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
+Using decision points only on calls to \x{mutex_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.
+
+In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
+Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two races and two deterministic errors.
+These bugs manifested as infinite loops, a kernel panic, and a use-after-free.
+
+% TODO: decide whether to put table 7.2 from the thesis here. i think probably not?

### 2013-01-07 edupar-13/landslide.tex
comment out ignore_sym paragraphs

@@
 We provide an interface for adding new decision points to refine the state space, described in Section~\ref{sec:decision}.
-In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction (DPOR) algorithm~\cite{dpor} to identify and prune redundant thread interleavings.
-While the specifics of the algorithm are beyond the educational scope of this work, we also provide an interface for helping Landslide further reduce the state space using domain knowledge about the kernel's implementation.
-We discuss this in Section~\ref{sec:focusing}.
+In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction (DPOR) algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the educational scope of this work.
+We also provide an interface for helping Landslide further reduce the state space by considering decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.
 
@@
 
-Additionally, the student may configure Landslide's state space reduction to ignore certain memory locations.
-DPOR's analysis requires a memory independence relation between thread transitions; the more transitions are independent from each other, the more reduction can be achieved.
-Unfortunately, in the kernel environment, every thread switch will execute through a common path that modifies shared scheduler data structures.
-Unchecked, this would cause every transition to conflict with each other and result in no reduction in execution time from DPOR.
-Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested; for example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
-
-We provide an interface with which the user can configure Landslide to ignore such unrelated conflicts to more efficiently test components of the kernel they care about.
-In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
+% Additionally, the student may configure Landslide's state space reduction to ignore certain memory locations.
+% DPOR's analysis requires a memory independence relation between thread transitions; the more transitions are independent from each other, the more reduction can be achieved.
+% Unfortunately, in the kernel environment, every thread switch will execute through a common path that modifies shared scheduler data structures.
+% Unchecked, this would cause every transition to conflict with each other and result in no reduction in execution time from DPOR.
+% Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested; for example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
+%
+% We provide an interface with which the user can configure Landslide to ignore such unrelated conflicts to more efficiently test components of the kernel they care about.
+% In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
 

### 2013-01-07 edupar-13/landslide.tex
fix \x{} size formatting

@@
 \begin{figure}[t]
+\small
 \begin{lstlisting}

### 2013-01-08 edupar-13/count.sh
Add count.sh

@@ -1,0 +1,46 @@
+#!/bin/bash
+
+count_lines() {
+	if [ -d "`echo $1 | sed 's/libthread.*/libthread/'`" ]; then # c.c.c.c
+		SLOC=`sloccount $1`
+		# filter out c++ groups
+		if echo $SLOC | grep '^cpp:' >/dev/null 2>/dev/null; then
+			echo -n "????,????" 
+		else
+			C=`  echo "$SLOC" | grep '^ansic:' | sed 's/ansic: *//' | sed 's/ *(.*)//'`
+			ASM=`echo "$SLOC" | grep '^asm:'   | sed 's/asm: *//'   | sed 's/ *(.*)//'`
+			if [ -z "$C" -o -z "$ASM" ]; then
+					echo -n "(huh? couldn't parse)"
+			else
+					echo -n "$C,$ASM"
+			fi
+		fi
+	else
+		echo -n "???,???"
+	fi
+}
+
+count_group() {
+	if [ -d "$1/p3extra" ]; then
+		P3DIR=p3extra
+	else
+		P3DIR=p3
+	fi
+	echo -n "$1," | sed 's/.*\///'
+	count_lines "$1/p2/user/libthread $1/p2/user/autostack.c"
+	echo -n ","
+	count_lines "$1/$P3DIR/kern"
+	echo "";
+}
+
+while [ ! -z "$1" ]; do
+		if [ ! -d "$1/user/group-00" ]; then
+				echo "$1 doesn't look like a 410 semester directory"
+				exit 1
+		fi
+		for i in `ls -d $1/user/group-*`; do
+				echo -n "`basename $1`,"
+				count_group $i;
+		done
+		shift
+done

### 2013-01-08 edupar-13/landslide.tex
Explain 'race condition' terminology

@@
 
+Note that our use of the term ``race condition'' refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
+In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs merely upon suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
 In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.

### 2013-01-08 edupar-13/count.sh

@@ -10,7 +10,7 @@ count_lines() {
 			C=`  echo "$SLOC" | grep '^ansic:' | sed 's/ansic: *//' | sed 's/ *(.*)//'`
 			ASM=`echo "$SLOC" | grep '^asm:'   | sed 's/asm: *//'   | sed 's/ *(.*)//'`
 			if [ -z "$C" -o -z "$ASM" ]; then
-					echo -n "(huh? couldn't parse)"
+					echo -n "(huh, couldn't parse??)"
 			else
 					echo -n "$C,$ASM"
 			fi

### 2013-01-08 edupar-13/landslide.tex
Move outcomes to after landslide

@@
 % In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Evaluation}
-
-We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
-
-We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
-We configured Landslide to search for five complicated already-known race conditions.
-In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
-Using decision points only on calls to \x{mutex_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.
-
-In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
-Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two races and two deterministic errors.
-These bugs manifested as infinite loops, a kernel panic, and a use-after-free.
-
-% TODO: decide whether to put table 7.2 from the thesis here. i think probably not?

### 2013-01-08 edupar-13/410.tex
Move outcomes to after landslide
write about grading

@@
 
+%=== Grading ===%
+\subsection{Grading}
+
+Pebbles kernels are graded in three ways: passing the ``hurdle'', running an automated test suite, and undergoing manual inspection by course staff.
+
+\subsubsection{Hurdle}
+%% TODO: Dave, write about the criteria (and what it means, i.e., "getting a shell up and running, getting processes to exit")
+The hurdle is a set of criteria that Pebbles implementations must meet to be considered sastisfactory
+(TODO: what are they?).
+Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
+Students who pass the hurdle are permitted to turn in their kernels and move on to the final, shorter, project; otherwise, they must spend the remaining two weeks continuing to work towards the hurdle.
+The hurdle criteria constitute 50\% of the kernel project's grade.
+
+\subsubsection{Automated tests}
+We run submitted kernels in an automated testing harness, called ``Fritz''.
+On each kernel, Fritz runs a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
+Passing Fritz's tests constitutes 30\% of the kernel project's grade.
+
+Unfortunately, due to the nondeterministic nature of many common bugs in student code, Fritz is unable to verify any submission's correctness with complete certainty.
+(The hit-or-miss nature of Fritz's bugfinding motivated our work on Landslide.)
+
+\subsubsection{Manual inspection}
+Finally, both to make up for Fritz's incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
+The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.
+After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
+Staff discretionary points constitute 20\% of the kernel project's grade.
+
 %=== Limitations ===%
@@
 reinforce that material.
-
-%=== Outcomes ===%
-\subsection{Outcomes}
-
-\framebox{Some text should go here!}
 

### 2013-01-08 edupar-13/410.tex
polish some text

@@
 discuss our use of Wind River Simics as the
-the basic simulation/debugging environment,
+the basic simulation/debugging environment
+and our approaches to grading,
 and close by mentioning current limitations in the scope
@@
 \subsubsection{Automated tests}
-We run submitted kernels in an automated testing harness, called ``Fritz''.
-On each kernel, Fritz runs a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
-Passing Fritz's tests constitutes 30\% of the kernel project's grade.
+We run submitted kernels in an automated testing harness.
+The test suite is a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
+Performance on these tests constitutes 30\% of the kernel project's grade.
 
-Unfortunately, due to the nondeterministic nature of many common bugs in student code, Fritz is unable to verify any submission's correctness with complete certainty.
-(The hit-or-miss nature of Fritz's bugfinding motivated our work on Landslide.)
+Of course, due to the nondeterministic nature of many common bugs in student code, these tests are unable to verify a submission's correctness with complete certainty.
+(This hit-or-miss nature motivated our work on Landslide.)
 
 \subsubsection{Manual inspection}
-Finally, both to make up for Fritz's incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
+Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
 The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.

### 2013-01-08 edupar-13/landslide.tex
polish some text

@@
 We provide an interface for adding new decision points to refine the state space, described in Section~\ref{sec:decision}.
-In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction (DPOR) algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the educational scope of this work.
+In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the educational scope of this work.
 We also provide an interface for helping Landslide further reduce the state space by considering decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.

### 2013-01-08 edupar-13/410.tex
Rearrange sections and table to make the 'Pebbles' section more respectable.

@@
 they implemented in the second project.
-%% TODO Space permitting, maybe have a table of system calls?
-%% The 410 section is otherwise lacking in visual aids. -- bblum
 These ingredients are combined to yield
@@
 console I/O, and miscellaneous),
-which are summarized in Figure~\ref{fig:syscalls}.
+which we discuss further in Section~\ref{sec:pebbles}.
 
@@
 
-\begin{figure}
-	\center \footnotesize
-	\begin{tabular}{|l|p{0.28\textwidth}|}
-		\hline
-		\bf System call name & \bf Summary \\
-		\hline
-		\multicolumn{2}{c}{\em Lifecycle management} \\
-		\hline
-		\x{fork} & Duplicates the invoking task, including all memory regions. \\
-		\x{thread_fork} & Creates a new thread in the current task.\\
-		\x{exec} & Replaces the program currently running in the invoking task with a new one specified. \\
-		\x{set_status} & Records the exit status of the current task. \\
-		\x{vanish} & Terminates execution of the calling thread. \\
-		\x{wait} & Blocks execution until another task terminates, and collects its exit status.\\
-		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
-		\hline
-		\multicolumn{2}{c}{\em Thread management} \\
-		\hline
-		\x{gettid} & Returns the ID of the invoking thread. \\
-		\x{yield} & Defers execution to a specified thread. \\
-		\x{deschedule} & Blocks execution of the invoking thread. \\
-		\x{make_runnable} & Wakes up another \x{deschedule}d thread. \\
-		\x{get_ticks} & Gets the number of timer ticks since bootup. \\
-		\x{sleep} & Blocks a thread for a given number of ticks. \\
-		\x{swexn} & Registers a given user-space function as a software exception handler.\\
-		\hline
-		\multicolumn{2}{c}{\em Memory management} \\
-		\hline
-		\x{new_pages} & Allocates a specified region of memory. \\
-		\x{remove_pages} & Deallocates same. \\
-		\hline
-		\multicolumn{2}{c}{\em Console I/O} \\
-		\hline
-		\x{getchar}* & Reads one character from keyboard input. \\
-		\x{readline} & Reads the next line from keyboard input. \\
-		\x{print} & Prints a given memory buffer to the console. \\
-		\x{set_term_color} & Sets the color for future console output. \\
-		\x{set_cursor_pos} & Sets the console cursor location. \\
-		\x{get_cursor_pos} & Retrieves the console cursor location. \\
-		\hline
-		\multicolumn{2}{c}{\em Miscellaneous} \\
-		\hline
-		\x{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system''. \\
-		\x{halt} & Ceases execution of the operating system. \\
-		\x{misbehave}* & Selects among several behaviors for thread interleaving in certain scenarios. \\
-		\hline
-	\end{tabular}
-	\caption{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
-	\label{fig:syscalls}
-\end{figure}
-
 \subsubsection{Pebbles Extension}
@@
 addding multi-processor support to their single-processor kernels.
-Meanwhile, students who were unable to complete
-the kernel project on time receive additional time
-to finish up and improve their code quality---this
-requirement is based on our belief that important
-learning experiences are obtained by solving the
-hard problems,
-which turn up at the end of the implementation.
 
@@
 % every time a RET instruction is executed)
-
-%=== Grading ===%
-\subsection{Grading}
-
-Pebbles kernels are graded in three ways: passing the ``hurdle'', running an automated test suite, and undergoing manual inspection by course staff.
-
-\subsubsection{Hurdle}
-%% TODO: Dave, write about the criteria (and what it means, i.e., "getting a shell up and running, getting processes to exit")
-The hurdle is a set of criteria that Pebbles implementations must meet to be considered sastisfactory
-(TODO: what are they?).
-Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
-Students who pass the hurdle are permitted to turn in their kernels and move on to the final, shorter, project; otherwise, they must spend the remaining two weeks continuing to work towards the hurdle.
-The hurdle criteria constitute 50\% of the kernel project's grade.
-
-\subsubsection{Automated tests}
-We run submitted kernels in an automated testing harness.
-The test suite is a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
-Performance on these tests constitutes 30\% of the kernel project's grade.
-
-Of course, due to the nondeterministic nature of many common bugs in student code, these tests are unable to verify a submission's correctness with complete certainty.
-(This hit-or-miss nature motivated our work on Landslide.)
-
-\subsubsection{Manual inspection}
-Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
-The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.
-After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
-Staff discretionary points constitute 20\% of the kernel project's grade.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Pebbles Kernel Specification}
+\section{The Pebbles Kernel}
 \label{sec:pebbles}
@@
 
-The kernel that students implement is loosely based on the
+\begin{figure}
+	\center \footnotesize
+	\begin{tabular}{|l|p{0.28\textwidth}|}
+		\hline
+		\bf System call name & \bf Summary \\
+		\hline
+		\multicolumn{2}{c}{\em Lifecycle management} \\
+		\hline
+		\x{fork} & Duplicates the invoking task, including all memory regions. \\
+		\x{thread_fork} & Creates a new thread in the current task.\\
+		\x{exec} & Replaces the program currently running in the invoking task with a new one specified. \\
+		\x{set_status} & Records the exit status of the current task. \\
+		\x{vanish} & Terminates execution of the calling thread. \\
+		\x{wait} & Blocks execution until another task terminates, and collects its exit status.\\
+		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
+		\hline
+		\multicolumn{2}{c}{\em Thread management} \\
+		\hline
+		\x{gettid} & Returns the ID of the invoking thread. \\
+		\x{yield} & Defers execution to a specified thread. \\
+		\x{deschedule} & Blocks execution of the invoking thread. \\
+		\x{make_runnable} & Wakes up another \x{deschedule}d thread. \\
+		\x{get_ticks} & Gets the number of timer ticks since bootup. \\
+		\x{sleep} & Blocks a thread for a given number of ticks. \\
+		\x{swexn} & Registers a given user-space function as a software exception handler.\\
+		\hline
+		\multicolumn{2}{c}{\em Memory management} \\
+		\hline
+		\x{new_pages} & Allocates a specified region of memory. \\
+		\x{remove_pages} & Deallocates same. \\
+		\hline
+		\multicolumn{2}{c}{\em Console I/O} \\
+		\hline
+		\x{getchar}* & Reads one character from keyboard input. \\
+		\x{readline} & Reads the next line from keyboard input. \\
+		\x{print} & Prints a given memory buffer to the console. \\
+		\x{set_term_color} & Sets the color for future console output. \\
+		\x{set_cursor_pos} & Sets the console cursor location. \\
+		\x{get_cursor_pos} & Retrieves the console cursor location. \\
+		\hline
+		\multicolumn{2}{c}{\em Miscellaneous} \\
+		\hline
+		\x{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system''. \\
+		\x{halt} & Ceases execution of the operating system. \\
+		\x{misbehave}* & Selects among several behaviors for thread interleaving in certain scenarios. \\
+		\hline
+	\end{tabular}
+	\caption{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
+	\label{fig:syscalls}
+\end{figure}
+
+\subsection{Specification}
+
+The kernel that students implement consists of twenty-five
+system calls, which are summarized in Figure~\ref{fig:syscalls}.
+%
+The process lifecycle is loosely based on the
 Unix model: \x{fork()}, \x{exec()}, and \x{wait()}.
@@
 of the kernel project.
+
+%=== Grading ===%
+\subsection{Grading}
+
+Pebbles kernels are graded in three ways: passing the ``hurdle'', running an automated test suite, and undergoing manual inspection by course staff.
+
+\subsubsection{Hurdle}
+%% TODO: Dave, write about the criteria (and what it means, i.e., "getting a shell up and running, getting processes to exit")
+The hurdle is a set of criteria that Pebbles implementations must meet to be considered sastisfactory
+(TODO: what are they?).
+Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
+Students who pass the hurdle turn in their kernels and move on to the shorter ``Pebbles Extension'' project.
+Meanwhile, students who were unable to meet the
+hurdle criteria on time receive additional time
+to finish up and improve their code quality---this
+requirement is based on our belief that important
+learning experiences are obtained by solving the
+hard problems,
+which turn up at the end of the kernel implementation.
+The hurdle criteria constitute 50\% of the kernel project's grade.
+
+\subsubsection{Automated tests}
+We run submitted kernels in an automated testing harness.
+The test suite is a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
+Of course, due to the nondeterministic nature of many common bugs in student code, these tests can only provide a rough estimate of a submission's correctness, rather than verifying it with certainty.
+Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.
+(This hit-or-miss nature motivated our work on Landslide.)
+
+\subsubsection{Manual inspection}
+Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
+The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.
+After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
+Staff discretionary points constitute 20\% of the kernel project's grade.

### 2013-01-08 edupar-13/intro.tex
Demote eval to subsection + fix a citation
Rearrange sections and table to make the 'Pebbles' section more respectable.

@@
 Section~\ref{sec:curriculum} gives an overview of 15-410's five programming projects.
-Section~\ref{sec:pebbles} discusses the Pebbles specification in detail.
-Section~\ref{sec:landslide} describes Landslide's testing mechanisms.
-Section~\ref{sec:eval} describes how we worked with students to evaluate Landslide.
+Section~\ref{sec:pebbles} discusses the Pebbles specification and grading process in detail.
+Section~\ref{sec:landslide} describes Landslide's testing mechanisms and how we worked with students to evaluate Landslide.
 Section~\ref{sec:future} concludes, discussing strategies for making Landslide more accessible to struggling students, and how Landslide might be applied to related operating systems projects at other universities.

### 2013-01-08 edupar-13/landslide.tex
clarify landslide-friendly test characteristics

@@
 If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
-Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code.
+Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code, but carefully written to bring about system call interactions known to be troublesome.
 

### 2013-01-09 edupar-13/intro.tex
Use garth's intro feedback & write abstract

@@
 Managing parallel computation, whether on thousand-node clusters or on one single-core machine, is often taught by resorting to powerful abstractions that simplify the hardest problems.
-Programming patterns such as divide-and-conquer and fork/join parallelism enable developers to reason about parallel algorithms at a high, abstract level.
-With the advent of massively multicore hardware and distributed systems, however, there is also a need for reasoning about parallelism at lower levels: those of operating system kernels and concurrent systems applications, which are built on (or even below) the abstractions of threads and processes.
-Low-level systems programmers must often reason about parallel code without the convenience of high-level simplifications, and hence are often susceptible to notoriously difficult concurrency errors.
+Programming patterns such as SPMD, divide-and-conquer, and fork/join parallelism enable developers to reason about parallel algorithms at a high, abstract level.
+The user-level libraries and distributed systems that provide these abstractions, in turn, rely on operating system kernels to simplify the hardware and communication mechanisms on which they are built.
+% garth's
+
+With the advent of massively multicore hardware,
+kernels themselves have increasing hardware concurrency and complexity. % garth's
+% there is also a need for reasoning about parallelism at lower levels: those of operating system kernels and concurrent systems applications, which are built on (or even below) the abstractions of threads and processes.
+% ben's old text
+Low-level systems programmers must reason about concurrent code without the convenience of high-level simplifications, and hence are often susceptible to notoriously difficult concurrency errors.
 
 In this paper, we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
-In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively challenging kernel project, in which they implement threads and processes in code that will boot on real hardware.
-Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in userspace on top of a Pebbles reference kernel in a two-week project, and subsequently, in a 6-week project, write their own kernel implementation.
-During these projects students face very demanding synchronization problems, including how to implement locking primitives and how to coordinate a parent and a child process that are exiting simultaneously.
+In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, in which they implement threads and processes to boot on real hardware.
+Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in userspace on top of a Pebbles reference kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
+During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate a parent and child process that exit simultaneously.
 Finally, in another two-week project, students implement a small extension to their Pebbles kernel.
-This project changes every semester, and in the past has included extending the kernel to be able to hypervise other kernels and to be able to run in SMP systems.
+This project changes every semester, and in the past has included extending the kernel to be able to hypervise other kernels and to be able to run on SMP systems.
 
-As we ask our students to write more and more difficult concurrent code in the same semester, we want to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time during the projects.
-However, we also want to avoid compromising the complexity and difficulty of the project's concurrency challenges. % how to say... "because it's the most valuable thing we do"
+As we ask our students to write more and more difficult concurrent code in one semester, we want to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
+However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project.
 Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads into execution sequences that expose concurrency bugs.

### 2013-01-09 edupar-13/conclusion.tex
conclusion.tex: acknowledge rpearl, jwise, David A. Wheeler (SLOCCount).

--- 2013-01-05 edupar-13/conclusion.tex
@@
 currently a Ph.D.\ candidate at Johns Hopkins,
-increased the sanity of the build infrastructure.
+increased the sanity of the build infrastructure,
+designed the \x{swexn()} system call, and helped critique this paper.
 The current reference kernel is named ``pathos''
@@
 and Elly Fong-Jones, currently at Google.
+The SMP infrastructure code was contributed by
+Ryan Pearl, currently at Mozilla.
+Joshua Wise, currently at NVIDIA,
+proposed and led several kernel-extension projects,
+including the hypervisor project.
 Over the past decade the course has benefitted
@@
 Wind River.
+Code-size numbers were generated using David A.\ Wheeler's
+``SLOCCount.''

### 2013-01-09 edupar-13/intro.tex
intro.tex: Clarification pass.

@@
 % ben's old text
-Low-level systems programmers must reason about concurrent code without the convenience of high-level simplifications, and hence are often susceptible to notoriously difficult concurrency errors.
+Low-level systems programmers must reason about concurrent code without the convenience of high-level simplifications, and hence often encounter notoriously difficult concurrency errors.
 
-In this paper, we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
-In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, in which they implement threads and processes to boot on real hardware.
-Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in userspace on top of a Pebbles reference kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
-During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate a parent and child process that exit simultaneously.
+In this paper we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
+In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, which runs on real hardware, in which they implement threads and processes.
+Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in user space on top of a Pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
+During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
 Finally, in another two-week project, students implement a small extension to their Pebbles kernel.
-This project changes every semester, and in the past has included extending the kernel to be able to hypervise other kernels and to be able to run on SMP systems.
+This project changes every semester, and in the past has included extending their kernel serve as a hypervisor, or to support symmetric multiprocessing.
 
-As we ask our students to write more and more difficult concurrent code in one semester, we want to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
-However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project.
-Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads into execution sequences that expose concurrency bugs.
+As we ask our students to write more and more difficult concurrent code in one semester, we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
+However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.
+Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads to take execution paths that expose concurrency bugs.
 We present our experience sharing a prototype of Landslide with volunteering students who used it to find bugs in their own code, and discuss the possibility for integrating it as a main component of the 15-410 curriculum.
@@
 Section~\ref{sec:curriculum} gives an overview of 15-410's five programming projects.
-Section~\ref{sec:pebbles} discusses the Pebbles specification and grading process in detail.
+Section~\ref{sec:pebbles} discusses the Pebbles kernel project and how it is graded.
 Section~\ref{sec:landslide} describes Landslide's testing mechanisms and how we worked with students to evaluate Landslide.

### 2013-01-09 edupar-13/landslide.tex
garth's feedback on landslide.tex

@@
 
-Landslide is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
-It follows in the footsteps of related tools such as CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter}, but is the first tool we know of to apply systematic exploration in a kernel environment.
+Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
+It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
+It follows in the footsteps of related tools such as CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter}, but is the first tool we know of to apply systematic exploration in a kernel development environment.
+
+Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
+Simics can load Landslide while booting a student's kernel, and calls into Landslide once per simulated instruction and once per simulated memory access.
+Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.
 In this section we give an overview of Landslide's design and interface, and point out the annotations students need to provide to make Landslide work with their own code.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Scope}
-
-\paragraph{Testing environment}
-Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
-Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and once per memory access of the simulated execution.
-Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.
-
-\paragraph{Limitations}
-Landslide assumes that timer interrupts are the only nondeterministic events that influence a kernel's concurrent execution.
-While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
-Landslide's model is also restricted to uniprocessor execution.
-Supporting multiprocessor environments and device driver testing is left to future work.
-% this used to say ", and production kernels"
-
-% TODO: This paragraph can be removed if we're hurting for space
-We also note that, in contrast with stress testing, Landslide works best with very small test cases.
-While a more conventional stress test achieves coverage by exercising many system calls (in parallel and/or in sequence), Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
-If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
-Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code, but carefully written to bring about system call interactions known to be troublesome.
 
@@
 This code is a simplified implementation of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID.
-However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run and invoke \x{exit}, causing \x{child} to become a dangling pointer.
+However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run long enough to invoke \x{exit}, causing \x{child} to become a dangling pointer.
 
@@
 
-Systematic exploration, Landslide's mechanism for testing different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings in a system.
+Systematic exploration, Landslide's mechanism for testing many different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings of a specific test.
 These interleavings are defined around {\em decision points}, which intuitively indicate points during execution at which a preemption could cause different behaviour to arise.
-Searching with few decision points results in coarser-grained interleavings, shorter test execution, and less likelihood of finding bugs; searching with more results in the opposite.
+Searching with few decision points results in coarser-grained interleavings, shorter test execution, and less likelihood of finding unexpected bugs; searching with more results in the opposite.
+The primary advantage of systematic testing is the ability to force execution sequences that might be very rare in simple repeated testing.
 
@@
 First, at each decision point, it can force an arbitary thread to run in place of the current one by injecting timer interrupts to force the kernel to context switch.
-Second, whenever the kernel finishes executing a particular interleaving of the test case, it reverts the system to an earlier state to explore additional interleavings using Simics's backtracking mechanism.
+Second, whenever the kernel finishes executing a particular interleaving of the test case, it reverts the system to an earlier state to explore additional interleavings using Simics's reverse execution.
 
-In our example bug, the necessary decision point for finding the bug is at line~\ref{line:hax}.
-With correct annotations (discussed in Section~\ref{sec:instrument} and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point.
-Landslide also automatically identifies a minimal set of decision points at voluntary reschedules between threads (e.g.
-\x{yield}).
+In Figure~\ref{fig:threadfork}, the necessary decision point for finding the bug is at line~\ref{line:hax}.
+With correct instrumentation (discussed in Section~\ref{sec:instrument} and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point when the child thread becomes runnable.
+Landslide also automatically identifies decision points at voluntary reschedules between threads (e.g.
+\x{yield()}).
 Together these decision points define an execution tree which exposes the bug, depicted in Figure~\ref{fig:tree}.
@@
 Of course, additional decision points are necessary to expose more subtle race conditions.
-We provide an interface for adding new decision points to refine the state space, described in Section~\ref{sec:decision}.
-In order to cope with the exponentially-sized nature of state spaces with many decision points, Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the educational scope of this work.
+We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
+In order to cope with the exponentially-growing nature of state spaces with many decision points,
+Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
 We also provide an interface for helping Landslide further reduce the state space by considering decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.
@@
 If a kernel has concurrency errors, they will show up in some branches, but not others.
-This tree shows the bug from Figure~\ref{fig:threadfork}, with decision points from the parent thread shaded.
+This tree shows the bug from Figure~\ref{fig:threadfork}, with operations from the parent thread shaded.
 }
@@
 Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.
-
 Landslide can optionally also check heuristically for infinite loops and livelock.
-This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
-We have never found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
+This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), we assume the kernel is stuck.
+We have not found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
 
 When Landslide identifies a bug, it outputs a {\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.
-This trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that we caused to preempt it.
+The trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces to show when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
 With this trace, the student can better understand the concurrent execution that led to the bug.
@@
 Note that our use of the term ``race condition'' refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
-In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs merely upon suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
-In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.
-However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that a race exists.
+In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs at suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
+In general,
+% with false-negative bug detection,
+a kernel might contain a buggy behaviour that Landslide could overlook (i.e., a false negative).
+However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that an error exists.
 
@@
 Users annotate their kernels to inform Landslide of certain important concurrency events during execution.
-We provide a set of annotation functions, called \x{tell_landslide}, for this purpose.
+We provide a set of annotation functions, named with the prefix \x{tell_landslide}, for this purpose.
 The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
-Figure~\ref{fig:annotation} shows the code from Figure~\ref{fig:threadfork} with \x{tell_landslide} annotations.
+Figure~\ref{fig:annotation} shows annotated code corresponding to Figure~\ref{fig:threadfork}.
 
-There is also a configuration file, \x{config.landslide}, in which the user must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which userspace test program Landslide should invoke.
+There is also a configuration file, \x{config.landslide}, in which the student must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which userspace test program Landslide should invoke.
 Finally, there are two short (nominally two-line) functions used within Landslide itself that the user must implement.
@@
 
-Using only the decision points automatically identified on voluntary reschedules will result in coarse-grained interleavings likely to overlook bugs.
+Using only decision points that Landslide automatically identifies on voluntary reschedules will result in coarse-grained interleavings likely to overlook bugs.
 We provide an extra annotation
-for the user to add more decision points for a finer-grained search, called
+for students to add more decision points for a finer-grained search, called
 \x{tell_landslide_decide()}.
-We recommend using it in concurrency primitives, such as at the start of \x{mutex_lock()} and at the end of \x{mutex_unlock()}.
+We recommend inserting it into concurrency primitives, such as at the start of \x{mutex_lock()} and at the end of \x{mutex_unlock()}.
 
@@
 
-The strategy we described above may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated system calls.
-We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively limit them.
-First, the student writes ``\x{DECISION_INFO_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
-Then, they write ``\x{within_function} \x{foo}'' to whitelist function \x{foo}, and ``\x{without_function} \x{bar}'' to blacklist function \x{bar}.
+The strategy we describe above may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated and/or already-trusted system calls.
+We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively eliminate them.
+% First, the student writes ``\x{DECISION_INFO_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
+% In this file, students can write ``\x{within_function} \x{foo}'' to whitelist function \x{foo}, and ``\x{without_function} \x{bar}'' to blacklist function \x{bar}.
+In this file, students can use the ``\x{within_function}'' and ``\x{without_function}'' directives to require that given functions are or are not on the call stack when decision points are identified.
+% make Landslide ignore any decision points that don't arise from the execution of specific functions.
 For example, if testing thread death and reaping, the student should write \x{within_function exit} and \x{within_function} \x{wait}, and also
@@
 % In \x{config.landslide}, the user writes ``\x{ignore_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Limitations}
+
+Landslide currently assumes that timer interrupts are the only nondeterministic events that influence a kernel's concurrent execution.
+While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
+Landslide's model is also currently restricted to uniprocessor execution.
+Supporting multiprocessor environments and device driver testing is left to future work.
+% this used to say ", and production kernels"
+
+% TODO: This paragraph can be removed if we're hurting for space
+We also note that, in contrast with conventional stress testing, in which the test program exercises many system calls (in parallel and/or in sequence), Landslide works best with very small test cases.
+Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
+If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
+Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code, but carefully written to bring about system call interactions known to be troublesome.

### 2013-01-09 edupar-13/410.tex
410.tex: local fix-ups to intro, project material (not grading).

@@
 Distributed Systems,
-and Computer Networks.
+and Computer Networks).
 The prerequisite course is 15-213,
@@
 
-%% TODO: Nwf says "vary widely" does not read as intended. -- bblum
-The course learning objectives vary widely.
+The course has many learning objectives,
+ranging from acquiring detailed factual knowledge about
+hardware features
+through practicing advanced cognitive processes
+such as open-ended design.
 %
@@
 analyzing and writing lock-based multi-threaded code and
-thread-synchronization objects, including detecting and documenting race conditions and deadlock,
-including
-the relationship between interrupts and other forms of concurrency.
+thread-synchronization objects.
+They practice detecting and documenting deadlock and race conditions,
+including both thread/thread concurrency and thread/interrupt concurrency.
 
@@
 \subsubsection{Stack Crawler}
-The first project is a ``stack crawler'':  invoked by
+The first project is a ``stack crawler'':  when invoked by
 a client program, it displays the program's stack
@@
 We provide students with selected C~library routines
-(\x{strcmp()}, \x{memmove()}, \x{printf()});
+(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()});
 they write a device-driver library consisting of
@@
 and booted via GRUB.
-The floppy-disk image can be copied to a real floppy disk
+If the floppy-disk image is copied to a real floppy disk
 or embedded into an ``El Torito'' bootable compact disc image,
-in which case it can be booted on standard PC hardware.
+it can be booted on standard PC hardware.
 However, students most often boot the floppy image
@@
 threads.
-Another goal is encouraging students to carefully
+A second goal is encouraging students to carefully
 consider how ingredients are assembled into
@@
 student code according to a variety of
-interleaving patterns.
+interleaving policies.
 In order to aid debugging of concurrency problems,
-% students can manually select among these
-% interleaving patterns.
-the kernel offers a variety of (specification-conformant, yet deliberately undocumented) interleaving policies;
-the \x{misbehave()} system call is provided to select the current global policy.
+the \x{misbehave()} system call enables
+students to manually select an interleaving policy.
 
@@
 For the fourth project, two-student teams
-implement a kernel which complies to the
+implement a kernel which complies with the
 same specification as the reference kernel
@@
 have produced.
-%% TODO Huh? I don't get how the sentence below is supposed to flow.
-%% Seems like it wants to elaborate on the concerns, but the next sentence
-%% is different. -- bblum
 Because the test suite and the grading criteria
@@
 there are many cross-cutting concerns.
-%
-%% TODO: Dave, is it OK to say the following sentence? -- bblum
-Widely regarded as the most difficult concurrency problem in the project
-is that of coordinating a parent and a child task that ``simultaneously''
-invoke \x{vanish()}: the parent task must reassign ownership of the child to
-\x{init}, the root process, while the child task must check its parent's state
-for any \x{wait()}ing threads which need to be woken up.
-Many solutions, due to design constraints imposed by other parts of the kernel,
-are prone to data races or deadlocks without careful thought.
 %
@@
 ``algorithm level'' and the register/bit-field level.
+
+Widely regarded as the most difficult concurrency problem in the project
+is that of coordinating a parent and a child task that ``simultaneously''
+invoke \x{vanish()}:
+when a task completes,
+live children and exited zombies must be handed off
+to the task's parent or to the system's \x{init} process,
+at a time when the task's parent may itself be
+exiting;
+meanwhile, threads in tasks that receive new children
+may need to be awakened from the \x{wait()} system call.
+Due to design constraints imposed by other parts of the kernel specification,
+solutions which are not carefully designed
+are prone to data races or deadlocks.
 
@@
 little care, it is possible to write a compliant
-kernel which runs on anything newer than a 486DX cpu,
+kernel which runs on anything newer than a 486DX processor,
 i.e., almost any PC hardware which is likely to be in
@@
 
-\subsubsection{32-Bit}
+\subsubsection{32-bit}
 
@@
 value is correct.
-Likewise, the old-fashioned focus on the stack
+Likewise, the calling convention's old-fashioned focus on the stack
 makes it easier to track function parameter values.
@@
 
-\subsubsection{Memory Model}
+\subsubsection{Memory model}
 
@@
 reinforce that material.
+
+\subsubsection{Reduced breadth of implementation}
+
+One cost of programming to the complexities of
+real hardware is that
+within the scope of a single semester,
+students do not have the opportunity to
+implement common features of many kernels,
+such as a file system, interprocess communication,
+a real-time scheduler,
+or a network protocol stack.
 
@@
 zero;
-in the old thread, the return value is the thread i.d.\
+in the old thread, the return value is the thread identifier
 of the newly created thread.
@@
 de-registering the handler,
-pushing the thread's register state onto the stack,
+pushing the thread's register state onto the handler stack,
 and invoking the handler code.
@@
 was done because it simplfies the kernel and
-``front-loads'' the policy question to an earlier
+``front-loads'' policy questions to an earlier
 part of the course,

### 2013-01-09 edupar-13/410.tex
410.tex: reorganized "Grading" and did initial work.

@@
 
-Pebbles kernels are graded in three ways: passing the ``hurdle'', running an automated test suite, and undergoing manual inspection by course staff.
+Pebbles kernels are graded through a combination
+of three methods:
+performance on a test suite,
+compliance with ``hurdle criteria,''
+and detailed code review by a member of the course staff.
+
+\subsubsection{Automated tests}
+
+While developing their code, students uncover both bugs
+and design misunderstandings by running test code we
+provide.
+%Some tests are simple, exercising one feature or condition
+%(e.g., allocating and freeing a large amount of memory),
+%while other tests exercise a mix of kernel features
+%simultaneously (creating and destroying processes,
+%some of which do I/O while others handle or are killed
+%by various exceptions) over an extended period of time.
+The test suite is a number of user-space programs which test
+functionality ranging from the basic process lifecycle, to robustness
+against corner-case security vulnerabilities, to long-term stability
+under large stress tests.
+%
+When students submit their kernels,
+we use an automated testing harness to
+run a superset of the public test suite.
+%
+Of course, due to the nondeterministic nature of many common bugs in student code, these tests can only provide a rough estimate of a submission's correctness, rather than verifying it with certainty.
+Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.
+(This hit-or-miss nature motivated our work on Landslide.)
 
 \subsubsection{Hurdle}
+
 %% TODO: Dave, write about the criteria (and what it means, i.e., "getting a shell up and running, getting processes to exit")
@@
 
-\subsubsection{Automated tests}
-We run submitted kernels in an automated testing harness.
-The test suite is a number of userspace programs which test functionality ranging from the basic process lifecycle, to robustness against corner-case security vulnerabilities, to long-term stability under large stress tests.
-Of course, due to the nondeterministic nature of many common bugs in student code, these tests can only provide a rough estimate of a submission's correctness, rather than verifying it with certainty.
-Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.
-(This hit-or-miss nature motivated our work on Landslide.)
-
 \subsubsection{Manual inspection}

### 2013-01-09 edupar-13/intro.tex
intro.tex: typo reported by Ben

@@
 Finally, in another two-week project, students implement a small extension to their Pebbles kernel.
-This project changes every semester, and in the past has included extending their kernel serve as a hypervisor, or to support symmetric multiprocessing.
+This project changes every semester, and in the past has included extending their kernel to serve as a hypervisor, or to support symmetric multiprocessing.
 

### 2013-01-10 edupar-13/conclusion.tex
conclusion.tex: place-holder for acknowledgment of Ben's funding.
Add garth's conclusion without reading it

@@
 
-we did stuff
+With the radical changes in computer architecture induced by the switch from speed through faster cycle time to speed through more cores, and more complex memory system semantics, operating systems kernels are experiencing rapid change and effective education of kernel
+programmers becomes much more labor intensive, especially in terms of concurrency error debugging, or much less detailed and hardware accurate.
+
+Carnegie Mellon University's undergraduate operating systems course has a tradition of requiring students to tackle highly hardware accurate kernel design and development.
+Rather than compromise the low level detail that our students overcome in a semester, as we include more hardware concurrency and the resulting complexity, we are experimenting with tools that simplify some of the most time consuming changes challenges our students face.
+
+Landslide is a tool that integrates into the Simics x86 hardware simulator and drives repeated execution of a given test with a unique scheduling of context switches on each execution, known as systematic testing, ensuring that extremely rare execution sequencings can be
+experienced by the test in a short period of time relative to simple repeated execution of tests, known as stress testing.
+
+Landslide automates a minimal set of context switch points, known as decision points, given annotations that expose invocation of fundamental operating system kernel operations, and offers student users additional annotations to add, and refine, decision points to focus
+exploration on user guided aspects of their kernels.
+
+Experiments with Landslide on single-core multi-threaded kernels, both completed in prior course offerings and during an ongoing course offering show promising results.
+Students were able to find unknown and unexpected bugs after a couple of hours of tool learning and kernel annotating.
+
+With this promising result, and our recent experience with a quad core symmetric multi-processor kernel extension project, we believe that systematic testing of kernel code can be useful for helping our students accomplish more complex concurrent hardware in the same time and effort.
+
+Landslide needs to be extended to cope with the multiple scheduler queues that occur in multi-core hardware, however, much of this can be managed with slight variations of the annotations student provide, naming the core a thread will runs on, for example.
+
+To fully realize this benefit Landslide usability should be improved.
+For example, our class notes communicate execution sequencing with a tabular representation of thread execution; Landslide's presentation of a buggy sequence could be tailored to match our teaching methodology.
+
+Tool usability is greatly improved by lowering the start up effort required of each student.
+Much of the annotation our students must add to their kernel could be eliminated if the course project specification required prescribed names for prescribed functions, so Landslide could easily identify what the student kernel is doing.
+There is a downside to doing too much of this, as the design of kernel interfaces is an important part of the course project, and prescribed templates could constrain the design space too much.
+
+Pebbles and Landslide have lots of future challenges as we introduce memory systems with complex consistency rules exposed to the kernel programmer, or decide to explore performance issues with data non-locality and sharing patterns.
+
+Landslide utility is also not restricted to Pebbles.
+It could be extracted from Simics and embedded in a virtual machine platform, to begin the transition to supporting production operating system kernel development.
 
@@
 ``SLOCCount.''
+
+\framebox{Ben's Navy funding should go here!}

### 2013-01-10 edupar-13/outcomes.tex
Move outcomes to after landslide
polish some text
garth's feedback on eval + army (not navy!) grant

--- 2013-01-02 edupar-13/eval.tex
@@
-\section{Evaluation}
-\label{sec:eval}
+\section{Outcomes}
 
-We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
+\subsection{Teaching with Pebbles}
+
+\framebox{Some text should go here!}
+
+\subsection{Evaluating Landslide}
+
+We evaluated Landslide in two ways: first, by instrumenting two prior student kernels to measure the exploration time needed to find different races, and second, by meeting with current students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
 
 We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
-We configured Landslide to search for five complicated already-known race conditions.
+We configured Landslide to search for five complicated well-known race conditions.
 In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
-Using decision points only on calls to \x{mutex_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.
+Using additional decision points only on calls to \x{mutex_lock}, Landslide found each of the six bugs in 11 to 57 seconds on a 2.6 GHz Xeon server, executing between 1 and 377 distinct interleavings for each.
 
 In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
-Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two races and two deterministic errors.
+Of the four groups who finished the required part, all four found previously unknown bugs in their kernels: two races and two deterministic errors.
 These bugs manifested as infinite loops, a kernel panic, and a use-after-free.
-
-% TODO: decide whether to put table 7.2 from the thesis here. i think probably not?
+Despite wishing the instrumentation were easier, the students reported that they found working with Landslide rewarding.

### 2013-01-10 edupar-13/410.tex
410.tex: rampaged through Grading section.

@@
 %
-Of course, due to the nondeterministic nature of many common bugs in student code, these tests can only provide a rough estimate of a submission's correctness, rather than verifying it with certainty.
+Of course, due to the nondeterministic nature of many common bugs in student code,
+these tests can provide only a rough estimate of a submission's correctness,
+rather than verifying it with certainty.
+(This hit-or-miss nature motivated our work on Landslide.)
 Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.
-(This hit-or-miss nature motivated our work on Landslide.)
 
-\subsubsection{Hurdle}
+\subsubsection{``Hurdle'' completion criteria}
 
-%% TODO: Dave, write about the criteria (and what it means, i.e., "getting a shell up and running, getting processes to exit")
-The hurdle is a set of criteria that Pebbles implementations must meet to be considered sastisfactory
-(TODO: what are they?).
-Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
-Students who pass the hurdle turn in their kernels and move on to the shorter ``Pebbles Extension'' project.
+Because of the ambitious learning objectives,
+we employ a very detailed notion of ``completeness''
+for the kernel project.
+The ``hurdle model'' is based on the following observations:
+
+\begin{itemize}
+
+\item We believe that students learn more from bugs
+that are harder to find.
+In extreme cases, a software component may do
+``almost everything right,'' but ``the last bug'' may
+demonstrate a deep misunderstanding which requires a re-design.
+
+\item Because concurrency bugs are triggered rarely,
+finding them requires a kernel capable of running
+for a long period without crashing.
+
+\item Robust code often requires a different structure
+than a ``usually works'' solution.
+The most straightforward way for an exiting task to
+convey its exit status to its parent is for it to
+\x{malloc()} an exit-status blob and enqueue it for
+later processing by the parent.
+However, this design suffers from a ``terminal irony'':
+a program wishing to exit,
+which will free up kernel memory,
+is unable to do so if kernel memory has been exhausted.
+
+\end{itemize}
+
+For these reasons, we believe much of the learning
+associated with the kernel project necessarily occurs
+at the end of the experience,
+so it is important for students to finish the process.
+% [suppressed: bad to build P4 on a shaky P3]
+We believe it would be particularly detrimental to
+assign ``partial credit'' to a large body of non-working
+code.
+Instead, we require student kernels to pass a
+specified percentage of each segment of the test
+suite to be considered complete.
+Furthermore,
+because the test suit cannot detect certain
+structural shortcuts (e.g., papering over a concurrency
+problem by disabling interrupts during \textit{all}
+kernel execution),
+we require students to certify, in writing, that
+their code possesses certain properties.
+
+A ``complete'' kernel can boot up and run the shell,
+can pass 80\% of the basic system-call sanity test,
+can pass 70\% of the ``solidity'' tests (which
+invoke system calls in illegal or non-trivial ways),
+and can pass two of the three long-running multi-feature tests.
+In addition,
+a complete kernel is certified by the authors to
+be free of ``concurrency shortcuts,''
+to handle resource exhaustion,
+and to avoid certain distasteful x86-specific code structures.
+%
+% The hurdle is a set of criteria that Pebbles implementations must meet to be considered sastisfactory
+% (TODO: what are they?).
+%Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
+Students who pass the completeness ``hurdle'' submit their kernels
+and can move on to the ``Pebbles Extension'' project.
 Meanwhile, students who were unable to meet the
 hurdle criteria on time receive additional time
-to finish up and improve their code quality---this
-requirement is based on our belief that important
-learning experiences are obtained by solving the
-hard problems,
-which turn up at the end of the kernel implementation.
+to finish up and improve their code quality.
+%---this
+%requirement is based on our belief that important
+%learning experiences are obtained by solving the
+%hard problems,
+%which turn up at the end of the kernel implementation.
 The hurdle criteria constitute 50\% of the kernel project's grade.
 
-\subsubsection{Manual inspection}
+\subsubsection{Code review and interview}
+
 Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.

### 2013-01-10 edupar-13/landslide.tex
landslide.tex: wrote "Landslide in one paragraph" paragraph.

@@
 
-Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
+Conceptually,
+Landslide is implemented as follows.
+Students annotate their kernel code so Landslide knows
+which kernel thread is currently running.
+After Landslide has run one kernel thread for some time,
+it feeds artificial clock interrupts to the kernel
+to force a switch to the thread Landslide wishes to execute.
+As the executing thread completes each instruction,
+Landslide applies multiple bug-detection predicates to
+the kernel's state to detect
+illegal heap references,
+deadlock, livelock, infinite loops, and kernel panics.
+If a given test scenario runs to completion according
+to one pattern of thread switches,
+Landslide rewinds the kernel's state to some previous time
+and executes forward according to a different thread interleaving.
+In theory,
+by forcing a thread switch after every non-scheduler instruction,
+Landslide could apply its bug-detection predicates to every reachable
+execution state.
+Because this would require prohibitive time to complete,
+in practice Landslide uses a variety of techniques to
+thread-switch less often and to avoid repeating
+bug-equivalent execution paths.
+%Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
 Simics can load Landslide while booting a student's kernel, and calls into Landslide once per simulated instruction and once per simulated memory access.
-Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.
+%Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.
 In this section we give an overview of Landslide's design and interface, and point out the annotations students need to provide to make Landslide work with their own code.
@@
 
-Figure~\ref{fig:threadfork} shows a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section.
+Figure~\ref{fig:threadfork} shows code containing a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section.
 This code is a simplified implementation of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID.
@@
 These interleavings are defined around {\em decision points}, which intuitively indicate points during execution at which a preemption could cause different behaviour to arise.
-Searching with few decision points results in coarser-grained interleavings, shorter test execution, and less likelihood of finding unexpected bugs; searching with more results in the opposite.
+Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
 The primary advantage of systematic testing is the ability to force execution sequences that might be very rare in simple repeated testing.

### 2013-01-10 edupar-13/outcomes.tex
outcomes.tex: trivial word-smithing in Landslide outcomes text.

@@
 
-We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
+In the first phase, we instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
 We configured Landslide to search for five complicated well-known race conditions.
@@
 
-In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
+In the user-study phase, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
 Of the four groups who finished the required part, all four found previously unknown bugs in their kernels: two races and two deterministic errors.

### 2013-01-10 edupar-13/410.tex
410.tex: compressed/fuzzed "terminal irony" description.

@@
 than a ``usually works'' solution.
-The most straightforward way for an exiting task to
-convey its exit status to its parent is for it to
-\x{malloc()} an exit-status blob and enqueue it for
-later processing by the parent.
-However, this design suffers from a ``terminal irony'':
+It is easy to write task-exit code which
+needs to allocate memory in order to notify its
+parent that it has exited.
+However, such designs suffer from ``terminal irony'':
 a program wishing to exit,
-which will free up kernel memory,
+which would free up kernel memory,
 is unable to do so if kernel memory has been exhausted.

### 2013-01-14 edupar-13/conclusion.tex
garth's feedback on eval + army (not navy!) grant
rewrite 1st half of conclusion

--- 2013-01-09 edupar-13/conclusion.tex
@@
-\section{Future Work and Conclusion}
+\section{Conclusions and Future Work}
 \label{sec:future}
 
-we did stuff
+Carnegie Mellon University's operating systems course offers a series of concurrency-focused projects to advanced undergraduates focused on low-level software development.
+These projects culminate in the development of a Pebbles kernel, a fully preemptible UNIX-like kernel which includes process lifecycle management, virtual memory, and some device drivers, during the course's six-week-long fourth project.
+% Dave: You could add here a brief blurb about what students accomplish, e.g. LOC, boot on real hardware, whatever. Or it would be fine without.
+In order to keep up with emerging trends in hardware architecture, especially the increasing focus on massively multi-core processing, we seek to include more hardware concurrency and the resulting complexity in the course's curriculum.
+
+% With the radical changes in computer architecture induced by the switch from speed through faster cycle time to speed through more cores, and more complex memory system semantics, operating systems kernels are experiencing rapid change and effective education of kernel programmers becomes much more labor intensive, especially in terms of concurrency error debugging, or much less detailed and hardware accurate.
+
+% Carnegie Mellon University's undergraduate operating systems course has a tradition of requiring students to tackle highly hardware accurate kernel design and development.
+% Rather than compromise the low level detail that our students overcome in a semester, as we include more hardware concurrency and the resulting complexity, we are experimenting with tools that simplify some of the most time consuming changes challenges our students face.
+
+Rather than making room for more material by reducing the projects' design complexity, which we feel would compromise the learning experience, we are experimenting with concurrency debugging tools to let students spend less time searching for errors in their code.
+We have presented Landslide, a tool that integrates with the Simics simulator students already use and provides a more principled approach to bug-finding than conventional stress testing.
+
+Due to the open-ended design problems students solve in their kernels, Landslide imposes a moderate annotation burden in order to instrument a kernel for systematic testing.
+However, by the very nature of having enough spare time to volunteer, the selection of students we worked with to use Landslide was necessarily biased towards the ``top of the class''.
+Though our experiments with these students show promising results, we need to reduce Landslide's annotation burden as much as possible to make it useful for struggling students as well.
+
+% some things we can automate -- inspired by automatic malloc instrumentation -- but some things require changing the spec
+
+\subsubsection{Automatic instrumentation}
+TODO
+
+\subsubsection{Mandating API names}
+For example, rather than providing \x{tell_landslide_thread_runnable()}, we could mandate that the scheduler's code path to making a thread runnable include a function with a specified name.
+
+\subsubsection{Mandating design}
+TODO
+
+% tighten
+% Landslide is a tool that integrates into the Simics x86 hardware simulator and drives repeated execution of a given test with a unique scheduling of context switches on each execution, known as systematic testing, ensuring that extremely rare execution sequencings can be experienced by the test in a short period of time relative to simple repeated execution of tests, known as stress testing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% rephrase
+% Landslide automates a minimal set of context switch points, known as decision points, given annotations that expose invocation of fundamental operating system kernel operations, and offers student users additional annotations to add, and refine, decision points to focus exploration on user-guided aspects of their kernels.
+
+% Experiments with Landslide on single-core multi-threaded kernels during a prior course offering show promising results. % rephrase to say "students from a prior semester"
+% Students were able to find unknown and unexpected bugs after a couple of hours of tool learning and kernel annotating.
+
+% tighten
+% With this promising result, and our recent experience with a multi-processor kernel extension project,
+% good
+% we believe systematic testing can be useful for helping our students learn more about concurrent programming with the same amount of time and effort.
+
+% Add: "However, ... top students vs struggling students" transition sentence
+% Replace following two paragraphs with a new one
+
+Landslide needs to be extended to cope with the multiple scheduler queues that occur in multi-core hardware, however, much of this can be managed with slight variations of the annotations student provide, naming the core a thread will runs on, for example.
+
+To fully realize this benefit Landslide usability should be improved.
+For example, our class notes communicate execution sequencing with a tabular representation of thread execution; Landslide's presentation of a buggy sequence could be tailored to match our teaching methodology.
+
+% this should all turn into a bulleted list of improvement areas
+Tool usability is greatly improved by lowering the start up effort required of each student.
+Much of the annotation our students must add to their kernel could be eliminated if the course project specification required prescribed names for prescribed functions, so Landslide could easily identify what the student kernel is doing.
+There is a downside to doing too much of this, as the design of kernel interfaces is an important part of the course project, and prescribed templates could constrain the design space too much.
+
+% condense below two into a single sentence
+
+% Pebbles and Landslide have lots of future challenges as we introduce memory systems with complex consistency rules exposed to the kernel programmer, or decide to explore performance issues with data non-locality and sharing patterns.
+
+% Landslide utility is also not restricted to Pebbles.
+% It could be extracted from Simics and embedded in a virtual machine platform, to begin the transition to supporting production operating system kernel development.
 
@@
 ``SLOCCount.''
+Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.

### 2013-01-15 edupar-13/landslide.tex
landslide.tex: I'll buy a paragraph break for $5.
shorten and polish landslide section

@@
 Conceptually,
-Landslide is implemented as follows.
-Students annotate their kernel code so Landslide knows
+Landslide is designed as follows.
+Students annotate their code so Landslide knows
 which kernel thread is currently running.
-After Landslide has run one kernel thread for some time,
-it feeds artificial clock interrupts to the kernel
-to force a switch to the thread Landslide wishes to execute.
-As the executing thread completes each instruction,
-Landslide applies multiple bug-detection predicates to
+After one kernel thread has run for some time,
+Landslide triggers artificial clock interrupts
+to force a new thread to interleave with the current one.
+When a test program finishes execution according
+to one pattern of thread switches,
+Landslide rewinds the kernel's state
+and resumes the test according to a different thread interleaving.
+After each instruction,
+Landslide applies several bug-detection predicates to
 the kernel's state to detect
-illegal heap references,
-deadlock, livelock, infinite loops, and kernel panics.
-If a given test scenario runs to completion according
-to one pattern of thread switches,
-Landslide rewinds the kernel's state to some previous time
-and executes forward according to a different thread interleaving.
+illegal heap accesses,
+deadlock, infinite loops, and panics.
+\shortversion{}{
 In theory,
@@
 bug-equivalent execution paths.
+}
 %Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
@@
 %Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.
+
 In this section we give an overview of Landslide's design and interface, and point out the annotations students need to provide to make Landslide work with their own code.
@@
 
-Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.
+Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other heap-related errors such as double-free and memory leaks), and
+\shortversion{
+deadlocks, and also includes a heuristic check for infinite loops and livelock.
+}{
+deadlocks.
 Landslide can optionally also check heuristically for infinite loops and livelock.
@@
 We have not found this technique to erroneously report false positives, although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
+}
 
@@
 
-Note that our use of the term ``race condition'' refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
+Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
 In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs at suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
+\shortversion{}{
 In general,
@@
 However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that an error exists.
+}
 
@@
 We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively eliminate them.
+Suppose a student were testing thread death and reaping, and wanted decision points to appear in \x{wait} and \x{vanish} but not if they're also associated with virtual memory operations.
+They should write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
+The \x{within_function} directive requires that at least one of the specified functions shall be on the call stack when decision points are identified, and \x{without_function} requires the opposite.
+
 % First, the student writes ``\x{DECISION_INFO_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
 % In this file, students can write ``\x{within_function} \x{foo}'' to whitelist function \x{foo}, and ``\x{without_function} \x{bar}'' to blacklist function \x{bar}.
-In this file, students can use the ``\x{within_function}'' and ``\x{without_function}'' directives to require that given functions are or are not on the call stack when decision points are identified.
+% In this file, students can use the ``\x{within_function}'' and ``\x{without_function}'' directives to require that given functions are or are not on the call stack when decision points are identified.
 % make Landslide ignore any decision points that don't arise from the execution of specific functions.
-For example, if testing thread death and reaping, the student should write \x{within_function exit} and \x{within_function} \x{wait}, and also
-(assuming they don't care about the virtual memory operations associated with \x{exit})
-\x{without_function} \x{destroy_address_space}.
+% For example, if testing thread death and reaping, the student should write \x{within_function exit} and \x{within_function} \x{wait}, and also
+% (assuming they don't care about the virtual memory operations associated with \x{exit})
+% \x{without_function} \x{destroy_address_space}.
 
@@
 
-% TODO: This paragraph can be removed if we're hurting for space
-We also note that, in contrast with conventional stress testing, in which the test program exercises many system calls (in parallel and/or in sequence), Landslide works best with very small test cases.
+In contrast with conventional stress testing, in which the test program exercises many system calls (in parallel and/or in sequence), Landslide works best with very small test cases.
 Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.

### 2013-01-15 edupar-13/intro.tex
Write related work

@@
 
+In comparison with other educational kernels,
+% such as Pintos~\cite{pintos} and Xv6~\cite{xv6},
+Pebbles allows students to design and implement their systems ``from the ground up''.
+At MIT, students study Xv6~\cite{xv6}, a kernel for multiprocessor x86 machines,
+and build their own in a series of labs.
+At Stanford, students build kernels following the Pintos architecture~\cite{pintos}.
+Both courses supply students with starter code, in which thread lifecycle, stack management, and synchronization primitives are already implemented.
+In 15-410, we give starter code only for the machine boot sequence and for dynamic memory allocation, leaving students to design threading and scheduling primitives from scratch.
+
 As we ask our students to write more and more difficult concurrent code in one semester, we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.

### 2013-01-15 edupar-13/landslide.tex
remove unnecessary citations

@@
 It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
-It follows in the footsteps of related tools such as CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter}, but is the first tool we know of to apply systematic exploration in a kernel development environment.
+It follows in the footsteps of related tools such as
+CHESS~\cite{chess} and dBug~\cite{dbug-ssv}, but is the first tool we know of to apply systematic exploration in a kernel development environment.
 
@@
 In order to cope with the exponentially-growing nature of state spaces with many decision points,
+\shortversion{
+we also provide an interface for making Landslide consider
+}{
 Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
-We also provide an interface for helping Landslide further reduce the state space by considering decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.
+We also provide an interface for helping Landslide further reduce the state space by considering
+}
+decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.
 

### 2013-01-15 edupar-13/410.tex
incorporate changes suggested in meeting

@@
 Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
-The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.
+The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for style, architectural issues, race conditions, and preemptibility.
 After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.

### 2013-01-15 edupar-13/conclusion.tex
incorporate changes suggested in meeting

@@
 currently a Ph.D.\ candidate at Johns Hopkins,
-increased the sanity of the build infrastructure,
-designed the \x{swexn()} system call, and helped critique this paper.
+% increased the sanity of the build infrastructure,
+designed the \x{swexn()} system call and helped critique this paper.
 The current reference kernel is named ``pathos''
@@
 Over the past decade the course has benefitted
-greatly from free educational Simics licenses,
-donated initially by Virtutech and recently by
-Wind River.
+greatly from free educational Simics licenses
+donated by Virtutech/Wind River.
 Code-size numbers were generated using David A.\ Wheeler's

### 2013-01-15 edupar-13/intro.tex
remove unnecessary citations
incorporate changes suggested in meeting

--- 2013-01-09 edupar-13/intro.tex
@@
 During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
-Finally, in another two-week project, students implement a small extension to their Pebbles kernel.
-This project changes every semester, and in the past has included extending their kernel serve as a hypervisor, or to support symmetric multiprocessing.
+Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing.
+% in the past has included extending their kernel to serve as a hypervisor, or to support symmetric multiprocessing.
 
-As we ask our students to write more and more difficult concurrent code in one semester, we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
+In comparison with other educational kernel projects,
+% such as Pintos~\cite{pintos} and Xv6~\cite{xv6},
+Pebbles focuses more on ground-up design than on teaching a broad feature-set.
+% Pebbles allows students to design and implement their systems ``from the ground up.''
+At Stanford, students build kernels following the Pintos architecture~\cite{pintos}, and at MIT, following the Xv6/JOS architecture~\cite{xv6}.
+In both courses, student kernels support multiprocessor execution;
+however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
+In 15-410, we give starter code only for the machine boot sequence and for a standard library including dynamic memory allocation, leaving students to design threading and scheduling primitives from scratch.
+
+% As we ask our students to write more and more difficult concurrent code in one semester,
+As we plan to incorporate SMP into 15-410's curriculum, expanding the challenges our students face in one semester,
+we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
 However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.

### 2013-01-15 edupar-13/landslide.tex
incorporate changes suggested in meeting

@@
 It follows in the footsteps of related tools such as
-CHESS~\cite{chess} and dBug~\cite{dbug-ssv}, but is the first tool we know of to apply systematic exploration in a kernel development environment.
+\shortversion{
+CHESS~\cite{chess},
+}{
+CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},
+}
+but is the first tool we know of to apply systematic exploration in a kernel development environment.
 
@@
 We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively eliminate them.
-Suppose a student were testing thread death and reaping, and wanted decision points to appear in \x{wait} and \x{vanish} but not if they're also associated with virtual memory operations.
+If a student were testing thread death and reaping, they might want decision points to appear in \x{wait} and \x{vanish} but not if unrelated virtual memory operations are also in progress.
 They should write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
@@
 
-In contrast with conventional stress testing, in which the test program exercises many system calls (in parallel and/or in sequence), Landslide works best with very small test cases.
+In contrast with conventional stress testing, in which the test program exercises many system calls, Landslide works best with very small test cases.
 Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.

### 2013-01-15 edupar-13/conclusion.tex
Finish conclusion

@@
 
-% some things we can automate -- inspired by automatic malloc instrumentation -- but some things require changing the spec
+We believe some annotations can be automatically detected;
+for example, rather than requiring students to indicate if their kernel has a mechanism for disabling preemption apart from disabling interrupts, Landslide could detect when a timer interrupt is received but doesn't induce a context switch.
+Other annotations, such as \x{tell_landslide_thread_runnable()}, could be replaced by mandating that the kernel's code paths in such cases include a function with a predetermined name.
+Landslide already automatically instruments the memory allocator by detecting function names in the starter code; similar could be done for thread lifecycle and context switch events if their APIs were specified.
+However, prescribing such templates has the downside of constrain the project's design space.
+As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.
 
-\subsubsection{Automatic instrumentation}
-TODO
-
-\subsubsection{Mandating API names}
-For example, rather than providing \x{tell_landslide_thread_runnable()}, we could mandate that the scheduler's code path to making a thread runnable include a function with a specified name.
-
-\subsubsection{Mandating design}
-TODO
+\shortversion{}{
+Additionally, various improvements in Landslide's interface could increase its accessibility. 
+For example, our class notes communicate execution sequencing with a tabular representation of thread execution; Landslide's presentation of a buggy sequence could be tailored to match our teaching methodology.
+Landslide could also be extended with a mechanism for minimizing execution traces after a buggy one is found, to more accurately reflect which context switches are most relevant to the bug at hand.
+}
 
@@
 
-Landslide needs to be extended to cope with the multiple scheduler queues that occur in multi-core hardware, however, much of this can be managed with slight variations of the annotations student provide, naming the core a thread will runs on, for example.
+% Landslide needs to be extended to cope with the multiple scheduler queues that occur in multi-core hardware, however, much of this can be managed with slight variations of the annotations student provide, naming the core a thread will runs on, for example.
 
-To fully realize this benefit Landslide usability should be improved.
-For example, our class notes communicate execution sequencing with a tabular representation of thread execution; Landslide's presentation of a buggy sequence could be tailored to match our teaching methodology.
+% To fully realize this benefit Landslide usability should be improved.
 
 % this should all turn into a bulleted list of improvement areas
-Tool usability is greatly improved by lowering the start up effort required of each student.
-Much of the annotation our students must add to their kernel could be eliminated if the course project specification required prescribed names for prescribed functions, so Landslide could easily identify what the student kernel is doing.
-There is a downside to doing too much of this, as the design of kernel interfaces is an important part of the course project, and prescribed templates could constrain the design space too much.
+% Tool usability is greatly improved by lowering the start up effort required of each student.
+% Much of the annotation our students must add to their kernel could be eliminated if the course project specification required prescribed names for prescribed functions, so Landslide could easily identify what the student kernel is doing.
+% There is a downside to doing too much of this, as the design of kernel interfaces is an important part of the course project, and prescribed templates could constrain the design space too much.
 

### 2013-01-16 edupar-13/410.tex
Add SMP4 paragraph and clean up various stuff

@@
 outline the programming projects,
-discuss our use of Wind River Simics as the
+discuss our use of Wind River Simics~\cite{simics} as the
 the basic simulation/debugging environment
@@
 between the library-level \x{thr_create()}
-call, analogous to \x{pthread_create()},
+call (analogous to \x{pthread_create()})
 and the \x{thread_fork} system call provided
@@
 writing a sound card driver,
+a filesystem,
 hibernation (suspend to disk),
 kernel profiling,
-and writing an in-kernel debugger.
-Two recent, more aggressive projects have been
+and an in-kernel debugger.
+Two recent, more aggressive, projects have been
 adding paravirtualization so that their kernels
@@
 addding multi-processor support to their single-processor kernels.
+
+In the multi-processor project, students extended their kernels to schedule multiple threads at once on up to eight cores.
+We provided base code for determining the system's number of CPUs, for booting the application processors (APs), and for sending inter-processor interrupts (IPIs).
+Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
+to implement moving threads between processors,
+and
+to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
+To allow struggling groups to make progress on the project without solving every synchronization challenge, we allowed the use of a ``big kernel lock'', and rewarded groups according to how little they needed to use it.
+Submissions were also graded according to
+correctness of the scheduler and spinlocks,
+stability on a suite of stress tests,
+and the use of IPIs for efficient processor coordination.
 
@@
 The process lifecycle is loosely based on the
-Unix model: \x{fork()}, \x{exec()}, and \x{wait()}.
+Unix model: \x{fork()}, \x{exec()}, \x{wait()},
+and \x{vanish()} (the Pebbles name for \x{exit()}).
 We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
@@
 This partitioning is a different approach from that of
-Plan~9's \x{rfork()} or Linux's \x{clone()},
+Plan~9's \x{rfork()}~\cite{plan9} or Linux's \x{clone()},
 which allow a continuum of sharing between
@@
 Furthermore,
-because the test suit cannot detect certain
+because the test suite cannot detect certain
 structural shortcuts (e.g., papering over a concurrency

### 2013-01-16 edupar-13/conclusion.tex
typo
Add SMP4 paragraph and clean up various stuff

@@
 Landslide already automatically instruments the memory allocator by detecting function names in the starter code; similar could be done for thread lifecycle and context switch events if their APIs were specified.
-However, prescribing such templates has the downside of constrain the project's design space.
+However, prescribing such templates has the downside of constraining the project's design space.
 As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.
@@
 The SMP infrastructure code was contributed by
-Ryan Pearl, currently at Mozilla.
+Ryan Pearl, currently at Dropbox.
 Joshua Wise, currently at NVIDIA,

### 2013-01-19 edupar-13/outcomes.tex
outcomes.tex: Lines-of-code table for P2, P3, P3extra
outcomes.tex: Ben says: SLOC table must go! P4 text added, 8.5 pp. -> 8.75 codesizes.png: screen shot of the SLOC table for future reference.

@@
 \section{Outcomes}
+
+In this section we briefly describe selected
+outcomes of the Spring 2012 offering of 15-410.
+We begin by discussing the results of the
+pre-Landslide educational experience and then
+report on the results of applying Landslide to
+kernels previously written by members of the course staff
+and kernels written that semester by students.
 
@@
 
-\framebox{Some text should go here!}
+At approximately the mid-point of the semester,
+thirty-nine thread-library projects were submitted
+on behalf of thirty-eight two-person teams and one
+auditing student who was working solo.
+%
+Submitted thread libraries were 800 lines of C code
+and 50 lines of x86 assembly-language code,
+though some were half that size and some were triple.
+%
+At this point, three teams dropped and the auditing
+student's requirements were complete;
+also, four teams fragmented---some students dropped and
+two new teams emerged.
+Thus, thirty-three teams attempted the kernel project.
+%
+Twenty-four teams submitted kernels on time and nine
+teams were forced to take extra time to complete the
+kernel project.
+Most kernels ranged in size from 2,500 to 3,500 lines
+of C code and 100 to 400 lines of assembly code,
+though some submissions were roughly twice as large
+(late kernels and on-time kernels were essentially the
+same size).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \pull \p4 \text \from \410.tex?
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The symmetric-multiprocessing kernel-extension project
+was a mixed success.
+Of the twenty-four teams that attempted it,
+one solo team was unable to submit a solution.
+Nineteen of the remaining twenty-three groups
+completed part of the project---at least enough to
+boot multiple cores (one, two, or eight depending
+on the simulator configuration) into their scheduler.
+Four of those nineteen groups were able to run
+almost all of our test suite of simple multi-threaded
+applications.
+Seven more groups were able to run roughly half of
+the combinations of core counts versus test applications.
+Finally, it was the sense of the grading team that
+most groups would have been able to substantially complete
+the project given a further week of development time.
+We view this outcome as an indication that feasible adjustments
+to the course design would bring this material within reach
+of most students.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \newcommand{\mathcols}{\multicolumn{1}{c}{$\downarrow$} & \multicolumn{1}{c}{$M$} & \multicolumn{1}{c}{$\mu$} & \multicolumn{1}{c}{$\sigma$} & \multicolumn{1}{c|}{$\uparrow$} }
+%
+% \begin{table*}[htb]
+% \typeout{Manually banging a table* to 9 points...}
+% \fontsize{9}{12}\selectfont
+% \centering
+% \begin{tabular}{|l|r||r r r r r|r r r r r|} 
+% \hline
+% Project   & Number    & \multicolumn{5}{c|}{C SLOC} & \multicolumn{5}{c|}{Assembly SLOC} \\
+% Type      & Submitted & \mathcols              & \mathcols                     \\
+% \hline
+% \hline
+% Thread libraries & 39 &  491 &  805 &  837 & (225) & 1693   &  29 &  43 & 52.5 & (25.0) & 142 \\
+% \hline
+% All kernels      & 31 & 2545 & 3408 & 3620 & (824) & 6697   & 107 & 365 &  391 & (198) & 829 \\
+% \hline
+% On-time kernels  & 24 & 2545 & 3459 & 3703 & (881) & 6697   & 107 & 373 &  391 & (208) & 829 \\
+% \hline
+% Late kernels     &  7 & 2576 & 3288 & 3336 & (490) & 4124   & 189 & 354 &  391 & (155) & 716 \\
+% \hline
+% \end{tabular}
+% \caption{Code-size statistics}
+% \label{table:sizes}
+% \end{table*}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2013-01-19 edupar-13/intro.tex
Finish conclusion
intro.tex: Minor compression tweaks.

@@
 
-Managing parallel computation, whether on thousand-node clusters or on one single-core machine, is often taught by resorting to powerful abstractions that simplify the hardest problems.
+Managing parallel computation, whether on thousand-node clusters or on a single-core machine, is often taught via powerful abstractions that simplify the hardest problems.
 Programming patterns such as SPMD, divide-and-conquer, and fork/join parallelism enable developers to reason about parallel algorithms at a high, abstract level.
@@
 
-In this paper we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
+In this paper we describe our experience teaching concurrent programming at the operating-system level to advanced undergraduate students at Carnegie Mellon University (CMU).
 In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, which runs on real hardware, in which they implement threads and processes.
-Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in user space on top of a Pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
+Presented with a kernel specification called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in user space on top of a Pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
 During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
-Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing.
+Finally, in another two-week project, students implement an extension to their Pebbles kernel.
+This project changes every semester, and recently has involved supporting symmetric multiprocessing (SMP).
 % in the past has included extending their kernel to serve as a hypervisor, or to support symmetric multiprocessing.
@@
 % such as Pintos~\cite{pintos} and Xv6~\cite{xv6},
-Pebbles focuses more on ground-up design than on teaching a broad feature-set.
+Pebbles focuses more on ground-up design than on teaching a broad feature set.
 % Pebbles allows students to design and implement their systems ``from the ground up.''
@@
 however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
-In 15-410, we give starter code only for the machine boot sequence and for a standard library including dynamic memory allocation, leaving students to design threading and scheduling primitives from scratch.
+In 15-410, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \xmalloc()}), leaving students to design threading and scheduling primitives from scratch.
 
-% As we ask our students to write more and more difficult concurrent code in one semester,
-As we plan to incorporate SMP into 15-410's curriculum, expanding the challenges our students face in one semester,
-we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
-However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.
+%% As we ask our students to write more and more difficult concurrent code in one semester,
+%As we plan to incorporate SMP into 15-410's curriculum, expanding the challenges our students face in one semester,
+%we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
+Incorporating SMP into the 15-410 curriculum would expand the challenges students face within a single semester.
+However, we wish to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.
+One solution would be easing the burden of debugging concurrency errors.
 Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads to take execution paths that expose concurrency bugs.
@@
 
+\shortversion{}{
 The rest of this paper is organized as follows.
@@
 Section~\ref{sec:future} concludes, discussing strategies for making Landslide more accessible to struggling students, and how Landslide might be applied to related operating systems projects at other universities.
+}

### 2013-01-19 edupar-13/410.tex
Fixed SIMFLEX citation.
410.tex: Compress wording in Pebbles syscall table.

@@
 %% could be left out (if we're in space trouble?) -- bblum
-features~\cite{UW-GEMS,SimFlex,FeS2}.
+features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
 Performance is very reasonable (Simics can simulate certain
@@
 		\x{sleep} & Blocks a thread for a given number of ticks. \\
-		\x{swexn} & Registers a given user-space function as a software exception handler.\\
+		\x{swexn} & Registers a user-space software exception handler.\\
 		\hline
@@
 		\x{new_pages} & Allocates a specified region of memory. \\
-		\x{remove_pages} & Deallocates same. \\
+		\x{remove_pages} & Deallocates a memory region. \\
 		\hline
@@
 		\hline
-		\x{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system''. \\
+		\x{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system.'' \\
 		\x{halt} & Ceases execution of the operating system. \\
-		\x{misbehave}* & Selects among several behaviors for thread interleaving in certain scenarios. \\
+		\x{misbehave}* & Selects among several thread-scheduling policies. \\
 		\hline

### 2013-01-19 edupar-13/410.tex
410.tex: Compression pass
* SMP-project description rewritten
* lots of \shortversion{}, especially in Limitations
* some "permanent" phrase shortening

@@
 For CS students it is one of five courses which fulfill
+\shortversion{a ``systems programming'' degree requirement.}{
 the ``Software Systems'' requirement (at present, the others
@@
 and Computer Networks).
+}
 The prerequisite course is 15-213,
@@
 outline the programming projects,
-discuss our use of Wind River Simics~\cite{simics} as the
-the basic simulation/debugging environment
-and our approaches to grading,
+discuss our use of Wind River Simics~\cite{simics} as
+the simulation/debugging environment,
+discuss our grading approach,
 and close by mentioning current limitations in the scope
@@
 Students study high-level concepts
-such as protection (least privilege, access control lists vs.
-capabilities)
-and
+such as
+\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
 file-system internals,
@@
 so they must synthesize design requirements from multiple sources
-and are provided with an opportunity to 
-decide on module boundaries and inter-module conventions
-themselves.
+and must
+choose their own module boundaries and inter-module conventions.
 %
-Due to the foundational nature of kernel code,
+\sortversion{The}{Due to the foundational nature of kernel code, the}
 assignment design and grading encourage students to
 think about corner cases, including resource exhaustion,
-instead of being satisfied by ``the right basic idea''
+instead of being satisfied by
+\shortversion{}{``the right basic idea''}
 implementations that handle only auspicious situations.
@@
 a client program, it displays the program's stack
-symbolically, rendering saved program-counter values
+symbolically,
+\shortversion
+{showing function names, strings, floating-point values, etc.}
+{rendering saved program-counter values
 as function names and printing function parameters
-in accordance with their types.
-This project meets a variety of goals:
-it enables students to review key process-model and
+in accordance with their types.}
+This project
+\shortversion{}{meets a variety of goals: it}
+enables students to review key process-model and
 language-runtime concepts from the
@@
 This project is built using a standard C tool chain
-and debugged as students see fit (generally using
-a combination of \x{printf()} and \x{gdb}).
+and debugged as students see fit (generally
+\shortversion{}{using a combination of}
+\x{printf()} and \x{gdb}).
 %% TODO I've been using "\x{printf()}" for code, an alias for \lstinline I
@@
 If the floppy-disk image is copied to a real floppy disk
-or embedded into an ``El Torito'' bootable compact disc image,
+or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
 it can be booted on standard PC hardware.
@@
 user-space programs.
-The library specification is essentially a stripped-down
+The \shortversion{}{library} specification is essentially a stripped-down
 version of POSIX Pthreads.
-The primary goal of the project is providing students
-with a laboratory for
-experimenting with concurrency and atomicity.
+The primary goal of the project is
+\shortversion{for students to experiment}{providing students with a laboratory for experimenting}
+with concurrency and atomicity.
 Students begin by designing mutexes using any
@@
 threads.
-A second goal is encouraging students to carefully
+A second goal is encouraging students to
 consider how ingredients are assembled into
@@
 %% Good rhetoric (below). -- bblum
-One example of this is a semantic gap
+\shortversion{There is a semantic gap}{One example of this is a semantic gap}
 between the library-level \x{thr_create()}
@@
 Because these abstractions are so different,
-students have the opportunity to
+students must
 understand each as an independent entity while
@@
 Student library code is linked against small
-test programs provided by the course staff,
+test programs provided by the course staff\shortversion{.}{,
 producing one statically linked ELF executable
-per test program.
+per test program.}
 The test programs are bundled into a RAM-disk
-image and linked against a kernel written by
+image and linked against a ``reference kernel'' written by
 the course staff
@@
 
-The behavior of the reference kernel is specified
+Pebbles kernel behavior is specified
 via a twelve-page document~\cite{kspec}.
-% the handout for the thread library project itself
-% is nineteen pages and includes not only a specification
-% of the thread-library primitives but also
-% a suggested implementation plan, grading guidelines,
-% optional design challenges, etc.
+\shortversion{}{The handout for the thread library project itself
+is nineteen pages and includes not only a specification
+of the thread-library primitives but also
+a suggested implementation plan, grading guidelines,
+optional design challenges, etc.}
 In addition to providing a reliable execution
 substrate,
-this ``reference kernel'' schedules
-the execution of user-space threads created by
+the reference kernel schedules
+\shortversion{}{the execution of} user-space threads created by
 student code according to a variety of
 interleaving policies.
-In order to aid debugging of concurrency problems,
+\shortversion{}{In order to aid debugging of concurrency problems,
 the \x{misbehave()} system call enables
-students to manually select an interleaving policy.
+students to manually select an interleaving policy.}
 
@@
 For the fourth project, two-student teams
-implement a kernel which complies with the
+implement a kernel which implements the
 same specification as the reference kernel
@@
 and code for setting up and tearing down
-threads and processes.
-Students re-use the basic device drivers
-they implemented in the second project.
-These ingredients are combined to yield
+threads and processes
+(they re-use their game-project device drivers).
+\shortversion{}{These ingredients are combined to yield
 twenty-five system calls
@@
 console I/O, and miscellaneous),
-which we discuss further in Section~\ref{sec:pebbles}.
+which we discuss further in Section~\ref{sec:pebbles}.}
 
@@
 can host guest kernels and
-addding multi-processor support to their single-processor kernels.
+adding multi-processor support to their single-processor kernels.
 
-In the multi-processor project, students extended their kernels to schedule multiple threads at once on up to eight cores.
-We provided base code for determining the system's number of CPUs, for booting the application processors (APs), and for sending inter-processor interrupts (IPIs).
-Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
-to implement moving threads between processors,
-and
-to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
-To allow struggling groups to make progress on the project without solving every synchronization challenge, we allowed the use of a ``big kernel lock'', and rewarded groups according to how little they needed to use it.
+In the multi-processor project, students extended their kernels to schedule multiple threads across up to eight cores.
+We provided base code for determining a system's CPU count, for booting the application processors (APs), and for sending inter-processor interrupts (IPIs).
+% Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
+% to implement moving threads between processors,
+% and
+% to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
+Students must move threads between processors,
+synchronize the removal of memory mappints from multiple TLBs,
+and must account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
+To allow struggling groups to make progress on the project without solving every synchronization challenge, we allowed the use of a ``big kernel lock,'' and rewarded groups according to how little they needed to use it.
 Submissions were also graded according to
@@
 provides very faithful bit-level support
-not only for code that behaves correctly but
-also for code that is ``incorrect within reason.''
+not only for code that behaves correctly
+\shortversion
+{but also for student kernels that accidentally ``abuse'' hardware.}
+{but also for code that is ``incorrect within reason.''
 Occasionally student code manages to invoke an unsupported
@@
 Meanwhile, inside its large envelope, Simics supports
-\textit{very} detailed models of various x86 processors.
+\textit{very} detailed models of various x86 processors.}
 Unlike hardware virtualization environments,
@@
 
-Students write code that runs on one specific
-hardware platform.
+\shortversion{}{Students write code that runs on one specific
+hardware platform.}
 We target x86 PCs because machines of this type are
@@
 % by labyrinthine Linux source code
-Finally, the platform is fairly stable---with only a
+\shortversion{}{Finally, the platform is fairly stable---with only a
 little care, it is possible to write a compliant
@@
 i.e., almost any PC hardware which is likely to be in
-working condition.
+working condition.}
 
@@
 frustrating for compiler writers,
-from the point of view of OS students,
+\shortversion{}{from the point of view of OS students,}
 the existence of fewer registers means less work
@@
 debugging time without imparting greater insight.
-Finally, since it is impossible for an x86 processor
+\shortversion{}{Finally, since it is impossible for an x86 processor
 to be in 64-bit mode with virtual memory off,
@@
 32-bit--to--64-bit transition or would need
-to be given a template virtual-memory implementation.
+to be given a template virtual-memory implementation.}
 All in all, at present we are satisifed with
@@
 architectures.
-Lecture material educates students about the
-general topic of out-of-order memory systems
+Lecture material educates students about
+%the general topic of
+out-of-order memory systems
 and fencing,
@@
 %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Warning: vanish() is *not* "the Pebbles name for exit()"--
+%% exit() used to mean one thing, and now means some other
+%% thing which is muddy.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The process lifecycle is loosely based on the
+%% Unix model: \x{fork()}, \x{exec()}, \x{wait()},
+%% and \x{vanish()} (the Pebbles name for \x{exit()}).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The process lifecycle is loosely based on the
-Unix model: \x{fork()}, \x{exec()}, \x{wait()},
-and \x{vanish()} (the Pebbles name for \x{exit()}).
+Unix \x{fork()}/\x{exec()}/\x{wait()} model.
 We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
@@
 its traditional position in Unix-like kernels
-was done because it simplfies the kernel and
+simplfies the kernel and
 ``front-loads'' policy questions to an earlier

### 2013-01-19 edupar-13/410.tex
Remember to run "make" BEFORE committing.

@@
 %
-\sortversion{The}{Due to the foundational nature of kernel code, the}
+\shortversion{The}{Due to the foundational nature of kernel code, the}
 assignment design and grading encourage students to

### 2013-01-19 edupar-13/intro.tex
Remember to run "make" BEFORE committing.

@@
 however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
-In 15-410, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \xmalloc()}), leaving students to design threading and scheduling primitives from scratch.
+In 15-410, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \x{malloc()}), leaving students to design threading and scheduling primitives from scratch.
 

### 2013-01-19 edupar-13/410.tex
410.tex: "invoke vanish()" -> "exit" to avoid a forward ref.

@@
 is that of coordinating a parent and a child task that ``simultaneously''
-invoke \x{vanish()}:
+exit:
+% invoke \x{vanish()}:
 when a task completes,

### 2013-01-19 edupar-13/410.tex
410.tex: Squeeze pass over "Grading" section.

@@
 The test suite is a number of user-space programs which test
-functionality ranging from the basic process lifecycle, to robustness
-against corner-case security vulnerabilities, to long-term stability
+functionality
+including the basic process lifecycle, robustness
+against corner-case security vulnerabilities, and long-term stability
 under large stress tests.
@@
 %
-Of course, due to the nondeterministic nature of many common bugs in student code,
+Of course, due to the nondeterministic nature of many bugs in student code,
 these tests can provide only a rough estimate of a submission's correctness,
@@
 
-\item We believe that students learn more from bugs
+\item \shortversion{Students}{We believe that students} learn more from bugs
 that are harder to find.
@@
 
-For these reasons, we believe much of the learning
+For these reasons, \shortversion{}{we believe} much of the learning
 associated with the kernel project necessarily occurs
@@
 % [suppressed: bad to build P4 on a shaky P3]
-We believe it would be particularly detrimental to
+\shortversion{It}{We believe it} would be particularly detrimental to
 assign ``partial credit'' to a large body of non-working
@@
 Students who pass the completeness ``hurdle'' submit their kernels
-and can move on to the ``Pebbles Extension'' project.
-Meanwhile, students who were unable to meet the
+and can move on to the kernel-extension project.
+\shortversion{Students}{Meanwhile, students} who were unable to meet the
 hurdle criteria on time receive additional time
@@
 Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
-The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for style, architectural issues, race conditions, and preemptibility.
-After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
+The instructors and teaching assistants print each group's submitted kernel and mark it with red ink, auditing the code for style, architectural issues, race conditions, and preemptibility.
+After the inking process, each group meets with \shortversion{their grader}{the staff member who graded them} for an hour-long ``kernel interview,'' essentially a debriefing in which student misconceptions can be rectified.
 Staff discretionary points constitute 20\% of the kernel project's grade.

### 2013-01-19 edupar-13/landslide.tex
shrink threadfork picture
Add SMP4 paragraph and clean up various stuff
landslide.tex: A couple initial text squeezes.

@@
 Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
-It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
+It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different thread interleavings in a concurrent system.
 It follows in the footsteps of related tools such as
@@
 Landslide triggers artificial clock interrupts
-to force a new thread to interleave with the current one.
+% to force a new thread to interleave with the current one.
+to force the scheduler to run a different thread.
 When a test program finishes execution according
@@
 
-Figure~\ref{fig:threadfork} shows code containing a timer-dependent bug common in many Pebbles implementations which we will use as a running example in the rest of this section.
+Figure~\ref{fig:threadfork} shows code containing a timer-dependent bug common in many Pebbles implementations which we will use as a running example\shortversion{.}{in the rest of this section.}
 This code is a simplified implementation of the \x{thread_fork} system call: it constructs data structures necessary for the new thread, then asks the scheduler to make the new thread runnable, then returns its assigned thread ID.
-However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run long enough to invoke \x{exit}, causing \x{child} to become a dangling pointer.
+However, if the timer preempts execution at line~\ref{line:hax}, the child thread might run long enough to exit, causing \x{child} to become a dangling pointer.
 
@@
 \begin{figure}[t]
-\includegraphics[width=0.48\textwidth]{threadfork/threadfork.pdf}
+	\centering
+\includegraphics[width=0.42\textwidth]{threadfork/threadfork.pdf}
 \caption{The state space of possible thread interleavings can be viewed as an {\em execution tree}.
@@
 If a student were testing thread death and reaping, they might want decision points to appear in \x{wait} and \x{vanish} but not if unrelated virtual memory operations are also in progress.
-They should write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
+Accordingly, they could write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
 The \x{within_function} directive requires that at least one of the specified functions shall be on the call stack when decision points are identified, and \x{without_function} requires the opposite.

### 2013-01-19 edupar-13/outcomes.tex
outcomes.tex: Trivial drive-by text squeezings.

@@
 %
-Submitted thread libraries were 800 lines of C code
+Submitted thread libraries were roughly 800 lines of C code
 and 50 lines of x86 assembly-language code,
@@
 
-The symmetric-multiprocessing kernel-extension project
+The SMP kernel-extension project
 was a mixed success.
@@
 completed part of the project---at least enough to
-boot multiple cores (one, two, or eight depending
-on the simulator configuration) into their scheduler.
+boot multiple cores
+\shortversion{}{(one, two, or eight depending on the simulator configuration)}
+into their scheduler.
 Four of those nineteen groups were able to run

### 2013-01-20 edupar-13/landslide.tex
landslide.tex: Squeezes that I hope Ben will sign off on.

@@
 Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
-The primary advantage of systematic testing is the ability to force execution sequences that might be very rare in simple repeated testing.
+% The primary advantage of systematic testing is the ability to force execution sequences that might be very rare in simple repeated testing.
+Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.
 
-Landslide ``explores'' decision trees by controlling the kernel's execution in two ways.
+Landslide ``explores'' an execution tree by controlling the kernel's execution in two ways.
 First, at each decision point, it can force an arbitary thread to run in place of the current one by injecting timer interrupts to force the kernel to context switch.
-Second, whenever the kernel finishes executing a particular interleaving of the test case, it reverts the system to an earlier state to explore additional interleavings using Simics's reverse execution.
+%Second, whenever the kernel finishes executing a particular interleaving of the test case, it reverts the system to an earlier state to explore additional interleavings using Simics's reverse execution.
+Second, when the kernel finishes executing a particular interleaving of the test case, it invokes Simics reverse execution to revert the system to an earlier state and explore the next interleaving.
 
 In Figure~\ref{fig:threadfork}, the necessary decision point for finding the bug is at line~\ref{line:hax}.
-With correct instrumentation (discussed in Section~\ref{sec:instrument} and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point when the child thread becomes runnable.
+With correct instrumentation (discussed in Section~\ref{sec:instrument} and shown in Figure~\ref{fig:annotation}), Landslide can automatically identify this decision point because a thread becomes runnable.
 Landslide also automatically identifies decision points at voluntary reschedules between threads (e.g.
 \x{yield()}).
-Together these decision points define an execution tree which exposes the bug, depicted in Figure~\ref{fig:tree}.
+Together these decision points define an execution tree which exposes this bug, depicted in Figure~\ref{fig:tree}.
 
-Of course, additional decision points are necessary to expose more subtle race conditions.
-We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
+Of course, additional decision points are necessary to expose more-subtle race conditions.
+\shortversion
+{In Section~\ref{sec:instrument} we describe an interface for adding new
+decision points to expand the state space,
+and also an interface for restricting Landslide's attention to certain kernel components,
+which helps cope with the exponentially-growing nature of state spaces with many decision points.}
+{We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
 In order to cope with the exponentially-growing nature of state spaces with many decision points,
-\shortversion{
-we also provide an interface for making Landslide consider
-}{
 Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
 We also provide an interface for helping Landslide further reduce the state space by considering
-}
-decision points only from certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.
+decision points from only certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.}
 
@@
 
-When Landslide identifies a bug, it outputs a {\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.
-The trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces to show when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
-With this trace, the student can better understand the concurrent execution that led to the bug.
+When Landslide identifies a bug, it outputs a
+\shortversion
+{{\em decision trace} (Figure~\ref{fig:trace}).}
+{{\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.}
+The trace reports what kind of bug was detected
+\shortversion
+{}
+{(for uses-after-free, it also prints stack traces to show when the block was allocated and freed),}
+and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
+With this trace, the student can better understand the concurrent execution that exposed the bug.
 
@@
 We provide a set of annotation functions, named with the prefix \x{tell_landslide}, for this purpose.
-The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
+The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
 Figure~\ref{fig:annotation} shows annotated code corresponding to Figure~\ref{fig:threadfork}.
 
-There is also a configuration file, \x{config.landslide}, in which the student must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which userspace test program Landslide should invoke.
+There is also a configuration file, \x{config.landslide}, in which the student must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which user-space test program Landslide should invoke.
 Finally, there are two short (nominally two-line) functions used within Landslide itself that the user must implement.
@@
 \end{lstlisting}
-\caption{Landslide requires the programmer to annotate important concurrency events in their kernel.
-\x{tell_landslide_thread_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \x{tell_landslide_forking()} announces a new thread's existence.}
+\caption{Landslide requires the programmer to annotate important concurrency events\shortversion{.}{in their kernel.} \x{tell_landslide_thread_runnable()} indicates that a thread can henceforth be forced to run with timer interrupts, and \x{tell_landslide_forking()} announces a new thread's existence.}
 \label{fig:annotation}

### 2013-01-20 edupar-13/outcomes.tex
outcomes.tex: This pass clarified (I hope), but produced no actual vertical shrinkage.

@@
 report on the results of applying Landslide to
-kernels previously written by members of the course staff
+both kernels previously written by members of the course staff
 and kernels written that semester by students.
@@
 Of the twenty-four teams that attempted it,
-one solo team was unable to submit a solution.
+one (solo) team was unable to submit a solution.
 Nineteen of the remaining twenty-three groups
@@
 boot multiple cores
-\shortversion{}{(one, two, or eight depending on the simulator configuration)}
 into their scheduler.
-Four of those nineteen groups were able to run
-almost all of our test suite of simple multi-threaded
-applications.
-Seven more groups were able to run roughly half of
-the combinations of core counts versus test applications.
+Four of those nineteen groups were able to pass
+almost all of our test suite,
+which consisted of simple multi-threaded applications
+and machines with various core counts\shortversion
+{.}
+{(one, two, or eight depending on the simulator configuration).}
+Seven more groups were able to pass roughly half of the test suite.
 Finally, it was the sense of the grading team that
@@
 
-We evaluated Landslide in two ways: first, by instrumenting two prior student kernels to measure the exploration time needed to find different races, and second, by meeting with current students of 15-410 in the Spring 2012 semester, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
+We evaluated Landslide in two ways: first, by instrumenting two prior-semester student kernels to measure the exploration time needed to find different races, and second, by meeting with current-semester students, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
 
-In the first phase, we instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
+In the first phase, we instrumented one kernel written by a teaching assistant in a previous year, and also one student kernel later graded by that TA.
 We configured Landslide to search for five complicated well-known race conditions.
-In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
-Using additional decision points only on calls to \x{mutex_lock}, Landslide found each of the six bugs in 11 to 57 seconds on a 2.6 GHz Xeon server, executing between 1 and 377 distinct interleavings for each.
+In addition to finding all five races, Landslide also found a sixth previously unknown race in the TA's own kernel.
+Using additional decision points only on calls to \x{mutex_lock}, Landslide found each of the six bugs in 11 to 57 seconds on a 2.6 GHz Xeon server, executing between 1 and 377 distinct interleavings per bug.
 

### 2013-01-20 edupar-13/outcomes.tex
outcomes.tex: Mention here that volunteers were "top students", not random.

@@
 
-We evaluated Landslide in two ways: first, by instrumenting two prior-semester student kernels to measure the exploration time needed to find different races, and second, by meeting with current-semester students, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
+We evaluated Landslide in two ways: first, by instrumenting two prior-semester student kernels to measure the exploration time needed to find different races, and second, by meeting with current-semester student volunteers, before they submitted their kernel for grading, to see if they could find bugs on their own with Landslide.
+(The volunteers were chosen from students with free time, i.e., were not chosen at random.)
 

### 2013-01-20 edupar-13/conclusion.tex
conclusion.tex: Squeeze pass.  Result of all squeezing: 8.6 pp. -> 8.2 pp.

@@
 
-Carnegie Mellon University's operating systems course offers a series of concurrency-focused projects to advanced undergraduates focused on low-level software development.
-These projects culminate in the development of a Pebbles kernel, a fully preemptible UNIX-like kernel which includes process lifecycle management, virtual memory, and some device drivers, during the course's six-week-long fourth project.
+Carnegie Mellon University's operating systems course offers
+advanced undergraduates
+a series of
+development projects focused on concurrency in low-level software.
+These projects culminate in the six-week development of a ``Pebbles'' kernel, a fully preemptible UNIX-like kernel\shortversion{.}{which includes process lifecycle management, virtual memory, and some device drivers}.
 % Dave: You could add here a brief blurb about what students accomplish, e.g. LOC, boot on real hardware, whatever. Or it would be fine without.
-In order to keep up with emerging trends in hardware architecture, especially the increasing focus on massively multi-core processing, we seek to include more hardware concurrency and the resulting complexity in the course's curriculum.
+In response to emerging trends in hardware architecture, especially the increasing focus on massively multi-core processing, we seek to include more hardware concurrency (and the resulting complexity) in the course's curriculum.
 
@@
 
-Rather than making room for more material by reducing the projects' design complexity, which we feel would compromise the learning experience, we are experimenting with concurrency debugging tools to let students spend less time searching for errors in their code.
+Rather than making room for more material by reducing the projects' design complexity, which we feel would compromise the learning experience, we are experimenting with concurrency debugging tools
+\shortversion
+{which help students find errors faster.}
+{to let students spend less time searching for errors in their code.}
 We have presented Landslide, a tool that integrates with the Simics simulator students already use and provides a more principled approach to bug-finding than conventional stress testing.
 
-Due to the open-ended design problems students solve in their kernels, Landslide imposes a moderate annotation burden in order to instrument a kernel for systematic testing.
-However, by the very nature of having enough spare time to volunteer, the selection of students we worked with to use Landslide was necessarily biased towards the ``top of the class''.
-Though our experiments with these students show promising results, we need to reduce Landslide's annotation burden as much as possible to make it useful for struggling students as well.
-
-We believe some annotations can be automatically detected;
-for example, rather than requiring students to indicate if their kernel has a mechanism for disabling preemption apart from disabling interrupts, Landslide could detect when a timer interrupt is received but doesn't induce a context switch.
-Other annotations, such as \x{tell_landslide_thread_runnable()}, could be replaced by mandating that the kernel's code paths in such cases include a function with a predetermined name.
-Landslide already automatically instruments the memory allocator by detecting function names in the starter code; similar could be done for thread lifecycle and context switch events if their APIs were specified.
-However, prescribing such templates has the downside of constraining the project's design space.
-As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.
+% Due to the open-ended design problems students solve in their kernels,
+% Landslide imposes a moderate annotation burden in order to instrument a kernel for systematic testing.
+% However, by the very nature of having enough spare time to volunteer, the selection of students we worked with to use Landslide was necessarily biased towards the ``top of the class''.
+In our experience,
+the annotation burden imposed on ``top of the class'' students by Landslide is modest.
+Though our experiments with these students show promising results, we seek to reduce Landslide's annotation burden to make it useful for struggling students as well.
+%
+We believe some annotations could be automatically detected;
+for example, rather than requiring students to declare if their kernel has a preemption disabling mechanism more elaborate than disabling interrupts, Landslide could detect when a timer interrupt doesn't induce a context switch.
+Other annotations, such as \x{tell_landslide_thread_runnable()},
+might be automated by requiring students to structure code in
+certain ways with pre-determined function names and parameters
+(Landslide already automatically instruments the memory allocator we provide).
+However, \shortversion{}{prescribing} such templates
+\shortversion{would constrain}{has the downside of constraining}
+the project's open-ended design space.
+We wish to better understand the tradeoffs between
+making Landslide more automatic and forcing particular design decisions on students.
+% As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.
 
@@
 15-410 over the past decade.
-Stephen Muckle, currently at Qualcomm, led
-the switch to x86 projects running in Simics.
+Stephen Muckle, currently at Qualcomm,
+led the switch to x86 projects running in Simics.
 Nathaniel Wesley Filardo,

### 2013-01-20 edupar-13/landslide.tex
respond to dave's edits to landslide.tex

@@
 Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
-% The primary advantage of systematic testing is the ability to force execution sequences that might be very rare in simple repeated testing.
-Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.
+The advantage of systematic testing is the ability to force execution sequences that might be rare in simple testing.
+% Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.
 
@@
 Of course, additional decision points are necessary to expose more-subtle race conditions.
-\shortversion
-{In Section~\ref{sec:instrument} we describe an interface for adding new
-decision points to expand the state space,
-and also an interface for restricting Landslide's attention to certain kernel components,
-which helps cope with the exponentially-growing nature of state spaces with many decision points.}
+% \shortversion
+% {In Section~\ref{sec:instrument} we describe an interface for adding new
+% decision points to expand the state space,
+% and also an interface for restricting Landslide's attention to certain kernel components,
+% which helps cope with the exponentially-growing nature of state spaces with many decision points.}
 {We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
 In order to cope with the exponentially-growing nature of state spaces with many decision points,
+\shortversion{
+we also provide an interface for making Landslide consider
+}{
 Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
 We also provide an interface for helping Landslide further reduce the state space by considering
+}
 decision points from only certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.}

### 2013-01-20 edupar-13/conclusion.tex
compress acknowledgements

@@
 designed the \x{swexn()} system call and helped critique this paper.
-The current reference kernel is named ``pathos''
-and was written by Michael J. Sullivan,
+The current reference kernel, called ``pathos,''
+was written by Michael J. Sullivan,
 currently a CMU CS Ph.D.\ candidate,
@@
 Joshua Wise, currently at NVIDIA,
-proposed and led several kernel-extension projects,
-including the hypervisor project.
-Over the past decade the course has benefitted
+led several kernel-extension projects.
+% including the hypervisor project.
+The course has benefitted
 greatly from free educational Simics licenses

### 2013-01-20 edupar-13/410.tex
It bloody fits on 8 pages (with garth's feedback)

@@
 assignment design and grading encourage students to
-think about corner cases, including resource exhaustion,
+think about corner cases, including resource exhaustion\shortversion{.}{,
 instead of being satisfied by
-\shortversion{}{``the right basic idea''}
-implementations that handle only auspicious situations.
+``the right basic idea''
+implementations that handle only auspicious situations.}
 %
@@
 The second project is a simple game, such as Hangman,
-which runs standalone (without an underlying
-operating system).
+which runs without an underlying
+operating system.
 We provide students with selected C~library routines
@@
 and a countdown timer.
-The game, which varies by semester, provides students
-with the ability to test their drivers.
+\shortversion{The game enables students to test their drivers.}
+{The game, which varies by semester, provides students
+with the ability to test their drivers.}
 %
@@
 and booted via GRUB.
-If the floppy-disk image is copied to a real floppy disk
+If the \shortversion{}{floppy-disk} image is copied to a real floppy
 or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
 it can be booted on standard PC hardware.
-However, students most often boot the floppy image
-using the Simics system emulator (described below).
+However, students most often \shortversion{use}{boot the floppy image
+using} the Simics system emulator (described below).
 
@@
 The third project is implementing a 1:1 thread library for
-user-space programs.
-The \shortversion{}{library} specification is essentially a stripped-down
+user-space programs,
+essentially a stripped-down
 version of POSIX Pthreads.
-The primary goal of the project is
+The primary goal is
 \shortversion{for students to experiment}{providing students with a laboratory for experimenting}
@@
 per test program.}
-The test programs are bundled into a RAM-disk
-image and linked against a ``reference kernel'' written by
+The test programs \shortversion{run on}{are bundled into a RAM-disk
+image and linked against} a ``reference kernel'' written by
 the course staff
 and provided in binary form.
-
+%
 Pebbles kernel behavior is specified
@@
 Students who complete the kernel project on time
-then work on a smaller kernel-extension project
-with various content depending on the semester.
+then work on a kernel-extension project\shortversion{.}
+{with various content depending on the semester.}
 Past projects have included
@@
 Students must move threads between processors,
-synchronize the removal of memory mappints from multiple TLBs,
-and must account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
-To allow struggling groups to make progress on the project without solving every synchronization challenge, we allowed the use of a ``big kernel lock,'' and rewarded groups according to how little they needed to use it.
+synchronize the removal of memory mappings from multiple TLBs,
+and account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
+To allow struggling groups to make progress without solving every synchronization challenge, we allowed the use of a ``big kernel lock,'' and rewarded groups according to how little they needed to use it.
 Submissions were also graded according to
@@
 
-As mentioned above, the main execution and debugging
+\shortversion{The}{As mentioned above, the} main execution and debugging
 platform used in the class is
@@
 \shortversion
-{but also for student kernels that accidentally ``abuse'' hardware.}
+{but also for kernels that accidentally ``abuse'' hardware.}
 {but also for code that is ``incorrect within reason.''
@@
 features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
-Performance is very reasonable (Simics can simulate certain
+\shortversion{}{Performance is very reasonable (Simics can simulate certain
 cases, such as a CPU halted awaiting a clock interrupt,
-much faster than real time).
+much faster than real time).}
 %% TODO: One huge advantage of simics over QEMU that we should mention is
@@
 up their code, type commands, and observe the results.
-Also, unlike many System-on-a-Chip platforms,
-detailed hardware descriptions are readily available.
+\violence{}{Also, unlike many System-on-a-Chip platforms,
+detailed hardware descriptions are readily available.}
 % rather than being proprietary or "documented"
@@
 The course still targets the 32-bit
-subset of the x86 platform, for three reasons.
+subset of the x86 platform, for multiple reasons.
 %% TODO I'm not sure I believe this is an advantage. Instead of having
@@
 within the scope of a single semester,
-students do not have the opportunity to
+students do not \violence{}{have the opportunity to}
 implement common features of many kernels,

### 2013-01-20 edupar-13/410.tex
feedback from munin and rpearl

@@
 Pebbles kernel behavior is specified
-via a twelve-page document~\cite{kspec}.
+in a twelve-page document~\cite{kspec}.
 \shortversion{}{The handout for the thread library project itself
@@
 For the fourth project, two-student teams
-implement a kernel which implements the
+produce a kernel which implements the
 same specification as the reference kernel
 they previously relied on.
-They implement some approach to
+They design and implement some approach to
 synchronizing and blocking threads while
-they are in kernel space,
+they are in kernel space\shortversion{,}{\footnote{
+Student designs for their kernel's synchronization primitives vary widely.
+For example, ``top of the class'' students who went on to become teaching assistants have alternately used condition variables but not semaphores, and semaphores but not condition variables.
+},}
 a simple round-robin scheduler,
@@
 Plan~9's \x{rfork()}~\cite{plan9} or Linux's \x{clone()},
-which allow a continuum of sharing between
+both of which allow a continuum of sharing between
 identical threads and independent processes,

### 2013-01-21 edupar-13/410.tex
Cite a different Plan 9 paper.

@@
 This partitioning is a different approach from that of
-Plan~9's \x{rfork()}~\cite{plan9} or Linux's \x{clone()},
+Plan~9's \x{rfork()}~\cite{Pike90plan9} or Linux's \x{clone()},
 both of which allow a continuum of sharing between

### 2013-01-21 edupar-13/410.tex
Copyedit pass on 410.tex

@@
 the simulation/debugging environment,
-discuss our grading approach,
+% discuss our grading approach,
 and close by mentioning current limitations in the scope
@@
 This project
-\shortversion{}{meets a variety of goals: it}
-enables students to review key process-model and
+\shortversion{enables}{meets a variety of goals: it enables}
+students to review key process-model and
 language-runtime concepts from the
@@
 This project is built using a standard C tool chain
-and debugged as students see fit (generally
-\shortversion{}{using a combination of}
-\x{printf()} and \x{gdb}).
-%% TODO I've been using "\x{printf()}" for code, an alias for \lstinline I
-%% adopted from the Rust paper template. We should be consistent because
-%% the font sizes are different (and maybe make \x{}'s font bigger too,
-%% what do you think?). -- bblum
+and debugged as students see fit (\shortversion{generally}
+{generally using a combination of}
+\x{printf()} and GDB).
 
@@
 We provide students with selected C~library routines
-(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()});
-they write a device-driver library consisting of
+(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()}),
+who then write a device-driver library consisting of
 console output (to a memory-mapped text-mode video display),
-keyboard input (interrupt driven),
+keyboard input (interrupt-driven),
 and a countdown timer.
@@
 C with a bit of x86-32 assembly code,
+which is then
 compiled and linked into an ELF executable,
@@
 and booted via GRUB.
-If the \shortversion{}{floppy-disk} image is copied to a real floppy
+If the \shortversion{image}{floppy-disk image} is copied to a real floppy
 or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
@@
 %
-Pebbles kernel behavior is specified
+The kernel's behavior is specified
 in a twelve-page document~\cite{kspec}.
-\shortversion{}{The handout for the thread library project itself
+\shortversion{In}{The handout for the thread library project itself
 is nineteen pages and includes not only a specification
@@
 a suggested implementation plan, grading guidelines,
-optional design challenges, etc.}
-In addition to providing a reliable execution
+optional design challenges, etc. In}
+addition to providing a reliable execution
 substrate,
-the reference kernel schedules
-\shortversion{}{the execution of} user-space threads created by
+the reference kernel
+\shortversion{schedules}{schedules the execution of} user-space threads created by
 student code according to a variety of
@@
 they gain experience with debugging at both the
-``algorithm level'' and the register/bit-field level.
+algorithm level and the register/bit-field level.
 
@@
 frameworks to evaluate proposed hardware
-%% TODO Fix citations. But also, maybe this mention isn't relevant and
-%% could be left out (if we're in space trouble?) -- bblum
 features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
@@
 much faster than real time).}
+\shortversion{}{A major advantage of using Simics over the QEMU emulator in particular
+is that QEMU only issues timer interrupts at basic-block boundaries,
+which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}
 %% TODO: One huge advantage of simics over QEMU that we should mention is
@@
 frustrating for compiler writers,
-\shortversion{}{from the point of view of OS students,}
-the existence of fewer registers means less work
+\shortversion{the}{from the point of view of OS students, the}
+existence of fewer registers means less work
 when it is necessary to verify that each register's
@@
 opposed to x86-32 requires more code and more
-debugging time without imparting greater insight.
-\shortversion{}{Finally, since it is impossible for an x86 processor
+debugging time without imparting greater
+\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
 to be in 64-bit mode with virtual memory off,
@@
 within the scope of a single semester,
-students do not \violence{}{have the opportunity to}
+students do \violence{not}{not have the opportunity to}
 implement common features of many kernels,
@@
 		\x{sleep} & Blocks a thread for a given number of ticks. \\
-		\x{swexn} & Registers a user-space software exception handler.\\
+		\x{swexn} & Registers a user-space function as a software exception handler.\\
 		\hline
@@
 		\x{new_pages} & Allocates a specified region of memory. \\
-		\x{remove_pages} & Deallocates a memory region. \\
+		\x{remove_pages} & Deallocates same. \\
 		\hline
@@
 simplfies the kernel and
-``front-loads'' policy questions to an earlier
+front-loads policy questions to an earlier
 part of the course,
@@
 
-\subsubsection{``Hurdle'' completion criteria}
+\subsubsection{Hurdle completion criteria}
 
 Because of the ambitious learning objectives,
-we employ a very detailed notion of ``completeness''
+we employ a very detailed notion of completeness
 for the kernel project.
-The ``hurdle model'' is based on the following observations:
+The ``hurdle'' model is based on the following observations:
 
@@
 In extreme cases, a software component may do
-``almost everything right,'' but ``the last bug'' may
+almost everything right, but the last bug may
 demonstrate a deep misunderstanding which requires a re-design.
@@
 
-For these reasons, \shortversion{}{we believe} much of the learning
+For these reasons, \shortversion{much}{we believe much} of the learning
 associated with the kernel project necessarily occurs
@@
 \shortversion{It}{We believe it} would be particularly detrimental to
-assign ``partial credit'' to a large body of non-working
+assign partial credit to a large body of non-working
 code.
@@
 can pass 80\% of the basic system-call sanity test,
-can pass 70\% of the ``solidity'' tests (which
+can pass 70\% of the solidity tests (which
 invoke system calls in illegal or non-trivial ways),
@@
 %Towards the end of the kernel project, students fill out the ``hurdle form'', a checklist of these criteria.
-Students who pass the completeness ``hurdle'' submit their kernels
+Students who pass the completeness hurdle submit their kernels
 and can move on to the kernel-extension project.

### 2013-01-21 edupar-13/conclusion.tex
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

@@
 We believe some annotations could be automatically detected;
-for example, rather than requiring students to declare if their kernel has a preemption disabling mechanism more elaborate than disabling interrupts, Landslide could detect when a timer interrupt doesn't induce a context switch.
+for example, rather than requiring students to declare if their kernel has a preemption-disabling mechanism more elaborate than disabling interrupts, Landslide could detect when a timer interrupt doesn't induce a context switch.
 Other annotations, such as \x{tell_landslide_thread_runnable()},
@@
 (Landslide already automatically instruments the memory allocator we provide).
-However, \shortversion{}{prescribing} such templates
+However, prescribing such templates
 \shortversion{would constrain}{has the downside of constraining}
@@
 We wish to better understand the tradeoffs between
-making Landslide more automatic and forcing particular design decisions on students.
+making Landslide more automatic and preserving the open-endedness of important design decisions.
 % As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.

### 2013-01-21 edupar-13/landslide.tex
clarify systematic vs stress sentence
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

@@
 It follows in the footsteps of related tools such as
-\shortversion{
-CHESS~\cite{chess},
-}{
-CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},
-}
+\shortversion{CHESS~\cite{chess},}
+{CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},}
 but is the first tool we know of to apply systematic exploration in a kernel development environment.
@@
 deadlock, infinite loops, and panics.
-\shortversion{}{
-In theory,
+\shortversion{}{In theory,
 by forcing a thread switch after every non-scheduler instruction,
@@
 thread-switch less often and to avoid repeating
-bug-equivalent execution paths.
-}
+bug-equivalent execution paths.}
 %Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels.
@@
 Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
-The advantage of systematic testing is the ability to force execution sequences that might be rare in simple testing.
+The primary advantage of systematic testing is the ability to force execution sequences that might be overlooked in simple stress testing.
 % Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.
@@
 
-Of course, additional decision points are necessary to expose more-subtle race conditions.
+Of course, additional decision points are often necessary to expose more-subtle race conditions.
 % \shortversion
@@
 % which helps cope with the exponentially-growing nature of state spaces with many decision points.}
-{We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
+We provide an interface for adding new decision points to expand the state space.
+%described in Section~\ref{sec:decision}.
 In order to cope with the exponentially-growing nature of state spaces with many decision points,
-\shortversion{
-we also provide an interface for making Landslide consider
-}{
-Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
-We also provide an interface for helping Landslide further reduce the state space by considering
-}
-decision points from only certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.}
+\shortversion{we also provide an interface for making Landslide consider}
+{Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
+We also provide an interface for helping Landslide further reduce the state space by considering}
+decision points from only certain components of the kernel.
+%which we discuss in Section~\ref{sec:focusing}.
+Section~\ref{sec:instrument} discusses these interface options.
 
@@
 \shortversion
-{}
-{(for uses-after-free, it also prints stack traces to show when the block was allocated and freed),}
-and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
+{and}
+{(for uses-after-free, it also prints stack traces to show when the block was allocated and freed), and}
+also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
 With this trace, the student can better understand the concurrent execution that exposed the bug.
@@
 In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs at suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
-\shortversion{}{
-In general,
+\shortversion{}{In general,
 % with false-negative bug detection,
@@
 We provide a set of annotation functions, named with the prefix \x{tell_landslide}, for this purpose.
-The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
+The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
 Figure~\ref{fig:annotation} shows annotated code corresponding to Figure~\ref{fig:threadfork}.

### 2013-01-21 edupar-13/outcomes.tex
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

@@
 pre-Landslide educational experience and then
-report on the results of applying Landslide to
-both kernels previously written by members of the course staff
-and kernels written that semester by students.
+report on the results of applying Landslide both to
+kernels previously written by members of the course staff
+and to kernels written that semester by students.
 
@@
 Seven more groups were able to pass roughly half of the test suite.
+None of the teams elected to use a ``big kernel lock'' to bypass any of the synchronization challenges.
 Finally, it was the sense of the grading team that

### 2015-09-20 pldi-16/paper.tex
Start writing PLDI paper.

%
% LaTeX template for prepartion of submissions to PLDI'15
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'15 web site.
% 
\documentclass[pldi]{sigplanconf-pldi15}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

\begin{document}

%
% any author declaration will be ignored  when using 'plid' option (for double blind review)
%

\title{Systematic Testing with Data-Race Preemption Points}
\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}

\maketitle
\begin{abstract}
Stateless model checking is a promising technique for testing concurrent programs,
but is vulnerable to exponential explosion of the state space when the test input parameters are too large.
Several partial-order reduction techniques exist for mitigating this explosion,
but even after pruning equivalent interleavings, the state space size for a fixed set of preemption points is unpredictable and often intractable.
%
Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
It avoids concerns of intractable state space size, but suffers from a high rate of false positives.

We present an {\em Iterative Deepening} framework for stateless model checking,
which manages the exploration of many state spaces using different subsets of preemption points.
It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
and it incorporates a data-race analysis to dynamically add new preemption points to verify each data race as buggy or benign.
%
Our evaluation shows this technique is
more effective than single-state-space model checking, both at finding more bugs and at completing more state spaces when no bug exists.

\end{abstract}

\section{Introduction}

% blah blah trite opening sentence
%As parallelism becomes ever more important for achieving high performance in modern-day programs,
%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
or verifying their absence,
%by systematically executing a program along as many distinct interleavings as possible,
by forcing a program to execute each distinct interleaving,
capturing and controlling this nondeterminism in a finite state space.
Unfortunately, these state spaces explode exponentially in the size of the input program.
Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.

% Can I even make a claim this broad to begin with?
However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
For example, CHESS \cite{chess} preemptions only on synchronization operations and library calls, which can miss lock-free shared memory races.
It provides an additional data-race analysis to report any violations of this model;
% TODO: Make sure that there are enough citations for this claim.
however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
%
On the other hand,
% TODO: Find a systematic tester that does this!
SKI \cite{ski}
is able to preempt threads around any shared resource access. Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
%
This work shows how to avoid making this tradeoff decision in advance.

\newcommand\landslideid{\textsc{Landslide-ID}}
We present \landslideid...

\section{Implementation}

Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.

% TODO
TODO

% TODO: make sure to talk about avoiding thrashing

\section{Evaluation}

Evaluation questions:
\begin{enumerate}
	\item Is ID more effective than 
\end{enumerate}

We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.

% TODO: Future work.
% Future work: Add a way to configure even smaller subsets (eg "only mutex_locks called from site X") for cases where mx_lock and mx_unlock alone are still too big. Count the number of kernels for which this was the case.

\section{Related Work}

None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.

% dbug-like systematic testers: dbug, chess, modist

% SKI-like systematic testers: SKI

% other: macemc(???), ETA (message passing only)

% TODO: Mention dbug+PARROT as a motivating argument for our work. The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.

\section{Conclusion}

We are
%overwhelmingly
great. Accept our paper.

% TODO: Publish source of ID wrapper alone, if can't publish landslide itself.
% TODO: Publish log files.

\section{Acknowledgements}

Thanks.

\bibliographystyle{abbrvnat}
\bibliography{citations}{}

\end{document}

### 2015-09-21 pldi-16/paper.tex
write more pldi

@@
 
+% FIXME: decide whether "stateless model checking" or "sys testing"
 \title{Systematic Testing with Data-Race Preemption Points}
@@
 However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
-For example, CHESS \cite{chess} preemptions only on synchronization operations and library calls, which can miss lock-free shared memory races.
+For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
 It provides an additional data-race analysis to report any violations of this model;
@@
 % TODO: Find a systematic tester that does this!
-SKI \cite{ski}
+\textsc{Ski} \cite{ski}
 is able to preempt threads around any shared resource access.
@@
 %
-This work shows how to avoid making this tradeoff decision in advance.
+This paper shows how to avoid making that tradeoff decision in advance.
 
 \newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid...
+We present \landslideid, a framework for {\em iterative deepening} (ID) of preemption points during stateless model checking.
+Named after the analogous technique in chess AI \cite{iterative-deepening-chess},
+
+This paper makes ELEVENTYBILLION contributions:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% TODO
-TODO
+% TODO: Talk about the multiple ways DR PPs are added (alone / in addition)
 
 % TODO: make sure to talk about avoiding thrashing
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{enumerate}
-	\item Is ID more effective than 
+	\item Easy question: Is it good to incorporate DR analysis for new PPs? (Do we find DR-PP-only bugs?)
+	\item How does this compare to static DR detection (how many DRs does ID verify as benign?)
+	\item DRs aside, is ID more effective than the maximal state space? (Do we find bugs in subset spaces, when the max state space would time out?)
+	\item Is ID useful for managing new DR PPs? (Are there some DRs where the PP alone finds the bug, and others where the old PP plus the new DR PP is needed?)
+		% How many DR bugs were found from DR-alone state space, and how many from sema_foo + DR state space?
+What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
+	\item How effective is the free-re-malloc technique at eliminating false positives? (Just turn the optimization off, and measure how many more SSes are generated.)
+
+% Figure out concretely what the data race tricks are that we do, so we can claim them as contributions in the paper. Then ACTUALLY EVALUATE THEM.
+%         - Speculative DR PPs.
+%                 Not a heuristic, rather how to make it work at all to begin with.
+%                 (Cite MS thesis, claim on backwards explorating finding bugs faster)
+%         - Free/re-malloc to eliminate some false positives. See #193.
+%                 Measure how many false positives are eliminated.
+%                 Check, ofc, to make ABSOLUTE SURE, that no bugs missed w/ this trick.
+%                         If there are, it could be because of the implementation
+%                         bug described in #193.
+%         - Using tid/last_call filtering because whole stack traces are too expensive.
+%                 Turn on/off and measure how resulting DR bug #s change.
+%         - Optional: Reprioritizing DRs based on "confirmed" / "suspected"
+%                 Shouldn't be hard just make ID wrapper print "s" or "c"!
+%                 Is it helpful for ID to put priorities on DR PPs?
+%                         Test by inverting the priority and see if fewer buges are found.
+%         // Super optional to talk about. Probably not worth the time.
+%         // - "Too suspicious" (during init/destroy)
+%         //      (Cite eraser, section 2.2)
+
 \end{enumerate}
@@
 Count the number of kernels for which this was the case.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.
+%Indeed, the low state space completion rate in dBug+Parrot's evaluation \cite{parrot} shows that for some tests even the set of all pthread API calls is too many PPs.
+
+% TODO: Address ICB. Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2015-10-02 pldi-16/paper.tex
write pldi related work

@@
 
-% FIXME: decide whether "stateless model checking" or "sys testing"
-\title{Systematic Testing with Data-Race Preemption Points}
+% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
+\newcommand\landslide{\textsc{Landslide}}
+\newcommand\quicksand{\textsc{Quicksand}}
+
+\title{Stateless Model Checking with Data-Race Preemption Points}
 \authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
@@
 Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.
+and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should they exist.
 
 % Can I even make a claim this broad to begin with?
-However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
+However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
 For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
@@
 is able to preempt threads around any shared resource access.
-Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
+Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for all but the most rudimentary test inputs.
 %
-This paper shows how to avoid making that tradeoff decision in advance.
+This work shows how to avoid making that tradeoff decision in advance.
 
-\newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid, a framework for {\em iterative deepening} (ID) of preemption points during stateless model checking.
-Named after the analogous technique in chess AI \cite{iterative-deepening-chess},
+We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
+Named after the analogous technique in chess AI \cite{iterative-deepening-chess}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
+Rather than attempting to search a single state space with every preemption point among a fixed set enabled (e.g., preempting on every pthread API call),
+\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
+It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.
 
-This paper makes ELEVENTYBILLION contributions:
+We evaluated \quicksand~by testing
+% TODO: What's the real number
+{\Large \em 31337}
+student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
+We show that \quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,
+and furthermore that adding data-race preemption points quickly exposes bugs missed by even the ``maximal'' state space of the conventional approach.
+
+% TODO fix dis when rest of paper is done
+This paper makes {\Large \em ELEVENTYBILLION} contributions:
+\begin{enumerate}
+	\item Iterative deepening wrapper
+	\item Data race tricks, including a false-positive reduction optimization that only works in the context of stateless model checking
+	\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
+		% don't forget the maybe_unblock cleverness
+	\item Evaluation (elaborate obv)
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Design}
+
+TODO
+
+Unlike in chess AI, however, adding more preemption points to a completed state space is not necessarily a linear process, as  % blah blah multiple combinations
+
+% define minimal state space and maximal state space
 
@@
 % TODO: make sure to talk about avoiding thrashing
+\subsection{Avoiding thrashing in \quicksand}
+
+(Here the reviewers will wonder if the whole point of the paper was just to build up to that one joke.)
 
@@
 \section{Evaluation}
+
+% TODO: Don't forget to add to quicksand
 
@@
 	\item DRs aside, is ID more effective than the maximal state space? (Do we find bugs in subset spaces, when the max state space would time out?)
+		% Make sure to argue how this shows ID is applicable to e.g. message-passing SMC tools. "Even when DR PPs aren't on the table, you should do our thing."
 	\item Is ID useful for managing new DR PPs? (Are there some DRs where the PP alone finds the bug, and others where the old PP plus the new DR PP is needed?)
@@
 
+%% Notes from jiri meeting. Some overlap with stuff above.
+% Evaluation in 2 major questions
+%         Data races (Sequence this first) vs no-DR-systesting
+%                 Sub questions: Compare to single-pass DRs
+%                         What false positive rate are we avoiding by comparison?
+%                         How often would the DR not even be caught on the 1st execution, requiring backtracks to expose?
+%                                 (don't forget set EXPLORE_BACKWARDS=0 here)
+%         "Control experiment vs ID" (PLDI cares less asbout "finding bugs faster")
+
 \end{enumerate}
@@
 
-None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
+%None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
 
-% dbug-like systematic testers: dbug, chess, modist
+\quicksand~is built upon many established concurrency testing techniques, dating back of course to Verisoft, the original C model checker \cite{verisoft}.
+We use \landslide~as our model checker \cite{landslide},
+% TODO: is DPOR abbreviated earlier in the paper
+which itself implements DPOR \cite{dpor},
+state space estimation \cite{estimation},
+and data race detection \cite{eraser}. Related model checkers can largely be classified based on their treatment of shared-memory thread communication.
 
-% SKI-like systematic testers: SKI
+{\bf Synchronization events only.} CHESS \cite{chess} and \textsc{dBug} \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
+Hence they will miss any crashes resulting from interleaving threads at instruction granularity during a data race.
+CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data race candidates as PPs in future tests.
 
-% other: macemc(???), ETA (message passing only)
+{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist}, and ETA \cite{parallel-dpor}, require thread communication to be restricted to a message-passing API to more effectively test distributed systems.
+This eliminates the need for data race analysis, but restricts the class of programs that can be tested.
+Nevertheless, Iterative Deepening is still applicable to these tools.
+
+{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
+SKI \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
+which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
+% TODO: Make sure you talk about your mutex experiment, or rephrase this.
+SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our mutex-testing experiment.
+However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.
+
+{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
+These are all orthogonal to our technique.
+\textsc{Parrot} \cite{parrot} combines model checking with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
+Finally, Iterative Context Bounding \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable.
+However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data race PPs during a test with ICB.
+
+% TODO: talk about data race detectors???
 
@@
 % TODO: Address ICB. Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.
+
+% IDK if i should mention it, but OOPSLA 2015, protocol based verification of MPI concurrency paper.
+Different verification approach entirely; doesn't suffer exponential explosion but limited to programs with no shared state and MPI communication only
+
+% Probably NOT worth a mention: OOPSLA 2015, stateless model checking of event driven applications.
+Turning timer-driven model on its head and checking single-threaded, but asynch-event-driven programs (i.e.
+device-like signal handlers)
+
+% TODO: Read OOPSLA 2015 "SATcheck, sat-directed SMC for SC/TSO"
 

### 2015-10-02 pldi-16/di.tex
split pldi multiple tex files

--- 2015-09-20 pldi-16/paper.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-% 
-\documentclass[pldi]{sigplanconf-pldi15}
+\section{Design}
 
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+TODO
 
-\begin{document}
+Unlike in chess AI, however, adding more preemption points to a completed state space is not necessarily a linear process, as  % blah blah multiple combinations
 
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
+% define minimal state space and maximal state space
 
-\title{Systematic Testing with Data-Race Preemption Points}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-Stateless model checking is a promising technique for testing concurrent programs,
-but is vulnerable to exponential explosion of the state space when the test input parameters are too large.
-Several partial-order reduction techniques exist for mitigating this explosion,
-but even after pruning equivalent interleavings, the state space size for a fixed set of preemption points is unpredictable and often intractable.
-%
-Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
-It avoids concerns of intractable state space size, but suffers from a high rate of false positives.
-
-We present an {\em Iterative Deepening} framework for stateless model checking,
-which manages the exploration of many state spaces using different subsets of preemption points.
-It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
-and it incorporates a data-race analysis to dynamically add new preemption points to verify each data race as buggy or benign.
-%
-Our evaluation shows this technique is
-more effective than single-state-space model checking, both at finding more bugs and at completing more state spaces when no bug exists.
-
-\end{abstract}
-
-\section{Introduction}
-
-% blah blah trite opening sentence
-%As parallelism becomes ever more important for achieving high performance in modern-day programs,
-%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
-Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
-{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
-or verifying their absence,
-%by systematically executing a program along as many distinct interleavings as possible,
-by forcing a program to execute each distinct interleaving,
-capturing and controlling this nondeterminism in a finite state space.
-Unfortunately, these state spaces explode exponentially in the size of the input program.
-Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.
-
-% Can I even make a claim this broad to begin with?
-However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
-For example, CHESS \cite{chess} preemptions only on synchronization operations and library calls, which can miss lock-free shared memory races.
-It provides an additional data-race analysis to report any violations of this model;
-% TODO: Make sure that there are enough citations for this claim.
-however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
-%
-On the other hand,
-% TODO: Find a systematic tester that does this!
-SKI \cite{ski}
-is able to preempt threads around any shared resource access.
-Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
-%
-This work shows how to avoid making this tradeoff decision in advance.
-
-\newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid...
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% TODO
-TODO
+% TODO: Talk about the multiple ways DR PPs are added (alone / in addition)
 
 % TODO: make sure to talk about avoiding thrashing
+\subsection{Avoiding thrashing in \quicksand}
 
-\section{Evaluation}
-
-Evaluation questions:
-\begin{enumerate}
-	\item Is ID more effective than 
-\end{enumerate}
-
-We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
-
-% TODO: Future work.
-% Future work: Add a way to configure even smaller subsets (eg "only mutex_locks called from site X") for cases where mx_lock and mx_unlock alone are still too big.
-Count the number of kernels for which this was the case.
-
-\section{Related Work}
-
-None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
-
-% dbug-like systematic testers: dbug, chess, modist
-
-% SKI-like systematic testers: SKI
-
-% other: macemc(???), ETA (message passing only)
-
-% TODO: Mention dbug+PARROT as a motivating argument for our work.
-The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.
-
-\section{Conclusion}
-
-We are
-%overwhelmingly
-great. Accept our paper.
-
-% TODO: Publish source of ID wrapper alone, if can't publish landslide itself.
-% TODO: Publish log files.
-
-\section{Acknowledgements}
-
-Thanks.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-
-\end{document}
+(Here the reviewers will wonder if the whole point of the paper was just to build up to that one joke.)

### 2015-10-02 pldi-16/evaluation.tex
split pldi multiple tex files

--- 2015-09-21 pldi-16/paper.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi15}
+\section{Evaluation}
 
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-% FIXME: decide whether "stateless model checking" or "sys testing"
-\title{Systematic Testing with Data-Race Preemption Points}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-Stateless model checking is a promising technique for testing concurrent programs,
-but is vulnerable to exponential explosion of the state space when the test input parameters are too large.
-Several partial-order reduction techniques exist for mitigating this explosion,
-but even after pruning equivalent interleavings, the state space size for a fixed set of preemption points is unpredictable and often intractable.
-%
-Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
-It avoids concerns of intractable state space size, but suffers from a high rate of false positives.
-
-We present an {\em Iterative Deepening} framework for stateless model checking,
-which manages the exploration of many state spaces using different subsets of preemption points.
-It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
-and it incorporates a data-race analysis to dynamically add new preemption points to verify each data race as buggy or benign.
-%
-Our evaluation shows this technique is
-more effective than single-state-space model checking, both at finding more bugs and at completing more state spaces when no bug exists.
-
-\end{abstract}
-
-\section{Introduction}
-
-% blah blah trite opening sentence
-%As parallelism becomes ever more important for achieving high performance in modern-day programs,
-%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
-Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
-{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
-or verifying their absence,
-%by systematically executing a program along as many distinct interleavings as possible,
-by forcing a program to execute each distinct interleaving,
-capturing and controlling this nondeterminism in a finite state space.
-Unfortunately, these state spaces explode exponentially in the size of the input program.
-Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.
-
-% Can I even make a claim this broad to begin with?
-However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
-For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
-It provides an additional data-race analysis to report any violations of this model;
-% TODO: Make sure that there are enough citations for this claim.
-however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
-%
-On the other hand,
-% TODO: Find a systematic tester that does this!
-\textsc{Ski} \cite{ski}
-is able to preempt threads around any shared resource access.
-Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
-%
-This paper shows how to avoid making that tradeoff decision in advance.
-
-\newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid, a framework for {\em iterative deepening} (ID) of preemption points during stateless model checking.
-Named after the analogous technique in chess AI \cite{iterative-deepening-chess},
-
-This paper makes ELEVENTYBILLION contributions:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Implementation}
-
-Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.
-
-% TODO: Talk about the multiple ways DR PPs are added (alone / in addition)
-
-% TODO: make sure to talk about avoiding thrashing
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Evaluation}
+% TODO: Don't forget to add to quicksand
 
@@
 Count the number of kernels for which this was the case.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Related Work}
-
-None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
-
-% dbug-like systematic testers: dbug, chess, modist
-
-% SKI-like systematic testers: SKI
-
-% other: macemc(???), ETA (message passing only)
-
-% TODO: Mention dbug+PARROT as a motivating argument for our work.
-The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.
-%Indeed, the low state space completion rate in dBug+Parrot's evaluation \cite{parrot} shows that for some tests even the set of all pthread API calls is too many PPs.
-
-% TODO: Address ICB. Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusion}
-
-We are
-%overwhelmingly
-great. Accept our paper.
-
-% TODO: Publish source of ID wrapper alone, if can't publish landslide itself.
-% TODO: Publish log files.
-
-\section{Acknowledgements}
-
-Thanks.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-
-\end{document}

### 2015-10-02 pldi-16/intro.tex
split pldi multiple tex files

--- 2015-09-20 pldi-16/paper.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-% 
-\documentclass[pldi]{sigplanconf-pldi15}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-\title{Systematic Testing with Data-Race Preemption Points}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-Stateless model checking is a promising technique for testing concurrent programs,
-but is vulnerable to exponential explosion of the state space when the test input parameters are too large.
-Several partial-order reduction techniques exist for mitigating this explosion,
-but even after pruning equivalent interleavings, the state space size for a fixed set of preemption points is unpredictable and often intractable.
-%
-Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
-It avoids concerns of intractable state space size, but suffers from a high rate of false positives.
-
-We present an {\em Iterative Deepening} framework for stateless model checking,
-which manages the exploration of many state spaces using different subsets of preemption points.
-It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
-and it incorporates a data-race analysis to dynamically add new preemption points to verify each data race as buggy or benign.
-%
-Our evaluation shows this technique is
-more effective than single-state-space model checking, both at finding more bugs and at completing more state spaces when no bug exists.
-
-\end{abstract}
-
 \section{Introduction}
@@
 Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.
+and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should they exist.
 
 % Can I even make a claim this broad to begin with?
-However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
-For example, CHESS \cite{chess} preemptions only on synchronization operations and library calls, which can miss lock-free shared memory races.
+However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
+% TODO: You repeat the following 1.5 sentence in the related work section; maybe you can eliminate this redundancy to save space?
+For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
 It provides an additional data-race analysis to report any violations of this model;
-% TODO: Make sure that there are enough citations for this claim.
 however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
 %
-On the other hand,
-% TODO: Find a systematic tester that does this!
-SKI \cite{ski}
-is able to preempt threads around any shared resource access.
-Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
+On the other hand, SPIN \cite{spin}
+is able to preempt threads around any shared memory access.
+Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for all but the most rudimentary test inputs.
 %
-This work shows how to avoid making this tradeoff decision in advance.
+This work shows how to avoid making that tradeoff decision in advance.
 
-\newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid...
+We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
+Named after the analogous technique in chess AI \cite{iterative-deepening-chess}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
+Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
+\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
+It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.
 
-\section{Implementation}
+We evaluated \quicksand~by testing
+% TODO: What's the real number
+{\Large \em 31337}
+student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
+We show that \quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,
+and furthermore that adding data-race preemption points quickly exposes bugs missed by even the ``maximal'' state space of the conventional approach.
 
-Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.
-
-% TODO
-TODO
-
-% TODO: make sure to talk about avoiding thrashing
-
-\section{Evaluation}
-
-Evaluation questions:
+% TODO fix dis when rest of paper is done
+This paper makes {\Large \em ELEVENTYBILLION} contributions:
 \begin{enumerate}
-	\item Is ID more effective than 
+	\item Iterative deepening wrapper
+	\item Data race tricks, including a false-positive reduction optimization that only works in the context of stateless model checking
+	\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
+		% don't forget the maybe_unblock cleverness
+	\item Evaluation (elaborate obv)
 \end{enumerate}
-
-We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
-
-% TODO: Future work.
-% Future work: Add a way to configure even smaller subsets (eg "only mutex_locks called from site X") for cases where mx_lock and mx_unlock alone are still too big.
-Count the number of kernels for which this was the case.
-
-\section{Related Work}
-
-None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
-
-% dbug-like systematic testers: dbug, chess, modist
-
-% SKI-like systematic testers: SKI
-
-% other: macemc(???), ETA (message passing only)
-
-% TODO: Mention dbug+PARROT as a motivating argument for our work.
-The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.
-
-\section{Conclusion}
-
-We are
-%overwhelmingly
-great. Accept our paper.
-
-% TODO: Publish source of ID wrapper alone, if can't publish landslide itself.
-% TODO: Publish log files.
-
-\section{Acknowledgements}
-
-Thanks.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-
-\end{document}

### 2015-10-02 pldi-16/paper.tex
split pldi multiple tex files

--- 2015-09-20 pldi-16/paper.tex
@@
 
-\title{Systematic Testing with Data-Race Preemption Points}
+% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
+\newcommand\landslide{\textsc{Landslide}}
+\newcommand\quicksand{\textsc{Quicksand}}
+
+\title{Stateless Model Checking with Data-Race Preemption Points}
 \authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
@@
 
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% blah blah trite opening sentence
-%As parallelism becomes ever more important for achieving high performance in modern-day programs,
-%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
-Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
-{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
-or verifying their absence,
-%by systematically executing a program along as many distinct interleavings as possible,
-by forcing a program to execute each distinct interleaving,
-capturing and controlling this nondeterminism in a finite state space.
-Unfortunately, these state spaces explode exponentially in the size of the input program.
-Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should the exist.
+\input{intro}
+\input{di}
+\input{evaluation}
+\input{related}
 
-% Can I even make a claim this broad to begin with?
-However, all stateless model checkers to date are bound by a fixed set of {\em preemption points} (PPs): code locations that define the granularity at which threads interleave.
-For example, CHESS \cite{chess} preemptions only on synchronization operations and library calls, which can miss lock-free shared memory races.
-It provides an additional data-race analysis to report any violations of this model;
-% TODO: Make sure that there are enough citations for this claim.
-however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
-%
-On the other hand,
-% TODO: Find a systematic tester that does this!
-SKI \cite{ski}
-is able to preempt threads around any shared resource access.
-Such fine granularity would automatically check if each data race is a real bug, but makes state space completion intractable for all but the most rudimentary test inputs.
-%
-This work shows how to avoid making this tradeoff decision in advance.
-
-\newcommand\landslideid{\textsc{Landslide-ID}}
-We present \landslideid...
-
-\section{Implementation}
-
-Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.
-
-% TODO
-TODO
-
-% TODO: make sure to talk about avoiding thrashing
-
-\section{Evaluation}
-
-Evaluation questions:
-\begin{enumerate}
-	\item Is ID more effective than 
-\end{enumerate}
-
-We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
-
-% TODO: Future work.
-% Future work: Add a way to configure even smaller subsets (eg "only mutex_locks called from site X") for cases where mx_lock and mx_unlock alone are still too big.
-Count the number of kernels for which this was the case.
-
-\section{Related Work}
-
-None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
-
-% dbug-like systematic testers: dbug, chess, modist
-
-% SKI-like systematic testers: SKI
-
-% other: macemc(???), ETA (message passing only)
-
-% TODO: Mention dbug+PARROT as a motivating argument for our work.
-The number of tests with INF state space, EVEN WITH ONLY PTHREAD API PPs, shows that fixed-in-advance-pp-set needs our work to improve.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Thanks.
+Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
+Thanks to Ji\v{r}\'{i} \v{S}im\v{s}a and the anonymous PLDI reviewers for their helpful comments.
+This work was supported in part by % TODO: funding sources
 

### 2015-10-02 pldi-16/related.tex
split pldi multiple tex files

\section{Related Work}

%None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.

\quicksand~is built upon many established concurrency testing techniques, dating back of course to Verisoft, the original C model checker \cite{verisoft}.
We use \landslide~as our model checker \cite{landslide},
% TODO: is DPOR abbreviated earlier in the paper
which itself implements DPOR \cite{dpor},
state space estimation \cite{estimation},
and data race detection \cite{eraser}. Related model checkers can largely be classified based on their treatment of shared-memory thread communication.

{\bf Synchronization events only.} CHESS \cite{chess} and \textsc{dBug} \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
Hence they will miss any crashes resulting from interleaving threads at instruction granularity during a data race. CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data race candidates as PPs in future tests.

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist}, and ETA \cite{parallel-dpor}, require thread communication to be restricted to a message-passing API to more effectively test distributed systems.
This eliminates the need for data race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is still applicable to these tools.

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
Ski \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
% TODO: Make sure you talk about your mutex experiment, or rephrase this.
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our mutex-testing experiment.
However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.

{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
These are all orthogonal to our technique.
\textsc{Parrot} \cite{parrot} combines model checking with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
Finally, Iterative Context Bounding \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable.
However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data race PPs during a test with ICB.

% TODO: talk about data race detectors???

% Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.

% IDK if i should mention it, but OOPSLA 2015, protocol based verification of MPI concurrency paper. Different verification approach entirely; doesn't suffer exponential explosion but limited to programs with no shared state and MPI communication only

% Probably NOT worth a mention: OOPSLA 2015, stateless model checking of event driven applications. Turning timer-driven model on its head and checking single-threaded, but asynch-event-driven programs (i.e. device-like signal handlers)

% TODO: Read OOPSLA 2015 "SATcheck, sat-directed SMC for SC/TSO"

### 2015-10-02 pldi-16/evaluation.tex
make structure for evaluation

@@
+% TODO: Abbreviate SMC, and fixed-PP approach.
+
 \section{Evaluation}
 
-% TODO: Don't forget to add to quicksand
+% TODO: Don't forget to add to quicksand a feature to print time-when-job-started (ie, tell total time from test start to bug found in a DR state space == job-start time + job-elapsed time)
+% TODO ******BEFORE****** you begin running experimence en masse!
 
-Evaluation questions:
+Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone in other model checkers.
+For example, a single-state-space tool could use data-race candidates during immediately subsequent interleavings, essentially changing the state space on the fly.
+Likewise, a message-passing-only tool could employ Iterative Deepening despite data races being absent from its concurrency model.
+Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
+we also sought to evaluate each technique individually.
+Our evaluation answers the following questions:
 \begin{enumerate}
-	\item Easy question: Is it good to incorporate DR analysis for new PPs? (Do we find DR-PP-only bugs?)
-	\item How does this compare to static DR detection (how many DRs does ID verify as benign?)
-	\item DRs aside, is ID more effective than the maximal state space? (Do we find bugs in subset spaces, when the max state space would time out?)
-		% Make sure to argue how this shows ID is applicable to e.g. message-passing SMC tools. "Even when DR PPs aren't on the table, you should do our thing."
-	\item Is ID useful for managing new DR PPs? (Are there some DRs where the PP alone finds the bug, and others where the old PP plus the new DR PP is needed?)
-		% How many DR bugs were found from DR-alone state space, and how many from sema_foo + DR state space?
-What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
-	\item How effective is the free-re-malloc technique at eliminating false positives? (Just turn the optimization off, and measure how many more SSes are generated.)
+	\item Is combining data-race analysis with stateless model checking productive?
+		\begin{enumerate}
+			\item Do data-race PPs expose new bugs that couldn't be found with single-state-space model checking?
+				% Elaborate later:
+				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
+			\item Do we avoid false positives compared to single-pass?
+				% Explain later as:
+				% How many data-race candidates were verified as benign
+				% Also Include:
+				% How many false positives does the free-re-malloc technique suppress?
+			% TODO: This one's optional. You can give up on it.
+			\item Do we avoid false negatives compared to single-pass?
+				% Explain later as: How many data-race bugs required backtracking to be identified as candidates in the first place
+				%(don't forget set EXPLORE_BACKWARDS=0 here)
+		\end{enumerate}
+	\item Is Iterative Deepening better than the single-state-space model?
+		\begin{enumerate}
+			\item Does Iterative Deepening find bugs
+				% in subset state spaces
+				faster than the single-state-space approach, even without data races?
+			\item Does Iterative Deepening help test data-race PPs?
+				% Are there some DRs where the PP alone finds the bug, while the superset SS cannot, and some vice versa?
+				% What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
+		\end{enumerate}
+\end{enumerate}
+
+\subsection{Evaluating Data-Race PPs}
+
+\subsection{Evaluating Iterative Deepening}
 
@@
 %         - Using tid/last_call filtering because whole stack traces are too expensive.
+%                 Moderately optional, 1st priority since theoretically interesting:
 %                 Turn on/off and measure how resulting DR bug #s change.
@@
 
-%% Notes from jiri meeting. Some overlap with stuff above.
-% Evaluation in 2 major questions
-%         Data races (Sequence this first) vs no-DR-systesting
-%                 Sub questions: Compare to single-pass DRs
-%                         What false positive rate are we avoiding by comparison?
-%                         How often would the DR not even be caught on the 1st execution, requiring backtracks to expose?
-%                                 (don't forget set EXPLORE_BACKWARDS=0 here)
-%         "Control experiment vs ID" (PLDI cares less asbout "finding bugs faster")
-
-\end{enumerate}
-
 We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.

### 2015-10-02 pldi-16/evaluation.tex
some touchups and talk about test suite

@@
 \end{enumerate}
+\subsection{Test Suite}
+% TODO: Maybe say how many lines of code total? How many lines on average per P2/pintos? (careful with pintos; lots of basecode)
+Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
+%across the Spring and Fall 2014 and Spring 2015 semesters;
+and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
+%
+The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condvars, semaphores, and r/w locks;
+all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
+%
+The Pintos kernel project
+involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
+using provided bare-bones mutex, context-switch, and virtal memory implementations
+\cite{pintos}.
+
+% TODO: List test programs that were run.
 

### 2015-10-02 pldi-16/intro.tex
some touchups and talk about test suite

@@
 
-We evaluated \quicksand~by testing
-% TODO: What's the real number
-{\Large \em 31337}
-student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
+We evaluated \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
 We show that \quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,

### 2015-10-02 pldi-16/related.tex
some touchups and talk about test suite

@@
 
-\quicksand~is built upon many established concurrency testing techniques, dating back of course to Verisoft, the original C model checker \cite{verisoft}.
-We use \landslide~as our model checker \cite{landslide},
+\quicksand~is built upon many established concurrency testing techniques, dating back
+% of course
+to Verisoft, the original C model checker \cite{verisoft}.
+Our model checker \landslide~\cite{landslide}
 % TODO: is DPOR abbreviated earlier in the paper
-which itself implements DPOR \cite{dpor},
+itself implements DPOR \cite{dpor},
 state space estimation \cite{estimation},
-and data race detection \cite{eraser}. Related model checkers can largely be classified based on their treatment of shared-memory thread communication.
+and data-race detection \cite{eraser}.
+We classify related model checkers largely based on their treatment of shared-memory thread communication.
 
 {\bf Synchronization events only.} CHESS \cite{chess} and \textsc{dBug} \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
-Hence they will miss any crashes resulting from interleaving threads at instruction granularity during a data race.
-CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data race candidates as PPs in future tests.
+Hence they will miss any crashes that require interleaving threads at instruction granularity during a data race.
+CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
 
 {\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist}, and ETA \cite{parallel-dpor}, require thread communication to be restricted to a message-passing API to more effectively test distributed systems.
-This eliminates the need for data race analysis, but restricts the class of programs that can be tested.
+This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
 Nevertheless, Iterative Deepening is still applicable to these tools.
@@
 Finally, Iterative Context Bounding \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable.
-However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data race PPs during a test with ICB.
+However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
 

### 2015-10-10 pldi-16/di.tex
talk about dr pps

@@
 
+% TODO: Overview
+
+\subsection{Managing multiple state spaces}
+
 Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.
 
-% TODO: Talk about the multiple ways DR PPs are added (alone / in addition)
+{\bf Avoiding thrashing in \quicksand.} % TODO
 
-% TODO: make sure to talk about avoiding thrashing
-\subsection{Avoiding thrashing in \quicksand}
+% TODO: decide which subsection should come first for better flow
 
-(Here the reviewers will wonder if the whole point of the paper was just to build up to that one joke.)
+\subsection{Data-race preemption points}
+
+Intuitively, % TODO describe what we're trying to achieve
+
+When \landslide~detects a data race, it produces a report for each of the two memory accesses involved in the race.
+Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
+(For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
+Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do as often as each shared memory access.
+Instead, \landslide~qualifies the PC with
+(a) the current thread ID and
+(b) the most recent {\tt call} instruction.
+% (a crude approximation of a stack trace)
+% which are much cheaper, as we carry them around all the time already
+Note that we do {\em not} qualify data races by {\em memory addresses},
+which change across thread interleavings.
+% especially when malloc is involved.
+Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
+
+% TODO: make dont-filter-dr-by-address example.
+
+When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
+a ``small'' job to preempt on the racing instruction only,
+and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
+%
+Hence each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
+Rather than adding a maximal job with both new PPs at once, we prefer to add smaller jobs which have a higher chance of completing in time.
+If those state spaces are bug-free, they will in turn add the maximal job later.
+%
+The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance: while the big job may not complete in time, the small job might miss the data race entirely if the original PPs were required to expose it.
+In practice, we have observed some bugs found quickly by the small job, and other bugs missed by the small job found eventually by the big job.
+% which justifies this design
+
+% TODO: make new-dr-jobs diagram
+
+%                                  (2)          landslide         landslide
+%  landslide                   "Preempt when-->  o  PPs=[foo]      o  PPs=[mutex_lock,
+%    o  PPs=[mutex_lock]        PC = FOO"       / \               / \      foo]
+%   / \                              |         o   o             o   o
+%  o  /\                             |                          /|\  |\
+%    o  o                                                      o o o o o
+%                             __QUICKSAND__
+%    |       (1)             |             |           (3)
+%    \  "Hey, an access      |  job1       | ----"Preempt when -->
+%     '- at FOO raced w/ --> |        job2 |      PC = BAR"
+%        one at BAR!         |    job4     |
+%                            |        job70|      [ same pictures for s/foo/bar/]
+%                            | job8        |
+%                            '-------------'
+%                maybe show some internal structore of QS as well

### 2015-10-10 pldi-16/evaluation.tex
talk about dr pps

@@
 				% Explain later as: How many data-race bugs required backtracking to be identified as candidates in the first place
-				%(don't forget set EXPLORE_BACKWARDS=0 here)
+				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
+				% TODO: And disable false-neg malloc-free technique
+				% TODO: And also disable the confirmed/suspected thing
+				% (where mem.c waits for reorder observed before
+				% messaging the latter half of the DR to QS)
 		\end{enumerate}
@@
 
+In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
+
+{\bf Finding bugs missed by SSS-MC.}
+% TODO describe
+
+{\bf Avoiding false positive data-race candidates.}
+% TODO describe
+
+{\bf Avoiding false negative data-race bugs.}
+Some pairs of unprotected shared-memory accesses may be hidden in some control flow path that requires a nondeterministic preemption to be executed.
+In such cases, a single-pass data-race detector
+%could not achieve the coverage necessary
+would fail
+to identify the access pair as a data-race candidate to begin with.
+%
+We counted how many such data-races, used as PPs, would lead to \landslide~finding new bugs,
+thereby making them {\em false negatives} of the single-pass approach.
+% TODO: Put a figure here giving an example of where e.g. a data race only shows up during the contention path of a mutex.
+We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).
+%
+To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
+For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must, rather than waiting until future interleavings to confirm them (see Section \ref{sec:free-re-malloc}).
+
 \subsection{Evaluating Iterative Deepening}

### 2015-10-10 pldi-16/related.tex
talk about dr pps

@@
 
-{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist}, and ETA \cite{parallel-dpor}, require thread communication to be restricted to a message-passing API to more effectively test distributed systems.
+{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist} \cite{modist}, and ETA \cite{dbug-retreat}, limits thread communication to a message-passing API to more effectively test distributed systems.
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
@@
 {\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
-Ski \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
+{\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
 which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
@@
 % TODO: talk about data race detectors???
+% eg Scalable Race Detection for Android Applications -- uses domain specific heuristics to filter out false positives
 

### 2015-10-10 pldi-16/di.tex
tighten up prior commit and write a conclusion

@@
 
+\subsection{Filtering false positive data races}
+
+% TODO: proof goes here
+% TODO: example codez with malloc re free go here
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Managing multiple state spaces}
+
+\quicksand~uses a workqueue model
+...
+priorities... data race priorities...
 
@@
 
-When \landslide~detects a data race, it produces a report for each of the two memory accesses involved in the race.
+When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
 Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
 (For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
-Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do as often as each shared memory access.
+Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
 Instead, \landslide~qualifies the PC with
@@
 % which are much cheaper, as we carry them around all the time already
-Note that we do {\em not} qualify data races by {\em memory addresses},
-which change across thread interleavings.
+Note that we do {\em not} qualify data races by the shared memory address,
+which can change based on different interleavings of previous code
+(for example, depending on the result of {\tt malloc()}).
 % especially when malloc is involved.
-Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
-% TODO: make dont-filter-dr-by-address example.
+% TODO: decide whether to make dont-filter-dr-by-address example.
+%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
 
@@
 %
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance: while the big job may not complete in time, the small job might miss the data race entirely if the original PPs were required to expose it.
-In practice, we have observed some bugs found quickly by the small job, and other bugs missed by the small job found eventually by the big job.
+The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
+while the big job risks not completing in time,
+the small job risks missing the data race entirely if the original PPs were required to expose it.
+In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
 % which justifies this design

### 2015-10-10 pldi-16/paper.tex
some touchups and talk about test suite
tighten up prior commit and write a conclusion

@@
 \newcommand\quicksand{\textsc{Quicksand}}
+
+% TODO: Finalize these numbers.
+\newcommand\numthrlibs{{\Large \bf NUMP2S}}
+\newcommand\numpintoses{{\Large \bf NUMPINTOSES}}
+\newcommand\numstudence{{\Large \bf NUMSTUDENCE}} % total pintoses plus p2s
 
@@
 
-% TODO: Publish source of ID wrapper alone, if can't publish landslide itself.
-% TODO: Publish log files.
+We have presented Iterative Deepening, a new technique for automating the choice of preemption points (PPs) during stateless model checking, and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
+\quicksand~incorporates \landslide's data-race analysis to create new PPs tailored specifically to the program under test,
+and manages multiple \landslide~instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
+
+Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
+We avoid all false-positive data race reports by verifying the absence of bugs in their associated state spaces, and find crashes caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
+Compared to SSS-MC, we find bugs faster in subset state spaces when the maximal state space is too large to explore, and find new bugs with data-race PPs that the maximal state space would not expose at all.
+
+Although \landslide~itself is not publicly available due to its dependency on the non-free simulator {\sc Simics}, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
+We have also posted the log files from all our experiments.
+They can be found here:
+
+\url{https://github.com/bblum/landslide} % TODO: put QS in its own repository
+
+\url{https://github.com/bblum/landslide} % TODO: put final logs somewhere
 
@@
 Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
+Thanks to Wind River Simics for %TODO
 Thanks to Ji\v{r}\'{i} \v{S}im\v{s}a and the anonymous PLDI reviewers for their helpful comments.

### 2015-10-11 pldi-16/evaluation.tex
get average LOC of p2s

@@
 \cite{pintos}.
+% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
+% All numbers, obtained with:
+% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
+% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
+% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
+% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
+% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
+% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
+Though not ``real world'' programs, both projects are quite large: % maybe "complex"?
+the P2s average 1807 lines of C and x86 assembly (stddev 489.5),
+% Pintos SLOC stats: TODO
+and the Pintoses average {\bf 9999999} % TODO
 

### 2015-10-13 pldi-16/evaluation.tex
nail down contributes and work on eval description a bit

@@
 \section{Evaluation}
-
-% TODO: Don't forget to add to quicksand a feature to print time-when-job-started (ie, tell total time from test start to bug found in a DR state space == job-start time + job-elapsed time)
-% TODO ******BEFORE****** you begin running experimence en masse!
 
@@
 				% How many data-race candidates were verified as benign
+				% But to be fair, you have to count how many DRs are reported as "couldn't test these, check yourself" at the end.
 				% Also Include:
@@
 
-% TODO: List test programs that were run.
+\newcommand\mxtest{\texttt{mx\_test}}
+\newcommand\tej{\texttt{thr\_exit\_join}}
+\newcommand\bct{\texttt{broadcast}}
+\newcommand\paraguay{\texttt{paraguay}}
+\newcommand\paradise{\texttt{paradise\_lost}}
+\newcommand\rwl{\texttt{rwl\_test}}
+We tested P2s with 6 multithreaded programs:
+% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
+% each tailored to exercise a different part of the P2 project
+\mxtest, for locking algorithm correctness, \tej, a test of thread lifecycle, \bct~and \paraguay, for condvars, \paradise~for semaphores, and \rwl~for r/w locks.
+For \mxtest~we enabled \landslide's special lock-testing option, which changes the lock-set tracking to consider accesses within {\tt mutex\_lock} and {\tt mutex\_unlock} as data-race candidates.
+(Normally, they are considered protected by the very lock they implement: assuming the locks are already correct to enable productive data-race analysis on the rest of the codebase.)
+\newcommand\prisema{\texttt{priority\_sema}}
+\newcommand\waitsimple{\texttt{wait\_simple}}
+% TODO: Add more test cases
+We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle. % TODO talk more about these.
+In total, this test suite comprises {\large \bf 99999} % TODO
+unique state spaces.
+
+All tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+\quicksand~tests were given 10 CPUs for 1 hour each.
+To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs enabled in advance (and no data-race PPs).
+% TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
+Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we leveled the playing field by giving each control 1 CPU for 10 hours and dividing all associated times by 10 (simulating perfect parallelism).
 
@@
 
+\begin{table}[t]
+	\begin{tabular}{l|l|l}
+			& QS bugs & SSS-MC bugs \\
+		\hline
+		\mxtest & eg 1000 & eg 0 \\
+		\bct & & \\
+		etc... & & \\
+		\hline
+		Total & & \\
+	\end{tabular}
+	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom {\bf \large 9999999})}
+	\label{tab:allbugs}
+\end{table}
+
+\begin{table}[t]
+	\small
+	\begin{tabular}{l|l|l||l|l}
+	& QS bug & \begin{tabular}{c} SSS-MC \\ completed\end{tabular}
+	& QS bug & \begin{tabular}{c}SSS-MC \\ timeout \end{tabular} \\
+		\hline
+		\mxtest & e.g. 5 & 10 & 0 & 0 \\
+		\bct & & & & \\
+		etc... & & & & \\
+		\hline
+		Total & & & & \\
+	\end{tabular}
+	\caption{Bugs requiring data-race PPs to expose, found by \quicksand but missed by the single-state-space approach.}
+	\label{tab:drbugs}
+\end{table}
+
 {\bf Finding bugs missed by SSS-MC.}
-% TODO describe
+Though state-of-the-art MCs preempt only on synchronization events, many serious concurrency bugs are caused by data races leading to corrupted shared state.
+In Table~\ref{tab:drbugs} we count ...... % TODO
+
+In Table~\ref{tab:allbugs} .... %TODO
 
 {\bf Avoiding false positive data-race candidates.}
-% TODO describe
+% Though we mechanically verify whether each data race candidate leads to a bug, each new PP can increase combinatorially..... obviously wish to avoid...
 
@@
 Count the number of kernels for which this was the case.
+
+\subsection{Discussion} % and future work?
+
+% TODO: Mention that SSSMC control never FAB in mutex_test whatsoever, while DR PPs enabled FABs.

### 2015-10-13 pldi-16/intro.tex
nail down contributes and work on eval description a bit

@@
 
-% TODO fix dis when rest of paper is done
-This paper makes {\Large \em ELEVENTYBILLION} contributions:
+This paper's contributions are as follows:
 \begin{enumerate}
-	\item Iterative deepening wrapper
-	\item Data race tricks, including a false-positive reduction optimization that only works in the context of stateless model checking
-	\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
-		% don't forget the maybe_unblock cleverness
-	\item Evaluation (elaborate obv)
+	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation;
+	\item A new tactic for eliminating false-positive data race reports, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
+	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
+	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches.
 \end{enumerate}

### 2015-10-13 pldi-16/paper.tex
get average LOC of p2s
random stuff

@@
 \newcommand\quicksand{\textsc{Quicksand}}
+\newcommand\simics{\textsc{Simics}}
 
 % TODO: Finalize these numbers.
-\newcommand\numthrlibs{{\Large \bf NUMP2S}}
+\newcommand\numthrlibs{79}
 \newcommand\numpintoses{{\Large \bf NUMPINTOSES}}
@@
 
-We are
-%overwhelmingly
-great. Accept our paper.
+%We are great. Accept our paper.
 
@@
 
-Although \landslide~itself is not publicly available due to its dependency on the non-free simulator {\sc Simics}, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
+Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
 We have also posted the log files from all our experiments.
@@
 Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
-Thanks to Wind River Simics for %TODO
+Thanks to Wind River for the use of their simulator \simics, upon which \landslide~is built.
 Thanks to Ji\v{r}\'{i} \v{S}im\v{s}a and the anonymous PLDI reviewers for their helpful comments.

### 2015-10-13 pldi-16/related.tex
random stuff

@@
 
+\subsection{Stateless Model Checking}
+
 {\bf Synchronization events only.} CHESS \cite{chess} and \textsc{dBug} \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
@@
 % TODO: Make sure you talk about your mutex experiment, or rephrase this.
-SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our mutex-testing experiment.
+SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
 However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.
@@
 However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+
+\subsection{Data Race Detection}
+
+blah blah blah eraser
+
+blah blah blah threadsanitizer, racePRO, etc
+
+\subsection{Other Concurrency Testing Approaches} % TODO: Well?
+
+blah blah pldi'15 symbiosis DSP
 

### 2015-10-16 pldi-16/di.tex
nail down contributes and work on eval description a bit
more writin'

@@
 Unlike in chess AI, however, adding more preemption points to a completed state space is not necessarily a linear process, as  % blah blah multiple combinations
+
+% TODO: Introduce abbreviation for SSS-MC here
 
@@
 
-% TODO: Overview
+\subsection{Landslide}
+\label{sec:landslide}
+
+We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for data-race analysis~\cite{eraser}.
+% optional: cite tsan above
+\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation \cite{estimation}, and data-race detection.
+It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) similarly to Fair-Bounded Search \cite{bpor}.
+% this line can be cut if space is needed
+It can test both user- and kernel-level code, though the latter requires user annotations and is limited to timer-driven nondeterminism.
+Its bug-detection metrics include assertion failures, deadlock, segfaults, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock checker.
+
+{\bf Terminology.}
+Importantly, we distinguish data-race {\em candidates} and data-race {\em bugs}.
+Because data-race analysis is false-positive-prone, Landslide reports unprotected access pairs separately from the problems listed above, calling the pair a {\em data race candidate}.
+Should a future interleaving, preempting threads during those accesses, expose e.g. a crash or deadlock, then we report a {\em data-race bug}.
+
+{\bf Data races in lock implementations.}
+Prior work data race tools recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself \cite{tsan}.
+Assuming the locks are already correct enables productive data-race analysis on the rest of the codebase.
+With testing limited to one execution, even if you wished to test for lock bugs, data-race analysis would still uselessly flag every single access pair in the lock implementation, requiring human attention to verify.
+However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
+Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
+(Accesses from other sync functions, such as {\tt cond\_wait()}, would be either included already, or protected by an internal mutex.)
+% TODO: Add a code sample showing how the lockset boundaries change? If there's room?
 
@@
 
-Landslide provides estimated completion time using the Recursive Estimator algorithm in \cite{estimation}.
-
 {\bf Avoiding thrashing in \quicksand.} % TODO
@@
 
-Intuitively, % TODO describe what we're trying to achieve
+% Intuitively, % TODO describe what we're trying to achieve % do this in design section
 
@@
 %                maybe show some internal structore of QS as well
+
+% TODO: Decide whether to discuss the fact that we don't add the 2nd of a "suspected" dr pair until it is confirmed.

### 2015-10-16 pldi-16/evaluation.tex
random stuff
more writin'

@@
 \begin{enumerate}
-	\item Is combining data-race analysis with stateless model checking productive?
+	\item Does Iterative Deepening find bugs faster than SSS-MC in subset state spaces, even without data-race PPs?
+	\item Is combining data-race analysis with MC productive?
 		\begin{enumerate}
-			\item Do data-race PPs expose new bugs that couldn't be found with single-state-space model checking?
+			\item Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
 				% Elaborate later:
 				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
-			\item Do we avoid false positives compared to single-pass?
+			\item Do we avoid false positives compared to a single-execution data race analysis?
 				% Explain later as:
@@
 				% How many false positives does the free-re-malloc technique suppress?
+				% TODO: If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
 			% TODO: This one's optional. You can give up on it.
 			\item Do we avoid false negatives compared to single-pass?
-				% Explain later as: How many data-race bugs required backtracking to be identified as candidates in the first place
 				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
@@
 				% messaging the latter half of the DR to QS)
-		\end{enumerate}
-	\item Is Iterative Deepening better than the single-state-space model?
-		\begin{enumerate}
-			\item Does Iterative Deepening find bugs
-				% in subset state spaces
-				faster than the single-state-space approach, even without data races?
-			\item Does Iterative Deepening help test data-race PPs?
-				% Are there some DRs where the PP alone finds the bug, while the superset SS cannot, and some vice versa?
-				% What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
 		\end{enumerate}
@@
 \mxtest, for locking algorithm correctness, \tej, a test of thread lifecycle, \bct~and \paraguay, for condvars, \paradise~for semaphores, and \rwl~for r/w locks.
-For \mxtest~we enabled \landslide's special lock-testing option, which changes the lock-set tracking to consider accesses within {\tt mutex\_lock} and {\tt mutex\_unlock} as data-race candidates.
-(Normally, they are considered protected by the very lock they implement: assuming the locks are already correct to enable productive data-race analysis on the rest of the codebase.)
+For \mxtest~we enabled \landslide's mutex-testing option described in \sect{\ref{sec:landslide}}.
+%, which changes the lock-set tracking to consider accesses within {\tt mutex\_lock} and {\tt mutex\_unlock} as data-race candidates.
+%(Normally, they are considered protected by the very lock they implement: assuming the locks are already correct to enable productive data-race analysis on the rest of the codebase.)
 \newcommand\prisema{\texttt{priority\_sema}}
@@
 Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we leveled the playing field by giving each control 1 CPU for 10 hours and dividing all associated times by 10 (simulating perfect parallelism).
-
-\subsection{Evaluating Data-Race PPs}
-
-In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
 
@@
 	\end{tabular}
-	\caption{Bugs requiring data-race PPs to expose, found by \quicksand but missed by the single-state-space approach.}
+	\caption{Bugs requiring data-race PPs to expose, found by \quicksand~but missed by the single-state-space approach.}
 	\label{tab:drbugs}
 \end{table}
+
+\subsection{Evaluating Iterative Deepening}
+
+asdfadsfasdf
+
+\subsection{Evaluating Data-Race PPs}
+
+In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
 
@@
 To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
-For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must, rather than waiting until future interleavings to confirm them (see Section \ref{sec:free-re-malloc}).
-
-\subsection{Evaluating Iterative Deepening}
+For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must, rather than waiting until future interleavings to confirm them (see \sect{\ref{sec:free-re-malloc}}).
 
@@
 
-We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
+\subsection{Discussion} % and future work?
+
+% TODO: Mention that SSSMC control never FAB in mutex_test whatsoever, while DR PPs enabled FABs.
+
+%We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
 
@@
 
-\subsection{Discussion} % and future work?
-
-% TODO: Mention that SSSMC control never FAB in mutex_test whatsoever, while DR PPs enabled FABs.
+% Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-10-16 pldi-16/intro.tex
more writin'

@@
 
+% TODO: ttuttle says this is too much of a jump -- that exactly what "subsets" means is not well defined (i.e, lock vs unlock, while NOT addressing the within/without_function problem)
+
 We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
@@
 	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation;
-	\item A new tactic for eliminating false-positive data race reports, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
+	\item A new tactic for eliminating one class of false-positive data races, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
 	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?

### 2015-10-16 pldi-16/paper.tex
more writin'

@@
 \newcommand\simics{\textsc{Simics}}
+\newcommand{\sect}[1]{\S #1}
 
@@
 
-We have presented Iterative Deepening, a new technique for automating the choice of preemption points (PPs) during stateless model checking, and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
+We have presented \quicksand, an implementation of Iterative Deepening, a new technique for automating the choice of PPs during stateless model checking.
+% , and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
 \quicksand~incorporates \landslide's data-race analysis to create new PPs tailored specifically to the program under test,
@@
 Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
-We avoid all false-positive data race reports by verifying the absence of bugs in their associated state spaces, and find crashes caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
-Compared to SSS-MC, we find bugs faster in subset state spaces when the maximal state space is too large to explore, and find new bugs with data-race PPs that the maximal state space would not expose at all.
+We dramatically reduce false-positive data race reports,
+%by verifying the absence of bugs in their associated state spaces
+and find crashes caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
+Compared to SSS-MC, we find bugs faster in smaller subset state spaces,
+%when the maximal state space is too large to explore
+and find new bugs with data-race PPs that the maximal state space would not expose at all.
 

### 2015-10-16 pldi-16/related.tex
more writin'

@@
 to Verisoft, the original C model checker \cite{verisoft}.
-Our model checker \landslide~\cite{landslide}
-% TODO: is DPOR abbreviated earlier in the paper
-itself implements DPOR \cite{dpor},
-state space estimation \cite{estimation},
-and data-race detection \cite{eraser}.
-We classify related model checkers largely based on their treatment of shared-memory thread communication.
+%Our model checker
+\landslide~\cite{landslide} itself implements many techniques from prior work (\sect{\ref{sec:landslide}}).
+%itself implements DPOR \cite{dpor},
+%state space estimation \cite{estimation},
+%and data-race detection \cite{eraser}.
+We compare related model checkers largely based on their treatment of shared-memory thread communication.
 
@@
 
-{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist} \cite{modist}, and ETA \cite{dbug-retreat}, limits thread communication to a message-passing API to more effectively test distributed systems.
+{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist} \cite{modist}, and ETA \cite{dbug-retreat}, limit thread communication to a message-passing API to more effectively test distributed systems.
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
@@
 SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.
+However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL.
+%so cannot check implementations directly.
 

### 2015-10-16 pldi-16/di.tex
Do some more stuff and write about portend

@@
 \subsection{Filtering false positive data races}
+
+{\bf Terminology.}
+Importantly, we distinguish data-race {\em candidates} and data-race {\em bugs}.
+Because data-race analysis is false-positive-prone, Landslide reports unprotected access pairs separately from the problems listed above, calling the pair a {\em data race candidate}.
+Should a future interleaving, preempting during those accesses, cause a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
+Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
+If they cannot be reordered at all, it is a {\em false positive}.
+For brevity, we classify benign data races as a subset of false positives.
 
@@
 
-{\bf Terminology.}
-Importantly, we distinguish data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is false-positive-prone, Landslide reports unprotected access pairs separately from the problems listed above, calling the pair a {\em data race candidate}.
-Should a future interleaving, preempting threads during those accesses, expose e.g. a crash or deadlock, then we report a {\em data-race bug}.
-
 {\bf Data races in lock implementations.}
@@
 priorities... data race priorities...
+
+% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
 

### 2015-10-16 pldi-16/evaluation.tex
Do some more stuff and write about portend

@@
 	\item Does Iterative Deepening find bugs faster than SSS-MC in subset state spaces, even without data-race PPs?
+			% Probably not... ICB is state of the art here.
+			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
 	\item Is combining data-race analysis with MC productive?
@@
 			\item Do we avoid false positives compared to a single-execution data race analysis?
+				% TODO: Argue:
+				% It is fair to compare multiple pass DR-analysis under Landslide against just a single execution because prior work DR detectors, being not integrated with a MC, are not intended to uncover different results under subsequent runs.
+				% Define a "stress tester" as a class of bug detectors where they are intended to [...]
+				% Maybe say: Do there exist any stress-tester DR detectors where they are intended to produce new results under reruns?
+
 				% Explain later as:
@@
 			% TODO: This one's optional. You can give up on it.
-			\item Do we avoid false negatives compared to single-pass?
+			\item Do we find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
 				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
@@
 {\bf Avoiding false negative data-race bugs.}
+\label{sec:eval-falseneg}
 Some pairs of unprotected shared-memory accesses may be hidden in some control flow path that requires a nondeterministic preemption to be executed.

### 2015-10-16 pldi-16/related.tex
Do some more stuff and write about portend

@@
 
-%None. We are the first ever to explore this area, and of course expect the Turing Award for our groundbreaking insights.
+% TODO: ", and related data race analyses based on ????"
 
-\quicksand~is built upon many established concurrency testing techniques, dating back
+\subsection{Stateless Model Checking}
+
+We build upon many established model-checking techniques, dating back
 % of course
 to Verisoft, the original C model checker \cite{verisoft}.
+%, and Eraser, the original data race detector \cite{eraser}.
 %Our model checker
-\landslide~\cite{landslide} itself implements many techniques from prior work (\sect{\ref{sec:landslide}}).
+%\landslide~\cite{landslide} itself implements many techniques from prior work (\sect{\ref{sec:landslide}}).
 %itself implements DPOR \cite{dpor},
@@
 %and data-race detection \cite{eraser}.
-We compare related model checkers largely based on their treatment of shared-memory thread communication.
-
-\subsection{Stateless Model Checking}
+We compare related model checkers by their treatment of shared-memory thread communication.
 
@@
 {\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
+% TODO: Be proud of them instead of being dismissive.
 {\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
@@
 
-blah blah blah eraser
+blah blah blah threadsanitizer, racePRO, etc
 
-blah blah blah threadsanitizer, racePRO, etc
+\newcommand\portend{\textsc{Portend}}
+\portend~\cite{portend} is the most closely related work we have found.
+It performs a single-pass analysis as normal, then explores a limited state space to classify candidates in a taxonomy of likely severity.
+Compared to us, \portend~can detect non-failing races which nevertheless cause suspiciously different program output,
+while we depend on directly detecting failures.
+It also uses symbolic execution to test input nondeterminism along with schedule nondeterminism,
+while we explore the latter only.
+However, \portend~does not perform MC-like exploration in advance of knowing data-race locations,
+which we showed in \sect{\ref{sec:eval-falseneg}} can miss some bugs.
+It is also limited to the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.
+Future work could combine the two approaches, using \landslide~to produce new traces for \portend~to classify, or using \portend's analysis to inform \quicksand's state space priorities.
 

### 2015-10-16 pldi-16/related.tex
rephrase

@@
 while we explore the latter only.
-However, \portend~does not perform MC-like exploration in advance of knowing data-race locations,
-which we showed in \sect{\ref{sec:eval-falseneg}} can miss some bugs.
+However, \portend~does not explore alternate interleavings in advance of knowing data-race locations,
+which is necessary to expose some bugs (\sect{\ref{sec:eval-falseneg}}).
 It is also limited to the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.

### 2015-10-17 pldi-16/di.tex
write about free-re-malloc and add figure 1

@@
+% TODO: get latex to put this on page 2 somehow
+\begin{figure*}[t]
+	\includegraphics[width=0.15\textwidth]{tree0.pdf}
+	\includegraphics[width=0.225\textwidth]{tree1.pdf}
+	\includegraphics[width=0.25\textwidth]{tree2.pdf}
+	\includegraphics[width=0.35\textwidth]{tree3.pdf}
+	\caption{Iterative Deepening visualization.
+		The minimal state space includes only mandatory thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
+Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone, {\tt mutex\_unlock} alone, or both together.
+Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
+Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
+}
+	% TODO: reference this label in first few paragraphs of design
+	\label{fig:id}
+\end{figure*}
 \section{Design}
@@
 % define minimal state space and maximal state space
-
-\subsection{Filtering false positive data races}
-
 {\bf Terminology.}
@@
 For brevity, we classify benign data races as a subset of false positives.
+
+\subsection{Verifying data races as buggy or benign}
+
+\subsection{Suppressing malloc-related false positives}
+
+% TODO: Understand what prior work means by 'happens before'. It's different from our MUST happen before relation.
+% TODO: Understand how it relates to this. Would it scoop us? Or are we outside their model?
+
+We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.
+Figure~\ref{fig:free-re-malloc} shows a common code pattern and interleaving which can expose such behavior.
+If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 6 will be flagged as a data race.
+To the human eye, this is obviously a false positive: reordering lines 4-6 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
+Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 4.
+
+\begin{figure}[t]
+\begin{tabular}{rll}
+	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \};}} \\
+	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \};~~~~~~~~~~}} \\
+	\\
+	& Thread 1 & Thread 2 \\
+	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	3 & & \texttt{// x's memory recycled} \\
+	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(...);} \\
+	5 & & \texttt{publish(y);} \\
+	6 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
+\end{tabular}
+\caption{A common pattern with {\tt malloc()} that produces false positive data race candidates.}
+\label{fig:free-re-malloc}
+\end{figure}
+
+% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
+By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
+
+It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
+\landslide~implements this check by adding a generation counter to its heap state tracking:
+each allocation is given a unique ID, and when evaluating whether two heap accesses can race, their generation counters must match.
+
+However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
+Consider the more unusual program in Figure~\ref{fig:frm-bug}:
+Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
+Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
+
+\begin{figure}[t]
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+	1 & \texttt{publish(x);} & \\
+	2 & & \texttt{x2 = get\_published\_x();} \\
+	3 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	4 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	5 & & \texttt{// x's memory recycled} \\
+	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(...);} \\
+	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+\end{tabular}
+\caption{blah blah}
+\label{fig:frm-bug}
+\end{figure}
+
+% TODO: Figure out why they claim "Happens before produces NO false positives, only benign races".
+% It seems impossible.
+% But if true, it means either (a) they can somehow identify FRM DRs on the 1st pass, not needing to replay,
+% or (b) reuse of memory by malloc is somehow outside their concurrency model.
+Note that while replay-oriented data race tools can already classify free-re-malloc candidates as false positives % benign? (not benign. they are ghosts. not even real.)
+\cite{recordreplaydrs},
+they require replaying the threads in a new interleaving.
+Moreover, \cite{portend} argues that accurate classification may require re-executing many times,
+according to many pre- and post-race sequences,
+which is tantamount to adding a new state space in \quicksand.
+Our goal here is to eliminate free-re-malloc false positives immediately upon detecting them, with no replay required.
 
@@
 \landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation \cite{estimation}, and data-race detection.
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) similarly to Fair-Bounded Search \cite{bpor}.
+It avoids state space cycles (e.g.
+ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) similarly to Fair-Bounded Search \cite{bpor}.
+% TODO: devote a paragraph to this, after talking about drs in locks
 % this line can be cut if space is needed
@@
 the small job risks missing the data race entirely if the original PPs were required to expose it.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
+In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
+which motivates the need for Iterative Deepening to decide at runtime which to prioritize.
+% TODO: Put numbers here.
 % which justifies this design

### 2015-10-17 pldi-16/evaluation.tex
write about free-re-malloc and add figure 1

@@
 			\item Do we avoid false positives compared to a single-execution data race analysis?
-				% TODO: Argue:
-				% It is fair to compare multiple pass DR-analysis under Landslide against just a single execution because prior work DR detectors, being not integrated with a MC, are not intended to uncover different results under subsequent runs.
-				% Define a "stress tester" as a class of bug detectors where they are intended to [...]
-				% Maybe say: Do there exist any stress-tester DR detectors where they are intended to produce new results under reruns?
-
 				% Explain later as:
@@
 % TODO: Add more test cases
-We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle. % TODO talk more about these.
-In total, this test suite comprises {\large \bf 99999} % TODO
-unique state spaces.
+We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle system calls.
+In total, this test suite comprises 632 unique state spaces.
 
@@
 \quicksand~tests were given 10 CPUs for 1 hour each.
-To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs enabled in advance (and no data-race PPs).
+To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs on sync primitives enabled in advance (and no data-race PPs).
 % TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
-Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we leveled the playing field by giving each control 1 CPU for 10 hours and dividing all associated times by 10 (simulating perfect parallelism).
+Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours, then dividing all associated times by 10 (simulating perfect parallelism).
 
@@
 	\end{tabular}
-	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom {\bf \large 9999999})}
+	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom 2)}
 	\label{tab:allbugs}
@@
 
-asdfadsfasdf
+% TODO: If you have time, rerun all the quicksand experiments JUST running the 4 base state spaces. Give it 10/4 hours of cpu budget on 4 cores.
 
-\subsection{Evaluating Data-Race PPs}
+First we compare \quicksand~to SSS-MC
+% mention exactly which state spaces we are comparing here
 
-In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
+% mention partial verification in terms of state space completion, when SSSMC times out.
 
-{\bf Finding bugs missed by SSS-MC.}
+\subsection{Finding data-race bugs missed by SSS-MC}
+
+%In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
+
 Though state-of-the-art MCs preempt only on synchronization events, many serious concurrency bugs are caused by data races leading to corrupted shared state.
-In Table~\ref{tab:drbugs} we count ...... % TODO
+Hence, in Table~\ref{tab:drbugs} we count how often \quicksand~uncovered a bug only in state spaces which included data-race PPs,
+while
 
@@
 
-{\bf Avoiding false positive data-race candidates.}
+\subsection{Avoiding false positive data-race candidates}
 % Though we mechanically verify whether each data race candidate leads to a bug, each new PP can increase combinatorially..... obviously wish to avoid...
 
-{\bf Avoiding false negative data-race bugs.}
+We also counted the number of free-re-malloc false positives that \landslide~suppressed.
+
+\subsection{Finding nondeterministic data-race candidates}
 \label{sec:eval-falseneg}
-Some pairs of unprotected shared-memory accesses may be hidden in some control flow path that requires a nondeterministic preemption to be executed.
-In such cases, a single-pass data-race detector
+Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
+In such cases, a single-pass dynamic data-race detector
 %could not achieve the coverage necessary
 would fail
-to identify the access pair as a data-race candidate to begin with.
+to identify a racing access pair as a candidate to begin with.
 %
-We counted how many such data-races, used as PPs, would lead to \landslide~finding new bugs,
+We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
 thereby making them {\em false negatives} of the single-pass approach.
@@
 We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).
-%
+
 To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
-For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must, rather than waiting until future interleavings to confirm them (see \sect{\ref{sec:free-re-malloc}}).
+For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must,
+rather than waiting until future interleavings to confirm them (as explained in \sect{\ref{sec:free-re-malloc}}).
+
+				% TODO: Argue:
+				% It is fair to compare multiple pass DR-analysis under Landslide against just a single execution because prior work DR detectors, being not integrated with a MC, are not intended to uncover different results under subsequent runs.
+				% Define a "stress tester" as a class of bug detectors where they are intended to [...]
+				% Maybe say: Do there exist any stress-tester DR detectors where they are intended to produce new results under reruns?
+
+One might also wonder: Why is it fair to compare the data race bugs \quicksand~finds (10 CPU-hours)
+against the candidates found by a prior work's single execution (less than a minute)?
+We argue this comparison is meaningful because prior work data-race tools, being not integrated with a MC,
+are not intended to uncover different results under subsequent runs.
+One could run a data-race tool repeatedly for 10 CPU-hours, but the advantage of stateless MC over stress testing is already well-understood.
+% TODO TODO TODO get a big citation for teh above sentence.
 
@@
 Count the number of kernels for which this was the case.
+However, larger state spaces often result from finer-grained synchronization,
+which can indicate a more complicated concurrent algorithm that would require more rigorous verification than a program with a single global lock.
+(analogous to a developer % changing an implementation to be finer graind and then re running the same unit test)
 

### 2015-10-17 pldi-16/intro.tex
write about free-re-malloc and add figure 1

@@
 Unfortunately, these state spaces explode exponentially in the size of the input program.
-Reduction techniques such as Dynamic Partial Order Reduction (DPOR) \cite{dpor} and Maximal Causality Reduction (MCR) \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding (ICB) \cite{chess} allow bugs to be found sooner in a given exploration should they exist.
+Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
+and search ordering strategies such as Iterative Context Bounding \cite{chess} allow bugs to be found sooner in a given exploration should they exist.
 
@@
 It provides an additional data-race analysis to report any violations of this model;
-however, data-race analyses are prone to report false positives and benign races which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
+however, data-race analyses are prone to report false positives
+%and benign races
+which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
 %

### 2015-10-17 pldi-16/related.tex
write about free-re-malloc and add figure 1

@@
 
-{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
-% TODO: Be proud of them instead of being dismissive.
+{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs.
+However, the resulting state space explosion demands that any such tool either
+choose a small subset of instructions to consider as PPs.
+or be limited to very small test inputs.
+%However, every such prior tool we know of has serious drawbacks.
 {\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
 which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
-% TODO: Make sure you talk about your mutex experiment, or rephrase this.
 SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL.
+However, it cannot practicelly test full-size programs, and requires code to be written in the PROMELA DSL.
 %so cannot check implementations directly.
@@
 It performs a single-pass analysis as normal, then explores a limited state space to classify candidates in a taxonomy of likely severity.
-Compared to us, \portend~can detect non-failing races which nevertheless cause suspiciously different program output,
+Compared to us, \portend~can also detect non-failing races which nevertheless cause suspiciously different program output,
 while we depend on directly detecting failures.
-It also uses symbolic execution to test input nondeterminism along with schedule nondeterminism,
+It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
 while we explore the latter only.
@@
 which is necessary to expose some bugs (\sect{\ref{sec:eval-falseneg}}).
-It is also limited to the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.
+It also assumes the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.
 Future work could combine the two approaches, using \landslide~to produce new traces for \portend~to classify, or using \portend's analysis to inform \quicksand's state space priorities.

### 2015-10-17 pldi-16/di.tex
random edits; rename FRM -> malloc-recycle; why am i still awake

@@
 If they cannot be reordered at all, it is a {\em false positive}.
-For brevity, we classify benign data races as a subset of false positives.
+For brevity, we count benign data races as a subset of false positives.
 
@@
 We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.
-Figure~\ref{fig:free-re-malloc} shows a common code pattern and interleaving which can expose such behavior.
+Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
 If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 6 will be flagged as a data race.
+We term this a {\em malloc-recycle data race}.
 To the human eye, this is obviously a false positive: reordering lines 4-6 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
@@
 \end{tabular}
-\caption{A common pattern with {\tt malloc()} that produces false positive data race candidates.}
-\label{fig:free-re-malloc}
+\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
+\label{fig:recycle}
 \end{figure}
@@
 each allocation is given a unique ID, and when evaluating whether two heap accesses can race, their generation counters must match.
-
 However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:frm-bug}:
+Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
 Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
@@
 \end{tabular}
-\caption{blah blah}
-\label{fig:frm-bug}
+\caption{If a single-pass data race detector discarded candidates matching the malloc-recycle pattern,
+it would miss the bug in this adversarial program.}
+\label{fig:recycle-bug}
 \end{figure}
@@
 % or (b) reuse of memory by malloc is somehow outside their concurrency model.
-Note that while replay-oriented data race tools can already classify free-re-malloc candidates as false positives % benign? (not benign. they are ghosts. not even real.)
+Note that while replay-oriented data race tools can already classify malloc-recycle candidates as false positives % benign? (not benign. they are ghosts. not even real.)
 \cite{recordreplaydrs},
@@
 which is tantamount to adding a new state space in \quicksand.
-Our goal here is to eliminate free-re-malloc false positives immediately upon detecting them, with no replay required.
+Our goal here is to eliminate malloc-recycle false positives immediately upon detecting them, with no replay required.
 

### 2015-10-17 pldi-16/di.tex
write about within_function and finish discussion/future section

@@
 
-\subsection{Suppressing malloc-related false positives}
+\subsection{Suppressing ``malloc-recycle'' false positives}
 
@@
 
+{\bf Restricting PPs with stack trace predicates.}
+Most MCs preempt indiscriminately on any sync API call, regardless of the call-site.
+However, when testing a particular module in a large codebase,
+the user is likely uninterested in PPs arising from other modules.
+\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
+Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
+The {\tt without\_function} directive works similarly, but as a blacklist.
+Multiple invocations can be used; later ones take precedence.
+%\cite{landslide} provides further detail on this feature.
+
 {\bf Data races in lock implementations.}

### 2015-10-17 pldi-16/evaluation.tex
write about within_function and finish discussion/future section

@@
 \end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Test Suite}
@@
 \mxtest, for locking algorithm correctness, \tej, a test of thread lifecycle, \bct~and \paraguay, for condvars, \paradise~for semaphores, and \rwl~for r/w locks.
-For \mxtest~we enabled \landslide's mutex-testing option described in \sect{\ref{sec:landslide}}.
-%, which changes the lock-set tracking to consider accesses within {\tt mutex\_lock} and {\tt mutex\_unlock} as data-race candidates.
-%(Normally, they are considered protected by the very lock they implement: assuming the locks are already correct to enable productive data-race analysis on the rest of the codebase.)
+For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
+and for \mxtest~we enabled \landslide's mutex-testing option
+(see \sect{\ref{sec:landslide}}).
 \newcommand\prisema{\texttt{priority\_sema}}
@@
 We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle system calls.
+For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} mutex.
 In total, this test suite comprises 632 unique state spaces.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Evaluating Iterative Deepening}
+\label{sec:eval-id}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Finding data-race bugs missed by SSS-MC}
+\label{sec:eval-drbugs}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Avoiding false positive data-race candidates}
+\label{sec:eval-falsepos}
 % Though we mechanically verify whether each data race candidate leads to a bug, each new PP can increase combinatorially..... obviously wish to avoid...
@@
 We also counted the number of free-re-malloc false positives that \landslide~suppressed.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+% TODO: restructure this paragraph to account for RaceFuzzer/Portend. Be like, "Yes they explore multiple schedules, but only AFTER finding a race." And/or make this argument in prior work
 One might also wonder: Why is it fair to compare the data race bugs \quicksand~finds (10 CPU-hours)
@@
 
-\subsection{Discussion} % and future work?
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% TODO: Mention that SSSMC control never FAB in mutex_test whatsoever, while DR PPs enabled FABs.
+\subsection{Discussion and future work}
+\label{sec:future}
 
-%We tested 87 pintos kernels and found races in 45-47 of them, among which 34-47 of those bugs required data-race preemption points to expose. That means we rock.
+{\bf Testing lock implementations.}
+In \sect{\ref{sec:eval-drbugs}}, using data-race PPs compares favourably across the board to SSS-MC.
+Observe in particular that in \mxtest, the control experiment found dramatically fewer bugs (just 1),
+even compared to the other test cases\footnote{
+% FIXME: "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
+	The one bug SSS-MC found was in a fully-assembly lock implementation.
+{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
+Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
+%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
+Though for many applications of MC, assuming a correct lock implementation is sufficient,
+we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.
+
+{\bf Partial verification.}
+We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-id}}).
+While we guarantee safety when preempted on certain combinations of PPs,
+ICB ~\cite{chess} guarantees safety under no more than a certain number of preemptions.
+%according to the maximum bound reached in the time limit.
+We imagine these two guarantees could be each be useful to developers in different scenarios,
+and are presently working to combine the two approaches to provide both at once.
+One benefit of our technique is that {\tt within\_function}-based Iterative Deepening (discussed above)
+would enable expert developers to configure custom subsets of PPs they are most interested in verifying,
+according to which modules of a codebase they wish to test.
 
@@
 Count the number of kernels for which this was the case.
-However, larger state spaces often result from finer-grained synchronization,
-which can indicate a more complicated concurrent algorithm that would require more rigorous verification than a program with a single global lock.
-(analogous to a developer % changing an implementation to be finer graind and then re running the same unit test)
+{\bf Finer-grained PP subsets.}
+\quicksand~was able to partially guarantee safety in {\large \bf 99999\%} of tests with large maximal state spaces.
+However, in {\large \bf 1337} tests, no more than the minimal state space could be verified,
+and in {\large \bf 42} tests, not even that much.
+Larger state spaces often result from finer-grained locking,
+which can indicate a more complicated concurrent algorithm requiring more rigorous verification than a program with a single global lock.
+Hence these corner cases are important to consider for future work.
+While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
+we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
+making partial verification of such large tests possible.
+%enabling partial verification of such large tests. % if need space
 

### 2015-10-17 pldi-16/related.tex
write about within_function and finish discussion/future section

@@
 
-We build upon many established model-checking techniques, dating back
+\landslide~uses many established model-checking techniques, dating back
 % of course
@@
 
-{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
+{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, MCR \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
 These are all orthogonal to our technique.
-\textsc{Parrot} \cite{parrot} combines model checking with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
-Finally, Iterative Context Bounding \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable.
+\textsc{Parrot} \cite{parrot} combines MC with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
+Finally, ICB \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
 However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

### 2015-10-18 pldi-16/evaluation.tex
minor fixes

@@
 % TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
-Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours, then dividing all associated times by 10 (simulating perfect parallelism).
+Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours,
+%then dividing all associated times by 10 (simulating perfect parallelism).
+and multiplying \quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
+% TODO: Can you make quicksand actually count the CPU hours. If so use the following instead:
+% and instrumenting \quicksand~to report total CPU-hours rather than wall-clock time.
 
@@
 %Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
-Though for many applications of MC, assuming a correct lock implementation is sufficient,
+%Though for many applications of MC, assuming a correct lock implementation is sufficient,
+Though it suffices for many applications of MC to assume correctly-implemented locks,
 we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.
+
+% TODO: get these numbers
+{\bf Finer-grained PP subsets.}
+\quicksand~was able to partially guarantee safety in {\large \bf 99999\%} of tests with large maximal state spaces.
+However, in {\large \bf 1337} tests, no more than the minimal state space could be verified,
+and in {\large \bf 42} tests, not even that much.
+Larger state spaces often result from finer-grained locking,
+which can indicate a more complicated concurrent algorithm requiring more rigorous verification than a program with a single global lock.
+Hence these corner cases are important to consider for future work.
+While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
+we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
+making partial verification of such large tests possible.
+%enabling partial verification of such large tests. % if need space
 
@@
 
-% TODO: Future work.
-% Future work: Add a way to configure even smaller subsets (eg "only mutex_locks called from site X") for cases where mx_lock and mx_unlock alone are still too big.
-Count the number of kernels for which this was the case.
-{\bf Finer-grained PP subsets.}
-\quicksand~was able to partially guarantee safety in {\large \bf 99999\%} of tests with large maximal state spaces.
-However, in {\large \bf 1337} tests, no more than the minimal state space could be verified,
-and in {\large \bf 42} tests, not even that much.
-Larger state spaces often result from finer-grained locking,
-which can indicate a more complicated concurrent algorithm requiring more rigorous verification than a program with a single global lock.
-Hence these corner cases are important to consider for future work.
-While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
-we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
-making partial verification of such large tests possible.
-%enabling partial verification of such large tests. % if need space
-
 % Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-10-18 pldi-16/related.tex
minor fixes

@@
 %and data-race detection \cite{eraser}.
-We compare related model checkers by their treatment of shared-memory thread communication.
+We compare related tools by their treatment of shared-memory thread communication.
 
@@
 SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, it cannot practicelly test full-size programs, and requires code to be written in the PROMELA DSL.
+However, it cannot practically test full-size programs, and requires code to be written in the PROMELA DSL.
 %so cannot check implementations directly.
 
+% TODO is it ok to abbreviate ICB and MCR, given we defined the abbrs all the way back in the intro?
 {\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, MCR \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.

### 2015-10-18 pldi-16/di.tex
change figures

@@
 % TODO: get latex to put this on page 2 somehow
-\begin{figure*}[t]
-	\includegraphics[width=0.15\textwidth]{tree0.pdf}
-	\includegraphics[width=0.225\textwidth]{tree1.pdf}
-	\includegraphics[width=0.25\textwidth]{tree2.pdf}
-	\includegraphics[width=0.35\textwidth]{tree3.pdf}
-	\caption{Iterative Deepening visualization.
-		The minimal state space includes only mandatory thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
-Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone, {\tt mutex\_unlock} alone, or both together.
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{trees.pdf}
+	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
+	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
+	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
+	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
+	\caption{Iterative Deepening example.
+		The minimal state space (top) includes only mandatory thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
+		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
 Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
@@
 	\label{fig:id}
-\end{figure*}
+\end{figure}
 \section{Design}
@@
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
-If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 6 will be flagged as a data race.
+If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
 We term this a {\em malloc-recycle data race}.
-To the human eye, this is obviously a false positive: reordering lines 4-6 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
-Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 4.
+To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
+Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
 
 \begin{figure}[t]
+	\small
 \begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \};}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \};~~~~~~~~~~}} \\
+	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
+	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
 	\\
@@
 	3 & & \texttt{// x's memory recycled} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(...);} \\
-	5 & & \texttt{publish(y);} \\
-	6 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
+	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	5 & & \texttt{// ...initialize...}\\
+	6 & & \texttt{publish(y);} \\
+	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
 \end{tabular}
@@
 \landslide~implements this check by adding a generation counter to its heap state tracking:
-each allocation is given a unique ID, and when evaluating whether two heap accesses can race, their generation counters must match.
+each allocation is given a unique ID,
+and when evaluating whether two heap accesses can race,
+the IDs of their containing blocks must match.
 However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
@@
 \begin{figure}[t]
+	\small
 \begin{tabular}{rll}
@@
 	5 & & \texttt{// x's memory recycled} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(...);} \\
+	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
 	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\

### 2015-10-20 pldi-16/di.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

@@
-% TODO: get latex to put this on page 2 somehow
 \begin{figure}[t]
@@
 	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only mandatory thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
+		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
 		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
@@
 }
-	% TODO: reference this label in first few paragraphs of design
 	\label{fig:id}
@@
 
-TODO
+Motivated by state-of-the-art tools' inflexibility to change their preemption points,
+Iterative Deepening searches among different combinations of such points during a test,
+deciding on-the-fly whether to pursue each resulting state space or abandon it in favour of smaller ones.
+Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}.
+We show a visualization in Figure~\ref{fig:id}.
 
-Unlike in chess AI, however, adding more preemption points to a completed state space is not necessarily a linear process, as  % blah blah multiple combinations
+Iterative Deepening depends heavily on state-space estimation \cite{estimation}
+to understand which state spaces are likely to complete on time,
+in advance of actually testing each interleaving within.
+The algorithm's purpose is to make decisions automatically about when to defer exploration of a state space,
+so an inexpert user need only provide their total CPU budget as an argument.
 
-% TODO: Introduce abbreviation for SSS-MC here
+Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
+A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
+Moreover, because Iterative Deepening treats the set of preemption points as mutable,
+it grants the potential to add new ones reactively based on any runtime analysis.
+In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
 
-% define minimal state space and maximal state space
-{\bf Terminology.}
-Importantly, we distinguish data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is false-positive-prone, Landslide reports unprotected access pairs separately from the problems listed above, calling the pair a {\em data race candidate}.
+\subsection{Terminology}
+
+For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
+{\em model checking} (MC),
+and {\em single-state-space MC} (SSS-MC) (i.e., the approach of prior work).
+
+Please note the difference between data-race {\em candidates} and data-race {\em bugs}.
+Because data-race analysis is false-positive-prone,
+we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
+calling such pairs {\em data race candidates}.
 Should a future interleaving, preempting during those accesses, cause a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
@@
 
-\subsection{Verifying data races as buggy or benign}
+We also identify the {\em minimal} and {\em maximal state space} for each test.
+The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}).
+The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
+However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
+
+\subsection{Verifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?
+
+\subsection{Choosing the best job}
+
+% TODO: should work block / find work algorthm here
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Implementation}
+
+\subsection{Landslide}
+\label{sec:landslide}
+
+We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for dynamic data-race analysis.
+\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation, and hybrid lockset/happens-before data-race detection.
+It avoids state space cycles (e.g.
+ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
+% TODO: devote a paragraph to this, after talking about drs in locks
+% this line can be cut if space is needed
+It can test both user- and kernel-level code, although is limited to timer-driven nondeterminism.
+Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
+
+{\bf Restricting PPs with stack trace predicates.}
+Most MCs preempt indiscriminately on any sync API call, regardless of the call-site.
+However, when testing a particular module in a large codebase,
+the user is likely uninterested in PPs arising from other modules.
+\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
+Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
+The {\tt without\_function} directive works similarly, but as a blacklist.
+Multiple invocations can be used; later ones take precedence.
+%\cite{landslide} provides further detail on this feature.
+
+{\bf Data races in lock implementations.}
+Prior work data race tools recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself \cite{tsan}.
+Assuming the locks are already correct enables productive data-race analysis on the rest of the codebase.
+With testing limited to one execution, even if you wished to test for lock bugs, data-race analysis would still uselessly flag every single access pair in the lock implementation, requiring human attention to verify.
+However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
+Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
+(Accesses from other sync functions, such as {\tt cond\_wait()}, would be either included already, or protected by an internal mutex.)
+% TODO: Add a code sample showing how the lockset boundaries change? If there's room?
+
+\subsection{Quicksand} %Managing multiple state spaces
+
+\quicksand~is an independent program that wraps the execution of several \landslide~instances.
+The implementation is roughly 3000 lines of C code.
+
+\quicksand~uses a workqueue model
+...
+priorities... data race priorities...
+
+{\bf Communication protocol.}
+The interface to \landslide~, which any similar MC could implement, is in two parts.
+First, when starting each job, it writes a configuration file declaring which PPs to use,
+% can lose this line due to space
+among other options such as mutex-testing mode,
+passed as an argument to \landslide.
+Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing on a FIFO pipe.
+\landslide~sends a message after each new interleaving is tested, to report updated progress and ETA,
+whenever a new data-race candidate is found, and whenever a bug is found.
+\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
+
+{\bf Heuristics.}
+
+% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
+
+{\bf Avoiding thrashing in \quicksand.} % TODO
+
+% TODO: decide which subsection should come first for better flow
+
+\subsection{Data-race preemption points}
+
+% Intuitively, % TODO describe what we're trying to achieve % do this in design section
+
+When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
+Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
+(For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
+Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
+Instead, \landslide~qualifies the PC with
+(a) the current thread ID and
+(b) the most recent {\tt call} instruction.
+% (a crude approximation of a stack trace)
+% which are much cheaper, as we carry them around all the time already
+Note that we do {\em not} qualify data races by the shared memory address,
+which can change based on different interleavings of previous code
+(for example, depending on the result of {\tt malloc()}).
+% especially when malloc is involved.
+% TODO: decide whether to make dont-filter-dr-by-address example.
+%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
+
+When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
+a ``small'' job to preempt on the racing instruction only,
+and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
+%
+Hence each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
+Rather than adding a maximal job with both new PPs at once, we prefer to add smaller jobs which have a higher chance of completing in time.
+If those state spaces are bug-free, they will in turn add the maximal job later.
+%
+The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
+while the big job risks not completing in time,
+the small job risks missing the data race entirely if the original PPs were required to expose it.
+In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
+which motivates the need for Iterative Deepening to decide at runtime which to prioritize.
+% TODO: Put numbers here.
+% which justifies this design
+
+% TODO: make new-dr-jobs diagram
+
+%                                  (2)          landslide         landslide
+%  landslide                   "Preempt when-->  o  PPs=[foo]      o  PPs=[mutex_lock,
+%    o  PPs=[mutex_lock]        PC = FOO"       / \               / \      foo]
+%   / \                              |         o   o             o   o
+%  o  /\                             |                          /|\  |\
+%    o  o                                                      o o o o o
+%                             __QUICKSAND__
+%    |       (1)             |             |           (3)
+%    \  "Hey, an access      |  job1       | ----"Preempt when -->
+%     '- at FOO raced w/ --> |        job2 |      PC = BAR"
+%        one at BAR!         |    job4     |
+%                            |        job70|      [ same pictures for s/foo/bar/]
+%                            | job8        |
+%                            '-------------'
+%                maybe show some internal structore of QS as well
+
+% TODO: Decide whether to discuss the fact that we don't add the 2nd of a "suspected" dr pair until it is confirmed.
 
@@
 % TODO: example codez with malloc re free go here
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Implementation}
-
-\subsection{Landslide}
-\label{sec:landslide}
-
-We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for data-race analysis~\cite{eraser}.
-% optional: cite tsan above
-\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation \cite{estimation}, and data-race detection.
-It avoids state space cycles (e.g.
-ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) similarly to Fair-Bounded Search \cite{bpor}.
-% TODO: devote a paragraph to this, after talking about drs in locks
-% this line can be cut if space is needed
-It can test both user- and kernel-level code, though the latter requires user annotations and is limited to timer-driven nondeterminism.
-Its bug-detection metrics include assertion failures, deadlock, segfaults, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock checker.
-
-{\bf Restricting PPs with stack trace predicates.}
-Most MCs preempt indiscriminately on any sync API call, regardless of the call-site.
-However, when testing a particular module in a large codebase,
-the user is likely uninterested in PPs arising from other modules.
-\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
-Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
-The {\tt without\_function} directive works similarly, but as a blacklist.
-Multiple invocations can be used; later ones take precedence.
-%\cite{landslide} provides further detail on this feature.
-
-{\bf Data races in lock implementations.}
-Prior work data race tools recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself \cite{tsan}.
-Assuming the locks are already correct enables productive data-race analysis on the rest of the codebase.
-With testing limited to one execution, even if you wished to test for lock bugs, data-race analysis would still uselessly flag every single access pair in the lock implementation, requiring human attention to verify.
-However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
-(Accesses from other sync functions, such as {\tt cond\_wait()}, would be either included already, or protected by an internal mutex.)
-% TODO: Add a code sample showing how the lockset boundaries change? If there's room?
-
-\subsection{Managing multiple state spaces}
-
-\quicksand~uses a workqueue model
-...
-priorities... data race priorities...
-
-% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-{\bf Avoiding thrashing in \quicksand.} % TODO
-
-% TODO: decide which subsection should come first for better flow
-
-\subsection{Data-race preemption points}
-
-% Intuitively, % TODO describe what we're trying to achieve % do this in design section
-
-When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
-Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-(For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
-Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
-Instead, \landslide~qualifies the PC with
-(a) the current thread ID and
-(b) the most recent {\tt call} instruction.
-% (a crude approximation of a stack trace)
-% which are much cheaper, as we carry them around all the time already
-Note that we do {\em not} qualify data races by the shared memory address,
-which can change based on different interleavings of previous code
-(for example, depending on the result of {\tt malloc()}).
-% especially when malloc is involved.
-% TODO: decide whether to make dont-filter-dr-by-address example.
-%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
-When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
-a ``small'' job to preempt on the racing instruction only,
-and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
-%
-Hence each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
-Rather than adding a maximal job with both new PPs at once, we prefer to add smaller jobs which have a higher chance of completing in time.
-If those state spaces are bug-free, they will in turn add the maximal job later.
-%
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
-while the big job risks not completing in time,
-the small job risks missing the data race entirely if the original PPs were required to expose it.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to decide at runtime which to prioritize.
-% TODO: Put numbers here.
-% which justifies this design
-
-% TODO: make new-dr-jobs diagram
-
-%                                  (2)          landslide         landslide
-%  landslide                   "Preempt when-->  o  PPs=[foo]      o  PPs=[mutex_lock,
-%    o  PPs=[mutex_lock]        PC = FOO"       / \               / \      foo]
-%   / \                              |         o   o             o   o
-%  o  /\                             |                          /|\  |\
-%    o  o                                                      o o o o o
-%                             __QUICKSAND__
-%    |       (1)             |             |           (3)
-%    \  "Hey, an access      |  job1       | ----"Preempt when -->
-%     '- at FOO raced w/ --> |        job2 |      PC = BAR"
-%        one at BAR!         |    job4     |
-%                            |        job70|      [ same pictures for s/foo/bar/]
-%                            | job8        |
-%                            '-------------'
-%                maybe show some internal structore of QS as well
-
-% TODO: Decide whether to discuss the fact that we don't add the 2nd of a "suspected" dr pair until it is confirmed.

### 2015-10-20 pldi-16/evaluation.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

@@
 
+%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
+% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
+% (as well as on the mutex API boundaries).
+% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
+
 {\bf Testing lock implementations.}
@@
 While we guarantee safety when preempted on certain combinations of PPs,
-ICB ~\cite{chess} guarantees safety under no more than a certain number of preemptions.
+CHESS
+guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.
 %according to the maximum bound reached in the time limit.

### 2015-10-20 pldi-16/intro.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

--- 2015-10-17 pldi-16/intro.tex
@@
 Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding \cite{chess} allow bugs to be found sooner in a given exploration should they exist.
+and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} allow bugs to be found sooner in a given exploration should they exist.
 
@@
 is able to preempt threads around any shared memory access.
-Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for all but the most rudimentary test inputs.
+Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for even modestly-sized tests.
 %
@@
 We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
-Named after the analogous technique in chess AI \cite{iterative-deepening-chess}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
+Named after the analogous technique in chess AI \cite{iterative-deepening-chess-ai}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
 Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),

### 2015-10-20 pldi-16/related.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

@@
 SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, it cannot practically test full-size programs, and requires code to be written in the PROMELA DSL.
+However, SPIN explicitly stores visited program states rather than using DPOR, limiting the size of programs that can be practically tested,
+and requires code to be written in the PROMELA DSL.
 %so cannot check implementations directly.
 
-% TODO is it ok to abbreviate ICB and MCR, given we defined the abbrs all the way back in the intro?
-{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, MCR \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
+{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
 These are all orthogonal to our technique.
 \textsc{Parrot} \cite{parrot} combines MC with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
-Finally, ICB \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
+Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
 However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

### 2015-10-20 pldi-16/di.tex
finish(?) related work

@@
 \subsection{Choosing the best job}
+
+This algorithm is the heart of Iterative Deepening; we list it in Figure~\ref{fig:shouldworkblock}.
+The algorithm takes into account... TODO
+Though it has worst-case performance of $O(mn)$ in the number of pending and blocked jobs,
+in practice we find the non-constant portion runs very infrequently and is negligible compared to the exponentially-sized state spaces.
 
@@
 \subsection{Suppressing ``malloc-recycle'' false positives}
+\label{sec:recycle}
 
@@
 % or (b) reuse of memory by malloc is somehow outside their concurrency model.
-Note that while replay-oriented data race tools can already classify malloc-recycle candidates as false positives % benign? (not benign. they are ghosts. not even real.)
-\cite{recordreplaydrs},
-they require replaying the threads in a new interleaving.
-Moreover, \cite{portend} argues that accurate classification may require re-executing many times,
-according to many pre- and post-race sequences,
-which is tantamount to adding a new state space in \quicksand.
-Our goal here is to eliminate malloc-recycle false positives immediately upon detecting them, with no replay required.
 

### 2015-10-20 pldi-16/related.tex
finish(?) related work

@@
 
-\subsection{Stateless Model Checking}
+\subsection{Model Checking}
 
@@
 
-{\bf Synchronization events only.} CHESS \cite{chess} and \textsc{dBug} \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
+{\bf Synchronization events only.} CHESS \cite{chess} and dBug \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
 Hence they will miss any crashes that require interleaving threads at instruction granularity during a data race.
@@
 
-{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist} \cite{modist}, and ETA \cite{dbug-retreat}, limit thread communication to a message-passing API to more effectively test distributed systems.
+{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, and ETA \cite{dbug-retreat}, limit thread communication to a message-passing API to more effectively test distributed systems.
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
@@
 %However, every such prior tool we know of has serious drawbacks.
-{\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
+SKI \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
 which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
@@
 These are all orthogonal to our technique.
-\textsc{Parrot} \cite{parrot} combines MC with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
-Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
-However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+Parrot \cite{parrot} combines MC with a partially-determinizing runtime, but still, fewer than half the non-trivial state spaces in their evaluation could be completed.
+%providing a strong argument for \quicksand.
+Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to Iterative Deepening,
+as both approaches provide a partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
+However, ICB is limited to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
 
@@
 
-blah blah blah threadsanitizer, racePRO, etc
+Too many related projects to list have made contributions to the false-positive problem first introduced in \cite{eraser}.
+\cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
+% TODO: any more?
+DroidRacer \cite{droidracer} and CAFA \cite{cafa} extend the analysis to event-driven Android applications, using domain-specific heuristics (orthogonal to our method) to reduce false positives.
+% cut for space?
 
-\newcommand\portend{\textsc{Portend}}
-\portend~\cite{portend} is the most closely related work we have found.
-It performs a single-pass analysis as normal, then explores a limited state space to classify candidates in a taxonomy of likely severity.
-Compared to us, \portend~can also detect non-failing races which nevertheless cause suspiciously different program output,
-while we depend on directly detecting failures.
+% TODO: Figure out why they claim "Happens before produces NO false positives, only benign races".
+% It seems impossible.
+% But if true, it means either (a) they can somehow identify FRM DRs on the 1st pass, not needing to replay,
+% or (b) reuse of memory by malloc is somehow outside their concurrency model.
+Closer to our work, replay analysis \cite{recordreplaydrs} also suppresses false positives by testing multiple thread interleavings.
+%after finding data race candidates.
+This work compares the immediately resulting program states for differences,
+preferring instead still to err on the side of false positives.
+RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited,
+although it uses random schedule fuzzing rather than stateless model checking.
+Note while these techniques can also classify our malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
+they require replaying the threads in a new interleaving.
+Moreover, \cite{portend} argues that accurate classification may require many re-executions,
+%according to many pre- and post-race sequences,
+which is tantamount to adding a new state space in \quicksand.
+Our proof in \sect{\ref{sec:recycle}} allows eliminating this special case with no replay required.
+
+Portend \cite{portend} is the most closely related work we have found.
+% FIXME: "limited"?
+Based on single-pass data race reports, it explores a limited state space to classify candidates in a taxonomy of likely severity.
+Compared to us, Portend additionally finds non-failing races which nevertheless cause
+%suspiciously
+different program output, while we depend on directly detecting failures.
 It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
 while we explore the latter only.
-However, \portend~does not explore alternate interleavings in advance of knowing data-race locations,
+However, Portend does not test alternate interleavings {\em in advance} of knowing data races,
 which is necessary to expose some bugs (\sect{\ref{sec:eval-falseneg}}).
 It also assumes the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.
-Future work could combine the two approaches, using \landslide~to produce new traces for \portend~to classify, or using \portend's analysis to inform \quicksand's state space priorities.
+Future work could combine the two approaches, using \landslide~to produce new traces for Portend to classify, or using Portend's analysis to inform \quicksand's state space priorities.
 
-\subsection{Other Concurrency Testing Approaches} % TODO: Well?
-
-blah blah pldi'15 symbiosis DSP
+% \subsection{Other Concurrency Testing Approaches} % TODO: Well?
+%
+% blah blah pldi'15 symbiosis DSP
 

### 2015-10-20 pldi-16/di.tex
write algorithm 1, GetBestJob

@@
 in advance of actually testing each interleaving within.
-The algorithm's purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user need only provide their total CPU budget as an argument.
+The purpose is to make decisions automatically about when to defer exploration of a state space,
+so an inexpert user need only provide their total CPU budget as an argument,
+and to enable completing appropriately-sized state spaces within that budget.
 
@@
 {\em model checking} (MC),
-and {\em single-state-space MC} (SSS-MC) (i.e., the approach of prior work).
+{\em single-state-space model checking} (SSS-MC) (i.e., the approach of prior work),
+and {\em state space estimate} (ETA).
 
@@
 The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
-However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
+%However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
 
@@
 
-This algorithm is the heart of Iterative Deepening; we list it in Figure~\ref{fig:shouldworkblock}.
+% TODO: put colour in this code
+\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
+\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
+\newcommand\GetETA[1]{ETA(#1)}
+\newcommand\GetPPSet[1]{PPSet(#1)}
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
+	\Input{$j_0$, the currently-running job}
+	%\Input{$eta$, $j_0$'s predicted completion time}
+	\Input{\PendingJobs, the list of pending jobs, sorted decreasingly by heuristic priority}
+	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted increasingly by ETA}
+	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
+		// common case: job will probably finish on time \\
+		return $j_0$
+	}
+	\ForEach{job $j_P \in$ \PendingJobs}{
+		// don't run a pending job if a subset of it is already suspended; it would find the same fate! \\
+		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
+			return $j_P$
+		}
+	}
+	%// no pending jobs; maybe resume a suspended job \\
+	\ForEach{job $j_S \in$ \SuspendedJobs}{
+		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
+			$\land$
+			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
+			// $j_S$ is tempting, but if a subset of it is also suspended, don't run the larger one first \\
+			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
+				return $j_S$
+			}
+		}
+	}
+	// $j_0$'s ETA was bad, but nothing else was better \\
+	return $j_0$
+	\caption{Suspending exploration of a too-large state space in favour of a potentially smaller one.}
+	\label{alg:shouldworkblock}
+\end{algorithm}
+
+With a limited CPU budget, we must avoid running tests likely to time out.
+When a MC instance reports too high ETA (greater than some heuristic factor times the time remaining),
+we compare other yet-untested PP sets to find another job more likely to complete in time.
+This algorithm is the heart of Iterative Deepening; we list it in Algorithm~\ref{alg:shouldworkblock}\footnote{
+Though its worst-case performance is $O(mn)$ in the number of pending and suspended jobs,
+in practice the non-constant portion runs very infrequently
+and is negligible compared to the exponentially-sized state spaces.}.
 The algorithm takes into account... TODO
-Though it has worst-case performance of $O(mn)$ in the number of pending and blocked jobs,
-in practice we find the non-constant portion runs very infrequently and is negligible compared to the exponentially-sized state spaces.
 

### 2015-10-20 pldi-16/di.tex
buncha minor revisions

@@
 With a limited CPU budget, we must avoid running tests likely to time out.
-When a MC instance reports too high ETA (greater than some heuristic factor times the time remaining),
+When a MC instance reports an ETA too high,
+% TODO: find some way to point out that heuristic factor is needed because the estimates are imperfect
+%(greater than some heuristic factor times the time remaining),
 we compare other yet-untested PP sets to find another job more likely to complete in time.
@@
 Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing on a FIFO pipe.
-\landslide~sends a message after each new interleaving is tested, to report updated progress and ETA,
+\landslide~sends a message after each new interleaving is tested to report updated progress and ETA,
 whenever a new data-race candidate is found, and whenever a bug is found.
@@
 
+%Note that Algorithm~\ref{alg:shouldworkblock} accounts for the bizarre possibility that given two jobs,
+%%$j_1,j_2$,
+%\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
+%but
+%\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
+%This can arise because the estimation gets more accurate over time \cite{estimation},
+%and one job perhaps ran much longer before suspending.
+%% In such scenarios,
+%We heuristically assume the smaller job's ETA is more accurate,
+
 % If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-{\bf Avoiding thrashing in \quicksand.} % TODO
 
@@
 Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-(For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
+(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
 Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
@@
 (a) the current thread ID and
+% FIXME: We don't actually do this.
 (b) the most recent {\tt call} instruction.
@@
 % especially when malloc is involved.
-% TODO: decide whether to make dont-filter-dr-by-address example.
 %Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
@@
 
+TODO: Here will be a proof sketch that, during MC only, that mechanism is safe, because in the Figure~\ref{fig:recycle-bug} scenario, we are sure DPOR will reorder away the {\tt free} in a future interleaving.
+
 % TODO: Figure out why they claim "Happens before produces NO false positives, only benign races".

### 2015-10-20 pldi-16/di.tex
It lets us avoid thrashing in Quicksand.

@@
 When a MC instance reports an ETA too high,
-% TODO: find some way to point out that heuristic factor is needed because the estimates are imperfect
-%(greater than some heuristic factor times the time remaining),
 we compare other yet-untested PP sets to find another job more likely to complete in time.
@@
 and is negligible compared to the exponentially-sized state spaces.}.
-The algorithm takes into account... TODO
+%
 
-% TODO: should work block / find work algorthm here
+Our method's main feature is that it understands that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+and $j_1$ is suspended, $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ would take at least as long.
+Hence we should avoid testing $j_2$ unless further time spent on $j_1$ reveals that it might finish in time after all.
+%
+We also account for the inherent inaccuracy of ETA estimates.
+Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
+when their ETAs are actually overestimates.
+Lines 12-14 account for the
+%bizarre
+possibility that among two suspended jobs,
+%given two jobs,
+%%$j_1,j_2$,
+\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
+but
+\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
+This can arise because estimates tend to get more accurate over time,
+and $j_1$ perhaps ran much longer before suspending.
+% In such scenarios,
+We heuristically assume the smaller job's ETA is more accurate
+to avoid repeatedly resuming larger jobs very briefly while their ETAs only become worse
+(it lets us avoid thrashing in \quicksand),
 
@@
 {\bf Heuristics.}
-
-%Note that Algorithm~\ref{alg:shouldworkblock} accounts for the bizarre possibility that given two jobs,
-%%$j_1,j_2$,
-%\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
-%but
-%\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
-%This can arise because the estimation gets more accurate over time \cite{estimation},
-%and one job perhaps ran much longer before suspending.
-%% In such scenarios,
-%We heuristically assume the smaller job's ETA is more accurate,
 

### 2015-10-21 pldi-16/di.tex
Add dr-jobs figure showing how dr jobs are added

@@
 %However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
+%% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.
 
@@
 and is negligible compared to the exponentially-sized state spaces.}.
-%
+Our method's main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+and $j_1$ is suspended,
+then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
+Hence we should avoid testing $j_2$ unless further time spent on $j_1$ reveals that it might finish in time after all.
+Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
 
-Our method's main feature is that it understands that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
-and $j_1$ is suspended, $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ would take at least as long.
-Hence we should avoid testing $j_2$ unless further time spent on $j_1$ reveals that it might finish in time after all.
 %
@@
 Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
-when their ETAs are actually overestimates.
+in case their ETAs are actually overestimates.
 Lines 12-14 account for the
@@
 We heuristically assume the smaller job's ETA is more accurate
-to avoid repeatedly resuming larger jobs very briefly while their ETAs only become worse
+to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
 (it lets us avoid thrashing in \quicksand),
@@
 \quicksand~is an independent program that wraps the execution of several \landslide~instances.
-The implementation is roughly 3000 lines of C code.
+The implementation is roughly 3000 lines of C.
+%Figure~\ref{fig:quicksand} depicts the major components.
+%and a registry of PPs to keep track of data race candidates.
 
@@
 \subsection{Data-race preemption points}
-
-% Intuitively, % TODO describe what we're trying to achieve % do this in design section
 
@@
 % which justifies this design
+
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
+	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
+		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
+	\label{fig:new-dr-jobs}
+\end{figure}
 

### 2015-10-21 pldi-16/di.tex
finish describing quicksand implementation

@@
 then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless further time spent on $j_1$ reveals that it might finish in time after all.
+Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves after further time spent on it.
+%reveals that it might finish in time after all.
 Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
@@
 The implementation is roughly 3000 lines of C.
-%Figure~\ref{fig:quicksand} depicts the major components.
-%and a registry of PPs to keep track of data race candidates.
-
-\quicksand~uses a workqueue model
-...
-priorities... data race priorities...
+It uses a thread pool to schedule the available state spaces,
+sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
+Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
 
@@
 {\bf Heuristics.}
+% List of all heuristix:
+% HOMESTRETCH - last 60sec of test, don't suspend
+% ETA_THRESH - "to let its ETA stabilize"
+% eta factor
+% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
+% priority change between suspected and confirmed dr
+Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
+to express how optimistic we are about the estimate's accuracy.
+We use 2 as this scaling factor based on the results in \cite{estimation},
+though we allow changing it via the command line.
+We also include a heuristic to
+%ignore ETAs entirely
+never suspend jobs before they pass a certain threshold of interleavings tested,
+for which we choose 32,
+so their ETAs have some time to stabilize.
+
+We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
+based on whether \landslide~observed the access pair ordered one or both ways while testing the original state space,
+Single-order candidates are much more likely to be false-positives
+(though we cannot say for sure before preempting during the access itself, hence the heuristic),
+so we prioritize jobs with both-order data-race PPs more highly.
+For single-order races, we do not add PPs for the later access at all;
+should they be needed, preempting on the first access will suffice to upgrade it to both-order.
 
@@
 % TODO: Put numbers here.
-% which justifies this design
 
@@
 \end{figure}
-
-% TODO: make new-dr-jobs diagram
-
-%                                  (2)          landslide         landslide
-%  landslide                   "Preempt when-->  o  PPs=[foo]      o  PPs=[mutex_lock,
-%    o  PPs=[mutex_lock]        PC = FOO"       / \               / \      foo]
-%   / \                              |         o   o             o   o
-%  o  /\                             |                          /|\  |\
-%    o  o                                                      o o o o o
-%                             __QUICKSAND__
-%    |       (1)             |             |           (3)
-%    \  "Hey, an access      |  job1       | ----"Preempt when -->
-%     '- at FOO raced w/ --> |        job2 |      PC = BAR"
-%        one at BAR!         |    job4     |
-%                            |        job70|      [ same pictures for s/foo/bar/]
-%                            | job8        |
-%                            '-------------'
-%                maybe show some internal structore of QS as well
 

### 2015-10-21 pldi-16/di.tex
minor revisions

@@
 deciding on-the-fly whether to pursue each resulting state space or abandon it in favour of smaller ones.
-Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}.
+Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
 We show a visualization in Figure~\ref{fig:id}.
@@
 The purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user need only provide their total CPU budget as an argument,
+so an inexpert user need only provide their total CPU budget as a test parameter,
 and to enable completing appropriately-sized state spaces within that budget.
@@
 Moreover, because Iterative Deepening treats the set of preemption points as mutable,
-it grants the potential to add new ones reactively based on any runtime analysis.
+it can add new ones reactively based on any runtime analysis.
 In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
@@
 calling such pairs {\em data race candidates}.
-Should a future interleaving, preempting during those accesses, cause a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
+Should a future interleaving, preempting during those accesses,
+lead to a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
 Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
@@
 we compare other yet-untested PP sets to find another job more likely to complete in time.
-This algorithm is the heart of Iterative Deepening; we list it in Algorithm~\ref{alg:shouldworkblock}\footnote{
-Though its worst-case performance is $O(mn)$ in the number of pending and suspended jobs,
-in practice the non-constant portion runs very infrequently
+Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
+Though its worst-case performance is $O(mn)$ in the
+%number of pending and suspended jobs,
+sizes of $\mathcal{P}$ and $\mathcal{S}$,
+in practice the non-constant portion beyond line 4 runs very infrequently
 and is negligible compared to the exponentially-sized state spaces.}.
-Our method's main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
 and $j_1$ is suspended,
@@
 
-We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for dynamic data-race analysis.
-\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation, and hybrid lockset/happens-before data-race detection.
+We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
+\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation, and a hybrid lockset/happens-before data-race analysis.
 It avoids state space cycles (e.g.
@@
 Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
-(Accesses from other sync functions, such as {\tt cond\_wait()}, would be either included already, or protected by an internal mutex.)
+(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
 % TODO: Add a code sample showing how the lockset boundaries change? If there's room?
@@
 
-% TODO: Decide whether to discuss the fact that we don't add the 2nd of a "suspected" dr pair until it is confirmed.
-
 \subsection{Suppressing ``malloc-recycle'' false positives}

### 2015-10-22 pldi-16/proof.tex
Add proof, yet unproved

%
% LaTeX template for prepartion of submissions to PLDI'15
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'15 web site.
%
\documentclass[pldi]{sigplanconf-pldi15}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{amsthm}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphicx}
%\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
%\doublespacing

\newtheorem{definition}{Definition}

\begin{document}

%
% any author declaration will be ignored  when using 'plid' option (for double blind review)
%

% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
\newcommand\landslide{\textsc{Landslide}}
\newcommand\quicksand{\textsc{Quicksand}}
\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1}
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}

% TODO: Finalize these numbers.
\newcommand\numthrlibs{79}
\newcommand\numpintoses{79}
\newcommand\numstudence{158} % total pintoses plus p2s

\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}

\maketitle
\begin{abstract}
In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
Despite many powerful reduction techniques, the state spaces are still exponentially sized,
so any way to avoid exploring some in advance is obviously beneficial.
We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.

This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
\end{abstract}

\section{Definitions}

\begin{definition}[Transition]
A sequence of instructions executed by the program between two preemption points (PPs).
\end{definition}
Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)

\begin{definition}[Must-happen-before (MHB)]
	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
\end{definition}
All transitions of the same thread are trivially MHB-related.
Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked. Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).

Note how our {\em must}-happen-before relation differs from the conventional definition of happens-before (``observed to happen before'') \cite{lamport-clocks}.
Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
the advantage of this over pure-happens-before detectors in producing fewer false negatives is well argued in those prior works\footnote{
Because pure-happens-before data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
Indeed, because most concurrent malloc implementations are protected by a lock, our malloc-recycle false positives fit that same pattern.
}.

\begin{definition}[Data race]
A pair of memory accesses between two threads such that:
\begin{itemize}
	\item At least one of them is a write,
	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
	\item The threads' transitions are not related by MHB.
\end{itemize}
\end{definition}

The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.

\begin{definition}[False positive data race]
	An apparent data race that cannot be observed in the opposite order from what was actually executed.
\end{definition}

False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
changes some variable values when the threads are reordered, such that the memory addresses no longer collide.

\begin{definition}[Malloc-recycle data race]
	A data race where the address is contained in some heap-allocated memory, and between the two accesses, that memory was passed to free() and returned again by a subsequent malloc().
\end{definition}

Figures~\ref{fig:recycle} and \ref{fig:recycle-bug} show an example.
In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.

\begin{figure}[t]
	\small
\begin{tabular}{rll}
	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
	\\
	& Thread 1 & Thread 2 \\
	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
	3 & & \texttt{// x's memory recycled} \\
	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
	5 & & \texttt{// ...initialize...}\\
	6 & & \texttt{publish(y);} \\
	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
\end{tabular}
\caption{False-positive malloc-recycle pattern. This is the common case for which we avoid creating new state spaces.}
\label{fig:recycle}
\end{figure}
\section{Intuition}

In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
then DPOR is guaranteed to ``reorder away the free and re-malloc''.
In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
while the access pair remains for the data-race detector to find, as showin in Figure~\ref{fig:reycle-goal}.
Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.

\begin{figure}[t]
	\small
\begin{tabular}{rll}
	& Thread 1 & Thread 2 \\
	1 & \texttt{publish(x);} & \\
	2 & & \texttt{x2 = get\_published\_x();} \\
	3 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
	4 & \texttt{\hilight{olivegreen}{free}(x);} \\
	5 & & \texttt{// x's memory recycled} \\
	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
\end{tabular}
\caption{Adversarial program which fits the malloc-recycle pattern, but nevertheless contains a true race.}
\label{fig:recycle-bug}
\end{figure}

\begin{figure}[t]
	\small
\begin{tabular}{rll}
	& Thread 1 & Thread 2 \\
	1 & \texttt{publish(x);} & \\
	2 & & \texttt{x2 = get\_published\_x();} \\
	3 & & \texttt{// x not free, so malloc's} \\
	4 & & \texttt{// return value changes!} \\
	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	7 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
	8 & \texttt{\hilight{olivegreen}{free}(x);} \\
\end{tabular}
\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
\label{fig:recycle-goal}
\end{figure}
\section{Assumptions}

We assume the model checker has all static/hard-coded PPs enabled during this state space exploration, and that we are not limited by a pressing CPU budget.
Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving.
However, Iterative Deepening aims to test the most important interleavings with the time available,
so in the case of not enough time, our point here is that continuing the current state space fits that goal best.

Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause MHB (either directly, or because they are built on top of mutexes).

We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection. \landslide~already tracks the heap state, so it can implement this check for free with a simple generation counter.

We assume that the only way the program can obtain heap addresses is through the return value of malloc().
Because we are testing C programs, any crazy violations of this assumption are technically possible,
but should you wish to check for bugs like this,
we would recommend a data-flow analysis which is much cheaper than model checking anyway.

\section{Proof}

Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.

This spells QED so we are done.

\bibliographystyle{abbrvnat}
\bibliography{citations}{}
\end{document}

### 2015-10-22 pldi-16/proof.tex
write all of proof besides han

@@
 
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
 \newtheorem{definition}{Definition}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Definitions}
+
+\subsection{Stateless model checking terms}
 
@@
 A sequence of instructions executed by the program between two preemption points (PPs).
+\label{def:transition}
 \end{definition}
@@
 Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
-the advantage of this over pure-happens-before detectors in producing fewer false negatives is well argued in those prior works\footnote{
+the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
 Because pure-happens-before data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
@@
 
+\begin{definition}[Shared memory conflict]
+A pair of memory accesses between two threads where at least one of them is a write.
+\end{definition}
+Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
+
+\begin{definition}[Dynamic Partial Order Reduction (DPOR)]
+	A state-space search algorithm for stateless model checkers,
+	guaranteed to reorder transitions of two threads
+	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
+	\label{def:dpor}
+\end{definition}
+
+\subsection{Data race and other memory terms}
+
 \begin{definition}[Data race]
-A pair of memory accesses between two threads such that:
+A shared memory conflict where furthermore:
 \begin{itemize}
-	\item At least one of them is a write,
 	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
@@
 The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
+In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
+It is up to \quicksand, outside the scope of this proof, to decide whether they are benign or buggy.
 
@@
 In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.
+
+\begin{definition}[Use after free]
+	Any read or write to heap memory which was once allocated, but no longer is.
+\end{definition}
+%Most commonly this refers to accesses to a region already freed, but for brevity we also include
+% TODO: need heap block overrun?
 
@@
 In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as showin in Figure~\ref{fig:reycle-goal}.
+while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
 Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Assumptions}
@@
 
+We do not require any particular implementation details of malloc; in fact, we instruct \landslide to
+ignore all potential PPs which might be inserted on malloc's internal mutex accesses;
+in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
+
 We assume that the only way the program can obtain heap addresses is through the return value of malloc().
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Proof}
 
-Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
+Though Figures~\ref{fig:recycle-bug} and \ref{fig:recycle-goal} show example programs, they do not capture all possible cases of how a true data race can fit the malloc-recycle pattern.
+We proceed by establishing what must be true of any such program, then casing on the ambiguous possibilities, and showing that PPs will exist where we need them to reorder the threads.
 
-This spells QED so we are done.
+For certain, there must be an access in one thread, followed by a free and malloc (we'll call them ``middle free'' and ``middle malloc''), each possibly from either thread, followed by an access from the other.
+If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
+WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
+
+\begin{lemma}
+	If DPOR will reorder {\tt x2} to before {\tt x1}, then a non-malloc-recycle data race or a use-after-free bug will be identified.
+	\label{lem:reorder}
+\end{lemma}
+\begin{proof}
+By case on which threads the middle free and middle malloc came from.
+\begin{itemize}
+	\item T1 free, T2 malloc (as shown in the figure).
+The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
+Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
+	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same allocation.
+	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
+	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
+\end{itemize}
+If either the middle free or middle malloc came from a third thread, the case is the same as if it belonged to T1.
+\end{proof}
+
+By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
+We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
+Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
+
+\subsection{T2 originally malloced x}
+
+\begin{lemma}[Greedo]
+	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads.
+	\label{lem:greedo}
+\end{lemma}
+\begin{proof}
+Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
+By Definition~\ref{def:transition}, each switch will be a PP.
+By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1}. Lemma~\ref{lem:reorder} finishes.
+\end{proof}
+
+This lemma also applies if there was a third thread, as there would still be a thread switch in the same spot.
+
+\subsection{T1 originally malloced x}
+
+This case is more tricky. We must prove there is a PP... TODO
+
+\begin{lemma}[Han]
+	TODO
+	\label{lem:han}
+\end{lemma}
+
+%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
+%As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
+
+First we must case on how T2 obtains the heap pointer
+
+\subsection{Conclusion}
+
+\begin{theorem}[Soundness of eliminating malloc-recycle races]
+	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
+\end{theorem}
+\begin{proof}
+	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases are covered.
+\end{proof}
+
+%\subsection{Malloc with internal mutexes}
+%
+%First we assume that the malloc implementation is guarded by
+%
+%
+%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
+%
+%
+%\subsection{Malloc as a magic primitive}
+%
+%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
+%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
+%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
+%
+%Hence, we provide another proof that does not require
+%
+%%\cite{landslide} % uncomment post-double-blind review
+%
+%This spells QED so we are done.
 

### 2015-10-22 pldi-16/proof.tex
begin to write about heap overruns in proof

@@
 \authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
 
@@
 However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock, our malloc-recycle false positives fit that same pattern.
+Indeed, because most concurrent malloc implementations are protected by a lock, our malloc-recycle false positives are indistinguishable from such false negatives.
 }.
@@
 \end{definition}
+
+These can immediately be identified as failures by a MC which tracks allocation state.
 %Most commonly this refers to accesses to a region already freed, but for brevity we also include
@@
 	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
 	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
@@
 	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x);} & \\
+	1 & \texttt{publish(x1);} & \\
 	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
 	4 & \texttt{\hilight{olivegreen}{free}(x);} \\
@@
 	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
 	8 & \texttt{\hilight{olivegreen}{free}(x);} \\
@@
 
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration, and that we are not limited by a pressing CPU budget.
+{\bf Maximal state space.}
+We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
+and that we are not limited by a pressing CPU budget.
+Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause MHB (either directly, or because they are built on top of mutexes).
+
 Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving.
@@
 
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause MHB (either directly, or because they are built on top of mutexes).
+{\bf Low overhead.}
+We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
+Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
 
-We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-\landslide~already tracks the heap state, so it can implement this check for free with a simple generation counter.
+{\bf Malloc is a magic black box.}
+We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as any tendency to reuse blocks or allocate adjacent ones.
+%particular allocation pattern regarding adjacency/coalescing/reuse.
+In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
+in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
+Furthermore, we configured DPOR to ignore the internal heap metadata accesses
+when tracking shared memory conflicts to achieve greater state space reduction
+(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
+%for the purpose of avoiding malloc-only shared memory conflicts.
+This is not without consequences; see section~\ref{sec:owned}.
 
-We do not require any particular implementation details of malloc; in fact, we instruct \landslide to
-ignore all potential PPs which might be inserted on malloc's internal mutex accesses;
-in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-
-We assume that the only way the program can obtain heap addresses is through the return value of malloc().
-Because we are testing C programs, any crazy violations of this assumption are technically possible,
+{\bf Sharing heap addresses.}
+Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
+Because we are testing C programs, any bizarre violations of this assumption are technically possible,
 but should you wish to check for bugs like this,
 we would recommend a data-flow analysis which is much cheaper than model checking anyway.
+
+For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the ``black box'' assumption.
 
@@
 \section{Proof}
+\label{sec:proof}
 
@@
 \begin{itemize}
-	\item T1 free, T2 malloc (as shown in the figure).
+	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
 The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
@@
 
-This lemma also applies if there was a third thread, as there would still be a thread switch in the same spot.
+This lemma also applies if a third thread was responsible for this malloc, as there would still be a thread switch in the same spot.
 
@@
 
-This case is more tricky. We must prove there is a PP... TODO
-
 \begin{lemma}[Han]
-	TODO
+	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads.
 	\label{lem:han}
@@
 
+\begin{proof}
+This is the hard case. We must prove there is a PP... TODO
+\end{proof}
+
 %We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
 %As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
-
-First we must case on how T2 obtains the heap pointer
 
@@
 \begin{proof}
-	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases are covered.
+	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases of possible program structure are covered.
 \end{proof}
+
+\section{Heap overruns}
+\label{sec:owned}
+
+Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
+Figure~\ref{fig:overrun} shows an example.
+Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
+So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+	1 & & \texttt{z = malloc(42);} \\
+	2 & & \texttt{// TODO bounds check??} \\
+	3 & & \texttt{x2 = \&z[50];} \\
+	4 & \texttt{x1 = malloc(...);} & \\
+	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	6 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	7 & & \texttt{// x's memory recycled} \\
+	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	9 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+\end{tabular}
+\caption{Final possibility for how T2 can share T1's allocation address, and probably a security vulnerability to boot!}
+\label{fig:overrun}
+\end{figure}
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+	1 & & \texttt{z = malloc(42);} \\
+	2 & & \texttt{// TODO bounds check??} \\
+	3 & & \texttt{x2 = \&z[50];} \\
+	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	6 & \texttt{x1 = malloc(...);} & \\
+	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	8 & \texttt{\hilight{olivegreen}{free}(x);} \\
+\end{tabular}
+\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
+Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
+\label{fig:overrun-notenough}
+\end{figure}
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+	1 & & \texttt{z = malloc(42);} \\
+	2 & & \texttt{// TODO bounds check??} \\
+	3 & & \texttt{x2 = \&z[50];} \\
+	4 & \texttt{x1 = malloc(...);} & \\
+	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	8 & \texttt{\hilight{olivegreen}{free}(x);} \\
+\end{tabular}
+\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
+\label{fig:overrun-goal}
+\end{figure}
+
+{\bf Being Owned.}
+The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
+Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
+
+The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
+If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
+this is because dpor ignores malloc's internal accesses from the shm conflict relation.
+
+what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
+BUT, the original one was not sound with respect to this bug to begin with.
+
+then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
+
+is that good enough?
+
+if not,
+Hopefully you can think of another way to resolve this later.
+
+\begin{lemma}
+	asdfa
+	\label{lem:leia} % "there is another"
+\end{lemma}
 

### 2015-10-26 pldi-16/di.tex
do some anonymizing

@@
 	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
+		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()} or {\tt cond\_wait()}.
 		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
@@
 It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
-\landslide~implements this check by adding a generation counter to its heap state tracking:
+We implemented this check by adding a generation counter to \landslide's heap tracking:
 each allocation is given a unique ID,

### 2015-10-26 pldi-16/evaluation.tex
buncha minor revisions
do some anonymizing

@@
 
-We also counted the number of free-re-malloc false positives that \landslide~suppressed.
+We also counted the number of malloc-recycle false positives that \landslide~suppressed.
 
@@
 To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
-For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must,
-rather than waiting until future interleavings to confirm them (as explained in \sect{\ref{sec:free-re-malloc}}).
+For example, we reported malloc-recycle data races during the first interleaving as {\em deterministic}, as a single-pass analysis must,
+rather than waiting until future interleavings to confirm them (as explained in \sect{\ref{sec:recycle}}).
 
@@
 which can indicate a more complicated concurrent algorithm requiring more rigorous verification than a program with a single global lock.
-Hence these corner cases are important to consider for future work.
+%Hence these corner cases are important to consider for future work.
 While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,

### 2015-10-26 pldi-16/paper.tex
write about free-re-malloc and add figure 1
write algorithm 1, GetBestJob
minor revisions
do some anonymizing

@@
 \usepackage{enumitem}      % adjust spacing in enums
+\usepackage[linesnumbered,ruled]{algorithm2e}
+\usepackage{algpseudocode}
+\usepackage{graphicx}
+%\usepackage{setspace}
 \usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
@@
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+%\doublespacing
 
@@
 \newcommand{\sect}[1]{\S #1}
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{brickred}{RGB}{192,0,0}
 
@@
 \newcommand\numthrlibs{79}
-\newcommand\numpintoses{{\Large \bf NUMPINTOSES}}
-\newcommand\numstudence{{\Large \bf NUMSTUDENCE}} % total pintoses plus p2s
+\newcommand\numpintoses{79}
+\newcommand\numstudence{158} % total pintoses plus p2s
 
@@
 and find crashes caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
+% TODO: Not true; Replace instead with a claim about partial verification.
 Compared to SSS-MC, we find bugs faster in smaller subset state spaces,
@@
 Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
-We have also posted the log files from all our experiments.
+We have also posted our evaluation's full data set and logs.
 They can be found here:
 
-\url{https://github.com/bblum/landslide} % TODO: put QS in its own repository
+% TODO: Make data set, at least, public.
+TODO: Data set at least can be public.
 
-\url{https://github.com/bblum/landslide} % TODO: put final logs somewhere
+{\em github links removed for double-blind review}
+%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put QS in its own repository
+%
+%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put final logs somewhere
+% POST-SUBMISSION TODO: Don't forget to anonymize the names of all your grupos.
 
@@
 
-Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
+%Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
+Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
 Thanks to Wind River for the use of their simulator \simics, upon which \landslide~is built.
-Thanks to Ji\v{r}\'{i} \v{S}im\v{s}a and the anonymous PLDI reviewers for their helpful comments.
+Thanks to
+% TODO: Post-double-blind-review
+% Ji\v{r}\'{i} \v{S}im\v{s}a and
+the anonymous PLDI reviewers for their helpful comments.
 This work was supported in part by % TODO: funding sources
+the U.S. Army Research Office under grant number W911NF0910273.
 

### 2015-10-26 pldi-16/proof.tex
Han shot first.

@@
 
-% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
-\newcommand\landslide{\textsc{Landslide}}
+%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
 \newcommand\quicksand{\textsc{Quicksand}}
@@
 \definecolor{brickred}{RGB}{192,0,0}
-
-% TODO: Finalize these numbers.
-\newcommand\numthrlibs{79}
-\newcommand\numpintoses{79}
-\newcommand\numstudence{158} % total pintoses plus p2s
 
@@
 the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
-Because pure-happens-before data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
+Because pure-HB data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
 However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock, our malloc-recycle false positives are indistinguishable from such false negatives.
+Indeed, because most concurrent malloc implementations are protected by a lock,
+our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
 }.
@@
 In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether they are benign or buggy.
+It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
 
@@
 	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 	3 & & \texttt{// x's memory recycled} \\
@@
 	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 	5 & & \texttt{// x's memory recycled} \\
@@
 	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x);} & \\
+	1 & \texttt{publish(x1);} & \\
 	2 & & \texttt{x2 = get\_published\_x();} \\
@@
 	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 \end{tabular}
@@
 and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause MHB (either directly, or because they are built on top of mutexes).
+Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
 
@@
 
-{\bf Low overhead.}
-We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
+%{\bf Low overhead.}
+%We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
+%Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
 
 {\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as any tendency to reuse blocks or allocate adjacent ones.
+We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
 %particular allocation pattern regarding adjacency/coalescing/reuse.
@@
 but should you wish to check for bugs like this,
-we would recommend a data-flow analysis which is much cheaper than model checking anyway.
+%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
+symbolic execution \cite{klee} would be more appropriate.
 
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the ``black box'' assumption.
+For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
 
@@
 \begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, then a non-malloc-recycle data race or a use-after-free bug will be identified.
+	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
+	then a non-malloc-recycle data race or a use-after-free bug will be identified.
 	\label{lem:reorder}
@@
 Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same allocation.
+	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
 	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
@@
 \end{proof}
+
+The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
+However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
+%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
+In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
+and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
+So, WLOG, we say that the only relevant events are the ones we mention explicitly.
 
@@
 \begin{lemma}[Greedo]
-	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads.
+	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
 	\label{lem:greedo}
@@
 By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1}. Lemma~\ref{lem:reorder} finishes.
+By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
+and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
+Lemma~\ref{lem:reorder} finishes.
 \end{proof}
@@
 \begin{lemma}[Han]
-	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads.
+	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
 	\label{lem:han}
@@
 \begin{proof}
-This is the hard case. We must prove there is a PP... TODO
+We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
+Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
+
+%\begin{figure}[t]
+%	\small
+%\begin{tabular}{rll}
+%	& Thread 1 & Thread 2 \\
+%	0 & \texttt{lock();} & \\
+%	1 & \texttt{publish(x1);} & \\
+%	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+%	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+%	0 & \texttt{unlock();} & \\
+%	0 & & \texttt{lock();} \\
+%	2 & & \texttt{x2 = get\_published\_x();} \\
+%	0 & & \texttt{unlock();} \\
+%	5 & & \texttt{// x's memory recycled} \\
+%	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+%	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+%\end{tabular}
+%\caption{If the locksets are the same in T1 during publish and x1, you can't avoid the data race on publish+get.}
+%\label{fig:recycle-bug}
+%\end{figure}
+
+We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
+If so, then by the maximal state space assumption, the event will be a PP, and we are done.
+If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
+For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
+which we'll call {\tt p}.
+By the previous two statements, {\tt p} must be a data race of its own.
+%(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+
+	%1 & \texttt{p1 = \hilight{olivegreen}{malloc}(...);} & \\
+	1 & \texttt{\hilight{brickred}{p1->ptr = x1;}} & \\
+	2 & \texttt{publish(p1);} & \\
+	3 & \texttt{\hilight{olivegreen}{free}(p1);} & \\
+	4 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	5 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+
+	6 & & \texttt{p2 = get\_published\_p();} \\
+	7 & & \texttt{// p's memory recycled} \\
+	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
+	9 & & \texttt{\hilight{brickred}{x2 = p2->ptr;}} \\
+
+	10 & & \texttt{// x's memory recycled} \\
+	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	12 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+\end{tabular}
+\caption{If the accesses used to publish {\tt x}'s address are a data race, their PPs may also be eliminated under the malloc-recycle pattern.
+Induction on the pointer structure leading to {\tt x} handles this case.}
+\label{fig:induction}
+\end{figure}
+
+Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
+as shown in Figure~\ref{fig:induction},
+we do not necessarily get the PP for free here.
+In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
+In \quicksand, data race PPs are not used immediately, but rather generate new state spaces to explore in the future. Anyway, under the maximal state space assumption, we will get to it eventually.}.
+We handle this with induction on T2's pointer chain leading to {\tt x}.
+
+\newcommand\publish[1]{{\tt p}$_{#1}$}
+\begin{itemize}
+	\item For the base case, the publish location \publish{0} is either in global memory, or shared directly using synchronization.
+		Non-heap memory data races are not subject to the malloc-recycle pattern, so will always get a data-race PP,
+		and use of synchronization always gets a PP in the maximal state space.
+	\item For the inductive step, a pointer \publish{n} is published in some heap memory \publish{n-1}\texttt{->ptr},
+		and we assume that however \publish{n-1} is shared to T2,
+		there will be a PP there
+		% where \publish{n-1} is shared
+		sufficient to make the \publish{n-1}\texttt{->ptr} access not malloc-recycle after DPOR.
+		%DPOR will test some interleaving in which T1's
+		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
+		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
+		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
+		% i.e., so that the p_n access no longer appears to be malloc-recycle.
+\end{itemize}
+
+Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
+a PP will be identified on the publish location {\tt p}.
+Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
+% XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
+As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
+Lemma~\ref{lem:reorder} finishes.
+%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
+
 \end{proof}
@@
 \end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	6 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 	7 & & \texttt{// x's memory recycled} \\
@@
 \end{figure}
+
+% TODO: Replace this with some real discussion.
 
@@
 
+% TODO
+% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
+% for the purpose of iterative deepening, MHB is
+% better because it will find the data races much sooner, rather than requiring you
+% to explore the maximal state space before the race can be uncovered
+
 %\subsection{Malloc with internal mutexes}

### 2015-10-28 pldi-16/di.tex
This spells QED so we are done.

@@
 If they cannot be reordered at all, it is a {\em false positive}.
-For brevity, we count benign data races as a subset of false positives.
+%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
+%For brevity, we count benign data races as a subset of false positives.
 
@@
 
-\subsection{Quicksand} %Managing multiple state spaces
+\subsection{Quicksand}
 
@@
 Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
+We seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}.
 
@@
 % If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-% TODO: decide which subsection should come first for better flow
 
@@
 
-TODO: Here will be a proof sketch that, during MC only, that mechanism is safe, because in the Figure~\ref{fig:recycle-bug} scenario, we are sure DPOR will reorder away the {\tt free} in a future interleaving.
+%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
+% FIXME: I'm really not satisfied with the flow of this paragraph.
+Note that under a {\em pure} happens-before analysis,
+these accesses are not concurrent at all
+because of {\tt malloc}'s internal locking events,
+and would not result in such false positives.
+However, as \cite{hybriddatarace} and \cite{tsan} argue,
+the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
+which is more appropriate in our context.
 
-% TODO: Figure out why they claim "Happens before produces NO false positives, only benign races".
-% It seems impossible.
-% But if true, it means either (a) they can somehow identify FRM DRs on the 1st pass, not needing to replay,
-% or (b) reuse of memory by malloc is somehow outside their concurrency model.
+Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning these false positives is sound even when such adversarial programs are considered.
+This makes it unnecessary to verify such data races by actively preempting threads during them,
+achieving a potentially combinatorial reduction in how many state spaces we generate.
+%Intuitively, we need not worry about cases such as Figure~\ref{fig:recycle-bug} because,
+%should they be true races,
+%DPOR will reorder threads sufficiently for the malloc-recycle pattern to disappear.
+We provide a proof sketch below.
 
-% TODO: proof goes here
-% TODO: example codez with malloc re free go here
+\renewcommand\proofname{Proof Sketch}
+
+\begin{theorem}[Soundness of eliminating malloc-recycle races]
+If a malloc-recycle data race is not a false positive,
+%DPOR will reorder threads such that
+DPOR will test an alternate thread interleaving in which
+%either
+the accesses still race without fitting the malloc-recycle pattern.
+%, or a use-after-free bug will be reported immediately.
+\end{theorem}
+
+\begin{proof}
+By definition of the malloc-recycle pattern,
+any such program must contain an access {\tt x1} by one thread T1,
+followed by a {\tt free} and a {\tt malloc} possibly by either thread,
+followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
+We say WLOG that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
+We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
+hence, there must also be some ``publish'' action {\tt p} by T1 which communicates the address to T2.
+Because this is a true data race, {\tt p} must occur before {\tt x1}, as {\tt x2} cannot be reordered before {\tt p}.
+
+We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
+The publish action must involved some thread communication, whether through a shared data structure or message-passing API.
+If locking or message-passing is used, our set of hard-coded PPs suffices.
+Otherwise, {\tt p} (and the corresponding read by T2) will be a data race, although it may itself be a malloc-recycle race.
+In this case we use induction on the pointer chain leading to {\tt x}:
+in the base case, {\tt p} is global or obtained via message-passing,
+and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on {\tt p}.
+Hence there will be a PP between {\tt p} and {\tt x1} no matter the mode of communication.
+
+By definition of DPOR, this PP causes {\tt x2} to be reordered before {\tt x1} while not changing {\tt x2}'s location.
+As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
+Hence {\tt x1} and {\tt x2} will be in the same allocation;
+hence the accesses can race without fitting the malloc-recycle pattern.
+% Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
+% This spells QED so we are done.
+\end{proof}
+
+Our full proof
+{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
+discusses our assumptions explicitly and provides the similar cases in more detail.
+\renewcommand\proofname{Proof}
+
+%For further discussion and for the full proof, refer to
+%{\em our supplementary material which we will cite as a tech report in the final version of the paper}.

### 2015-10-29 pldi-16/di.tex
touch up proof

@@
 
-For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
+For the rest of the paper, we abbreviate {\em preemption point} (PP),
+%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
 {\em model checking} (MC),
-{\em single-state-space model checking} (SSS-MC) (i.e., the approach of prior work),
-and {\em state space estimate} (ETA).
+{\em single-state-space MC} (SSS-MC) (i.e., the approach of prior work),
+Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
 
@@
 We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
-\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation, and a hybrid lockset/happens-before data-race analysis.
+\landslide~features DPOR,
+%\cite{dpor},
+state space estimation, and a hybrid lockset/happens-before data-race analysis.
 It avoids state space cycles (e.g.
@@
 	1 & \texttt{publish(x);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+	3 & \texttt{\hilight{olivegreen}{free}(x);} \\
+	4 & & \texttt{x2 = get\_published\_x();} \\
 	5 & & \texttt{// x's memory recycled} \\
@@
 %Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
-% FIXME: I'm really not satisfied with the flow of this paragraph.
-Note that under a {\em pure} happens-before analysis,
-these accesses are not concurrent at all
+Note that under a {\em pure happens-before} analysis,
+these accesses are not considered concurrent % at all
 because of {\tt malloc}'s internal locking events,
 and would not result in such false positives.
-However, as \cite{hybriddatarace} and \cite{tsan} argue,
-the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
-which is more appropriate in our context.
+However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
+so in our context it is more appropriate to use the
+{\em limited happens-before} relation in a hybrid approach with lockset tracking.
+%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
 
@@
 We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
-The publish action must involved some thread communication, whether through a shared data structure or message-passing API.
+The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
 If locking or message-passing is used, our set of hard-coded PPs suffices.
@@
 
+Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
+which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
 Our full proof

### 2015-10-29 pldi-16/evaluation.tex
Write eval-SSS-MC and put preliminary graph

@@
 \begin{enumerate}
-	\item Does Iterative Deepening find bugs faster than SSS-MC in subset state spaces, even without data-race PPs?
+
+	\item Does \quicksand~improve upon state-of-the-art MC?
+		\begin{enumerate}
+			\item Does Iterative Deepening find bugs faster
+				%than SSS-MC
+				in subset state spaces, even without data-race PPs?
 			% Probably not... ICB is state of the art here.
 			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
-	\item Is combining data-race analysis with MC productive?
-		\begin{enumerate}
 			\item Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
@@
 				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
+		\end{enumerate}
+	\item Does MC improve the accuracy of data-race detection?
+		\begin{enumerate}
 			\item Do we avoid false positives compared to a single-execution data race analysis?
@@
 For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} mutex.
+% XXX: Some pintoses can't run all the tests. So this number is too high.
 In total, this test suite comprises 632 unique state spaces.
-
 All tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
-\quicksand~tests were given 10 CPUs for 1 hour each.
-To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs on sync primitives enabled in advance (and no data-race PPs).
-% TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
-Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours,
-%then dividing all associated times by 10 (simulating perfect parallelism).
-and multiplying \quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
-% TODO: Can you make quicksand actually count the CPU hours. If so use the following instead:
-% and instrumenting \quicksand~to report total CPU-hours rather than wall-clock time.
 
@@
 
-\subsection{Evaluating Iterative Deepening}
-\label{sec:eval-id}
+\subsection{Comparing Iterative Deepening to SSS-MC}
+\label{sec:eval-sssmc}
 
-% TODO: If you have time, rerun all the quicksand experiments JUST running the 4 base state spaces. Give it 10/4 hours of cpu budget on 4 cores.
+% TODO: If you have time, rerun all the quicksand experiments JUST running the 4 base state spaces. Give it 10/4 hours of cpu budget on 4 cores. John's expt.
 
-First we compare \quicksand~to SSS-MC
 % mention exactly which state spaces we are comparing here
-
 % mention partial verification in terms of state space completion, when SSSMC times out.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Finding data-race bugs missed by SSS-MC}
-\label{sec:eval-drbugs}
 
@@
 
+To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs on sync primitives enabled in advance (and no data-race PPs).
+We gave each \quicksand~test 10 CPUs for 1 hour each. % XXX
+% TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
+Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours,
+%then dividing all associated times by 10 (simulating perfect parallelism).
+and instrumenting \quicksand~to report total CPU-hours rather than wall-clock time.
+%\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
+Figure~\ref{fig:dowefindbugsfaster} plots the bug-finding speed of SSS-MC against that of two different \quicksand~experiments:
+%which we explain presently.
+
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster.pdf}
+	\caption{Comparison of bug-finding performance
+	by several configurations of \quicksand~and the SSS-MC control.
+	\quicksand~finds 169\% as many bugs with data-race PPs.}
+	\label{fig:dowefindbugsfaster}
+\end{figure}
+
+{\bf Finding the same bugs faster.}
+To show that Iterative Deepening is effective even for MC domains without data races, such as message-passing distributed systems,
+we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs (i.e., ignoring all data-race reports)\footnote{
+Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
+we ran these tests for 2.5 hours on 4 CPUs each.
+Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
+%
+The line QS-no-DR-PPs represents this experiment;
+% TODO: rephrase based on result of expt
+we see that even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
+QS-no-DR-PPs finds many more of the bugs much sooner.
+From this we conclude that for smaller arbitrary CPU budgets,
+Iterative Deepening is likely to find bugs SSS-MC will miss,
+and programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
+
+{\bf Finding new data-race bugs.}
 Though state-of-the-art MCs preempt only on synchronization events, many serious concurrency bugs are caused by data races leading to corrupted shared state.
-Hence, in Table~\ref{tab:drbugs} we count how often \quicksand~uncovered a bug only in state spaces which included data-race PPs,
-while
+The line QS-DRs represents our ``no holds barred'' \quicksand~tests:
+we quickly pull ahead of SSS-MC, and ultimately conclude with 69\% more bugs in total.
+The break-even point is at a negligible 45 seconds.
 
-In Table~\ref{tab:allbugs} .... %TODO
+Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
+which represents only the bugs found in state spaces without data-race PPs (like QS-no-DR-PPs, but even when data-race PPs are enabled).
+Intuitively, this line shows that for programs with only benign data races,
+\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC\footnote{
+The initial perfect overlap between QS-DRs and QS-no-DR-bugs indicates how long it takes before the first data-race bug is found.}.
+%even after the extra overhead of verifying them, \quicksand~still slightly edges out SSS-MC
+
+{\bf Partial verification guarantees.}
+TODO % TODO
+
+%Hence, in Table~\ref{tab:drbugs} we count how often \quicksand~uncovered a bug only in state spaces which included data-race PPs, while
+%
+%In Table~\ref{tab:allbugs} ....
 
@@
 {\bf Testing lock implementations.}
-In \sect{\ref{sec:eval-drbugs}}, using data-race PPs compares favourably across the board to SSS-MC.
+In \sect{\ref{sec:eval-sssmc}}, using data-race PPs compares favourably across the board to SSS-MC.
 Observe in particular that in \mxtest, the control experiment found dramatically fewer bugs (just 1),
@@
 {\bf Partial verification.}
-We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-id}}).
+We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
 While we guarantee safety when preempted on certain combinations of PPs,

### 2015-10-29 pldi-16/evaluation.tex
write about partial veirf

@@
 \cite{pintos}.
+
 % P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
@@
 For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} mutex.
+All tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+
+We chose P2s and Pintoses for our test suite because of the relative ease of generating hundreds of unique state spaces,
+varied in size and correctness, and with a diverse set of bug types.
 % XXX: Some pintoses can't run all the tests. So this number is too high.
-In total, this test suite comprises 632 unique state spaces.
-All tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+In total, the test suite comprises 632 unique state spaces, at least 106 of which contain bugs.
+% TODO: Is this ok? Too arrogant?
+We believe this test suite's size allows for a more statistically significant comparison among MC strategies,
+compared to anecdotally showing a small handful of new bugs that could be found in real-world programs.
 
@@
 {\bf Partial verification guarantees.}
-TODO % TODO
+We know of no approach to quantify the probability, when a state space times out,
+that a bug would be exposed by an untested interleaving.
+\quicksand~does not attempt any such quantitative partial verification guarantee,
+but can at least report which subsets of PPs produced state spaces that could be completed.
+
+The SSS-MC control experiment timed out in 10 hours on 77 tests.
+For these, we counted the number of state spaces \quicksand~was able to complete.
+In two cases, \quicksand~also failed to complete anything; beyond these,
+between 1 and 508 state spaces could be completed, with a mean of 72.8 and median of 30.
+These completed state spaces guarantee that, if the test case could expose a bug,
+% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
+it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
 

### 2015-11-01 pldi-16/di.tex
buncha rephrasing

@@
 Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning these false positives is sound even when such adversarial programs are considered.
-This makes it unnecessary to verify such data races by actively preempting threads during them,
+This makes it unnecessary to verify such data races by actively adding more preemptions,
 achieving a potentially combinatorial reduction in how many state spaces we generate.
@@
 followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
-We say WLOG that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
+For brevity we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
 We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};

### 2015-11-01 pldi-16/evaluation.tex
buncha rephrasing

--- 2015-10-29 pldi-16/evaluation.tex
@@
 % 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
-Though not ``real world'' programs, both projects are quite large: % maybe "complex"?
+Though not ``real world'' programs, both projects are quite complex:
 the P2s average 1807 lines of C and x86 assembly (stddev 489.5),
@@
 and the Pintoses average {\bf 9999999} % TODO
+
+We chose P2s and Pintoses for our test suite because of the relative ease of generating hundreds of unique state spaces,
+varied in size and correctness, and with a diverse set of bug types\footnote{
+Many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested);
+these we fixed by hand in advance to ensure that every bug in our study required some meaningful work by the MC.}.
+% TODO: Is this ok? Too arrogant?
+We believe that merely finding a small handful of new real-world bugs is too anecdotal,
+and that our test suite's size allows for a more statistically significant comparison among MC and data-race testing strategies,
+%compared to anecdotally showing a small handful of new bugs that could be found in real-world programs.
 
@@
 For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} mutex.
+Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
 % XXX: Some pintoses can't run all the tests. So this number is too high.
-In total, this test suite comprises 632 unique state spaces.
-All tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+In total, the test suite comprises 632 unique state spaces, at least 106 of which contain bugs.
 
-\begin{table}[t]
-	\begin{tabular}{l|l|l}
-			& QS bugs & SSS-MC bugs \\
-		\hline
-		\mxtest & eg 1000 & eg 0 \\
-		\bct & & \\
-		etc... & & \\
-		\hline
-		Total & & \\
-	\end{tabular}
-	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom 2)}
-	\label{tab:allbugs}
-\end{table}
-
-\begin{table}[t]
-	\small
-	\begin{tabular}{l|l|l||l|l}
-	& QS bug & \begin{tabular}{c} SSS-MC \\ completed\end{tabular}
-	& QS bug & \begin{tabular}{c}SSS-MC \\ timeout \end{tabular} \\
-		\hline
-		\mxtest & e.g. 5 & 10 & 0 & 0 \\
-		\bct & & & & \\
-		etc... & & & & \\
-		\hline
-		Total & & & & \\
-	\end{tabular}
-	\caption{Bugs requiring data-race PPs to expose, found by \quicksand~but missed by the single-state-space approach.}
-	\label{tab:drbugs}
-\end{table}
+%\begin{table}[t]
+%	\begin{tabular}{l|l|l}
+%			& QS bugs & SSS-MC bugs \\
+%		\hline
+%		\mxtest & eg 1000 & eg 0 \\
+%		\bct & & \\
+%		etc... & & \\
+%		\hline
+%		Total & & \\
+%	\end{tabular}
+%	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom 2)}
+%	\label{tab:allbugs}
+%\end{table}
+%
+%\begin{table}[t]
+%	\small
+%	\begin{tabular}{l|l|l||l|l}
+%	& QS bug & \begin{tabular}{c} SSS-MC \\ completed\end{tabular}
+%	& QS bug & \begin{tabular}{c}SSS-MC \\ timeout \end{tabular} \\
+%		\hline
+%		\mxtest & e.g. 5 & 10 & 0 & 0 \\
+%		\bct & & & & \\
+%		etc... & & & & \\
+%		\hline
+%		Total & & & & \\
+%	\end{tabular}
+%	\caption{Bugs requiring data-race PPs to expose, found by \quicksand~but missed by the single-state-space approach.}
+%	\label{tab:drbugs}
+%\end{table}
 
@@
 
+% XXX: I hate this paragraph. I have no idea what order to put the sentences in.
 To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs on sync primitives enabled in advance (and no data-race PPs).
 We gave each \quicksand~test 10 CPUs for 1 hour each. % XXX
-% TODO figure out somewhere to mention what landslide's pps are: mx lock/unlock (aka sema up/down)
 Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours,
@@
 %\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
-Figure~\ref{fig:dowefindbugsfaster} plots the bug-finding speed of SSS-MC against that of two different \quicksand~experiments:
-%which we explain presently.
+Figure~\ref{fig:dowefindbugsfaster} plots a cumulative distribution of bugs found before a given elapsed CPU-time. % during its associated test.
+We tested several configurations of \quicksand~to compare against SSS-MC from multiple angles,
+which we explain presently.
 
@@
 To show that Iterative Deepening is effective even for MC domains without data races, such as message-passing distributed systems,
-we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs (i.e., ignoring all data-race reports)\footnote{
+we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
+(i.e., ignoring all data-race candidates)\footnote{
 Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
@@
 %
-The line QS-no-DR-PPs represents this experiment;
-% TODO: rephrase based on result of expt
-we see that even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
+The line QS-no-DR-PPs represents this experiment.
+Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
 QS-no-DR-PPs finds many more of the bugs much sooner.
@@
 Iterative Deepening is likely to find bugs SSS-MC will miss,
-and programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
+and for more ambitiously-sized tests,
+programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
+%We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
+%Iterative Deepening is likely to find bugs SSS-MC will miss.
+%Moreover, it is also easy to imagine scaling up the size of each test case to test ,
+%using more threads or longer sequences of API calls.
+%We hope that is compelling even to users willing to spend many CPU-hours on testing.
+%These results show explicitly that for arbitrary CPU-time budgets
+
+%Note that even users willing to spend 10 CPU-hours should find this ``early advantage'' compelling:
 
@@
 Intuitively, this line shows that for programs with only benign data races,
-\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC\footnote{
-The initial perfect overlap between QS-DRs and QS-no-DR-bugs indicates how long it takes before the first data-race bug is found.}.
+\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
+%\footnote{The initial perfect overlap between QS-DRs and QS-no-DR-bugs indicates how long it takes before the first data-race bug is found.}
 %even after the extra overhead of verifying them, \quicksand~still slightly edges out SSS-MC
 
-{\bf Partial verification guarantees.}
-TODO % TODO
+{\bf Partial verification.}
+% TODO: But what about state space estimation? Does that count as probability? I don't think it does. But can you mention it?
+When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
+Beyond using the state space estimator to guess the percent coverage,
+we know of no approach to quantify the probability
+that a bug would be exposed by an untested interleaving.
+\quicksand~does not attempt any such quantitative guarantee,
+but can at least report which subsets of PPs produced state spaces that completed in time.
+
+The SSS-MC control experiment timed out in 10 hours on 77 tests.
+Among these, \quicksand~found data-race bugs in 15.
+For the other 62, we counted the number of state spaces \quicksand~was able to complete and the number of unique PPs among them.
+In two cases, \quicksand~also failed to complete anything; beyond these,
+between 1 and 253 state spaces were completed, with a mean of 58.9 and median of 29.
+Between 0 and 233 unique PPs were verified among some subsets, with a mean of 27.2 and median of 7.5.
+These completions guarantee that, if the test case could expose a bug,
+% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
+it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
 
@@
 We also counted the number of malloc-recycle false positives that \landslide~suppressed.
+
+TODO
 
@@
 
+TODO: Present results.
+
 % TODO: restructure this paragraph to account for RaceFuzzer/Portend. Be like, "Yes they explore multiple schedules, but only AFTER finding a race." And/or make this argument in prior work
@@
 We argue this comparison is meaningful because prior work data-race tools, being not integrated with a MC,
-are not intended to uncover different results under subsequent runs.
-One could run a data-race tool repeatedly for 10 CPU-hours, but the advantage of stateless MC over stress testing is already well-understood.
+are not intended to discover new data race candidates under subsequent runs.
+Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
+but the reliability problem of stress testing compared to stateless MC is already well-understood
+% TODO: 1 more good citation here
+\cite{chess-icb,gambit}.
+%One could run a data-race tool repeatedly for 10 CPU-hours, but the advantage of stateless MC over stress testing is already well-understood.
 % TODO TODO TODO get a big citation for teh above sentence.

### 2015-11-01 pldi-16/proof.tex
more rephrasing

@@
 so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
+An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
+In that case, our point is to justify deprioritizing those jobs so aggressively.
 
@@
 
-{\bf Being Owned.}
-The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
+%{\bf Being Owned.}
+%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
 Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-
-The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
+%
+%
+%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
 If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
@@
 then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
-
-is that good enough?
-
-if not,
-Hopefully you can think of another way to resolve this later.
 

### 2015-11-02 pldi-16/proof.tex
touch up proof at carlo's suggestion

@@
 \begin{definition}[Shared memory conflict]
-A pair of memory accesses between two threads where at least one of them is a write.
+A pair of memory accesses between two threads to the same address where at least one of them is a write.
 \end{definition}
@@
 \label{sec:proof}
+
+We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
+Hence, our soundness statement is as follows:
+
+\begin{theorem}[Soundness of eliminating malloc-recycle races]
+	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
+\end{theorem}
 
@@
 
+\setcounter{theorem}{0}
 \begin{theorem}[Soundness of eliminating malloc-recycle races]
@@
 
+\section{Acknowledgements}
+
+{\em removed for double blind review}
+%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
+
 \bibliographystyle{abbrvnat}

### 2015-11-02 pldi-16/di.tex
Talk about cli/sti which I still need to actually implement.

@@
 state space estimation, and a hybrid lockset/happens-before data-race analysis.
+% TODO: \cite{fairstatelessmc} instead or in addition?
 It avoids state space cycles (e.g.
@@
 Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
-We seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}.
+
+{\bf Initial PPs.}
+For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}.
+By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
+such as condvars, semaphores, and r/w locks.
+
+For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
+so we also preempt just before {\tt cli} and just after {\tt sti}\footnote{
+During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock,
+as {\tt cli}'d memory accesses can still race with others that have interrupts on.}.
+% FIXME: "What happened to yield?" "Yield is implicitly included in each subset."
+\quicksand~is configured to use the subsets
+$\{ \}$,
+$\{lock\}$,
+$\{unlock\}$,
+$\{cli\}$,
+$\{sti\}$,
+and $\{lock,unlock,cli,sti\}$.
+Other subsets are obviously possible, but as a heuristic, we don't bother with them.
 

### 2015-11-02 pldi-16/evaluation.tex
Some rephrasings in eval; talk about avoiding repeated work

@@
 We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle system calls.
-For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} mutex.
+For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
 Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
@@
 Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
-but the reliability problem of stress testing compared to stateless MC is already well-understood
+but stress testing's comparative problem with achieving reliable coverage is already well-understood
 % TODO: 1 more good citation here
 \cite{chess-icb,gambit}.
+% TODO: read gambit paper
 %One could run a data-race tool repeatedly for 10 CPU-hours, but the advantage of stateless MC over stress testing is already well-understood.
@@
 
+% TODO Rename to "Limitations and Future Work"
 \subsection{Discussion and future work}
@@
 
+% TODO: Move this paragraph up into evaluation; this section should only talk about PROBLEMS FROM THE FUTURE
 {\bf Testing lock implementations.}
@@
 even compared to the other test cases\footnote{
-% FIXME: "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
+% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
 	The one bug SSS-MC found was in a fully-assembly lock implementation.
@@
 
-% TODO: get these numbers
+% TODO: Measure how much subset overlap there is.
+{\bf Avoiding redundant work.}
+When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
+Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
+Because we prioritize completing small state spaces before extending them with more PPs,
+the superset state spaces we run later are guaranteed to repeat each branch of their already-completed subsets.
+Among completed state spaces across our test suite, the proportion of repeated work was {\bf \large 999\%}.
+Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,
+but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
+to prune some or all of this repeated work.
+
 {\bf Finer-grained PP subsets.}
-\quicksand~was able to partially guarantee safety in {\large \bf 99999\%} of tests with large maximal state spaces.
-However, in {\large \bf 1337} tests, no more than the minimal state space could be verified,
-and in {\large \bf 42} tests, not even that much.
+% TODO: update these numbers
+\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
+However, in 2 tests, no more than the minimal state space could be verified,
+and in 2 other tests, not even that much.
 Larger state spaces often result from finer-grained locking,

### 2015-11-03 pldi-16/evaluation.tex
move mxtest discussion to eval not futurework

@@
 Though not ``real world'' programs, both projects are quite complex:
-the P2s average 1807 lines of C and x86 assembly (stddev 489.5),
-% Pintos SLOC stats: TODO
-and the Pintoses average {\bf 9999999} % TODO
+the P2s average 1807 lines of code,%C and x86 assembly (stddev 489.5),
+% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
+% Obtained with:
+%function getloc {
+%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
+%}
+%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
+%#for i in /tmp/shit/pintoses/daniel-deniz*; do
+%	echo -ne "$i";
+%	tloc=`getloc $i threads`
+%	uloc=`getloc $i userprog`
+%	dloc=`getloc $i devices`
+%	echo -e "\t$tloc\t$uloc\t$dloc"
+%done
+and the Pintoses average 718.
 
@@
 
+% TODO: Make a table with a breakdown by test case.
+
+%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
+% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
+% (as well as on the mutex API boundaries).
+% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
+In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
+even compared to the other test cases\footnote{
+% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
+	The one bug SSS-MC found was in a fully-assembly lock implementation.
+{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
+Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
+%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
+%Though for many applications of MC, assuming a correct lock implementation is sufficient,
+Though it suffices for many applications of MC to assume correctly-implemented locks,
+we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.
+
 {\bf Partial verification.}
@@
 % TODO Rename to "Limitations and Future Work"
-\subsection{Discussion and future work}
+\section{Discussion}
 \label{sec:future}
 
-%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
-% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
-% (as well as on the mutex API boundaries).
-% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
-
-% TODO: Move this paragraph up into evaluation; this section should only talk about PROBLEMS FROM THE FUTURE
-{\bf Testing lock implementations.}
-In \sect{\ref{sec:eval-sssmc}}, using data-race PPs compares favourably across the board to SSS-MC.
-Observe in particular that in \mxtest, the control experiment found dramatically fewer bugs (just 1),
-even compared to the other test cases\footnote{
-% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
-	The one bug SSS-MC found was in a fully-assembly lock implementation.
-{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
-Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
-%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
-%Though for many applications of MC, assuming a correct lock implementation is sufficient,
-Though it suffices for many applications of MC to assume correctly-implemented locks,
-we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.
+In this section, we discuss \quicksand's limitations and opportunities for future improvement.
 
@@
 Because we prioritize completing small state spaces before extending them with more PPs,
-the superset state spaces we run later are guaranteed to repeat each branch of their already-completed subsets.
-Among completed state spaces across our test suite, the proportion of repeated work was {\bf \large 999\%}.
+the superset state spaces we run later will always repeat each branch of their already-completed subsets.
+%
+We measured the proportion of repeated work among completed state spaces across our test suite;
+on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
 Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,
@@
 %enabling partial verification of such large tests. % if need space
+%Because our current implementation does not avoid repeated interleavings across state spaces,
+%as discussed above,
+%we were limited to a small number of very basic PP subsets to statically seed the exploration
+
+%{\bf Small test cases.}
 

### 2015-11-03 pldi-16/di.tex
some di rephrase

@@
 such as condvars, semaphores, and r/w locks.
+Preempting on voluntary switches such as {\tt yield} is necessary for the invariant that only one thread runs between consecutive PPs,
+so the {\tt yield} PP is always implicitly enabled.
 
 For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so we also preempt just before {\tt cli} and just after {\tt sti}\footnote{
+so (e.g. on x86) we also preempt just before {\tt cli} and just after {\tt sti}\footnote{
 During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock,
@@
 % FIXME: "What happened to yield?" "Yield is implicitly included in each subset."
-\quicksand~is configured to use the subsets
+\quicksand~is configured to begin with the subsets
 $\{ \}$,
@@
 and $\{lock,unlock,cli,sti\}$.
-Other subsets are obviously possible, but as a heuristic, we don't bother with them.
+As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
+which could potentially be improved in future work (\sect{\ref{sec:future}}).
 

### 2015-11-03 pldi-16/di.tex
Implement jiri's comments

@@
 The purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user need only provide their total CPU budget as a test parameter,
+so an inexpert user must provide only their total CPU budget as a test parameter,
 and to enable completing appropriately-sized state spaces within that budget.
@@
 {\em model checking} (MC),
-{\em single-state-space MC} (SSS-MC) (i.e., the approach of prior work),
+{\em single-state-space model checking} (SSS-MC),% (i.e., the approach of prior work),
 Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
 
+SSS-MC indicates the approach of prior model checking tools:
+the set of PPs is fixed in advance, and the tool commits to testing every interleaving available via those PPs.
+%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
+Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
+but dynamic analysis cannot add new PPs or remove ineffective ones.
+
 Please note the difference between data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is false-positive-prone,
+Because data-race analysis is prone to false-positives,
 we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
@@
 
-\subsection{Verifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?
+\subsection{Classifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?
 
@@
 	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
-		// common case: job will probably finish on time \\
+		// Common case: Job is expected to complete. \\
 		return $j_0$
@@
 	\ForEach{job $j_P \in$ \PendingJobs}{
-		// don't run a pending job if a subset of it is already suspended; it would find the same fate! \\
+		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
 		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
@@
 			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
-			// $j_S$ is tempting, but if a subset of it is also suspended, don't run the larger one first \\
+			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
 			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
@@
 	}
-	// $j_0$'s ETA was bad, but nothing else was better \\
+	// $j_0$'s ETA was bad, but nothing else was better. \\
 	return $j_0$
-	\caption{Suspending exploration of a too-large state space in favour of a potentially smaller one.}
+	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
 	\label{alg:shouldworkblock}
@@
 
-With a limited CPU budget, we must avoid running tests likely to time out.
+With a limited CPU budget, we must avoid running tests that are likely to time out.
 When a MC instance reports an ETA too high,
@@
 then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves after further time spent on it.
+Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
 %reveals that it might finish in time after all.
@@
 Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
-in case their ETAs are actually overestimates.
+in case their ETAs are actually overestimated.
 Lines 12-14 account for the
@@
 to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
-(it lets us avoid thrashing in \quicksand),
+(it lets us avoid thrashing in \quicksand).
 
@@
 We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
-\landslide~features DPOR,
+\landslide~implements DPOR,
 %\cite{dpor},
@@
 % this line can be cut if space is needed
-It can test both user- and kernel-level code, although is limited to timer-driven nondeterminism.
+It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
 Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
@@
 Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
-The {\tt without\_function} directive works similarly, but as a blacklist.
+%The {\tt without\_function} directive works similarly, but as a blacklist.
+The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
 Multiple invocations can be used; later ones take precedence.
@@
 {\bf Data races in lock implementations.}
-Prior work data race tools recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself \cite{tsan}.
+Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself.
 Assuming the locks are already correct enables productive data-race analysis on the rest of the codebase.
@@
 {\bf Communication protocol.}
-The interface to \landslide~, which any similar MC could implement, is in two parts.
-First, when starting each job, it writes a configuration file declaring which PPs to use,
+The interface to \landslide~, which any similar MC could implement, has two parts.
+First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
 % can lose this line due to space
@@
 passed as an argument to \landslide.
-Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing on a FIFO pipe.
-\landslide~sends a message after each new interleaving is tested to report updated progress and ETA,
+Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing.%on a FIFO pipe.
+\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
 whenever a new data-race candidate is found, and whenever a bug is found.
@@
 (though we cannot say for sure before preempting during the access itself, hence the heuristic),
-so we prioritize jobs with both-order data-race PPs more highly.
+so we prioritize jobs with both-order data-race PPs.
 For single-order races, we do not add PPs for the later access at all;
@@
 In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to decide at runtime which to prioritize.
+which motivates the need for Iterative Deepening to decide at runtime which jobs to prioritize.
 % TODO: Put numbers here.

### 2015-11-03 pldi-16/evaluation.tex
Implement jiri's comments

@@
 \newcommand\mxtest{\texttt{mx\_test}}
-\newcommand\tej{\texttt{thr\_exit\_join}}
-\newcommand\bct{\texttt{broadcast}}
-\newcommand\paraguay{\texttt{paraguay}}
-\newcommand\paradise{\texttt{paradise\_lost}}
+\newcommand\tej{\texttt{join\_test}}
+\newcommand\bct{\texttt{broadcast\_test}}
+\newcommand\paraguay{\texttt{signal\_test}}
+\newcommand\paradise{\texttt{sem\_test}}
 \newcommand\rwl{\texttt{rwl\_test}}
@@
 Larger state spaces often result from finer-grained locking,
-which can indicate a more complicated concurrent algorithm requiring more rigorous verification than a program with a single global lock.
+which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
+Such programs may require even more rigorous verification than a program with a single global lock.
 %Hence these corner cases are important to consider for future work.

### 2015-11-03 pldi-16/intro.tex
Implement jiri's comments

@@
 by forcing a program to execute each distinct interleaving,
-capturing and controlling this nondeterminism in a finite state space.
+capturing and controlling this nondeterminism using a finite state space.
 Unfortunately, these state spaces explode exponentially in the size of the input program.
@@
 %and benign races
-which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
+which require annotations or imprecise heuristics to suppress \cite{racerx,tsan,datacollider}.
 %
@@
 %
+Configuring a model checker is a tradeoff between schedule coverage and feasibility of completion.
 This work shows how to avoid making that tradeoff decision in advance.
@@
 \begin{enumerate}
-	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation;
-	\item A new tactic for eliminating one class of false-positive data races, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
+	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
+	\item A new tactic for eliminating one class of false-positive data races,
+		which cannot soundly be used in a single-pass analysis,
+		but which we prove correct when used with Iterative Deepening;
+		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
 	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?

### 2015-11-03 pldi-16/paper.tex
This spells QED so we are done.
more rephrasing
change num pintoses
Implement jiri's comments

@@
 \usepackage{graphicx}
+\usepackage{amsthm}
 %\usepackage{setspace}
@@
 %\doublespacing
+
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
 
@@
 
-% TODO: Finalize these numbers.
 \newcommand\numthrlibs{79}
-\newcommand\numpintoses{79}
-\newcommand\numstudence{158} % total pintoses plus p2s
+\newcommand\numpintoses{78} % 'di' is basecode. durr
+\newcommand\numstudence{157} % total pintoses plus p2s
 
@@
 Stateless model checking is a promising technique for testing concurrent programs,
-but is vulnerable to exponential explosion of the state space when the test input parameters are too large.
+but suffers from exponential explosion of the state space when the test input parameters are too large.
 Several partial-order reduction techniques exist for mitigating this explosion,
@@
 Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
-It avoids concerns of intractable state space size, but suffers from a high rate of false positives.
+It avoids concerns of intractable state space size, but is prone to false positives.
 
@@
 It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
-and it incorporates a data-race analysis to dynamically add new preemption points to verify each data race as buggy or benign.
+and it incorporates a data-race analysis to dynamically add new preemption points to classify each data race as buggy or benign.
 %
-Our evaluation shows this technique is
-more effective than single-state-space model checking, both at finding more bugs and at completing more state spaces when no bug exists.
+Our evaluation shows that combining data-race detection with stateless model checking can find new bugs neither technique would find on their own.
+%Our evaluation shows this technique is
+%more effective than single-state-space model checking
+%%both at finding more bugs and at completing more state spaces when no bug exists.
+%both at finding the same bugs faster and at finding new bugs entirely.
 
@@
 %by verifying the absence of bugs in their associated state spaces
-and find crashes caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
+and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
 % TODO: Not true; Replace instead with a claim about partial verification.
@@
 
-Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
+%Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
+\quicksand~is open-source and its interface can be adapted to fit any MC tool similar to \landslide.
 We have also posted our evaluation's full data set and logs.
@@
 Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
-Thanks to Wind River for the use of their simulator \simics, upon which \landslide~is built.
+Thanks to Wind River for the use of their simulator \simics.
 Thanks to
 % TODO: Post-double-blind-review
-% Ji\v{r}\'{i} \v{S}im\v{s}a and
+% Ji\v{r}\'{i} \v{S}im\v{s}a, John Wilkes, and
 the anonymous PLDI reviewers for their helpful comments.

### 2015-11-03 pldi-16/related.tex
minor revisions
Write eval-SSS-MC and put preliminary graph
buncha rephrasing
Implement jiri's comments

@@
 {\bf Synchronization events only.} CHESS \cite{chess} and dBug \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
-Hence they will miss any crashes that require interleaving threads at instruction granularity during a data race.
+Hence they will miss any bugs that require interleaving threads at instruction granularity during a data race.
 CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
@@
 However, the resulting state space explosion demands that any such tool either
-choose a small subset of instructions to consider as PPs.
+choose a small subset of instructions to consider as PPs
 or be limited to very small test inputs.
@@
 SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, SPIN explicitly stores visited program states rather than using DPOR, limiting the size of programs that can be practically tested,
-and requires code to be written in the PROMELA DSL.
+However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
+SPIN also requires code to be written in the PROMELA DSL.
 %so cannot check implementations directly.
@@
 
-Too many related projects to list have made contributions to the false-positive problem first introduced in \cite{eraser}.
+%Too many related projects to list have made contributions to the
+Many advances have been made on the false-positive data race problem first introduced in \cite{eraser}.
 \cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
@@
 % cut for space?
+% No, IDGAF about pure happens before.
+%FastTrack \cite{fasttrack} optimizes the performance of pure happens-
 
@@
 This work compares the immediately resulting program states for differences,
-preferring instead still to err on the side of false positives.
+preferring to err on the side of false positives.
 RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited,
@@
 which is tantamount to adding a new state space in \quicksand.
-Our proof in \sect{\ref{sec:recycle}} allows eliminating this special case with no replay required.
+Our proof in \sect{\ref{sec:recycle}} allows eliminating this special case with no additional replay beyond what DPOR already requires.
 
@@
 which is necessary to expose some bugs (\sect{\ref{sec:eval-falseneg}}).
-It also assumes the POSIX synchronization API, so cannot verify lock implementations such as we do with \mxtest.
-Future work could combine the two approaches, using \landslide~to produce new traces for Portend to classify, or using Portend's analysis to inform \quicksand's state space priorities.
+% XXX: Is this true??
+It also assumes the POSIX synchronization API, so cannot verify arbitrary synchronization algorithms such as we do with \mxtest.
+Future work could combine the two approaches, using MC to produce new data-race traces for Portend to classify, or using Portend's analysis to inform \quicksand's state space priorities.
 

### 2015-11-03 pldi-16/di.tex
write about classifying-drs, rearrange figures

@@
 SSS-MC indicates the approach of prior model checking tools:
-the set of PPs is fixed in advance, and the tool commits to testing every interleaving available via those PPs.
+the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
 %Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
 Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
-but dynamic analysis cannot add new PPs or remove ineffective ones.
+but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.
 
@@
 
+Iterative Deepening must be seeded with some initial state spaces,
+which can be any number of subsets of the statically-available PPs that SSS-MC would use.
+While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
+Because data races indicate access pairs that can interleave at instruction granularity,
+it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
+
+With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread.
+That state space may expose a failure, in which case we report a data-race bug,
+or complete successfully, which indicates a benign or false-positive data race.
+It may also uncover a new data-race candidate in some alternate interleaving,
+in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
+Because Iterative Deepening is constrained by a CPU budget,
+we may time out before completing a data race's associated state space,
+in which case it remains a potential false positive that the user must handle.
+
 \subsection{Choosing the best job}
 
-% TODO: put colour in this code
 \newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
@@
 \newcommand\GetPPSet[1]{PPSet(#1)}
+With a limited CPU budget, we must avoid running tests that are likely to time out.
+Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
+and a set of {\em pending} jobs (untested ones with unknown ETAs).
+When a MC instance reports an ETA too high,
+we compare other pending jobs to find another one more likely to complete in time.
+%
+Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
+Though its worst-case performance is $O(mn)$ in the
+%number of pending and suspended jobs,
+sizes of $\mathcal{P}$ and $\mathcal{S}$,
+in practice the non-constant portion beyond line 4 runs very infrequently
+and is negligible compared to the exponentially-sized state spaces.}.
+Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+and $j_1$ is suspended,
+then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
+Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
+%reveals that it might finish in time after all.
+Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
+
+% TODO: put colour in this code
 \begin{algorithm}[t]
@@
 \end{algorithm}
-
-With a limited CPU budget, we must avoid running tests that are likely to time out.
-When a MC instance reports an ETA too high,
-we compare other yet-untested PP sets to find another job more likely to complete in time.
-Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
-Though its worst-case performance is $O(mn)$ in the
-%number of pending and suspended jobs,
-sizes of $\mathcal{P}$ and $\mathcal{S}$,
-in practice the non-constant portion beyond line 4 runs very infrequently
-and is negligible compared to the exponentially-sized state spaces.}.
-Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
-and $j_1$ is suspended,
-then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
-%reveals that it might finish in time after all.
-Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
 
@@
 
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
+	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
+		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
+	\label{fig:new-dr-jobs}
+\end{figure}
+
 When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
@@
 
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
-	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
-		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
-	\label{fig:new-dr-jobs}
-\end{figure}
-
 \subsection{Suppressing ``malloc-recycle'' false positives}
@@
 Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
+
+% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
+By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
+
+It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
+We implemented this check by adding a generation counter to \landslide's heap tracking:
+each allocation is given a unique ID,
+and when evaluating whether two heap accesses can race,
+the IDs of their containing blocks must match.
+However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
+Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
+Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
+Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
 
@@
 \end{figure}
-
-% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
-By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
-
-It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
-We implemented this check by adding a generation counter to \landslide's heap tracking:
-each allocation is given a unique ID,
-and when evaluating whether two heap accesses can race,
-the IDs of their containing blocks must match.
-However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
-Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
-Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
-
 \begin{figure}[t]

### 2015-11-03 pldi-16/evaluation.tex
some rephrasing and start to make a big table

@@
 %
-The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condvars, semaphores, and r/w locks;
+The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
 all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
@@
 varied in size and correctness, and with a diverse set of bug types\footnote{
-Many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested);
-these we fixed by hand in advance to ensure that every bug in our study required some meaningful work by the MC.}.
+Many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested).
+We fixed these by hand in advance to ensure that every bug in our study required some meaningful work by the MC.}.
 % TODO: Is this ok? Too arrogant?
@@
 \newcommand\tej{\texttt{join\_test}}
-\newcommand\bct{\texttt{broadcast\_test}}
+\newcommand\bct{\texttt{bcast\_test}}
 \newcommand\paraguay{\texttt{signal\_test}}
@@
 % each tailored to exercise a different part of the P2 project
-\mxtest, for locking algorithm correctness, \tej, a test of thread lifecycle, \bct~and \paraguay, for condvars, \paradise~for semaphores, and \rwl~for r/w locks.
+\mxtest, for locking algorithm correctness,
+\tej, a test of thread lifecycle,
+\bct~and \paraguay~for condvars,
+\paradise~for semaphores,
+and \rwl~for r/w locks.
 For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
@@
 (see \sect{\ref{sec:landslide}}).
-\newcommand\prisema{\texttt{priority\_sema}}
-\newcommand\waitsimple{\texttt{wait\_simple}}
+\newcommand\prisema{\texttt{sched\_test}}
+\newcommand\waitsimple{\texttt{wait\_test}}
 % TODO: Add more test cases
@@
 Though state-of-the-art MCs preempt only on synchronization events, many serious concurrency bugs are caused by data races leading to corrupted shared state.
-The line QS-DRs represents our ``no holds barred'' \quicksand~tests:
+The line QS-DRs represents our ``no holds barred'' \quicksand~tests,
+with dynamically-discovered data-race PPs enabled:
 we quickly pull ahead of SSS-MC, and ultimately conclude with 69\% more bugs in total.
@@
 
+\begin{table*}[t]
+	\begin{center}
+		%\small
+	\begin{tabular}{r||c|c|c||c|c|c|c}
+		& {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
+		{\bf Test name} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf recycle FPs} & {\bf DR PPs} & {\bf DR bugs} \\
+		\hline
+		{\bct		} & 7	& 11	& 4	&	& 65	&	& 4	\\
+		{\tej		} & 14 	& 25	& 11	&	& 471	&	& 6	\\
+		{\mxtest	} & 1	& 10	& 9	&	& 7	&	& 1	\\
+		{\paradise	} & 10	& 17	& 7	&	& 268	&	& 3	\\
+		{\paraguay	} & 4	& 12	& 9	&	& 188	&	& 0	\\
+		{\rwl		} & 26	& 30	& 3	&	& 157	&	& 3	\\
+		\hline
+		{\prisema	} &	&	&	&	&	&	&	\\
+		{\waitsimple	} &	&	&	&	&	&	&	\\
+		\hline
+		{\bf Total}	&	&	&	&	&	&	&
+	\end{tabular}
+	\end{center}
+	\caption{Breakdown of number and types of bugs found by test case.
+		SSS-MC is the control; QS is \quicksand.
+		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose.
+	``Nondeterministic DRs'' counts among the DR-only bugs how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
+		TODO: more explanation}
+	\label{tab:drbugs}
+\end{table*}
+
+The left half of Table~\ref{tab:drbugs} breaks down the number and types of bugs found by each test case.
+% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
 %TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.

### 2015-11-03 pldi-16/intro.tex
add structural text

@@
 \end{enumerate}
+
+The remainder of the paper is organized as follows.
+\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening,
+\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it,
+\sect{\ref{sec:eval}} presents our evaluation,
+\sect{\ref{sec:future}} discusses limitations and future work,
+\sect{\ref{sec:related}} surveys the related work,
+and \sect{\ref{sec:conclusion}} concludes.

### 2015-11-04 pldi-16/proof.tex
Start writing about convergence in proof.tex

@@
 
-\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
+%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
+\title{Soundness Proofs for Iterative Deepening}
 \authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
@@
 Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
+This document presents two proofs concerning the technique.
+
+First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a naive MC preempting on every single instruction.
+In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
+it serves as a full verification of {\em all possible} thread schedules of the given test case.
+
+Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
 Despite many powerful reduction techniques, the state spaces are still exponentially sized,
@@
 
+\begin{definition}[Reachable data race]
+A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
+or transitively also configured to use data-race PPs of other (already) reachable data races.
+\end{definition}
+
+%To prove a theory, generally sized super mario bros is not possible to complete, there is two step, one step for each F.
+To prove Iterative Deepening's convergence to full coverage,
+we are concerned with showing that any data race PP necessary to expose a bug will be reachable.
+
 \begin{definition}[False positive data race]
@@
 \end{figure}
-\section{Intuition}
-
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
-Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
 
@@
 
+\section{Intuition}
+
+In this section we provide (hopefully) intuitive summaries of both our proof goals,
+for readers who might be satisfied without necessarily following the proofs' internal structure.
+
+{\bf Intuition for Iterative Deepening convergence.}
+In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
+that set represents every single code location where a preemption could possibly affect the program's behaviour,
+and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
+Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
+but the technique's iterative nature will eventually find it.
+On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour.
+
+{\bf Intuition for malloc-recycle soundness.}
+In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
+then DPOR is guaranteed to ``reorder away the free and re-malloc''.
+In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
+while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
+Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
+So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Assumptions}
+
+This section documents the assumptions we make about the concurrency model, language model, and test environment,
+and discusses some limitations that may arise from these assumptions.
+
+\subsection{Assumptions for both proofs}
 
@@
 
-Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving.
+Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
 However, Iterative Deepening aims to test the most important interleavings with the time available,
 so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
-An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
-In that case, our point is to justify deprioritizing those jobs so aggressively.
+%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
+%In that case, our point is to justify deprioritizing those jobs so aggressively.
 
@@
 %Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
+
+{\bf Shared memory thread communication.}
+We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
+is through either shared memory or a correctly-instrumented sync API. Both DPOR and data-race detec
+
+{\bf Schedule nondeterminism only.}
+We discount the possibility of other types of nondeterminism,
+such as program input nondeterminism (including randomness/timestamps) or
+store-buffer nondeterminism on weak-memory architectures.
+We refer the reader to \cite{klee,portend} for related work on the former, and to \cite{tsopso} for the latter.
+
+\subsection{Assumptions for Iterative Deepening convergence}
+
+{\bf Locks are correct.}
+Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
+we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
+This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
+although we do not model RCU.
+Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
+
+Should the user wish to verify these properties of locks,
+they could either run a locking test separately (as we suggest in our paper),
+which would cheaply test the locks to an extent limited by the separate test case;
+or they could remove the data-race analysis's lockset tracking,
+which would expensively test all the main program's required locking properties in tandem with the program itself.
+
+\subsection{Assumptions for malloc-recycle soundness}
 
@@
 
-\section{Proof}
+\section{Proof of Iterative Deepening convergence}
+
+% TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Proof of malloc-recycle soundness}
 \label{sec:proof}
@@
 
-\section{Heap overruns}
+\subsection{Heap overruns}
 \label{sec:owned}

### 2015-11-04 pldi-16/di.tex
add structural text
elaborate more about seed pp sets in di.tex

@@
 \section{Design}
+\label{sec:design}
 
@@
 \section{Implementation}
+\label{sec:implementation}
 
@@
 {\bf Initial PPs.}
-For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}.
+For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}:
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+and $\{yield,lock,unlock\}$,
 By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
-such as condvars, semaphores, and r/w locks.
-Preempting on voluntary switches such as {\tt yield} is necessary for the invariant that only one thread runs between consecutive PPs,
-so the {\tt yield} PP is always implicitly enabled.
+such as condition variables, semaphores, and r/w locks.
+Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
+%so the {\tt yield} PP is always implicitly enabled.
 
 For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so (e.g. on x86) we also preempt just before {\tt cli} and just after {\tt sti}\footnote{
+so we also preempt just before {\tt cli} and just after {\tt sti} (to appropriate the names of the x86 instructions)\footnote{
 During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock,
-as {\tt cli}'d memory accesses can still race with others that have interrupts on.}.
-% FIXME: "What happened to yield?" "Yield is implicitly included in each subset."
+as {\tt cli}'d memory accesses can still race with others that have interrupts on.
+Some kernels disable preemption without disabling interrupts,
+which can be modelled the same way using manual annotations of that API.}.
 \quicksand~is configured to begin with the subsets
-$\{ \}$,
-$\{lock\}$,
-$\{unlock\}$,
-$\{cli\}$,
-$\{sti\}$,
-and $\{lock,unlock,cli,sti\}$.
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+$\{yield,cli\}$,
+$\{yield,sti\}$,
+and $\{yield,lock,$ $unlock,cli,sti\}$.
 As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
@@
 We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether \landslide~observed the access pair ordered one or both ways while testing the original state space,
-Single-order candidates are much more likely to be false-positives
-(though we cannot say for sure before preempting during the access itself, hence the heuristic),
-so we prioritize jobs with both-order data-race PPs.
+based on whether the MC observed the access pair ordered one or both ways while testing the original state space,
+We prioritize jobs with both-order data-race PPs,
+because single-order candidates are much more likely to be false-positives
+(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
+%so we prioritize jobs with both-order data-race PPs.
 For single-order races, we do not add PPs for the later access at all;

### 2015-11-04 pldi-16/proof.tex
Add skeleton for convergence proof

@@
 In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily following the proofs' internal structure.
+for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
 
@@
 but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour.
+On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
+so any PPs beyond the ones we consider are irrelevant.
 
@@
 This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
-although we do not model RCU.
+although we do not model RCU \cite{rcu}.
 Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
@@
 
-% TODO
+We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
+We'll call the latter the {\em naive maximal state space}, and call the condition of testing all its interleavings {\em convergence}.
+(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
+Hence, our convergence statement is as follows:
+
+\begin{theorem}[Total verification]
+	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
+	it serves as a verification of all possible thread schedules of the given test program.
+	\label{thm:totalverif}
+\end{theorem}
+
+The contrapositive statement offers more structure for an easier proof\footnote{
+For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
+note that we are using the constructive contrapositive direction:
+Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
+
+%DPOR + Iterative Deepening will eventually test every nonequivalent thread interleaving that would be exposed by preempting at every single instruction
+\begin{theorem}[Convergence of Iterative Deepening]
+	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+	\label{thm:convergence}
+\end{theorem}
+
+First we will prove a slightly weaker statement, limiting the naive approach's PPs to data races,
+then strengthen it by showing that the all interleavings in the naive maximal state space
+are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
+
+\subsection{Saturation of data-race PPs}
+
+\begin{lemma}
+	If a bug is exposed by any interleaving comprising any combination of data-race PPs and sync API PPs,
+	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
+		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
+		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
+	\label{lem:saturation}
+\end{lemma}
+
+\begin{proof}
+TODO % TODO
+\end{proof}
+
+\subsection{Saturation of data-race PPs}
+
+\begin{lemma}
+	For any interleaving in the naive maximal state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
+	\label{lem:equivalent}
+\end{lemma}
+
+\begin{proof}
+TODO % TODO
+\end{proof}
+
+\subsection{Conclusion}
+
+\setcounter{theorem}{1}
+\begin{theorem}[Convergence of Iterative Deepening]
+	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+\end{theorem}
+
+\begin{proof}
+By Lemma~\ref{lem:equivalent}, the naive interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
+By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
+Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
+\end{proof}
 
@@
 
-\setcounter{theorem}{0}
+\setcounter{theorem}{2}
 \begin{theorem}[Soundness of eliminating malloc-recycle races]

### 2015-11-05 pldi-16/proof.tex
Add some structure to convergence proof, but it requires rethinking

@@
 
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a naive MC preempting on every single instruction.
+First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve MC preempting on every single instruction.
 In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
@@
 We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
-is through either shared memory or a correctly-instrumented sync API. Both DPOR and data-race detec
+is through either shared memory or a correctly-instrumented sync API.
+Both DPOR and data-race detection rely on this assumption,
+as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
+For brevity in these proofs, we refer to all thread communication as shared memory,
+and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.
 
@@
 which would expensively test all the main program's required locking properties in tandem with the program itself.
+
+{\bf No forcibly blocking other threads.}
+We assume the sync API behaves much like pthreads and/or message-passing:
+that is, a thread can cause itself to block on some condition,
+and other threads may wake it up,
+but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
 
@@
 We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em naive maximal state space}, and call the condition of testing all its interleavings {\em convergence}.
+We'll call the latter the {\em na\"{i}ve state space}, and call the condition of testing all its interleavings {\em convergence}.
 (It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
@@
 \begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
 	\label{thm:convergence}
@@
 
-First we will prove a slightly weaker statement, limiting the naive approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the naive maximal state space
+First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
+then strengthen it by showing that the all interleavings in the na\"{i}ve state space
 are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
@@
 \begin{lemma}
-	If a bug is exposed by any interleaving comprising any combination of data-race PPs and sync API PPs,
+	If a bug can be exposed by some combination of preemptions on any data-race instruction and sync API boundaries,
 	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
@@
 
+The main challenge of this proof is that some data-race candidates are nondeterministic;
+i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
+as shown in Figure~\ref{fig:nondet-dr}.
+Hence, the maximal state space of statically-available sync API PPs will not necessarily uncover all possible data-race PPs;
+Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
+However, the proof is surprisingly straightforward.
+
 \begin{proof}
-TODO % TODO
+Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
+We induct on the $p$s to show that all data-race PPs will be reachable.
+\begin{itemize}
+	\item For the base case, $p_0$ required 
+	\item
+\end{itemize}
 \end{proof}
 
-\subsection{Saturation of data-race PPs}
+% TODO make this be on the right column
+\begin{figure}[t]
+	\small
+	\begin{tabular}{rl}
+	& \multicolumn{1}{c}{~\texttt{int x = 0, y = 0;~~~~~~~~~~~~~~~~}} \\
+	& \multicolumn{1}{c}{\texttt{bool t1\_x = false, t1\_y = false;}} \\
+	& \multicolumn{1}{c}{\texttt{bool t2\_x = false, t2\_y = false;}} \\
+\\
+		& Thread 1 (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
+	1 & \texttt{\hilight{brickred}{x = x + 1;}} \\
+	2 & \texttt{t1\_x = true;} \\
+	3 & \texttt{// "if x raced"} \\
+	4 & \texttt{if (x == 1 \&\& t2\_x) \{} \\
+	5 & \texttt{~~~~\hilight{brickred}{y = y + 1;}} \\
+	6 & \texttt{~~~~t1\_y = true;} \\
+	7 & \texttt{~~~~// "if y raced"} \\
+	8 & \texttt{~~~~if (y == 1 \&\& t2\_y) \{} \\
+	9 & \texttt{~~~~~~~~panic();} \\
+	10 & \texttt{~~~~\}} \\
+	11 & \texttt{\}} \\
+	\end{tabular}
+	\caption{Not all data races will immediately be uncovered by interleaving threads around sync API PPs alone.
+Here, preempting during the race on line 1 is necessary to force the program into the control flow which can race on line 5.}
+	\label{fig:nondet-dr}
+\end{figure}
+
+\subsection{Equivalence of irrelevant PPs}
 
 \begin{lemma}
-	For any interleaving in the naive maximal state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
+	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
 	\label{lem:equivalent}
@@
 \begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
 \end{theorem}
@@
 \begin{proof}
-By Lemma~\ref{lem:equivalent}, the naive interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
+By Lemma~\ref{lem:equivalent}, the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
 By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.

### 2015-11-07 pldi-16/proof.tex
Prove equivalence of irrelevant PPs

@@
 \begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
 	\label{thm:convergence}
@@
 
-First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the na\"{i}ve state space
-are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
+%First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
+%then strengthen it by showing that the all interleavings in the na\"{i}ve state space
+%are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
+
+%%%%
+
+\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
+\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
+\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
+\newcommand\pppfx[1]{\ensuremath{\mathsf{pfx}(#1)}}
+\newcommand\conflicts[2]{\ensuremath{\mathsf{conflicts}(#1,#2)}}
+\newcommand\pai{\ensuremath{p_{\alpha{}i}}}
+Let
+\[
+	\mathcal{I} = \{(t_\alpha, p_\alpha)_1, (t_\beta, p_\beta)_1, ... (t_\alpha, p_\alpha)_n, ...\}
+\]
+be an interleaving trace, where the element $(t_\alpha, p_\alpha)_i$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
+We will use the following notation:
+\begin{itemize}
+	\item $\ppnext{p_{\alpha{}i}}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
+	\item $\ppinstr{p_{\alpha{}i}}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
+	\item $\ppothers{p_{\alpha{}i}}$ indicates the set of all transitions by other threads between $t_{\alpha{}i}$ and $t_{\alpha{}(i+1)}$.
+	\item $\conflicts{t_\alpha,t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
+\end{itemize}
+
+\subsection{Equivalence of irrelevant PPs}
+
+Iterative Deepening will not, of course, preempt on exactly the same instructions that an arbitrary naive interleaving would,
+as it is only capable of preempting on data races and sync API boundaries.
+Thus, our first task is to show that all naive interleavings have an equivalent interleaving with only data-race and sync API PPs.
+
+% TODO: come up with a better name
+\begin{definition}[Relevant preemption point]
+	We say a PP $\pai$ is {\em relevant} if either $\ppnext{\pai}$ is a sync API boundary (including a voluntary yield),
+	or if
+	\[
+		\conflicts{\ppothers{\pai}}{\ppinstr{\pai}} \ne \emptyset
+	\]
+	i.e., the instruction by thread $\alpha$ immediately after $\pai$ has a memory conflict with some other thread interleaved during $\pai$.
+\end{definition}
+\begin{definition}[Fully-relevant interleaving]
+	An interleaving comprised only of relevant PPs.
+\end{definition}
+
+\begin{lemma}
+	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only relevant PPs.
+	%data-race PPs and sync API PPs.
+	\label{lem:equivalent}
+\end{lemma}
+
+\begin{proof}
+	Let $\pai$ be the first irrelevant PP of a naive interleaving $\mathcal{I}$.
+	We ask, did some thread among $\ppothers{\pai}$ conflict with any instruction from $\ppnext{\pai}$, even though there was no conflict with $\ppinstr{\pai}$?
+	\begin{itemize}
+			% rho for relevant.
+		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $t_{\alpha{}i}$ contains a sync API boundary instruction that's not already a PP,
+			then let $\rho$ be either the first instruction by $\ppnext{\pai}$ among the conflicts or the first sync API instruction, whichever comes first.
+			Let $\pppfx{\rho}$ denote the instruction sequence between $\ppinstr{\pai}$ and $\rho$, including the former but not the latter.
+			Now, we output the new interleaving:
+
+			\begin{tabular}{lll}
+				\\
+				$\mathcal{I}' = \{...,$&$(t_{\alpha{}i} \cup \pppfx{\rho}, \pai'),$& \\
+									   &$\ppothers{\pai},$& \\
+					       &$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
+				\\
+			\end{tabular}
+
+			In English, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
+			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
+			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR \cite{dpor}, $\mathcal{I}' \equiv \mathcal{I}$.
+		\item Otherwise, $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} = \emptyset$, and no instructions among $\ppnext{\pai}$ are a sync API boundary.
+			Then we output the new interleaving:
+			\[
+				\mathcal{I}' = \{..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
+			\]
+			In English, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
+			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
+					       %&$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
+	\end{itemize}
+	This constitutes an algorithm for inductively converting all irrelevant PPs to relevant ones.
+\end{proof}
+
+%%%%
 
@@
 \begin{lemma}
-	If a bug can be exposed by some combination of preemptions on any data-race instruction and sync API boundaries,
+	If a bug can be exposed by any fully-relevant interleaving, %some combination of preemptions on any data-race instruction and sync API boundaries,
 	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
@@
 
-The main challenge of this proof is that some data-race candidates are nondeterministic;
+The main challenge of this proof is that some data-race candidates are nondeterministic to find;
 i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
@@
 Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
-However, the proof is surprisingly straightforward.
 
@@
 Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
+% TODO: define the special way of ordering the pps
 We induct on the $p$s to show that all data-race PPs will be reachable.
@@
 
-\subsection{Equivalence of irrelevant PPs}
-
-\begin{lemma}
-	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:equivalent}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
+%%%%
 

### 2015-11-10 pldi-16/proof/proof.tex
Move proof.tex to its own subdir

--- 2015-11-07 pldi-16/proof.tex
@@
 %
-\documentclass[pldi]{sigplanconf-pldi15}
+\documentclass[pldi]{../sigplanconf-pldi15}
 
@@
 It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
-
-\begin{definition}[Reachable data race]
-A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
-or transitively also configured to use data-race PPs of other (already) reachable data races.
-\end{definition}
-
-%To prove a theory, generally sized super mario bros is not possible to complete, there is two step, one step for each F.
-To prove Iterative Deepening's convergence to full coverage,
-we are concerned with showing that any data race PP necessary to expose a bug will be reachable.
 
@@
 \[
-	\mathcal{I} = \{(t_\alpha, p_\alpha)_1, (t_\beta, p_\beta)_1, ... (t_\alpha, p_\alpha)_n, ...\}
+	\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), (t_{\beta{}1}, p_{\beta{}1}), ... (t_{\alpha{}n}, p_{\alpha{}n}), ...\}
 \]
-be an interleaving trace, where the element $(t_\alpha, p_\alpha)_i$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
+be an interleaving trace, where the element $(t_{\alpha{}i}, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
 We will use the following notation:
@@
 
-			In English, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
+			In other words, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
 			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
@@
 			\]
-			In English, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
+			In other words, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
 			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
@@
 	\end{itemize}
-	This constitutes an algorithm for inductively converting all irrelevant PPs to relevant ones.
+	This constitutes an algorithm for inductively converting (or removing) all irrelevant PPs to relevant ones.
 \end{proof}
@@
 
-\begin{lemma}
-	If a bug can be exposed by any fully-relevant interleaving, %some combination of preemptions on any data-race instruction and sync API boundaries,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-	\label{lem:saturation}
-\end{lemma}
-
-The main challenge of this proof is that some data-race candidates are nondeterministic to find;
+Now we must show that, starting from a sync-PP-only state space,
+Iterative Deepening will eventually interleave threads sufficiently to expose all data races which are used as PPs in any buggy fully-relevant interleaving.
+The challenge of this proof is that some data-race candidates are nondeterministic to find;
 i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
@@
 
-\begin{proof}
-Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
-% TODO: define the special way of ordering the pps
-We induct on the $p$s to show that all data-race PPs will be reachable.
-\begin{itemize}
-	\item For the base case, $p_0$ required 
-	\item
-\end{itemize}
-\end{proof}
-
-% TODO make this be on the right column
 \begin{figure}[t]
@@
 
+\begin{definition}[Reachable data race]
+A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
+or transitively also configured to use data-race PPs of other (already) reachable data races.
+\end{definition}
+
+\begin{definition}[Reachable preemption point]
+A PP $\pai$ such that either $\ppinstr{\pai}$ is part of a reachable data race, or a sync API boundary.
+\end{definition}
+
+\begin{lemma}
+	Given a fully-relevant interleaving
+	\[
+		\mathcal{I} = \{ ..., (t_{\alpha{}i},\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
+	\]
+	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
+	\[
+		\mathcal{J} = \{ ..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
+	\]
+	are reachable, then $\pai$ is reachable.
+	\label{lem:coalesce}
+\end{lemma}
+
+Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
+i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
+and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
+Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
+Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.
+
+% TODO: What if ppothers changes based on ppinstr(pai)?
+You say, there must exist some J where ppothers has fewer PPs.
+You just stop preempting!
+Those preemptions don't matter.
+Gotta go fast.
+Gotta keep the induction metric of getting fewer pps in J than in I.
+
+%%%
+
+\begin{lemma}
+	If a bug can be exposed by any fully-relevant interleaving, %some combination of preemptions on any data-race instruction and sync API boundaries,
+	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
+		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
+		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
+	\label{lem:saturation}
+\end{lemma}
+
+% TODO: Here's what to do.
+% Induct on the PPs by the order of next(my_PP) in the trace.
+% Inductive assumption gives that all prior other data race PPs with next(other_PP) before next(my_PP) are reachable,
+% Is there a prior data race PP with next(other_PP) *AFTER* next(my_PP)?
+% If not, then we are done, and this one is reachable.
+% If so, then we ask, which transition among others(other_PP) (this includes us) does instr(other_PP) conflict with?
+% <vague handwavy part>
+% Then we show that there must be a trace with next(other_PP) interleaved to be before next(my_PP) where, even if the bug doesn't show up, the data race still shows up, which makes other_PP reachable.
+% </vague handwavy part>
+% Since other_PP is reachable then my_PP is reachable.
+
+\begin{proof}
+Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
+We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
+Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
+i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions.
+%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
+%%% well, the way we do it, might not be the first reachable one either...
+
+\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
+For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj}$ occurring before $\ppnext{\pai}$ in $\mathcal{I}$, $\pbj$ is reachable.
+% Now to show \pai is reachable.
+
+\begin{itemize}
+	\item
+	\item
+\end{itemize}
+\end{proof}
+
 %%%%
@@
 \begin{proof}
-By Lemma~\ref{lem:equivalent}, the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
+By Lemma~\ref{lem:equivalent}, there must be some equivalent fully-relevant interleaving.
+%the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
 By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
@@
 \bibliographystyle{abbrvnat}
-\bibliography{citations}{}
+\bibliography{../citations}{}
 \end{document}

### 2015-11-10 pldi-16/proof/assumptions.tex
split proof into files

--- 2015-11-05 pldi-16/proof.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi15}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{amsthm}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-\usepackage{graphicx}
-%\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-%\doublespacing
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
-\newcommand\quicksand{\textsc{Quicksand}}
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-
-%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
-\title{Soundness Proofs for Iterative Deepening}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
-Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
-This document presents two proofs concerning the technique.
-
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve MC preempting on every single instruction.
-In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
-it serves as a full verification of {\em all possible} thread schedules of the given test case.
-
-Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
-Despite many powerful reduction techniques, the state spaces are still exponentially sized,
-so any way to avoid exploring some in advance is obviously beneficial.
-We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.
-
-This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Definitions}
-
-\subsection{Stateless model checking terms}
-
-\begin{definition}[Transition]
-A sequence of instructions executed by the program between two preemption points (PPs).
-\label{def:transition}
-\end{definition}
-Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-
-\begin{definition}[Must-happen-before (MHB)]
-	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
-\end{definition}
-All transitions of the same thread are trivially MHB-related.
-Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
-Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
-% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
-% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).
-
-Note how our {\em must}-happen-before relation differs from the conventional definition of happens-before (``observed to happen before'') \cite{lamport-clocks}.
-Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
-the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
-Because pure-HB data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
-However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock,
-our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
-}.
-
-\begin{definition}[Shared memory conflict]
-A pair of memory accesses between two threads to the same address where at least one of them is a write.
-\end{definition}
-Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
-
-\begin{definition}[Dynamic Partial Order Reduction (DPOR)]
-	A state-space search algorithm for stateless model checkers,
-	guaranteed to reorder transitions of two threads
-	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
-	\label{def:dpor}
-\end{definition}
-
-\subsection{Data race and other memory terms}
-
-\begin{definition}[Data race]
-A shared memory conflict where furthermore:
-\begin{itemize}
-	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
-	\item The threads' transitions are not related by MHB.
-\end{itemize}
-\end{definition}
-
-The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
-
-\begin{definition}[Reachable data race]
-A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
-or transitively also configured to use data-race PPs of other (already) reachable data races.
-\end{definition}
-
-%To prove a theory, generally sized super mario bros is not possible to complete, there is two step, one step for each F.
-To prove Iterative Deepening's convergence to full coverage,
-we are concerned with showing that any data race PP necessary to expose a bug will be reachable.
-
-\begin{definition}[False positive data race]
-	An apparent data race that cannot be observed in the opposite order from what was actually executed.
-\end{definition}
-
-False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
-changes some variable values when the threads are reordered, such that the memory addresses no longer collide.
-
-\begin{definition}[Malloc-recycle data race]
-	A data race where the address is contained in some heap-allocated memory, and between the two accesses, that memory was passed to free() and returned again by a subsequent malloc().
-\end{definition}
-
-Figures~\ref{fig:recycle} and \ref{fig:recycle-bug} show an example.
-In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.
-
-\begin{definition}[Use after free]
-	Any read or write to heap memory which was once allocated, but no longer is.
-\end{definition}
-
-These can immediately be identified as failures by a MC which tracks allocation state.
-%Most commonly this refers to accesses to a region already freed, but for brevity we also include
-% TODO: need heap block overrun?
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	3 & & \texttt{// x's memory recycled} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{False-positive malloc-recycle pattern. This is the common case for which we avoid creating new state spaces.}
-\label{fig:recycle}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	5 & & \texttt{// x's memory recycled} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Adversarial program which fits the malloc-recycle pattern, but nevertheless contains a true race.}
-\label{fig:recycle-bug}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & & \texttt{// x not free, so malloc's} \\
-	4 & & \texttt{// return value changes!} \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
-\label{fig:recycle-goal}
-\end{figure}
-
-\section{Intuition}
-
-In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
-
-{\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
-that set represents every single code location where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
-Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
-but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any PPs beyond the ones we consider are irrelevant.
-
-{\bf Intuition for malloc-recycle soundness.}
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
-Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Assumptions}
@@
 For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of Iterative Deepening convergence}
-
-We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em na\"{i}ve state space}, and call the condition of testing all its interleavings {\em convergence}.
-(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
-Hence, our convergence statement is as follows:
-
-\begin{theorem}[Total verification]
-	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
-	it serves as a verification of all possible thread schedules of the given test program.
-	\label{thm:totalverif}
-\end{theorem}
-
-The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
-note that we are using the constructive contrapositive direction:
-Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
-
-%DPOR + Iterative Deepening will eventually test every nonequivalent thread interleaving that would be exposed by preempting at every single instruction
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-	\label{thm:convergence}
-\end{theorem}
-
-First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the na\"{i}ve state space
-are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
-
-\subsection{Saturation of data-race PPs}
-
-\begin{lemma}
-	If a bug can be exposed by some combination of preemptions on any data-race instruction and sync API boundaries,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-	\label{lem:saturation}
-\end{lemma}
-
-The main challenge of this proof is that some data-race candidates are nondeterministic;
-i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
-as shown in Figure~\ref{fig:nondet-dr}.
-Hence, the maximal state space of statically-available sync API PPs will not necessarily uncover all possible data-race PPs;
-Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
-However, the proof is surprisingly straightforward.
-
-\begin{proof}
-Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
-We induct on the $p$s to show that all data-race PPs will be reachable.
-\begin{itemize}
-	\item For the base case, $p_0$ required 
-	\item
-\end{itemize}
-\end{proof}
-
-% TODO make this be on the right column
-\begin{figure}[t]
-	\small
-	\begin{tabular}{rl}
-	& \multicolumn{1}{c}{~\texttt{int x = 0, y = 0;~~~~~~~~~~~~~~~~}} \\
-	& \multicolumn{1}{c}{\texttt{bool t1\_x = false, t1\_y = false;}} \\
-	& \multicolumn{1}{c}{\texttt{bool t2\_x = false, t2\_y = false;}} \\
-\\
-		& Thread 1 (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
-	1 & \texttt{\hilight{brickred}{x = x + 1;}} \\
-	2 & \texttt{t1\_x = true;} \\
-	3 & \texttt{// "if x raced"} \\
-	4 & \texttt{if (x == 1 \&\& t2\_x) \{} \\
-	5 & \texttt{~~~~\hilight{brickred}{y = y + 1;}} \\
-	6 & \texttt{~~~~t1\_y = true;} \\
-	7 & \texttt{~~~~// "if y raced"} \\
-	8 & \texttt{~~~~if (y == 1 \&\& t2\_y) \{} \\
-	9 & \texttt{~~~~~~~~panic();} \\
-	10 & \texttt{~~~~\}} \\
-	11 & \texttt{\}} \\
-	\end{tabular}
-	\caption{Not all data races will immediately be uncovered by interleaving threads around sync API PPs alone.
-Here, preempting during the race on line 1 is necessary to force the program into the control flow which can race on line 5.}
-	\label{fig:nondet-dr}
-\end{figure}
-
-\subsection{Equivalence of irrelevant PPs}
-
-\begin{lemma}
-	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:equivalent}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{1}
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-\end{theorem}
-
-\begin{proof}
-By Lemma~\ref{lem:equivalent}, the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
-Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of malloc-recycle soundness}
-\label{sec:proof}
-
-We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
-Hence, our soundness statement is as follows:
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-Though Figures~\ref{fig:recycle-bug} and \ref{fig:recycle-goal} show example programs, they do not capture all possible cases of how a true data race can fit the malloc-recycle pattern.
-We proceed by establishing what must be true of any such program, then casing on the ambiguous possibilities, and showing that PPs will exist where we need them to reorder the threads.
-
-For certain, there must be an access in one thread, followed by a free and malloc (we'll call them ``middle free'' and ``middle malloc''), each possibly from either thread, followed by an access from the other.
-If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
-WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
-
-\begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
-	then a non-malloc-recycle data race or a use-after-free bug will be identified.
-	\label{lem:reorder}
-\end{lemma}
-\begin{proof}
-By case on which threads the middle free and middle malloc came from.
-\begin{itemize}
-	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
-The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
-Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
-	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
-	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
-\end{itemize}
-If either the middle free or middle malloc came from a third thread, the case is the same as if it belonged to T1.
-\end{proof}
-
-The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
-However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
-%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
-and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
-So, WLOG, we say that the only relevant events are the ones we mention explicitly.
-
-By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
-We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
-Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
-
-\subsection{T2 originally malloced x}
-
-\begin{lemma}[Greedo]
-	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:greedo}
-\end{lemma}
-\begin{proof}
-Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
-By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
-and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
-Lemma~\ref{lem:reorder} finishes.
-\end{proof}
-
-This lemma also applies if a third thread was responsible for this malloc, as there would still be a thread switch in the same spot.
-
-\subsection{T1 originally malloced x}
-
-\begin{lemma}[Han]
-	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:han}
-\end{lemma}
-
-\begin{proof}
-We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
-Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
-
-%\begin{figure}[t]
-%	\small
-%\begin{tabular}{rll}
-%	& Thread 1 & Thread 2 \\
-%	0 & \texttt{lock();} & \\
-%	1 & \texttt{publish(x1);} & \\
-%	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-%	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-%	0 & \texttt{unlock();} & \\
-%	0 & & \texttt{lock();} \\
-%	2 & & \texttt{x2 = get\_published\_x();} \\
-%	0 & & \texttt{unlock();} \\
-%	5 & & \texttt{// x's memory recycled} \\
-%	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-%	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-%\end{tabular}
-%\caption{If the locksets are the same in T1 during publish and x1, you can't avoid the data race on publish+get.}
-%\label{fig:recycle-bug}
-%\end{figure}
-
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
-If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
-For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
-which we'll call {\tt p}.
-By the previous two statements, {\tt p} must be a data race of its own.
-%(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-
-	%1 & \texttt{p1 = \hilight{olivegreen}{malloc}(...);} & \\
-	1 & \texttt{\hilight{brickred}{p1->ptr = x1;}} & \\
-	2 & \texttt{publish(p1);} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(p1);} & \\
-	4 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	5 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-
-	6 & & \texttt{p2 = get\_published\_p();} \\
-	7 & & \texttt{// p's memory recycled} \\
-	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
-	9 & & \texttt{\hilight{brickred}{x2 = p2->ptr;}} \\
-
-	10 & & \texttt{// x's memory recycled} \\
-	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	12 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If the accesses used to publish {\tt x}'s address are a data race, their PPs may also be eliminated under the malloc-recycle pattern.
-Induction on the pointer structure leading to {\tt x} handles this case.}
-\label{fig:induction}
-\end{figure}
-
-Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
-as shown in Figure~\ref{fig:induction},
-we do not necessarily get the PP for free here.
-In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
-In \quicksand, data race PPs are not used immediately, but rather generate new state spaces to explore in the future. Anyway, under the maximal state space assumption, we will get to it eventually.}.
-We handle this with induction on T2's pointer chain leading to {\tt x}.
-
-\newcommand\publish[1]{{\tt p}$_{#1}$}
-\begin{itemize}
-	\item For the base case, the publish location \publish{0} is either in global memory, or shared directly using synchronization.
-		Non-heap memory data races are not subject to the malloc-recycle pattern, so will always get a data-race PP,
-		and use of synchronization always gets a PP in the maximal state space.
-	\item For the inductive step, a pointer \publish{n} is published in some heap memory \publish{n-1}\texttt{->ptr},
-		and we assume that however \publish{n-1} is shared to T2,
-		there will be a PP there
-		% where \publish{n-1} is shared
-		sufficient to make the \publish{n-1}\texttt{->ptr} access not malloc-recycle after DPOR.
-		%DPOR will test some interleaving in which T1's
-		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
-		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
-		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
-		% i.e., so that the p_n access no longer appears to be malloc-recycle.
-\end{itemize}
-
-Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location {\tt p}.
-Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
-% XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
-As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
-Lemma~\ref{lem:reorder} finishes.
-%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
-
-\end{proof}
-
-%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
-%As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{2}
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-\begin{proof}
-	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases of possible program structure are covered.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Heap overruns}
-\label{sec:owned}
-
-Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
-Figure~\ref{fig:overrun} shows an example.
-Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
-So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	7 & & \texttt{// x's memory recycled} \\
-	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	9 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Final possibility for how T2 can share T1's allocation address, and probably a security vulnerability to boot!}
-\label{fig:overrun}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{x1 = malloc(...);} & \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
-\label{fig:overrun-notenough}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
-\label{fig:overrun-goal}
-\end{figure}
-
-% TODO: Replace this with some real discussion.
-
-%{\bf Being Owned.}
-%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
-Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-%
-%
-%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
-If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
-this is because dpor ignores malloc's internal accesses from the shm conflict relation.
-
-what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
-BUT, the original one was not sound with respect to this bug to begin with.
-
-then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
-
-\begin{lemma}
-	asdfa
-	\label{lem:leia} % "there is another"
-\end{lemma}
-
-% TODO
-% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
-% for the purpose of iterative deepening, MHB is
-% better because it will find the data races much sooner, rather than requiring you
-% to explore the maximal state space before the race can be uncovered
-
-%\subsection{Malloc with internal mutexes}
-%
-%First we assume that the malloc implementation is guarded by
-%
-%
-%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-%
-%
-%\subsection{Malloc as a magic primitive}
-%
-%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
-%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
-%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
-%
-%Hence, we provide another proof that does not require
-%
-%%\cite{landslide} % uncomment post-double-blind review
-%
-%This spells QED so we are done.
-
-\section{Acknowledgements}
-
-{\em removed for double blind review}
-%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-\end{document}

### 2015-11-10 pldi-16/proof/convergence.tex
split proof into files

--- 2015-11-10 pldi-16/proof/proof.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-%
-\documentclass[pldi]{../sigplanconf-pldi15}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{amsthm}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-\usepackage{graphicx}
-%\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-%\doublespacing
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
-\newcommand\quicksand{\textsc{Quicksand}}
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-
-%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
-\title{Soundness Proofs for Iterative Deepening}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
-Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
-This document presents two proofs concerning the technique.
-
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve MC preempting on every single instruction.
-In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
-it serves as a full verification of {\em all possible} thread schedules of the given test case.
-
-Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
-Despite many powerful reduction techniques, the state spaces are still exponentially sized,
-so any way to avoid exploring some in advance is obviously beneficial.
-We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.
-
-This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Definitions}
-
-\subsection{Stateless model checking terms}
-
-\begin{definition}[Transition]
-A sequence of instructions executed by the program between two preemption points (PPs).
-\label{def:transition}
-\end{definition}
-Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-
-\begin{definition}[Must-happen-before (MHB)]
-	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
-\end{definition}
-All transitions of the same thread are trivially MHB-related.
-Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
-Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
-% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
-% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).
-
-Note how our {\em must}-happen-before relation differs from the conventional definition of happens-before (``observed to happen before'') \cite{lamport-clocks}.
-Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
-the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
-Because pure-HB data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
-However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock,
-our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
-}.
-
-\begin{definition}[Shared memory conflict]
-A pair of memory accesses between two threads to the same address where at least one of them is a write.
-\end{definition}
-Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
-
-\begin{definition}[Dynamic Partial Order Reduction (DPOR)]
-	A state-space search algorithm for stateless model checkers,
-	guaranteed to reorder transitions of two threads
-	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
-	\label{def:dpor}
-\end{definition}
-
-\subsection{Data race and other memory terms}
-
-\begin{definition}[Data race]
-A shared memory conflict where furthermore:
-\begin{itemize}
-	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
-	\item The threads' transitions are not related by MHB.
-\end{itemize}
-\end{definition}
-
-The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
-
-\begin{definition}[False positive data race]
-	An apparent data race that cannot be observed in the opposite order from what was actually executed.
-\end{definition}
-
-False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
-changes some variable values when the threads are reordered, such that the memory addresses no longer collide.
-
-\begin{definition}[Malloc-recycle data race]
-	A data race where the address is contained in some heap-allocated memory, and between the two accesses, that memory was passed to free() and returned again by a subsequent malloc().
-\end{definition}
-
-Figures~\ref{fig:recycle} and \ref{fig:recycle-bug} show an example.
-In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.
-
-\begin{definition}[Use after free]
-	Any read or write to heap memory which was once allocated, but no longer is.
-\end{definition}
-
-These can immediately be identified as failures by a MC which tracks allocation state.
-%Most commonly this refers to accesses to a region already freed, but for brevity we also include
-% TODO: need heap block overrun?
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	3 & & \texttt{// x's memory recycled} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{False-positive malloc-recycle pattern. This is the common case for which we avoid creating new state spaces.}
-\label{fig:recycle}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	5 & & \texttt{// x's memory recycled} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Adversarial program which fits the malloc-recycle pattern, but nevertheless contains a true race.}
-\label{fig:recycle-bug}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & & \texttt{// x not free, so malloc's} \\
-	4 & & \texttt{// return value changes!} \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
-\label{fig:recycle-goal}
-\end{figure}
-
-\section{Intuition}
-
-In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
-
-{\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
-that set represents every single code location where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
-Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
-but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any PPs beyond the ones we consider are irrelevant.
-
-{\bf Intuition for malloc-recycle soundness.}
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
-Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Assumptions}
-
-This section documents the assumptions we make about the concurrency model, language model, and test environment,
-and discusses some limitations that may arise from these assumptions.
-
-\subsection{Assumptions for both proofs}
-
-{\bf Maximal state space.}
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
-and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
-
-Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
-However, Iterative Deepening aims to test the most important interleavings with the time available,
-so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
-%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
-%In that case, our point is to justify deprioritizing those jobs so aggressively.
-
-%{\bf Low overhead.}
-%We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-%Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
-
-{\bf Shared memory thread communication.}
-We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
-is through either shared memory or a correctly-instrumented sync API.
-Both DPOR and data-race detection rely on this assumption,
-as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
-For brevity in these proofs, we refer to all thread communication as shared memory,
-and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.
-
-{\bf Schedule nondeterminism only.}
-We discount the possibility of other types of nondeterminism,
-such as program input nondeterminism (including randomness/timestamps) or
-store-buffer nondeterminism on weak-memory architectures.
-We refer the reader to \cite{klee,portend} for related work on the former, and to \cite{tsopso} for the latter.
-
-\subsection{Assumptions for Iterative Deepening convergence}
-
-{\bf Locks are correct.}
-Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
-we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
-This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
-although we do not model RCU \cite{rcu}.
-Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
-
-Should the user wish to verify these properties of locks,
-they could either run a locking test separately (as we suggest in our paper),
-which would cheaply test the locks to an extent limited by the separate test case;
-or they could remove the data-race analysis's lockset tracking,
-which would expensively test all the main program's required locking properties in tandem with the program itself.
-
-{\bf No forcibly blocking other threads.}
-We assume the sync API behaves much like pthreads and/or message-passing:
-that is, a thread can cause itself to block on some condition,
-and other threads may wake it up,
-but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
-
-\subsection{Assumptions for malloc-recycle soundness}
-
-{\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
-%particular allocation pattern regarding adjacency/coalescing/reuse.
-In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
-in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-Furthermore, we configured DPOR to ignore the internal heap metadata accesses
-when tracking shared memory conflicts to achieve greater state space reduction
-(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
-%for the purpose of avoiding malloc-only shared memory conflicts.
-This is not without consequences; see section~\ref{sec:owned}.
-
-{\bf Sharing heap addresses.}
-Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
-Because we are testing C programs, any bizarre violations of this assumption are technically possible,
-but should you wish to check for bugs like this,
-%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
-symbolic execution \cite{klee} would be more appropriate.
-
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Proof of Iterative Deepening convergence}
@@
 \end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of malloc-recycle soundness}
-\label{sec:proof}
-
-We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
-Hence, our soundness statement is as follows:
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-Though Figures~\ref{fig:recycle-bug} and \ref{fig:recycle-goal} show example programs, they do not capture all possible cases of how a true data race can fit the malloc-recycle pattern.
-We proceed by establishing what must be true of any such program, then casing on the ambiguous possibilities, and showing that PPs will exist where we need them to reorder the threads.
-
-For certain, there must be an access in one thread, followed by a free and malloc (we'll call them ``middle free'' and ``middle malloc''), each possibly from either thread, followed by an access from the other.
-If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
-WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
-
-\begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
-	then a non-malloc-recycle data race or a use-after-free bug will be identified.
-	\label{lem:reorder}
-\end{lemma}
-\begin{proof}
-By case on which threads the middle free and middle malloc came from.
-\begin{itemize}
-	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
-The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
-Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
-	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
-	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
-\end{itemize}
-If either the middle free or middle malloc came from a third thread, the case is the same as if it belonged to T1.
-\end{proof}
-
-The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
-However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
-%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
-and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
-So, WLOG, we say that the only relevant events are the ones we mention explicitly.
-
-By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
-We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
-Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
-
-\subsection{T2 originally malloced x}
-
-\begin{lemma}[Greedo]
-	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:greedo}
-\end{lemma}
-\begin{proof}
-Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
-By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
-and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
-Lemma~\ref{lem:reorder} finishes.
-\end{proof}
-
-This lemma also applies if a third thread was responsible for this malloc, as there would still be a thread switch in the same spot.
-
-\subsection{T1 originally malloced x}
-
-\begin{lemma}[Han]
-	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:han}
-\end{lemma}
-
-\begin{proof}
-We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
-Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
-
-%\begin{figure}[t]
-%	\small
-%\begin{tabular}{rll}
-%	& Thread 1 & Thread 2 \\
-%	0 & \texttt{lock();} & \\
-%	1 & \texttt{publish(x1);} & \\
-%	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-%	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-%	0 & \texttt{unlock();} & \\
-%	0 & & \texttt{lock();} \\
-%	2 & & \texttt{x2 = get\_published\_x();} \\
-%	0 & & \texttt{unlock();} \\
-%	5 & & \texttt{// x's memory recycled} \\
-%	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-%	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-%\end{tabular}
-%\caption{If the locksets are the same in T1 during publish and x1, you can't avoid the data race on publish+get.}
-%\label{fig:recycle-bug}
-%\end{figure}
-
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
-If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
-For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
-which we'll call {\tt p}.
-By the previous two statements, {\tt p} must be a data race of its own.
-%(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-
-	%1 & \texttt{p1 = \hilight{olivegreen}{malloc}(...);} & \\
-	1 & \texttt{\hilight{brickred}{p1->ptr = x1;}} & \\
-	2 & \texttt{publish(p1);} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(p1);} & \\
-	4 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	5 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-
-	6 & & \texttt{p2 = get\_published\_p();} \\
-	7 & & \texttt{// p's memory recycled} \\
-	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
-	9 & & \texttt{\hilight{brickred}{x2 = p2->ptr;}} \\
-
-	10 & & \texttt{// x's memory recycled} \\
-	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	12 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If the accesses used to publish {\tt x}'s address are a data race, their PPs may also be eliminated under the malloc-recycle pattern.
-Induction on the pointer structure leading to {\tt x} handles this case.}
-\label{fig:induction}
-\end{figure}
-
-Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
-as shown in Figure~\ref{fig:induction},
-we do not necessarily get the PP for free here.
-In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
-In \quicksand, data race PPs are not used immediately, but rather generate new state spaces to explore in the future. Anyway, under the maximal state space assumption, we will get to it eventually.}.
-We handle this with induction on T2's pointer chain leading to {\tt x}.
-
-\newcommand\publish[1]{{\tt p}$_{#1}$}
-\begin{itemize}
-	\item For the base case, the publish location \publish{0} is either in global memory, or shared directly using synchronization.
-		Non-heap memory data races are not subject to the malloc-recycle pattern, so will always get a data-race PP,
-		and use of synchronization always gets a PP in the maximal state space.
-	\item For the inductive step, a pointer \publish{n} is published in some heap memory \publish{n-1}\texttt{->ptr},
-		and we assume that however \publish{n-1} is shared to T2,
-		there will be a PP there
-		% where \publish{n-1} is shared
-		sufficient to make the \publish{n-1}\texttt{->ptr} access not malloc-recycle after DPOR.
-		%DPOR will test some interleaving in which T1's
-		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
-		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
-		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
-		% i.e., so that the p_n access no longer appears to be malloc-recycle.
-\end{itemize}
-
-Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location {\tt p}.
-Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
-% XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
-As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
-Lemma~\ref{lem:reorder} finishes.
-%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
-
-\end{proof}
-
-%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
-%As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{2}
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-\begin{proof}
-	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases of possible program structure are covered.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Heap overruns}
-\label{sec:owned}
-
-Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
-Figure~\ref{fig:overrun} shows an example.
-Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
-So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	7 & & \texttt{// x's memory recycled} \\
-	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	9 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Final possibility for how T2 can share T1's allocation address, and probably a security vulnerability to boot!}
-\label{fig:overrun}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{x1 = malloc(...);} & \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
-\label{fig:overrun-notenough}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
-\label{fig:overrun-goal}
-\end{figure}
-
-% TODO: Replace this with some real discussion.
-
-%{\bf Being Owned.}
-%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
-Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-%
-%
-%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
-If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
-this is because dpor ignores malloc's internal accesses from the shm conflict relation.
-
-what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
-BUT, the original one was not sound with respect to this bug to begin with.
-
-then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
-
-\begin{lemma}
-	asdfa
-	\label{lem:leia} % "there is another"
-\end{lemma}
-
-% TODO
-% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
-% for the purpose of iterative deepening, MHB is
-% better because it will find the data races much sooner, rather than requiring you
-% to explore the maximal state space before the race can be uncovered
-
-%\subsection{Malloc with internal mutexes}
-%
-%First we assume that the malloc implementation is guarded by
-%
-%
-%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-%
-%
-%\subsection{Malloc as a magic primitive}
-%
-%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
-%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
-%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
-%
-%Hence, we provide another proof that does not require
-%
-%%\cite{landslide} % uncomment post-double-blind review
-%
-%This spells QED so we are done.
-
-\section{Acknowledgements}
-
-{\em removed for double blind review}
-%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
-
-\bibliographystyle{abbrvnat}
-\bibliography{../citations}{}
-\end{document}

### 2015-11-10 pldi-16/proof/di.tex
split proof into files

--- 2015-11-05 pldi-16/proof.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi15}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{amsthm}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-\usepackage{graphicx}
-%\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-%\doublespacing
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
-\newcommand\quicksand{\textsc{Quicksand}}
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-
-%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
-\title{Soundness Proofs for Iterative Deepening}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
-Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
-This document presents two proofs concerning the technique.
-
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve MC preempting on every single instruction.
-In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
-it serves as a full verification of {\em all possible} thread schedules of the given test case.
-
-Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
-Despite many powerful reduction techniques, the state spaces are still exponentially sized,
-so any way to avoid exploring some in advance is obviously beneficial.
-We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.
-
-This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Definitions}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Intuition}
@@
 So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Assumptions}
-
-This section documents the assumptions we make about the concurrency model, language model, and test environment,
-and discusses some limitations that may arise from these assumptions.
-
-\subsection{Assumptions for both proofs}
-
-{\bf Maximal state space.}
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
-and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
-
-Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
-However, Iterative Deepening aims to test the most important interleavings with the time available,
-so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
-%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
-%In that case, our point is to justify deprioritizing those jobs so aggressively.
-
-%{\bf Low overhead.}
-%We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-%Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
-
-{\bf Shared memory thread communication.}
-We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
-is through either shared memory or a correctly-instrumented sync API.
-Both DPOR and data-race detection rely on this assumption,
-as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
-For brevity in these proofs, we refer to all thread communication as shared memory,
-and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.
-
-{\bf Schedule nondeterminism only.}
-We discount the possibility of other types of nondeterminism,
-such as program input nondeterminism (including randomness/timestamps) or
-store-buffer nondeterminism on weak-memory architectures.
-We refer the reader to \cite{klee,portend} for related work on the former, and to \cite{tsopso} for the latter.
-
-\subsection{Assumptions for Iterative Deepening convergence}
-
-{\bf Locks are correct.}
-Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
-we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
-This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
-although we do not model RCU \cite{rcu}.
-Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
-
-Should the user wish to verify these properties of locks,
-they could either run a locking test separately (as we suggest in our paper),
-which would cheaply test the locks to an extent limited by the separate test case;
-or they could remove the data-race analysis's lockset tracking,
-which would expensively test all the main program's required locking properties in tandem with the program itself.
-
-{\bf No forcibly blocking other threads.}
-We assume the sync API behaves much like pthreads and/or message-passing:
-that is, a thread can cause itself to block on some condition,
-and other threads may wake it up,
-but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
-
-\subsection{Assumptions for malloc-recycle soundness}
-
-{\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
-%particular allocation pattern regarding adjacency/coalescing/reuse.
-In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
-in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-Furthermore, we configured DPOR to ignore the internal heap metadata accesses
-when tracking shared memory conflicts to achieve greater state space reduction
-(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
-%for the purpose of avoiding malloc-only shared memory conflicts.
-This is not without consequences; see section~\ref{sec:owned}.
-
-{\bf Sharing heap addresses.}
-Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
-Because we are testing C programs, any bizarre violations of this assumption are technically possible,
-but should you wish to check for bugs like this,
-%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
-symbolic execution \cite{klee} would be more appropriate.
-
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of Iterative Deepening convergence}
-
-We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em na\"{i}ve state space}, and call the condition of testing all its interleavings {\em convergence}.
-(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
-Hence, our convergence statement is as follows:
-
-\begin{theorem}[Total verification]
-	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
-	it serves as a verification of all possible thread schedules of the given test program.
-	\label{thm:totalverif}
-\end{theorem}
-
-The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
-note that we are using the constructive contrapositive direction:
-Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
-
-%DPOR + Iterative Deepening will eventually test every nonequivalent thread interleaving that would be exposed by preempting at every single instruction
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-	\label{thm:convergence}
-\end{theorem}
-
-First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the na\"{i}ve state space
-are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
-
-\subsection{Saturation of data-race PPs}
-
-\begin{lemma}
-	If a bug can be exposed by some combination of preemptions on any data-race instruction and sync API boundaries,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-	\label{lem:saturation}
-\end{lemma}
-
-The main challenge of this proof is that some data-race candidates are nondeterministic;
-i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
-as shown in Figure~\ref{fig:nondet-dr}.
-Hence, the maximal state space of statically-available sync API PPs will not necessarily uncover all possible data-race PPs;
-Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
-However, the proof is surprisingly straightforward.
-
-\begin{proof}
-Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
-We induct on the $p$s to show that all data-race PPs will be reachable.
-\begin{itemize}
-	\item For the base case, $p_0$ required 
-	\item
-\end{itemize}
-\end{proof}
-
-% TODO make this be on the right column
-\begin{figure}[t]
-	\small
-	\begin{tabular}{rl}
-	& \multicolumn{1}{c}{~\texttt{int x = 0, y = 0;~~~~~~~~~~~~~~~~}} \\
-	& \multicolumn{1}{c}{\texttt{bool t1\_x = false, t1\_y = false;}} \\
-	& \multicolumn{1}{c}{\texttt{bool t2\_x = false, t2\_y = false;}} \\
-\\
-		& Thread 1 (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
-	1 & \texttt{\hilight{brickred}{x = x + 1;}} \\
-	2 & \texttt{t1\_x = true;} \\
-	3 & \texttt{// "if x raced"} \\
-	4 & \texttt{if (x == 1 \&\& t2\_x) \{} \\
-	5 & \texttt{~~~~\hilight{brickred}{y = y + 1;}} \\
-	6 & \texttt{~~~~t1\_y = true;} \\
-	7 & \texttt{~~~~// "if y raced"} \\
-	8 & \texttt{~~~~if (y == 1 \&\& t2\_y) \{} \\
-	9 & \texttt{~~~~~~~~panic();} \\
-	10 & \texttt{~~~~\}} \\
-	11 & \texttt{\}} \\
-	\end{tabular}
-	\caption{Not all data races will immediately be uncovered by interleaving threads around sync API PPs alone.
-Here, preempting during the race on line 1 is necessary to force the program into the control flow which can race on line 5.}
-	\label{fig:nondet-dr}
-\end{figure}
-
-\subsection{Equivalence of irrelevant PPs}
-
-\begin{lemma}
-	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:equivalent}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{1}
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-\end{theorem}
-
-\begin{proof}
-By Lemma~\ref{lem:equivalent}, the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
-Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of malloc-recycle soundness}
-\label{sec:proof}
-
-We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
-Hence, our soundness statement is as follows:
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-Though Figures~\ref{fig:recycle-bug} and \ref{fig:recycle-goal} show example programs, they do not capture all possible cases of how a true data race can fit the malloc-recycle pattern.
-We proceed by establishing what must be true of any such program, then casing on the ambiguous possibilities, and showing that PPs will exist where we need them to reorder the threads.
-
-For certain, there must be an access in one thread, followed by a free and malloc (we'll call them ``middle free'' and ``middle malloc''), each possibly from either thread, followed by an access from the other.
-If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
-WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
-
-\begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
-	then a non-malloc-recycle data race or a use-after-free bug will be identified.
-	\label{lem:reorder}
-\end{lemma}
-\begin{proof}
-By case on which threads the middle free and middle malloc came from.
-\begin{itemize}
-	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
-The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
-Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
-	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
-	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
-\end{itemize}
-If either the middle free or middle malloc came from a third thread, the case is the same as if it belonged to T1.
-\end{proof}
-
-The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
-However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
-%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
-and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
-So, WLOG, we say that the only relevant events are the ones we mention explicitly.
-
-By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
-We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
-Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
-
-\subsection{T2 originally malloced x}
-
-\begin{lemma}[Greedo]
-	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:greedo}
-\end{lemma}
-\begin{proof}
-Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
-By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
-and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
-Lemma~\ref{lem:reorder} finishes.
-\end{proof}
-
-This lemma also applies if a third thread was responsible for this malloc, as there would still be a thread switch in the same spot.
-
-\subsection{T1 originally malloced x}
-
-\begin{lemma}[Han]
-	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:han}
-\end{lemma}
-
-\begin{proof}
-We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
-Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
-
-%\begin{figure}[t]
-%	\small
-%\begin{tabular}{rll}
-%	& Thread 1 & Thread 2 \\
-%	0 & \texttt{lock();} & \\
-%	1 & \texttt{publish(x1);} & \\
-%	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-%	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-%	0 & \texttt{unlock();} & \\
-%	0 & & \texttt{lock();} \\
-%	2 & & \texttt{x2 = get\_published\_x();} \\
-%	0 & & \texttt{unlock();} \\
-%	5 & & \texttt{// x's memory recycled} \\
-%	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-%	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-%\end{tabular}
-%\caption{If the locksets are the same in T1 during publish and x1, you can't avoid the data race on publish+get.}
-%\label{fig:recycle-bug}
-%\end{figure}
-
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
-If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
-For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
-which we'll call {\tt p}.
-By the previous two statements, {\tt p} must be a data race of its own.
-%(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-
-	%1 & \texttt{p1 = \hilight{olivegreen}{malloc}(...);} & \\
-	1 & \texttt{\hilight{brickred}{p1->ptr = x1;}} & \\
-	2 & \texttt{publish(p1);} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(p1);} & \\
-	4 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	5 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-
-	6 & & \texttt{p2 = get\_published\_p();} \\
-	7 & & \texttt{// p's memory recycled} \\
-	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
-	9 & & \texttt{\hilight{brickred}{x2 = p2->ptr;}} \\
-
-	10 & & \texttt{// x's memory recycled} \\
-	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	12 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If the accesses used to publish {\tt x}'s address are a data race, their PPs may also be eliminated under the malloc-recycle pattern.
-Induction on the pointer structure leading to {\tt x} handles this case.}
-\label{fig:induction}
-\end{figure}
-
-Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
-as shown in Figure~\ref{fig:induction},
-we do not necessarily get the PP for free here.
-In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
-In \quicksand, data race PPs are not used immediately, but rather generate new state spaces to explore in the future. Anyway, under the maximal state space assumption, we will get to it eventually.}.
-We handle this with induction on T2's pointer chain leading to {\tt x}.
-
-\newcommand\publish[1]{{\tt p}$_{#1}$}
-\begin{itemize}
-	\item For the base case, the publish location \publish{0} is either in global memory, or shared directly using synchronization.
-		Non-heap memory data races are not subject to the malloc-recycle pattern, so will always get a data-race PP,
-		and use of synchronization always gets a PP in the maximal state space.
-	\item For the inductive step, a pointer \publish{n} is published in some heap memory \publish{n-1}\texttt{->ptr},
-		and we assume that however \publish{n-1} is shared to T2,
-		there will be a PP there
-		% where \publish{n-1} is shared
-		sufficient to make the \publish{n-1}\texttt{->ptr} access not malloc-recycle after DPOR.
-		%DPOR will test some interleaving in which T1's
-		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
-		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
-		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
-		% i.e., so that the p_n access no longer appears to be malloc-recycle.
-\end{itemize}
-
-Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location {\tt p}.
-Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
-% XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
-As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
-Lemma~\ref{lem:reorder} finishes.
-%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
-
-\end{proof}
-
-%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
-%As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{2}
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-\begin{proof}
-	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases of possible program structure are covered.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Heap overruns}
-\label{sec:owned}
-
-Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
-Figure~\ref{fig:overrun} shows an example.
-Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
-So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	7 & & \texttt{// x's memory recycled} \\
-	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	9 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Final possibility for how T2 can share T1's allocation address, and probably a security vulnerability to boot!}
-\label{fig:overrun}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{x1 = malloc(...);} & \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
-\label{fig:overrun-notenough}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
-\label{fig:overrun-goal}
-\end{figure}
-
-% TODO: Replace this with some real discussion.
-
-%{\bf Being Owned.}
-%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
-Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-%
-%
-%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
-If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
-this is because dpor ignores malloc's internal accesses from the shm conflict relation.
-
-what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
-BUT, the original one was not sound with respect to this bug to begin with.
-
-then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
-
-\begin{lemma}
-	asdfa
-	\label{lem:leia} % "there is another"
-\end{lemma}
-
-% TODO
-% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
-% for the purpose of iterative deepening, MHB is
-% better because it will find the data races much sooner, rather than requiring you
-% to explore the maximal state space before the race can be uncovered
-
-%\subsection{Malloc with internal mutexes}
-%
-%First we assume that the malloc implementation is guarded by
-%
-%
-%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-%
-%
-%\subsection{Malloc as a magic primitive}
-%
-%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
-%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
-%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
-%
-%Hence, we provide another proof that does not require
-%
-%%\cite{landslide} % uncomment post-double-blind review
-%
-%This spells QED so we are done.
-
-\section{Acknowledgements}
-
-{\em removed for double blind review}
-%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-\end{document}

### 2015-11-10 pldi-16/proof/proof.tex
split proof into files

--- 2015-11-05 pldi-16/proof.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Definitions}
-
-\subsection{Stateless model checking terms}
-
-\begin{definition}[Transition]
-A sequence of instructions executed by the program between two preemption points (PPs).
-\label{def:transition}
-\end{definition}
-Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-
-\begin{definition}[Must-happen-before (MHB)]
-	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
-\end{definition}
-All transitions of the same thread are trivially MHB-related.
-Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
-Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
-% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
-% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).
-
-Note how our {\em must}-happen-before relation differs from the conventional definition of happens-before (``observed to happen before'') \cite{lamport-clocks}.
-Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
-the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
-Because pure-HB data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
-However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock,
-our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
-}.
-
-\begin{definition}[Shared memory conflict]
-A pair of memory accesses between two threads to the same address where at least one of them is a write.
-\end{definition}
-Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
-
-\begin{definition}[Dynamic Partial Order Reduction (DPOR)]
-	A state-space search algorithm for stateless model checkers,
-	guaranteed to reorder transitions of two threads
-	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
-	\label{def:dpor}
-\end{definition}
-
-\subsection{Data race and other memory terms}
-
-\begin{definition}[Data race]
-A shared memory conflict where furthermore:
-\begin{itemize}
-	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
-	\item The threads' transitions are not related by MHB.
-\end{itemize}
-\end{definition}
-
-The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
-
-\begin{definition}[Reachable data race]
-A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
-or transitively also configured to use data-race PPs of other (already) reachable data races.
-\end{definition}
-
-%To prove a theory, generally sized super mario bros is not possible to complete, there is two step, one step for each F.
-To prove Iterative Deepening's convergence to full coverage,
-we are concerned with showing that any data race PP necessary to expose a bug will be reachable.
-
-\begin{definition}[False positive data race]
-	An apparent data race that cannot be observed in the opposite order from what was actually executed.
-\end{definition}
-
-False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
-changes some variable values when the threads are reordered, such that the memory addresses no longer collide.
-
-\begin{definition}[Malloc-recycle data race]
-	A data race where the address is contained in some heap-allocated memory, and between the two accesses, that memory was passed to free() and returned again by a subsequent malloc().
-\end{definition}
-
-Figures~\ref{fig:recycle} and \ref{fig:recycle-bug} show an example.
-In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.
-
-\begin{definition}[Use after free]
-	Any read or write to heap memory which was once allocated, but no longer is.
-\end{definition}
-
-These can immediately be identified as failures by a MC which tracks allocation state.
-%Most commonly this refers to accesses to a region already freed, but for brevity we also include
-% TODO: need heap block overrun?
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	3 & & \texttt{// x's memory recycled} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{False-positive malloc-recycle pattern. This is the common case for which we avoid creating new state spaces.}
-\label{fig:recycle}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	5 & & \texttt{// x's memory recycled} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Adversarial program which fits the malloc-recycle pattern, but nevertheless contains a true race.}
-\label{fig:recycle-bug}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & & \texttt{// x not free, so malloc's} \\
-	4 & & \texttt{// return value changes!} \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
-\label{fig:recycle-goal}
-\end{figure}
-
-\section{Intuition}
-
-In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
-
-{\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
-that set represents every single code location where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
-Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
-but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any PPs beyond the ones we consider are irrelevant.
-
-{\bf Intuition for malloc-recycle soundness.}
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
-Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Assumptions}
-
-This section documents the assumptions we make about the concurrency model, language model, and test environment,
-and discusses some limitations that may arise from these assumptions.
-
-\subsection{Assumptions for both proofs}
-
-{\bf Maximal state space.}
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
-and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
-
-Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
-However, Iterative Deepening aims to test the most important interleavings with the time available,
-so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
-%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
-%In that case, our point is to justify deprioritizing those jobs so aggressively.
-
-%{\bf Low overhead.}
-%We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-%Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
-
-{\bf Shared memory thread communication.}
-We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
-is through either shared memory or a correctly-instrumented sync API.
-Both DPOR and data-race detection rely on this assumption,
-as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
-For brevity in these proofs, we refer to all thread communication as shared memory,
-and assume that other mechanisms, such as system calls that access the filesystem, could be instrumented to fit the same model.
-
-{\bf Schedule nondeterminism only.}
-We discount the possibility of other types of nondeterminism,
-such as program input nondeterminism (including randomness/timestamps) or
-store-buffer nondeterminism on weak-memory architectures.
-We refer the reader to \cite{klee,portend} for related work on the former, and to \cite{tsopso} for the latter.
-
-\subsection{Assumptions for Iterative Deepening convergence}
-
-{\bf Locks are correct.}
-Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
-we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
-This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
-although we do not model RCU \cite{rcu}.
-Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
-
-Should the user wish to verify these properties of locks,
-they could either run a locking test separately (as we suggest in our paper),
-which would cheaply test the locks to an extent limited by the separate test case;
-or they could remove the data-race analysis's lockset tracking,
-which would expensively test all the main program's required locking properties in tandem with the program itself.
-
-{\bf No forcibly blocking other threads.}
-We assume the sync API behaves much like pthreads and/or message-passing:
-that is, a thread can cause itself to block on some condition,
-and other threads may wake it up,
-but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
-
-\subsection{Assumptions for malloc-recycle soundness}
-
-{\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
-%particular allocation pattern regarding adjacency/coalescing/reuse.
-In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
-in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-Furthermore, we configured DPOR to ignore the internal heap metadata accesses
-when tracking shared memory conflicts to achieve greater state space reduction
-(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
-%for the purpose of avoiding malloc-only shared memory conflicts.
-This is not without consequences; see section~\ref{sec:owned}.
-
-{\bf Sharing heap addresses.}
-Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
-Because we are testing C programs, any bizarre violations of this assumption are technically possible,
-but should you wish to check for bugs like this,
-%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
-symbolic execution \cite{klee} would be more appropriate.
-
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of Iterative Deepening convergence}
-
-We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em na\"{i}ve state space}, and call the condition of testing all its interleavings {\em convergence}.
-(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
-Hence, our convergence statement is as follows:
-
-\begin{theorem}[Total verification]
-	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
-	it serves as a verification of all possible thread schedules of the given test program.
-	\label{thm:totalverif}
-\end{theorem}
-
-The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
-note that we are using the constructive contrapositive direction:
-Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
-
-%DPOR + Iterative Deepening will eventually test every nonequivalent thread interleaving that would be exposed by preempting at every single instruction
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-	\label{thm:convergence}
-\end{theorem}
-
-First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the na\"{i}ve state space
-are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
-
-\subsection{Saturation of data-race PPs}
-
-\begin{lemma}
-	If a bug can be exposed by some combination of preemptions on any data-race instruction and sync API boundaries,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-	\label{lem:saturation}
-\end{lemma}
-
-The main challenge of this proof is that some data-race candidates are nondeterministic;
-i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
-as shown in Figure~\ref{fig:nondet-dr}.
-Hence, the maximal state space of statically-available sync API PPs will not necessarily uncover all possible data-race PPs;
-Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
-However, the proof is surprisingly straightforward.
-
-\begin{proof}
-Let $\mathcal{I} = \{p_0, p_1 \dots p_n\}$ be the interleaving and sequence of preemption points in the premise of the lemma which exposed a bug.
-We induct on the $p$s to show that all data-race PPs will be reachable.
-\begin{itemize}
-	\item For the base case, $p_0$ required 
-	\item
-\end{itemize}
-\end{proof}
-
-% TODO make this be on the right column
-\begin{figure}[t]
-	\small
-	\begin{tabular}{rl}
-	& \multicolumn{1}{c}{~\texttt{int x = 0, y = 0;~~~~~~~~~~~~~~~~}} \\
-	& \multicolumn{1}{c}{\texttt{bool t1\_x = false, t1\_y = false;}} \\
-	& \multicolumn{1}{c}{\texttt{bool t2\_x = false, t2\_y = false;}} \\
-\\
-		& Thread 1 (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
-	1 & \texttt{\hilight{brickred}{x = x + 1;}} \\
-	2 & \texttt{t1\_x = true;} \\
-	3 & \texttt{// "if x raced"} \\
-	4 & \texttt{if (x == 1 \&\& t2\_x) \{} \\
-	5 & \texttt{~~~~\hilight{brickred}{y = y + 1;}} \\
-	6 & \texttt{~~~~t1\_y = true;} \\
-	7 & \texttt{~~~~// "if y raced"} \\
-	8 & \texttt{~~~~if (y == 1 \&\& t2\_y) \{} \\
-	9 & \texttt{~~~~~~~~panic();} \\
-	10 & \texttt{~~~~\}} \\
-	11 & \texttt{\}} \\
-	\end{tabular}
-	\caption{Not all data races will immediately be uncovered by interleaving threads around sync API PPs alone.
-Here, preempting during the race on line 1 is necessary to force the program into the control flow which can race on line 5.}
-	\label{fig:nondet-dr}
-\end{figure}
-
-\subsection{Equivalence of irrelevant PPs}
-
-\begin{lemma}
-	For any interleaving in the na\"{i}ve state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:equivalent}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{1}
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-\end{theorem}
-
-\begin{proof}
-By Lemma~\ref{lem:equivalent}, the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
-Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of malloc-recycle soundness}
-\label{sec:proof}
-
-We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
-Hence, our soundness statement is as follows:
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-Though Figures~\ref{fig:recycle-bug} and \ref{fig:recycle-goal} show example programs, they do not capture all possible cases of how a true data race can fit the malloc-recycle pattern.
-We proceed by establishing what must be true of any such program, then casing on the ambiguous possibilities, and showing that PPs will exist where we need them to reorder the threads.
-
-For certain, there must be an access in one thread, followed by a free and malloc (we'll call them ``middle free'' and ``middle malloc''), each possibly from either thread, followed by an access from the other.
-If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
-WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
-
-\begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
-	then a non-malloc-recycle data race or a use-after-free bug will be identified.
-	\label{lem:reorder}
-\end{lemma}
-\begin{proof}
-By case on which threads the middle free and middle malloc came from.
-\begin{itemize}
-	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
-The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
-Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
-	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
-	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
-\end{itemize}
-If either the middle free or middle malloc came from a third thread, the case is the same as if it belonged to T1.
-\end{proof}
-
-The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
-However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
-%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
-and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
-So, WLOG, we say that the only relevant events are the ones we mention explicitly.
-
-By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
-We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
-Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
-
-\subsection{T2 originally malloced x}
-
-\begin{lemma}[Greedo]
-	If T2 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:greedo}
-\end{lemma}
-\begin{proof}
-Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
-By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
-and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
-Lemma~\ref{lem:reorder} finishes.
-\end{proof}
-
-This lemma also applies if a third thread was responsible for this malloc, as there would still be a thread switch in the same spot.
-
-\subsection{T1 originally malloced x}
-
-\begin{lemma}[Han]
-	If T1 originally malloced the block containing {\tt x}, DPOR will reorder the threads to uncover a non-malloc-recycle race or a use-after-free bug.
-	\label{lem:han}
-\end{lemma}
-
-\begin{proof}
-We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
-Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
-
-%\begin{figure}[t]
-%	\small
-%\begin{tabular}{rll}
-%	& Thread 1 & Thread 2 \\
-%	0 & \texttt{lock();} & \\
-%	1 & \texttt{publish(x1);} & \\
-%	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-%	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-%	0 & \texttt{unlock();} & \\
-%	0 & & \texttt{lock();} \\
-%	2 & & \texttt{x2 = get\_published\_x();} \\
-%	0 & & \texttt{unlock();} \\
-%	5 & & \texttt{// x's memory recycled} \\
-%	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-%	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-%\end{tabular}
-%\caption{If the locksets are the same in T1 during publish and x1, you can't avoid the data race on publish+get.}
-%\label{fig:recycle-bug}
-%\end{figure}
-
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
-If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
-For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
-which we'll call {\tt p}.
-By the previous two statements, {\tt p} must be a data race of its own.
-%(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-
-	%1 & \texttt{p1 = \hilight{olivegreen}{malloc}(...);} & \\
-	1 & \texttt{\hilight{brickred}{p1->ptr = x1;}} & \\
-	2 & \texttt{publish(p1);} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(p1);} & \\
-	4 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	5 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-
-	6 & & \texttt{p2 = get\_published\_p();} \\
-	7 & & \texttt{// p's memory recycled} \\
-	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
-	9 & & \texttt{\hilight{brickred}{x2 = p2->ptr;}} \\
-
-	10 & & \texttt{// x's memory recycled} \\
-	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	12 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If the accesses used to publish {\tt x}'s address are a data race, their PPs may also be eliminated under the malloc-recycle pattern.
-Induction on the pointer structure leading to {\tt x} handles this case.}
-\label{fig:induction}
-\end{figure}
-
-Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
-as shown in Figure~\ref{fig:induction},
-we do not necessarily get the PP for free here.
-In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
-In \quicksand, data race PPs are not used immediately, but rather generate new state spaces to explore in the future. Anyway, under the maximal state space assumption, we will get to it eventually.}.
-We handle this with induction on T2's pointer chain leading to {\tt x}.
-
-\newcommand\publish[1]{{\tt p}$_{#1}$}
-\begin{itemize}
-	\item For the base case, the publish location \publish{0} is either in global memory, or shared directly using synchronization.
-		Non-heap memory data races are not subject to the malloc-recycle pattern, so will always get a data-race PP,
-		and use of synchronization always gets a PP in the maximal state space.
-	\item For the inductive step, a pointer \publish{n} is published in some heap memory \publish{n-1}\texttt{->ptr},
-		and we assume that however \publish{n-1} is shared to T2,
-		there will be a PP there
-		% where \publish{n-1} is shared
-		sufficient to make the \publish{n-1}\texttt{->ptr} access not malloc-recycle after DPOR.
-		%DPOR will test some interleaving in which T1's
-		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
-		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
-		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
-		% i.e., so that the p_n access no longer appears to be malloc-recycle.
-\end{itemize}
-
-Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location {\tt p}.
-Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
-% XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
-As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
-Lemma~\ref{lem:reorder} finishes.
-%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
-
-\end{proof}
-
-%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
-%As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{2}
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
-\end{theorem}
-\begin{proof}
-	Between Lemmas \ref{lem:greedo} and \ref{lem:han}, all cases of possible program structure are covered.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Heap overruns}
-\label{sec:owned}
-
-Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
-Figure~\ref{fig:overrun} shows an example.
-Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
-So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	7 & & \texttt{// x's memory recycled} \\
-	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	9 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Final possibility for how T2 can share T1's allocation address, and probably a security vulnerability to boot!}
-\label{fig:overrun}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{x1 = malloc(...);} & \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
-\label{fig:overrun-notenough}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
-\label{fig:overrun-goal}
-\end{figure}
-
-% TODO: Replace this with some real discussion.
-
-%{\bf Being Owned.}
-%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
-Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-%
-%
-%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
-If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
-this is because dpor ignores malloc's internal accesses from the shm conflict relation.
-
-what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
-BUT, the original one was not sound with respect to this bug to begin with.
-
-then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
-
-\begin{lemma}
-	asdfa
-	\label{lem:leia} % "there is another"
-\end{lemma}
+\input{di}
+\input{assumptions}
+\input{convergence}
+\input{recycle}
 
@@
 
-%\subsection{Malloc with internal mutexes}
-%
-%First we assume that the malloc implementation is guarded by
-%
-%
-%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-%
-%
-%\subsection{Malloc as a magic primitive}
-%
-%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
-%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
-%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
-%
-%Hence, we provide another proof that does not require
-%
-%%\cite{landslide} % uncomment post-double-blind review
-%
-%This spells QED so we are done.
-
 \section{Acknowledgements}

### 2015-11-10 pldi-16/proof/recycle.tex
split proof into files

--- 2015-11-04 pldi-16/proof.tex
@@
-%
-% LaTeX template for prepartion of submissions to PLDI'15
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi15}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{amsthm}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-\usepackage{graphicx}
-%\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-%\doublespacing
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-
-\begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
-
-%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
-\newcommand\quicksand{\textsc{Quicksand}}
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-
-%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
-\title{Soundness Proofs for Iterative Deepening}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
-Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
-This document presents two proofs concerning the technique.
-
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a naive MC preempting on every single instruction.
-In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
-it serves as a full verification of {\em all possible} thread schedules of the given test case.
-
-Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
-Despite many powerful reduction techniques, the state spaces are still exponentially sized,
-so any way to avoid exploring some in advance is obviously beneficial.
-We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.
-
-This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Definitions}
-
-\subsection{Stateless model checking terms}
-
-\begin{definition}[Transition]
-A sequence of instructions executed by the program between two preemption points (PPs).
-\label{def:transition}
-\end{definition}
-Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-
-\begin{definition}[Must-happen-before (MHB)]
-	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
-\end{definition}
-All transitions of the same thread are trivially MHB-related.
-Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
-Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
-% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
-% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).
-
-Note how our {\em must}-happen-before relation differs from the conventional definition of happens-before (``observed to happen before'') \cite{lamport-clocks}.
-Our use of MHB matches the ``limited happens-before'' used in \cite{hybriddatarace} and \cite{tsan};
-the advantage of this over pure-happens-before detectors in producing fewer false negatives is well-argued in those prior works\footnote{
-Because pure-HB data race detectors avoid false positives altogether, they would have no trouble avoiding our malloc-recycle false positives.
-However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock,
-our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
-}.
-
-\begin{definition}[Shared memory conflict]
-A pair of memory accesses between two threads to the same address where at least one of them is a write.
-\end{definition}
-Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
-
-\begin{definition}[Dynamic Partial Order Reduction (DPOR)]
-	A state-space search algorithm for stateless model checkers,
-	guaranteed to reorder transitions of two threads
-	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
-	\label{def:dpor}
-\end{definition}
-
-\subsection{Data race and other memory terms}
-
-\begin{definition}[Data race]
-A shared memory conflict where furthermore:
-\begin{itemize}
-	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
-	\item The threads' transitions are not related by MHB.
-\end{itemize}
-\end{definition}
-
-The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
-
-\begin{definition}[Reachable data race]
-A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
-or transitively also configured to use data-race PPs of other (already) reachable data races.
-\end{definition}
-
-%To prove a theory, generally sized super mario bros is not possible to complete, there is two step, one step for each F.
-To prove Iterative Deepening's convergence to full coverage,
-we are concerned with showing that any data race PP necessary to expose a bug will be reachable.
-
-\begin{definition}[False positive data race]
-	An apparent data race that cannot be observed in the opposite order from what was actually executed.
-\end{definition}
-
-False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
-changes some variable values when the threads are reordered, such that the memory addresses no longer collide.
-
-\begin{definition}[Malloc-recycle data race]
-	A data race where the address is contained in some heap-allocated memory, and between the two accesses, that memory was passed to free() and returned again by a subsequent malloc().
-\end{definition}
-
-Figures~\ref{fig:recycle} and \ref{fig:recycle-bug} show an example.
-In the case of malloc-recycle false positives, the allocation heap is the ``other shared state'' mentioned in the previous definition, and malloc's return value is the variable value that changed.
-
-\begin{definition}[Use after free]
-	Any read or write to heap memory which was once allocated, but no longer is.
-\end{definition}
-
-These can immediately be identified as failures by a MC which tracks allocation state.
-%Most commonly this refers to accesses to a region already freed, but for brevity we also include
-% TODO: need heap block overrun?
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	3 & & \texttt{// x's memory recycled} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{False-positive malloc-recycle pattern. This is the common case for which we avoid creating new state spaces.}
-\label{fig:recycle}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	5 & & \texttt{// x's memory recycled} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{Adversarial program which fits the malloc-recycle pattern, but nevertheless contains a true race.}
-\label{fig:recycle-bug}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x1);} & \\
-	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & & \texttt{// x not free, so malloc's} \\
-	4 & & \texttt{// return value changes!} \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
-\label{fig:recycle-goal}
-\end{figure}
-
-\section{Intuition}
-
-In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
-
-{\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
-that set represents every single code location where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
-Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
-but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any PPs beyond the ones we consider are irrelevant.
-
-{\bf Intuition for malloc-recycle soundness.}
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
-while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
-Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Assumptions}
-
-This section documents the assumptions we make about the concurrency model, language model, and test environment,
-and discusses some limitations that may arise from these assumptions.
-
-\subsection{Assumptions for both proofs}
-
-{\bf Maximal state space.}
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
-and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
-
-Using only a subset of PPs or aborting early due to time-out could each ruin our ability to reach the goal interleaving or goal state space.
-However, Iterative Deepening aims to test the most important interleavings with the time available,
-so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
-%An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
-%In that case, our point is to justify deprioritizing those jobs so aggressively.
-
-%{\bf Low overhead.}
-%We assume the model checker can identify malloc-recycle data races with little or no overhead beyond what's already associated with data race detection.
-%Our MC already tracks the heap state, so we implement this check for free with a simple generation counter.
-
-{\bf Shared memory thread communication.}
-We assume that the only way for two threads' transitions to affect each other's behaviour, should they be reordered,
-is through either shared memory or a correctly-instrumented sync API. Both DPOR and data-race detec
-
-{\bf Schedule nondeterminism only.}
-We discount the possibility of other types of nondeterminism,
-such as program input nondeterminism (including randomness/timestamps) or
-store-buffer nondeterminism on weak-memory architectures.
-We refer the reader to \cite{klee,portend} for related work on the former, and to \cite{tsopso} for the latter.
-
-\subsection{Assumptions for Iterative Deepening convergence}
-
-{\bf Locks are correct.}
-Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
-we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
-This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
-although we do not model RCU \cite{rcu}.
-Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
-
-Should the user wish to verify these properties of locks,
-they could either run a locking test separately (as we suggest in our paper),
-which would cheaply test the locks to an extent limited by the separate test case;
-or they could remove the data-race analysis's lockset tracking,
-which would expensively test all the main program's required locking properties in tandem with the program itself.
-
-\subsection{Assumptions for malloc-recycle soundness}
-
-{\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
-%particular allocation pattern regarding adjacency/coalescing/reuse.
-In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
-in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-Furthermore, we configured DPOR to ignore the internal heap metadata accesses
-when tracking shared memory conflicts to achieve greater state space reduction
-(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
-%for the purpose of avoiding malloc-only shared memory conflicts.
-This is not without consequences; see section~\ref{sec:owned}.
-
-{\bf Sharing heap addresses.}
-Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
-Because we are testing C programs, any bizarre violations of this assumption are technically possible,
-but should you wish to check for bugs like this,
-%we would recommend a data-flow analysis which is much cheaper than model checking anyway.
-symbolic execution \cite{klee} would be more appropriate.
-
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of Iterative Deepening convergence}
-
-We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em naive maximal state space}, and call the condition of testing all its interleavings {\em convergence}.
-(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
-Hence, our convergence statement is as follows:
-
-\begin{theorem}[Total verification]
-	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
-	it serves as a verification of all possible thread schedules of the given test program.
-	\label{thm:totalverif}
-\end{theorem}
-
-The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
-note that we are using the constructive contrapositive direction:
-Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
-
-%DPOR + Iterative Deepening will eventually test every nonequivalent thread interleaving that would be exposed by preempting at every single instruction
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-	\label{thm:convergence}
-\end{theorem}
-
-First we will prove a slightly weaker statement, limiting the naive approach's PPs to data races,
-then strengthen it by showing that the all interleavings in the naive maximal state space
-are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
-
-\subsection{Saturation of data-race PPs}
-
-\begin{lemma}
-	If a bug is exposed by any interleaving comprising any combination of data-race PPs and sync API PPs,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-	\label{lem:saturation}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
-
-\subsection{Saturation of data-race PPs}
-
-\begin{lemma}
-	For any interleaving in the naive maximal state space, there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:equivalent}
-\end{lemma}
-
-\begin{proof}
-TODO % TODO
-\end{proof}
-
-\subsection{Conclusion}
-
-\setcounter{theorem}{1}
-\begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the naive maximal state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
-\end{theorem}
-
-\begin{proof}
-By Lemma~\ref{lem:equivalent}, the naive interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
-Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Proof of malloc-recycle soundness}
@@
 
-% TODO
-% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
-% for the purpose of iterative deepening, MHB is
-% better because it will find the data races much sooner, rather than requiring you
-% to explore the maximal state space before the race can be uncovered
-
 %\subsection{Malloc with internal mutexes}
@@
 %This spells QED so we are done.
-
-\section{Acknowledgements}
-
-{\em removed for double blind review}
-%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
-
-\bibliographystyle{abbrvnat}
-\bibliography{citations}{}
-\end{document}

### 2015-11-10 pldi-16/proof/assumptions.tex
start proving saturation for real

@@
 
+{\bf Halting.}
+From a certain perspective, proving convergence is the same as proving completeness,
+which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
+However, being already limited by practical real-world CPU budgets,
+we are already accepting that many tests will time out.
+We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
+Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
+If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
+In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
+i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.
+% , such as for a program trying to find a counterexample for Sally's Conjecture.
+% TODO: cite some unproved mathematical conjecture by a woman, as an example of an undecidable program?
+
 \subsection{Assumptions for malloc-recycle soundness}

### 2015-11-10 pldi-16/proof/convergence.tex
start proving saturation for real

@@
 
-%First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
-%then strengthen it by showing that the all interleavings in the na\"{i}ve state space
-%are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.
-
-%%%%
-
 \newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
@@
 \newcommand\pai{\ensuremath{p_{\alpha{}i}}}
+\newcommand\tai{\ensuremath{t_{\alpha{}i}}}
 Let
@@
 \]
-be an interleaving trace, where the element $(t_{\alpha{}i}, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
+be an interleaving trace, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
 We will use the following notation:
 \begin{itemize}
-	\item $\ppnext{p_{\alpha{}i}}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
-	\item $\ppinstr{p_{\alpha{}i}}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
-	\item $\ppothers{p_{\alpha{}i}}$ indicates the set of all transitions by other threads between $t_{\alpha{}i}$ and $t_{\alpha{}(i+1)}$.
+	\item $\ppnext{\pai}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
+	\item $\ppinstr{\pai}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
+	\item $\ppothers{\pai}$ indicates the set of all transitions by other threads between $\tai$ and $t_{\alpha{}(i+1)}$.
 	\item $\conflicts{t_\alpha,t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% TODO: come up with a better name
 \begin{definition}[Relevant preemption point]
@@
 			% rho for relevant.
-		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $t_{\alpha{}i}$ contains a sync API boundary instruction that's not already a PP,
+		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $\tai$ contains a sync API boundary instruction that's not already a PP,
 			then let $\rho$ be either the first instruction by $\ppnext{\pai}$ among the conflicts or the first sync API instruction, whichever comes first.
@@
 				\\
-				$\mathcal{I}' = \{...,$&$(t_{\alpha{}i} \cup \pppfx{\rho}, \pai'),$& \\
+				$\mathcal{I}' = \{...,$&$(\tai \cup \pppfx{\rho}, \pai'),$& \\
 									   &$\ppothers{\pai},$& \\
@@
 			\[
-				\mathcal{I}' = \{..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
+				\mathcal{I}' = \{..., (\tai \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
 			\]
@@
 
-%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
-or transitively also configured to use data-race PPs of other (already) reachable data races.
+or transitively also configured to use data-race PPs of other
+%already
+reachable data races.
 \end{definition}
@@
 
-\begin{lemma}
-	Given a fully-relevant interleaving
-	\[
-		\mathcal{I} = \{ ..., (t_{\alpha{}i},\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
-	\]
-	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
-	\[
-		\mathcal{J} = \{ ..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
-	\]
-	are reachable, then $\pai$ is reachable.
-	\label{lem:coalesce}
-\end{lemma}
-
-Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
-i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
-and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
-Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
-Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.
-
-% TODO: What if ppothers changes based on ppinstr(pai)?
-You say, there must exist some J where ppothers has fewer PPs.
-You just stop preempting!
-Those preemptions don't matter.
-Gotta go fast.
-Gotta keep the induction metric of getting fewer pps in J than in I.
-
-%%%
+%%% weird backwards-induction thingie i came up with at the airport that doesn't work
+%\begin{lemma}
+%	Given a fully-relevant interleaving
+%	\[
+%		\mathcal{I} = \{ ..., (\tai,\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
+%	\]
+%	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
+%	\[
+%		\mathcal{J} = \{ ..., (\tai \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
+%	\]
+%	are reachable, then $\pai$ is reachable.
+%	\label{lem:coalesce}
+%\end{lemma}
+%
+%Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
+%i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
+%and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
+%Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
+%Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.
 
 \begin{lemma}
-	If a bug can be exposed by any fully-relevant interleaving, %some combination of preemptions on any data-race instruction and sync API boundaries,
-	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
+	All PPs of any fully-relevant interleaving are reachable.
 	\label{lem:saturation}
@@
 % If not, then we are done, and this one is reachable.
-% If so, then we ask, which transition among others(other_PP) (this includes us) does instr(other_PP) conflict with?
-% <vague handwavy part>
-% Then we show that there must be a trace with next(other_PP) interleaved to be before next(my_PP) where, even if the bug doesn't show up, the data race still shows up, which makes other_PP reachable.
-% </vague handwavy part>
-% Since other_PP is reachable then my_PP is reachable.
+% I
+
+\newcommand\pbh{\ensuremath{p_{\beta{}h}}}
+\newcommand\paj{\ensuremath{p_{\alpha{}j}}}
+\newcommand\pbk{\ensuremath{p_{\beta{}k}}}
+%\newcommand\tak{\ensuremath{t_{\alpha{}k}}}
+%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
+\newcommand\tbh{\ensuremath{t_{\beta{}h}}}
+\newcommand\tbk{\ensuremath{t_{\beta{}k}}}
+\newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}
 
@@
 Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
-We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
-Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
-i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions.
+We proceed by inducting on the PPs according to the order of their preemptions.
+(Note that this is not necessarily the same as the order of the racing instructions $\ppinstr\pai$, which occur in $\ppnext\pai$, not in \tai.)
+For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbh$ with $\tbh \prec \tai \in \mathcal{I}$, $\pbh$ is reachable.
+We must show that a data race involving $\ppinstr\pai$
+is reachable, and we are not allowed to use $\pai$ until we find the data race between $\ppinstr\pai$ and $\ppothers\pai$.
+
+Consider the alternate interleaving prefix:
+\[
+	\mathcal{I}' = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
+\]
+where we have reordered $\ppnext\pai$ to before the execution of $\ppothers\pai$.
+Here, $\paj$ is the first sync API PP in $\alpha$ after $\pai$ (as $p_{\alpha(i+1)}$ may be a not-yet-reachable data race PP),
+and $\ppnext\pai'$ may include transitions of $\alpha$ beyond $\ppnext\pai$ itself.
+%and for brevity, we will abbreviate the union of transitions as $\tak$.
+Then, $\ppothers\pai'$ are the other threads' transitions from our target interleaving.
+except they may be altered by the presence of some {\em other} data race in $\ppnext\pai'$ occurring after $\ppinstr\pai$.
+Likewise, $\ppnext\pai'$ may be altered by some other data race in $\ppothers\pai'$.
+The only certainty so far is that $\ppnext\pai'$ begins with $\ppinstr\pai$.
+
+It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
+could be reordered to after $\ppothers\pai'$,
+eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
+Unfortunately, $\mathcal{I}'$ is not reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
+which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
+Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
+which we will abbreviate:
+\[
+	\mathcal{I}'' = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
+\]
+For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
+starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
+and ending with the next sync API PP, which we'll call $\pbk$.
+It follows from sync API PPs being reachable by definition,
+and from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
+that $\mathcal{I''}$ is reachable.
+
+%%%%%%%%%%%%%%%%% this didn't work out
+%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
+%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
+%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
+%Why use this strange ordering, rather than simply inducting by the order of preemptions?
 %Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
-%%% well, the way we do it, might not be the first reachable one either...
+%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
+%
+%\begin{figure}[t]
+%	TODO - example of why ppnext
+%	\caption{TODO}
+%	\label{fig:why-ppnext}
+%\end{figure}
+%%%% well, the way we do it, might not be the first reachable one either...
+%
+%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
+%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
+%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
+%To show $\pai$ is reachable, the proof has two parts.
+%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
+%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
+%%%%%%%%%%%%%%%%%%%
 
-\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
-For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj}$ occurring before $\ppnext{\pai}$ in $\mathcal{I}$, $\pbj$ is reachable.
-% Now to show \pai is reachable.
+% TODO
 
@@
 %the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
+By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption\footnote{
+		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
+		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
 Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.

### 2015-11-10 pldi-16/proof/convergence.tex
finish convergence proof, i think

@@
 
-% TODO: Here's what to do.
-% Induct on the PPs by the order of next(my_PP) in the trace.
-% Inductive assumption gives that all prior other data race PPs with next(other_PP) before next(my_PP) are reachable,
-% Is there a prior data race PP with next(other_PP) *AFTER* next(my_PP)?
-% If not, then we are done, and this one is reachable.
-% I
-
 \newcommand\pbh{\ensuremath{p_{\beta{}h}}}
@@
 \newcommand\pbk{\ensuremath{p_{\beta{}k}}}
-%\newcommand\tak{\ensuremath{t_{\alpha{}k}}}
-%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
 \newcommand\tbh{\ensuremath{t_{\beta{}h}}}
-\newcommand\tbk{\ensuremath{t_{\beta{}k}}}
+\newcommand\tgk{\ensuremath{t_{\gamma{}k}}}
+\newcommand\tgka{\ensuremath{t_{\gamma{}k1}}}
+\newcommand\tgkb{\ensuremath{t_{\gamma{}k2}}}
+\newcommand\tdl{\ensuremath{t_{\delta{}l}}}
 \newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}
@@
 eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
-Unfortunately, $\mathcal{I}'$ is not reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
+Unfortunately, $\mathcal{I}'$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
 which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
@@
 starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
-and ending with the next sync API PP, which we'll call $\pbk$.
-It follows from sync API PPs being reachable by definition,
-and from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
+and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
+It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
+and from sync API PPs being reachable by definition,
 that $\mathcal{I''}$ is reachable.
+
+Again, the only certainly-executed instruction by any thread after $\tai$ is $\ppinstr\pai$.
+It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
+until a different data race PP between two other threads is discovered,
+as shown in Figure~\ref{fig:threethreads}.
+
+\begin{figure}[t]
+	TODO % TODO
+	\caption{In this example, even though thread 1's data-race PP must occur first in a buggy interleaving, it cannot be reached until after Iterative Deepening first reaches the later-occurring data-race PP between threads 2 and 3.}
+	\label{fig:threethreads}
+\end{figure}
+
+Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
+We assume that a state space $\mathcal{S}$ containing $\mathcal{I}''$,
+plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
+will be reachable.
+(In the base case, the number of new data-race PPs is 0, and $\mathcal{I}''$'s reachability is justified above.)
+We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
+a new unique data-race PP nevertheless will be.
+
+By the definition of fully-relevant interleaving, it must be possible for some other thread $\beta$ to execute an instruction that conflicts with $\ppinstr\pai$.
+%
+If this conflict is not observed in $\mathcal{S}$,
+it means that some preceding thread communication (involving any two threads $\gamma$ and $\delta$, either potentially the same or different as $\alpha$ or $\beta$) must exist to have suppressed it,
+and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
+%
+By the converse of our definition of DPOR, if a possible (conflicting, concurrent) thread interleaving is not tested,
+% XXX: Handwave?
+I'm not entirely convinced by this step.
+How does it justify skipping all the way to the first possible "reachable" data race among the chain of PPs needed to expose \beta's communication with \ppinstr\pai?
+then there must exist some transition with a necessary site of interleaving in the middle of it, rather than at its boundaries.
+
+Let $\tgk$ be this transition, $\tgka$ and $\tgkb$ be the result of splitting at that site, and $\tdl$ be the communicating transition of the other thread.
+%
+If the interleaving $\{\tgka, \tdl', \tgkb'\}$ can produce a different program behaviour than $\{\tgka, \tgkb, \tdl\}$,
+then again by the soundness of DPOR, $\tgkb$ and $\tdl$ must contain some shared memory conflict.
+By the maximal state space assumption, all sync API PPs are enabled, so the locks held by $\gamma$ cannot change during $\tgk$,
+and if $\tdl$ can be reordered to the middle of $\tgk$, the locksets of $\gamma$ and $\delta$ cannot overlap,
+and likewise there is no sync-enforced MHB relation.
+Hence the memory conflict between $\tdl$ and $\tgkb$ will be identified as a data-race.
+Because $\tgk$ was split such that the first instruction of $\tgkb$ was not previously a PP, the data-race was not already enabled as a PP.
+
+Hence either a data-race including $\ppinstr\pai$ will be identified,
+or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
+but by the halting assumption, this would constitute an infinite loop bug, which is sufficient.
+(Alternatively, for any program with a finitely-sized instruction listing, the number of unique instruction pairs is finite.)
+This finishes our inner induction.
+Hence $\pai$ is reachable, finishing our outer induction.
+Hence all PPs in $\mathcal{I}$ are reachable.
+\end{proof}
 
@@
 %\begin{figure}[t]
-%	TODO - example of why ppnext
-%	\caption{TODO}
+%	\caption{to do}
 %	\label{fig:why-ppnext}
@@
 %%%%%%%%%%%%%%%%%%%
-
-% TODO
-
-\begin{itemize}
-	\item
-	\item
-\end{itemize}
-\end{proof}
 

### 2015-11-10 pldi-16/proof/convergence.tex
clarify convergence proof

@@
 and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
-It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
-and from sync API PPs being reachable by definition,
-that $\mathcal{I''}$ is reachable.
-
 Again, the only certainly-executed instruction by any thread after $\tai$ is $\ppinstr\pai$.
@@
 as shown in Figure~\ref{fig:threethreads}.
+
+It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
+and from sync API PPs being reachable by definition,
+that $\mathcal{I''}$ is reachable.
 
@@
 If this conflict is not observed in $\mathcal{S}$,
-it means that some preceding thread communication (involving any two threads $\gamma$ and $\delta$, either potentially the same or different as $\alpha$ or $\beta$) must exist to have suppressed it,
-and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
-%
-By the converse of our definition of DPOR, if a possible (conflicting, concurrent) thread interleaving is not tested,
 % XXX: Handwave?
@@
 How does it justify skipping all the way to the first possible "reachable" data race among the chain of PPs needed to expose \beta's communication with \ppinstr\pai?
-then there must exist some transition with a necessary site of interleaving in the middle of it, rather than at its boundaries.
+it means that some preceding thread communication (involving $\beta$ and any number of other threads, possibly including $\alpha$)
+must exist to have suppressed it,
+and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
+%
+By the soundness of DPOR \cite{dpor}, if a program behaviour is possible by interleaving threads at the boundaries of
+the given transitions, that interleaving will be tested.
+By contrapositive\footnote{
+Still constructive: again, going from $A \rightarrow B$ to $\neg B \rightarrow \neg A$.},
+because it was not tested, a necessary site of interleaving must be in the middle of some transition, rather than at its boundaries.
 

### 2015-11-10 pldi-16/proof/convergence.tex
Change notation I' -> J, I'' -> K

@@
 \[
-	\mathcal{I}' = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
+	\mathcal{J} = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
 \]
@@
 eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
-Unfortunately, $\mathcal{I}'$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
+Unfortunately, $\mathcal{J}$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
 which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
@@
 \[
-	\mathcal{I}'' = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
+	\mathcal{K} = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
 \]
@@
 Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
-We assume that a state space $\mathcal{S}$ containing $\mathcal{I}''$,
+We assume that a state space $\mathcal{S}$ containing $\mathcal{K}$,
 plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
 will be reachable.
-(In the base case, the number of new data-race PPs is 0, and $\mathcal{I}''$'s reachability is justified above.)
+(In the base case, the number of new data-race PPs is 0, and $\mathcal{K}$'s reachability is justified above.)
 We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,

### 2015-11-11 pldi-16/proof/convergence.tex
add colour and define soundness of dpor

@@
 	2 & \texttt{t1\_x = true;} \\
-	3 & \texttt{// "if x raced"} \\
+	3 & \texttt{\hilight{commentblue}{// "if x raced"}} \\
 	4 & \texttt{if (x == 1 \&\& t2\_x) \{} \\
@@
 	6 & \texttt{~~~~t1\_y = true;} \\
-	7 & \texttt{~~~~// "if y raced"} \\
+	7 & \texttt{~~~~\hilight{commentblue}{// "if y raced"}} \\
 	8 & \texttt{~~~~if (y == 1 \&\& t2\_y) \{} \\

### 2015-11-11 pldi-16/proof/di.tex
add colour and define soundness of dpor

@@
 \end{definition}
+
+The soundness of DPOR guarantees that if a program behaviour can possibly be exposed by any thread interleaving around the given transitions/PPs,
+that interleaving will eventually be tested by reordering only such conflicting transitions.
+In other words, reordering memory-independent thread transitions cannot possibly affect program behaviour.
 
@@
 	2 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	3 & & \texttt{// x's memory recycled} \\
+	3 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
+	5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
 	6 & & \texttt{publish(y);} \\
@@
 	4 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	5 & & \texttt{// x's memory recycled} \\
+	5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
@@
 	2 & & \texttt{x2 = get\_published\_x();} \\
-	3 & & \texttt{// x not free, so malloc's} \\
-	4 & & \texttt{// return value changes!} \\
+	3 & & \texttt{\hilight{commentblue}{// x not free, so malloc's}} \\
+	4 & & \texttt{\hilight{commentblue}{// return value changes!}} \\
 	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\

### 2015-11-11 pldi-16/proof/recycle.tex
add colour and define soundness of dpor

@@
 	6 & & \texttt{p2 = get\_published\_p();} \\
-	7 & & \texttt{// p's memory recycled} \\
+	7 & & \texttt{\hilight{commentblue}{// p's memory recycled}} \\
 	8 & & \texttt{q = \hilight{olivegreen}{malloc}(sizeof *q);} \\
@@
 
-	10 & & \texttt{// x's memory recycled} \\
+	10 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	11 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
@@
 	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{// TODO bounds check??} \\
+	2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
 	3 & & \texttt{x2 = \&z[50];} \\
@@
 	6 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-	7 & & \texttt{// x's memory recycled} \\
+	7 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	8 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\

### 2015-11-11 pldi-16/proof/convergence.tex
clean things up in convergence; clarify the xxx

@@
 
-%%% weird backwards-induction thingie i came up with at the airport that doesn't work
-%\begin{lemma}
-%	Given a fully-relevant interleaving
-%	\[
-%		\mathcal{I} = \{ ..., (\tai,\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
-%	\]
-%	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
-%	\[
-%		\mathcal{J} = \{ ..., (\tai \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
-%	\]
-%	are reachable, then $\pai$ is reachable.
-%	\label{lem:coalesce}
-%\end{lemma}
-%
-%Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
-%i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
-%and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
-%Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
-%Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.
-
 \begin{lemma}
@@
 
+{\bf Coalescing not-yet-reachable data race PPs.}
 Consider the alternate interleaving prefix:
@@
 and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
-Again, the only certainly-executed instruction by any thread after $\tai$ is $\ppinstr\pai$.
+Again, the only certainly-executed instruction by $\alpha$ after $\tai$ is $\ppinstr\pai$.
 It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
@@
 and from sync API PPs being reachable by definition,
-that $\mathcal{I''}$ is reachable.
+that $\mathcal{K}$ is reachable.
 
@@
 
+{\bf Finding the next reachable data-race PP.}
 Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
@@
 If this conflict is not observed in $\mathcal{S}$,
-% XXX: Handwave?
-I'm not entirely convinced by this step.
-How does it justify skipping all the way to the first possible "reachable" data race among the chain of PPs needed to expose \beta's communication with \ppinstr\pai?
 it means that some preceding thread communication (involving $\beta$ and any number of other threads, possibly including $\alpha$)
@@
 Still constructive: again, going from $A \rightarrow B$ to $\neg B \rightarrow \neg A$.},
+% XXX: Handwave.
+This neglects the possiblity that TWO necessariy sites of interleaving are as such, and that they are dependent (both preemptions must occur before either data race can be identified).
+% TODO: Prove that there must exist one single transition with next paragraph's property, without needing to preempt on any other such transition.
 because it was not tested, a necessary site of interleaving must be in the middle of some transition, rather than at its boundaries.
@@
 
-Hence either a data-race including $\ppinstr\pai$ will be identified,
+{\bf Reaching $\mathbf{p_{\alpha{}i}}$.}
+Hence either a data-race with $\ppinstr\pai$ will be identified,
 or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
@@
 \end{proof}
-
-%%%%%%%%%%%%%%%%% this didn't work out
-%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
-%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
-%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
-%Why use this strange ordering, rather than simply inducting by the order of preemptions?
-%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
-%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
-%
-%\begin{figure}[t]
-%	\caption{to do}
-%	\label{fig:why-ppnext}
-%\end{figure}
-%%%% well, the way we do it, might not be the first reachable one either...
-%
-%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
-%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
-%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
-%To show $\pai$ is reachable, the proof has two parts.
-%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
-%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
-%%%%%%%%%%%%%%%%%%%
 

### 2015-11-11 pldi-16/proof/convergence.tex
Admit that it's nonconstructive to conjure the existence of the first-reachable data race. Really finish proof.

@@
 \newcommand\tbh{\ensuremath{t_{\beta{}h}}}
+\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
 \newcommand\tgk{\ensuremath{t_{\gamma{}k}}}
@@
 We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
-a new unique data-race PP nevertheless will be.
+a new unique data-race PP will instead be reachable.
 
-By the definition of fully-relevant interleaving, it must be possible for some other thread $\beta$ to execute an instruction that conflicts with $\ppinstr\pai$.
+By the definition of fully-relevant interleaving, $\mathcal{I}$ guarantees that some other thread $\omega$ can execute a data-racing instruction with $\ppinstr\pai$.
 %
-If this conflict is not observed in $\mathcal{S}$,
-it means that some preceding thread communication (involving $\beta$ and any number of other threads, possibly including $\alpha$)
-must exist to have suppressed it,
-and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
+By the soundness of DPOR, if a program behaviour is possible by interleaving threads at the boundaries of
+the given transitions, that interleaving will be tested.
+By contrapositive,
+because $\omega$'s data-racing instruction was not tested in $\mathcal{S}$, one or more necessary sites of interleaving must be in the middle of some transition, rather than at its boundaries.
+
+However, we do not get a {\em single} such transition for free.
+We have arrived at the crux of the proof:
+to show that there cannot be multiple data-race PPs which must both be enabled before either data race can be identified.
+Such a circular dependency seems intuitively impossible, but to actually find the ``first reachable'' PP is not straightforward.
+We require that in $\mathcal{S}$, there exists a single transition $\tgk$ that can alone be split into $\{\tgka,\tgkb\}$,
+and some other communicating transition $\tdl$ which conflicts with $\tgkb$.
+%i.e., $\{\tgka, \tdl', \tgkb'\}$ exposes new program behaviour.
+
+% TODO: rename gamma->beta, delta->gamma
+We show this by contradiction\footnote{
+Regrettably, we could not devise a constructive algorithm for this part of the proof.
+We leave actually identifying the first-reachable data-race PP to future work,
+and trust that merely showing one must exist will satisfy the demands of modern verification.}:
 %
-By the soundness of DPOR \cite{dpor}, if a program behaviour is possible by interleaving threads at the boundaries of
-the given transitions, that interleaving will be tested.
-By contrapositive\footnote{
-Still constructive: again, going from $A \rightarrow B$ to $\neg B \rightarrow \neg A$.},
-% XXX: Handwave.
-This neglects the possiblity that TWO necessariy sites of interleaving are as such, and that they are dependent (both preemptions must occur before either data race can be identified).
-% TODO: Prove that there must exist one single transition with next paragraph's property, without needing to preempt on any other such transition.
-because it was not tested, a necessary site of interleaving must be in the middle of some transition, rather than at its boundaries.
+Assume that for all $\tgk \in \mathcal{S}$,
+and all possible points $\pgk'$ at which to split it into $\{\tgka,\tgkb\}$,
+and all other non-MHB transitions $\tdl$,
+$\tdl$ has no shared memory conflicts with $\tgkb$.
+%
+Let $\mathcal{S}' = \mathcal{S} \cup \pgk'$, i.e., the state space obtained by adding any such $\pgk'$ to $\mathcal{S}$'s PP-set.
+By the soundness of DPOR, because any such $\tdl$ is independent with $\tgkb$,
+\[
+	\{..., \tgka, \tgkb, \tdl\}
+	\equiv
+	\{..., \tgka, \tdl', \tgkb'\}
+\]
+Hence, $\mathcal{S} \equiv \mathcal{S}'$. % This jump to full-state-space-equivalence seems a bit big, but the middle steps are all just more equivalence, obviously.
+Then, the above assumption also applies to $\mathcal{S}'$,
+which shows that for any {\em pair} of transitions such as $\tgk$, adding two new PPs cannot expose new program behaviour.
+Hence, inductively, no set of new PPs of any size would expose new program behaviour not already exposed in $\mathcal{S}$.
+However, the instruction by $\omega$ which conflicts with $\ppinstr\pai$ was not observed in $\mathcal{S}$,
+so we have our contradiction.
 
-Let $\tgk$ be this transition, $\tgka$ and $\tgkb$ be the result of splitting at that site, and $\tdl$ be the communicating transition of the other thread.
-%
-If the interleaving $\{\tgka, \tdl', \tgkb'\}$ can produce a different program behaviour than $\{\tgka, \tgkb, \tdl\}$,
-then again by the soundness of DPOR, $\tgkb$ and $\tdl$ must contain some shared memory conflict.
-By the maximal state space assumption, all sync API PPs are enabled, so the locks held by $\gamma$ cannot change during $\tgk$,
-and if $\tdl$ can be reordered to the middle of $\tgk$, the locksets of $\gamma$ and $\delta$ cannot overlap,
-and likewise there is no sync-enforced MHB relation.
+Hence, if $\mathcal{S}$ does not expose the $\ppinstr\pai$ data race directly,
+there must exist some transitions $\tgk = \{\tgka,\tgkb\}$ and $\tdl$ such that $\tdl$ shared-memory-conflicts with $\tgkb$ and could be interleaved immediately before it.
+By the maximal state space assumption, all sync API PPs are already enabled, so the locksets of $\beta$ and $\gamma$ cannot overlap and there is no MHB relation.
+%%% more verbose version
+%so the locks held by $\gamma$ cannot change during $\tgk$,
+%and if $\tdl$ can be reordered to the middle of $\tgk$, the locksets of $\gamma$ and $\delta$ cannot overlap,
+%and likewise there is no sync-enforced MHB relation.
 Hence the memory conflict between $\tdl$ and $\tgkb$ will be identified as a data-race.
-Because $\tgk$ was split such that the first instruction of $\tgkb$ was not previously a PP, the data-race was not already enabled as a PP.
+Finally, because $\tgk$ was not previously split by a PP, the data-race was not already discovered.
 

### 2015-11-11 pldi-16/proof/convergence.tex
alpha very

@@
 			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
-			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR \cite{dpor}, $\mathcal{I}' \equiv \mathcal{I}$.
+			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR, $\mathcal{I}' \equiv \mathcal{I}$.
 		\item Otherwise, $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} = \emptyset$, and no instructions among $\ppnext{\pai}$ are a sync API boundary.
@@
 \newcommand\tbh{\ensuremath{t_{\beta{}h}}}
-\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
-\newcommand\tgk{\ensuremath{t_{\gamma{}k}}}
-\newcommand\tgka{\ensuremath{t_{\gamma{}k1}}}
-\newcommand\tgkb{\ensuremath{t_{\gamma{}k2}}}
-\newcommand\tdl{\ensuremath{t_{\delta{}l}}}
 \newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}
@@
 Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
-We proceed by inducting on the PPs according to the order of their preemptions.
+We proceed by induction on the PPs according to the order of their preemptions.
 (Note that this is not necessarily the same as the order of the racing instructions $\ppinstr\pai$, which occur in $\ppnext\pai$, not in \tai.)
@@
 
+\newcommand\tbk{\ensuremath{t_{\beta{}k}}}
+\newcommand\tbka{\ensuremath{t_{\beta{}k1}}}
+\newcommand\tbkb{\ensuremath{t_{\beta{}k2}}}
+\newcommand\tgl{\ensuremath{t_{\gamma{}l}}}
+
 {\bf Finding the next reachable data-race PP.}
@@
 Such a circular dependency seems intuitively impossible, but to actually find the ``first reachable'' PP is not straightforward.
-We require that in $\mathcal{S}$, there exists a single transition $\tgk$ that can alone be split into $\{\tgka,\tgkb\}$,
-and some other communicating transition $\tdl$ which conflicts with $\tgkb$.
-%i.e., $\{\tgka, \tdl', \tgkb'\}$ exposes new program behaviour.
+We require that in $\mathcal{S}$, there exists a single transition $\tbk$ that can alone be split into $\{\tbka,\tbkb\}$,
+and some other communicating transition $\tgl$ which conflicts with $\tbkb$.
+%i.e., $\{\tbka, \tgl', \tbkb'\}$ exposes new program behaviour.
 
-% TODO: rename gamma->beta, delta->gamma
 We show this by contradiction\footnote{
@@
 %
-Assume that for all $\tgk \in \mathcal{S}$,
-and all possible points $\pgk'$ at which to split it into $\{\tgka,\tgkb\}$,
-and all other non-MHB transitions $\tdl$,
-$\tdl$ has no shared memory conflicts with $\tgkb$.
+Assume that for all $\tbk \in \mathcal{S}$,
+and all possible points $\pbk'$ at which to split it into $\{\tbka,\tbkb\}$,
+and all other non-MHB transitions $\tgl$,
+$\tgl$ has no shared memory conflicts with $\tbkb$.
 %
-Let $\mathcal{S}' = \mathcal{S} \cup \pgk'$, i.e., the state space obtained by adding any such $\pgk'$ to $\mathcal{S}$'s PP-set.
-By the soundness of DPOR, because any such $\tdl$ is independent with $\tgkb$,
+Let $\mathcal{S}' = \mathcal{S} \cup \pbk'$, i.e., the state space obtained by adding any such $\pbk'$ to $\mathcal{S}$'s PP-set.
+By the soundness of DPOR, because any such $\tgl$ is independent with $\tbkb$,
 \[
-	\{..., \tgka, \tgkb, \tdl\}
+	\{..., \tbka, \tbkb, \tgl\}
 	\equiv
-	\{..., \tgka, \tdl', \tgkb'\}
+	\{..., \tbka, \tgl', \tbkb'\}
 \]
@@
 Then, the above assumption also applies to $\mathcal{S}'$,
-which shows that for any {\em pair} of transitions such as $\tgk$, adding two new PPs cannot expose new program behaviour.
+which shows that for any {\em pair} of transitions such as $\tbk$, adding two new PPs cannot expose new program behaviour.
 Hence, inductively, no set of new PPs of any size would expose new program behaviour not already exposed in $\mathcal{S}$.
@@
 Hence, if $\mathcal{S}$ does not expose the $\ppinstr\pai$ data race directly,
-there must exist some transitions $\tgk = \{\tgka,\tgkb\}$ and $\tdl$ such that $\tdl$ shared-memory-conflicts with $\tgkb$ and could be interleaved immediately before it.
+there must exist some transitions $\tbk = \{\tbka,\tbkb\}$ and $\tgl$ such that $\tgl$ shared-memory-conflicts with $\tbkb$ and could be interleaved immediately before it.
 By the maximal state space assumption, all sync API PPs are already enabled, so the locksets of $\beta$ and $\gamma$ cannot overlap and there is no MHB relation.
 %%% more verbose version
-%so the locks held by $\gamma$ cannot change during $\tgk$,
-%and if $\tdl$ can be reordered to the middle of $\tgk$, the locksets of $\gamma$ and $\delta$ cannot overlap,
+%so the locks held by $\beta$ cannot change during $\tbk$,
+%and if $\tgl$ can be reordered to the middle of $\tbk$, the locksets of $\beta$ and $\gamma$ cannot overlap,
 %and likewise there is no sync-enforced MHB relation.
-Hence the memory conflict between $\tdl$ and $\tgkb$ will be identified as a data-race.
-Finally, because $\tgk$ was not previously split by a PP, the data-race was not already discovered.
+Hence the memory conflict between $\tgl$ and $\tbkb$ will be identified as a data-race.
+Finally, because $\tbk$ was not previously split by a PP, the data-race was not already discovered.
 
@@
 (Alternatively, for any program with a finitely-sized instruction listing, the number of unique instruction pairs is finite.)
-This finishes our inner induction.
-Hence $\pai$ is reachable, finishing our outer induction.
-Hence all PPs in $\mathcal{I}$ are reachable.
+Hence by the ``next reachable data-race'' induction, $\pai$ is reachable.
+Hence by the ``$\pai$ is reachable'' induction, all PPs in $\mathcal{I}$ are reachable.
 \end{proof}
@@
 		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
-Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
+Then DPOR will find the buggy interleaving within this state space.
 \end{proof}

### 2015-11-11 pldi-16/proof/convergence.tex
minor rephrase

@@
 Regrettably, we could not devise a constructive algorithm for this part of the proof.
-We leave actually identifying the first-reachable data-race PP to future work,
+We leave constructively identifying the first-reachable data-race PP to future work,
 and trust that merely showing one must exist will satisfy the demands of modern verification.}:
@@
 		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
+		Because we are concerned with verifying the absence of bugs, this also suffices for convergence, although we have left it implicit for brevity.}.
 Then DPOR will find the buggy interleaving within this state space.

### 2015-11-11 pldi-16/di.tex
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

@@
 %% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.
-
-\subsection{Classifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?
-
-Iterative Deepening must be seeded with some initial state spaces,
-which can be any number of subsets of the statically-available PPs that SSS-MC would use.
-While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
-Because data races indicate access pairs that can interleave at instruction granularity,
-it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
-
-With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread.
-That state space may expose a failure, in which case we report a data-race bug,
-or complete successfully, which indicates a benign or false-positive data race.
-It may also uncover a new data-race candidate in some alternate interleaving,
-in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
-Because Iterative Deepening is constrained by a CPU budget,
-we may time out before completing a data race's associated state space,
-in which case it remains a potential false positive that the user must handle.
 
@@
 (it lets us avoid thrashing in \quicksand).
+
+\subsection{Classifying data races as buggy or benign}
+
+Iterative Deepening must be seeded with some initial state spaces,
+which can be any number of subsets of the statically-available PPs that SSS-MC would use.
+While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
+Because data races indicate access pairs that can interleave at instruction granularity,
+it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
+
+With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread.
+That state space may expose a failure, in which case we report a data-race bug,
+or complete successfully, which indicates a benign or false-positive data race.
+It may also uncover a new data-race candidate in some alternate interleaving,
+in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
+Because Iterative Deepening is constrained by a CPU budget,
+we may time out before completing a data race's associated state space,
+in which case it remains a potential false positive that the user must handle.
+
+\subsection{Convergence to total verification}
+\label{sec:totalverif}
+
+Although Iterative Deepening's main purpose is to heuristically choose the most effective PP subsets to test
+when the maximal state space is too large,
+some tests may be small enough that their maximal state spaces can also be completed in time.
+For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
+In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
+%In other words, contrapositively,
+%if it is possible to find a bug with any sequence of preemptions on any instruction whatsoever,
+%an equivalent thread interleaving will be reachable using only data-race PPs and synchronization API PPs.
+A proof sketch of the contrapositive statement follows.
+
+\begin{theorem}
+If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
+Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
+\end{theorem}
+\begin{proof}
+The proof has two parts:
+first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
+second, we show that Iterative Deepening will eventually detect all such data races.
+
+\begin{lemma}[Equivalence of irrelevant PPs]
+For any thread interleaving possible by preempting on any instruction,
+there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
+	\label{lem:relevant}
+\end{lemma}
+
+asdfadsf % TODO
+
+\begin{definition}[Reachable data race]
+A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
+\end{definition}
+Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
+\begin{lemma}[Saturation of data-race PPs]
+	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
+	\label{lem:saturation}
+\end{lemma}
+
+asdfasdfas % TODO
+
+To conclude,
+for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
+and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
+Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
+\end{proof}
+
+Our full proof
+{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
+discusses our assumptions explicitly and includes more formal definitions and structure.
 
@@
 Our full proof
-{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
-discusses our assumptions explicitly and provides the similar cases in more detail.
+{\em [see supplementary material]}
+provides the similar cases in more detail.
 \renewcommand\proofname{Proof}

### 2015-11-11 pldi-16/evaluation.tex
add structural text
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

@@
 \section{Evaluation}
+\label{sec:eval}
 
@@
 
+% TODO: You should introduce this in the narrative as a key contribution way earlier.
 \subsection{Finding nondeterministic data-race candidates}
@@
 on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
+
+In some cases, this repeated work can cause Iterative Deepening to be slower than SSS-MC to find certain bugs,
+for example, if both {\tt mutex\_lock} and {\tt mutex\_lock} preemptions together would expose a bug, but not either alone.
+Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
+if the state space resulting from preempting on every instruction could be completed,
+SSS-MC will achieve that verification faster, as Iterative Deepening will test many subsets of data-race PPs first.
 Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,

### 2015-11-11 pldi-16/intro.tex
Add narrative text about convergence; add proof sketch skeleton (lemmas unproved) for convergence

@@
 It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.
+%In fact, if given enough time to fully test all discovered data-race preemption points,
+%Iterative Deepening provides a full verification of all thread schedules that could arise from preempting anywhere.
+In fact, Iterative Deepening is fully general:
+we have proved that if it completes all state spaces resulting from data-race preemption points,
+that serves as a total verification of all possible thread interleavings of the given test program.
 
@@
 	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
+	\item A proof of convergence, which shows that fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules, should that be possible in the given CPU budget;
 	\item A new tactic for eliminating one class of false-positive data races,
@@
 The remainder of the paper is organized as follows.
-\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening,
-\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it,
+\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening, including our proof of convergence,
+\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it, including our new false-positive data-race tactic,
 \sect{\ref{sec:eval}} presents our evaluation,

### 2015-11-11 pldi-16/paper.tex
add structural text
somecleanup

@@
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
-We dramatically reduce false-positive data race reports,
-%by verifying the absence of bugs in their associated state spaces
-and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
+%We dramatically reduce false-positive data race reports,
+%%by verifying the absence of bugs in their associated state spaces
+%and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
 % TODO: Not true; Replace instead with a claim about partial verification.
@@
 and find new bugs with data-race PPs that the maximal state space would not expose at all.
+Moreover, for tests where all state spaces with data-race PPs can be completed, we provide an equally strong verification as if every single instruction had been used as a PP.
 
@@
 % TODO: Make data set, at least, public.
-TODO: Data set at least can be public.
 
@@
 
-%Many thanks to David A. Eckhardt, Vaishaal Shankar, and Haryadi Gunawi for generously providing student implementations from CMU's, Berkeley's, and U. of Chicago's OS classes respectively.
 Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
@@
 the anonymous PLDI reviewers for their helpful comments.
-This work was supported in part by % TODO: funding sources
+This work was supported in part by
 the U.S. Army Research Office under grant number W911NF0910273.

### 2015-11-11 pldi-16/related.tex
some rephrasing and start to make a big table
add structural text
somecleanup

@@
 \section{Related Work}
+\label{sec:related}
 
@@
 %Too many related projects to list have made contributions to the
-Many advances have been made on the false-positive data race problem first introduced in \cite{eraser}.
+Many advances have been made on the false-positive data race problem since it was first introduced in \cite{eraser}.
 \cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
@@
 Portend \cite{portend} is the most closely related work we have found.
-% FIXME: "limited"?
-Based on single-pass data race reports, it explores a limited state space to classify candidates in a taxonomy of likely severity.
-Compared to us, Portend additionally finds non-failing races which nevertheless cause
-%suspiciously
-different program output, while we depend on directly detecting failures.
+Based on single-pass data race reports, it tests alternate program executions to classify candidates in a taxonomy of likely severity.
 It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
 while we explore the latter only.
+Portend additionally reports non-failing races which nevertheless cause
+%suspiciously
+different program output, which is orthogonal to our technique.
 However, Portend does not test alternate interleavings {\em in advance} of knowing data races,
-which is necessary to expose some bugs (\sect{\ref{sec:eval-falseneg}}).
+which is necessary to expose certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
 % XXX: Is this true??
@@
 % blah blah pldi'15 symbiosis DSP
-
-% TODO: talk about data race detectors???
-% eg Scalable Race Detection for Android Applications -- uses domain specific heuristics to filter out false positives
 

### 2015-11-12 pldi-16/di.tex
somecleanup
add the brief algorithm 2

@@
 	}
-	// $j_0$'s ETA was bad, but nothing else was better. \\
-	return $j_0$
+	return $j_0$ // ETA was bad, but no other job was better.
 	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
@@
 
-With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread.
-That state space may expose a failure, in which case we report a data-race bug,
+\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	\Input{$j_0$, the currently-running job}
+	\Input{\AllJobs, the set of all existing jobs}
+	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
+	\If{$\forall j \in \AllJobs,$
+	\GetPPSet{$j_0$} $\cup$ $\alpha$
+	$\not\subseteq$
+	\GetPPSet{$j$}
+	}{
+		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$))
+	}
+	\caption{Adding new jobs with data-race PPs.}
+	\label{alg:handledatarace}
+\end{algorithm}
+
+With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
+We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
+The new space may expose a failure, in which case we report a data-race bug,
 or complete successfully, which indicates a benign or false-positive data race.
@@
 when the maximal state space is too large,
-some tests may be small enough that their maximal state spaces can also be completed in time.
+some tests may be small enough that even their maximal state spaces could be completed in time.
 For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
@@
 
-\begin{lemma}[Equivalence of irrelevant PPs]
+\begin{lemma}[Equivalence of non-data-race PPs]
 For any thread interleaving possible by preempting on any instruction,
@@
 
-asdfadsf % TODO
+TODO: sketch the proof of this lemma % TODO
 
@@
 \end{lemma}
+% TODO: Put a figure here giving an example of where a data race is only transitively reachable, not initially reachable.
 
-asdfasdfas % TODO
+TODO: sketch the proof of this lemma % TODO
+\\
 

### 2015-11-12 pldi-16/evaluation.tex
some rephrasing

@@
 
+% TODO: Make a CDF graph of how many tests got total verification here.
+
+% TODO: Figure out how to visualize how many unique data race PPs were verified, and how many remained as unclassified false positives
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 thereby making them {\em false negatives} of the single-pass approach.
-% TODO: Put a figure here giving an example of where e.g. a data race only shows up during the contention path of a mutex.
 We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).
@@
 
-				% TODO: Argue:
-				% It is fair to compare multiple pass DR-analysis under Landslide against just a single execution because prior work DR detectors, being not integrated with a MC, are not intended to uncover different results under subsequent runs.
-				% Define a "stress tester" as a class of bug detectors where they are intended to [...]
-				% Maybe say: Do there exist any stress-tester DR detectors where they are intended to produce new results under reruns?
-
 TODO: Present results.
 
-% TODO: restructure this paragraph to account for RaceFuzzer/Portend. Be like, "Yes they explore multiple schedules, but only AFTER finding a race." And/or make this argument in prior work
-One might also wonder: Why is it fair to compare the data race bugs \quicksand~finds (10 CPU-hours)
-against the candidates found by a prior work's single execution (less than a minute)?
-We argue this comparison is meaningful because prior work data-race tools, being not integrated with a MC,
-are not intended to discover new data race candidates under subsequent runs.
+At first, it might seem unfair to compare the data race bugs \quicksand~finds (10 CPU-hours)
+against the candidates found by a prior work's single execution (less than a minute).
+We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
+are not intended to discover new data race candidates under subsequent runs. %before discovering any initial candidates.
 Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
@@
 \cite{chess-icb,gambit}.
+%Prior work data-race tools \cite{portend} could be configured to have the same sort of feedback loop as Iterative Deepening,
+%i.e., discovering transitively-reachable data races while testing initial ones,
+%although that would still not simultaneously be able to provide total verifications.
 % TODO: read gambit paper
-%One could run a data-race tool repeatedly for 10 CPU-hours, but the advantage of stateless MC over stress testing is already well-understood.
-% TODO TODO TODO get a big citation for teh above sentence.
 
@@
 
-% TODO Rename to "Limitations and Future Work"
 \section{Discussion}
@@
 if the state space resulting from preempting on every instruction could be completed,
-SSS-MC will achieve that verification faster, as Iterative Deepening will test many subsets of data-race PPs first.
+an SSS-MC tool such as \cite{spin} might achieve that verification faster,
+as Iterative Deepening will test many subsets of data-race PPs first.
 Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,

### 2015-11-12 pldi-16/intro.tex
some rephrasing

@@
 	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
-	\item A proof of convergence, which shows that fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules, should that be possible in the given CPU budget;
+	\item A proof of convergence, which shows that should it be possible in the given CPU budget,
+		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
 	\item A new tactic for eliminating one class of false-positive data races,

### 2015-11-13 pldi-16/evaluation.tex
some more polish; clean up stale todos; widen the table

@@
 \newcommand\paradise{\texttt{sem\_test}}
-\newcommand\rwl{\texttt{rwl\_test}}
+\newcommand\rwldgr{\texttt{rwlock\_test}}
 We tested P2s with 6 multithreaded programs:
@@
 \tej, a test of thread lifecycle,
-\bct~and \paraguay~for condvars,
+\bct~and \paraguay~for cvars,
 \paradise~for semaphores,
-and \rwl~for r/w locks.
+and \rwldgr~for r/w locks.
 For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
@@
 \newcommand\waitsimple{\texttt{wait\_test}}
+\newcommand\alarmsimul{\texttt{alarm\_test}}
 % TODO: Add more test cases
-We tested the Pintoses with 2 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm, and \waitsimple, a test of process lifecycle system calls.
+We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
+\alarmsimul, for the timer sleep routine,
+and \waitsimple, for process lifecycle system calls\footnote{
+	Some of the Pintoses were partially implemented, so each test could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
+}
+The source code of all 9 tests is available at
+{\em [submitted as supplementary material]}.
 For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
 Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
-% XXX: Some pintoses can't run all the tests. So this number is too high.
-In total, the test suite comprises 632 unique state spaces, at least 106 of which contain bugs.
+In total, the test suite comprises 629 unique state spaces,
+% TODO: Add number of pintos bugs.
+at least 106 of which contain bugs.
 
@@
 		%\small
-	\begin{tabular}{r||c|c|c||c|c|c|c}
-		& {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
-		{\bf Test name} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf recycle FPs} & {\bf DR PPs} & {\bf DR bugs} \\
+	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
+		% TODO: Make the "bugs" left half of this graph into a bar graph.
+		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Untested} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
+		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf DR PPs} & {\bf recycle DRs} & {\bf DR PPs} & {\bf DR bugs} \\
 		\hline
-		{\bct		} & 7	& 11	& 4	&	& 65	&	& 4	\\
-		{\tej		} & 14 	& 25	& 11	&	& 471	&	& 6	\\
-		{\mxtest	} & 1	& 10	& 9	&	& 7	&	& 1	\\
-		{\paradise	} & 10	& 17	& 7	&	& 268	&	& 3	\\
-		{\paraguay	} & 4	& 12	& 9	&	& 188	&	& 0	\\
-		{\rwl		} & 26	& 30	& 3	&	& 157	&	& 3	\\
+		{\bct		} & 79	& 7	& 11	& 4	&	&	& 65	&	& 4	\\
+		{\tej		} & 79	& 14 	& 25	& 11	&	&	& 471	&	& 6	\\
+		{\mxtest	} & 79	& 1	& 10	& 9	&	&	& 7	&	& 1	\\
+		{\paradise	} & 79	& 10	& 17	& 7	&	&	& 268	&	& 3	\\
+		{\paraguay	} & 79	& 4	& 12	& 9	&	&	& 188	&	& 0	\\
+		{\rwldgr	} & 79	& 26	& 30	& 3	&	&	& 157	&	& 3	\\
 		\hline
-		{\prisema	} &	&	&	&	&	&	&	\\
-		{\waitsimple	} &	&	&	&	&	&	&	\\
+		{\prisema	} & 59	&	&	&	&	&	&	&	&	\\
+		{\waitsimple	} & 52	&	&	&	&	&	&	&	&	\\
+		{\alarmsimul	} & 44	&	&	&	&	&	&	&	&	\\
 		\hline
-		{\bf Total}	&	&	&	&	&	&	&
+		{\bf Total}	& 629	&	&	&	&	&	&	&	&
 	\end{tabular}
 	\end{center}
-	\caption{Breakdown of number and types of bugs found by test case.
+	\caption{Summary of number and types of bugs found by each test case.
 		SSS-MC is the control; QS is \quicksand.
-		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose.
+		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}}).
 	``Nondeterministic DRs'' counts among the DR-only bugs how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
@@
 {\bf Partial verification.}
-% TODO: But what about state space estimation? Does that count as probability? I don't think it does. But can you mention it?
 When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
@@
 
+TODO
+
+{\bf Full verification.}
 TODO
@@
 \cite{chess-icb,gambit}.
-%Prior work data-race tools \cite{portend} could be configured to have the same sort of feedback loop as Iterative Deepening,
-%i.e., discovering transitively-reachable data races while testing initial ones,
+This experiment also suggests that
+other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use the same sort of feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
+%i.e., discovering transitively-reachable data races while testing initial ones.
 %although that would still not simultaneously be able to provide total verifications.
@@
 
-In some cases, this repeated work can cause Iterative Deepening to be slower than SSS-MC to find certain bugs,
-for example, if both {\tt mutex\_lock} and {\tt mutex\_lock} preemptions together would expose a bug, but not either alone.
+This repeated work can cause Iterative Deepening to be slower than SSS-MC to find certain bugs,
+for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} preemptions together would expose a bug, but not either alone.
 Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
@@
 However, in 2 tests, no more than the minimal state space could be verified,
-and in 2 other tests, not even that much.
+and in 3 other tests, not even that much.
 Larger state spaces often result from finer-grained locking,

### 2015-11-13 pldi-16/testcases/testcases.tex
add test cases tech report / suppl material

%
% LaTeX template for prepartion of submissions to PLDI'15
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'15 web site.
%
\documentclass[pldi]{../sigplanconf-pldi15}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{amsthm}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphicx}
%\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
%\doublespacing

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

\begin{document}

%
% any author declaration will be ignored  when using 'plid' option (for double blind review)
%

%\newcommand\landslide{\textsc{Landslide}} % Don't mention landslide in the proof.
\newcommand\quicksand{\textsc{Quicksand}}
\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1}
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}
\definecolor{commentblue}{RGB}{0,0,255}

%\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
\title{Source Code of Test Cases for Iterative Deepening}
\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}

\maketitle
\begin{abstract}
In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
This document presents the source code of the 9 test cases we used, as supplemental material to our main paper.
\end{abstract}

\section{``P2'' thread library tests}

The following tests were used to test the user-space POSIX-like thread libraries implemented by students in CMU's undergrad OS class, 15-410.
They use the thread library API defined in \cite{thrlib} and the kernel system-call API defined in \cite{kspec}.

\begin{figure*}
\begin{verbatim}
#include <syscall.h>
#include <stdlib.h>
#include <thread.h>
#include <mutex.h>
#include <cond.h>
#include "410_tests.h"
#include <report.h>
#include <test.h>

#define STACK_SIZE 4096
#define ERR REPORT_FAILOUT_ON_ERR

mutex_t lock;
cond_t cvar;
static int ready = 0;

void *waiter(void *dummy)
{
        mutex_lock(&lock);
        while (ready == 0) {
                cond_wait(&cvar, &lock);
        }
        mutex_unlock(&lock);
        return NULL;
}

int main(void)
{
        report_start(START_CMPLT);

        ERR(thr_init(STACK_SIZE));
        ERR(mutex_init(&lock));
        ERR(cond_init(&cvar));

        ERR(thr_create(waiter, NULL));

        mutex_lock(&lock);
        ready = 1;
        cond_broadcast(&cvar);
        mutex_unlock(&lock);

        return 0;
}
\end{verbatim}
	\caption{{\tt broadcast\_test}, a test of {\tt cond\_signal}/{\tt cond\_broadcast} interaction.}
	\label{fig:bct}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <syscall.h>
#include <stdlib.h>
#include <thread.h>
#include <mutex.h>
#include <cond.h>
#include <assert.h>
#include "410_tests.h"
#include <report.h>
#include <test.h>

#define STACK_SIZE 4096
#define ERR REPORT_FAILOUT_ON_ERR

mutex_t lock;
static int num_in_section = 0;

void critical_section()
{
        mutex_lock(&lock);
        num_in_section++;
        yield(-1);
        assert(num_in_section == 1);
        num_in_section--;
        mutex_unlock(&lock);
}

void *contender(void *dummy)
{
        ERR(swexn(NULL, NULL, NULL, NULL));
        critical_section();
        // Do it again, in case lock is correct but unlock corrupts something.
        critical_section();
        vanish();
        return NULL;
}

int main(void)
{
        ERR(thr_init(STACK_SIZE));
        ERR(mutex_init(&lock));
        ERR(thr_create(contender, NULL));
        ERR(swexn(NULL, NULL, NULL, NULL));
        critical_section();
        vanish(); // bypass thr_exit
        return 0;
}
\end{verbatim}
	\caption{{\tt mutex\_test}, a test of {\tt mutex\_lock}/{\tt mutex\_unlock} correctness.
	Used with the special mutex-testing mode of data-race detection, described in the main paper (Section 3.1).}
	\label{fig:mx}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <syscall.h>
#include <stdlib.h>
#include <thread.h>
#include <mutex.h>
#include <cond.h>
#include <sem.h>
#include <assert.h>
#include "410_tests.h"
#include <report.h>
#include <test.h>

#define STACK_SIZE 4096
#define ERR REPORT_FAILOUT_ON_ERR

mutex_t little_lock; /* used to avoid spurious data race reports */
sem_t lock;
int num_in_section = 0;

// note: the mutex accesses in this function are ignored as PPs
// by use of without_function; they are here to prevent
// num_in_section acesses from being identified as data races.
void __attribute__((noinline)) critical_section() {
        mutex_lock(&little_lock);
        num_in_section++;
        mutex_unlock(&little_lock);
        yield(-1);
        mutex_lock(&little_lock);
        assert(num_in_section == 1);
        num_in_section--;
        mutex_unlock(&little_lock);
}
void *consumer(void *dummy) {
        ERR(swexn(NULL, NULL, NULL, NULL));
        sem_wait(&lock);
        critical_section();
        sem_signal(&lock);
        vanish();
        return NULL;
}
void producer() {
        ERR(mutex_init(&little_lock));
        ERR(sem_init(&lock, 0));
        ERR(thr_create(consumer, NULL));
        ERR(thr_create(consumer, NULL));
        sem_signal(&lock);
}
int main(void) {
        ERR(thr_init(STACK_SIZE));
        ERR(swexn(NULL, NULL, NULL, NULL));
        producer();
        vanish(); // bypass thr_exit
        return 0;
}
\end{verbatim}
	\caption{{\tt sem\_test},
	%known to 15-410 students as {\tt paradise\_lost}, % TODO CAMREADY
	a test of {\tt sem\_wait}/{\tt sem\_signal} interaction.}
	\label{fig:paradise}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <syscall.h>
#include <stdlib.h>
#include <thread.h>
#include <mutex.h>
#include <cond.h>
#include "410_tests.h"
#include <report.h>
#include <test.h>

#define STACK_SIZE 4096
#define ERR REPORT_FAILOUT_ON_ERR

mutex_t lock1, lock2;
cond_t cvar1, cvar2;
int slept1 = 0; /* Whether thread1 has gotten to sleep on cvar1 */
int signaled1 = 0; /* Set right before main thread signals cvar1 */
int slept2 = 0; /* Whether thread1 has gotten to sleep on cvar2 */
int signaled2 = 0; /* Set right before main thread signals cvar2 */

void *thread1(void *dummy) {
        /* go to sleep on cvar1 */
        mutex_lock(&lock1);
        slept1 = 1;
        cond_wait(&cvar1, &lock1);
        if (!signaled1) {
                report_end(END_FAIL);
        }   
        mutex_unlock(&lock1);

        /* go to sleep on cvar1 */
        mutex_lock(&lock2);
        slept2 = 1;
        cond_wait(&cvar2, &lock2);
        if (!signaled2) {
                report_end(END_FAIL);
        }   
        mutex_unlock(&lock2);
        return NULL;
}
\end{verbatim}
	\caption{{\tt signal\_test},
	%known to 15-410 students as {\tt paraguay}, % TODO CAMREADY
	a test for a deep bug in {\tt cond\_signal} (Part 1; see part 2 in Figure~\ref{fig:paraguay-pt2}).}
	\label{fig:paraguay}
\end{figure*}
\begin{figure*}
\begin{verbatim}
int main(void) {
        ERR(thr_init(STACK_SIZE));
        ERR(mutex_init(&lock1));
        ERR(mutex_init(&lock2));
        ERR(cond_init(&cvar1));
        ERR(cond_init(&cvar2));
        ERR(thr_create(thread1, NULL));

        /* Wait for thread1 to get to sleep on cvar1. */
        mutex_lock(&lock1);
        while (!slept1) {
                mutex_unlock(&lock1);
                thr_yield(-1);
                mutex_lock(&lock1);
        }
        /* Indicate that we are about to signal */
        signaled1 = 1;
        mutex_unlock(&lock1);

        /* Signal. Note that we know for sure that thread1 is asleep on
         * cvar1 (assuming correct cond vars and mutexes...) */
        cond_signal(&cvar1);

        /* Now do it all again for the second set of things. */
        /* Wait for thread1 to get to sleep on cvar2. */
        mutex_lock(&lock2);
        while (!slept2) {
                mutex_unlock(&lock2);
                thr_yield(-1);
                mutex_lock(&lock2);
        }
        /* Indicate that we are about to signal */
        signaled2 = 1;
        mutex_unlock(&lock2);

        /* Signal. Note that we know for sure that thread1 is asleep on
         * cvar1 (assuming correct cond vars and mutexes...) */
        cond_signal(&cvar2);

        return 0;
}
\end{verbatim}
	\caption{{\tt signal\_test},
	%known to 15-410 students as {\tt paraguay}, % TODO CAMREADY
	a test for a deep bug in {\tt cond\_signal}
	(Part 2; see part 1 in Figure~\ref{fig:paraguay}).}
	\label{fig:paraguay-pt2}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <thread.h>
#include <mutex.h>
#include <cond.h>
#include <rwlock.h>
#include <syscall.h>
#include <stdlib.h>
#include <stdio.h>
#include "410_tests.h"
#include <test.h>
#define STACK_SIZE 4096

int read_count = 0;
mutex_t read_count_lock;
rwlock_t lock;
int pass = -1;

void g() {
        mutex_lock(&read_count_lock);
        read_count++;
        if (read_count == 2) {
                pass = 0;
        }   
        mutex_unlock(&read_count_lock);
}
void *f(void *arg) {
        rwlock_lock(&lock, RWLOCK_READ);
        g();
        return((void *)0);
}
int main() {
        int tid1;
        thr_init(STACK_SIZE);

        REPORT_ON_ERR(rwlock_init(&lock));
        REPORT_ON_ERR(mutex_init(&read_count_lock));
        rwlock_lock(&lock, RWLOCK_WRITE);

        if ((tid1 = thr_create(f, NULL)) < 0) {
                REPORT_END_FAIL;
                return -1; 
        }   
        rwlock_downgrade(&lock);
        g();
        if (thr_join(tid1, NULL) < 0) {
                REPORT_END_FAIL;
                return -1;
        }
        if (pass != 0)
                REPORT_END_FAIL;
        return pass;
}
\end{verbatim}
	\caption{{\tt rwlock\_test},
	%known to 15-410 students as {\tt rwlock\_downgrade\_read\_test}, % TODO CAMREADY
	a test of all {\tt rwlock} functions.}
	\label{fig:rwldgr}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <stdio.h>
#include <thread.h>
#include <syscall.h> /* for PAGE_SIZE */

void *waiter(void *p) {
  int status;
  thr_join((int)p, (void **)&status);
  thr_exit((void *) 0); 
}

int main() {
  thr_init(16 * PAGE_SIZE);
  (void) thr_create(waiter, (void *) thr_getid());
  thr_exit((void *)'!');
}
\end{verbatim}
	\caption{{\tt join\_test},
	% known to 15-410 students as {\tt thr\_exit\_join}, % TODO CAMREADY
	a test of {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join}.}
	\label{fig:tej}
\end{figure*}

\begin{enumerate}
	\item {\tt broadcast\_test}: Figure \ref{fig:bct}
	\item {\tt mutex\_test}: Figure \ref{fig:mx}
	\item {\tt sem\_test}: Figure \ref{fig:paradise}
	\item {\tt signal\_test}: Figures \ref{fig:paraguay} and \ref{fig:paraguay-pt2}
	\item {\tt rwlock\_test}: Figure \ref{fig:rwldgr}
	\item {\tt join\_test}: Figure \ref{fig:tej}
\end{enumerate}

\section{``Pintos'' kernel tests}

The following tests were used to test the first two Pintos kernel projects, ``threads'' and ``userprog'',
implemented by students in Berkeley's and the University of Chicago's undergrad OS classes,
CS162 and CS230 respectively \cite{pintos}.
They use the internal kernel API provided by the basecode, available online at \cite{pintos-github}.

\begin{figure*}
\begin{verbatim}
#include <stdio.h>
#include "tests/threads/tests.h"
#include "threads/init.h"
#include "threads/malloc.h"
#include "threads/synch.h"
#include "threads/thread.h"
#include "devices/timer.h"

static struct semaphore sema;

static void priority_sema_thread (void *aux UNUSED) 
{
  sema_down (&sema);
}

void test_priority_sema (void) 
{
  int i;

  sema_init (&sema, 0); 
  thread_set_priority (PRI_DEFAULT);
  for (i = 0; i < 2; i++)
    {   
      int priority = PRI_DEFAULT;
      char name[16];
      snprintf (name, sizeof name, "priority %d", priority);
      thread_create (name, priority, priority_sema_thread, NULL);
    }

  for (i = 0; i < 2; i++)
    {   
      sema_up (&sema);
    }   
}
\end{verbatim}
	\caption{{\tt sched\_test},
	% known to CS162 and CS230 students as {\tt priority\_sema}, % TODO CAMREADY
	a test of {\tt thread\_create}, {\tt sema\_up}, and {\tt sema\_down}.}
	\label{fig:prisema}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <stdio.h>
#include "tests/threads/tests.h"
#include "threads/init.h"
#include "threads/synch.h"
#include "threads/thread.h"
#include "devices/timer.h"

struct semaphore done_sema;

// semaphore PPs here are ignored using without_function
static __attribute__((noinline)) void child_done() {
        sema_up(&done_sema);
}
// semaphore PPs here are ignored using without_function
static __attribute__((noinline)) void parent_done() {
        sema_down(&done_sema);
        sema_down(&done_sema);
}

static void sleeper (void *hax) {
        timer_sleep ((int)hax);
        child_done();
        thread_yield();
}
void test_alarm_simultaneous() {
        sema_init(&done_sema, 0); 
        thread_create ("Patrice Godefroid", PRI_DEFAULT, sleeper, (void *)3);
        thread_create ("Cormac Flanagan", PRI_DEFAULT, sleeper, (void *)4);
        thread_yield();
        timer_sleep (2);
        parent_done();
        thread_yield();
}
\end{verbatim}
	\caption{{\tt alarm\_test},
	% known to CS162 and CS230 students as {\tt alarm\_simultaneous}, % TODO CAMREADY
	a test of {\tt timer\_sleep}.}
	\label{fig:alarm}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <syscall.h>
#include "tests/lib.h"
#include "tests/main.h"

void test_main (void) 
{
  msg ("wait(exec()) = %d", wait (exec ("child-simple")));
}
\end{verbatim}
	\caption{{\tt wait\_test},
	% known to CS162 and CS230 students as {\tt wait\_simple}, % TODO CAMREADY
	a test of {\tt exec} and of {\tt wait}/{\tt exit} interaction.}
	\label{fig:waitsimple}
\end{figure*}
\begin{figure*}
\begin{verbatim}
#include <stdio.h>
#include "tests/lib.h"

int main (void) 
{
  msg ("run");
  return 81; 
}
\end{verbatim}
	\caption{{\tt child\_simple}, a helper program used by {\tt wait\_test} (Figure~\ref{fig:waitsimple}).}
	\label{fig:childsimple}
\end{figure*}

\begin{enumerate}
	\item {\tt sched\_test}: Figure \ref{fig:prisema}
	\item {\tt alarm\_test}: Figure \ref{fig:alarm}
	\item {\tt wait\_test}: Figures \ref{fig:waitsimple} and \ref{fig:childsimple}
\end{enumerate}

\bibliographystyle{abbrvnat}
\bibliography{../citations}{}
\end{document}

### 2015-11-13 pldi-16/di.tex
implement joshua's revisions pt1

@@
 	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()} or {\tt cond\_wait()}.
+		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
 		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
@@
 Iterative Deepening searches among different combinations of such points during a test,
-deciding on-the-fly whether to pursue each resulting state space or abandon it in favour of smaller ones.
+deciding on-the-fly whether to pursue each resulting state space or, or to defer it in favour of smaller ones.
 Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
@@
 {\em model checking} (MC),
-{\em single-state-space model checking} (SSS-MC),% (i.e., the approach of prior work),
+{\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
 Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
@@
 
-Please note the difference between data-race {\em candidates} and data-race {\em bugs}.
+We distinguish between data-race {\em candidates} and data-race {\em bugs}.
 Because data-race analysis is prone to false-positives,
@@
 Should a future interleaving, preempting during those accesses,
+% TODO: joshua wants 'failure' to be explicitly defined
 lead to a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
@@
 \newcommand\GetPPSet[1]{PPSet(#1)}
-With a limited CPU budget, we must avoid running tests that are likely to time out.
+With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
 Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
@@
 \subsection{Classifying data races as buggy or benign}
+\label{sec:classifying}
 
@@
 With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
-We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
-The new space may expose a failure, in which case we report a data-race bug,
+Note that even though a data race may involve two different instructions, we add new state spaces with only one new PP at a time.
+Rather than adding a single state space configured to preempt on both involved instructions,
+i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
+we prefer to add multiple smaller jobs which have a higher chance of completing in time. i.e.,
+$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
+$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
+If $A$ and $B$ bug-free, they will in turn add $AB$ later.
+%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
+The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
+for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.
+
+The new spaces may expose a failure, in which case we report a data-race bug,
 or complete successfully, which indicates a benign or false-positive data race.
-It may also uncover a new data-race candidate in some alternate interleaving,
+They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
 in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
-Because Iterative Deepening is constrained by a CPU budget,
+%Because Iterative Deepening is
+Being constrained by a CPU budget,
 we may time out before completing a data race's associated state space,
@@
 We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
-\landslide~implements DPOR,
+\landslide~implements DPOR \cite{dpor},
 %\cite{dpor},
-state space estimation, and a hybrid lockset/happens-before data-race analysis.
+state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
 % TODO: \cite{fairstatelessmc} instead or in addition?
@@
 It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
+% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
 Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
@@
 {\bf Restricting PPs with stack trace predicates.}
-Most MCs preempt indiscriminately on any sync API call, regardless of the call-site.
+Most model checkers (MCs) preempt indiscriminately on any sync API call, regardless of the call-site.
 However, when testing a particular module in a large codebase,
@@
 Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself.
-Assuming the locks are already correct enables productive data-race analysis on the rest of the codebase.
-With testing limited to one execution, even if you wished to test for lock bugs, data-race analysis would still uselessly flag every single access pair in the lock implementation, requiring human attention to verify.
+The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
+Otherwise, with testing limited to one execution,
+%even if one wishes to test for lock bugs,
+data-race analysis would flag every access pair in the lock implementation, requiring human attention to verify.
 However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
@@
 {\bf Initial PPs.}
-For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs shown in Figure~\ref{fig:id}:
+For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs that we showed in Figure~\ref{fig:id}:
 $\{yield\}$,
@@
 For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so we also preempt just before {\tt cli} and just after {\tt sti} (to appropriate the names of the x86 instructions)\footnote{
-During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock,
-as {\tt cli}'d memory accesses can still race with others that have interrupts on.
+so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
+%(to appropriate the names of the x86 instructions)
+During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
+%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
 Some kernels disable preemption without disabling interrupts,
-which can be modelled the same way using manual annotations of that API.}.
+which can be modelled the same way using manual annotations. %of that API.
+This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
 \quicksand~is configured to begin with the subsets
@@
 passed as an argument to \landslide.
-Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing.%on a FIFO pipe.
+Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
 \landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
@@
 \quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
+We implement suspending jobs simply by making \landslide~wait on a message-passing call.
+Should \quicksand~later resume a suspended job, we send a message to continue,
+causing the \landslide~instance to resume exploring where it left off;
+otherwise, we send a message only after time runs out, causing it to exit.
 
@@
 for which we choose 32,
-so their ETAs have some time to stabilize.
+so that their ETAs have some time to stabilize.
 
 We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether the MC observed the access pair ordered one or both ways while testing the original state space,
-We prioritize jobs with both-order data-race PPs,
-because single-order candidates are much more likely to be false-positives
+based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
+Jobs with both-order data-race PPs are prioritized higher,
+because single-order candidates are more likely to be false-positives
 (though before preempting during the access itself, we cannot say for sure, hence the heuristic).
 %so we prioritize jobs with both-order data-race PPs.
-For single-order races, we do not add PPs for the later access at all;
-should they be needed, preempting on the first access will suffice to upgrade it to both-order.
+For single-order races, we do not add a PP for the later access at all;
+should it be needed, preempting on the first access will suffice to upgrade the race to both-order.
 
@@
 %
-Hence each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
-Rather than adding a maximal job with both new PPs at once, we prefer to add smaller jobs which have a higher chance of completing in time.
-If those state spaces are bug-free, they will in turn add the maximal job later.
+Hence, together with the logic in \sect{\ref{sec:classifying}}, each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
 %
@@
 In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to decide at runtime which jobs to prioritize.
+which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
 % TODO: Put numbers here.

### 2015-11-13 pldi-16/intro.tex
implement joshua's revisions pt1

@@
 In fact, Iterative Deepening is fully general:
-we have proved that if it completes all state spaces resulting from data-race preemption points,
+we prove that if it completes all state spaces resulting from data-race preemption points,
 that serves as a total verification of all possible thread interleavings of the given test program.
 
-We evaluated \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
-We show that \quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,
-and furthermore that adding data-race preemption points quickly exposes bugs missed by even the ``maximal'' state space of the conventional approach.
+We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
+\quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,
+% joshua wants me to say "conventional approachES" here
+and furthermore, adding data-race preemption points quickly exposes bugs missed by even the ``maximal'' state space of the conventional approach.
 
@@
 \begin{enumerate}
-	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open source implementation of the technique;
+	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation of the technique;
 	\item A proof of convergence, which shows that should it be possible in the given CPU budget,

### 2015-11-14 pldi-16/di.tex
touch up di a bit

@@
 Iterative Deepening searches among different combinations of such points during a test,
-deciding on-the-fly whether to pursue each resulting state space or, or to defer it in favour of smaller ones.
+deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
 Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
@@
 
-For the rest of the paper, we abbreviate {\em preemption point} (PP),
+For the rest of the paper, we will abbreviate {\em preemption point} (PP),
 %For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
@@
 $B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
-If $A$ and $B$ bug-free, they will in turn add $AB$ later.
+If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
 %We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
@@
 	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
-	3 & & \texttt{// x's memory recycled} \\
+	3 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{// ...initialize...}\\
+	5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
 	6 & & \texttt{publish(y);} \\
@@
 	4 & & \texttt{x2 = get\_published\_x();} \\
-	5 & & \texttt{// x's memory recycled} \\
+	5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
 	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\

### 2015-11-14 pldi-16/evaluation.tex
implement joshua's revisions pt1
Write eval for free-re-mallocs and totalverif

@@
 In total, the test suite comprises 629 unique state spaces,
-% TODO: Add number of pintos bugs.
-at least 106 of which contain bugs.
+at least 165 of which contain bugs.
 
@@
 		\hline
-		{\prisema	} & 59	&	&	&	&	&	&	&	&	\\
-		{\waitsimple	} & 52	&	&	&	&	&	&	&	&	\\
-		{\alarmsimul	} & 44	&	&	&	&	&	&	&	&	\\
+		{\prisema	} & 59	&	&	&	&	&	& 0	&	&	\\
+		{\alarmsimul	} & 44	&	&	&	&	&	& 37	&	&	\\
+		{\waitsimple	} & 52	&	&	&	&	&	& 57	&	&	\\
 		\hline
-		{\bf Total}	& 629	&	&	&	&	&	&	&	&
+		{\bf Total}	& 629	&	&	&	&	&	& 1250	&	&
 	\end{tabular}
@@
 
-The left half of Table~\ref{tab:drbugs} breaks down the number and types of bugs found by each test case.
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{bargraph.pdf}
+	\caption{{\bf TODO:} Is this graph productive to include in the paper? Or should I just leave the numbers in table format (the left half of Table~\ref{tab:drbugs})?}
+	\label{fig:bargraph}
+\end{figure}
+
+%Figure~\ref{fig:bargraph}
+The left half of Table~\ref{tab:drbugs}
+breaks down the number and types of bugs found by each test case.
 % The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
@@
 
-We also counted the number of malloc-recycle false positives that \landslide~suppressed.
+Beyond finding new bugs with data-race PPs, we evaluated \quicksand's performance in classifying data-race candidates in two ways.
 
-TODO
+{\bf Suppressing ``malloc-recycle'' false positives.}
+In \sect{\ref{sec:recycle}} we showed the soundness of suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
+In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test case.
+In total, 1250 data-races fit the malloc-recycle pattern across all tests,
+only 69 of which were observed to avoid the re-allocation in an alternate interleaving.
+Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 1181 other state spaces.
+
+Among the 69 true data-races which initially fit the malloc-recycle pattern,
+none of them exposed a new bug when used as a PP.
+This shows that for other data-race tools,
+suppressing malloc-recycle candidates may be a productive heuristic,
+even if unsound without Iterative Deepening.
+However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 tests,
+and fall back to classifying them with DPOR,
+which we consider a much stronger verification.
 
 {\bf Full verification.}
-TODO
+For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
+In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU-time.
+Among these, 36 tests contained no data-race candidates whatsoever, 
+so the same verification could be provided by SSS-MC.
+We plot these for comparison,
+according to the time taken by that test's SSS-MC control experiment,.
+When a test can be totally verified without any data-race PPs,
+\quicksand~will be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
+Of course, using data-race PPs increases our capacity for total verification by 4.25x.
+
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{totalverifs.pdf}
+	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
+	Some tests had no data-race candidates, and hence could also be verified by SSS-MC.}
+	\label{fig:totalverif}
+\end{figure}
 
@@
 
-% TODO: Measure how much subset overlap there is.
+% TODO CAMREADY: Measure how much subset overlap there is.
 {\bf Avoiding redundant work.}
@@
 Because we prioritize completing small state spaces before extending them with more PPs,
-the superset state spaces we run later will always repeat each branch of their already-completed subsets.
+the superset state spaces we run later will repeat each branch of their already-completed subsets.
 %
-We measured the proportion of repeated work among completed state spaces across our test suite;
-on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
+%We measured the proportion of repeated work among completed state spaces across our test suite;
+%on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
 
@@
 
+Likewise, for test cases where full verification is not computationally feasible,
+the state spaces associated with many data-race candidates will time out.
+We cannot guarantee that those races are false positives or benign, even though no bug was found.
+% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
+In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test case.
+For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
+which could be combined with Iterative Deepening in future work.
 % Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-11-14 pldi-16/evaluation.tex
clean up stale TODOs

@@
-% TODO: Abbreviate SMC, and fixed-PP approach.
-
 \section{Evaluation}
@@
 				% How many false positives does the free-re-malloc technique suppress?
-				% TODO: If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
-			% TODO: This one's optional. You can give up on it.
+				% to do (done) If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
+				% Answer: Just 1. (for p2s at least; as of time of writing, pintos dr-falsenegs not run yet)
 			\item Do we find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
-				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
-				% TODO: And disable false-neg malloc-free technique
-				% TODO: And also disable the confirmed/suspected thing
-				% (where mem.c waits for reorder observed before
-				% messaging the latter half of the DR to QS)
 		\end{enumerate}
@@
 \subsection{Test Suite}
-% TODO: Maybe say how many lines of code total? How many lines on average per P2/pintos? (careful with pintos; lots of basecode)
 Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
-%across the Spring and Fall 2014 and Spring 2015 semesters;
 and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
@@
 \newcommand\alarmsimul{\texttt{alarm\_test}}
-% TODO: Add more test cases
+% TODO CAMREADY: Add another test case
 We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
@@
 \label{sec:eval-sssmc}
-
-% TODO: If you have time, rerun all the quicksand experiments JUST running the 4 base state spaces. Give it 10/4 hours of cpu budget on 4 cores. John's expt.
 
@@
 
-% TODO: Make a table with a breakdown by test case.
-
 \begin{table*}[t]
@@
 % The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
-%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
+%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
 % Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
@@
 even if unsound without Iterative Deepening.
-However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 tests,
+However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 distinct tests,
 and fall back to classifying them with DPOR,
@@
 
-% TODO: Make a CDF graph of how many tests got total verification here.
-
 % TODO: Figure out how to visualize how many unique data race PPs were verified, and how many remained as unclassified false positives
@@
 
-% TODO: You should introduce this in the narrative as a key contribution way earlier.
 \subsection{Finding nondeterministic data-race candidates}
@@
 but stress testing's comparative problem with achieving reliable coverage is already well-understood
-% TODO: 1 more good citation here
 \cite{chess-icb,gambit}.

### 2015-11-14 pldi-16/paper.tex
implement joshua's revisions pt1
touch up di a bit
clean up stale TODOs

@@
 \definecolor{brickred}{RGB}{192,0,0}
+\definecolor{commentblue}{RGB}{0,0,255}
 
@@
 %
-Our evaluation shows that combining data-race detection with stateless model checking can find new bugs neither technique would find on their own.
+Our evaluation shows that combining data-race detection with stateless model checking can find new bugs that neither technique would find on their own.
 %Our evaluation shows this technique is
@@
 %and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
-% TODO: Not true; Replace instead with a claim about partial verification.
-Compared to SSS-MC, we find bugs faster in smaller subset state spaces,
+Compared to SSS-MC,
+%, we find bugs faster in smaller subset state spaces, %% not true, or at least, not by very much
 %when the maximal state space is too large to explore
-and find new bugs with data-race PPs that the maximal state space would not expose at all.
+we find new bugs with data-race PPs that the maximal state space would not expose at all.
 Moreover, for tests where all state spaces with data-race PPs can be completed, we provide an equally strong verification as if every single instruction had been used as a PP.

### 2015-11-14 pldi-16/di.tex
clean up stale TODOs
more small edits

@@
 
-% TODO: put colour in this code
 \begin{algorithm}[t]
@@
 i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
-we prefer to add multiple smaller jobs which have a higher chance of completing in time. i.e.,
+we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
 $A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
@@
 \end{lemma}
-% TODO: Put a figure here giving an example of where a data race is only transitively reachable, not initially reachable.
 
@@
 % TODO: \cite{fairstatelessmc} instead or in addition?
-It avoids state space cycles (e.g.
-ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
-% TODO: devote a paragraph to this, after talking about drs in locks
+It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
 % this line can be cut if space is needed
@@
 (Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
-% TODO: Add a code sample showing how the lockset boundaries change? If there's room?
 
@@
 
-% TODO: Understand what prior work means by 'happens before'. It's different from our MUST happen before relation.
-% TODO: Understand how it relates to this. Would it scoop us? Or are we outside their model?
-
 We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.

### 2015-11-14 pldi-16/evaluation.tex
more small edits

@@
 \begin{figure}[t]
+	% TODO: update this with pintos buges
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster.pdf}
@@
 \quicksand~does not attempt any such quantitative guarantee,
-but can at least report which subsets of PPs produced state spaces that completed in time.
+but can at least report which subsets of PPs produced state spaces that did complete in time.
 
@@
 For the other 62, we counted the number of state spaces \quicksand~was able to complete and the number of unique PPs among them.
-In two cases, \quicksand~also failed to complete anything; beyond these,
+In three cases, \quicksand~also failed to complete anything; beyond these,
 between 1 and 253 state spaces were completed, with a mean of 58.9 and median of 29.
-Between 0 and 233 unique PPs were verified among some subsets, with a mean of 27.2 and median of 7.5.
+Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
 These completions guarantee that, if the test case could expose a bug,
@@
 even if unsound without Iterative Deepening.
-However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 distinct tests,
+However, \quicksand~was able to correctly identify the 69 violations of that heuristic among 30 distinct tests,
 and fall back to classifying them with DPOR,

### 2015-11-15 pldi-16/evaluation.tex
add pintos bugs into dowefindbugsfaster.pdf

@@
 involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
-using provided bare-bones mutex, context-switch, and virtal memory implementations
+using provided bare-bones mutex, context-switch, and virtual memory implementations
 \cite{pintos}.
@@
 
-\newcommand\mxtest{\texttt{mx\_test}}
+\newcommand\mxtest{\texttt{mx\_test}} % TODO: find a way to change name to mutex_test without overflowing lines in the next paragraph
 \newcommand\tej{\texttt{join\_test}}
@@
 \begin{figure}[t]
-	% TODO: update this with pintos buges
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster.pdf}
@@
 	by several configurations of \quicksand~and the SSS-MC control.
-	\quicksand~finds 169\% as many bugs with data-race PPs.}
+	\quicksand~finds 179\% as many bugs with data-race PPs.}
 	\label{fig:dowefindbugsfaster}
@@
 with dynamically-discovered data-race PPs enabled:
-we quickly pull ahead of SSS-MC, and ultimately conclude with 69\% more bugs in total.
-The break-even point is at a negligible 45 seconds.
+we quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
+The break-even point is at a negligible 90 seconds.
 

### 2015-11-15 pldi-16/paper.tex
anonymize acknowledgemence

@@
 
-Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
-Thanks to Wind River for the use of their simulator \simics.
-Thanks to
-% TODO: Post-double-blind-review
-% Ji\v{r}\'{i} \v{S}im\v{s}a, John Wilkes, and
-the anonymous PLDI reviewers for their helpful comments.
-This work was supported in part by
-the U.S. Army Research Office under grant number W911NF0910273.
+% TODO CAMREADY
+{\em removed for double-blind review}
+%Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
+%Thanks to Wind River for the use of their simulator \simics.
+%Thanks to
+%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, John Wilkes, and
+%the anonymous PLDI reviewers for their helpful comments.
+%This work was supported in part by
+%the U.S. Army Research Office under grant number W911NF0910273.
 

### 2015-11-15 pldi-16/evaluation.tex
eval present dr-falsenegs result; not entirely happy with explanation

@@
 
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{bargraph.pdf}
-	\caption{{\bf TODO:} Is this graph productive to include in the paper? Or should I just leave the numbers in table format (the left half of Table~\ref{tab:drbugs})?}
-	\label{fig:bargraph}
-\end{figure}
+%\begin{figure}[t]
+%	\includegraphics[width=0.48\textwidth]{bargraph.pdf}
+%	\caption{{\bf TODO:} Is this graph productive to include in the paper? Or should I just leave the numbers in table format (the left half of Table~\ref{tab:drbugs})?}
+%	\label{fig:bargraph}
+%\end{figure}
 
@@
 thereby making them {\em false negatives} of the single-pass approach.
-We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).
+We classified each data-race candidate according to whether \landslide~reported them during the first interleaving,
+before any backtracking or preempting:
+if so, they were {\em single-pass data races}; otherwise, {\em nondeterministic}.
 
 To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
-For example, we reported malloc-recycle data races during the first interleaving as {\em deterministic}, as a single-pass analysis must,
-rather than waiting until future interleavings to confirm them (as explained in \sect{\ref{sec:recycle}}).
+For example, we reported malloc-recycle data races during the first interleaving, as a single-pass analysis must,
+rather than waiting until future interleavings to witness them without the malloc-recycle pattern
+(\sect{\ref{sec:recycle}}).
 
-TODO: Present results.
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{nondet-drs-1-v2.pdf}
+	\caption{Some data-race candidates may not be identified during a single program execution.
+		Using nondeterministic data races as PPs,
+		\quicksand~found 189\% as many bugs compared to the approach of prior work \cite{portend}.
+	}
+	\label{fig:dr-falsenegs}
+\end{figure}
+Figure~\ref{fig:dr-falsenegs} compares the types of data-race candidates necessary to expose each data-race bug in our test suite.
+The first series represents the bugs found using PPs from single-pass data-race candidates,
+% TODO: re read racefuzzer/portend paper to confirm this is true.
+i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
+The second series shows all the data-race bugs \quicksand~found,
+which includes the former type as well as new bugs involving nondeterministic data-races.
+In total, among \quicksand's 68 total data-race bugs, just 36 could be found with single-pass data-race candidates alone.
+%In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.
 
@@
 We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
-are not intended to discover new data race candidates under subsequent runs. %before discovering any initial candidates.
+are not intended to discover new candidates under subsequent runs. %before discovering any initial candidates.
 Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
@@
 \cite{chess-icb,gambit}.
-This experiment also suggests that
-other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use the same sort of feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
+This result suggests that
+other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use a similar feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
 %i.e., discovering transitively-reachable data races while testing initial ones.

### 2015-11-16 pldi-16/di.tex
implement pt2 of joshua comments

@@
 With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
-Note that even though a data race may involve two different instructions, we add new state spaces with only one new PP at a time.
-Rather than adding a single state space configured to preempt on both involved instructions,
+Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
+Rather than adding a single large state space, %configured to preempt on both involved instructions,
 i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
@@
 
-%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
-Note that under a {\em pure happens-before} analysis,
-these accesses are not considered concurrent % at all
-because of {\tt malloc}'s internal locking events,
-and would not result in such false positives.
-However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
-so in our context it is more appropriate to use the
-{\em limited happens-before} relation in a hybrid approach with lockset tracking.
+%%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
+%Note that under a {\em pure happens-before} analysis,
+%these accesses are not considered concurrent % at all
+%because of {\tt malloc}'s internal locking events,
+%and would not result in such false positives.
+%However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
+%so in our context it is more appropriate to use the
+%{\em limited happens-before} relation in a hybrid approach with lockset tracking.
 %the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,

### 2015-11-16 pldi-16/evaluation.tex
implement pt2 of joshua comments

@@
 Though not ``real world'' programs, both projects are quite complex:
-the P2s average 1807 lines of code,%C and x86 assembly (stddev 489.5),
+the P2s average 1807 lines of code, %C and x86 assembly (stddev 489.5),
 % Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
@@
 \newcommand\alarmsimul{\texttt{alarm\_test}}
-% TODO CAMREADY: Add another test case
 We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
@@
 and \waitsimple, for process lifecycle system calls\footnote{
-	Some of the Pintoses were partially implemented, so each test could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
+	Some of the Pintoses were partially implemented, so each test program could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
 }
-The source code of all 9 tests is available at
+The source code of all 9 test cases is available at
 {\em [submitted as supplementary material]}.
 For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
-Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
-In total, the test suite comprises 629 unique state spaces,
-at least 165 of which contain bugs.
+We ran all tests on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+In total, our evaluation comprises 629 tests (pairs of a test program and a Pintos or P2),
+at least 163 of which could expose bugs.
 
@@
 we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
-(i.e., ignoring all data-race candidates)\footnote{
-Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
-we ran these tests for 2.5 hours on 4 CPUs each.
-Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
+(i.e., ignoring all data-race candidates).
+%\footnote{
+%Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
+%we ran these tests for 2.5 hours on 4 CPUs each.
+%Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
 %
@@
 QS-no-DR-PPs finds many more of the bugs much sooner.
-From this we conclude that for smaller arbitrary CPU budgets,
-Iterative Deepening is likely to find bugs SSS-MC will miss,
-and for more ambitiously-sized tests,
-programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
+Hence, for smaller arbitrary CPU budgets,
+Iterative Deepening is likely to find bugs SSS-MC will miss.
+%and for more ambitiously-sized tests,
+%programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
 %We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
@@
 	\end{center}
-	\caption{Summary of number and types of bugs found by each test case.
+	\caption{Summary of number and types of bugs found by each test program.
 		SSS-MC is the control; QS is \quicksand.
@@
 The left half of Table~\ref{tab:drbugs}
-breaks down the number and types of bugs found by each test case.
+breaks down the number and types of bugs found by each test program.
 % The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
@@
 In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
-even compared to the other test cases\footnote{
-% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
+even compared to the other test programs\footnote{
 	The one bug SSS-MC found was in a fully-assembly lock implementation.
@@
 Though it suffices for many applications of MC to assume correctly-implemented locks,
-we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.
+we consider this strong evidence that low-level synchronization code must be verified with data-race PPs.
 
+% TODO: maybe the whole story about partial verification can be cut
 {\bf Partial verification.}
@@
 
-The SSS-MC control experiment timed out in 10 hours on 77 tests.
+On 77 tests, the SSS-MC control experiment timed out in 10 hours with no bugs found.
 Among these, \quicksand~found data-race bugs in 15.
@@
 Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
-These completions guarantee that, if the test case could expose a bug,
+These completions guarantee that, if the test program could expose a bug,
 % Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
@@
 In \sect{\ref{sec:recycle}} we showed the soundness of suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
-In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test case.
+In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
 In total, 1250 data-races fit the malloc-recycle pattern across all tests,
@@
 even if unsound without Iterative Deepening.
-However, \quicksand~was able to correctly identify the 69 violations of that heuristic among 30 distinct tests,
-and fall back to classifying them with DPOR,
-which we consider a much stronger verification.
+However, \quicksand~was able to correctly identify the 69 violations of that heuristic (among 30 distinct tests),
+and fall back to classifying them with DPOR.
+%which we consider a much stronger verification.
 
@@
 When a test can be totally verified without any data-race PPs,
-\quicksand~will be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
+\quicksand~may be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
 Of course, using data-race PPs increases our capacity for total verification by 4.25x.
@@
 \label{sec:eval-falseneg}
+
 Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
@@
 		Using nondeterministic data races as PPs,
-		\quicksand~found 189\% as many bugs compared to the approach of prior work \cite{portend}.
+		\quicksand~found 189\% as many bugs as could be found using single-pass data races alone.
 	}
@@
 i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
-The second series shows all the data-race bugs \quicksand~found,
+The second series shows all data-race bugs \quicksand~found,
 which includes the former type as well as new bugs involving nondeterministic data-races.
-In total, among \quicksand's 68 total data-race bugs, just 36 could be found with single-pass data-race candidates alone.
+\quicksand~found 68 data-race bugs in total, only 36 of which could be found with single-pass data-race candidates alone.
 %In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.
 
-At first, it might seem unfair to compare the data race bugs \quicksand~finds (10 CPU-hours)
-against the candidates found by a prior work's single execution (less than a minute).
-We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
-are not intended to discover new candidates under subsequent runs. %before discovering any initial candidates.
-Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
+Note that we are not comparing how much testing time is required before identifying the data-race candidates involved in each bug.
+Single-pass data races can all be found after a single program execution,
+while \quicksand~may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
+However, prior work data-race tools \cite{tsan}, being not integrated with a MC,
+are not intended to discover new candidates under subsequent runs.
+Running a single-pass data-race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
 but stress testing's comparative problem with achieving reliable coverage is already well-understood
 \cite{chess-icb,gambit}.
-This result suggests that
-other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use a similar feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
+Likewise, replay-based tools \cite{portend} are dependent upon the data-race detector to provide an execution trace leading to each candidate.
+This experiment suggests that
+such tools should be configured to use a similar feedback loop as Iterative Deepening.
 %i.e., discovering transitively-reachable data races while testing initial ones.
@@
 However, in 2 tests, no more than the minimal state space could be verified,
-and in 3 other tests, not even that much.
+and in 3 other tests, no state spaces could be completed at all.
 Larger state spaces often result from finer-grained locking,
@@
 
-Likewise, for test cases where full verification is not computationally feasible,
+Likewise, for tests where full verification is not computationally feasible,
 the state spaces associated with many data-race candidates will time out.
@@
 % FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
-In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test case.
+In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test program.
 For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},

### 2015-11-16 pldi-16/paper.tex
implement pt2 of joshua comments

@@
 
-We have presented \quicksand, an implementation of Iterative Deepening, a new technique for automating the choice of PPs during stateless model checking.
+We have presented Iterative Deepening and \quicksand, a new technique and tool for automating the choice of PPs during stateless model checking.
 % , and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
-\quicksand~incorporates \landslide's data-race analysis to create new PPs tailored specifically to the program under test,
-and manages multiple \landslide~instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
+\quicksand~incorporates data-race analysis to create new PPs tailored specifically to the program under test,
+and manages multiple model checker instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
 
@@
 we find new bugs with data-race PPs that the maximal state space would not expose at all.
-Moreover, for tests where all state spaces with data-race PPs can be completed, we provide an equally strong verification as if every single instruction had been used as a PP.
+Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide an equally strong verification as if every single instruction had been used as a PP.
 

### 2015-11-16 pldi-16/proof/di.tex
implement pt2 of joshua comments

@@
 
+Recent work \cite{sparc-ssm} has proposed hardware techniques for detecting many classes of stale heap pointer accesses, including the one shown in Figure \ref{fig:recycle-bug}.
+This approach could be combined with stateless model checking in future work to identify such bugs immediately,
+rather than requiring Iterative Deepening to explore new state spaces corresponding to the data race.
+However, if the {\tt malloc} call were in thread 1 instead of thread 2, the bug would still be nondeterministic, requiring stateless model checking to expose.
+
 \begin{definition}[Use after free]
@@
 
-In this section we provide (hopefully) intuitive summaries of both our proof goals,
-for readers who might be satisfied without necessarily double-checking the proofs' internal structure.
+In this section we provide (hopefully) intuitive summaries of both our proof goals.
+%for readers not interested in double-checking the proofs' internal structure.
 

### 2015-11-16 pldi-16/related.tex
some more polish; clean up stale todos; widen the table
clean up stale TODOs
one more case of anonymizing
implement pt2 of joshua comments

@@
 
-% TODO: ", and related data race analyses based on ????"
-
 \subsection{Model Checking}
 
-\landslide~uses many established model-checking techniques, dating back
+We build upon many established model-checking techniques, dating back
 % of course
@@
 {\bf Synchronization events only.} CHESS \cite{chess} and dBug \cite{dbug-ssv} instrument the thread library API, and can preempt programs only during calls to this API.
-Hence they will miss any bugs that require interleaving threads at instruction granularity during a data race.
+Hence, they will miss any bugs that require interleaving threads at instruction granularity during a data race.
 CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
@@
 \subsection{Data Race Detection}
+\label{sec:related-dr}
 
@@
 
-% TODO: Figure out why they claim "Happens before produces NO false positives, only benign races".
-% It seems impossible.
-% But if true, it means either (a) they can somehow identify FRM DRs on the 1st pass, not needing to replay,
-% or (b) reuse of memory by malloc is somehow outside their concurrency model.
 Closer to our work, replay analysis \cite{recordreplaydrs} also suppresses false positives by testing multiple thread interleavings.
@@
 
-% \subsection{Other Concurrency Testing Approaches} % TODO: Well?
+% \subsection{Other Concurrency Testing Approaches}
 %

### 2015-11-16 pldi-16/di.tex
implement jean's comments for abstrax & intro

@@
+\section{Design}
+\label{sec:design}
+
+Motivated by state-of-the-art tools' inflexibility to change their preemption points,
+we introduce a new technique called {\em Iterative Deepening} for adjusting the preemption points at runtime.
+Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
+this approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
+Rather than attempting to search a single state space with every available preemption point enabled
+(e.g., preempting on every pthread API call),
+Iterative Deepening searches among different combinations of such points during the test,
+deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
+Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
+We show a visualization in Figure~\ref{fig:id}.
+
 \begin{figure}[t]
@@
 \end{figure}
-\section{Design}
-\label{sec:design}
 
-Motivated by state-of-the-art tools' inflexibility to change their preemption points,
-Iterative Deepening searches among different combinations of such points during a test,
-deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
-Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-We show a visualization in Figure~\ref{fig:id}.
-
-Iterative Deepening depends heavily on state-space estimation \cite{estimation}
+Iterative Deepening relies on state-space estimation \cite{estimation}
 to understand which state spaces are likely to complete on time,

### 2015-11-16 pldi-16/intro.tex
clean up stale TODOs
implement jean's comments for abstrax & intro

@@
 However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
-% TODO: You repeat the following 1.5 sentence in the related work section; maybe you can eliminate this redundancy to save space?
+Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion.
 For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
-It provides an additional data-race analysis to report any violations of this model;
-however, data-race analyses are prone to report false positives
-%and benign races
-which require annotations or imprecise heuristics to suppress \cite{racerx,tsan,datacollider}.
 %
@@
 %
-Configuring a model checker is a tradeoff between schedule coverage and feasibility of completion.
 This work shows how to avoid making that tradeoff decision in advance.
 
-% TODO: ttuttle says this is too much of a jump -- that exactly what "subsets" means is not well defined (i.e, lock vs unlock, while NOT addressing the within/without_function problem)
-
-We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
-Named after the analogous technique in chess AI \cite{iterative-deepening-chess-ai}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
-Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
-\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
-It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.
-%In fact, if given enough time to fully test all discovered data-race preemption points,
-%Iterative Deepening provides a full verification of all thread schedules that could arise from preempting anywhere.
-In fact, Iterative Deepening is fully general:
-we prove that if it completes all state spaces resulting from data-race preemption points,
-that serves as a total verification of all possible thread interleavings of the given test program.
+We present \quicksand,
+a model checking framework for deciding at runtime which preemption points to test,
+according to which resulting state spaces are most likely to fit a prescribed CPU budget.
+It uses data-race analysis to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
+When prior work would time out on large tests by trying several preemption points simultaneously,
+\quicksand~tests smaller state spaces based on subsets of those points, which often finds the same bugs sooner.
+On the other hand, when the CPU budget is large enough to fully test all data-race preemption points,
+we prove that this constitutes a total verification of all possible thread schedules.
+Short of deciding in advance to preempt on every single instruction \cite{spin}, this was not previously possible without data-race preemption points.
 
 We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
-\quicksand~finds more bugs than the conventional stateless model checking approach given the same CPU budget,
-% joshua wants me to say "conventional approachES" here
-and furthermore, adding data-race preemption points quickly exposes bugs missed by even the ``maximal'' state space of the conventional approach.
+We show many advantages over both conventional stateless model checking and single-pass data-race analysis.
+We find that many of the bugs found by the conventional approach can be found faster using subsets of preemption points,
+and that data-race preemption points quickly expose many new bugs that prior model checkers could not find at all.
 
-This paper's contributions are as follows:
+Our contributions are as follows:
 \begin{enumerate}
-	\item Iterative Deepening, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation of the technique;
-	\item A proof of convergence, which shows that should it be possible in the given CPU budget,
+	\item {\em Iterative Deepening}, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation of the technique;
+	\item A proof of convergence, showing that should it be possible in the given CPU budget,
 		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
@@
 		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
-	%\item Techniques for detecting and flattening cyclic state spaces resulting from ad-hoc while-loop synchronization % TODO: is there actually room for this in the paper?
-	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches.
+	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
 \end{enumerate}

### 2015-11-16 pldi-16/paper.tex
implement jean's comments for abstrax & intro

@@
 but suffers from exponential explosion of the state space when the test input parameters are too large.
-Several partial-order reduction techniques exist for mitigating this explosion,
-but even after pruning equivalent interleavings, the state space size for a fixed set of preemption points is unpredictable and often intractable.
+Several reduction techniques can mitigate this explosion,
+but even after pruning equivalent interleavings, the state space size
+%for a fixed set of preemption points
+is often intractable.
+Prior tools are limited to preempting only on synchronization APIs,
+which reduces the space further, but can miss unsynchronized thread communication bugs.
 %
-Data race detection, another concurrency testing approach, focuses on identifying suspicious memory access pairs during a single test execution.
-It avoids concerns of intractable state space size, but is prone to false positives.
+Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
+It avoids concerns of state space size, but is prone to false positives.
 
 We present an {\em Iterative Deepening} framework for stateless model checking,
-which manages the exploration of many state spaces using different subsets of preemption points.
+which manages the exploration of many state spaces using different preemption points.
 It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
-and it incorporates a data-race analysis to dynamically add new preemption points to classify each data race as buggy or benign.
+and it incorporates data-race analysis to add new preemption points on the fly.
+Preempting threads during a data race's instructions
+%These preemption points allow us to check whether each data-race can produce an observable failure, and
+can automatically classify the race as buggy or benign,
+and uncovers new bugs not reachable by prior model checkers.
 %
-Our evaluation shows that combining data-race detection with stateless model checking can find new bugs that neither technique would find on their own.
+It also enables full verification of all possible schedules when every data race is verified as benign within the CPU budget.
+In our evaluation, Iterative Deepening
+%with data-race preemption points
+found 79\% more bugs than the best prior stateless model checking approach.
 %Our evaluation shows this technique is

### 2015-11-16 pldi-16/design.tex
begin major swizzling around of sections according to joan's suggestion

--- 2015-11-16 pldi-16/di.tex
@@
 in which case it remains a potential false positive that the user must handle.
-
-\subsection{Convergence to total verification}
-\label{sec:totalverif}
-
-Although Iterative Deepening's main purpose is to heuristically choose the most effective PP subsets to test
-when the maximal state space is too large,
-some tests may be small enough that even their maximal state spaces could be completed in time.
-For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
-In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
-%In other words, contrapositively,
-%if it is possible to find a bug with any sequence of preemptions on any instruction whatsoever,
-%an equivalent thread interleaving will be reachable using only data-race PPs and synchronization API PPs.
-A proof sketch of the contrapositive statement follows.
-
-\begin{theorem}
-If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
-Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
-\end{theorem}
-\begin{proof}
-The proof has two parts:
-first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
-second, we show that Iterative Deepening will eventually detect all such data races.
-
-\begin{lemma}[Equivalence of non-data-race PPs]
-For any thread interleaving possible by preempting on any instruction,
-there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:relevant}
-\end{lemma}
-
-TODO: sketch the proof of this lemma % TODO
-
-\begin{definition}[Reachable data race]
-A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
-\end{definition}
-Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
-\begin{lemma}[Saturation of data-race PPs]
-	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
-	\label{lem:saturation}
-\end{lemma}
-
-TODO: sketch the proof of this lemma % TODO
-\\
-
-To conclude,
-for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
-and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
-Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
-\end{proof}
-
-Our full proof
-{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
-discusses our assumptions explicitly and includes more formal definitions and structure.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Implementation}
-\label{sec:implementation}
-
-\subsection{Landslide}
-\label{sec:landslide}
-
-We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
-\landslide~implements DPOR \cite{dpor},
-%\cite{dpor},
-state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
-% TODO: \cite{fairstatelessmc} instead or in addition?
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
-% this line can be cut if space is needed
-It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
-% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
-Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
-
-{\bf Restricting PPs with stack trace predicates.}
-Most model checkers (MCs) preempt indiscriminately on any sync API call, regardless of the call-site.
-However, when testing a particular module in a large codebase,
-the user is likely uninterested in PPs arising from other modules.
-\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
-Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
-%The {\tt without\_function} directive works similarly, but as a blacklist.
-The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
-Multiple invocations can be used; later ones take precedence.
-%\cite{landslide} provides further detail on this feature.
-
-{\bf Data races in lock implementations.}
-Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself.
-The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
-Otherwise, with testing limited to one execution,
-%even if one wishes to test for lock bugs,
-data-race analysis would flag every access pair in the lock implementation, requiring human attention to verify.
-However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
-(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
-
-\subsection{Quicksand}
-
-\quicksand~is an independent program that wraps the execution of several \landslide~instances.
-The implementation is roughly 3000 lines of C.
-It uses a thread pool to schedule the available state spaces,
-sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
-Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
-
-{\bf Initial PPs.}
-For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs that we showed in Figure~\ref{fig:id}:
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-and $\{yield,lock,unlock\}$,
-By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
-such as condition variables, semaphores, and r/w locks.
-Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
-%so the {\tt yield} PP is always implicitly enabled.
-
-For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
-%(to appropriate the names of the x86 instructions)
-During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
-%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
-Some kernels disable preemption without disabling interrupts,
-which can be modelled the same way using manual annotations. %of that API.
-This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
-\quicksand~is configured to begin with the subsets
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-$\{yield,cli\}$,
-$\{yield,sti\}$,
-and $\{yield,lock,$ $unlock,cli,sti\}$.
-As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
-which could potentially be improved in future work (\sect{\ref{sec:future}}).
-
-{\bf Communication protocol.}
-The interface to \landslide~, which any similar MC could implement, has two parts.
-First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
-% can lose this line due to space
-among other options such as mutex-testing mode,
-passed as an argument to \landslide.
-Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
-\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
-whenever a new data-race candidate is found, and whenever a bug is found.
-\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
-We implement suspending jobs simply by making \landslide~wait on a message-passing call.
-Should \quicksand~later resume a suspended job, we send a message to continue,
-causing the \landslide~instance to resume exploring where it left off;
-otherwise, we send a message only after time runs out, causing it to exit.
-
-{\bf Heuristics.}
-% List of all heuristix:
-% HOMESTRETCH - last 60sec of test, don't suspend
-% ETA_THRESH - "to let its ETA stabilize"
-% eta factor
-% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
-% priority change between suspected and confirmed dr
-Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
-to express how optimistic we are about the estimate's accuracy.
-We use 2 as this scaling factor based on the results in \cite{estimation},
-though we allow changing it via the command line.
-We also include a heuristic to
-%ignore ETAs entirely
-never suspend jobs before they pass a certain threshold of interleavings tested,
-for which we choose 32,
-so that their ETAs have some time to stabilize.
-
-We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
-Jobs with both-order data-race PPs are prioritized higher,
-because single-order candidates are more likely to be false-positives
-(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
-%so we prioritize jobs with both-order data-race PPs.
-For single-order races, we do not add a PP for the later access at all;
-should it be needed, preempting on the first access will suffice to upgrade the race to both-order.
-
-% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-\subsection{Data-race preemption points}
-
-When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
-Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
-Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
-Instead, \landslide~qualifies the PC with
-(a) the current thread ID and
-% FIXME: We don't actually do this.
-(b) the most recent {\tt call} instruction.
-% (a crude approximation of a stack trace)
-% which are much cheaper, as we carry them around all the time already
-Note that we do {\em not} qualify data races by the shared memory address,
-which can change based on different interleavings of previous code
-(for example, depending on the result of {\tt malloc()}).
-% especially when malloc is involved.
-%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
-	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
-		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
-	\label{fig:new-dr-jobs}
-\end{figure}
-
-When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
-a ``small'' job to preempt on the racing instruction only,
-and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
-%
-Hence, together with the logic in \sect{\ref{sec:classifying}}, each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
-%
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
-while the big job risks not completing in time,
-the small job risks missing the data race entirely if the original PPs were required to expose it.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
-% TODO: Put numbers here.
-
-\subsection{Suppressing ``malloc-recycle'' false positives}
-\label{sec:recycle}
-
-We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.
-Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
-If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
-We term this a {\em malloc-recycle data race}.
-To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
-Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
-
-% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
-By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
-
-It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
-We implemented this check by adding a generation counter to \landslide's heap tracking:
-each allocation is given a unique ID,
-and when evaluating whether two heap accesses can race,
-the IDs of their containing blocks must match.
-However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
-Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
-Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
-	3 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
-\label{fig:recycle}
-\end{figure}
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x);} & \\
-	2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(x);} \\
-	4 & & \texttt{x2 = get\_published\_x();} \\
-	5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If a single-pass data race detector discarded candidates matching the malloc-recycle pattern,
-it would miss the bug in this adversarial program.}
-\label{fig:recycle-bug}
-\end{figure}
-
-%%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
-%Note that under a {\em pure happens-before} analysis,
-%these accesses are not considered concurrent % at all
-%because of {\tt malloc}'s internal locking events,
-%and would not result in such false positives.
-%However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
-%so in our context it is more appropriate to use the
-%{\em limited happens-before} relation in a hybrid approach with lockset tracking.
-%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
-
-Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning these false positives is sound even when such adversarial programs are considered.
-This makes it unnecessary to verify such data races by actively adding more preemptions,
-achieving a potentially combinatorial reduction in how many state spaces we generate.
-%Intuitively, we need not worry about cases such as Figure~\ref{fig:recycle-bug} because,
-%should they be true races,
-%DPOR will reorder threads sufficiently for the malloc-recycle pattern to disappear.
-We provide a proof sketch below.
-
-\renewcommand\proofname{Proof Sketch}
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-If a malloc-recycle data race is not a false positive,
-%DPOR will reorder threads such that
-DPOR will test an alternate thread interleaving in which
-%either
-the accesses still race without fitting the malloc-recycle pattern.
-%, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-\begin{proof}
-By definition of the malloc-recycle pattern,
-any such program must contain an access {\tt x1} by one thread T1,
-followed by a {\tt free} and a {\tt malloc} possibly by either thread,
-followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
-For brevity we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
-We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
-hence, there must also be some ``publish'' action {\tt p} by T1 which communicates the address to T2.
-Because this is a true data race, {\tt p} must occur before {\tt x1}, as {\tt x2} cannot be reordered before {\tt p}.
-
-We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
-The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
-If locking or message-passing is used, our set of hard-coded PPs suffices.
-Otherwise, {\tt p} (and the corresponding read by T2) will be a data race, although it may itself be a malloc-recycle race.
-In this case we use induction on the pointer chain leading to {\tt x}:
-in the base case, {\tt p} is global or obtained via message-passing,
-and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on {\tt p}.
-Hence there will be a PP between {\tt p} and {\tt x1} no matter the mode of communication.
-
-By definition of DPOR, this PP causes {\tt x2} to be reordered before {\tt x1} while not changing {\tt x2}'s location.
-As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
-Hence {\tt x1} and {\tt x2} will be in the same allocation;
-hence the accesses can race without fitting the malloc-recycle pattern.
-% Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-% This spells QED so we are done.
-\end{proof}
-
-Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
-which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
-Our full proof
-{\em [see supplementary material]}
-provides the similar cases in more detail.
-\renewcommand\proofname{Proof}
-
-%For further discussion and for the full proof, refer to
-%{\em our supplementary material which we will cite as a tech report in the final version of the paper}.

### 2015-11-16 pldi-16/implementation.tex
begin major swizzling around of sections according to joan's suggestion

--- 2015-11-14 pldi-16/di.tex
@@
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees.pdf}
-	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
-	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
-	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
-	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
-	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
-		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
-Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
-Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
-}
-	\label{fig:id}
-\end{figure}
-\section{Design}
-\label{sec:design}
-
-Motivated by state-of-the-art tools' inflexibility to change their preemption points,
-Iterative Deepening searches among different combinations of such points during a test,
-deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
-Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-We show a visualization in Figure~\ref{fig:id}.
-
-Iterative Deepening depends heavily on state-space estimation \cite{estimation}
-to understand which state spaces are likely to complete on time,
-in advance of actually testing each interleaving within.
-The purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user must provide only their total CPU budget as a test parameter,
-and to enable completing appropriately-sized state spaces within that budget.
-
-Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
-A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
-Moreover, because Iterative Deepening treats the set of preemption points as mutable,
-it can add new ones reactively based on any runtime analysis.
-In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
-
-\subsection{Terminology}
-
-For the rest of the paper, we will abbreviate {\em preemption point} (PP),
-%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
-{\em model checking} (MC),
-{\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
-Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
-
-SSS-MC indicates the approach of prior model checking tools:
-the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
-%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
-Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
-but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.
-
-We distinguish between data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is prone to false-positives,
-we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
-calling such pairs {\em data race candidates}.
-Should a future interleaving, preempting during those accesses,
-% TODO: joshua wants 'failure' to be explicitly defined
-lead to a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
-If they cannot be reordered at all, it is a {\em false positive}.
-%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
-%For brevity, we count benign data races as a subset of false positives.
-
-We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}).
-The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
-%However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
-%% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.
-
-\subsection{Choosing the best job}
-
-\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
-\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
-\newcommand\GetETA[1]{ETA(#1)}
-\newcommand\GetPPSet[1]{PPSet(#1)}
-With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
-Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
-and a set of {\em pending} jobs (untested ones with unknown ETAs).
-When a MC instance reports an ETA too high,
-we compare other pending jobs to find another one more likely to complete in time.
-%
-Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
-Though its worst-case performance is $O(mn)$ in the
-%number of pending and suspended jobs,
-sizes of $\mathcal{P}$ and $\mathcal{S}$,
-in practice the non-constant portion beyond line 4 runs very infrequently
-and is negligible compared to the exponentially-sized state spaces.}.
-Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
-and $j_1$ is suspended,
-then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
-%reveals that it might finish in time after all.
-Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
-
-\begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
-	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
-	\Input{$j_0$, the currently-running job}
-	%\Input{$eta$, $j_0$'s predicted completion time}
-	\Input{\PendingJobs, the list of pending jobs, sorted decreasingly by heuristic priority}
-	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted increasingly by ETA}
-	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
-		// Common case: Job is expected to complete. \\
-		return $j_0$
-	}
-	\ForEach{job $j_P \in$ \PendingJobs}{
-		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
-		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
-			return $j_P$
-		}
-	}
-	%// no pending jobs; maybe resume a suspended job \\
-	\ForEach{job $j_S \in$ \SuspendedJobs}{
-		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
-			$\land$
-			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
-			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
-			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
-				return $j_S$
-			}
-		}
-	}
-	return $j_0$ // ETA was bad, but no other job was better.
-	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
-	\label{alg:shouldworkblock}
-\end{algorithm}
-
-%
-We also account for the inherent inaccuracy of ETA estimates.
-Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
-in case their ETAs are actually overestimated.
-Lines 12-14 account for the
-%bizarre
-possibility that among two suspended jobs,
-%given two jobs,
-%%$j_1,j_2$,
-\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
-but
-\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
-This can arise because estimates tend to get more accurate over time,
-and $j_1$ perhaps ran much longer before suspending.
-% In such scenarios,
-We heuristically assume the smaller job's ETA is more accurate
-to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
-(it lets us avoid thrashing in \quicksand).
-
-\subsection{Classifying data races as buggy or benign}
-\label{sec:classifying}
-
-Iterative Deepening must be seeded with some initial state spaces,
-which can be any number of subsets of the statically-available PPs that SSS-MC would use.
-While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
-Because data races indicate access pairs that can interleave at instruction granularity,
-it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
-
-\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
-\begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
-	\Input{$j_0$, the currently-running job}
-	\Input{\AllJobs, the set of all existing jobs}
-	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
-	\If{$\forall j \in \AllJobs,$
-	\GetPPSet{$j_0$} $\cup$ $\alpha$
-	$\not\subseteq$
-	\GetPPSet{$j$}
-	}{
-		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$))
-	}
-	\caption{Adding new jobs with data-race PPs.}
-	\label{alg:handledatarace}
-\end{algorithm}
-
-With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
-Note that even though a data race may involve two different instructions, we add new state spaces with only one new PP at a time.
-Rather than adding a single state space configured to preempt on both involved instructions,
-i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
-we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
-$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
-$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
-If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
-%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
-The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
-for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.
-
-The new spaces may expose a failure, in which case we report a data-race bug,
-or complete successfully, which indicates a benign or false-positive data race.
-They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
-in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
-%Because Iterative Deepening is
-Being constrained by a CPU budget,
-we may time out before completing a data race's associated state space,
-in which case it remains a potential false positive that the user must handle.
-
-\subsection{Convergence to total verification}
-\label{sec:totalverif}
-
-Although Iterative Deepening's main purpose is to heuristically choose the most effective PP subsets to test
-when the maximal state space is too large,
-some tests may be small enough that even their maximal state spaces could be completed in time.
-For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
-In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
-%In other words, contrapositively,
-%if it is possible to find a bug with any sequence of preemptions on any instruction whatsoever,
-%an equivalent thread interleaving will be reachable using only data-race PPs and synchronization API PPs.
-A proof sketch of the contrapositive statement follows.
-
-\begin{theorem}
-If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
-Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
-\end{theorem}
-\begin{proof}
-The proof has two parts:
-first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
-second, we show that Iterative Deepening will eventually detect all such data races.
-
-\begin{lemma}[Equivalence of non-data-race PPs]
-For any thread interleaving possible by preempting on any instruction,
-there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
-	\label{lem:relevant}
-\end{lemma}
-
-TODO: sketch the proof of this lemma % TODO
-
-\begin{definition}[Reachable data race]
-A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
-\end{definition}
-Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
-\begin{lemma}[Saturation of data-race PPs]
-	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
-	\label{lem:saturation}
-\end{lemma}
-
-TODO: sketch the proof of this lemma % TODO
-\\
-
-To conclude,
-for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
-and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
-Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
-\end{proof}
-
-Our full proof
-{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
-discusses our assumptions explicitly and includes more formal definitions and structure.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Implementation}
@@
 % TODO: Put numbers here.
-
-\subsection{Suppressing ``malloc-recycle'' false positives}
-\label{sec:recycle}
-
-We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.
-Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
-If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
-We term this a {\em malloc-recycle data race}.
-To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
-Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
-
-% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
-By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
-
-It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
-We implemented this check by adding a generation counter to \landslide's heap tracking:
-each allocation is given a unique ID,
-and when evaluating whether two heap accesses can race,
-the IDs of their containing blocks must match.
-However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
-Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
-Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-	2 & \texttt{\hilight{olivegreen}{free}(x);} \\
-	3 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
-	6 & & \texttt{publish(y);} \\
-	7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
-\end{tabular}
-\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
-\label{fig:recycle}
-\end{figure}
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
-	1 & \texttt{publish(x);} & \\
-	2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-	3 & \texttt{\hilight{olivegreen}{free}(x);} \\
-	4 & & \texttt{x2 = get\_published\_x();} \\
-	5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
-	6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-\end{tabular}
-\caption{If a single-pass data race detector discarded candidates matching the malloc-recycle pattern,
-it would miss the bug in this adversarial program.}
-\label{fig:recycle-bug}
-\end{figure}
-
-%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
-Note that under a {\em pure happens-before} analysis,
-these accesses are not considered concurrent % at all
-because of {\tt malloc}'s internal locking events,
-and would not result in such false positives.
-However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
-so in our context it is more appropriate to use the
-{\em limited happens-before} relation in a hybrid approach with lockset tracking.
-%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
-
-Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning these false positives is sound even when such adversarial programs are considered.
-This makes it unnecessary to verify such data races by actively adding more preemptions,
-achieving a potentially combinatorial reduction in how many state spaces we generate.
-%Intuitively, we need not worry about cases such as Figure~\ref{fig:recycle-bug} because,
-%should they be true races,
-%DPOR will reorder threads sufficiently for the malloc-recycle pattern to disappear.
-We provide a proof sketch below.
-
-\renewcommand\proofname{Proof Sketch}
-
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-If a malloc-recycle data race is not a false positive,
-%DPOR will reorder threads such that
-DPOR will test an alternate thread interleaving in which
-%either
-the accesses still race without fitting the malloc-recycle pattern.
-%, or a use-after-free bug will be reported immediately.
-\end{theorem}
-
-\begin{proof}
-By definition of the malloc-recycle pattern,
-any such program must contain an access {\tt x1} by one thread T1,
-followed by a {\tt free} and a {\tt malloc} possibly by either thread,
-followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
-For brevity we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
-We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
-hence, there must also be some ``publish'' action {\tt p} by T1 which communicates the address to T2.
-Because this is a true data race, {\tt p} must occur before {\tt x1}, as {\tt x2} cannot be reordered before {\tt p}.
-
-We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
-The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
-If locking or message-passing is used, our set of hard-coded PPs suffices.
-Otherwise, {\tt p} (and the corresponding read by T2) will be a data race, although it may itself be a malloc-recycle race.
-In this case we use induction on the pointer chain leading to {\tt x}:
-in the base case, {\tt p} is global or obtained via message-passing,
-and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on {\tt p}.
-Hence there will be a PP between {\tt p} and {\tt x1} no matter the mode of communication.
-
-By definition of DPOR, this PP causes {\tt x2} to be reordered before {\tt x1} while not changing {\tt x2}'s location.
-As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
-Hence {\tt x1} and {\tt x2} will be in the same allocation;
-hence the accesses can race without fitting the malloc-recycle pattern.
-% Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-% This spells QED so we are done.
-\end{proof}
-
-Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
-which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
-Our full proof
-{\em [see supplementary material]}
-provides the similar cases in more detail.
-\renewcommand\proofname{Proof}
-
-%For further discussion and for the full proof, refer to
-%{\em our supplementary material which we will cite as a tech report in the final version of the paper}.

### 2015-11-16 pldi-16/intro.tex
begin major swizzling around of sections according to joan's suggestion

@@
 The remainder of the paper is organized as follows.
-\sect{\ref{sec:design}} discusses the background and design of Iterative Deepening, including our proof of convergence,
-\sect{\ref{sec:implementation}} explains \quicksand's approach to implementing it, including our new false-positive data-race tactic,
+\sect{\ref{sec:overview}} reviews the background on stateless model checking and data-race analysis,
+\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation of Iterative Deepening,
+\sect{\ref{sec:soundness}} presents our proofs of convergence and of the soundness of our new false-positive data-race tactic,
 \sect{\ref{sec:eval}} presents our evaluation,

### 2015-11-16 pldi-16/soundness.tex
begin major swizzling around of sections according to joan's suggestion

--- 2015-11-16 pldi-16/di.tex
@@
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees.pdf}
-	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
-	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
-	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
-	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
-	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
-		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
-Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
-Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
-}
-	\label{fig:id}
-\end{figure}
-\section{Design}
-\label{sec:design}
+\section{Proofs (TODO: need better name)}
+\label{sec:soundness}
 
-Motivated by state-of-the-art tools' inflexibility to change their preemption points,
-Iterative Deepening searches among different combinations of such points during a test,
-deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
-Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-We show a visualization in Figure~\ref{fig:id}.
+In this section we present two theorems concerning Iterative Deepening's correctness.
+Our full proofs
+{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
+discuss our assumptions explicitly and include more formal definitions and structure.
 
-Iterative Deepening depends heavily on state-space estimation \cite{estimation}
-to understand which state spaces are likely to complete on time,
-in advance of actually testing each interleaving within.
-The purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user must provide only their total CPU budget as a test parameter,
-and to enable completing appropriately-sized state spaces within that budget.
+\renewcommand\proofname{Proof Sketch}
 
-Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
-A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
-Moreover, because Iterative Deepening treats the set of preemption points as mutable,
-it can add new ones reactively based on any runtime analysis.
-In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
-
-\subsection{Terminology}
-
-For the rest of the paper, we will abbreviate {\em preemption point} (PP),
-%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
-{\em model checking} (MC),
-{\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
-Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
-
-SSS-MC indicates the approach of prior model checking tools:
-the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
-%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
-Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
-but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.
-
-We distinguish between data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is prone to false-positives,
-we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
-calling such pairs {\em data race candidates}.
-Should a future interleaving, preempting during those accesses,
-% TODO: joshua wants 'failure' to be explicitly defined
-lead to a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
-If they cannot be reordered at all, it is a {\em false positive}.
-%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
-%For brevity, we count benign data races as a subset of false positives.
-
-We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}).
-The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
-%However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
-%% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.
-
-\subsection{Choosing the best job}
-
-\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
-\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
-\newcommand\GetETA[1]{ETA(#1)}
-\newcommand\GetPPSet[1]{PPSet(#1)}
-With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
-Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
-and a set of {\em pending} jobs (untested ones with unknown ETAs).
-When a MC instance reports an ETA too high,
-we compare other pending jobs to find another one more likely to complete in time.
-%
-Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
-Though its worst-case performance is $O(mn)$ in the
-%number of pending and suspended jobs,
-sizes of $\mathcal{P}$ and $\mathcal{S}$,
-in practice the non-constant portion beyond line 4 runs very infrequently
-and is negligible compared to the exponentially-sized state spaces.}.
-Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
-and $j_1$ is suspended,
-then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
-%reveals that it might finish in time after all.
-Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
-
-\begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
-	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
-	\Input{$j_0$, the currently-running job}
-	%\Input{$eta$, $j_0$'s predicted completion time}
-	\Input{\PendingJobs, the list of pending jobs, sorted decreasingly by heuristic priority}
-	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted increasingly by ETA}
-	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
-		// Common case: Job is expected to complete. \\
-		return $j_0$
-	}
-	\ForEach{job $j_P \in$ \PendingJobs}{
-		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
-		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
-			return $j_P$
-		}
-	}
-	%// no pending jobs; maybe resume a suspended job \\
-	\ForEach{job $j_S \in$ \SuspendedJobs}{
-		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
-			$\land$
-			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
-			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
-			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
-				return $j_S$
-			}
-		}
-	}
-	return $j_0$ // ETA was bad, but no other job was better.
-	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
-	\label{alg:shouldworkblock}
-\end{algorithm}
-
-%
-We also account for the inherent inaccuracy of ETA estimates.
-Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
-in case their ETAs are actually overestimated.
-Lines 12-14 account for the
-%bizarre
-possibility that among two suspended jobs,
-%given two jobs,
-%%$j_1,j_2$,
-\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
-but
-\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
-This can arise because estimates tend to get more accurate over time,
-and $j_1$ perhaps ran much longer before suspending.
-% In such scenarios,
-We heuristically assume the smaller job's ETA is more accurate
-to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
-(it lets us avoid thrashing in \quicksand).
-
-\subsection{Classifying data races as buggy or benign}
-\label{sec:classifying}
-
-Iterative Deepening must be seeded with some initial state spaces,
-which can be any number of subsets of the statically-available PPs that SSS-MC would use.
-While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
-Because data races indicate access pairs that can interleave at instruction granularity,
-it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
-
-\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
-\begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
-	\Input{$j_0$, the currently-running job}
-	\Input{\AllJobs, the set of all existing jobs}
-	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
-	\If{$\forall j \in \AllJobs,$
-	\GetPPSet{$j_0$} $\cup$ $\alpha$
-	$\not\subseteq$
-	\GetPPSet{$j$}
-	}{
-		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$))
-	}
-	\caption{Adding new jobs with data-race PPs.}
-	\label{alg:handledatarace}
-\end{algorithm}
-
-With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
-Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
-Rather than adding a single large state space, %configured to preempt on both involved instructions,
-i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
-we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
-$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
-$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
-If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
-%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
-The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
-for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.
-
-The new spaces may expose a failure, in which case we report a data-race bug,
-or complete successfully, which indicates a benign or false-positive data race.
-They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
-in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
-%Because Iterative Deepening is
-Being constrained by a CPU budget,
-we may time out before completing a data race's associated state space,
-in which case it remains a potential false positive that the user must handle.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Our full proof
-{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
-discusses our assumptions explicitly and includes more formal definitions and structure.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Implementation}
-\label{sec:implementation}
-
-\subsection{Landslide}
-\label{sec:landslide}
-
-We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
-\landslide~implements DPOR \cite{dpor},
-%\cite{dpor},
-state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
-% TODO: \cite{fairstatelessmc} instead or in addition?
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
-% this line can be cut if space is needed
-It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
-% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
-Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
-
-{\bf Restricting PPs with stack trace predicates.}
-Most model checkers (MCs) preempt indiscriminately on any sync API call, regardless of the call-site.
-However, when testing a particular module in a large codebase,
-the user is likely uninterested in PPs arising from other modules.
-\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
-Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
-%The {\tt without\_function} directive works similarly, but as a blacklist.
-The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
-Multiple invocations can be used; later ones take precedence.
-%\cite{landslide} provides further detail on this feature.
-
-{\bf Data races in lock implementations.}
-Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself.
-The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
-Otherwise, with testing limited to one execution,
-%even if one wishes to test for lock bugs,
-data-race analysis would flag every access pair in the lock implementation, requiring human attention to verify.
-However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
-(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
-
-\subsection{Quicksand}
-
-\quicksand~is an independent program that wraps the execution of several \landslide~instances.
-The implementation is roughly 3000 lines of C.
-It uses a thread pool to schedule the available state spaces,
-sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
-Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
-
-{\bf Initial PPs.}
-For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs that we showed in Figure~\ref{fig:id}:
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-and $\{yield,lock,unlock\}$,
-By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
-such as condition variables, semaphores, and r/w locks.
-Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
-%so the {\tt yield} PP is always implicitly enabled.
-
-For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
-%(to appropriate the names of the x86 instructions)
-During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
-%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
-Some kernels disable preemption without disabling interrupts,
-which can be modelled the same way using manual annotations. %of that API.
-This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
-\quicksand~is configured to begin with the subsets
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-$\{yield,cli\}$,
-$\{yield,sti\}$,
-and $\{yield,lock,$ $unlock,cli,sti\}$.
-As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
-which could potentially be improved in future work (\sect{\ref{sec:future}}).
-
-{\bf Communication protocol.}
-The interface to \landslide~, which any similar MC could implement, has two parts.
-First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
-% can lose this line due to space
-among other options such as mutex-testing mode,
-passed as an argument to \landslide.
-Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
-\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
-whenever a new data-race candidate is found, and whenever a bug is found.
-\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
-We implement suspending jobs simply by making \landslide~wait on a message-passing call.
-Should \quicksand~later resume a suspended job, we send a message to continue,
-causing the \landslide~instance to resume exploring where it left off;
-otherwise, we send a message only after time runs out, causing it to exit.
-
-{\bf Heuristics.}
-% List of all heuristix:
-% HOMESTRETCH - last 60sec of test, don't suspend
-% ETA_THRESH - "to let its ETA stabilize"
-% eta factor
-% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
-% priority change between suspected and confirmed dr
-Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
-to express how optimistic we are about the estimate's accuracy.
-We use 2 as this scaling factor based on the results in \cite{estimation},
-though we allow changing it via the command line.
-We also include a heuristic to
-%ignore ETAs entirely
-never suspend jobs before they pass a certain threshold of interleavings tested,
-for which we choose 32,
-so that their ETAs have some time to stabilize.
-
-We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
-Jobs with both-order data-race PPs are prioritized higher,
-because single-order candidates are more likely to be false-positives
-(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
-%so we prioritize jobs with both-order data-race PPs.
-For single-order races, we do not add a PP for the later access at all;
-should it be needed, preempting on the first access will suffice to upgrade the race to both-order.
-
-% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-\subsection{Data-race preemption points}
-
-When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
-Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
-Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
-Instead, \landslide~qualifies the PC with
-(a) the current thread ID and
-% FIXME: We don't actually do this.
-(b) the most recent {\tt call} instruction.
-% (a crude approximation of a stack trace)
-% which are much cheaper, as we carry them around all the time already
-Note that we do {\em not} qualify data races by the shared memory address,
-which can change based on different interleavings of previous code
-(for example, depending on the result of {\tt malloc()}).
-% especially when malloc is involved.
-%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
-	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
-		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
-	\label{fig:new-dr-jobs}
-\end{figure}
-
-When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
-a ``small'' job to preempt on the racing instruction only,
-and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
-%
-Hence, together with the logic in \sect{\ref{sec:classifying}}, each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
-%
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
-while the big job risks not completing in time,
-the small job risks missing the data race entirely if the original PPs were required to expose it.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
-% TODO: Put numbers here.
 
@@
 
-\renewcommand\proofname{Proof Sketch}
-
 \begin{theorem}[Soundness of eliminating malloc-recycle races]
@@
 \end{proof}
+\renewcommand\proofname{Proof}
 
@@
 which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
-Our full proof
-{\em [see supplementary material]}
-provides the similar cases in more detail.
-\renewcommand\proofname{Proof}
-
-%For further discussion and for the full proof, refer to
-%{\em our supplementary material which we will cite as a tech report in the final version of the paper}.

### 2015-11-16 pldi-16/related.tex
add inspect to related work

@@
 which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
-SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
-However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
-SPIN also requires code to be written in the PROMELA DSL.
+%
+SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment,
+although it requires code to be written in the PROMELA DSL.
+%However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
+%
+Inspect \cite{inspect} instruments source code by inserting wrapper calls around all accesses to potentially-shared data.
+It identifies such instructions in advance with an over-approximating alias analysis,
+while \landslide~\cite{landslide} traces the memory locations of accesses at runtime.
+Both SPIN and Inspect fix their set of PPs in advance, so could be combined with \quicksand~in future work.
 %so cannot check implementations directly.

### 2015-11-16 pldi-16/design.tex
Write overview section

@@
 
-Motivated by state-of-the-art tools' inflexibility to change their preemption points,
-we introduce a new technique called {\em Iterative Deepening} for adjusting the preemption points at runtime.
 Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
-this approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
-Rather than attempting to search a single state space with every available preemption point enabled
-(e.g., preempting on every pthread API call),
-Iterative Deepening searches among different combinations of such points during the test,
-deciding on-the-fly whether to pursue each resulting state space, or to defer it in favour of smaller ones.
-Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-We show a visualization in Figure~\ref{fig:id}.
-
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees.pdf}
-	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
-	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
-	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
-	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
-	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
-		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
-Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
-Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
-}
-	\label{fig:id}
-\end{figure}
-
-Iterative Deepening relies on state-space estimation \cite{estimation}
+Iterative Deepening likewise makes progressively deeper searches of the state space until the CPU budget is exhausted.
+In our context, the ``depth'' is the number of PPs used.
+We rely on state-space estimation \cite{estimation}
 to understand which state spaces are likely to complete on time,
 in advance of actually testing each interleaving within.
-The purpose is to make decisions automatically about when to defer exploration of a state space,
+The overall purpose is to make decisions automatically about when to defer exploration of a state space,
 so an inexpert user must provide only their total CPU budget as a test parameter,
@@
 
-\subsection{Terminology}
+\subsection{Choosing the best job}
 
-For the rest of the paper, we will abbreviate {\em preemption point} (PP),
-%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
-{\em model checking} (MC),
-{\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
-Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
-
-SSS-MC indicates the approach of prior model checking tools:
-the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
-%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
-Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
-but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.
-
-We distinguish between data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is prone to false-positives,
-we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
-calling such pairs {\em data race candidates}.
-Should a future interleaving, preempting during those accesses,
-% TODO: joshua wants 'failure' to be explicitly defined
-lead to a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
-If they cannot be reordered at all, it is a {\em false positive}.
-%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
-%For brevity, we count benign data races as a subset of false positives.
-
-We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}).
-The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
-%However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
-%% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.
-
-\subsection{Choosing the best job}
+% TODO: Move the 'seed sets' discussion up here, from implementation.
 

### 2015-11-16 pldi-16/evaluation.tex
Write overview section

@@
 {\bf Finding new data-race bugs.}
-Though state-of-the-art MCs preempt only on synchronization events, many serious concurrency bugs are caused by data races leading to corrupted shared state.
 The line QS-DRs represents our ``no holds barred'' \quicksand~tests,

### 2015-11-16 pldi-16/intro.tex
Write overview section

@@
 %
-On the other hand, SPIN \cite{spin}
+On the other hand, Inspect \cite{inspect} %SPIN \cite{spin}
 is able to preempt threads around any shared memory access.
@@
 \begin{enumerate}
-	\item {\em Iterative Deepening}, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation of the technique;
+	\item {\em Iterative Deepening}, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
 	\item A proof of convergence, showing that should it be possible in the given CPU budget,
@@
 The remainder of the paper is organized as follows.
-\sect{\ref{sec:overview}} reviews the background on stateless model checking and data-race analysis,
-\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation of Iterative Deepening,
+\sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,
+\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation, %of Iterative Deepening,
 \sect{\ref{sec:soundness}} presents our proofs of convergence and of the soundness of our new false-positive data-race tactic,

### 2015-11-16 pldi-16/overview.tex
Write overview section

\section{Overview}
\label{sec:overview}

We review the background in stateless model checking and data-race analysis using the example program in Figure~\ref{fig:example}.

\begin{figure}[t]
	\small
\begin{tabular}{rll}
	& \multicolumn{2}{c}{\texttt{int x = 0; mutex\_t mx;}} \\
	& {\bf Thread 1} & {\bf Thread 2} \\
	1 & \texttt{\hilight{orange}{mutex\_lock(\&mx);}} & \\
	2 & \texttt{int tmp = x;} &\\
	3 & & \texttt{atomic\_xadd(\&x, 1);} \\
	4 & & \texttt{\hilight{olivegreen}{yield();}} \\
	5 & & \texttt{atomic\_xadd(\&x, 1);} \\
	6 & \texttt{x = tmp + 1;} & \\
	7 & \texttt{\hilight{olivegreen}{yield();}} & \\
	8 & \texttt{\hilight{commentblue}{mutex\_unlock(\&mx);}} & \\
	9 & \texttt{assert(x >= 1);}  & \\
	10 & & \texttt{assert(x >= 2);} \\
	%1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
	%2 & \texttt{\hilight{olivegreen}{free}(x);} \\
	%3 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
	%4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
	%5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
	%6 & & \texttt{publish(y);} \\
	%7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
\end{tabular}
\caption{Example program with a data-race bug. In this interleaving, the assertion on line 10 will fail. Two data-race preemptions are required to expose the bug: one just before thread 1's line 6, and one just before thread 2's line 10.}
\label{fig:example}
\end{figure}

\subsection{Stateless Model Checking}

Stateless model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
stateless model checkers focus on concrete observable failures such as assertions, deadlocks, and segfaults.
A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.

% TODO: Fix picture.
\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{trees.pdf}
	\caption{Iterative Deepening example.
		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
Each option increases the state space size unpredictably, so multiple state spaces should be tested in parallel.
Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
}
	\label{fig:id}
\end{figure}

The checker defins the granularity of thread interleavings by which {\em preemption points} it uses to switch threads.
Most model checkers \cite{chess} choose synchronization/thread library API boundaries for these points;
in our example program, these would be lines 1, 4, 7, and 8.
Figure~\ref{fig:id} shows several possible resulting state spaces.
The approach of prior work is to enable all preemption points simultaneously, i.e., to test only the bottom state space.

To mitigate the exponential explosion,
Dynamic Partial Order Reduction \cite{dpor} identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving.
Nevertheless, state spaces are still exponentially-sized in the number of conflicting transitions.

This motivates {\em Iterative Deepening}, our new technique for adjusting the preemption points at runtime.
Rather than committing to one state space with every available preemption point enabled,
Iterative Deepening searches among different {\em subsets} of the points;
for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
Hence, we will test all the state spaces shown in Figure~\ref{fig:id} in parallel,
and decide on-the-fly whether to pursue each test, or to defer it in favour of smaller ones.

\subsection{Data Race Detection}

Data race detection \cite{eraser} is an analysis which identifies pairs of unsynchronized memory accesses between threads.
Two instructions are said to race if
they both access the same memory address,
at least one is a write,
the threads do not hold the same lock,
and no synchronization enforces the order of the thread transitions.
In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 9, and line 6 further races with 10.

Though state-of-the-art MCs preempt only on synchronization events,
many serious concurrency bugs are caused by data races leading to corrupted shared state.
Figure~\ref{fig:example}'s thread interleaving is possible only with {\em data-race preemption points}:
preempting a thread just before an instruction identified as part of a data race.
None of the state spaces in Figure~\ref{fig:id} contain this interleaving.

\subsection{Terminology}

For the rest of the paper, we will abbreviate {\em preemption point} (PP),
%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
{\em model checking} (MC),
{\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).

SSS-MC indicates the approach of prior model checking tools:
the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.

We distinguish between data-race {\em candidates} and data-race {\em bugs}.
Because data-race analysis is prone to false-positives,
we classify unprotected access pairs separately from bugs,
calling such pairs {\em data race candidates}.
Should a future interleaving, preempting during those accesses,
lead to a failure, then we report a {\em data-race bug}.
Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
If they cannot be reordered at all, it is a {\em false positive}.
%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
%For brevity, we count benign data races as a subset of false positives.

We also identify the {\em minimal} and {\em maximal state space} for each test.
The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}, top).
The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}, bottom).
%However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.
%% don't say this ^ -- because when you add dr pps, they add in pairs and there would be multiple maximals, at least until they each get explored and subseuqnetly add a pp of the other of the pair.

### 2015-11-16 pldi-16/overview.tex
Fix pictures

@@
 	6 & \texttt{x = tmp + 1;} & \\
-	7 & \texttt{\hilight{olivegreen}{yield();}} & \\
-	8 & \texttt{\hilight{commentblue}{mutex\_unlock(\&mx);}} & \\
-	9 & \texttt{assert(x >= 1);}  & \\
-	10 & & \texttt{assert(x >= 2);} \\
+	7 & \texttt{\hilight{commentblue}{mutex\_unlock(\&mx);}} & \\
+	8 & \texttt{assert(x >= 1);}  & \\
+	9 & & \texttt{assert(x >= 2);} \\
 	%1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
@@
 \caption{Example program with a data-race bug.
-In this interleaving, the assertion on line 10 will fail.
-Two data-race preemptions are required to expose the bug: one just before thread 1's line 6, and one just before thread 2's line 10.}
+In this interleaving, the assertion on line 9 will fail.
+Two data-race preemptions are required to expose the bug: one just before thread 1's line 6, and one just before thread 2's line 9.}
 \label{fig:example}
@@
 
-% TODO: Fix picture.
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees.pdf}
+	\includegraphics[width=0.48\textwidth]{trees-v2.pdf}
 	\caption{Iterative Deepening example.
@@
 Most model checkers \cite{chess} choose synchronization/thread library API boundaries for these points;
-in our example program, these would be lines 1, 4, 7, and 8.
+in our example program, these would be lines 1, 4, and 7.
 Figure~\ref{fig:id} shows several possible resulting state spaces.
@@
 and no synchronization enforces the order of the thread transitions.
-In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 9, and line 6 further races with 10.
+In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 8, and line 6 further races with 9.
 
@@
 preempting a thread just before an instruction identified as part of a data race.
-None of the state spaces in Figure~\ref{fig:id} contain this interleaving.
+None of the state spaces in Figure~\ref{fig:id} contain this interleaving,
+as none of the mutex/yield preemptions split lines 2 and 6 across different transitions.
 

### 2015-11-17 pldi-16/design.tex
Move a lot of stuff from i to d

@@
 Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
-Iterative Deepening likewise makes progressively deeper searches of the state space until the CPU budget is exhausted.
-In our context, the ``depth'' is the number of PPs used.
+Iterative Deepening
+%likewise
+makes progressively deeper searches of the state space until the CPU budget is exhausted.
+In this context, the depth is the number of PPs used.
+Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs
+, which we refer to as {\em jobs}.
+It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
+
 We rely on state-space estimation \cite{estimation}
-to understand which state spaces are likely to complete on time,
-in advance of actually testing each interleaving within.
-The overall purpose is to make decisions automatically about when to defer exploration of a state space,
-so an inexpert user must provide only their total CPU budget as a test parameter,
-and to enable completing appropriately-sized state spaces within that budget.
+to understand which jobs are likely to complete on time,
+before actually testing each interleaving within.
+The overall purpose is to decide automatically when to defer testing a state space,
+so an inexpert user can provide only their total CPU budget as a test parameter,
+and to enable completing appropriately-sized jobs within that budget.
 
-Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
-A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
+Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
+A MC tool is still used to test each state space, and other reduction techniques are still applicable.
 Moreover, because Iterative Deepening treats the set of preemption points as mutable,
 it can add new ones reactively based on any runtime analysis.
-In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
+We focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Initial PP configuration}
+
+Iterative Deepening must be seeded with some initial state spaces,
+which can be any number of subsets of the statically-available PPs that SSS-MC would use.
+For testing user-space code, we begin with the four subsets of locking API PPs that we showed in Figure~\ref{fig:id}:
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+and $\{yield,lock,unlock\}$,
+By extension, these also introduce PPs on any other sync primitives implemented with internal locks,
+such as condvars or semaphores.
+Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
+%so the {\tt yield} PP is always implicitly enabled.
+
+For kernel-level testing, we consider interrupt-disabling analogous to locking,
+so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
+%(to appropriate the names of the x86 instructions)
+During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
+%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
+Some kernels disable preemption without disabling interrupts,
+which can be modelled the same way using manual annotations. %of that API.
+This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
+\quicksand~is configured to begin with
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+$\{yield,cli\}$,
+$\{yield,sti\}$,
+and $\{yield,lock,$ $unlock,cli,sti\}$.
+As a heuristic, we don't test every intermediate subset such as $\{lock,sti\}$,
+which could potentially be improved in future work (\sect{\ref{sec:future}}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %
-Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
-Though its worst-case performance is $O(mn)$ in the
-%number of pending and suspended jobs,
-sizes of $\mathcal{P}$ and $\mathcal{S}$,
-in practice the non-constant portion beyond line 4 runs very infrequently
-and is negligible compared to the exponentially-sized state spaces.}.
+Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
+%\footnote{
+%Though its worst-case performance is $O(mn)$ in the
+%%number of pending and suspended jobs,
+%sizes of $\mathcal{P}$ and $\mathcal{S}$,
+%in practice the non-constant portion beyond line 4 runs very infrequently
+%and is negligible compared to the exponentially-sized state spaces.}.
 Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
@@
 
-\subsection{Classifying data races as buggy or benign}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Data-race preemption points}
 \label{sec:classifying}
 
-Iterative Deepening must be seeded with some initial state spaces,
-which can be any number of subsets of the statically-available PPs that SSS-MC would use.
-While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
+During stateless MC, runtime data-race detection may find data-race candidates that we wish to investigate further.
 Because data races indicate access pairs that can interleave at instruction granularity,
@@
 
+Furthermore, we do not always increase the number of PPs when we find a new data race.
+%When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
+For each instruction involved in a data race, \quicksand~adds two new jobs:
+a ``small'' job to preempt on that instruction only,
+and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
+%
+Hence,
+%together with the logic in \sect{\ref{sec:classifying}},
+each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
+%
+The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
+while the big job risks not completing in time,
+the small job risks missing the data race entirely if the original PPs were required to expose it.
+% TODO CAMREADY: Put numbers here.
+In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
+which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
+
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
+	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
+		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
+	\label{fig:new-dr-jobs}
+\end{figure}
+
 The new spaces may expose a failure, in which case we report a data-race bug,
@@
 in which case it remains a potential false positive that the user must handle.
+
+%When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
+%Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
+%(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
+%Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
+%Instead, \landslide~qualifies the PC with
+%(a) the current thread ID and
+%% FIXME: We don't actually do this.
+%(b) the most recent {\tt call} instruction.
+%% (a crude approximation of a stack trace)
+%% which are much cheaper, as we carry them around all the time already
+%Note that we do {\em not} qualify data races by the shared memory address,
+%which can change based on different interleavings of previous code
+%(for example, depending on the result of {\tt malloc()}).
+%% especially when malloc is involved.
+%%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Heuristics}
+% List of all heuristix:
+% HOMESTRETCH - last 60sec of test, don't suspend
+% ETA_THRESH - "to let its ETA stabilize"
+% eta factor
+% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
+% priority change between suspected and confirmed dr
+Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
+to express how optimistic we are about the estimate's accuracy.
+We use 2 as this scaling factor based on the results in \cite{estimation}.
+%though we allow changing it via the command line.
+We also include a heuristic to
+%ignore ETAs entirely
+never suspend jobs before they pass a certain threshold of interleavings tested,
+for which we choose 32,
+so that their ETAs have some time to stabilize.
+
+We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
+based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
+Jobs with both-order data-race PPs are prioritized higher,
+because single-order candidates are more likely to be false-positives
+(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
+%so we prioritize jobs with both-order data-race PPs.
+For single-order races, we do not add a PP for the later access at all;
+should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

### 2015-11-17 pldi-16/implementation.tex
Fix pictures
Move a lot of stuff from i to d

@@
 state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
-% TODO: \cite{fairstatelessmc} instead or in addition?
 It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
 % this line can be cut if space is needed
-It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
+%It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
 % joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
@@
 {\bf Restricting PPs with stack trace predicates.}
-Most model checkers (MCs) preempt indiscriminately on any sync API call, regardless of the call-site.
+Most MC tools preempt indiscriminately on any sync API call, regardless of the call-site.
 However, when testing a particular module in a large codebase,
@@
 The implementation is roughly 3000 lines of C.
-It uses a thread pool to schedule the available state spaces,
-sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
-Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
-
-{\bf Initial PPs.}
-For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs that we showed in Figure~\ref{fig:id}:
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-and $\{yield,lock,unlock\}$,
-By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
-such as condition variables, semaphores, and r/w locks.
-Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
-%so the {\tt yield} PP is always implicitly enabled.
-
-For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
-so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
-%(to appropriate the names of the x86 instructions)
-During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
-%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
-Some kernels disable preemption without disabling interrupts,
-which can be modelled the same way using manual annotations. %of that API.
-This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
-\quicksand~is configured to begin with the subsets
-$\{yield\}$,
-$\{yield,lock\}$,
-$\{yield,unlock\}$,
-$\{yield,cli\}$,
-$\{yield,sti\}$,
-and $\{yield,lock,$ $unlock,cli,sti\}$.
-As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
-which could potentially be improved in future work (\sect{\ref{sec:future}}).
-
-{\bf Communication protocol.}
-The interface to \landslide~, which any similar MC could implement, has two parts.
+%It uses a thread pool to schedule the available state spaces,
+%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
+%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
+The interface to \landslide~ has two parts. %, which any similar MC could implement, has two parts.
 First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
@@
 Should \quicksand~later resume a suspended job, we send a message to continue,
-causing the \landslide~instance to resume exploring where it left off;
-otherwise, we send a message only after time runs out, causing it to exit.
-
-{\bf Heuristics.}
-% List of all heuristix:
-% HOMESTRETCH - last 60sec of test, don't suspend
-% ETA_THRESH - "to let its ETA stabilize"
-% eta factor
-% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
-% priority change between suspected and confirmed dr
-Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
-to express how optimistic we are about the estimate's accuracy.
-We use 2 as this scaling factor based on the results in \cite{estimation},
-though we allow changing it via the command line.
-We also include a heuristic to
-%ignore ETAs entirely
-never suspend jobs before they pass a certain threshold of interleavings tested,
-for which we choose 32,
-so that their ETAs have some time to stabilize.
-
-We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
-Jobs with both-order data-race PPs are prioritized higher,
-because single-order candidates are more likely to be false-positives
-(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
-%so we prioritize jobs with both-order data-race PPs.
-For single-order races, we do not add a PP for the later access at all;
-should it be needed, preempting on the first access will suffice to upgrade the race to both-order.
-
-% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.
-
-\subsection{Data-race preemption points}
-
-When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
-Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
-Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
-Instead, \landslide~qualifies the PC with
-(a) the current thread ID and
-% FIXME: We don't actually do this.
-(b) the most recent {\tt call} instruction.
-% (a crude approximation of a stack trace)
-% which are much cheaper, as we carry them around all the time already
-Note that we do {\em not} qualify data races by the shared memory address,
-which can change based on different interleavings of previous code
-(for example, depending on the result of {\tt malloc()}).
-% especially when malloc is involved.
-%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
-	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
-		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
-	\label{fig:new-dr-jobs}
-\end{figure}
-
-When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
-a ``small'' job to preempt on the racing instruction only,
-and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
-%
-Hence, together with the logic in \sect{\ref{sec:classifying}}, each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
-%
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
-while the big job risks not completing in time,
-the small job risks missing the data race entirely if the original PPs were required to expose it.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
-% TODO: Put numbers here.
+resuming the \landslide~instance where it left off;
+otherwise, we resume it only after time runs out, causing it to exit.

### 2015-11-17 pldi-16/implementation.tex
chop some lines off impl

@@
 %even if one wishes to test for lock bugs,
-data-race analysis would flag every access pair in the lock implementation, requiring human attention to verify.
-However, Iterative Deepening can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the analysis.
+data-race analysis would flag every access pair in the lock implementation. %, requiring human attention to verify.
+However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
+Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the data race analysis.
 (Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
@@
 \landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
-whenever a new data-race candidate is found, and whenever a bug is found.
-\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
-We implement suspending jobs simply by making \landslide~wait on a message-passing call.
+and whenever it finds a new data-race candidate or bug.
+\quicksand~in turn replies whether to suspend/resume due to too high ETA, or quit due to timeout.
+We suspend jobs simply by making \landslide~wait on a message-passing call.
 Should \quicksand~later resume a suspended job, we send a message to continue,
-resuming the \landslide~instance where it left off;
-otherwise, we resume it only after time runs out, causing it to exit.
+resuming the \landslide~instance where it left off.
+%otherwise, we resume it only after time runs out, causing it to exit.

### 2015-11-17 pldi-16/design.tex
trim some lines

@@
 It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
-
 We rely on state-space estimation \cite{estimation}
@@
 which can be any number of subsets of the statically-available PPs that SSS-MC would use.
-For testing user-space code, we begin with the four subsets of locking API PPs that we showed in Figure~\ref{fig:id}:
+For testing user-space code, we begin with the four PP sets from Figure~\ref{fig:id}:
 $\{yield\}$,

### 2015-11-17 pldi-16/evaluation.tex
minor rephrasing

@@
 {\bf Finding the same bugs faster.}
-To show that Iterative Deepening is effective even for MC domains without data races, such as message-passing distributed systems,
+To show that Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
 we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
@@
 {\bf Finding new data-race bugs.}
-The line QS-DRs represents our ``no holds barred'' \quicksand~tests,
-with dynamically-discovered data-race PPs enabled:
-we quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
+The line QS-DRs represents \quicksand's bug-finding performance
+with data-race PPs enabled.
+We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
 The break-even point is at a negligible 90 seconds.
@@
 	\begin{center}
-		%\small
+		\small
 	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
@@
 
-%\begin{figure}[t]
-%	\includegraphics[width=0.48\textwidth]{bargraph.pdf}
-%	\caption{{\bf TODO:} Is this graph productive to include in the paper? Or should I just leave the numbers in table format (the left half of Table~\ref{tab:drbugs})?}
-%	\label{fig:bargraph}
-%\end{figure}
-
 %Figure~\ref{fig:bargraph}
+% TODO: deal with the table
 The left half of Table~\ref{tab:drbugs}
@@
 		Using nondeterministic data races as PPs,
-		\quicksand~found 189\% as many bugs as could be found using single-pass data races alone.
+		\quicksand~found 189\% as many data-race bugs compared to using single-pass candidates alone.
 	}

### 2015-11-17 pldi-16/soundness.tex
minor rephrasing

@@
 	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
-	\\
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & \texttt{publish(x);} & \\

### 2015-11-17 pldi-16/soundness.tex
He could put on a graph! To colored it! Q D E!

@@
-\section{Proofs (TODO: need better name)}
+\section{Soundness}
 \label{sec:soundness}
@@
 some tests may be small enough that even their maximal state spaces could be completed in time.
-For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin} would provide a total verification of all possible thread schedules, if it could complete in time.
+For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin,inspect} would provide a total verification of all possible thread schedules, if it could complete in time.
 In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
@@
 second, we show that Iterative Deepening will eventually detect all such data races.
+\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
+\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
+\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
+Given a PP $p$, let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,
+let $\ppinstr{p}$ denote the first instruction of $\ppnext{p}$,
+and let $\ppothers{p}$ denote the transitions by other threads between $p$ and $\ppnext{p}$.
 
@@
 
-TODO: sketch the proof of this lemma % TODO
+Let $p$ be the first PP in the given interleaving such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a sync API boundary.
+Because $\ppinstr{p}$ is not a sync API boundary,
+no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
+Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
+Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a sync API boundary.
+If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
+We modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
+By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
+In other words, we have transformed $p$ into $p'$ such that $\ppnext{p'} = i$, which is a data race or sync API boundary.
+All PPs in the input trace can be inductively converted in the same manner.
 
@@
 
-TODO: sketch the proof of this lemma % TODO
+We induct on the PPs according to the order of their preemptions.
+Given that the interleaving prefix preceding some PP $p$ is reachable,
+we will show that either $p$ is reachable, or a new data race among $\ppothers{p}$, not previously reachable, is now reachable.
+The latter condition suffices because in a finitely-sized program, there must be finitely many unique racing instruction pairs, so induction on the number of new data-race PPs among $\ppothers{p}$ will make $p$ itself reachable.
+
+First, we must ``coalesce'' away $p$, as well as any other not-yet-reachable PPs in $\ppothers{p}$.
+Consider the alternate interleaving in which the first thread executes past $p$ until the first already-reachable PP,
+then the other threads among $\ppothers{p}$ execute the same way.
+This interleaving's PPs are all reachable, so a state space $\mathcal{S}$ containing it will be tested.
+
+If $p$ is a not-yet-reachable data-race PP,
+it must be possible for some other thread to execute a data-racing instruction with $\ppinstr{p}$.
+If this conflict was observed in the state space containing our coalesced interleaving, we have reached $p$.
+Otherwise, we appeal to the soundness of DPOR:
+If a program behaviour is possible by interleaving threads at the boundaries of the given transitions,
+it will be tested in the containing state space.
+By contrapositive, to expose this behaviour, one or more preemptions must occur in the middle of some transition, rather than at the boundaries.
+
+We now show by contradiction there cannot be {\em multiple} data-race PPs which must both be enabled before either data race can be identified.
+Assume there does not exist a single transition $t_1 \in \mathcal{S}$ which alone can be split into $\{t_1',t_1''\}$ by a PP $q$ such that another thread's concurrent transition $t_2$ conflicts with $t_1''$.
+By the soundness of DPOR, because all $t_2$s are independent with $t_1''$, $\mathcal{S} \equiv \mathcal{S} \cup q$.
+Replacing $\mathcal{S}$ with $\mathcal{S} \cup q$ in the above assumption shows that no {\em pair} of new $q$s would expose new program behaviour, and inductively, no set of $q$s of any size, which contradicts the previous paragraph.
+
+Hence, a single new not-yet-reachable data race is reachable in $\mathcal{S}$. Hence $p$ will be reached.
 \\

### 2015-11-17 pldi-16/design.tex
minor rephrasing
TRIMTRIMTRIM. rip in pieces, author-year citations :(

@@
 it can add new ones reactively based on any runtime analysis.
-We focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.
+We focus on dynamic data-race detection~\cite{hybriddatarace,tsan} as the mechanism for finding new preemption candidates.
 
@@
 
-Furthermore, we do not always increase the number of PPs when we find a new data race.
+Furthermore, we do not always strictly increase the number of PPs when we find a new data race.
 %When \quicksand~receives a data race report, it adds two new jobs to its workqueue:

### 2015-11-17 pldi-16/evaluation.tex
TRIMTRIMTRIM. rip in pieces, author-year citations :(

@@
 The Pintos kernel project
-involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
-using provided bare-bones mutex, context-switch, and virtual memory implementations
+involves priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
+using provided mutex, context-switch, and virtual memory implementations
 \cite{pintos}.
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster.pdf}
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf}
 	\caption{Comparison of bug-finding performance
@@
 %Though for many applications of MC, assuming a correct lock implementation is sufficient,
-Though it suffices for many applications of MC to assume correctly-implemented locks,
+Though it often suffices to assume correctly-implemented locks,
 we consider this strong evidence that low-level synchronization code must be verified with data-race PPs.
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{totalverifs.pdf}
+	\includegraphics[width=0.48\textwidth]{totalverifs-squashed.pdf}
 	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
@@
 {\bf Partial verification.}
+% TODO: Figure out how to cut this paragrph.
 We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
@@
 
-Likewise, for tests where full verification is not computationally feasible,
+Likewise, when full verification is not computationally feasible,
 the state spaces associated with many data-race candidates will time out.
-We cannot guarantee that those races are false positives or benign, even though no bug was found.
+We cannot guarantee that those races are
+%false positives or
+benign, even though no bug was found.
 % FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.

### 2015-11-17 pldi-16/overview.tex
Move a lot of stuff from i to d
trim some lines
minor rephrasing
TRIMTRIMTRIM. rip in pieces, author-year citations :(

@@
-\section{Overview}
+\section{Background}
 \label{sec:overview}
@@
 Stateless model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
+A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
 Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
 stateless model checkers focus on concrete observable failures such as assertions, deadlocks, and segfaults.
-A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
 
@@
 
-Data race detection \cite{eraser} is an analysis which identifies pairs of unsynchronized memory accesses between threads.
+Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
 Two instructions are said to race if
@@
 and no synchronization enforces the order of the thread transitions.
-In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 8, and line 6 further races with 9.
+Data races are not necessarily bugs, but represent suspicious violations of the locking discipline.
+In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 8, and line 6 races with 9.
 

### 2015-11-17 pldi-16/related.tex
TRIMTRIMTRIM. rip in pieces, author-year citations :(

@@
 %However, every such prior tool we know of has serious drawbacks.
-SKI \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
+SKI \cite{ski} approaches kernel code by choosing a random set of instructions in advance, %offsets from the start of the test,
 which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
@@
 % XXX: Is this true??
-It also assumes the POSIX synchronization API, so cannot verify arbitrary synchronization algorithms such as we do with \mxtest.
-Future work could combine the two approaches, using MC to produce new data-race traces for Portend to classify, or using Portend's analysis to inform \quicksand's state space priorities.
+%It also assumes the POSIX synchronization API, so cannot verify arbitrary synchronization algorithms such as we do with \mxtest.
+Future work could combine the two approaches, using MC to produce new data-race traces for Portend to classify, or using Portend's analysis to inform \quicksand's heuristic priorities.
 

### 2015-11-17 pldi-16/soundness.tex
TRIMTRIMTRIM. rip in pieces, author-year citations :(

@@
 some tests may be small enough that even their maximal state spaces could be completed in time.
-For such tests, a SSS-MC tool configured to preempt on every shared memory access \cite{spin,inspect} would provide a total verification of all possible thread schedules, if it could complete in time.
+For such tests, preempting on every shared memory access \cite{spin,inspect} would provide a total verification of all possible thread schedules, if it could complete in time.
 In this section, we show that Iterative Deepening provides a verification of the same strength if it completes the state spaces associated with every discovered data-race PP.
@@
 We now show by contradiction there cannot be {\em multiple} data-race PPs which must both be enabled before either data race can be identified.
-Assume there does not exist a single transition $t_1 \in \mathcal{S}$ which alone can be split into $\{t_1',t_1''\}$ by a PP $q$ such that another thread's concurrent transition $t_2$ conflicts with $t_1''$.
+Assume there does not exist a single transition $t_1 \in \mathcal{S}$ which alone can be split into $\{t_1',t_1''\}$ by a PP $q$,
+such that another thread's concurrent transition $t_2$ conflicts with $t_1''$.
 By the soundness of DPOR, because all $t_2$s are independent with $t_1''$, $\mathcal{S} \equiv \mathcal{S} \cup q$.
@@
 
-We identify a particular class of false positive data-race candidate in which the associated memory is recycled by {\tt malloc} between the two accesses.
+We identify a particular class of false positive data-race candidate in which the memory is recycled by {\tt malloc} between the two accesses.
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.

### 2015-11-17 pldi-16/soundness.tex
fix formatting error

@@
 
+\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
+\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
+\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
 \begin{theorem}
@@
 second, we show that Iterative Deepening will eventually detect all such data races.
-\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
-\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
-\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
 Given a PP $p$, let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,

### 2015-11-17 pldi-16/evaluation.tex
TRIMTRIMTRIM. back on 11 pages woo!

@@
 \subsection{Test Suite}
+\label{sec:testsuite}
 Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
@@
 
-\newcommand\mxtest{\texttt{mx\_test}} % TODO: find a way to change name to mutex_test without overflowing lines in the next paragraph
+\newcommand\mxtest{\texttt{mx\_test}}
 \newcommand\tej{\texttt{join\_test}}
@@
 	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
-		% TODO: Make the "bugs" left half of this graph into a bar graph.
 		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Untested} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
@@
 		\hline
-		{\prisema	} & 59	&	&	&	&	&	& 0	&	&	\\
-		{\alarmsimul	} & 44	&	&	&	&	&	& 37	&	&	\\
-		{\waitsimple	} & 52	&	&	&	&	&	& 57	&	&	\\
+		{\prisema	} & 59	& \multicolumn{5}{c}{TODO: finish this table} & 0	&	&	\\
+		{\alarmsimul	} & 44	& \multicolumn{5}{c}{haven't yet decided how to organize this data;} & 37	&	&	\\
+		{\waitsimple	} & 52	& \multicolumn{5}{c}{suggestions welcome} & 57	&	&	\\
 		\hline
@@
 
-% TODO: maybe the whole story about partial verification can be cut
 {\bf Partial verification.}
@@
 \end{figure}
-
-% TODO: Figure out how to visualize how many unique data race PPs were verified, and how many remained as unclassified false positives
 
@@
 Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
-Because we prioritize completing small state spaces before extending them with more PPs,
-the superset state spaces we run later will repeat each branch of their already-completed subsets.
+%Because we prioritize completing small state spaces before extending them with more PPs,
+%the superset state spaces we run later will repeat each branch of their already-completed subsets.
 %
@@
 %on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
-
-This repeated work can cause Iterative Deepening to be slower than SSS-MC to find certain bugs,
-for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} preemptions together would expose a bug, but not either alone.
-Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
-if the state space resulting from preempting on every instruction could be completed,
-an SSS-MC tool such as \cite{spin} might achieve that verification faster,
-as Iterative Deepening will test many subsets of data-race PPs first.
-Predicting whether each still-untested interleaving will uncover new PPs is not straightforward,
+This may make Iterative Deepening slower than SSS-MC to find certain bugs,
+for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} PPs together expose a bug, but not either alone.
+%Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
+%if the state space resulting from preempting on every instruction could be completed,
+%an SSS-MC tool such as \cite{spin} might achieve that verification faster,
+%as Iterative Deepening will test many subsets of data-race PPs first.
+Predicting whether each future interleaving has been tested before is not straightforward,
 but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
-to prune some or all of this repeated work.
+to prune some or all such repeated work.
 
@@
 and in 3 other tests, no state spaces could be completed at all.
-Larger state spaces often result from finer-grained locking,
-which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
-Such programs may require even more rigorous verification than a program with a single global lock.
+%Larger state spaces often result from finer-grained locking,
+%which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
+%Such programs may require even more rigorous verification than a program with a single global lock.
 %Hence these corner cases are important to consider for future work.
-While this work used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
-we envision future Iterative Deepening implementations could incorporate this method to {\em dynamically} subset PPs further,
+While we used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
+%we envision
+future
+%Iterative Deepening
+implementations could use this mechanism to {\em dynamically} subset PPs further,
 making partial verification of such large tests possible.
@@
 {\bf Partial verification.}
-% TODO: Figure out how to cut this paragrph.
-We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
-While we guarantee safety when preempted on certain combinations of PPs,
+%We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
+While we guarantee safety when preempted on certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
 CHESS
@@
 %according to the maximum bound reached in the time limit.
-We imagine these two guarantees could be each be useful to developers in different scenarios,
-and are presently working to combine the two approaches to provide both at once.
-One benefit of our technique is that {\tt within\_function}-based Iterative Deepening (discussed above)
-would enable expert developers to configure custom subsets of PPs they are most interested in verifying,
-according to which modules of a codebase they wish to test.
+%We imagine
+These two guarantees could be each be useful to developers in different scenarios,
+and future work could combine the two approaches to provide both at once.
+One benefit of our technique is that {\tt within\_function} %-based Iterative Deepening (discussed above)
+would enable expert developers to
+%configure custom subsets of PPs they are most interested in verifying,
+%according to which modules of a codebase they wish to test.
+restrict Iterative Deepening to only the modules of a codebase they wish to test.
 
 Likewise, when full verification is not computationally feasible,
-the state spaces associated with many data-race candidates will time out.
+some jobs with data-race PPs will time out.
 We cannot guarantee that those races are

### 2015-11-17 pldi-16/implementation.tex
TRIMTRIMTRIM. rip in pieces, author-year citations :(
TRIMTRIMTRIM. back on 11 pages woo!

@@
 state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
+It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
 % this line can be cut if space is needed
@@
 {\bf Data races in lock implementations.}
-Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses resulting from the lock implementation itself.
+Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses therein. %resulting from the lock implementation itself.
 The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
@@
 However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with a custom option to change the lock-set tracking to include accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()} in the data race analysis.
-(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
+Hence, we extended \landslide~with an option to make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
+which we use in \sect{\ref{sec:testsuite}}.
+%(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
 

### 2015-11-17 pldi-16/soundness.tex
TRIMTRIMTRIM. back on 11 pages woo!

@@
 
-By definition of DPOR, this PP causes {\tt x2} to be reordered before {\tt x1} while not changing {\tt x2}'s location.
+With this PP, DPOR will reorder {\tt x2} before {\tt x1}, while not changing {\tt x2}'s location.
 As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
@@
 
-Note especially that our reasoning does not require PPs on the internal locking of {\tt malloc} or {\tt free},
+Note that we do not require PPs on the internal locking of {\tt malloc} or {\tt free},
 which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.

### 2015-11-17 pldi-16/evaluation.tex
split discussion into own tex file

@@
 %         //      (Cite eraser, section 2.2)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion}
-\label{sec:future}
-
-In this section, we discuss \quicksand's limitations and opportunities for future improvement.
-
-% TODO CAMREADY: Measure how much subset overlap there is.
-{\bf Avoiding redundant work.}
-When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
-Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
-%Because we prioritize completing small state spaces before extending them with more PPs,
-%the superset state spaces we run later will repeat each branch of their already-completed subsets.
-%
-%We measured the proportion of repeated work among completed state spaces across our test suite;
-%on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
-This may make Iterative Deepening slower than SSS-MC to find certain bugs,
-for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} PPs together expose a bug, but not either alone.
-%Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
-%if the state space resulting from preempting on every instruction could be completed,
-%an SSS-MC tool such as \cite{spin} might achieve that verification faster,
-%as Iterative Deepening will test many subsets of data-race PPs first.
-Predicting whether each future interleaving has been tested before is not straightforward,
-but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
-to prune some or all such repeated work.
-
-{\bf Finer-grained PP subsets.}
-% TODO: update these numbers
-\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
-However, in 2 tests, no more than the minimal state space could be verified,
-and in 3 other tests, no state spaces could be completed at all.
-%Larger state spaces often result from finer-grained locking,
-%which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
-%Such programs may require even more rigorous verification than a program with a single global lock.
-%Hence these corner cases are important to consider for future work.
-While we used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
-%we envision
-future
-%Iterative Deepening
-implementations could use this mechanism to {\em dynamically} subset PPs further,
-making partial verification of such large tests possible.
-%enabling partial verification of such large tests. % if need space
-%Because our current implementation does not avoid repeated interleavings across state spaces,
-%as discussed above,
-%we were limited to a small number of very basic PP subsets to statically seed the exploration
-
-%{\bf Small test cases.}
-
-{\bf Partial verification.}
-%We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
-While we guarantee safety when preempted on certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
-CHESS
-guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.
-%according to the maximum bound reached in the time limit.
-%We imagine
-These two guarantees could be each be useful to developers in different scenarios,
-and future work could combine the two approaches to provide both at once.
-One benefit of our technique is that {\tt within\_function} %-based Iterative Deepening (discussed above)
-would enable expert developers to
-%configure custom subsets of PPs they are most interested in verifying,
-%according to which modules of a codebase they wish to test.
-restrict Iterative Deepening to only the modules of a codebase they wish to test.
-
-Likewise, when full verification is not computationally feasible,
-some jobs with data-race PPs will time out.
-We cannot guarantee that those races are
-%false positives or
-benign, even though no bug was found.
-% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
-In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test program.
-For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
-which could be combined with Iterative Deepening in future work.
-% Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-11-17 pldi-16/warpzone.tex
split discussion into own tex file

\section{Discussion}
\label{sec:future}

In this section, we discuss \quicksand's limitations and opportunities for future improvement.

% TODO CAMREADY: Measure how much subset overlap there is.
{\bf Avoiding redundant work.}
When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
%Because we prioritize completing small state spaces before extending them with more PPs,
%the superset state spaces we run later will repeat each branch of their already-completed subsets.
%
%We measured the proportion of repeated work among completed state spaces across our test suite;
%on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
This may make Iterative Deepening slower than SSS-MC to find certain bugs,
for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} PPs together expose a bug, but not either alone.
%Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
%if the state space resulting from preempting on every instruction could be completed,
%an SSS-MC tool such as \cite{spin} might achieve that verification faster,
%as Iterative Deepening will test many subsets of data-race PPs first.
Predicting whether each future interleaving has been tested before is not straightforward,
but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
to prune some or all such repeated work.

{\bf Finer-grained PP subsets.}
% TODO: update these numbers
\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
However, in 2 tests, no more than the minimal state space could be verified,
and in 3 other tests, no state spaces could be completed at all.
%Larger state spaces often result from finer-grained locking,
%which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design.
%Such programs may require even more rigorous verification than a program with a single global lock.
%Hence these corner cases are important to consider for future work.
While we used {\tt within\_function} (\sect{\ref{sec:landslide}}) {\em statically} to restrict where PPs could arise in advance of the test,
%we envision
future
%Iterative Deepening
implementations could use this mechanism to {\em dynamically} subset PPs further,
making partial verification of such large tests possible.
%enabling partial verification of such large tests. % if need space
%Because our current implementation does not avoid repeated interleavings across state spaces,
%as discussed above,
%we were limited to a small number of very basic PP subsets to statically seed the exploration

%{\bf Small test cases.}

{\bf Partial verification.}
%We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
While we guarantee safety when preempted on certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
CHESS
guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.
%according to the maximum bound reached in the time limit.
%We imagine
These two guarantees could be each be useful to developers in different scenarios,
and future work could combine the two approaches to provide both at once.
One benefit of our technique is that {\tt within\_function} %-based Iterative Deepening (discussed above)
would enable expert developers to
%configure custom subsets of PPs they are most interested in verifying,
%according to which modules of a codebase they wish to test.
restrict Iterative Deepening to only the modules of a codebase they wish to test.

Likewise, when full verification is not computationally feasible,
some jobs with data-race PPs will time out.
We cannot guarantee that those races are
%false positives or
benign, even though no bug was found.
% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test program.
For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
which could be combined with Iterative Deepening in future work.
% Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-11-17 pldi-16/evaluation.tex
Fix the story about partial verification. Fill out the table, mostly.

@@
 and \waitsimple, for process lifecycle system calls\footnote{
-	Some of the Pintoses were partially implemented, so each test program could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
+	Some of the Pintoses were partially implemented,
+	so each test could only be run on a subset of the 78 submissions; see ``Number tested'' in Table~\ref{tab:drbugs}.
 }
@@
 		\small
-	\begin{tabular}{r|c||c|c|c||c|c|c|c|c}
-		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Total} & {\bf Untested} & {\bf Malloc-} & {\bf Nondet.} & {\bf Nondet.} \\
-		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR PPs} & {\bf DR PPs} & {\bf recycle DRs} & {\bf DR PPs} & {\bf DR bugs} \\
+	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
+		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
+		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf recycle DRs} &
+		{\bf time-outs} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
 		\hline
-		{\bct		} & 79	& 7	& 11	& 4	&	&	& 65	&	& 4	\\
-		{\tej		} & 79	& 14 	& 25	& 11	&	&	& 471	&	& 6	\\
-		{\mxtest	} & 79	& 1	& 10	& 9	&	&	& 7	&	& 1	\\
-		{\paradise	} & 79	& 10	& 17	& 7	&	&	& 268	&	& 3	\\
-		{\paraguay	} & 79	& 4	& 12	& 9	&	&	& 188	&	& 0	\\
-		{\rwldgr	} & 79	& 26	& 30	& 3	&	&	& 157	&	& 3	\\
+				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
+		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 65	& 0	& -	&	&	\\
+		{\tt join} 	& 79	& 14 	& 25	& 10	& 6	& 471	& 5	& 60.8	&	&	\\
+		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	&	&	\\
+		{\tt sem} 	& 79	& 10	& 17	& 6	& 5	& 268	& 29	& 73.3	&	&	\\
+		{\tt signal} 	& 79	& 4	& 12	& 7	& 6	& 188	& 10	& 54.2	&	&	\\
+		{\tt rwlock} 	& 79	& 26	& 30	& 2	& 2	& 157	& 19	& 29.6	&	&	\\
 		\hline
-		{\prisema	} & 59	& \multicolumn{5}{c}{TODO: finish this table} & 0	&	&	\\
-		{\alarmsimul	} & 44	& \multicolumn{5}{c}{haven't yet decided how to organize this data;} & 37	&	&	\\
-		{\waitsimple	} & 52	& \multicolumn{5}{c}{suggestions welcome} & 57	&	&	\\
+		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	&	&	\\
+		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 37	& 22	& 9.1	&	&	\\
+		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 57	& 8	& 28.0	&	&	\\
 		\hline
-		{\bf Total}	& 629	&	&	&	&	&	& 1250	&	&
+		{\bf Total}	& 629	& 91	& 163	& 68	& 32	& 1250	& 93	& 42.6 	&	&	\\
 	\end{tabular}
 	\end{center}
-	\caption{Summary of number and types of bugs found by each test program.
+	\caption{Summary of bugs and data races found by each test program.
 		SSS-MC is the control; QS is \quicksand.
-		``DR-only'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}}).
-	``Nondeterministic DRs'' counts among the DR-only bugs how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
-		TODO: more explanation}
+		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
+	``Malloc-recycle DRs'' counts how many false positives we suppressed (\sect{\ref{sec:eval-falsepos}}).
+	``Mutual time-outs'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
+	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
+	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests with no bugs found;
+	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
+		}
 	\label{tab:drbugs}
@@
 When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
-Beyond using the state space estimator to guess the percent coverage,
-we know of no approach to quantify the probability
-that a bug would be exposed by an untested interleaving.
-\quicksand~does not attempt any such quantitative guarantee,
-but can at least report which subsets of PPs produced state spaces that did complete in time.
-
-On 77 tests, the SSS-MC control experiment timed out in 10 hours with no bugs found.
-Among these, \quicksand~found data-race bugs in 15.
-For the other 62, we counted the number of state spaces \quicksand~was able to complete and the number of unique PPs among them.
-In three cases, \quicksand~also failed to complete anything; beyond these,
-between 1 and 253 state spaces were completed, with a mean of 58.9 and median of 29.
-Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
+%Beyond using the state space estimator to guess the percent coverage,
+We know of no approach to quantify the probability
+that a bug would be exposed by an untested interleaving,
+but \quicksand~at least reports which subsets of PPs produced state spaces that did complete in time.
+%\quicksand~does not attempt any such quantitative guarantee,
+%but can at least report
+On 129 tests, the SSS-MC control experiment timed out in 10 hours with no bugs found.
+Among these, \quicksand~found bugs in 36.
+For the other 93, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
+%and the number of unique PPs among them.
+%In 6 cases, \quicksand~also failed to complete anything; beyond these,
+%between 1 and 253 state spaces were completed for each test
+%Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
 These completions guarantee that, if the test program could expose a bug,
@@
 
-Among the 69 true data-races which initially fit the malloc-recycle pattern,
+Among those 69 true data-races, %which initially fit the malloc-recycle pattern,
 none of them exposed a new bug when used as a PP.

### 2015-11-17 pldi-16/paper.tex
begin major swizzling around of sections according to joan's suggestion
Write overview section
TRIMTRIMTRIM. rip in pieces, author-year citations :(
split discussion into own tex file
Fix the story about partial verification. Fill out the table, mostly.

@@
 \newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{orange}{RGB}{192,96,0}
 \definecolor{olivegreen}{RGB}{0,127,0}
 \definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,255}
+\definecolor{commentblue}{RGB}{0,0,192}
 
@@
 %with data-race preemption points
-found 79\% more bugs than the best prior stateless model checking approach.
+found 1.79x as many bugs and verified 4.25x as many tests compared to stateless model checking without data-race preemption points.
 %Our evaluation shows this technique is
@@
 \input{intro}
-\input{di}
+\input{overview}
+\input{design}
+\input{implementation}
+\input{soundness}
 \input{evaluation}
+\input{warpzone}
 \input{related}
@@
 
-We have presented Iterative Deepening and \quicksand, a new technique and tool for automating the choice of PPs during stateless model checking.
+We have presented Iterative Deepening and \quicksand, a new technique and tool for automating the choice of preemption points (PPs) during stateless model checking.
 % , and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
@@
 
-Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
+Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking alone.
 %We dramatically reduce false-positive data race reports,
@@
 %and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
-Compared to SSS-MC,
+%Compared to prior work,
 %, we find bugs faster in smaller subset state spaces, %% not true, or at least, not by very much
 %when the maximal state space is too large to explore
-we find new bugs with data-race PPs that the maximal state space would not expose at all.
+We find new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
 Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide an equally strong verification as if every single instruction had been used as a PP.

### 2015-11-17 pldi-16/warpzone.tex
Fix the story about partial verification. Fill out the table, mostly.

@@
 {\bf Finer-grained PP subsets.}
-% TODO: update these numbers
-\quicksand~was able to partially guarantee safety in 97\% of tests with large maximal state spaces.
-However, in 2 tests, no more than the minimal state space could be verified,
-and in 3 other tests, no state spaces could be completed at all.
+\quicksand~was able to partially guarantee safety for some PPs in 89\% of tests with too-large maximal state spaces.
+However, in 4 tests, no more than the minimal state space could be verified,
+and in 6 other tests, no state spaces were completed at all.
 %Larger state spaces often result from finer-grained locking,
@@
 %We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-sssmc}}).
-While we guarantee safety when preempted on certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
-CHESS
+While we guarantee safety when using certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
+Iterative Context Bounding
 guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.
@@
 %We imagine
-These two guarantees could be each be useful to developers in different scenarios,
+These guarantees could each be useful to developers in different scenarios,
 and future work could combine the two approaches to provide both at once.
@@
 % FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
-In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we count the number of such unverified candidates from each test program.
+In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we show how many such candidates we could not verify.
 For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},

### 2015-11-18 pldi-16/proof/di.tex
improve proof defintions

@@
 \subsection{Stateless model checking terms}
+
+\begin{definition}[Preemption point (PP)]
+	A PP is a code location between two instructions at which we may force threads to switch.
+\end{definition}
+
+In the main paper, we identified PPs by predicates on the instruction pointer and current thread ID.
+Hence, the ``same'' PP may occur multiple times during an execution; for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
+For the sake of this proof, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
 
@@
 \end{definition}
+
+\begin{definition}[Independent transitions]
+Two transitions between different threads are independent if the intersection of their shared memory accesses contains no conflicts.
+\end{definition}
+
+\begin{definition}[Equivalent interleaving]
+Two interleavings are equivalent if one can be transformed into the other by permuting only independent transitions.
+\end{definition}
+
 Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
+All possible interleavings of a program can be partitioned into equivalence classes,
+so only one interleaving from each equivalence class need be tested to ensure total schedule coverage \cite{mazurkiewicz}.
 

### 2015-11-18 pldi-16/evaluation.tex
Finish the table. Paper can now be called 100% complete, in a pinch.

@@
 		\hline
+		% FIXME: update FRMs
 				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
-		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 65	& 0	& -	&	&	\\
-		{\tt join} 	& 79	& 14 	& 25	& 10	& 6	& 471	& 5	& 60.8	&	&	\\
-		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	&	&	\\
-		{\tt sem} 	& 79	& 10	& 17	& 6	& 5	& 268	& 29	& 73.3	&	&	\\
-		{\tt signal} 	& 79	& 4	& 12	& 7	& 6	& 188	& 10	& 54.2	&	&	\\
-		{\tt rwlock} 	& 79	& 26	& 30	& 2	& 2	& 157	& 19	& 29.6	&	&	\\
+		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
+		{\tt join} 	& 79	& 14 	& 25	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
+		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
+		{\tt sem} 	& 79	& 10	& 17	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
+		{\tt signal} 	& 79	& 4	& 12	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
+		{\tt rwlock} 	& 79	& 26	& 30	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
 		\hline
-		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	&	&	\\
-		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 37	& 22	& 9.1	&	&	\\
-		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 57	& 8	& 28.0	&	&	\\
+		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
+		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
+		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
 		\hline
-		{\bf Total}	& 629	& 91	& 163	& 68	& 32	& 1250	& 93	& 42.6 	&	&	\\
+		{\bf Total}	& 629	& 91	& 163	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
 	\end{tabular}
@@
 %Figure~\ref{fig:bargraph}
-% TODO: deal with the table
 The left half of Table~\ref{tab:drbugs}
@@
 In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
-In total, 1250 data-races fit the malloc-recycle pattern across all tests,
+In total, 902 data-races fit the malloc-recycle pattern across all tests,
 only 69 of which were observed to avoid the re-allocation in an alternate interleaving.
-Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 1181 other state spaces.
+Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 833 other state spaces.
 

### 2015-11-18 pldi-16/warpzone.tex
Finish the table. Paper can now be called 100% complete, in a pinch.

@@
 some jobs with data-race PPs will time out.
-We cannot guarantee that those races are
+We cannot guarantee those races are
 %false positives or
 benign, even though no bug was found.
-% FIXME: You need a separate table counting only the state spaces WITHOUT BUGES.
-In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we show how many such candidates we could not verify.
+In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we show how many such candidates we could not verify (32\%).
 For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},
-which could be combined with Iterative Deepening in future work.
+which could be combined with \quicksand~in future work.
 % Future work: Add parallel DPOR so you can fill your spare CPUs when there are fewer than the max number of jobs.

### 2015-11-18 pldi-16/evaluation.tex
minor touchup

@@
 	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
-	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests with no bugs found;
+	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
 	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).

### 2015-11-18 pldi-16/design.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 	}{
-		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$))
+		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
+	}
+	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{\alpha\}$}{
+		AddNewJob($\{\alpha\}$, HeuristicPriority($\alpha$))
 	}
@@
 For each instruction involved in a data race, \quicksand~adds two new jobs:
-a ``small'' job to preempt on that instruction only,
-and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
+a ``small'' job to preempt on that instruction only (line 5),
+and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job (line 2).
 %
@@
 
-The new spaces may expose a failure, in which case we report a data-race bug,
+The new state spaces may expose a failure, in which case we report a data-race bug,
 or complete successfully, which indicates a benign or false-positive data race.

### 2015-11-18 pldi-16/implementation.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 % this line can be cut if space is needed
-%It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
+It can test both user-level and kernel-level code, although is limited to timer nondeterminism.
 % joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
-Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.
+Its bug-detection metrics are assertion failure, deadlock, segfault, heap checking \cite{valgrind}, and a heuristic infinite loop check.
 

### 2015-11-18 pldi-16/intro.tex
trim some lines
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 %
-On the other hand, Inspect \cite{inspect} %SPIN \cite{spin}
-is able to preempt threads around any shared memory access.
+On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
+are able to preempt threads around any shared memory access.
 Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for even modestly-sized tests.
@@
 
-The remainder of the paper is organized as follows.
+The rest of the paper is organized as follows.
 \sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,

### 2015-11-18 pldi-16/overview.tex
TRIMTRIMTRIM. back on 11 pages woo!
Finish the table. Paper can now be called 100% complete, in a pinch.
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees-v2.pdf}
+	\includegraphics[width=0.48\textwidth]{trees-v2-squashed.pdf}
 	\caption{Iterative Deepening example.
@@
 The checker defins the granularity of thread interleavings by which {\em preemption points} it uses to switch threads.
-Most model checkers \cite{chess} choose synchronization/thread library API boundaries for these points;
+Most model checkers \cite{chess,dbug-ssv} choose synchronization/thread library API boundaries for these points;
 in our example program, these would be lines 1, 4, and 7.
@@
 
-This motivates {\em Iterative Deepening}, our new technique for adjusting the preemption points at runtime.
+This motivates {\em Iterative Deepening}, our new technique for heuristically adjusting the preemption points at runtime.
 Rather than committing to one state space with every available preemption point enabled,
@@
 
+Prior work \cite{hybriddatarace} distinguishes between {\em pure happens-before} \cite{lamport-clocks},
+in which accesses with a lock release/acquire in between are not considered concurrent,
+and {\em limited happens-before},
+in which only blocking operations like {\tt cond\_wait} or {\tt join} enforce the order.
+Compared to the former, the latter may output false positives, but avoids false negatives \cite{tsan},
+which is more appropriate for this work (\sect{\ref{sec:totalverif}}).
+
 \subsection{Terminology}
@@
 %Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
-Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit},
+Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit,smc-empirical-study},
 but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.
 
-We distinguish between data-race {\em candidates} and data-race {\em bugs}.
+Although all data races are undefined behaviour in C and C++ \cite{boehm-memorymodels},
+we distinguish between data-race {\em candidates} and data-race {\em bugs}.
 Because data-race analysis is prone to false-positives,

### 2015-11-18 pldi-16/paper.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 %Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
-\quicksand~is open-source and its interface can be adapted to fit any MC tool similar to \landslide.
-We have also posted our evaluation's full data set and logs.
-They can be found here:
+\quicksand~is open-source and its interface can be adapted to fit any tool similar to \landslide.
+We have also posted our evaluation's full data set. %and logs.
+They are available at:
 
@@
 
-\section{Acknowledgements}
+%\section{Acknowledgements}
 
 % TODO CAMREADY
-{\em removed for double-blind review}
+%{\em removed for double-blind review}
 %Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
@@
 %Thanks to
-%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, John Wilkes, and
+%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, John Wilkes, and
 %the anonymous PLDI reviewers for their helpful comments.

### 2015-11-18 pldi-16/related.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 
-{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
+{\bf Other techniques.} Various improvements or alternatives to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, DPOR for relaxed memory models \cite{tsopso}, and SAT-directed MC \cite{satcheck}.
 These are all orthogonal to our technique.
@@
 %providing a strong argument for \quicksand.
-Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to Iterative Deepening,
-as both approaches provide a partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
+Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to our work,
+as both approaches provide a partial verification
+%on some subset of interleavings
+when full completion is intractable (\sect{\ref{sec:future}}).
 However, ICB is limited to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
@@
 \cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
-% TODO: any more?
-DroidRacer \cite{droidracer} and CAFA \cite{cafa} extend the analysis to event-driven Android applications, using domain-specific heuristics (orthogonal to our method) to reduce false positives.
-% cut for space?
+DroidRacer \cite{droidracer} and CAFA \cite{cafa} target
+%event-driven
+Android applications, using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
+Like \landslide, DataCollider \cite{datacollider} offers data-race techniques for kernel code.
+IFRit \cite{ifrit}
+improves performance using an interference analysis,
+which would allow future work to avoid tracing every memory access.
+%although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
 % No, IDGAF about pure happens before.
@@
 It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
-while we explore the latter only.
-Portend additionally reports non-failing races which nevertheless cause
+%while we explore the latter only.
+and additionally reports non-failing races which nevertheless cause
 %suspiciously
-different program output, which is orthogonal to our technique.
+different program output. %, which is orthogonal to our technique.
 However, Portend does not test alternate interleavings {\em in advance} of knowing data races,
 which is necessary to expose certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
-% XXX: Is this true??
+% Not really true.
 %It also assumes the POSIX synchronization API, so cannot verify arbitrary synchronization algorithms such as we do with \mxtest.

### 2015-11-18 pldi-16/soundness.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 \begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
+	%& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
+	%& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
 	& {\bf Thread 1} & {\bf Thread 2} \\

### 2015-11-18 pldi-16/testcases/testcases.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops

@@
 \begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
+In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor,chess}.
 Our approach involves adding new state spaces to explore each time a new data-race candidate is found.

### 2015-11-18 pldi-16/evaluation.tex
implement brandon's comments, and trim a bit; back onto 12th page, oops
trimtrimtrim

@@
 
-Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone in other model checkers.
-For example, a single-state-space tool could use data-race candidates during immediately subsequent interleavings, essentially changing the state space on the fly.
-Likewise, a message-passing-only tool could employ Iterative Deepening despite data races being absent from its concurrency model.
+Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone as well. %in other model checkers.
+For example, a single-state-space tool could use data-race PPs during immediately subsequent interleavings,
+%essentially
+changing the state space on the fly.
+Likewise, a message-passing-only tool could use Iterative Deepening despite a concurrency model lacking data races. % being absent from its concurrency model.
 Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
-we also sought to evaluate each technique individually.
+we also evaluated each technique individually.
 Our evaluation answers the following questions:
@@
 \label{sec:eval-falseneg}
+%For the [NONDET] expt, you don't need to additionally count how many DRs
+%needed to be found in a DR-PP state space, because when EXPLORE_BACKWARDS=0,
+%you'll need to preempt on the DR PP to find the new DR anyway, hence they will
+%be nondet. Unless you want to separately count the subset of nondet drs that
+%are also dr-pp-ss-only (the reviewers may ask for this before cam-ready?).
 
@@
 
+%When proving that [NONDET] dr buges exist, make sure to mention that, although
+%their frequency varies depending on test case (mx test: almost none; bct:
+%almost all), they are still PRESENT in all (or almost all) test cases, meaning
+%it is not just a matter of writing better test cases.
+
 Note that we are not comparing how much testing time is required before identifying the data-race candidates involved in each bug.

### 2015-11-18 pldi-16/intro.tex
trimtrimtrim

@@
 by forcing a program to execute each distinct interleaving,
-capturing and controlling this nondeterminism using a finite state space.
+capturing
+%and controlling
+this nondeterminism in a finite state space.
 Unfortunately, these state spaces explode exponentially in the size of the input program.
 Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} allow bugs to be found sooner in a given exploration should they exist.
+and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} allow bugs to be found sooner in a given space should they exist.
 
@@
 We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
-We show many advantages over both conventional stateless model checking and single-pass data-race analysis.
-We find that many of the bugs found by the conventional approach can be found faster using subsets of preemption points,
+%We show many advantages over both conventional stateless model checking and single-pass data-race analysis.
+We find many of the same bugs found by the conventional approach faster using subsets of preemption points,
 and that data-race preemption points quickly expose many new bugs that prior model checkers could not find at all.

### 2015-11-18 pldi-16/overview.tex
trimtrimtrim

@@
 	7 & \texttt{\hilight{commentblue}{mutex\_unlock(\&mx);}} & \\
-	8 & \texttt{assert(x >= 1);}  & \\
-	9 & & \texttt{assert(x >= 2);} \\
+	8 & & \texttt{assert(x >= 2);} \\
 	%1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
@@
 \caption{Example program with a data-race bug.
-In this interleaving, the assertion on line 9 will fail.
-Two data-race preemptions are required to expose the bug: one just before thread 1's line 6, and one just before thread 2's line 9.}
+In this interleaving, the assertion on line 8 will fail.
+Two data-race preemptions are required to expose the bug: one just before thread 1's line 6, and one just before thread 2's line 8.}
 \label{fig:example}
@@
 Data races are not necessarily bugs, but represent suspicious violations of the locking discipline.
-In Figure~\ref{fig:example}, lines 3 and 5 each race with 2, 6, and 8, and line 6 races with 9.
+In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
 
@@
 {\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
-Dynamic Partial Order Reduction (DPOR), and {\em state space estimate} (ETA).
+{\em Dynamic Partial Order Reduction} (DPOR), and {\em state space estimate} (ETA).
 
-SSS-MC indicates the approach of prior model checking tools:
+SSS-MC indicates the approach of prior tools:
 the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
 %Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
-Many techniques exist to skip equivalent interleavings and order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit,smc-empirical-study},
+Many techniques can skip equivalent interleavings or order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit,smc-empirical-study},
 but new PPs cannot be added, nor ineffective ones removed, by any dynamic analysis.

### 2015-11-19 pldi-16/design.tex
garth's comments, pt 1.

@@
 We rely on state-space estimation \cite{estimation}
-to understand which jobs are likely to complete on time,
-before actually testing each interleaving within.
-The overall purpose is to decide automatically when to defer testing a state space,
+to predict which jobs are likely to complete within a reasonable time,
+before actually testing each interleaving.
+The overall goal is to decide automatically when to defer testing a state space,
 so an inexpert user can provide only their total CPU budget as a test parameter,
@@
 Moreover, because Iterative Deepening treats the set of preemption points as mutable,
-it can add new ones reactively based on any runtime analysis.
-We focus on dynamic data-race detection~\cite{hybriddatarace,tsan} as the mechanism for finding new preemption candidates.
+it can add new preemption points reactively based on any runtime analysis.
+We focus on run-time data-race detection~\cite{hybriddatarace,tsan,ifrit} as the mechanism for finding new preemption candidates.
 
@@
 
-Iterative Deepening must be seeded with some initial state spaces,
-which can be any number of subsets of the statically-available PPs that SSS-MC would use.
+Iterative Deepening must be seeded with a set of initial state spaces,
+which can be any number of subsets of the statically-available PPs SSS-MC would use.
 For testing user-space code, we begin with the four PP sets from Figure~\ref{fig:id}:
@@
 and a set of {\em pending} jobs (untested ones with unknown ETAs).
-When a MC instance reports an ETA too high,
-we compare other pending jobs to find another one more likely to complete in time.
+When the MC reports an ETA too high for some job,
+we compare with other pending and suspended jobs to find another one more likely to complete in time.
 %
@@
 then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
+Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
 %reveals that it might finish in time after all.
-Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
+Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
 
@@
 
-With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
+With Iterative Deepening, this is a simple matter of creating a new state space with an additional PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
 Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
@@
 The new state spaces may expose a failure, in which case we report a data-race bug,
-or complete successfully, which indicates a benign or false-positive data race.
+or complete successfully, indicating a benign or false-positive data race.
 They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
-in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
+in which case we may iteratively advance to a superset state space containing PPs for both racing access pairs.
 %Because Iterative Deepening is
@@
 we may time out before completing a data race's associated state space,
-in which case it remains a potential false positive that the user must handle.
+in which case we report a potential false positive that the user must handle (\sect{\ref{sec:future}}).
 

### 2015-11-19 pldi-16/intro.tex
garth's comments, pt 1.

@@
 However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
-Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion.
 For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
@@
 On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
-are able to preempt threads around any shared memory access.
-Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for even modestly-sized tests.
+are able to preempt threads before any shared memory access.
+Such fine granularity would automatically check each data race for the possibility of failure, but risks timing out before the state space can be completed.
+%makes full state space completion intractable for even modestly-sized tests.
+Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion,
+but even with state-of-the-art reduction techniques,
+any choice made in advance must necessarily leave some tests impossibly large \cite{parrot,mcr}.
 %
-This work shows how to avoid making that tradeoff decision in advance.
+%This work shows how to avoid making that tradeoff decision in advance.
 
@@
 according to which resulting state spaces are most likely to fit a prescribed CPU budget.
-It uses data-race analysis to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
+It uses data-race analysis \cite{eraser} to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
 When prior work would time out on large tests by trying several preemption points simultaneously,

### 2015-11-19 pldi-16/overview.tex
garth's comments, pt 1.

@@
 We review the background in stateless model checking and data-race analysis using the example program in Figure~\ref{fig:example}.
+
+\subsection{Stateless Model Checking}
+
+Stateless model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
+A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
+Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
+stateless model checkers focus on concrete observable failures such as assertions, deadlocks, and segfaults.
 
@@
 
-\subsection{Stateless Model Checking}
-
-Stateless model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
-A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
-Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
-stateless model checkers focus on concrete observable failures such as assertions, deadlocks, and segfaults.
-
 \begin{figure}[t]
@@
 and tests at least one execution from each equivalence class.
-Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving.
+Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
 Nevertheless, state spaces are still exponentially-sized in the number of conflicting transitions.
@@
 for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-Hence, we will test all the state spaces shown in Figure~\ref{fig:id} in parallel,
-and decide on-the-fly whether to pursue each test, or to defer it in favour of smaller ones.
+Hence, we will test all the state spaces illustrated in Figure~\ref{fig:id} in parallel,
+and decide on-the-fly whether to pursue each test, or to defer it in favour of others.
 
@@
 Because data-race analysis is prone to false-positives,
-we classify unprotected access pairs separately from bugs,
+we classify unprotected memory access pairs separately from bugs,
 calling such pairs {\em data race candidates}.
-Should a future interleaving, preempting during those accesses,
+Should a future interleaving, preempting during such accesses,
 lead to a failure, then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
-If they cannot be reordered at all, it is a {\em false positive}.
+Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
+If they cannot be reordered at all, due to some other communication, it is a {\em false positive}.
 %TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.

### 2015-11-19 pldi-16/paper.tex
garth's comments, pt 1.

@@
 \begin{abstract}
-Stateless model checking is a promising technique for testing concurrent programs,
-but suffers from exponential explosion of the state space when the test input parameters are too large.
+Stateless model checking is a powerful technique for testing concurrent programs,
+but suffers from exponential state space explosion when the test input parameters are too large.
 Several reduction techniques can mitigate this explosion,

### 2015-11-19 pldi-16/related.tex
trimtrimtrim
garth's comments, pt 1.

@@
 
-{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, and ETA \cite{dbug-retreat}, limit thread communication to a message-passing API to more effectively test distributed systems.
+{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, ETA \cite{dbug-retreat}, and Concuerror \cite{optimal-dpor},
+limit thread communication to a message-passing API to more effectively test distributed systems.
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
@@
 while \landslide~\cite{landslide} traces the memory locations of accesses at runtime.
-Both SPIN and Inspect fix their set of PPs in advance, so could be combined with \quicksand~in future work.
+Both SPIN and Inspect fix their set of PPs in advance, so could be
+%combined with \quicksand~
+extended with Iterative Deepening in future work.
 %so cannot check implementations directly.

### 2015-11-19 pldi-16/design.tex
garth's comments for design/soundness

@@
 Some kernels disable preemption without disabling interrupts,
-which can be modelled the same way using manual annotations. %of that API.
+which can be communicated to the MC using manual annotations. %of that API.
 This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
@@
 	%\Input{$eta$, $j_0$'s predicted completion time}
-	\Input{\PendingJobs, the list of pending jobs, sorted decreasingly by heuristic priority}
-	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted increasingly by ETA}
+	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
+	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
 	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
@@
 	}
-	return $j_0$ // ETA was bad, but no other job was better.
+	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
 	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
@@
 Because data races indicate access pairs that can interleave at instruction granularity,
-it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
+it is logical to re-execute the test and issue preemptions just before those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.
 
@@
 With Iterative Deepening, this is a simple matter of creating a new state space with an additional PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
+We call these {\em data-race PPs}.
 Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
@@
 %
-The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
+The rationale of spawning multiple jobs is that we don't know in advance which will be most fruitful: %which will be most fruitful is not known in advance:
 while the big job risks not completing in time,
@@
 % TODO CAMREADY: Put numbers here.
-In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
-which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
+In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
+%which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
+This phenomenon motivates Iterative Deepening to prioritize the jobs at run-time.
 
@@
 	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
-		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
+		When an access pair is reported as a data race candidate, we generate a new PP for each access, and add new jobs corresponding to different combinations of those with the existing PPs.}
 	\label{fig:new-dr-jobs}
@@
 Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
-to express how optimistic we are about the estimate's accuracy.
-We use 2 as this scaling factor based on the results in \cite{estimation}.
+to express how pessimistic we are about the estimate's accuracy.
+We use a scaling factor defaulting to 2 based on the results in \cite{estimation}.
 %though we allow changing it via the command line.
@@
 based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
-Jobs with both-order data-race PPs are prioritized higher,
-because single-order candidates are more likely to be false-positives
-(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
+Single-order candidates are more likely to be false positives,
+although preempting during the access itself is necessary to say for sure.
+Hence, we add PPs for both types of candiates, and heuristically prioritize jobs with both-order data-race PPs
+(indicated by the HeuristicPriority($\alpha$) call in Algorithm~\ref{alg:handledatarace}).
+%Jobs with both-order data-race PPs are prioritized higher,
+%because single-order candidates are more likely to be false-positives.
+%(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
 %so we prioritize jobs with both-order data-race PPs.
-For single-order races, we do not add a PP for the later access at all;
-should it be needed, preempting on the first access will suffice to upgrade the race to both-order.
+For single-order races, we do not initially add a PP for the later access at all:
+if preempting on the first access can reorder the race, it will be upgraded to both-order in the new state space, and we will add the second PP then.
+%should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

### 2015-11-19 pldi-16/implementation.tex
garth's comments for design/soundness

@@
 % this line can be cut if space is needed
-It can test both user-level and kernel-level code, although is limited to timer nondeterminism.
+It can test both user-level and kernel-level code (although it is limited to timer nondeterminism).
 % joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
@@
 {\bf Restricting PPs with stack trace predicates.}
-Most MC tools preempt indiscriminately on any sync API call, regardless of the call-site.
-However, when testing a particular module in a large codebase,
-the user is likely uninterested in PPs arising from other modules.
-\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
+When testing a particular module in a large codebase,
+the user is likely to be uninterested in PPs arising from other modules.
+Rather than preempting indiscriminately on any synchronization call, regardless of the call-site,
+\landslide~provides a configuration command, {\tt within\_function}, for identifying which call-sites matter.
+%Most MC tools preempt indiscriminately on any synchronization call, regardless of the call-site.
+%However, when testing a particular module in a large codebase,
+%the user is likely to be uninterested in PPs arising from other modules.
+%\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
 Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
-%The {\tt without\_function} directive works similarly, but as a blacklist.
-The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
-Multiple invocations can be used; later ones take precedence.
+%Similarly, the {\tt without\_function} directive indicates a blacklist,
+%serving as the dual of {\tt within\_function}.
+The {\tt without\_\allowbreak{}function} directive is the dual of {\tt within\_function}, indicating a blacklist.
+We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}).
+%Multiple invocations can be used; later ones take precedence.
 %\cite{landslide} provides further detail on this feature.
@@
 {\bf Data races in lock implementations.}
-Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses therein. %resulting from the lock implementation itself.
-The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
+Data race tools in prior work \cite{tsan,portend} recognize the implementations of synchronization primitives to avoid spuriously flagging memory accesses therein.
+%resulting from the lock implementation itself.
+Assuming that the lock implementation is already correct enables more productive data-race analysis on the rest of the codebase.
 Otherwise, with testing limited to one execution,
@@
 However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~with an option to make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
+Hence, we extended \landslide~to
+%with an option
+optionally make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
 which we use in \sect{\ref{sec:testsuite}}.
-%(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
+%(Accesses from other synchronization functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
 
@@
 %Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
-The interface to \landslide~ has two parts. %, which any similar MC could implement, has two parts.
+The interface with the MC has two parts. %, which any similar MC could implement, has two parts.
 First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
 % can lose this line due to space
-among other options such as mutex-testing mode,
-passed as an argument to \landslide.
-Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
-\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
-and whenever it finds a new data-race candidate or bug.
+plus other MC-specific options such as our modifications to \landslide~for testing mutexes. %mutex-testing mode,
+%passed as an argument to \landslide.
+Then, a dedicated \quicksand~thread communicates with the MC process via message-passing. %on a FIFO pipe.
+%\landslide~messages \quicksand~
+The MC messages after testing each new interleaving to report updated progress and ETA
+and whenever a new data-race candidate or bug is found.
 \quicksand~in turn replies whether to suspend/resume due to too high ETA, or quit due to timeout.
-We suspend jobs simply by making \landslide~wait on a message-passing call.
-Should \quicksand~later resume a suspended job, we send a message to continue,
-resuming the \landslide~instance where it left off.
+We suspend jobs simply by making the MC wait on a message-passing call.
+Should \quicksand~later re-schedule a suspended job, we send a message to continue,
+resuming it where it left off.
 %otherwise, we resume it only after time runs out, causing it to exit.

### 2015-11-19 pldi-16/soundness.tex
garth's comments for design/soundness

@@
 \begin{theorem}
-If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
+If a bug can be exposed by any thread interleaving possible by preempting on all instructions,
 Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
@@
 For any thread interleaving possible by preempting on any instruction,
-there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
+there exists an equivalent interleaving which uses only data-race PPs and synchronization API PPs.
 	\label{lem:relevant}
@@
 
-Let $p$ be the first PP in the given interleaving such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a sync API boundary.
-Because $\ppinstr{p}$ is not a sync API boundary,
+Let $p$ be the first PP in the given interleaving such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a synchronization API boundary.
+Because $\ppinstr{p}$ is not a synchronization boundary,
 no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
 Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
-Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a sync API boundary.
+Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a synchronization boundary.
 If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
@@
 By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
-In other words, we have transformed $p$ into $p'$ such that $\ppnext{p'} = i$, which is a data race or sync API boundary.
+In other words, we have transformed $p$ into $p'$ such that $\ppnext{p'} = i$, which is a data race or synchronization boundary.
 All PPs in the input trace can be inductively converted in the same manner.
@@
 \begin{definition}[Reachable data race]
-A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
+A data race candidate (and associated PPs) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
 \end{definition}
-Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
+Initially, the statically-available synchronization API PPs are reachable. Reachability of data-race PPs is transitive.
 \begin{lemma}[Saturation of data-race PPs]
-	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
+	Given any interleaving comprising only data-race PPs and synchronization API PPs, all involved PPs are reachable.
 	\label{lem:saturation}
@@
 To conclude,
-for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
+for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and synchronization PPs,
 and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
@@
 
-We identify a particular class of false positive data-race candidate in which the memory is recycled by {\tt malloc} between the two accesses.
+We identify a particular class of false positive data-race candidates in which the memory was recycled by {\tt malloc} between the two accesses.
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
 If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
-We term this a {\em malloc-recycle data race}.
+We call this a {\em malloc-recycle data race}.
 To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
-Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
+Here, Thread 2's logic usually corresponds to an initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
 
@@
 
-It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
-We implemented this check by adding a generation counter to \landslide's heap tracking:
+We mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address,
+%We implemented this check
+by adding a generation counter to \landslide's heap tracking:
 each allocation is given a unique ID,
@@
 the IDs of their containing blocks must match.
-However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
-Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
-Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.
+%However,
+When limited to a single test execution, suppressing any data race matching this pattern is unsound.
+Consider the more unusual program in Figure~\ref{fig:recycle-bug},
+in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
+Here, reordering lines 6-7 before line 3 will allow {\tt x} and {\tt x2} to race.
 
@@
 By definition of the malloc-recycle pattern,
-any such program must contain an access {\tt x1} by one thread T1,
+any such program must contain an access $a_1$ by one thread T1,
 followed by a {\tt free} and a {\tt malloc} possibly by either thread,
-followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
-For brevity we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
+followed by an access $a_2$ by the other thread T2. % not depending on the result of the middle malloc.
+Without loss of generality, we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}. %; the other cases are similar.
 We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
-hence, there must also be some ``publish'' action {\tt p} by T1 which communicates the address to T2.
-Because this is a true data race, {\tt p} must occur before {\tt x1}, as {\tt x2} cannot be reordered before {\tt p}.
+hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
+Because this is a true data race, $p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.
 
-We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
+We now show that a PP will be identified during T1 between $p$ and $a_1$.
 The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
-If locking or message-passing is used, our set of hard-coded PPs suffices.
-Otherwise, {\tt p} (and the corresponding read by T2) will be a data race, although it may itself be a malloc-recycle race.
-In this case we use induction on the pointer chain leading to {\tt x}:
-in the base case, {\tt p} is global or obtained via message-passing,
-and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on {\tt p}.
-Hence there will be a PP between {\tt p} and {\tt x1} no matter the mode of communication.
+If locking or message-passing is used, our set of hard-coded PPs suffices to provide a PP.
+Otherwise, $p$ (and the corresponding read by T2) will be a data race, although that may itself be a malloc-recycle race.
+In this case we use induction on the pointer chain leading to the shared address containing $p$:
+in the base case, $p$ is communicated via global data or message-passing,
+and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on $p$.
+Hence there will be a PP between $p$ and $a_1$ no matter the mode of communication.
 
-With this PP, DPOR will reorder {\tt x2} before {\tt x1}, while not changing {\tt x2}'s location.
+With this PP, DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
 As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
-Hence {\tt x1} and {\tt x2} will be in the same allocation;
+Hence $a_1$ and $a_2$ will be in the same allocation;
 hence the accesses can race without fitting the malloc-recycle pattern.
@@
 Note that we do not require PPs on the internal locking of {\tt malloc} or {\tt free},
-which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
+which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}, \sect{\ref{sec:future}}) to reduce state space size.

### 2015-11-19 pldi-16/evaluation.tex
garth's comments, pt 1.
garth's comments pt3, eval and related

@@
 
-Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone as well. %in other model checkers.
+Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could be employed alone as well. %in other model checkers.
 For example, a single-state-space tool could use data-race PPs during immediately subsequent interleavings,
@@
 changing the state space on the fly.
-Likewise, a message-passing-only tool could use Iterative Deepening despite a concurrency model lacking data races. % being absent from its concurrency model.
+Likewise, a message-passing-only tool could use Iterative Deepening despite a concurrency model lacking data races, % being absent from its concurrency model.
+to promote completing subset PP sets for large tests.
 Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
@@
 \label{sec:testsuite}
-Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
-and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
+Our test suite consists of \numthrlibs~``P2'' student thread libraries, from Carnegie Mellon's 15-410 operating systems class,
+and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's CS230 operating systems classes.
 %
-The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
+The P2 project comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and reader/writer locks;
 all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
@@
 The Pintos kernel project
-involves priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
+comprises priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
 using provided mutex, context-switch, and virtual memory implementations
@@
 Many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested).
-We fixed these by hand in advance to ensure that every bug in our study required some meaningful work by the MC.}.
+We fixed these by hand, before running tests, to ensure that every bug in our study required meaningful work by the MC.}.
 % TODO: Is this ok? Too arrogant?
-We believe that merely finding a small handful of new real-world bugs is too anecdotal,
+We believe that merely finding a small handful of new real-world bugs is distressingly anecdotal,
 and that our test suite's size allows for a more statistically significant comparison among MC and data-race testing strategies,
@@
 \newcommand\rwldgr{\texttt{rwlock\_test}}
+\newcommand\prisema{\texttt{sched\_test}}
+\newcommand\waitsimple{\texttt{wait\_test}}
+\newcommand\alarmsimul{\texttt{alarm\_test}}
+
 We tested P2s with 6 multithreaded programs:
@@
 (see \sect{\ref{sec:landslide}}).
-\newcommand\prisema{\texttt{sched\_test}}
-\newcommand\waitsimple{\texttt{wait\_test}}
-\newcommand\alarmsimul{\texttt{alarm\_test}}
-We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
+We tested Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
 \alarmsimul, for the timer sleep routine,
@@
 	so each test could only be run on a subset of the 78 submissions; see ``Number tested'' in Table~\ref{tab:drbugs}.
-}
+}.
 The source code of all 9 test cases is available at
 {\em [submitted as supplementary material]}.
-For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
-We ran all tests on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
+For all tests, we used {\tt without\_function} to blacklist PPs on {\tt malloc}'s internal lock.
+We ran all tests on 12-core 3.2 GHz Xeon machines. %with 12GB of RAM.
 In total, our evaluation comprises 629 tests (pairs of a test program and a Pintos or P2),
@@
 % XXX: I hate this paragraph. I have no idea what order to put the sentences in.
-To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all PPs on sync primitives enabled in advance (and no data-race PPs).
+To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all synchronization PPs enabled in advance (and no data-race PPs).
 We gave each \quicksand~test 10 CPUs for 1 hour each. % XXX
@@
 and instrumenting \quicksand~to report total CPU-hours rather than wall-clock time.
+%Pragmatically, with the growing importance of multicore, \quicksand's inherent parallelism is convenient to say the least.
+%Pragmatically, with most future performance improvements coming from increased core counts, \quicksand's inherent parallelism is convenient to say the least.
+(With the growing importance of multicore for performance, \quicksand's inherent parallelism is another convenient benefit.)
+
 %\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
@@
 We tested several configurations of \quicksand~to compare against SSS-MC from multiple angles,
+indicated by each of the ``QS'' lines,
 which we explain presently.
@@
 {\bf Finding the same bugs faster.}
-To show that Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
+To show that Iterative Deepening is effective even for MC testing domains without data races, %such as message-passing distributed systems,
 we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
@@
 QS-no-DR-PPs finds many more of the bugs much sooner.
-Hence, for smaller arbitrary CPU budgets,
+Hence, for arbitrary small CPU budgets,
 Iterative Deepening is likely to find bugs SSS-MC will miss.
@@
 with data-race PPs enabled.
-We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
-The break-even point is at a negligible 90 seconds.
+%We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
+%The break-even point is at a negligible 90 seconds.
+We find significantly more bugs within any CPU budget greater than 90 seconds,
+and ultimately conclude 10 CPU-hours with 179\% as many bugs in total.
 
 Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
-which represents only the bugs found in state spaces without data-race PPs (like QS-no-DR-PPs, but even when data-race PPs are enabled).
+which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
 Intuitively, this line shows that for programs with only benign data races,
@@
 % Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
-In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
-even compared to the other test programs\footnote{
-	The one bug SSS-MC found was in a fully-assembly lock implementation.
+%In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
+%even compared to the other test programs
+In \mxtest, in which we stop trusting the lock implementation's correctness,
+the control experiment found dramatically fewer bugs (just 1)\footnote{
+	The one bug SSS-MC found was in a fully-assembler lock implementation.
 {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
-Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
+Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
 %Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
@@
 {\bf Partial verification.}
-When a state space times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
+When a MC job times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
 %Beyond using the state space estimator to guess the percent coverage,
@@
 that a bug would be exposed by an untested interleaving,
-but \quicksand~at least reports which subsets of PPs produced state spaces that did complete in time.
+but \quicksand~at least reports which subsets of PPs resulted in state spaces that did complete in time.
 %\quicksand~does not attempt any such quantitative guarantee,
 %but can at least report
-On 129 tests, the SSS-MC control experiment timed out in 10 hours with no bugs found.
-Among these, \quicksand~found bugs in 36.
+On 129 tests, the SSS-MC control experiment timed out after 10 hours with no bugs found.
+Among these tests, \quicksand~found bugs in 36.
 For the other 93, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
@@
 
-Beyond finding new bugs with data-race PPs, we evaluated \quicksand's performance in classifying data-race candidates in two ways.
+Beyond finding new bugs with data-race PPs, we evaluated \quicksand's performance for classifying data-race candidates in two ways.
 
@@
 none of them exposed a new bug when used as a PP.
-This shows that for other data-race tools,
+This suggests that for other data-race tools,
 suppressing malloc-recycle candidates may be a productive heuristic,
@@
 We plot these for comparison,
-according to the time taken by that test's SSS-MC control experiment,.
+according to the time taken by that test's SSS-MC control experiment.
 When a test can be totally verified without any data-race PPs,
@@
 %could not achieve the coverage necessary
-would fail
+could fail
 to identify a racing access pair as a candidate to begin with.
@@
 To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
-For example, we reported malloc-recycle data races during the first interleaving, as a single-pass analysis must,
-rather than waiting until future interleavings to witness them without the malloc-recycle pattern
+For example, we reported malloc-recycle data races during the first interleaving, as a single-pass analysis must
+%rather than waiting until future interleavings to witness them without the malloc-recycle pattern
 (\sect{\ref{sec:recycle}}).
+This prevents \landslide~from suppressing an observed data race on the first interleaving,
+which would falsely classify it as nondeterministic.
 
@@
 This experiment suggests that
-such tools should be configured to use a similar feedback loop as Iterative Deepening.
+such tools could benefit from a similar feedback loop as Iterative Deepening.
 %i.e., discovering transitively-reachable data races while testing initial ones.

### 2015-11-19 pldi-16/paper.tex
garth's comments pt3, eval and related

@@
 We find new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
-Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide an equally strong verification as if every single instruction had been used as a PP.
+Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.
 

### 2015-11-19 pldi-16/related.tex
garth's comments pt3, eval and related

@@
 
-\subsection{Model Checking}
+\subsection{Stateless Model Checking}
 
@@
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
-Nevertheless, Iterative Deepening is still applicable to these tools.
+Nevertheless, Iterative Deepening is applicable to these tools.
 
@@
 choose a small subset of instructions to consider as PPs
-or be limited to very small test inputs.
+or be limited to very small test programs.
 %However, every such prior tool we know of has serious drawbacks.
-SKI \cite{ski} approaches kernel code by choosing a random set of instructions in advance, %offsets from the start of the test,
-which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
+SKI \cite{ski} approaches kernel code by statically choosing a random set of instructions in advance, %offsets from the start of the test,
+which is perhaps more similar to
+%stress testing or
+schedule fuzzing \cite{randomized-scheduler} than to exhaustive state space exploration.
 %
-SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment,
-although it requires code to be written in the PROMELA DSL.
+SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU \cite{rcu}, which is similar to our \mxtest~experiment,
+although it requires code to be written in the PROMELA language.
 %However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
@@
 
-{\bf Other techniques.} Various improvements or alternatives to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, DPOR for relaxed memory models \cite{tsopso}, and SAT-directed MC \cite{satcheck}.
-These are all orthogonal to our technique.
-Parrot \cite{parrot} combines MC with a partially-determinizing runtime, but still, fewer than half the non-trivial state spaces in their evaluation could be completed.
+{\bf Other techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
+%DPOR for relaxed memory models \cite{tsopso},
+and SAT-directed MC \cite{satcheck}.
+These are all compatible with our technique.
+Recent work \cite{tsopso} has extended DPOR for relaxed memory models \cite{memory-consistency-models},
+which we do not yet account for in our proofs (\sect{\ref{sec:soundness}}).
+Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial state spaces in their evaluation could be completed.
 %providing a strong argument for \quicksand.
@@
 when full completion is intractable (\sect{\ref{sec:future}}).
-However, ICB is limited to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+However, ICB is limited to a fixed set of PPs, and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
 
@@
 preferring to err on the side of false positives.
-RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited,
-although it uses random schedule fuzzing rather than stateless model checking.
-Note while these techniques can also classify our malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
+RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
+although it uses random schedule fuzzing rather than stateless MC.
+While this technique can also classify malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
 they require replaying the threads in a new interleaving.
@@
 which is tantamount to adding a new state space in \quicksand.
-Our proof in \sect{\ref{sec:recycle}} allows eliminating this special case with no additional replay beyond what DPOR already requires.
+Our proof in \sect{\ref{sec:recycle}} allows us to eliminate this special case with no additional replay beyond what DPOR already requires.
 
@@
 different program output. %, which is orthogonal to our technique.
-However, Portend does not test alternate interleavings {\em in advance} of knowing data races,
+However, Portend does not test alternate interleavings {\em in advance} of knowing any specific data races,
 which is necessary to expose certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).

### 2015-11-19 pldi-16/soundness.tex
garth's comments pt3, eval and related

@@
 
-\begin{definition}[Reachable data race]
-A data race candidate (and associated PPs) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
+\begin{definition}[Reachability]
+A data race candidate, and its associated PPs, are reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
 \end{definition}
@@
 Given that the interleaving prefix preceding some PP $p$ is reachable,
-we will show that either $p$ is reachable, or a new data race among $\ppothers{p}$, not previously reachable, is now reachable.
-The latter condition suffices because in a finitely-sized program, there must be finitely many unique racing instruction pairs, so induction on the number of new data-race PPs among $\ppothers{p}$ will make $p$ itself reachable.
+we require that either $p$ is reachable, or a new data race among $\ppothers{p}$ will be newly reachable. %, not previously reachable, is now reachable.
+The latter condition suffices because in a finitely-sized codebase, there must be finitely many unique racing instruction pairs.
+%, so induction on the number of new data-race PPs among $\ppothers{p}$ will make $p$ itself reachable.
 
@@
 
-We mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address,
-%We implemented this check
-by adding a generation counter to \landslide's heap tracking:
-each allocation is given a unique ID,
-and when evaluating whether two heap accesses can race,
-the IDs of their containing blocks must match.
 %However,
@@
 \begin{proof}
-By definition of the malloc-recycle pattern,
-any such program must contain an access $a_1$ by one thread T1,
+%By definition of the malloc-recycle pattern,
+Any such program must contain an access $a_1$ by one thread T1,
 followed by a {\tt free} and a {\tt malloc} possibly by either thread,
@@
 
-We now show that a PP will be identified during T1 between $p$ and $a_1$.
+We require that a PP will be identified during T1 between $p$ and $a_1$.
 The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
@@
 As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
-Hence $a_1$ and $a_2$ will be in the same allocation;
-hence the accesses can race without fitting the malloc-recycle pattern.
+Hence $a_1$ and $a_2$ %will be in the same allocation;
+can race without fitting the malloc-recycle pattern.
 % Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
@@
 
-Note that we do not require PPs on the internal locking of {\tt malloc} or {\tt free},
-which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}, \sect{\ref{sec:future}}) to reduce state space size.
+Note that we do not require PPs on the internal lock of {\tt malloc} or {\tt free},
+which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
+
+We implemented a simple check in \landslide~to recognize the malloc-recycle pattern:
+%We mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address,
+%We implemented this check
+%by adding a generation counter to \landslide's heap tracking:
+%when tracking heap state,
+each heap allocation is given a unique ID,
+and when evaluating whether two heap accesses can race,
+the IDs of their containing blocks must match.

### 2015-11-19 pldi-16/warpzone.tex
garth's comments pt3, eval and related

@@
 {\bf Avoiding redundant work.}
-When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of thread interleavings compared to the old one.
+When we extend a small state space with more PPs, the new state space is guaranteed to test a superset of interleavings compared to the old one.
 Any interleaving which does not preempt threads on any of the new PPs will be repeated work.
@@
 %on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
-This may make Iterative Deepening slower than SSS-MC to find certain bugs,
-for example, if both {\tt mutex\_lock} and {\tt mutex\_unlock} PPs together expose a bug, but not either alone.
+This may make us slower than SSS-MC to find certain bugs,
+for example, if both {\tt lock} and {\tt unlock} PPs together expose a bug, but not either alone.
 %Similarly, when pursuing total verification (\sect{\ref{sec:totalverif}}),
@@
 %as Iterative Deepening will test many subsets of data-race PPs first.
-Predicting whether each future interleaving has been tested before is not straightforward,
-but we believe future implementations of Iterative Deepening could incorporate cross-job memoization
+Predicting whether an upcoming interleaving has already been tested is not straightforward,
+but we believe future implementations
+%of Iterative Deepening
+could incorporate cross-job memoization
 to prune some or all such repeated work.
@@
 \quicksand~was able to partially guarantee safety for some PPs in 89\% of tests with too-large maximal state spaces.
-However, in 4 tests, no more than the minimal state space could be verified,
-and in 6 other tests, no state spaces were completed at all.
+However, in 4 cases, no more than the minimal state space could be verified,
+and in 6 others, no state spaces were completed at all.
 %Larger state spaces often result from finer-grained locking,
@@
 implementations could use this mechanism to {\em dynamically} subset PPs further,
-making partial verification of such large tests possible.
+making partial verification of larger tests possible.
 %enabling partial verification of such large tests. % if need space

### 2015-11-19 pldi-16/design.tex
some of haryadi's commence

@@
 In this context, the depth is the number of PPs used.
-Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs
-, which we refer to as {\em jobs}.
+Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs,
+which we refer to as {\em jobs}.
 It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
@@
 which can be any number of subsets of the statically-available PPs SSS-MC would use.
+For completeness (\sect{\ref{sec:totalverif}}), the maximal state space must be included among these.
+
 For testing user-space code, we begin with the four PP sets from Figure~\ref{fig:id}:
@@
 and $\{yield,lock,unlock\}$,
-By extension, these also introduce PPs on any other sync primitives implemented with internal locks,
+By extension, these also introduce PPs on any other sync primitives
+%implemented
+with internal locks,
 such as condvars or semaphores.
@@
 	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
-		// Common case: Job is expected to complete. \\
-		return $j_0$
+		return $j_0$ // Common case: job is expected to finish.
 	}

### 2015-11-19 pldi-16/evaluation.tex
some of haryadi's commence

@@
 Our test suite consists of \numthrlibs~``P2'' student thread libraries, from Carnegie Mellon's 15-410 operating systems class,
-and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's CS230 operating systems classes.
+and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's
+% TODO CAMREADY: CMSC 23000 to be precise
+CS230 operating systems classes.
 %
@@
 % 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
-Though not ``real world'' programs, both projects are quite complex:
+%Though not ``real world'' programs,
+Both projects are quite complex:
 the P2s average 1807 lines of code, %C and x86 assembly (stddev 489.5),
@@
 %done
-and the Pintoses average 718.
+and the Pintoses average 718 lines, for a total of 198,772.
 
@@
 We believe that merely finding a small handful of new real-world bugs is distressingly anecdotal,
-and that our test suite's size allows for a more statistically significant comparison among MC and data-race testing strategies,
+and that our test suite's size allows for a more statistically significant comparison among MC and data-race testing strategies.
 %compared to anecdotally showing a small handful of new bugs that could be found in real-world programs.

### 2015-11-19 pldi-16/overview.tex
some of haryadi's commence

@@
 
-The checker defins the granularity of thread interleavings by which {\em preemption points} it uses to switch threads.
+The checker defines the granularity of thread interleavings by which {\em preemption points} it uses to switch threads.
 Most model checkers \cite{chess,dbug-ssv} choose synchronization/thread library API boundaries for these points;
@@
 Rather than committing to one state space with every available preemption point enabled,
-Iterative Deepening searches among different {\em subsets} of the points;
-for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
-Hence, we will test all the state spaces illustrated in Figure~\ref{fig:id} in parallel,
+we will search among different {\em subsets} of the points.
+%for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
+Hence, we will test all the state spaces shown in Figure~\ref{fig:id} in parallel,
 and decide on-the-fly whether to pursue each test, or to defer it in favour of others.

### 2015-11-20 pldi-16/overview.tex
put subfigure labels on figure 2

@@
 	\caption{Iterative Deepening example.
-		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
-		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
+		The minimal state space (A) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
+		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (B), {\tt mutex\_unlock} alone (C), or both together (D).
 Each option increases the state space size unpredictably, so multiple state spaces should be tested in parallel.
@@
 Figure~\ref{fig:id} shows several possible resulting state spaces.
-The approach of prior work is to enable all preemption points simultaneously, i.e., to test only the bottom state space.
+The approach of prior work is to enable all preemption points simultaneously, i.e., to test only the state space denoted (D).
 
@@
 We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}, top).
-The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}, bottom).
+The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}, A).
+The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}, D).
 %However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.

### 2015-11-20 pldi-16/proof/convergence.tex
handle nwf's proof comments and fix notation scope issue

@@
 
-\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
-\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
-\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
-\newcommand\pppfx[1]{\ensuremath{\mathsf{pfx}(#1)}}
+%\newcommand\ppnext[3]{\ensuremath{\mathsf{next}_{#1}(#2,#3)}}
+\newcommand\ppinstr[3]{\ensuremath{\mathsf{nextinstr}_{#1}(#2,#3)}}
+\newcommand\ppothers[3]{\ensuremath{\mathsf{others}_{#1}(#2,#3)}}
+\newcommand\pppfx[2]{\ensuremath{\mathsf{pfx}(#1,#2)}}
 \newcommand\conflicts[2]{\ensuremath{\mathsf{conflicts}(#1,#2)}}
@@
 \[
-	\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), (t_{\beta{}1}, p_{\beta{}1}), ... (t_{\alpha{}n}, p_{\alpha{}n}), ...\}
+	\mathcal{I} = [ (t_{\alpha{}1}, p_{\alpha{}1}), (t_{\beta{}1}, p_{\beta{}1}), ... (t_{\alpha{}n}, p_{\alpha{}n}), ... ]
 \]
-be an interleaving trace, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
+be an interleaving trace, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$ at its end.
 We will use the following notation:
 \begin{itemize}
-	\item $\ppnext{\pai}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
-	\item $\ppinstr{\pai}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
-	\item $\ppothers{\pai}$ indicates the set of all transitions by other threads between $\tai$ and $t_{\alpha{}(i+1)}$.
-	\item $\conflicts{t_\alpha,t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
+	\item $\ppinstr{\mathcal{I}}{\alpha}{i}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$ (the {\em next} transition), i.e., the instruction of $\alpha$ that \pai preempted just before.
+	\item $\ppothers{\mathcal{I}}{\alpha}{i}$ indicates the set of all transitions by other threads between $\tai$ and $t_{\alpha{}(i+1)}$.
+	\item $\conflicts{t_\alpha}{t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
 \end{itemize}
@@
 \begin{definition}[Relevant preemption point]
-	We say a PP $\pai$ is {\em relevant} if either $\ppnext{\pai}$ is a sync API boundary (including a voluntary yield),
+	We say a PP $\pai$ is {\em relevant} if either $\ppinstr{\mathcal{I}}{\alpha}{i}$ is a sync API boundary (including a voluntary yield),
 	or if
 	\[
-		\conflicts{\ppothers{\pai}}{\ppinstr{\pai}} \ne \emptyset
+		\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{\ppinstr{\mathcal{I}}{\alpha}{i}} \ne \emptyset
 	\]
-	i.e., the instruction by thread $\alpha$ immediately after $\pai$ has a memory conflict with some other thread interleaved during $\pai$.
+	i.e., the instruction by thread $\alpha$ immediately after $\pai$ has a memory conflict with some other thread interleaved between $\tai$ and that instruction.
 \end{definition}
@@
 	Let $\pai$ be the first irrelevant PP of a naive interleaving $\mathcal{I}$.
-	We ask, did some thread among $\ppothers{\pai}$ conflict with any instruction from $\ppnext{\pai}$, even though there was no conflict with $\ppinstr{\pai}$?
+	We ask, did some thread among $\ppothers{\mathcal{I}}{\alpha}{i}$ conflict with any instruction from $t_{\alpha{}(i+1)}$, even though there was no conflict with $\ppinstr{\mathcal{I}}{\alpha}{i}$?
 	\begin{itemize}
 			% rho for relevant.
-		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $\tai$ contains a sync API boundary instruction that's not already a PP,
-			then let $\rho$ be either the first instruction by $\ppnext{\pai}$ among the conflicts or the first sync API instruction, whichever comes first.
-			Let $\pppfx{\rho}$ denote the instruction sequence between $\ppinstr{\pai}$ and $\rho$, including the former but not the latter.
+			% no, scratch that.
+			% chi for conflict.
+		\item If $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} \ne \emptyset$, or if $\tai$ contains a sync API boundary instruction that's not already a PP,
+			then let $\chi$ be either the first instruction by $t_{\alpha{}(i+1)}$ among the conflicts or the first sync API instruction, whichever comes first.
+			Let $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ denote the instruction sequence between $\ppinstr{\mathcal{I}}{\alpha}{i}$ and $\chi$, including the former but not the latter.
 			Now, we output the new interleaving:
@@
 				\\
-				$\mathcal{I}' = \{...,$&$(\tai \cup \pppfx{\rho}, \pai'),$& \\
-									   &$\ppothers{\pai},$& \\
-					       &$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
+				$\mathcal{I}' = [...,$&$(\tai \cup \pppfx{t_{\alpha{}(i+1)}}{\chi}, \pai'),$& \\
+																		 &$\ppothers{\mathcal{I}}{\alpha}{i},$& \\
+						       &$(t_{\alpha{}(i+1)} \setminus \pppfx{t_{\alpha{}(i+1)}}{\chi}, p_{\alpha{}(i+1)}),$ & $ ...]$ \\
 				\\
@@
 
-			In other words, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
-			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
-			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR, $\mathcal{I}' \equiv \mathcal{I}$.
-		\item Otherwise, $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} = \emptyset$, and no instructions among $\ppnext{\pai}$ are a sync API boundary.
+			In other words, we have simply reordered $\ppothers{\mathcal{I}}{\alpha}{i}$ to between $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ and $\chi$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\chi$.
+			We know it must be possible to reorder $\ppothers{\mathcal{I}}{\alpha}{i}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{t_{\alpha{}(i+1)}}{\chi}$, hence the other threads' runnability cannot be affected.
+			Likewise $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ is independent with $\ppothers{\mathcal{I}}{\alpha}{i}$, so by the soundness of DPOR, $\mathcal{I}' \equiv \mathcal{I}$.
+		\item Otherwise, $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} = \emptyset$, and no instructions among $t_{\alpha{}(i+1)}$ are a sync API boundary.
 			Then we output the new interleaving:
 			\[
-				\mathcal{I}' = \{..., (\tai \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
+				\mathcal{I}' = [..., (\tai \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\mathcal{I}}{\alpha}{i},...]
 			\]
-			In other words, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
+			In other words, we have reordered {\ppothers{\mathcal{I}}{\alpha}{i}} with the entire next transition by thread $\alpha$.
 			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
-					       %&$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
+			%&$(t_{\alpha{}(i+1)}) \setminus \pppfx{t_{\alpha{}(i+1)}}{\chi}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
 	\end{itemize}
@@
 \begin{definition}[Reachable preemption point]
-A PP $\pai$ such that either $\ppinstr{\pai}$ is part of a reachable data race, or a sync API boundary.
+	A PP $\pai$ such that either $\ppinstr{\mathcal{I}}{\alpha}{i}$ is part of a reachable data race, or a sync API boundary.
 \end{definition}
@@
 \begin{proof}
-Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
+Let $\mathcal{I} = [(t_{\alpha{}1}, p_{\alpha{}1}), ...]$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
 We proceed by induction on the PPs according to the order of their preemptions.
-(Note that this is not necessarily the same as the order of the racing instructions $\ppinstr\pai$, which occur in $\ppnext\pai$, not in \tai.)
+(Note that this is not necessarily the same as the order of the racing instructions $\ppinstr{\mathcal{I}}{\alpha}{i}$, which occur in $t_{\alpha{}(i+1)}$, not in \tai.)
 For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbh$ with $\tbh \prec \tai \in \mathcal{I}$, $\pbh$ is reachable.
-We must show that a data race involving $\ppinstr\pai$
-is reachable, and we are not allowed to use $\pai$ until we find the data race between $\ppinstr\pai$ and $\ppothers\pai$.
+We must show that a data race involving $\ppinstr{\mathcal{I}}{\alpha}{i}$
+is reachable, and we are not allowed to use $\pai$ until we find the data race between $\ppinstr{\mathcal{I}}{\alpha}{i}$ and $\ppothers{\mathcal{I}}{\alpha}{i}$.
 
@@
 \[
-	\mathcal{J} = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
+	\mathcal{J} = [..., (\tai \cup t_{\alpha{}(i+1)}', \paj), \ppothers{\mathcal{I}}{\alpha}{i}' ]
 \]
-where we have reordered $\ppnext\pai$ to before the execution of $\ppothers\pai$.
+where we have reordered $t_{\alpha{}(i+1)}$ to before the execution of $\ppothers{\mathcal{I}}{\alpha}{i}$.
 Here, $\paj$ is the first sync API PP in $\alpha$ after $\pai$ (as $p_{\alpha(i+1)}$ may be a not-yet-reachable data race PP),
-and $\ppnext\pai'$ may include transitions of $\alpha$ beyond $\ppnext\pai$ itself.
+and $t_{\alpha{}(i+1)}'$ may include transitions of $\alpha$ beyond $t_{\alpha{}(i+1)}$ itself.
 %and for brevity, we will abbreviate the union of transitions as $\tak$.
-Then, $\ppothers\pai'$ are the other threads' transitions from our target interleaving.
-except they may be altered by the presence of some {\em other} data race in $\ppnext\pai'$ occurring after $\ppinstr\pai$.
-Likewise, $\ppnext\pai'$ may be altered by some other data race in $\ppothers\pai'$.
-The only certainty so far is that $\ppnext\pai'$ begins with $\ppinstr\pai$.
+Then, $\ppothers{\mathcal{I}}{\alpha}{i}'$ are the other threads' transitions from our target interleaving.
+except they may be altered by the presence of some {\em other} data race in $t_{\alpha{}(i+1)}'$ occurring after $\ppinstr{\mathcal{I}}{\alpha}{i}$.
+Likewise, $t_{\alpha{}(i+1)}'$ may be altered by some other data race in $\ppothers{\mathcal{I}}{\alpha}{i}'$.
+The only certainty so far is that $t_{\alpha{}(i+1)}'$ begins with $\ppinstr{\mathcal{I}}{\alpha}{i}$.
 
-It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
-could be reordered to after $\ppothers\pai'$,
-eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
-Unfortunately, $\mathcal{J}$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
+It would be straightforward to show that any other data races in $t_{\alpha{}(i+1)}'$, which would be discovered in this interleaving,
+could be reordered to after $\ppothers{\mathcal{I}}{\alpha}{i}'$,
+eventually transforming $t_{\alpha{}(i+1)}'$ to contain no conflicting memory accesses but $\ppinstr{\mathcal{I}}{\alpha}{i}$ itself.
+Unfortunately, $\mathcal{J}$ is not necessarily reachable, as $\ppothers{\mathcal{I}}{\alpha}{i}'$ still includes any data-race PPs from $\ppothers{\mathcal{I}}{\alpha}{i}$,
 which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
-Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
+Hence, we must perform the same ``coalescing'' for each thread in $\ppothers{\mathcal{I}}{\alpha}{i}'$ that we did for $\alpha$,
 which we will abbreviate:
 \[
-	\mathcal{K} = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
+	\mathcal{K} = [..., (\tai \cup t_{\alpha{}(i+1)}', \paj), \coalesce{\ppothers{\mathcal{I}}{\alpha}{i}'} ]
 \]
-For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
-starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
+For each other thread $\beta$, $\coalesce{\ppothers{\mathcal{I}}{\alpha}{i}'}$ will contain a single transition,
+starting with the first instruction of $\beta$'s corresponding transition in $\ppothers{\mathcal{I}}{\alpha}{i}$,
 and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
-Again, the only certainly-executed instruction by $\alpha$ after $\tai$ is $\ppinstr\pai$.
-It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
+Again, the only certainly-executed instruction by $\alpha$ after $\tai$ is $\ppinstr{\mathcal{I}}{\alpha}{i}$.
+It is even possible that neither $\ppinstr{\mathcal{I}}{\alpha}{i}$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
 until a different data race PP between two other threads is discovered,
@@
 {\bf Finding the next reachable data-race PP.}
-Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
+Now, we show by induction that a data race on $\ppinstr{\mathcal{I}}{\alpha}{i}$ will be reachable.
 We assume that a state space $\mathcal{S}$ containing $\mathcal{K}$,
-plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
+plus $n$ unique data race PPs among $t_{\alpha{}(i+1)}'$ and $\ppothers{\mathcal{I}}{\alpha}{i}$,
 will be reachable.
 (In the base case, the number of new data-race PPs is 0, and $\mathcal{K}$'s reachability is justified above.)
-We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
+We must show that in this state space, if the $\ppinstr{\mathcal{I}}{\alpha}{i}$ data race is not reachable,
 a new unique data-race PP will instead be reachable.
 
-By the definition of fully-relevant interleaving, $\mathcal{I}$ guarantees that some other thread $\omega$ can execute a data-racing instruction with $\ppinstr\pai$.
+By the definition of fully-relevant interleaving, $\mathcal{I}$ guarantees that some other thread $\omega$ can execute a data-racing instruction with $\ppinstr{\mathcal{I}}{\alpha}{i}$.
 %
@@
 Such a circular dependency seems intuitively impossible, but to actually find the ``first reachable'' PP is not straightforward.
-We require that in $\mathcal{S}$, there exists a single transition $\tbk$ that can alone be split into $\{\tbka,\tbkb\}$,
+We require that in $\mathcal{S}$, there exists a single transition $\tbk$ that can alone be split into $[\tbka,\tbkb]$,
 and some other communicating transition $\tgl$ which conflicts with $\tbkb$.
-%i.e., $\{\tbka, \tgl', \tbkb'\}$ exposes new program behaviour.
+%i.e., $[\tbka, \tgl', \tbkb']$ exposes new program behaviour.
 
-We show this by contradiction\footnote{
-Regrettably, we could not devise a constructive algorithm for this part of the proof.
-We leave constructively identifying the first-reachable data-race PP to future work,
-and trust that merely showing one must exist will satisfy the demands of modern verification.}:
+We show this by contradiction:
+% Sully points out that it's constructive because the set of interleavings / data races is finite.
+%\footnote{
+%Regrettably, we could not devise a constructive algorithm for this part of the proof.
+%We leave constructively identifying the first-reachable data-race PP to future work,
+%and trust that merely showing one must exist will satisfy the demands of modern verification.}:
 %
 Assume that for all $\tbk \in \mathcal{S}$,
-and all possible points $\pbk'$ at which to split it into $\{\tbka,\tbkb\}$,
+and all possible points $\pbk'$ at which to split it into $[\tbka,\tbkb]$,
 and all other non-MHB transitions $\tgl$,
@@
 \[
-	\{..., \tbka, \tbkb, \tgl\}
+	[..., \tbka, \tbkb, \tgl]
 	\equiv
-	\{..., \tbka, \tgl', \tbkb'\}
+	[..., \tbka, \tgl', \tbkb']
 \]
@@
 Hence, inductively, no set of new PPs of any size would expose new program behaviour not already exposed in $\mathcal{S}$.
-However, the instruction by $\omega$ which conflicts with $\ppinstr\pai$ was not observed in $\mathcal{S}$,
+However, the instruction by $\omega$ which conflicts with $\ppinstr{\mathcal{I}}{\alpha}{i}$ was not observed in $\mathcal{S}$,
 so we have our contradiction.
 
-Hence, if $\mathcal{S}$ does not expose the $\ppinstr\pai$ data race directly,
-there must exist some transitions $\tbk = \{\tbka,\tbkb\}$ and $\tgl$ such that $\tgl$ shared-memory-conflicts with $\tbkb$ and could be interleaved immediately before it.
+Hence, if $\mathcal{S}$ does not expose the $\ppinstr{\mathcal{I}}{\alpha}{i}$ data race directly,
+there must exist some transitions $\tbk = [\tbka,\tbkb]$ and $\tgl$ such that $\tgl$ shared-memory-conflicts with $\tbkb$ and could be interleaved immediately before it.
 By the maximal state space assumption, all sync API PPs are already enabled, so the locksets of $\beta$ and $\gamma$ cannot overlap and there is no MHB relation.
@@
 {\bf Reaching $\mathbf{p_{\alpha{}i}}$.}
-Hence either a data-race with $\ppinstr\pai$ will be identified,
+Hence either a data-race with $\ppinstr{\mathcal{I}}{\alpha}{i}$ will be identified,
 or an infinite/nonterminating sequence of other unique data-race PPs will be identified,

### 2015-11-20 pldi-16/proof/di.tex
handle nwf's proof comments and fix notation scope issue

@@
 Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
+We also include a trace of all memory accesses in the representation of transitions.
 
@@
 Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
-% FIXME: Address concern about using blocking sync primitives, eg sem or rwlock, in a mutex like fashion.
-% FIXME: Even though MHB(T1_0, T1_1) and MHB(T1_1, T2_0), then still not necessarily MHB(T1_0, T2_0).
 
@@
 }.
+We illustrate the difference in Figure~\ref{fig:mhb}.
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& Thread 1 & Thread 2 \\
+	1 & \texttt{\hilight{brickred}{my\_x->foo = ...;}} & \\
+	2 & \texttt{\hilight{olivegreen}{mutex\_lock(...);}} &\\
+	3 & \texttt{global\_x = my\_x;} & \\
+	%4 & \texttt{\hilight{olivegreen}{yield();}} & \\
+	4 & \texttt{\hilight{olivegreen}{mutex\_unlock(...);}} & \\
+	5 & & \texttt{\hilight{olivegreen}{mutex\_lock(...);}} \\
+	6 & & \texttt{my\_x = global\_x;} \\
+	7 & & \texttt{\hilight{olivegreen}{mutex\_unlock(...);}} \\
+	8 & & \texttt{if (my\_x != NULL)} \\
+	9 & & \texttt{\hilight{brickred}{~~~~my\_x->foo = ...;}} \\
+\end{tabular}
+	\caption{Example program to illustrate the difference between {\em pure happens-before} and {\em must-happen-before}.
+	Under pure happens-before (which does not identify false positives), lines 1 and 9 are not a data race candidate.
+	Under MHB, they are; although after trying to reorder them, it will be classified as a false positive.}
+	\label{fig:mhb}
+\end{figure}
+Intuitively, $A$ MHB $B$ can be thought to mean ``$B$ cannot be reordered before $A$''.
+Note also that, while lock-protected critical sections can be reordered around each other (i.e., line 1 not MHB lines 8-9),
+one cannot be reordered to be in the middle of the other (i.e, lines 3-4 MHB line 6).
+Hence, MHB is not necessarily transitive.
+In the latter case, the MHB relation is established by the mutex's blocking mechanism used during contention.
 
@@
 A pair of memory accesses between two threads to the same address where at least one of them is a write.
+\end{definition}
+
+\begin{definition}[Interleaving]
+	An ordered list of transitions and preemption points between them.
 \end{definition}
@@
 
+\begin{definition}[State space]
+	A set of interleavings subject to the constraint that, given the preemption points used, all equivalence classes of possible interleavings are represented by at least one member.
+\end{definition}
+
 \begin{definition}[Dynamic Partial Order Reduction (DPOR)]
@@
 	guaranteed to reorder transitions of two threads
-	iff they have a shared memory conflict and are not related by MHB \cite{dpor}.
+	iff they are not independent and are not related by MHB \cite{dpor}.
 	\label{def:dpor}

### 2015-11-20 pldi-16/proof/proof.tex
add colour and define soundness of dpor
handle nwf's proof comments and fix notation scope issue

@@
 
-First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve MC preempting on every single instruction.
+First, we prove that given enough time, Iterative Deepening will converge to the same degree of schedule coverage that would be provided by a na\"{i}ve model checker (MC) preempting on every single instruction.
 In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,

### 2015-11-20 pldi-16/proof/convergence.tex
make the convergence figure but comment it out

@@
 The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO: uncomment for camready/TR
+For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO CAMREADY: uncomment
 note that we are using the constructive contrapositive direction:
@@
 It is even possible that neither $\ppinstr{\mathcal{I}}{\alpha}{i}$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
-until a different data race PP between two other threads is discovered,
-as shown in Figure~\ref{fig:threethreads}.
+until a different data race PP between two other threads is discovered.
+% TODO CAMREADY
+%as shown in Figure~\ref{fig:threethreads}.
 
@@
 
-\begin{figure}[t]
-	TODO % TODO
-	\caption{In this example, even though thread 1's data-race PP must occur first in a buggy interleaving, it cannot be reached until after Iterative Deepening first reaches the later-occurring data-race PP between threads 2 and 3.}
-	\label{fig:threethreads}
-\end{figure}
+%\begin{figure*}[t]
+%	\small
+%	\centering
+%\begin{tabular}{rlll}
+%	& \multicolumn{3}{c}{\texttt{~~int x = 0, y = 0;~~~~~~~~~~~~~~~~~~~~~~~~~~~}} \\
+%	& \multicolumn{3}{c}{\texttt{bool run\_racy\_code = false;~~~~~~~~~~~~~~~~}} \\
+%	\\
+%	& Thread $\alpha$ & Thread $\beta$ & Thread $\gamma$ \\
+%& \texttt{run\_racy\_code = true;} \\
+%& \texttt{int tmp = x;} \\
+%& // preempted ($p_{\alpha{}1}$) \\
+%& & & \texttt{if (run\_racy\_code) \{} \\
+%& & & \texttt{~~~~int tmp = y;} \\
+%& & & \texttt{~~~~}// preempted ($p_{\gamma{}1}$) \\
+%& & \texttt{int tmp = y;} \\
+%& & // no preemption ($p_{\beta{}1}$) \\
+%& & // $\ppinstr{}{\beta}{1}$ \\
+%& & \texttt{y = tmp + 1;} \\
+%& & & \texttt{~~~~}// $\ppinstr{}{\gamma}{1}$ \\
+%& & & \texttt{~~~~y = tmp + 1;} \\
+%& & & \texttt{~~~~if (/* y raced */) \{} \\
+%& & & \texttt{~~~~~~~~int tmp = x;} \\
+%& & & \texttt{~~~~~~~~}// preempted ($p_{\gamma{}2}$) \\
+%& // $\ppinstr{}{\alpha}{1}$ \\
+%& \texttt{x = tmp + 1;} \\
+%& & & \texttt{~~~~~~~~}// $\ppinstr{}{\gamma}{2}$ \\
+%& & & \texttt{~~~~~~~~x = tmp + 1;} \\
+%& & & \texttt{~~~~~~~~assert(!(/* x raced */));} \\
+%& & & \texttt{~~~~\}} \\
+%& & & \texttt{\}} \\
+%
+%
+%\end{tabular}
+%	\caption{In this example, even though thread 1's data-race PP must occur first in a buggy interleaving, it cannot be reached until after Iterative Deepening first reaches the later-occurring data-race PP between threads 2 and 3.}
+%	\label{fig:threethreads}
+%\end{figure*}
 

### 2015-11-20 pldi-16/proof/di.tex
clean up todos

@@
 The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and judging whether they are true data races or false positives.
-It is up to \quicksand, outside the scope of this proof, to decide whether true data races are benign or buggy.
+In this proof we are concerned solely with candidates, and wheter they can be observed to race or are false positives.
+It is up to the MC to decide whether true data races are benign or buggy.
 

### 2015-11-20 pldi-16/design.tex
clean up todos
garth's final comments, pt1, up thru implementation

@@
 In this context, the depth is the number of PPs used.
-Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs,
-which we refer to as {\em jobs}.
-It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
+Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs.
+%which we refer to as {\em jobs}.
+We refer to each unique set of PPs as a {\em job},
+and prioritize them based on number of PPs, ETA, and whether they include data-race PPs,
 We rely on state-space estimation \cite{estimation}
 to predict which jobs are likely to complete within a reasonable time,
-before actually testing each interleaving.
+before actually testing a large fraction of interleavings for each.
 The overall goal is to decide automatically when to defer testing a state space,
@@
 and to enable completing appropriately-sized jobs within that budget.
+We seek to maximize completed state spaces,
+as each one serves as a guarantee that all interleavings possible with its PPs were tested.
 
@@
 and $\{yield,lock,unlock\}$,
-By extension, these also introduce PPs on any other sync primitives
-%implemented
-with internal locks,
+By extension, these also introduce PPs on any other primitives
+which use
+internal locks,
 such as condvars or semaphores.
@@
 \subsection{Choosing the best job}
-
-% TODO: Move the 'seed sets' discussion up here, from implementation.
 
@@
 	\Input{$j_0$, the currently-running job}
-	\Input{\AllJobs, the set of all existing jobs}
+	\Input{\AllJobs, the set of all existing (or completed) jobs}
 	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
@@
 	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{\alpha\}$}{
-		AddNewJob($\{\alpha\}$, HeuristicPriority($\alpha$))
+		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
 	}
@@
 never suspend jobs before they pass a certain threshold of interleavings tested,
-for which we choose 32,
+with a default of 32,
 so that their ETAs have some time to stabilize.

### 2015-11-20 pldi-16/implementation.tex
garth's final comments, pt1, up thru implementation

@@
 optionally make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
-which we use in \sect{\ref{sec:testsuite}}.
+which we use to test mutexes (\sect{\ref{sec:testsuite}}).
 %(Accesses from other synchronization functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)
@@
 
-\quicksand~is an independent program that wraps the execution of several \landslide~instances.
+\quicksand~is an independent program that wraps the execution of several \landslide~MC instances.
 The implementation is roughly 3000 lines of C.
@@
 %\landslide~messages \quicksand~
-The MC messages after testing each new interleaving to report updated progress and ETA
+The MC messages after testing each interleaving to report updated progress and ETA
 and whenever a new data-race candidate or bug is found.
-\quicksand~in turn replies whether to suspend/resume due to too high ETA, or quit due to timeout.
-We suspend jobs simply by making the MC wait on a message-passing call.
+\quicksand~in turn replies whether to resume/suspend (due to too high ETA), or quit (due to timeout).
+We suspend jobs simply by making the MC wait on a message-passing reply.
 Should \quicksand~later re-schedule a suspended job, we send a message to continue,

### 2015-11-20 pldi-16/intro.tex
garth's final comments, pt1, up thru implementation

@@
 %makes full state space completion intractable for even modestly-sized tests.
-Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion,
+Choosing preemption points is a tradeoff between schedule coverage and feasibility of completion,
 but even with state-of-the-art reduction techniques,
-any choice made in advance must necessarily leave some tests impossibly large \cite{parrot,mcr}.
+any choice made in advance must necessarily leave some tests unaffordably large \cite{parrot,mcr}.
 %
@@
 
-We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate OS classes at Carnegie Mellon, Berkeley, and the University of Chicago.
+We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate operating systems classes at Carnegie Mellon, University of California at Berkeley, and University of Chicago.
 %We show many advantages over both conventional stateless model checking and single-pass data-race analysis.
@@
 
-The rest of the paper is organized as follows.
-\sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,
-\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation, %of Iterative Deepening,
-\sect{\ref{sec:soundness}} presents our proofs of convergence and of the soundness of our new false-positive data-race tactic,
-\sect{\ref{sec:eval}} presents our evaluation,
-\sect{\ref{sec:future}} discusses limitations and future work,
-\sect{\ref{sec:related}} surveys the related work,
-and \sect{\ref{sec:conclusion}} concludes.
+% TODO CAMREADY: uncomment if a reviewer complains about the absence of this
+%The rest of the paper is organized as follows.
+%\sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,
+%\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation, %of Iterative Deepening,
+%\sect{\ref{sec:soundness}} presents our proofs of convergence and of the soundness of our new false-positive data-race tactic,
+%\sect{\ref{sec:eval}} presents our evaluation,
+%\sect{\ref{sec:future}} discusses limitations and future work,
+%\sect{\ref{sec:related}} surveys the related work,
+%and \sect{\ref{sec:conclusion}} concludes.

### 2015-11-20 pldi-16/overview.tex
clean up todos
garth's final comments, pt1, up thru implementation

@@
 Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
-stateless model checkers focus on concrete observable failures such as assertions, deadlocks, and segfaults.
+stateless model checkers focus on concrete observed failures such as assertions, deadlocks, and segfaults.
 
@@
 
-The checker defines the granularity of thread interleavings by which {\em preemption points} it uses to switch threads.
+The checker defines the granularity of thread interleavings by the {\em preemption points} it uses to switch threads.
 Most model checkers \cite{chess,dbug-ssv} choose synchronization/thread library API boundaries for these points;
@@
 the threads do not hold the same lock,
-and no synchronization enforces the order of the thread transitions.
-Data races are not necessarily bugs, but represent suspicious violations of the locking discipline.
+and no synchronization enforces an order on the thread transitions.
+Data races are not necessarily bugs, but represent suspicious violations of common locking discipline.
 In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
 
-Though state-of-the-art MCs preempt only on synchronization events,
+Though state-of-the-art model checkers preempt only on synchronization events,
 many serious concurrency bugs are caused by data races leading to corrupted shared state.
-Figure~\ref{fig:example}'s thread interleaving is possible only with {\em data-race preemption points}:
-preempting a thread just before an instruction identified as part of a data race.
+Figure~\ref{fig:example}'s buggy interleaving is possible only with {\em data-race preemption points}:
+preempting just before an instruction identified as part of a data race.
 None of the state spaces in Figure~\ref{fig:id} contain this interleaving,
@@
 Compared to the former, the latter may output false positives, but avoids false negatives \cite{tsan},
-which is more appropriate for this work (\sect{\ref{sec:totalverif}}).
+in which a potential racing access pair is not reported, even though it could lead to a bug.
+We use the limited happens-before analysis, accepting some false positive overhead for the sake of total verification (\sect{\ref{sec:totalverif}}).
 
@@
 If they cannot be reordered at all, due to some other communication, it is a {\em false positive}.
-%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
-%For brevity, we count benign data races as a subset of false positives.
 

### 2015-11-20 pldi-16/paper.tex
some of haryadi's commence
clean up todos
garth's final comments, pt1, up thru implementation

@@
 
-% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
+% TODO CAMREADY: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
 \newcommand\landslide{\textsc{Landslide}}
@@
 is often intractable.
-Prior tools are limited to preempting only on synchronization APIs,
+Most prior tools are limited to preempting only on synchronization APIs,
 which reduces the space further, but can miss unsynchronized thread communication bugs.
@@
 Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
-It avoids concerns of state space size, but is prone to false positives.
+It avoids concerns of state space size, but is prone to false positives:
+spurious reports of access pairs that cannot be reordered to produce a bug.
 
@@
 %with data-race preemption points
-found 1.79x as many bugs and verified 4.25x as many tests compared to stateless model checking without data-race preemption points.
+found 1.8x as many bugs and verified 4.3x as many tests compared to stateless model checking without data-race preemption points.
 %Our evaluation shows this technique is
@@
 
-% TODO: Make data set, at least, public.
+% TODO CAMREADY: Make data set, at least, public.
 
@@
 %\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put final logs somewhere
-% POST-SUBMISSION TODO: Don't forget to anonymize the names of all your grupos.
+% TODO CAMREADY: Don't forget to anonymize the names of all your grupos.
 
@@
 %{\em removed for double-blind review}
-%Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
+%Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of chicago's, and CMU's OS classes respectively.
 %Thanks to Wind River for the use of their simulator \simics.
 %Thanks to
-%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, John Wilkes, and
+%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
 %the anonymous PLDI reviewers for their helpful comments.

### 2015-11-20 pldi-16/evaluation.tex
put subfigure labels on figure 2
clean up todos
garth's final comments pt2, all except the tough ones

@@
 % TODO: Is this ok? Too arrogant?
-We believe that merely finding a small handful of new real-world bugs is distressingly anecdotal,
+We believe that merely finding a small handful of new real-world bugs is largely anecdotal,
 and that our test suite's size allows for a more statistically significant comparison among MC and data-race testing strategies.
@@
 %\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
-Figure~\ref{fig:dowefindbugsfaster} plots a cumulative distribution of bugs found before a given elapsed CPU-time. % during its associated test.
+Figure~\ref{fig:dowefindbugsfaster} plots a cumulative distribution of bugs found before a given elapsed CPU consumption. % during its associated test.
 We tested several configurations of \quicksand~to compare against SSS-MC from multiple angles,
@@
 \end{figure}
+
+% TODO CAMREADY: ICB
+
+% TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
 
@@
 %even compared to the other test programs
-In \mxtest, in which we stop trusting the lock implementation's correctness,
+In \mxtest, in which we do not trust the lock implementation's correctness,
 the control experiment found dramatically fewer bugs (just 1)\footnote{
@@
 Though it often suffices to assume correctly-implemented locks,
-we consider this strong evidence that low-level synchronization code must be verified with data-race PPs.
+we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
 
@@
 Among those 69 true data-races, %which initially fit the malloc-recycle pattern,
-none of them exposed a new bug when used as a PP.
+none exposed a new bug when used as a PP.
 This suggests that for other data-race tools,
@@
 For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
-In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU-time.
+In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU consumption.
 Among these, 36 tests contained no data-race candidates whatsoever,
@@
 The first series represents the bugs found using PPs from single-pass data-race candidates,
-% TODO: re read racefuzzer/portend paper to confirm this is true.
+% not entirely true, as portend could be given data-race traces from an MC,
+%but they don't do it in their paper, so i feel comfortable making this claim
 i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
@@
 This experiment suggests that
-such tools could benefit from a similar feedback loop as Iterative Deepening.
+such tools could benefit from a similar feedback loop as used in Iterative Deepening.
 %i.e., discovering transitively-reachable data races while testing initial ones.

### 2015-11-20 pldi-16/intro.tex
garth's final comments pt2, all except the tough ones

@@
 
-We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate operating systems classes at Carnegie Mellon, University of California at Berkeley, and University of Chicago.
+We evaluate \quicksand~by testing \numstudence~student thread libraries and kernels from the undergraduate operating systems classes at Carnegie Mellon University, University of California at Berkeley, and University of Chicago.
 %We show many advantages over both conventional stateless model checking and single-pass data-race analysis.

### 2015-11-20 pldi-16/overview.tex
garth's final comments pt2, all except the tough ones

@@
 in which a potential racing access pair is not reported, even though it could lead to a bug.
-We use the limited happens-before analysis, accepting some false positive overhead for the sake of total verification (\sect{\ref{sec:totalverif}}).
+We use the limited happens-before analysis, accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
 

### 2015-11-20 pldi-16/paper.tex
garth's final comments pt2, all except the tough ones

@@
 \quicksand~incorporates data-race analysis to create new PPs tailored specifically to the program under test,
-and manages multiple model checker instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
+and automatically finds state spaces that are appropriately sized to complete in a given CPU budget.
+%and manages multiple model checker instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
 
-Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking alone.
+We achieve better bug-finding results than either single-pass data-race detection or single-state-space model checking alone,
 %We dramatically reduce false-positive data race reports,
@@
 %when the maximal state space is too large to explore
-We find new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
+finding new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
+By using \numstudence~student projects as our test suite, we also show the potential benefit as a debugging platform in educational settings.
 Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.
@@
 \quicksand~is open-source and its interface can be adapted to fit any tool similar to \landslide.
-We have also posted our evaluation's full data set. %and logs.
+We have also posted our evaluation's data set. %and logs.
 They are available at:

### 2015-11-20 pldi-16/related.tex
clean up todos
garth's final comments pt2, all except the tough ones

@@
 Inspect \cite{inspect} instruments source code by inserting wrapper calls around all accesses to potentially-shared data.
-It identifies such instructions in advance with an over-approximating alias analysis,
+It identifies such instructions in advance with an over-approximating static alias analysis,
 while \landslide~\cite{landslide} traces the memory locations of accesses at runtime.
@@
 improves performance using an interference analysis,
-which would allow future work to avoid tracing every memory access.
+which could allow future work to avoid tracing every memory access.
 %although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
@@
 Portend \cite{portend} is the most closely related work we have found.
-Based on single-pass data race reports, it tests alternate program executions to classify candidates in a taxonomy of likely severity.
+Based on reports from single-pass data-race analysis, it tests alternate executions to classify candidates in a taxonomy of likely severity.
 It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
@@
 However, Portend does not test alternate interleavings {\em in advance} of knowing any specific data races,
-which is necessary to expose certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
+which is necessary to find certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
 % Not really true.
@@
 device-like signal handlers)
-
-% TODO: Read OOPSLA 2015 "SATcheck, sat-directed SMC for SC/TSO"

### 2015-11-20 pldi-16/soundness.tex
garth's final comments pt2, all except the tough ones

@@
 \begin{theorem}
-If a bug can be exposed by any thread interleaving possible by preempting on all instructions,
+If a bug can be exposed by any thread interleaving possible by preempting on all instructions during a specific test,
 Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
@@
 
-We now show by contradiction there cannot be {\em multiple} data-race PPs which must both be enabled before either data race can be identified.
+We now show by contradiction there cannot be {\em multiple} data-race PPs which must all be enabled before either data race can be identified.
 Assume there does not exist a single transition $t_1 \in \mathcal{S}$ which alone can be split into $\{t_1',t_1''\}$ by a PP $q$,
@@
 
-We identify a particular class of false positive data-race candidates in which the memory was recycled by {\tt malloc} between the two accesses.
+We identify a particular class of false positive data-race candidates in which the associated memory was recycled by re-allocation between the two accesses.
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
@@
 
-Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning these false positives is sound even when such adversarial programs are considered.
+Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning all malloc-recycle candidates is sound, even considering adversarial programs such as Figure~\ref{fig:recycle-bug}.
 This makes it unnecessary to verify such data races by actively adding more preemptions,

### 2015-11-20 pldi-16/evaluation.tex
last of garth's commence

@@
 {\bf Finding the same bugs faster.}
-To show that Iterative Deepening is effective even for MC testing domains without data races, %such as message-passing distributed systems,
-we ran the test suite with \quicksand~configured to explore only subsets of the hard-coded mutex PPs
+To show that Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
+we configured \quicksand~to explore only subsets of the synchronization PPs
 (i.e., ignoring all data-race candidates).
@@
 Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
-QS-no-DR-PPs finds many more of the bugs much sooner.
-Hence, for arbitrary small CPU budgets,
-Iterative Deepening is likely to find bugs SSS-MC will miss.
+and is faster in the first 60 seconds due to less start-up overhead,
+QS-no-DR-PPs finds more of the bugs sooner thereafter.
+Hence, for modest CPU budgets,
+\quicksand~is likely to find bugs SSS-MC will miss.
 %and for more ambitiously-sized tests,

### 2015-11-20 pldi-16/overview.tex
last of garth's commence

@@
 in which only blocking operations like {\tt cond\_wait} or {\tt join} enforce the order.
-Compared to the former, the latter may output false positives, but avoids false negatives \cite{tsan},
+% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
+Compared to the former, the latter may report false positives, but avoids false negatives \cite{tsan},
 in which a potential racing access pair is not reported, even though it could lead to a bug.

### 2015-11-20 pldi-16/related.tex
last of garth's commence

@@
 %after finding data race candidates.
-This work compares the immediately resulting program states for differences,
+This work compares the program states immediately after the access pair for differences,
 preferring to err on the side of false positives.
@@
 although it uses random schedule fuzzing rather than stateless MC.
-While this technique can also classify malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
+While this technique can classify malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
 they require replaying the threads in a new interleaving.

### 2015-11-20 pldi-16/soundness.tex
name convergence lemma in paper

@@
 \newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
-\begin{theorem}
+\begin{theorem}[Convergence]
 If a bug can be exposed by any thread interleaving possible by preempting on all instructions during a specific test,
@@
 Initially, the statically-available synchronization API PPs are reachable. Reachability of data-race PPs is transitive.
+
 \begin{lemma}[Saturation of data-race PPs]

### 2016-03-10 oopsla-16/implementation.tex
Begin banging PLDI paper into shape for OOPSLA.

--- 2015-11-20 pldi-16/implementation.tex
@@
 
-We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
+We chose \landslide~\cite{landslide} as our MC tool due to its ability to trace execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
 \landslide~implements DPOR \cite{dpor},
 %\cite{dpor},
-state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
-% this line can be cut if space is needed
-It can test both user-level and kernel-level code (although it is limited to timer nondeterminism).
+state space estimation \cite{estimation}, and a hybrid
+%lockset/happens-before
+data-race analysis \cite{hybriddatarace}.
+It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
 % joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
 Its bug-detection metrics are assertion failure, deadlock, segfault, heap checking \cite{valgrind}, and a heuristic infinite loop check.
+It can test both userspace and kernel code (although it is limited to timer nondeterminism),
+and runs programs in a full-system hardware simulator \cite{simics}.
 

### 2016-03-10 oopsla-16/intro.tex
Begin banging PLDI paper into shape for OOPSLA.

--- 2015-11-20 pldi-16/intro.tex
@@
 		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
-	\item A large evaluation in which \quicksand~compares favorably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
+	\item A large evaluation in which \quicksand~compares favourably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
 \end{enumerate}

### 2016-03-10 oopsla-16/paper.tex
Begin banging PLDI paper into shape for OOPSLA.

--- 2015-11-20 pldi-16/paper.tex
@@
+%-----------------------------------------------------------------------------
 %
-% LaTeX template for prepartion of submissions to PLDI'15
+%               Template for sigplanconf LaTeX Class
 %
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'15 web site.
+% Name:         sigplanconf-template.tex
 %
-\documentclass[pldi]{sigplanconf-pldi15}
+% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
+%               file for SIGPLAN conference proceedings.
+%
+% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
+%               sigplanconf-guide.pdf
+%
+% Author:       Paul C. Anagnostopoulos
+%               Windfall Software
+%               978 371-2316
+%               paul@windfall.com
+%
+% Created:      15 February 2005
+%
+%-----------------------------------------------------------------------------
 
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                  % format URLs
-\usepackage{listings}          % format code
-\usepackage{enumitem}      % adjust spacing in enums
+\documentclass[10pt,times,numbers]{sigplanconf}
+\usepackage[1stsubmission]{oopsla2016}
+
+% The following \documentclass options may be useful:
+
+% preprint      Remove this option only once the paper is in final form.
+% 10pt          To set in 10-point type instead of 9-point.
+% 11pt          To set in 11-point type instead of 9-point.
+% numbers       To obtain numeric citation style instead of author/year.
+
+% template-provided usepackages
+\usepackage{amsmath}
+% custom ones copied from pldi template
+\usepackage{url}
+\usepackage{enumitem}
 \usepackage[linesnumbered,ruled]{algorithm2e}
@@
 \usepackage{amsthm}
-%\usepackage{setspace}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-%\doublespacing
+\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}
+\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
 
@@
 
+% ???
+%\newcommand{\cL}{{\cal L}}
+
 \begin{document}
-
-%
-% any author declaration will be ignored  when using 'plid' option (for double blind review)
-%
 
@@
 
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{OOPSLA '16}{November 2--4, 2016, Amsterdam, the Netherlands}
+\copyrightyear{2016}
+\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
+\copyrightdoi{nnnnnnn.nnnnnnn}
+
+% Uncomment the publication rights you want to use.
+%\publicationrights{transferred}
+\publicationrights{licensed}     % this is the default
+%\publicationrights{author-pays}
+
+%\titlebanner{banner above paper title}        % These are ignored unless
+%\preprintfooter{short description of paper}   % 'preprint' option specified.
+
 \title{Stateless Model Checking with Data-Race Preemption Points}
+
 \authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
 \authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
+%\authorinfo{\vspace{24pt}}{}{}
 
 \maketitle
+
 \begin{abstract}
@@
 %both at finding the same bugs faster and at finding new bugs entirely.
+\end{abstract}
 
-\end{abstract}
+\category{D.2.4}{Software Engineering}{Software/Program Verification}
+\category{D.2.5}{Software Engineering}{Testing and Debugging}
+%
+%% general terms are not compulsory anymore,
+%% you may leave them out
+%\terms
+%term1, term2
+%
+%\keywords
+%keyword1, keyword2
 
@@
 
+% TODO: proofs->appeendix
+\appendix
+\section{Appendix Title}
+
+This is the text of the appendix, if you need one.
+
+\acks
+
+Acknowledgments, if needed.
+
+% We recommend abbrvnat bibliography style.
+
 \bibliographystyle{abbrvnat}

### 2016-03-10 oopsla-16/soundness.tex
Begin banging PLDI paper into shape for OOPSLA.

--- 2015-11-20 pldi-16/soundness.tex
@@
 The proof has two parts:
-first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
+first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behaviour;
 second, we show that Iterative Deepening will eventually detect all such data races.
@@
 We identify a particular class of false positive data-race candidates in which the associated memory was recycled by re-allocation between the two accesses.
-Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
+Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
 If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.

### 2016-03-15 oopsla-16/proof/recycle.tex
Import proof

--- 2015-11-11 pldi-16/proof/recycle.tex
@@
 If the data race is not a false positive, then the second access must not change locations based on the middle malloc's return value.
-WLOG, we say that thread 1 (T1) does the first access, called {\tt x1}, and thread 2 (T2) does the second, {\tt x2}.
+WLOG, we say that thread 1 (T1) does the first access, called $a_1$, and thread 2 (T2) does the second, $a_2$.
 
 \begin{lemma}
-	If DPOR will reorder {\tt x2} to before {\tt x1}, and the location of access {\tt x2} doesn't change,
+	If DPOR will reorder $a_2$ to before $a_1$, and the location of access $a_2$ doesn't change,
 	then a non-malloc-recycle data race or a use-after-free bug will be identified.
@@
 	\item T1 free, T2 malloc (as shown in Figure~\ref{fig:recycle-bug}).
-The malloc will go with {\tt x2} to before the free, and because the allocation of concern has not been freed yet, will return a different value.
-Hence {\tt x1} and {\tt x2} will be in the same allocation; hence the race is not malloc-recycle anymore.
-	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but {\tt x2} now occurs before then, being in the same, older, allocation.
-	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either {\tt x1} or {\tt x2}. The memory will be recycled and both accesses will appear to be in the later allocation.
+The malloc will go with $a_2$ to before the free, and because the allocation of concern has not been freed yet, will return a different value.
+Hence $a_1$ and $a_2$ will be in the same allocation; hence the race is not malloc-recycle anymore.
+	\item T1 free, T1 malloc. Same as above, but the malloc does not move. The middle malloc will still recycle the memory, but $a_2$ now occurs before then, being in the same, older, allocation.
+	\item T2 free, T2 malloc. Both the free and re-malloc will occur before either $a_1$ or $a_2$. The memory will be recycled and both accesses will appear to be in the later allocation.
 	\item T2 free, T1 malloc. The free gets reordered earlier, the malloc stays put, and the accesses go in between. This will be a use-after-free bug.
@@
 
-The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to {\tt x1}) that affect what T2's malloc returns after being reordered?
+The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to $a_1$) that affect what T2's malloc returns after being reordered?
 However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
-%If such a malloc could affect {\tt x2}, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect {\tt x1} or {\tt x2} could only convert the program back into a false-positive scenario;
+%If such a malloc could affect $a_2$, the data race would be a false positive after all.
+In general, adding spurious mallocs that could affect $a_1$ or $a_2$ could only convert the program back into a false-positive scenario;
 and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
@@
 By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
-We must ask, which thread did the malloc which returned {\tt x1}'s address in the first place?
+We must ask, which thread did the malloc which returned $a_1$'s address in the first place?
 Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
@@
 \begin{proof}
-Because T1 had the first access, there was a thread switch between the original malloc and {\tt x1}, as well as between {\tt x1} and {\tt x2}.
-By Definition~\ref{def:transition}, each switch will be a PP.
-By Definition~\ref{def:dpor}, DPOR will reorder {\tt x2} to before {\tt x1},
-and because T1 is not involved in the logic determining {\tt x2}, the access's location stays the same.
+Because T1 had the first access, there was a thread switch between the original malloc and $a_1$, as well as between $a_1$ and $a_2$. By Definition~\ref{def:transition}, each switch will be a PP.
+By Definition~\ref{def:dpor}, DPOR will reorder $a_2$ to before $a_1$,
+and because T1 is not involved in the logic determining $a_2$, the access's location stays the same.
 Lemma~\ref{lem:reorder} finishes.
@@
 \begin{proof}
-We must guarantee there will be a PP during T1 before its {\tt x1} access, but after whatever action it took to communicate the heap address to T2\footnote{
-Note why we assert the publish action must come before {\tt x1}: otherwise, T2 couldn't be reordered to race {\tt x2} with {\tt x1} before T1 communicated the address, and it would be a false positive after all.}.
+We must guarantee there will be a PP during T1 before its $a_1$ access, but after whatever action it took to communicate the heap address to T2\footnote{
+Note why we assert the publish action must come before $a_1$: otherwise, T2 couldn't be reordered to race $a_2$ with $a_1$ before T1 communicated the address, and it would be a false positive after all.}.
 
@@
 
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and {\tt x1}?
+We ask: was there a synchronization event (unlock or message-pass) between the publish action and $a_1$?
 If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during {\tt x1} (and the fact that there is no MHB cannot change).
-For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing {\tt x1}; T2 doesn't have that yet),
-which we'll call {\tt p}.
-By the previous two statements, {\tt p} must be a data race of its own.
+If not, then  T1's lockset during publish will be the same as the lockset during $a_1$ (and the fact that there is no MHB cannot change).
+For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing $a_1$; T2 doesn't have that yet),
+which we'll call $p$.
+By the previous two statements, $p$ must be a data race of its own.
 %(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
@@
 
-Because it's possible for {\tt p} to {\em also} be a malloc-recycle data race,
+Because it's possible for $p$ to {\em also} be a malloc-recycle data race,
 as shown in Figure~\ref{fig:induction},
@@
 
-\newcommand\publish[1]{{\tt p}$_{#1}$}
+\newcommand\publish[1]{$p_{#1}$}
 \begin{itemize}
@@
 Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location {\tt p}.
+a PP will be identified on the publish location $p$.
 Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
 % XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
-As long as T2's execution occurs after the publish, it will receive the same value for its {\tt x2}, so the location of the data race does not change.
+As long as T2's execution occurs after the publish, it will receive the same value for its $a_2$, so the location of the data race does not change.
 Lemma~\ref{lem:reorder} finishes.
-%Hence for any type of publish action, there will be a PP between it and {\tt x1}, hence DPOR will reorder {\tt x2} to before {\tt x1} (without changing the address at which {\tt x2} occurs), and Lemma~\ref{lem:reorder} finishes.
+%Hence for any type of publish action, there will be a PP between it and $a_1$, hence DPOR will reorder $a_2$ to before $a_1$ (without changing the address at which $a_2$ occurs), and Lemma~\ref{lem:reorder} finishes.
 
@@
 
-%We will also say that T1 does the initial malloc which first returns {\tt x1}'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
+%We will also say that T1 does the initial malloc which first returns $a_1$'s address; although certainly T2 or even a third thread could have malloced it, that would involve a thread switch to thread 1, and hence a PP by Definition~\ref{def:transition}.
 %As we will see, the main challenge of the proof is showing that enough PPs will exist at appropriate points, so adding more PPs by considering other threads for the initial malloc can only make the proof easier.
@@
 \begin{theorem}[Soundness of eliminating malloc-recycle races]
+	\label{thm:recycle}
 	If a malloc-recycle data race is not a false positive, DPOR will reorder threads such that either the same accesses will still race without fitting the malloc-recycle pattern, or a use-after-free bug will be reported immediately.
@@
 
-Actually, there is another way to share the allocation's address without T1 and T2 communicating outside of {\tt x1}/{\tt x2}: one thread overruns a {\em different} heap block which happened to be adjacent to the one containing {\tt x1}/{\tt x2} (call them the ``neighbour block'' and ``real block'' respectively).
+If we relax the ``sharing heap addresses'' assumption, there is another way to share the allocation's address without T1 and T2 communicating outside of $a_1$/$a_2$.
+One thread can overrun a {\em different} heap block which happened to be adjacent to the one containing $a_1$/$a_2$ (call them the ``neighbour block'' and ``real block'' respectively).
 Figure~\ref{fig:overrun} shows an example.
-Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our model.
-So we must also consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
+Heap overrun bugs are quite serious \cite{eternal-war}, so we do not wish to exclude them from our proof.
+
+In the evaluation of our main paper, we used the full ``sharing heap addresses'' assumption to heuristically reduce state space size,
+and not test reorderings of threads which could only change the addresses allocated by {\tt malloc}.
+This restricted the scope of our tests to exclude testing for heap-overflow bugs.
+We consider this justified because new techniques \cite{sparc-ssm} have recently emerged for finding such bugs quickly without the need for data-race PPs.
+However, for users wishing to test for this class of bug simultaneously with other concurrency bugs, we show in this section how to strengthen the configuration of DPOR to cover the weakened assumption.
+
+Note also that even if malloc-recycle candidates are {\em not} suppressed,
+a DPOR which ignores malloc's internal metadata accesses would still be unsound with respect to these bugs.
+We illustrate this in Figure~\ref{fig:overrun-notenough}: even if there were PPs in arbitrary places, the two threads' transitions do not collide in this interleaving on anything but {\tt malloc}'s internal metadata, and hence DPOR would not attempt to reorder them.
+Hence, what we have already proved so far can be viewed as showing that suppressing malloc-recycle candidates is sound with respect only to the classes of bugs which DPOR is already sound to.
 
@@
 \end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls which produce {\tt x1} and {\tt x2} must not be disrupted.}
+\caption{Goal interleaving of Figure~\ref{fig:overrun}.
+To ensure collision, the sequence of malloc calls which produce $a_1$ and $a_2$ must not be disrupted, compared to the trace which originally exposed the malloc-recycle race.}
 \label{fig:overrun-goal}
@@
 
-% TODO: Replace this with some real discussion.
-
-%{\bf Being Owned.}
-%The second best thing I can think to do is say, put "speculative" PPs on the end of malloc, and enable them only when that malloc is involved in a FRM situtation.
-Then, measure among all FRM sitations (gs12642/landslide/pldi/p2-live/frms) how much bigger the state spaces get (and whether that affects our bugfinding ability/cputimes).
-%
-%
-%The first best thing I can think to do is to argue that this case doesn't disrupt soundness because the original model we were working with was already unsound.
-If the other interleaving ("the best dpor could find") occurred first, dpor would not identify a memory conflict at all, and would not reorder to find the original interleaving (with the FRM false pos) at all.
-this is because dpor ignores malloc's internal accesses from the shm conflict relation.
-
-what we are trying to prove is that, if we START WITH a sound dpor, and add this FRM-elimination, you end up with a sound dpor.
-BUT, the original one was not sound with respect to this bug to begin with.
-
-then you can simply point out that if you want dpor to be sound wrt this bug, you have to put a PP on the end of malloc (other spots, beginning of malloc and either end of free, are not needed), and also make the shm conflix relation account for the way malloc reordering can change the address collision issue.
+% For shown it is in complexity class, not possible to complete,
+To soundly suppress heap-overrun malloc-recycle candidates,
+%we must consider the cases where T1 mallocs the real block and T2 overflows a neighbour, and vice versa.
+we must strengthen our configuration of DPOR as follows:
+accesses from {\tt malloc}'s internal implementation will not be ignored when computing the shared memory conflict relation,
+and we will automatically insert a synchronization API PP at the start and end of each {\tt malloc} and {\tt free} call.
+% TODO CAMREADY: uncomment footnote (it's commented because it seems like it's advertising landslide too much)
+%\footnote{
+%	As a practical matter, in a large codebase where parts of it are trusted to avoid heap overflows, while others are not,
+%	a user could use {\sc Landslide}'s {\tt within\_function} command to restrict such PPs, as we describe in the main paper.
+%}.
 
 \begin{lemma}
-	asdfa
+	If T1 and T2 each malloced neighbouring blocks, and collided based on pointer arithmetic involving no shared memory accesses,
+	%did not communicate via any publish action $p$,
+	%and the DPOR configuration is strengthened as above,
+	DPOR will reorder the threads to uncover a non-malloc-recycle data race.
 	\label{lem:leia} % "there is another"
@@
 
-%\subsection{Malloc with internal mutexes}
-%
-%First we assume that the malloc implementation is guarded by
-%
-%
-%Mario man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
-%
-%
-%\subsection{Malloc as a magic primitive}
-%
-%Actually, regardless of malloc's internal synchronization details, in virtually all cases we wish to avoid unnecessary PPs on its boundaries during MC, because unlike other uses of mutexes, it doesn't indicate any concurrent behaviour is going on.
-%Much like the {\tt mx\_test} in our paper, the malloc implementation could itself be tested for concurrency bugs, but in all other cases,
-%we drastically reduce the state space size by assuming it is correct and avoiding PPs on it, treating it instead as a ``magic primitive''.
-%
-%Hence, we provide another proof that does not require
-%
+\begin{proof}
+WLOG, let T1's access in the original malloc-recycle race have come first.
+We require that our strengthened DPOR will reorder T2's racing access to before T1's, such that both still occur on the same address.
+There will be a PP in T1's execution between its {\tt malloc} call and the subsequent racing access.
+If there are no other memory conflicts between T1 and T2, then by the soundness of DPOR, this PP suffices to reorder without changing the address.
+Otherwise, let $p$ be the latest conflicting access by T1 before its access $a_1$.
+By the same inductive reasoning as we used in Lemma~\ref{lem:han}, Iterative Deepening will add a data-race PP on $p$.
+As T1 has no further conflicts between $p$ and $a_1$, T2's $a_2$ will be reordered between them without changing the address.
+Lemma~\ref{lem:reorder} finishes.
+\end{proof}
+
+Theorem~\ref{thm:recycle} is modified to simply include this lemma in addition to Lemmas~\ref{lem:greedo} and \ref{lem:han}.
+
+% TODO CAMREADY cite landslide
 %%\cite{landslide} % uncomment post-double-blind review

### 2016-03-15 oopsla-16/paper.tex
start doing revisions for oopsla

@@
 
-% template-provided usepackages
 \usepackage{amsmath}
-% custom ones copied from pldi template
 \usepackage{url}
@@
 \usepackage{graphicx}
+\usepackage{courier} % TODO: remove this? and get nicer fonts?
 \usepackage{amsthm}
@@
 \newtheorem{definition}{Definition}
-
-% ???
-%\newcommand{\cL}{{\cal L}}
 
@@
 
+% TODO: remove all changebars before submission
+%\setcounter{changebargrey}{0}
+%\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
+%\newcommand\revision[1]{\cbstart#1\cbend}
+%\newcommand\revision[1]{#1}
+\definecolor{purple}{RGB}{192,0,192}
+\newcommand\revision[1]{\hilight{purple}{#1}}
+
 \title{Stateless Model Checking with Data-Race Preemption Points}
@@
 
-We present an {\em Iterative Deepening} framework for stateless model checking,
-which manages the exploration of many state spaces using different preemption points.
+\revision{We present \quicksand, a new stateless model checking framework}~which
+manages the exploration of many state spaces using different preemption points.
 It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
@@
 
-\category{D.2.4}{Software Engineering}{Software/Program Verification}
-\category{D.2.5}{Software Engineering}{Testing and Debugging}
+\revision{\category{D.2.4}{Software Engineering}{Software/Program Verification}}
+%\category{D.2.5}{Software Engineering}{Testing and Debugging}
 %

### 2016-03-15 oopsla-16/related.tex
start doing revisions for oopsla

--- 2015-11-20 pldi-16/related.tex
@@
 Hence, they will miss any bugs that require interleaving threads at instruction granularity during a data race.
-CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
+%CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
+\revision{\cite{chess} mentions the ability of CHESS to add PPs using a single-pass data-race analysis,
+% FIXME... idk, "discusses (without explicit evaluation)..., but, as our evaluation shows..."?
+but does not discuss either the increase in bug-finding capacity or the soundness properties.
+Our convergence theorem (\sect{\ref{sec:soundness}}) is a natural extension of \cite{chess-icb}'s Theorem 3,
+which alone provides soundness only for race-free programs.}
 
@@
 %DPOR for relaxed memory models \cite{tsopso},
-and SAT-directed MC \cite{satcheck}.
+SAT-directed MC \cite{satcheck},
+\revision{and R4 for event-driven applications \cite{r4}}.
 These are all compatible with our technique.
@@
 \cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
-DroidRacer \cite{droidracer} and CAFA \cite{cafa} target
-%event-driven
-Android applications, using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
+\revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
+%DroidRacer \cite{droidracer} and CAFA \cite{cafa} target
+%%event-driven
+%Android applications,
+using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
 Like \landslide, DataCollider \cite{datacollider} offers data-race techniques for kernel code.

### 2016-03-16 oopsla-16/intro.tex
Write expanded explanation of HB for background section.

@@
 \begin{enumerate}
-	\item {\em Iterative Deepening}, a new technique for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
+	\item {\em Iterative Deepening}, a new \revision{algorithm}~for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
 	\item A proof of convergence, showing that should it be possible in the given CPU budget,

### 2016-03-16 oopsla-16/overview.tex
Write expanded explanation of HB for background section.

--- 2015-11-20 pldi-16/overview.tex
@@
 the threads do not hold the same lock,
-and no synchronization enforces an order on the thread transitions.
-Data races are not necessarily bugs, but represent suspicious violations of common locking discipline.
+and no synchronization enforces an order on the thread transitions \revision{(the {\em Happens-Before} relation)}.
+Data races \revision{do not always lead to failures}, but represent suspicious violations of common locking discipline.
 In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
@@
 
-Prior work \cite{hybriddatarace} distinguishes between {\em pure happens-before} \cite{lamport-clocks},
-in which accesses with a lock release/acquire in between are not considered concurrent,
-and {\em limited happens-before},
-in which only blocking operations like {\tt cond\_wait} or {\tt join} enforce the order.
-Compared to the former, the latter may output false positives, but avoids false negatives \cite{tsan},
-in which a potential racing access pair is not reported, even though it could lead to a bug.
-We use the limited happens-before analysis, accepting some false positive overhead for the sake of total verification (\sect{\ref{sec:totalverif}}).
+% TODO: if you need, you can recover a line of text by abbreviating "HB".
+\revision{
+	{\bf Variants of Happens-Before (HB).}
+	Most prior work focuses on {\em Pure HB} \cite{lamport-clocks} as the order relation between accesses.
+\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
+it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
+as shown in the example program in Figure~\ref{fig:hb-example}(a). % TODO
+We call such unreported access pairs {\em false negatives}.
+\cite{hybriddatarace} introduces the {\em Limited HB} relation,
+which will report such potential races
+by considering only blocking operations like {\tt cond\_wait} to enforce the order.
+However, consider the similar program in Figure~\ref{fig:hb-example}(b), % TODO
+in which the access pair ceases to exist in the alternate interleaving.
+Limited HB will report all potential races, avoiding false negatives \cite{tsan},
+but at the cost of necessarily reporting some such {\em false positives}.
+Finally, the {\em Causally-Precedes} relation \cite{predictive-dr} strikes a middle ground,
+extending Pure HB to additionally report a subset of potential races while soundly avoiding false positives.
+
+Stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx}.
+However, this work incorporates data-race analysis in an internal feedback loop,
+using MC to automatically check each potential race
+and reporting to the user only of directly observed failures.
+%Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})
+%% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
+Hence, we employ Limited HB in this paper},
+accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{c}
+	\revision{
+\begin{tabular}{rll}
+	& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~// A1} & \\
+	2 & \texttt{mutex\_lock(\&mx);} & \\
+	3 & \texttt{mutex\_unlock(\&mx);} & \\
+	4 & & \texttt{mutex\_lock(\&mx);} \\
+	5 & & \texttt{mutex\_unlock(\&mx);} \\
+	6 & & \texttt{\hilight{brickred}{x++;}~// A2} \\
+\end{tabular}
+}
+\\
+\revision{\normalsize (a) True potential data race.}
+\\
+\\
+\revision{
+\begin{tabular}{rll}
+	%& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+	%& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~// B1} & \\
+	2 & \texttt{mutex\_lock(\&mx);} & \\
+	3 & \texttt{y = true;} & \\
+	4 & \texttt{mutex\_unlock(\&mx);} & \\
+	5 & & \texttt{mutex\_lock(\&mx);} \\
+	6 & & \texttt{bool tmp = y;} \\
+	7 & & \texttt{mutex\_unlock(\&mx);} \\
+	8 & & \texttt{if (tmp)} \\
+	9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
+\end{tabular}
+}
+\\
+\revision{\normalsize (b) No data race in any interleaving.}
+\end{tabular}
+\caption{\revision{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
+Applying Pure HB to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
+while using Limited HB on (b) will produce a false alarm on B1/B2.}}
+\label{fig:hb-example}
+\end{figure}
+
+% Old explanation.
+%Prior work \cite{hybriddatarace} distinguishes between {\em pure happens-before} \cite{lamport-clocks},
+%in which accesses with a lock release/acquire in between are not considered concurrent,
+%and {\em limited happens-before},
+%in which only blocking operations like {\tt cond\_wait} or {\tt join} enforce the order.
+%Compared to the former, the latter may report false positives, but avoids false negatives \cite{tsan},
+%in which a potential racing access pair is not reported, even though it could lead to a bug.
+%We use the limited happens-before analysis, accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
 
@@
 we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-Because data-race analysis is prone to false-positives,
-we classify unprotected memory access pairs separately from bugs,
-calling such pairs {\em data race candidates}.
+% old explanation
+%Because data-race analysis is prone to false-positives,
+%we classify unprotected memory access pairs separately from bugs,
+\revision{We refer to racing (or potentially-racing) memory access pairs as}~{\em data race candidates}.
 Should a future interleaving, preempting during such accesses,
-lead to a failure, then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
-If they cannot be reordered at all, due to some other communication, it is a {\em false positive}.
+lead to \revision{an observable}~failure, then we report a {\em data-race bug}.
+Otherwise, if the access pair can be reordered \revision{to be simultaneous}, but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
+If they cannot be reordered at all, due to some other communication \revision{such as in Figure~\ref{fig:hb-example}(b)}, it is a {\em false positive}.
 

### 2016-03-17 oopsla-16/evaluation.tex
Expand related work to address PLDI reviewer concerns.

--- 2015-11-20 pldi-16/evaluation.tex
@@
 and \rwldgr~for r/w locks.
-For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
+For \mxtest, \paradise, and \paraguay, we used the {\tt without\_function} command to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
 and for \mxtest~we enabled \landslide's mutex-testing option
 (see \sect{\ref{sec:landslide}}).
-We tested Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
+We tested Pintoses with 3 programs from the class's provided test suite: \prisema, a test of the kernel scheduling algorithm,
 \alarmsimul, for the timer sleep routine,
@@
 We ran all tests on 12-core 3.2 GHz Xeon machines. %with 12GB of RAM.
-In total, our evaluation comprises 629 tests (pairs of a test program and a Pintos or P2),
+In total, our evaluation comprises 629 unique tests (pairs of a test program and a Pintos or P2),
 at least 163 of which could expose bugs.

### 2016-03-17 oopsla-16/overview.tex
Expand related work to address PLDI reviewer concerns.

@@
 \subsection{Data Race Detection}
+\label{sec:overview-dr}
 
@@
 using MC to automatically check each potential race
-and reporting to the user only of directly observed failures.
+and reporting to the user only directly observed failures.
 %Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})

### 2016-03-17 oopsla-16/related.tex
Expand related work to address PLDI reviewer concerns.

@@
 
-{\bf Other techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
+{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
 %DPOR for relaxed memory models \cite{tsopso},
@@
 %providing a strong argument for \quicksand.
+
+\revision{{\bf Restricting preemptions.}
+Preemption Sealing \cite{sealing} pre\-sents a mechanism similar to the {\tt within\_function} command (\sect{\ref{sec:landslide}}) for users to manually restrict preemptions.
+It demonstrates the need to consider subsets of PPs,
+as well as developers' willingness to limit a test case's scope so the resulting state space may be fully verified (\sect{\ref{sec:totalverif}}).
+%It justifies our assumption that expert
+%programmers can customize small test cases which, when fully verified,
+%provide a correctness assurance much stronger than the assurance from
+%a partially-complete exploration of a huge state space.
+Probabilistic Concurrency Testing (PCT) \cite{randomized-scheduler}
+is a randomized algorithm that can quantify the probability of uncovering bugs.
+PCT is optimized for tests with impossibly large state spaces,
+while for more approachable tests,
+its incompatibility with DPOR would significantly delay completion.
+Future work could use ETAs to heuristically switch between DPOR and PCT}.
 Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to our work,
@@
 when full completion is intractable (\sect{\ref{sec:future}}).
-However, ICB is limited to a fixed set of PPs, and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+However, ICB is limited to a fixed set of PPs,
+\revision{cannot estimate remaining time to total verification},
+and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
 
@@
 Many advances have been made on the false-positive data race problem since it was first introduced in \cite{eraser}.
-\cite{hybriddatarace} and \cite{tsan} combine the lockset and happens-before analyses into a hybrid technique, which we employ.
+\cite{hybriddatarace} and \cite{tsan}
+%combine the lockset and happens-before analyses into a hybrid technique,
+\revision{introduce Limited Happens-Before}, %(\sect{\ref{sec:overview-dr}})},
+which we employ,
+\revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.
 \revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
@@
 using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
-Like \landslide, DataCollider \cite{datacollider} offers data-race techniques for kernel code.
+Like \landslide, DataCollider \cite{datacollider} \revision{finds data races in}~kernel code.
 IFRit \cite{ifrit}

### 2016-03-17 oopsla-16/related.tex
more relwork

@@
 %CHESS provides a data-race analysis to report any such violations of its concurrency model to the user, but does not incorporate data-race candidates as PPs in future tests.
-\revision{\cite{chess} mentions the ability of CHESS to add PPs using a single-pass data-race analysis,
+\revision{\cite{chess} discusses the ability of CHESS to add PPs using a single-pass data-race analysis,
 % FIXME... idk, "discusses (without explicit evaluation)..., but, as our evaluation shows..."?
-but does not discuss either the increase in bug-finding capacity or the soundness properties.
+but does not evaluate either the increase in bug-finding capacity or the soundness properties.
 Our convergence theorem (\sect{\ref{sec:soundness}}) is a natural extension of \cite{chess-icb}'s Theorem 3,
@@
 which we do not yet account for in our proofs (\sect{\ref{sec:soundness}}).
-Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial state spaces in their evaluation could be completed.
+Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial \revision{tests}~in their evaluation could be completed,
+\revision{which motivates \quicksand's CPU-budget-oriented approach.}
 %providing a strong argument for \quicksand.

### 2016-03-19 oopsla-16/evaluation.tex
Fix miscounting in QS bugs, 163->157; and make ICB bug total the new denominator for claimed x-factor

@@
 In total, our evaluation comprises 629 unique tests (pairs of a test program and a Pintos or P2),
-at least 163 of which could expose bugs.
+% at least 157 of which could expose bugs.
+at least 165 of which could expose bugs. % QS bugs + bugs where SSSMC found but QS didnt
+% TODO: update this number with any new bugs found by PREEMPT_EVERYWHERE
 
@@
 
+% TODO: Nix discussion of QS-no-DR-bugs. Just do QS-no-DR-PPs.
+
+% TODO: Nix this discussion.
 {\bf Finding the same bugs faster.}
@@
 We find significantly more bugs within any CPU budget greater than 90 seconds,
-and ultimately conclude 10 CPU-hours with 179\% as many bugs in total.
+and ultimately conclude 10 CPU-hours with 151\% as many bugs in total.
 
@@
 
+% TODO: Incorporate the following point.
+% "Before the break-even point at 45 seconds (or whatever), SSS-MC approaches seem to outperform \quicksand,
+% but converting this CPU-time advantage into faster wall-clock performance is itself an open research problem \cite{parallel-dpor}.
+
+% TODO
+% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
+Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
+
 \begin{table*}[t]
@@
 	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+		% TODO: Update SSSMC column to have more favourable ICB numbers.
 		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
@@
 		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
-		{\tt join} 	& 79	& 14 	& 25	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
+		{\tt join} 	& 79	& 14 	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
 		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
-		{\tt sem} 	& 79	& 10	& 17	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
-		{\tt signal} 	& 79	& 4	& 12	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
-		{\tt rwlock} 	& 79	& 26	& 30	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
+		{\tt sem} 	& 79	& 10	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
+		{\tt signal} 	& 79	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
+		{\tt rwlock} 	& 79	& 26	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
 		\hline
@@
 		\hline
-		{\bf Total}	& 629	& 91	& 163	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
+		{\bf Total}	& 629	& 91	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
 	\end{tabular}

### 2016-03-21 oopsla-16/evaluation.tex
Write about experimental setup.

@@
 				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
+			\revision{
+			\item Does Iterative Deepening provide more full verifications of bug-free programs more quickly?
+			}
 		\end{enumerate}
@@
 For all tests, we used {\tt without\_function} to blacklist PPs on {\tt malloc}'s internal lock.
-We ran all tests on 12-core 3.2 GHz Xeon machines. %with 12GB of RAM.
 In total, our evaluation comprises 629 unique tests (pairs of a test program and a Pintos or P2),
@@
 
+\revision{
+\subsection{Experimental Setup}
+
+To evaluate the benefits of data-race PPs and Iterative Deepening separately, we ran the test suite under \quicksand~in two different configurations, each of which was given a 1-hour budget and 10 CPUs for each test.
+\begin{itemize}
+	\item {\bf QS-DR}: \quicksand~with data-race PPs enabled, and
+	\item {\bf QS-Sync-Only}: \quicksand~with PPs still seeded as described in \sect{\ref{sec:initial-pp}}, but never adding new PPs from reported data-race candidates.
+\end{itemize}
+
+We represented the MC state-of-the-art with 3
+%different
+configurations of stand-alone \landslide~on the same test suite:
+\begin{itemize}
+	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with plain DPOR \cite{dpor},
+	\item {\bf SSS-MC-ICB}: With PPs as above, but using ICB \cite{chess-icb} with BPOR \cite{bpor} instead of DPOR to find bugs faster, and
+	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
+		Using ICB+BPOR, configured to insert PPs on any shared memory access
+		(determined at runtime, excluding threads' accesses to their own stacks),
+		which in principle includes all possible data-race PPs.
+\end{itemize}
+Because parallelizing DPOR/ICB during SSS-MC is an open research problem \cite{parallel-dpor},
+we instead gave control experiments 10 hours per test with 1 CPU.
+%To match this,
+\quicksand~reports the CPU-time spent alongside the wall-clock time for a direct comparison
+(although, with the growing importance of multicore, \quicksand's inherent parallelism is a convenient benefit).
+%Finally,
+All tests ran on 12-core 3.2 GHz Xeon W3670 machines. %with 12GB of RAM.
+}
+
 \subsection{Comparing Iterative Deepening to SSS-MC}
@@
 
-% XXX: I hate this paragraph. I have no idea what order to put the sentences in.
-To compare to SSS-MC, we ran a control experiment for each test, running \landslide~on a single state space with all synchronization PPs enabled in advance (and no data-race PPs).
-We gave each \quicksand~test 10 CPUs for 1 hour each. % XXX
-Though \landslide~does not implement parallel DPOR \cite{parallel-dpor}, we compensated by giving each control test 1 CPU for 10 hours,
-%then dividing all associated times by 10 (simulating perfect parallelism).
-and instrumenting \quicksand~to report total CPU-hours rather than wall-clock time.
-%Pragmatically, with the growing importance of multicore, \quicksand's inherent parallelism is convenient to say the least.
-%Pragmatically, with most future performance improvements coming from increased core counts, \quicksand's inherent parallelism is convenient to say the least.
-(With the growing importance of multicore for performance, \quicksand's inherent parallelism is another convenient benefit.)
-
+% TODO: Cut/rearrange.
 %\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
@@
 
-% TODO CAMREADY: ICB
-
 % TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
-
 % TODO: Nix discussion of QS-no-DR-bugs. Just do QS-no-DR-PPs.

### 2016-03-21 oopsla-16/implementation.tex
another mention of sealing
Write about experimental setup.

@@
 data-race analysis \cite{hybriddatarace}.
+\revision{It can optionally replace DPOR with Iterative Context Bounding (ICB) \cite{chess-icb},
+for which it uses Bounded Partial Order Reduction (BPOR) \cite{bpor} for a similar sound reduction,
+although we leave incorporating ICB with Iterative Deepening to future work (\sect{\ref{sec:future}})}.
 It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
@@
 The {\tt without\_\allowbreak{}function} directive is the dual of {\tt within\_function}, indicating a blacklist.
-We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}).
+We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}),
+\revision{similarly to Preemption Sealing from prior work \cite{sealing}}.
 %Multiple invocations can be used; later ones take precedence.

### 2016-03-22 oopsla-16/evaluation.tex
Write loads more for eval. Getting there.

@@
 		\begin{enumerate}
-			\item Does Iterative Deepening find bugs faster
+			\revision{\item} Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
+				% Elaborate later:
+				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
+			\revision{\item} Does Iterative Deepening find bugs faster
 				%than SSS-MC
@@
 			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
-			\item Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
-				% Elaborate later:
-				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
 			\revision{
@@
 and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's
-% TODO CAMREADY: CMSC 23000 to be precise
-CS230 operating systems classes.
+\revision{CMSC 23000}~operating systems classes.
 %
@@
 	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with plain DPOR \cite{dpor},
-	\item {\bf SSS-MC-ICB}: With PPs as above, but using ICB \cite{chess-icb} with BPOR \cite{bpor} instead of DPOR to find bugs faster, and
+	\item {\bf SSS-MC-ICB}: With PPs as above, but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster, and
 	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
@@
 All tests ran on 12-core 3.2 GHz Xeon W3670 machines. %with 12GB of RAM.
+
+% TODO: Where should this paragraph be?
+{\bf Variance.}
+Because \quicksand~can be nondeterministic in which state spaces get scheduled for testing more often,
+we quantified the variance in the distribution of QS-DR's bugs with a permutation test.
+We re-ran
+% TODO: X=? 100?
+%X\% of the tests in
+the QS-DR experiment and compared the two observed distributions,
+finding the observed area between the curves to be among the middle
+% TODO
+Y\%
+% FIXME: How many randomly sampled permutations?
+of possible areas between permuted distributions,
+and furthermore finding the observed area to be
+% TODO
+Z\%
+of the area between QS-DR and SSS-MC-ICB (the next closest).
 }
 
-\subsection{Comparing Iterative Deepening to SSS-MC}
+\subsection{\revision{Comparison to state-of-the-art MC}}%Comparing Iterative Deepening to SSS-MC
 \label{sec:eval-sssmc}
 
-% mention exactly which state spaces we are comparing here
-% mention partial verification in terms of state space completion, when SSSMC times out.
-
-%In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.
-
-% TODO: Cut/rearrange.
-%\quicksand's times by 10 to convert from wall-clock time to CPU-hours (even though it sometimes falls short of 100\% parallelism).
-Figure~\ref{fig:dowefindbugsfaster} plots a cumulative distribution of bugs found before a given elapsed CPU consumption. % during its associated test.
-We tested several configurations of \quicksand~to compare against SSS-MC from multiple angles,
-indicated by each of the ``QS'' lines,
-which we explain presently.
+Figure~\ref{fig:dowefindbugsfaster} plots the cumulative distribution of bugs found \revision{by each experiment}.
+\revision{
+Figure~\ref{fig:dowefindbugsfaster}(a) is the main comparison by CPU-time;
+we additionally show a wall-clock comparison in (b) to highlight the impact of \quicksand's parallelism.
+}
 
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf}
+	\hspace{-1em}
+		\revision{
+	\begin{tabular}{p{0.48\textwidth}}
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
+		(a) Bugs found by elapsed CPU time. A more direct comparison than (b),
+		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
+		\\
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-wallclock.pdf} \\
+		(b) Bugs found by elapsed wall-clock time.
+		\quicksand~is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
+	\end{tabular}
+	}
 	\caption{Comparison of bug-finding performance
 	by several configurations of \quicksand~and the SSS-MC control.
-	\quicksand~finds 179\% as many bugs with data-race PPs.}
+	\quicksand~finds \revision{151}\% as many bugs with data-race PPs.}
 	\label{fig:dowefindbugsfaster}
@@
 
+{\bf Finding new data-race bugs.}
+%We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
+%The break-even point is at a negligible 90 seconds.
+\revision{Compared to the best SSS-MC series, \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds,
+and ultimately concludes 10 CPU-hours with \revision{151}\% as many bugs in total.
+\revision{Before the break-even point at 200 seconds,
+% TODO
+% (for SSS-MC-ICB) or XXX seconds (for SSS-MC-Shared-Mem)
+\quicksand~lags behind the SSS-MC approaches due to additional start-up overhead from its tenfold parallelism.
+However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
+In Figure~\ref{fig:dowefindbugsfaster}(b), we give \quicksand~full credit for its inherent parallelism, and it outperforms SSS-MC for any fixed budget of wall-clock time.}
+
+% TODO: Fix up this paragraph after preempt-everywhere results come in.
+The left half of Table~\ref{tab:drbugs}
+breaks down the number and types of bugs found by each test program.
+In \mxtest, in which we do not trust the lock implementation's correctness,
+the control experiment found dramatically fewer bugs (just 1)\footnote{
+	The one bug SSS-MC found was in a fully-assembler lock implementation.
+{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
+Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
+%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
+%Though for many applications of MC, assuming a correct lock implementation is sufficient,
+Though it often suffices to assume correctly-implemented locks,
+we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
+%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
+% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
+% (as well as on the mutex API boundaries).
+% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
+
+%% RIP this.
+%Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
+%which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
+%Intuitively, this line shows that for programs with only benign data races,
+%\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
+
 % TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
 % TODO: Nix discussion of QS-no-DR-bugs. Just do QS-no-DR-PPs.
-
-% TODO: Nix this discussion.
 {\bf Finding the same bugs faster.}
-To show that Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
-we configured \quicksand~to explore only subsets of the synchronization PPs
-(i.e., ignoring all data-race candidates).
+%To test whether Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
+\revision{The QS-Sync-Only experiment tests whether}~Iterative Deepening is effective even for MC domains without data races.
+\revision{When \quicksand~ignores all data-race candidates,
+its results are competitive with SSS-MC-DPOR, but SSS-MC-ICB outperforms it.
+This is unsurprising: the seed subsets of PPs (\sect{\ref{sec:initial-pp}}) QS-Sync-Only is limited to are much less flexible than ICB.
+This result suggests that in future work, \quicksand~should consider using ICB in parallel with its initial PP configuration when it finds no data-race candidates to test.}
 %\footnote{
@@
 %Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
-%
-The line QS-no-DR-PPs represents this experiment.
-Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
-and is faster in the first 60 seconds due to less start-up overhead,
-QS-no-DR-PPs finds more of the bugs sooner thereafter.
-Hence, for modest CPU budgets,
-\quicksand~is likely to find bugs SSS-MC will miss.
+
+%Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
+%and is faster in the first 60 seconds due to less start-up overhead,
+%QS-no-DR-PPs finds more of the bugs sooner thereafter.
+%Hence, for modest CPU budgets,
+%\quicksand~is likely to find bugs SSS-MC will miss.
 %and for more ambitiously-sized tests,
@@
 %These results show explicitly that for arbitrary CPU-time budgets
-
-%Note that even users willing to spend 10 CPU-hours should find this ``early advantage'' compelling:
-
-{\bf Finding new data-race bugs.}
-The line QS-DRs represents \quicksand's bug-finding performance
-with data-race PPs enabled.
-%We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
-%The break-even point is at a negligible 90 seconds.
-We find significantly more bugs within any CPU budget greater than 90 seconds,
-and ultimately conclude 10 CPU-hours with 151\% as many bugs in total.
-
-Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
-which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
-Intuitively, this line shows that for programs with only benign data races,
-\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
 %\footnote{The initial perfect overlap between QS-DRs and QS-no-DR-bugs indicates how long it takes before the first data-race bug is found.}
@@
 
-% TODO: Incorporate the following point.
-% "Before the break-even point at 45 seconds (or whatever), SSS-MC approaches seem to outperform \quicksand,
-% but converting this CPU-time advantage into faster wall-clock performance is itself an open research problem \cite{parallel-dpor}.
+\revision{
+On the other hand,
+comparing QS-DR to SSS-MC-Shared-Mem shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
+%We attribute this to the fact that
+Statically configuring a PP for every shared memory access in advance
+produces orders of magnitude more PPs than
+waiting for an access to be identified as part of a (potential) data race at runtime.
+%
+In principle, DPOR and BPOR should identify and prune any equivalences arising from extraneous PPs on non-conflicting accesses.
+However, in practice,
+%we found that
+the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking\footnote{
+	\revision{The bug-finding performance of SSS-MC-ICB approach could be improved by heuristically
+	%preempting only on data-race candidates found in advance by a single-pass analysis
+	using a single-pass analysis to find data-race PP candidates in advance
+	\cite{portend}, but this sacrifices soundness, as shown in \sect{\ref{sec:eval-falseneg}}.}
+}.
+Iterative Deepening avoids this overhead by waiting until runtime to identify fewer, more relevant PPs dynamically,
+and is hence more suitable for MC with data-race PPs.}
 
-% TODO
-% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
-Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
-
+% TODO: fix this table.
 \begin{table*}[t]
@@
 
-%Figure~\ref{fig:bargraph}
-The left half of Table~\ref{tab:drbugs}
-breaks down the number and types of bugs found by each test program.
-% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
-%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
-% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
-% (as well as on the mutex API boundaries).
-% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
-%In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
-%even compared to the other test programs
-In \mxtest, in which we do not trust the lock implementation's correctness,
-the control experiment found dramatically fewer bugs (just 1)\footnote{
-	The one bug SSS-MC found was in a fully-assembler lock implementation.
-{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
-Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
-%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
-%Though for many applications of MC, assuming a correct lock implementation is sufficient,
-Though it often suffices to assume correctly-implemented locks,
-we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
-
 {\bf Partial verification.}
@@
 %In Table~\ref{tab:allbugs} ....
+
+{\bf Full verification.}
+For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
+In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU consumption.
+Among these, 36 tests contained no data-race candidates whatsoever,
+so the same verification could be provided by SSS-MC.
+We plot these for comparison,
+according to the time taken by that test's SSS-MC control experiment.
+When a test can be totally verified without any data-race PPs,
+\quicksand~may be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
+Of course, using data-race PPs increases our capacity for total verification by 4.25x.
+
+\begin{figure}[t]
+	\includegraphics[width=0.48\textwidth]{totalverifs-squashed.pdf}
+	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
+	Some tests had no data-race candidates, and hence could also be verified by SSS-MC.}
+	\label{fig:totalverif}
+\end{figure}
+% TODO
+% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
+Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
 
@@
 %which we consider a much stronger verification.
-
-{\bf Full verification.}
-For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
-In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU consumption.
-Among these, 36 tests contained no data-race candidates whatsoever,
-so the same verification could be provided by SSS-MC.
-We plot these for comparison,
-according to the time taken by that test's SSS-MC control experiment.
-When a test can be totally verified without any data-race PPs,
-\quicksand~may be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
-Of course, using data-race PPs increases our capacity for total verification by 4.25x.
-
-\begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{totalverifs-squashed.pdf}
-	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
-	Some tests had no data-race candidates, and hence could also be verified by SSS-MC.}
-	\label{fig:totalverif}
-\end{figure}
 

### 2016-03-22 oopsla-16/overview.tex
Write loads more for eval. Getting there.

@@
 	7 & & \texttt{mutex\_unlock(\&mx);} \\
-	8 & & \texttt{if (tmp)} \\
-	9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
+	8 & & \texttt{if (tmp) \hilight{brickred}{x++;}~// B2} \\
+	%8 & & \texttt{if (tmp)} \\
+	%9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
 \end{tabular}

### 2016-03-22 oopsla-16/evaluation.tex
Write about full verif (and update grafs, and move the table)

@@
 
-% TODO: Where should this paragraph be?
-{\bf Variance.}
-Because \quicksand~can be nondeterministic in which state spaces get scheduled for testing more often,
-we quantified the variance in the distribution of QS-DR's bugs with a permutation test.
-We re-ran
-% TODO: X=? 100?
-%X\% of the tests in
-the QS-DR experiment and compared the two observed distributions,
-finding the observed area between the curves to be among the middle
-% TODO
-Y\%
-% FIXME: How many randomly sampled permutations?
-of possible areas between permuted distributions,
-and furthermore finding the observed area to be
-% TODO
-Z\%
-of the area between QS-DR and SSS-MC-ICB (the next closest).
-}
-
 \subsection{\revision{Comparison to state-of-the-art MC}}%Comparing Iterative Deepening to SSS-MC
@@
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
-		(a) Bugs found by elapsed CPU time. A more direct comparison than (b),
+		(a) Bugs found by elapsed CPU time. Overall, a more direct comparison than (b),
 		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
@@
 
+%{\bf Variance.}
+% TODO CAMREADY: Probably will need at least a 3rd run of dis.
+Because \quicksand~can be nondeterministic in its scheduling of state spaces,
+we ran the QS-DR tests a second time to quantify the variance.
+We found the area between the two QS-DR curves to be
+% TODO
+{\bf XX\%}
+of the area between QS-DR and SSS-MC-ICB (the next closest),
+% TODO cite
+which we consider an insignificant amount of variance.
+}
+
 % TODO: Fix up this paragraph after preempt-everywhere results come in.
@@
 
+% TODO: fix this table.
+\begin{table*}[t]
+	\begin{center}
+		\small
+	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+		% TODO: Update SSSMC column to have more favourable ICB numbers.
+		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
+		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
+		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf recycle DRs} &
+		{\bf time-outs} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
+		\hline
+		% FIXME: update FRMs
+				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
+		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
+		{\tt join} 	& 79	& 14 	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
+		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
+		{\tt sem} 	& 79	& 10	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
+		{\tt signal} 	& 79	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
+		{\tt rwlock} 	& 79	& 26	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
+		\hline
+		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
+		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
+		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
+		\hline
+		{\bf Total}	& 629	& 91	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
+	\end{tabular}
+	\end{center}
+	\caption{Summary of bugs and data races found by each test program.
+		SSS-MC is the control; QS is \quicksand.
+		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
+	``Malloc-recycle DRs'' counts how many false positives we suppressed (\sect{\ref{sec:eval-dr}}).
+	``Mutual time-outs'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
+	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
+	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
+	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
+		}
+	\label{tab:drbugs}
+\end{table*}
+
 % TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
@@
 its results are competitive with SSS-MC-DPOR, but SSS-MC-ICB outperforms it.
-This is unsurprising: the seed subsets of PPs (\sect{\ref{sec:initial-pp}}) QS-Sync-Only is limited to are much less flexible than ICB.
-This result suggests that in future work, \quicksand~should consider using ICB in parallel with its initial PP configuration when it finds no data-race candidates to test.}
+This is unsurprising: the seed subsets of PPs (\sect{\ref{sec:initial-pp}}) QS-Sync-Only is limited to are much less flexible than ICB's preemption strategy.
+This result suggests that in future work, \quicksand~should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.}
 %\footnote{
@@
 	using a single-pass analysis to find data-race PP candidates in advance
-	\cite{portend}, but this sacrifices soundness, as shown in \sect{\ref{sec:eval-falseneg}}.}
+	\cite{portend}, but this sacrifices soundness, as shown in \sect{\ref{sec:eval-dr}}.}
 }.
@@
 and is hence more suitable for MC with data-race PPs.}
-
-% TODO: fix this table.
-\begin{table*}[t]
-	\begin{center}
-		\small
-	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
-		% TODO: Update SSSMC column to have more favourable ICB numbers.
-		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
-		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
-		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf recycle DRs} &
-		{\bf time-outs} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
-		\hline
-		% FIXME: update FRMs
-				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
-		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
-		{\tt join} 	& 79	& 14 	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
-		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
-		{\tt sem} 	& 79	& 10	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
-		{\tt signal} 	& 79	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
-		{\tt rwlock} 	& 79	& 26	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
-		\hline
-		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
-		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
-		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
-		\hline
-		{\bf Total}	& 629	& 91	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
-	\end{tabular}
-	\end{center}
-	\caption{Summary of bugs and data races found by each test program.
-		SSS-MC is the control; QS is \quicksand.
-		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
-	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-falseneg}}).
-	``Malloc-recycle DRs'' counts how many false positives we suppressed (\sect{\ref{sec:eval-falsepos}}).
-	``Mutual time-outs'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
-	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
-	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
-	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
-		}
-	\label{tab:drbugs}
-\end{table*}
 
@@
 {\bf Full verification.}
+% TODO: Discuss how ICB sucks even more.
 For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
-In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications before a given elapsed CPU consumption.
-Among these, 36 tests contained no data-race candidates whatsoever,
-so the same verification could be provided by SSS-MC.
-We plot these for comparison,
-according to the time taken by that test's SSS-MC control experiment.
-When a test can be totally verified without any data-race PPs,
-\quicksand~may be slower than SSS-MC due to redundant work (\sect{\ref{sec:future}}).
-Of course, using data-race PPs increases our capacity for total verification by 4.25x.
+In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications
+%before a given elapsed CPU consumption.
+\revision{against those provided by SSS-MC-Shared-Mem\footnote{
+	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
+	For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
+	SSS-MC-Shared-Mem's verifications are only sound in the absence of such sharing!
+	\quicksand~detects these dynamically, without needing to discriminate between stack and heap.}
+	%Expanding the statically-coded PPs to include stack accesses would make SSS-MC-Shared-Mem's performance even worse.
+	% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
+Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
+}.
+\quicksand~outperforms SSS-MC, being able to verify safety for
+% TODO
+{\bf 99x}
+as many tests in 10 CPU-hours.
+We attribute this again to the overhead of the sheer number of statically-coded PPs that SSS-MC-Shared-Mem must use,
+as well as to the work which ICB necessarily repeats when increasing its preemption bound.}
+
+Among these verified tests, 36 contained no data-race candidates whatsoever,
+so the same verification could be provided
+%by SSS-MC
+\revision{with synchronization PPs only.
+We plot the verifications by QS-Sync-Only, SSS-MC-DPOR and SSS-MC-ICB as well:
+among these, the otherwise more antiquated SSS-MC-DPOR performs best,
+while the other two lag behind
+}
+%\quicksand~is slower than SSS-MC-DPOR
+due to redundant work (\sect{\ref{sec:future}}).
+\revision{While QS-Sync-Only and SSS-MC-ICB are competitive with each other},
+using data-race PPs increases our \revision{verification capacity}~by 4.25x.
 
@@
 \end{figure}
-% TODO
-% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
-Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
 
@@
 
-\subsection{Avoiding false positive data-race candidates}
-\label{sec:eval-falsepos}
+\subsection{Comparing to single-pass data-race analysis}
+\label{sec:eval-dr}
 % Though we mechanically verify whether each data race candidate leads to a bug, each new PP can increase combinatorially..... obviously wish to avoid...
@@
 
-\subsection{Finding nondeterministic data-race candidates}
-\label{sec:eval-falseneg}
 %For the [NONDET] expt, you don't need to additionally count how many DRs
@@
 %are also dr-pp-ss-only (the reviewers may ask for this before cam-ready?).
-
+{\bf \revision{Finding nondeterministic data-race candidates}.}
 Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.

### 2016-03-22 oopsla-16/related.tex
Write about full verif (and update grafs, and move the table)

@@
 However, Portend does not test alternate interleavings {\em in advance} of knowing any specific data races,
-which is necessary to find certain bugs (\sect{\ref{sec:eval-falseneg}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
+which is necessary to find certain bugs (\sect{\ref{sec:eval-dr}}) or to provide full verification (\sect{\ref{sec:totalverif}}).
 % Not really true.

### 2016-03-22 oopsla-16/evaluation.tex
clean up todos a bit

@@
 		% TODO: Update SSSMC column to have more favourable ICB numbers.
-		& {\bf Num.} & {\bf SSS-MC} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
+		& {\bf Num.} & {\bf DPOR} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
 		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
@@
 	\caption{Summary of bugs and data races found by each test program.
-		SSS-MC is the control; QS is \quicksand.
+		% TODO
+		DPOR is the control; QS is \quicksand.
 		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
@@
 % TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
-% TODO: Nix discussion of QS-no-DR-bugs. Just do QS-no-DR-PPs.
 {\bf Finding the same bugs faster.}

### 2016-03-22 oopsla-16/overview.tex
clean up todos a bit

@@
 
-% TODO: if you need, you can recover a line of text by abbreviating "HB".
 \revision{
@@
 it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
-as shown in the example program in Figure~\ref{fig:hb-example}(a). % TODO
+as shown in the example program in Figure~\ref{fig:hb-example}(a).
 We call such unreported access pairs {\em false negatives}.
@@
 by considering only blocking operations like {\tt cond\_wait} to enforce the order.
-However, consider the similar program in Figure~\ref{fig:hb-example}(b), % TODO
+However, consider the similar program in Figure~\ref{fig:hb-example}(b),
 in which the access pair ceases to exist in the alternate interleaving.

### 2016-03-22 oopsla-16/evaluation.tex
Move discussion of static DR as seed sets to warp zone.

@@
 All tests ran on 12-core 3.2 GHz Xeon W3670 machines. %with 12GB of RAM.
+}
 
@@
 
+\revision{
 %{\bf Variance.}
@@
 		\small
-	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+	\begin{tabular}{r|c||c|c|c|c|c|c||c|c||c|c}
 		% TODO: Update SSSMC column to have more favourable ICB numbers.
-		& {\bf Num.} & {\bf DPOR} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
-		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
-		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf recycle DRs} &
-		{\bf time-outs} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
+		& {\bf Num.} & {\bf DPOR} & {\bf ICB} & {\bf QS} & {\bf DR} & {\bf Nondet.} & {\bf Malloc} & {\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
+		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf DRs} & {\bf timeouts} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
 		\hline
-		% FIXME: update FRMs
-				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
-		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
-		{\tt join} 	& 79	& 14 	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
-		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
-		{\tt sem} 	& 79	& 10	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
-		{\tt signal} 	& 79	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
-		{\tt rwlock} 	& 79	& 26	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
+		% FIXME CAMREADY: update FRMs
+					% TODO: update to be ShMem not DPOR
+											% TODO: recount with ICB
+				% #test  sssmc   icb      qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
+		{\tt bcast}	& 79	& 7	& TODO	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
+		{\tt join} 	& 79	& 14 	&	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
+		{\tt mx} 	& 79	& 1	&	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
+		{\tt sem} 	& 79	& 10	&	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
+		{\tt signal} 	& 79	& 4	&	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
+		{\tt rwlock} 	& 79	& 26	&	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
 		\hline
-		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
-		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
-		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
+		{\tt sched} 	& 59	& 1	&	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
+		{\tt alarm} 	& 44	& 5	&	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
+		{\tt wait} 	& 52	& 23	&	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
 		\hline
-		{\bf Total}	& 629	& 91	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
+		{\bf Total}	& 629	& 91	&	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
 	\end{tabular}
@@
 		DPOR is the control; QS is \quicksand.
-		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+		``DR bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
 	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
-	``Malloc-recycle DRs'' counts how many false positives we suppressed (\sect{\ref{sec:eval-dr}}).
-	``Mutual time-outs'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
+	``Malloc DRs'' counts how many malloc-recycle false positives we suppressed (\sect{\ref{sec:eval-dr}}).
+	``Mutual timeouts'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
 	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
@@
 %we found that
-the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking\footnote{
-	\revision{The bug-finding performance of SSS-MC-ICB approach could be improved by heuristically
-	%preempting only on data-race candidates found in advance by a single-pass analysis
-	using a single-pass analysis to find data-race PP candidates in advance
-	\cite{portend}, but this sacrifices soundness, as shown in \sect{\ref{sec:eval-dr}}.}
-}.
+the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking.
 Iterative Deepening avoids this overhead by waiting until runtime to identify fewer, more relevant PPs dynamically,

### 2016-03-22 oopsla-16/related.tex
Move discussion of static DR as seed sets to warp zone.

@@
 
+% TODO: Expand on MCR and SATcheck.
 {\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
@@
 However, ICB is limited to a fixed set of PPs,
-\revision{cannot estimate remaining time to total verification},
-and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+\revision{cannot estimate remaining time to total verification,
+and cannot incorporate data-race PPs not statically coded in advance}.
+%and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
 

### 2016-03-22 oopsla-16/warpzone.tex
Move discussion of static DR as seed sets to warp zone.

--- 2015-11-19 pldi-16/warpzone.tex
@@
 
+\revision{
+{\bf Integration with static data-race analysis.}
+In \sect{\ref{sec:eval-sssmc}}, we evaluated SSS-MC's ability to find data-race-induced failures
+by configuring a static predicate to preempt on any non-stack memory access.
+This introduced up to thousands of new PPs on each new test execution, with a prohibitive performance impact.
+While this performance could be improved by
+%relaxing the preemption strategy,
+instead using a static or single-pass analysis to find data-race candidate PPs in advance \cite{portend},
+this strategy sacrifices the soundness of the verification guarantee, as shown in \sect{\ref{sec:eval-dr}}.
+However, \quicksand~itself could employ static data-race analysis \cite{racerx} in future work.
+Statically-identified data race candidates could heuristically be included in our ``seed'' PP sets (\sect{\ref{sec:initial-pp}}),
+enabling \quicksand~to focus on the most suspicious races immediately, rather than waiting for them to be identified after potentially many iterations of MC.
+}
+
 %{\bf Small test cases.}

### 2016-03-22 oopsla-16/area-between-curves.c
area between curves program

@@ -1,0 +1,214 @@
+#include <assert.h>
+#include <stdio.h>
+#include <math.h>
+
+#define END (36000.0)
+// QS DRs
+long double series1[] = {
+36.510254, 37.739562, 38.365079, 38.637311, 39.170384, 39.501006, 39.662938,
+39.833846, 40.255752, 40.328943, 40.933243, 41.157442, 41.211145, 41.228979,
+41.351358, 41.796638, 42.160382, 42.214631, 42.510143, 43.070664, 43.583659,
+44.369137, 44.512721, 44.728749, 44.854787, 45.177707, 46.237215, 46.265118,
+46.58655, 47.360211, 48.021923, 48.124353, 51.690667, 51.710857, 52.561983,
+54.016317, 54.171573, 56.185903, 57.138399, 60.047792, 61.982603, 64.056988,
+65.164125, 65.258305, 65.601548, 68.292868, 68.769434, 69.493605, 69.775006,
+70.402518, 72.45192, 75.730529, 79.877823, 80.11299, 80.529691, 82.429194,
+84.657224, 86.851432, 90.712605, 93.975184, 96.353075, 99.706684, 101.141091,
+101.62696, 103.092741, 103.523732, 104.287254, 104.334906, 104.576109,
+106.496646, 106.648468, 106.702727, 107.72336, 115.586888, 120.696353,
+122.54649, 127.810423, 136.250888, 141.992309, 152.042405, 158.834142,
+163.826769, 168.454978, 174.54547, 177.181624, 181.328461, 181.653364,
+185.581022, 187.422864, 200.170433, 203.316413, 204.883637, 210.411072,
+212.247479, 246.775512, 322.956047, 331.17364, 369.822964, 375.260412,
+380.221346, 445.226695, 453.693388, 457.974405, 463.664797, 593.610861,
+617.736286, 772.679389, 816.080472, 833.232277, 887.320373, 1257.8181,
+1266.503171, 1524.251812, 1532.174158, 1551.012061, 1590.47937, 1715.713368,
+1781.956105, 1922.114035, 1939.692742, 2012.371056, 2054.098519, 2261.579858,
+2865.213544, 3069.225804, 3321.310137, 3891.428744, 4084.332501, 4402.315102,
+5106.131582, 5446.954874, 5568.344117, 5671.15995, 6164.389778, 6662.561184,
+6716.330609, 7555.430816, 7787.214391, 8030.477676, 8181.063836, 8697.435133,
[…]

### 2016-03-23 oopsla-16/area-between-curves.c
test case for area

@@ -2,6 +2,7 @@
 #include <stdio.h>
 #include <math.h>
 
+#if 1
 #define END (36000.0)
 // QS DRs
 long double series1[] = {
@@ -52,6 +53,7 @@ long double series2[] = {
 795.725917, 843.174624, 970.285112, 1387.089948, 1410.237694, 1822.463231,
 2582.530643, 20959.641808, END
 };
+#endif
 
 
 #if 0
@@ -93,6 +95,25 @@ long double series2[] = {
 	1.0,
 	2.0,
 	3.0,
+	END
+};
+#endif
+#if 0
+// Test case -- expect total logarithmic area to be 5
+#define END M_E * M_E * M_E * M_E * M_E
+long double series1[] = {
+	1,
+	M_E,
+	M_E * M_E,
+	M_E * M_E * M_E,
+	M_E * M_E * M_E * M_E,
+	END
+};
+long double series2[] = {
+	M_E,
+	M_E * M_E,
+	M_E * M_E * M_E,
+	M_E * M_E * M_E * M_E,
 	END
 };
 #endif

### 2016-03-23 oopsla-16/overview.tex
More defensive prose in anticipation of data race definition wankery.

@@
 In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
+
+\revision{
+A data race analysis may be either {\em static} (inspecting source code) \cite{racerx} or {\em dynamic} (tracking individual accesses arising at run-time) \cite{tsan}.
+This paper focuses exclusively on dynamic analysis,
+so although our example refers to numbered source lines for ease of explanation,
+in practice we are actually classifying the individual memory access events corresponding to those lines during execution.
+}
+
+%\revision{Among these, only the races involving lines 3 and 5 are relevant to the possible assertion failure;
+%permuting the order of line 8 alone
+%% relative to 2 and 6
+%will not change the program's behaviour.}
 

### 2016-03-23 oopsla-16/soundness.tex
More defensive prose in anticipation of data race definition wankery.

@@
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
-If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
-We call this a {\em malloc-recycle data race}.
+If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a \revision{potential}~data race.
+We call this a {\em malloc-recycle data race \revision{candidiate}}.
 To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
@@
 %However,
-When limited to a single test execution, suppressing any data race matching this pattern is unsound.
+When limited to a single test execution, suppressing any data race \revision{candidate}~matching this pattern is unsound.
 Consider the more unusual program in Figure~\ref{fig:recycle-bug},
@@
 \end{tabular}
-\caption{If a single-pass data race detector discarded candidates matching the malloc-recycle pattern,
+	\caption{If a single-pass \revision{Limited HB analysis}~discarded candidates matching the malloc-recycle pattern,
 it would miss the bug in this adversarial program.}
@@
 Fortunately, when data-race detection is combined with DPOR and Iterative Deepening, pruning all malloc-recycle candidates is sound, even considering adversarial programs such as Figure~\ref{fig:recycle-bug}.
-This makes it unnecessary to verify such data races by actively adding more preemptions,
+This makes it unnecessary to verify such \revision{candidates}~by actively adding more preemptions,
 achieving a potentially combinatorial reduction in how many state spaces we generate.
@@
 
-\begin{theorem}[Soundness of eliminating malloc-recycle races]
-If a malloc-recycle data race is not a false positive,
+\begin{theorem}[Soundness of eliminating malloc-recycle candidates]
+	If a malloc-recycle \revision{candidate}~is not a false positive,
 %DPOR will reorder threads such that
@@
 %either
-the accesses still race without fitting the malloc-recycle pattern.
+the accesses can race without fitting the malloc-recycle pattern.
 %, or a use-after-free bug will be reported immediately.
@@
 hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
-Because this is a true data race, $p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.
+Because this is a true \revision{potential}~data race, $p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.
 
@@
 If locking or message-passing is used, our set of hard-coded PPs suffices to provide a PP.
-Otherwise, $p$ (and the corresponding read by T2) will be a data race, although that may itself be a malloc-recycle race.
+	Otherwise, $p$ (and the corresponding read by T2) will be a \revision{potential}~data race, although that may itself be a malloc-recycle \revision{candidate}.
 In this case we use induction on the pointer chain leading to the shared address containing $p$:

### 2016-03-23 oopsla-16/intro.tex
Write about experimental setup.
Wow factor of verif

@@
 
+% TODO: Incorporate the following point
+% On the other hand, techniques such as ICB assume all state spaces will be too big, and optimize the search to uncover bugs more quickly, but at the expense of / by sacrificing the ability to complete a total verification quickly when possible.
+
 We present \quicksand,
@@
 When prior work would time out on large tests by trying several preemption points simultaneously,
-\quicksand~tests smaller state spaces based on subsets of those points, which often finds the same bugs sooner.
+\quicksand~\revision{identifies these pitfalls in advance using state space estimation \cite{estimation},
+and instead}~tests smaller state spaces based on subsets of those points.
+\revision{Often, testing these}~smaller state spaces can even find the same bugs sooner.
+
 On the other hand, when the CPU budget is large enough to fully test all data-race preemption points,
 we prove that this constitutes a total verification of all possible thread schedules.
-Short of deciding in advance to preempt on every single instruction \cite{spin}, this was not previously possible without data-race preemption points.
+%Short of deciding in advance to preempt on every single instruction \cite{spin}, this was not previously possible without data-race preemption points.
+\revision{To achieve the same level of verification, prior model checkers must decide}~in advance to preempt on every single memory access \cite{spin}, \revision{which is computationally prohibitive for
+%even moderately-sized
+larger tests.
+Our approach provides the best of both worlds:
+by estimating the size of the test on-the-fly, \quicksand~can find bugs quickly in large tests {\em and} provide fast total verification for small ones.}
 

### 2016-03-23 oopsla-16/design.tex
Write about experimental setup.
Clean up old commented prose, and expand prose in some places, especially malloc-recycle.

--- 2015-11-20 pldi-16/design.tex
@@
 Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs.
-%which we refer to as {\em jobs}.
 We refer to each unique set of PPs as a {\em job},
@@
 \subsection{Initial PP configuration}
+\label{sec:initial-pp}
 
@@
 	}
-	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{\alpha\}$}{
+	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
 		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
@@
 If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
-%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
 The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
@@
 Furthermore, we do not always strictly increase the number of PPs when we find a new data race.
-%When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
 For each instruction involved in a data race, \quicksand~adds two new jobs:
@@
 %
-The rationale of spawning multiple jobs is that we don't know in advance which will be most fruitful: %which will be most fruitful is not known in advance:
+The rationale of spawning multiple jobs is that we don't know in advance which will be most fruitful:
 while the big job risks not completing in time,
@@
 In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
-%which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
 This phenomenon motivates Iterative Deepening to prioritize the jobs at run-time.
@@
 in which case we may iteratively advance to a superset state space containing PPs for both racing access pairs.
-%Because Iterative Deepening is
 Being constrained by a CPU budget,
@@
 based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
-Single-order candidates are more likely to be false positives,
+Single-order candidates are more likely to be false positives (\sect{\ref{sec:overview-dr}}),
 although preempting during the access itself is necessary to say for sure.
@@
 (indicated by the HeuristicPriority($\alpha$) call in Algorithm~\ref{alg:handledatarace}).
-%Jobs with both-order data-race PPs are prioritized higher,
-%because single-order candidates are more likely to be false-positives.
-%(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
-%so we prioritize jobs with both-order data-race PPs.
 For single-order races, we do not initially add a PP for the later access at all:

### 2016-03-23 oopsla-16/implementation.tex
Clean up old commented prose, and expand prose in some places, especially malloc-recycle.

@@
 \landslide~implements DPOR \cite{dpor},
-%\cite{dpor},
 state space estimation \cite{estimation}, and a hybrid
-%lockset/happens-before
 data-race analysis \cite{hybriddatarace}.
@@
 %serving as the dual of {\tt within\_function}.
-The {\tt without\_\allowbreak{}function} directive is the dual of {\tt within\_function}, indicating a blacklist.
+The {\tt without\_\allowbreak{}function} directive \revision{serves as}~the dual of {\tt within\_function}, indicating a blacklist.
 We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}),

### 2016-03-23 oopsla-16/intro.tex
Clean up old commented prose, and expand prose in some places, especially malloc-recycle.

@@
 
-% Can I even make a claim this broad to begin with?
 However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
@@
 any choice made in advance must necessarily leave some tests unaffordably large \cite{parrot,mcr}.
-%
 %This work shows how to avoid making that tradeoff decision in advance.
-
-% TODO: Incorporate the following point
-% On the other hand, techniques such as ICB assume all state spaces will be too big, and optimize the search to uncover bugs more quickly, but at the expense of / by sacrificing the ability to complete a total verification quickly when possible.
 
@@
 It uses data-race analysis \cite{eraser} to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
-When prior work would time out on large tests by trying several preemption points simultaneously,
-\quicksand~\revision{identifies these pitfalls in advance using state space estimation \cite{estimation},
+When prior \revision{approaches}~would time out on large tests by trying several preemption points simultaneously,
+\quicksand~\revision{identifies this pitfall in advance using state space estimation \cite{estimation},
 and instead}~tests smaller state spaces based on subsets of those points.
 \revision{Often, testing these}~smaller state spaces can even find the same bugs sooner.
+
+%\revision{Hence, recent approaches favor
+%heuristic
+%%search ordering
+%strategies \cite{chess-icb,randomized-scheduler}
+%to prioritize fast bug-finding over ultimate completion time,
+%%assuming state spaces will always be too big and
+%%at the expense of ultimate completion time,
+%which in turn sacrifices the potential for a full safety guarantee when no bugs exist.}
+%%search ordering strategies such as Iterative Context Bounding \cite{chess-icb}
+%%assume all state spaces will be too big,
+%%and optimize the search to uncover bugs more quickly at the expense of ultimate completion time.}
+%%/ by sacrificing the ability to complete a total verification quickly when possible.
 
@@
 		but which we prove correct when used with Iterative Deepening;
-		%, unsound in single-pass analysis but which we prove sound when used with Iterative Deepening;
 	\item A large evaluation in which \quicksand~compares favourably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
@@
 
-% TODO CAMREADY: uncomment if a reviewer complains about the absence of this
-%The rest of the paper is organized as follows.
-%\sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,
-%\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation, %of Iterative Deepening,
-%\sect{\ref{sec:soundness}} presents our proofs of convergence and of the soundness of our new false-positive data-race tactic,
-%\sect{\ref{sec:eval}} presents our evaluation,
-%\sect{\ref{sec:future}} discusses limitations and future work,
-%\sect{\ref{sec:related}} surveys the related work,
-%and \sect{\ref{sec:conclusion}} concludes.
+% Comment this out for space?
+The rest of the paper is organized as follows.
+\sect{\ref{sec:overview}} reviews the background material, %on stateless model checking and data-race analysis,
+\sect{\ref{sec:design}} and \sect{\ref{sec:implementation}} discuss our design and implementation, %of Iterative Deepening,
+\sect{\ref{sec:soundness}} provides our proofs of soundness, %convergence and of the soundness of our new false-positive data-race tactic,
+\sect{\ref{sec:eval}} presents our evaluation,
+\sect{\ref{sec:future}} discusses limitations and future work,
+\sect{\ref{sec:related}} surveys the related work,
+and \sect{\ref{sec:conclusion}} concludes.

### 2016-03-23 oopsla-16/overview.tex
Clean up old commented prose, and expand prose in some places, especially malloc-recycle.

@@
 For the rest of the paper, we will abbreviate {\em preemption point} (PP),
-%For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
+\revision{{\em happens-before} (HB)},
 {\em model checking} (MC),
@@
 the set of PPs is fixed in advance, and the tool commits to testing every interleaving available with those PPs.
-%Reduction techniques are often used to skip equivalent interleavings \cite{dpor}, and search-ordering strategies
 Many techniques can skip equivalent interleavings or order the search to uncover bugs faster \cite{dpor,demeter,chess-icb,gambit,smc-empirical-study},
@@
 we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-% old explanation
-%Because data-race analysis is prone to false-positives,
-%we classify unprotected memory access pairs separately from bugs,
 \revision{We refer to racing (or potentially-racing) memory access pairs as}~{\em data race candidates}.

### 2016-03-23 oopsla-16/soundness.tex
Clean up old commented prose, and expand prose in some places, especially malloc-recycle.

@@
 Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
+	\label{thm:convergence}
 \end{theorem}
@@
 
-% This class of false positive is unique to heap-allocated memory, among all ways threads could communicate.
-By contrast, global memory has unlimited lifetime, and message-passing primitives enforce a must-happens-before relationship which precludes the race.
-
 %However,
@@
 \begin{tabular}{rll}
-	%& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-	%& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
+	% TODO: Can cut struct defns for space.
+	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
+	& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
+	\\
 	& {\bf Thread 1} & {\bf Thread 2} \\
@@
 \end{figure}
-
-%%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
-%Note that under a {\em pure happens-before} analysis,
-%these accesses are not considered concurrent % at all
-%because of {\tt malloc}'s internal locking events,
-%and would not result in such false positives.
-%However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
-%so in our context it is more appropriate to use the
-%{\em limited happens-before} relation in a hybrid approach with lockset tracking.
-%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,
 
@@
 
-Note that we do not require PPs on the internal lock of {\tt malloc} or {\tt free},
-which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
-
 We implemented a simple check in \landslide~to recognize the malloc-recycle pattern:
@@
 the IDs of their containing blocks must match.
+Note that \revision{this proof does}~not require PPs on \revision{{\tt malloc}'s internal lock},
+%the internal lock of {\tt malloc} or {\tt free},
+which is an ideal candidate to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.
+
+\revision{This class of false positive is unique to heap-allocated memory among all ways threads could communicate.
+By contrast, global memory has unlimited lifetime,
+and message-passing primitives enforce an ordering which precludes the race. %(even under Limited HB).
+Finally, note that as long as concurrent {\tt malloc} is implemented with an internal lock,
+these false positives are of concern only under a Limited HB analysis (\sect{\ref{sec:overview-dr}}).
+Nevertheless,
+%our previous proof's
+Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over Pure HB.
+% TODO CAMREADY: Cite the tech report proofs, which I think call out the use of limited HB in the proof more explicitly.
+}
+
+%%Because concurrent {\tt malloc} is often implemented with an internal lock, under a {\em pure} happens-before analysis,
+%Note that under a {\em pure happens-before} analysis,
+%these accesses are not considered concurrent % at all
+%because of {\tt malloc}'s internal locking events,
+%and would not result in such false positives.
+%However, pure happens-before can miss many real bugs \cite{hybriddatarace,tsan},
+%so in our context it is more appropriate to use the
+%{\em limited happens-before} relation in a hybrid approach with lockset tracking.
+%the hybrid approach combining lockset tracking and the {\em limited} happens-before relation is not vulnerable to false negatives,

### 2016-03-23 oopsla-16/evaluation.tex
Back up shaky partial verif results with a citation to sealing.
minor
Fill in icb column on table. Back out variance paragraph. some other commence

@@
 
-\revision{
-%{\bf Variance.}
-% TODO CAMREADY: Probably will need at least a 3rd run of dis.
-Because \quicksand~can be nondeterministic in its scheduling of state spaces,
-we ran the QS-DR tests a second time to quantify the variance.
-We found the area between the two QS-DR curves to be
-% TODO
-{\bf XX\%}
-of the area between QS-DR and SSS-MC-ICB (the next closest),
-% TODO cite
-which we consider an insignificant amount of variance.
-}
+% TODO CAMREADY
+%\revision{
+%%{\bf Variance.}
+%% TODO CAMREADY: Probably will need at least a 3rd run of dis.
+%Because \quicksand~can be nondeterministic in its scheduling of state spaces,
+%we ran the QS-DR tests a second time to quantify the variance.
+%% TODO CAMREADY
+%% We found the maximum area between any two QS-DR curves to be
+%We found the area between the two QS-DR curves to be
+%% TODO
+%{\bf XX\%}
+%of the area between QS-DR and SSS-MC-ICB (the next closest),
+%% TODO cite
+%which we consider an insignificant amount of variance.
+%}
 
@@
 				% #test  sssmc   icb      qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
-		{\tt bcast}	& 79	& 7	& TODO	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
-		{\tt join} 	& 79	& 14 	&	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
-		{\tt mx} 	& 79	& 1	&	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
-		{\tt sem} 	& 79	& 10	&	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
-		{\tt signal} 	& 79	& 4	&	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
-		{\tt rwlock} 	& 79	& 26	&	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
+			% TODO: If possible, add a column for how many false negatives (incorrect verif claims) ShMem had.
+		{\tt bcast}	& 79	& 7	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
+		{\tt join} 	& 79	& 14 	& 14	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
+		{\tt mx} 	& 79	& 1	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
+		{\tt sem} 	& 79	& 10	& 11	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
+		{\tt signal} 	& 79	& 4	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
+		{\tt rwlock} 	& 79	& 26	& 27	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
 		\hline
-		{\tt sched} 	& 59	& 1	&	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
-		{\tt alarm} 	& 44	& 5	&	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
-		{\tt wait} 	& 52	& 23	&	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
+		{\tt sched} 	& 59	& 1	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
+		{\tt alarm} 	& 44	& 5	& 16	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
+		{\tt wait} 	& 52	& 23	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
 		\hline
-		{\bf Total}	& 629	& 91	&	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
+		{\bf Total}	& 629	& 91	& 104	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
 	\end{tabular}
@@
 	\caption{Summary of bugs and data races found by each test program.
-		% TODO
-		DPOR is the control; QS is \quicksand.
+		% TODO: Add Shared-Mem
+		DPOR/ICB are the controls; QS is \quicksand.
 		``DR bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
@@
 	``Malloc DRs'' counts how many malloc-recycle false positives we suppressed (\sect{\ref{sec:eval-dr}}).
-	``Mutual timeouts'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
+	% TODO CAMREADY: count timeouts for ICB instead?
+	``Mutual timeouts'' counts how often both SSS-MC-DPOR and \quicksand~timed out with no bug found;
+	% TODO CAMREADY: update avg tested subsets to be from the ICB timeouts
 	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
@@
 it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
+\revision{
+%While obviously not as strong as full verification,
+Prior work \cite{sealing} has argued the value of similar {\em compositional testing} when full verification is intractable,
+deferring to the user's expertise to judge the value of each subset of PPs verified.
+}
 
@@
 using data-race PPs increases our \revision{verification capacity}~by 4.25x.
+
+% TODO
+% Actually, while SSS-MC-Shared-Mem proved competitive with \quicksand~at finding data-race-related bugs,
+% its heuristic to ignore threads' own stack accesses introduces unsoundness.
+% <Move contents of footnote here.>
+% Indeed, SSS-MC-Shared-Mem claimed a full verification on 999999 tests for which \quicksand~found bugs!
+% % Hence, no SSS-MC strategy competes with quicksand for truly sound verification
+% Hence, only SSS-MC-DPOR and SSS-MC-ICB compete with \quicksand for sound verification, and they are thoroughly outperformed.
+% Between data-race bugs and verifications, \quicksand~is shown to provide the best of both worlds.
 

### 2016-03-24 oopsla-16/area-between-curves.c
add prelim variance qs-dr curve to area.c, but it sucks

@@ -4,8 +4,11 @@
 
 #if 1
 #define END (36000.0)
+
+#define series2_variance series2
+
 // QS DRs
-long double series1[] = {
+long double series1_qs[] = {
 36.510254, 37.739562, 38.365079, 38.637311, 39.170384, 39.501006, 39.662938,
 39.833846, 40.255752, 40.328943, 40.933243, 41.157442, 41.211145, 41.228979,
 41.351358, 41.796638, 42.160382, 42.214631, 42.510143, 43.070664, 43.583659,
@@ -35,7 +38,7 @@ long double series1[] = {
 };
 
 // ICB
-long double series2[] = {
+long double series2_icb[] = {
 17.994204, 18.004315, 19.40409, 20.560611, 21.067809, 21.254569, 22.135241,
 22.368192, 22.427883, 22.467139, 22.543117, 22.555297, 22.608603, 22.766646,
 22.79467, 23.05492, 23.166092, 23.389737, 23.605885, 24.214014, 24.654485,
@@ -52,6 +55,33 @@ long double series2[] = {
 473.037868, 569.814699, 579.142859, 582.016615, 639.690507, 673.07157,
 795.725917, 843.174624, 970.285112, 1387.089948, 1410.237694, 1822.463231,
 2582.530643, 20959.641808, END
+};
+
+
+long double series2_variance[] = {
+36.962755, 37.41802, 37.977435, 38.219288, 38.629402, 38.774045, 39.060517,
+39.087839, 39.136187, 39.496591, 40.389329, 41.456918, 41.462233, 42.187643,
+43.127638, 43.631639, 44.323396, 45.875653, 47.684681, 48.368754, 48.528354,
+48.912373, 50.635048, 51.098988, 51.646919, 52.328304, 52.659676, 52.883668,
+55.973556, 56.2661, 57.273289, 57.389903, 58.571541, 60.232219, 61.685885,
+62.206952, 63.768558, 65.638704, 66.584283, 68.599373, 68.983534, 71.039208,
+73.682596, 74.514733, 74.58467, 74.94964, 75.105398, 77.943361, 81.04122,
+82.639926, 83.678052, 84.918299, 87.300981, 93.755081, 97.191752, 101.413719,
+102.938904, 103.014602, 103.985437, 107.775351, 112.670877, 119.428875,
+122.245833, 126.767798, 128.089589, 134.723934, 145.504192, 158.995358,
[…]

### 2016-03-24 oopsla-16/design.tex
Garth's comments thru soundness

@@
 %
-Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
+Our method,
+%for doing so,
+listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
 %\footnote{
@@
 %reveals that it might finish in time after all.
-Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they would find the same bug.
+Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.
 
@@
 	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
-	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
+	\Input{$T$, the remaining time in the CPU budget}
+	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
 		return $j_0$ // Common case: job is expected to finish.
@@
 In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
-This phenomenon motivates Iterative Deepening to prioritize the jobs at run-time.
+This phenomenon motivates Iterative Deepening to prioritize jobs at run-time.
 
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
+	% TODO CAMREADY: increase back to 0.48
+	\hspace{1em}\includegraphics[width=0.45\textwidth]{dr-jobs-v2.pdf}
 	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
@@
 We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
-based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
+based on whether the MC observed the racing instructions ordered one or both ways in the original state space.
 Single-order candidates are more likely to be false positives (\sect{\ref{sec:overview-dr}}),

### 2016-03-24 oopsla-16/implementation.tex
Garth's comments thru soundness

--- 2016-03-10 oopsla-16/implementation.tex
@@
 \landslide~implements DPOR \cite{dpor},
-%\cite{dpor},
 state space estimation \cite{estimation}, and a hybrid
-%lockset/happens-before
+lockset/Limited HB
 data-race analysis \cite{hybriddatarace}.
-It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
+\revision{It can optionally replace DPOR with ICB \cite{chess-icb},
+for which it uses Bounded Partial Order Reduction \cite{bpor} for a similar reduction,
+%although we leave incorporating ICB with Iterative Deepening to future work (\sect{\ref{sec:future}})}.
+though \quicksand~does not employ this feature (\sect{\ref{sec:future}})}.
+It avoids state space cycles
+%(e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops)
+\revision{(e.g. spin
+%or {\tt yield}
+loops)}~with
+a heuristic similar to Fair-Bounded Search \cite{bpor}.
 % joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
-Its bug-detection metrics are assertion failure, deadlock, segfault, heap checking \cite{valgrind}, and a heuristic infinite loop check.
-It can test both userspace and kernel code (although it is limited to timer nondeterminism),
+Its bug-detection metrics are assertion failure, deadlock, segfault, use-after-free \cite{valgrind}, and (heuristically) infinite loops.
+%a heuristic infinite loop check.
+
+\revision{\landslide}~can test both userspace and kernel code (although it is limited to timer nondeterminism),
 and runs programs in a full-system hardware simulator \cite{simics}.
+\revision{The simulator allows \landslide~to track memory accesses and check for heap errors on uninstrumented binaries,
+although for performance, a similar MC under \quicksand~could use compiler instrumentation instead.
+It also provides a convenient backtracking mechanism to avoid the need to re-simulate common execution prefixes among many interleavings.}
 
@@
 Rather than preempting indiscriminately on any synchronization call, regardless of the call-site,
-\landslide~provides a configuration command, {\tt within\_function}, for identifying which call-sites matter.
+prior work introduced Preemption Sealing \cite{sealing} for identifying which call-sites matter.
+\landslide~provides this feature through a configuration command, {\tt within\_function}.
 %Most MC tools preempt indiscriminately on any synchronization call, regardless of the call-site.
@@
 The {\tt without\_\allowbreak{}function} directive is the dual of {\tt within\_function}, indicating a blacklist.
-We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}).
+We use these to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}).
+%\revision{similarly to Preemption Sealing from prior work \cite{sealing}}.
 %Multiple invocations can be used; later ones take precedence.
@@
 {\bf Data races in lock implementations.}
-Data race tools in prior work \cite{tsan,portend} recognize the implementations of synchronization primitives to avoid spuriously flagging memory accesses therein.
+Data race tools in prior work \cite{tsan,portend} recognize the implementations of synchronization primitives to avoid spuriously flagging memory accesses %therein.
 %resulting from the lock implementation itself.
-Assuming that the lock implementation is already correct enables more productive data-race analysis on the rest of the codebase.
-Otherwise, with testing limited to one execution,
-%even if one wishes to test for lock bugs,
-data-race analysis would flag every access pair in the lock implementation. %, requiring human attention to verify.
-However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
-Hence, we extended \landslide~to
+that implement them.
+Assuming that the lock implementation is already correct enables more productive data-race analysis on the rest of the codebase,
+% TODO CAMREADY: Cite simsa thesis
+while the locks themselves can be verified separately.
+%Otherwise, %with testing limited to one execution,
+%%even if one wishes to test for lock bugs,
+%\revision{a single-pass}~data-race analysis would flag every access pair in the lock implementation. %, requiring human attention to verify.
+%However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
+%While most tests in our evaluation assume correct locks,
+We included a mutex test in our evaluation
+to showcase \quicksand's ability to verify synchronization primitives with data-race PPs
+(\sect{\ref{sec:testsuite}}).
+To support this test, we extended \landslide~to
 %with an option
-optionally make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
-which we use to test mutexes (\sect{\ref{sec:testsuite}}).
+optionally make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()}.
+%which we use to test mutexes (\sect{\ref{sec:testsuite}}).
 %(Accesses from other synchronization functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)

### 2016-03-24 oopsla-16/intro.tex
Garth's comments thru soundness

@@
 Unfortunately, these state spaces explode exponentially in the size of the input program.
-Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
-and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} allow bugs to be found sooner in a given space should they exist.
+%Reduction
+Techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
+and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} \revision{encourage}~bugs to be found sooner in a given space should they exist.
 
@@
 but even with state-of-the-art reduction techniques,
-any choice made in advance must necessarily leave some tests unaffordably large \cite{parrot,mcr}.
+\revision{fixing the degree of coverage}~in advance necessarily leaves some tests unaffordably large \cite{parrot,mcr}.
 %This work shows how to avoid making that tradeoff decision in advance.
@@
 When prior \revision{approaches}~would time out on large tests by trying several preemption points simultaneously,
-\quicksand~\revision{identifies this pitfall in advance using state space estimation \cite{estimation},
-and instead}~tests smaller state spaces based on subsets of those points.
+\quicksand~\revision{identifies this pitfall in advance using state space size estimation \cite{estimation},
+and instead}~tests smaller state spaces based on subsets of those preemption points.
 \revision{Often, testing these}~smaller state spaces can even find the same bugs sooner.
@@
 		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
-	\item A new tactic for eliminating one class of false-positive data races,
+	\item A new tactic for eliminating one class of false-positive data race candidates,
 		which cannot soundly be used in a single-pass analysis,

### 2016-03-24 oopsla-16/overview.tex
Garth's comments thru soundness

@@
 A stateless model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
+During each execution, it forces threads to execute serially, thereby confining the program's nondeterminism to controlled thread switches.
+
 Rather than identifying suspicious conditions which may include false alarms, the approach of many static analyses \cite{racerx,coverity},
-stateless model checkers focus on concrete observed failures such as assertions, deadlocks, and segfaults.
+stateless model checkers focus on concrete observed failures such as assertions, deadlocks, segfaults, and use-after-free.
+For example, data races \revision{do not always lead to failures}, but represent suspicious violations of common locking discipline.
+Although C++ defines all data races as undefined behaviour \cite{boehm-memorymodels},
+this work focuses on reporting races to the user only when accompanied by direct evidence of a failure.
 
@@
 Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
+Iterative Context Bounding \cite{chess-icb}, another popular technique, heuristically reorders the search to prioritize interleavings with fewer preemptions first, according to the insight that most bugs require few preemptions to uncover.
 Nevertheless, state spaces are still exponentially-sized in the number of conflicting transitions.
@@
 
+\begin{figure}[t]
+	\small
+\begin{tabular}{c}
+	\revision{
+\begin{tabular}{rll}
+	& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~// A1} & \\
+	2 & \texttt{mutex\_lock(\&mx);} & \\
+	3 & \texttt{mutex\_unlock(\&mx);} & \\
+	4 & & \texttt{mutex\_lock(\&mx);} \\
+	5 & & \texttt{mutex\_unlock(\&mx);} \\
+	6 & & \texttt{\hilight{brickred}{x++;}~// A2} \\
+\end{tabular}
+}
+\\
+\revision{\normalsize (a) True potential data race.}
+% TODO CAMREADY: put another line break here
+\\
+\revision{
+\begin{tabular}{rll}
+	%& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+	%& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~// B1} & \\
+	2 & \texttt{mutex\_lock(\&mx);} & \\
+	3 & \texttt{y = true;} & \\
+	4 & \texttt{mutex\_unlock(\&mx);} & \\
+	5 & & \texttt{mutex\_lock(\&mx);} \\
+	6 & & \texttt{bool tmp = y;} \\
+	7 & & \texttt{mutex\_unlock(\&mx);} \\
+	8 & & \texttt{if (tmp) \hilight{brickred}{x++;}~// B2} \\
+	%8 & & \texttt{if (tmp)} \\
+	%9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
+\end{tabular}
+}
+\\
+\revision{\normalsize (b) No data race in any interleaving.}
+\end{tabular}
+\caption{\revision{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
+Applying HB to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
+while using Limited HB on (b) will produce a false alarm on B1/B2.}}
+\label{fig:hb-example}
+\end{figure}
+
 Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
@@
 and no synchronization enforces an order on the thread transitions \revision{(the {\em Happens-Before} relation)}.
-Data races \revision{do not always lead to failures}, but represent suspicious violations of common locking discipline.
 In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
@@
 \revision{
-	{\bf Variants of Happens-Before (HB).}
-	Most prior work focuses on {\em Pure HB} \cite{lamport-clocks} as the order relation between accesses.
+	{\bf Variants of Happens-Before.}
+	Most prior work focuses on {\em Happens-Before} (HB) \cite{lamport-clocks} as the order relation between accesses.
 \cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
@@
 but at the cost of necessarily reporting some such {\em false positives}.
-Finally, the {\em Causally-Precedes} relation \cite{predictive-dr} strikes a middle ground,
-extending Pure HB to additionally report a subset of potential races while soundly avoiding false positives.
+
+Finally, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
+extends HB to additionally report a subset of potential races while soundly avoiding false positives.
+It tracks conflicting accesses in intervening
+critical sections to determine whether lock events are unrelated to a potential race.
+Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
+although it can still miss true potential races in other cases.
 
 Stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx}.
-However, this work incorporates data-race analysis in an internal feedback loop,
-using MC to automatically check each potential race
-and reporting to the user only directly observed failures.
+%However,
+This work incorporates data-race analysis in an internal feedback loop,
+using model checking to automatically test each potential race
+and report only directly observed failures to the user.
 %Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})
@@
 accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{c}
-	\revision{
-\begin{tabular}{rll}
-	& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
-	& {\bf Thread 1} & {\bf Thread 2} \\
-	1 & \texttt{\hilight{brickred}{x++;}~// A1} & \\
-	2 & \texttt{mutex\_lock(\&mx);} & \\
-	3 & \texttt{mutex\_unlock(\&mx);} & \\
-	4 & & \texttt{mutex\_lock(\&mx);} \\
-	5 & & \texttt{mutex\_unlock(\&mx);} \\
-	6 & & \texttt{\hilight{brickred}{x++;}~// A2} \\
-\end{tabular}
-}
-\\
-\revision{\normalsize (a) True potential data race.}
-\\
-\\
-\revision{
-\begin{tabular}{rll}
-	%& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
-	%& {\bf Thread 1} & {\bf Thread 2} \\
-	1 & \texttt{\hilight{brickred}{x++;}~// B1} & \\
-	2 & \texttt{mutex\_lock(\&mx);} & \\
-	3 & \texttt{y = true;} & \\
-	4 & \texttt{mutex\_unlock(\&mx);} & \\
-	5 & & \texttt{mutex\_lock(\&mx);} \\
-	6 & & \texttt{bool tmp = y;} \\
-	7 & & \texttt{mutex\_unlock(\&mx);} \\
-	8 & & \texttt{if (tmp) \hilight{brickred}{x++;}~// B2} \\
-	%8 & & \texttt{if (tmp)} \\
-	%9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
-\end{tabular}
-}
-\\
-\revision{\normalsize (b) No data race in any interleaving.}
-\end{tabular}
-\caption{\revision{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
-Applying Pure HB to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
-while using Limited HB on (b) will produce a false alarm on B1/B2.}}
-\label{fig:hb-example}
-\end{figure}
 
@@
 {\em single-state-space model checking} (SSS-MC), % (i.e., the approach of prior work),
-{\em Dynamic Partial Order Reduction} (DPOR), and {\em state space estimate} (ETA).
+{\em Dynamic Partial Order Reduction} (DPOR),
+{\em Iterative Context Bounding} (ICB),
+and {\em state space estimate} (ETA).
 
@@
 
-Although all data races are undefined behaviour in C and C++ \cite{boehm-memorymodels},
-we distinguish between data-race {\em candidates} and data-race {\em bugs}.
+We distinguish between data-race {\em candidates} and data-race {\em bugs}.
 \revision{We refer to racing (or potentially-racing) memory access pairs as}~{\em data race candidates}.
-Should a future interleaving, preempting during such accesses,
+Should %a future interleaving,
+preempting during such accesses
 lead to \revision{an observable}~failure, then we report a {\em data-race bug}.
-Otherwise, if the access pair can be reordered \revision{to be simultaneous}, but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
+Otherwise, if the access pair can be reordered,
+%\revision{to be simultaneous},
+but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
 If they cannot be reordered at all, due to some other communication \revision{such as in Figure~\ref{fig:hb-example}(b)}, it is a {\em false positive}.
@@
 We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}, A).
+The {\em minimal state space} includes only thread switches arising from no preemptions (Figure~\ref{fig:id}, A).
 The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}, D).

### 2016-03-24 oopsla-16/warpzone.tex
Garth's comments thru soundness

@@
 While we guarantee safety when using certain combinations of PPs (\sect{\ref{sec:eval-sssmc}}),
-Iterative Context Bounding
+%Iterative Context Bounding
+ICB
 guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.

### 2016-03-24 oopsla-16/evaluation.tex
Last of garth's commence

@@
 to promote completing subset PP sets for large tests.
-Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
-we also evaluated each technique individually.
+Hence,
+we also evaluated each technique individually,
+though many of our experiments compare \quicksand~to the state-of-the-art as a whole.
 Our evaluation answers the following questions:
@@
 				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
-			\revision{\item} Does Iterative Deepening find bugs faster
+			\revision{\item} Does \quicksand~find bugs faster
 				%than SSS-MC
@@
 			\revision{
-			\item Does Iterative Deepening provide more full verifications of bug-free programs more quickly?
+			\item Does \quicksand~provide more full verifications of bug-free programs more quickly?
 			}
@@
 		\begin{enumerate}
-			\item Do we avoid false positives compared to a single-execution data race analysis?
+			\item Does \quicksand~avoid false positives compared to a single-execution data race analysis?
 				% Explain later as:
@@
 				% Answer: Just 1. (for p2s at least; as of time of writing, pintos dr-falsenegs not run yet)
-			\item Do we find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
+			\item Does \quicksand~find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
 		\end{enumerate}
@@
 %done
-and the Pintoses average 718 lines, for a total of 198,772.
+and the Pintoses average 718 lines, for a total of 198,772 lines tested for this paper.
 
@@
 \tej, a test of thread lifecycle,
-\bct~and \paraguay~for cvars,
+\bct~and \paraguay~for condition variables,
 \paradise~for semaphores,
-and \rwldgr~for r/w locks.
+% XXX f u latex
+and \rwldgr\\for r/w locks.
 For \mxtest, \paradise, and \paraguay, we used the {\tt without\_function} command to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
@@
 % at least 157 of which could expose bugs.
-at least 165 of which could expose bugs. % QS bugs + bugs where SSSMC found but QS didnt
+at least 165 of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
 % TODO: update this number with any new bugs found by PREEMPT_EVERYWHERE
@@
 \begin{itemize}
-	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with plain DPOR \cite{dpor},
+	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with DPOR \cite{dpor},
 	\item {\bf SSS-MC-ICB}: With PPs as above, but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster, and
@@
 Because parallelizing DPOR/ICB during SSS-MC is an open research problem \cite{parallel-dpor},
-we instead gave control experiments 10 hours per test with 1 CPU.
+we optimistically gave control experiments a linear speedup of 10 hours per test with 1 CPU.
 %To match this,
-\quicksand~reports the CPU-time spent alongside the wall-clock time for a direct comparison
-(although, with the growing importance of multicore, \quicksand's inherent parallelism is a convenient benefit).
+\quicksand~reports the CPU-time spent in addition to the wall-clock time for a direct comparison
+(although, with the growing importance of multicore for performance, \quicksand's inherent parallelism is a convenient benefit).
 %Finally,
@@
 
-Figure~\ref{fig:dowefindbugsfaster} plots the cumulative distribution of bugs found \revision{by each experiment}.
+Figure~\ref{fig:dowefindbugsfaster} plots the cumulative distribution of bugs found \revision{by each experiment}
+against the time taken to find each bug.
 \revision{
-Figure~\ref{fig:dowefindbugsfaster}(a) is the main comparison by CPU-time;
+Figure~\ref{fig:dowefindbugsfaster}(a) is the main, resource-fair comparison by CPU-time;
 we additionally show a wall-clock comparison in (b) to highlight the impact of \quicksand's parallelism.
@@
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
-		(a) Bugs found by elapsed CPU time. Overall, a more direct comparison than (b),
+		(a) Bugs found as a function of elapsed CPU time. Overall, a more resource-fair comparison than (b),
 		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
@@
 However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
-In Figure~\ref{fig:dowefindbugsfaster}(b), we give \quicksand~full credit for its inherent parallelism, and it outperforms SSS-MC for any fixed budget of wall-clock time.}
+In Figure~\ref{fig:dowefindbugsfaster}(b), we give \quicksand~full credit for its inherent parallelism,
+and with a 10 core allocation, it outperforms SSS-MC for any fixed budget of wall-clock time.}
 
@@
 %Though for many applications of MC, assuming a correct lock implementation is sufficient,
+% TODO CAMREADY cite simsa thesis
 Though it often suffices to assume correctly-implemented locks,
@@
 its results are competitive with SSS-MC-DPOR, but SSS-MC-ICB outperforms it.
-This is unsurprising: the seed subsets of PPs (\sect{\ref{sec:initial-pp}}) QS-Sync-Only is limited to are much less flexible than ICB's preemption strategy.
+This is unsurprising: the seed subsets of PPs
+QS-Sync-Only is limited to
+(\sect{\ref{sec:initial-pp}})
+are much less flexible than ICB's preemption strategy (\sect{\ref{sec:overview-sssmc}}).
 This result suggests that in future work, \quicksand~should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.}
@@
 %we found that
-the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking.
+the sheer number of accesses during each new execution
+% TODO
+%(often hundreds)
+added significant performance overhead to the MC when computing DPOR and backtracking.
 Iterative Deepening avoids this overhead by waiting until runtime to identify fewer, more relevant PPs dynamically,
@@
 %before a given elapsed CPU consumption.
+% TODO TODO TODO decide whether to include dis or not
 \revision{against those provided by SSS-MC-Shared-Mem\footnote{
+	% TODO TODO TODO if not plot this curve, move this discussion into main body
 	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
@@
 % TODO
-{\bf 99x}
+{\bf 9999x}
 as many tests in 10 CPU-hours.
@@
 
-Beyond finding new bugs with data-race PPs, we evaluated \quicksand's performance for classifying data-race candidates in two ways.
+Beyond finding new bugs and completing full verifications with data-race PPs, we evaluated \quicksand's performance for classifying data-race candidates in two ways.
 
@@
 could fail
-to identify a racing access pair as a candidate to begin with.
+to identify a racing access pair as a candidate at all. %to begin with.
 %

### 2016-03-24 oopsla-16/paper.tex
Write expanded explanation of HB for background section.
Expand related work to address PLDI reviewer concerns.
Fix miscounting in QS bugs, 163->157; and make ICB bug total the new denominator for claimed x-factor
Garth's comments thru soundness
Last of garth's commence

@@
 \usepackage{graphicx}
-\usepackage{courier} % TODO: remove this? and get nicer fonts?
 \usepackage{amsthm}
@@
 %\newcommand\revision[1]{\cbstart#1\cbend}
-%\newcommand\revision[1]{#1}
+\newcommand\revision[1]{#1}
 \definecolor{purple}{RGB}{192,0,192}
-\newcommand\revision[1]{\hilight{purple}{#1}}
+%\newcommand\revision[1]{\hilight{purple}{#1}}
 
@@
 It also enables full verification of all possible schedules when every data race is verified as benign within the CPU budget.
-In our evaluation, Iterative Deepening
+In our evaluation, \quicksand~found
 %with data-race preemption points
-found 1.8x as many bugs and verified 4.3x as many tests compared to stateless model checking without data-race preemption points.
+1.5x as many bugs and verified 4.3x as many tests compared to \revision{prior model checking approaches}. %stateless model checking without data-race preemption points.
 %Our evaluation shows this technique is
@@
 finding new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
-By using \numstudence~student projects as our test suite, we also show the potential benefit as a debugging platform in educational settings.
+By using \numstudence~student operating system implementation projects as our test suite, we also show the potential benefit as a debugging platform in educational settings.
 Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.

### 2016-03-24 oopsla-16/related.tex
Garth's comments thru soundness
Last of garth's commence

@@
 Future work could use ETAs to heuristically switch between DPOR and PCT}.
-Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to our work,
+Finally, ICB \cite{chess-icb} is most similar to our work,
 as both approaches provide a partial verification
@@
 %Too many related projects to list have made contributions to the
-Many advances have been made on the false-positive data race problem since it was first introduced in \cite{eraser}.
+Many advances have been made on the false-positive potential data race problem since it was first introduced in \cite{eraser}.
 \cite{hybriddatarace} and \cite{tsan}
@@
 \revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.
+
 \revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
@@
 IFRit \cite{ifrit}
-improves performance using an interference analysis,
+improves the performance of HB using an interference analysis,
 which could allow future work to avoid tracing every memory access.

### 2016-03-24 oopsla-16/related.tex
Expand on SATcheck and MCR.

@@
 % TODO: Expand on MCR and SATcheck.
-{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
+{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter},
+%Maximal Causality Reduction \cite{mcr},
 %DPOR for relaxed memory models \cite{tsopso},
-SAT-directed MC \cite{satcheck},
+%SAT-directed MC \cite{satcheck},
 \revision{and R4 for event-driven applications \cite{r4}}.
 These are all compatible with our technique.
+%
 Recent work \cite{tsopso} has extended DPOR for relaxed memory models \cite{memory-consistency-models},
 which we do not yet account for in our proofs (\sect{\ref{sec:soundness}}).
+%
+SATCheck \cite{satcheck} and Maximal Causality Reduction \cite{mcr} replace DPOR by
+%encoding the program's behaviour as a SAT or SMT formula,
+%then using a solver to search for schedules guaranteed to expose new behaviour.
+using a SAT or SMT solver to search for schedules guaranteed to expose new behaviour.
+%These approaches provide improved
+They improve
+reduction by considering values read and written to identify additional independences,
+while DPOR considers only addresses.
+They generate new schedules at
+%the granularity of each memory access,
+memory access granularity,
+which Iterative Deepening could allow to be relaxed for large tests in future work.
+%
 Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial \revision{tests}~in their evaluation could be completed,
 \revision{which motivates \quicksand's CPU-budget-oriented approach.}
-%providing a strong argument for \quicksand.
 

### 2016-03-24 oopsla-16/proof/di.tex
Import proof
Fix definitions in proof for oopsla crowd

--- 2015-11-20 pldi-16/proof/di.tex
@@
 
+In our main paper, we refer to this relation as Limited HB.
+
 \begin{definition}[Shared memory conflict]
@@
 
-The same as in the paper, we distinguish between data-race {\em candidates} and data-race {\em bugs}.
-In this proof we are concerned solely with candidates, and wheter they can be observed to race or are false positives.
+The same as in the paper, we distinguish between data-race {\em candidates} (or {\em potential} data races) and data-race {\em bugs}.
+For brevity, we now use ``data race'' to refer both to true races and to potential data-race candidates identified by MHB.
+In this proof we are concerned solely with candidates, and whether they can be observed to race or are false positives.
 It is up to the MC to decide whether true data races are benign or buggy.
@@
 %Most commonly this refers to accesses to a region already freed, but for brevity we also include
-% TODO: need heap block overrun?
 

### 2016-03-24 oopsla-16/evaluation.tex
some tiny things
present sssmc-shmem results, and clean up some todos

@@
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
-		(a) Bugs found as a function of elapsed CPU time. Overall, a more resource-fair comparison than (b),
+		% TODO CAMREADY "as a function of elapsed"
+		(a) Bugs found by elapsed CPU time. Overall,
+		% TODO CAMREADY "a more resource-fair"
+		a more direct comparison than (b),
 		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
@@
 \revision{Before the break-even point at 200 seconds,
-% TODO
-% (for SSS-MC-ICB) or XXX seconds (for SSS-MC-Shared-Mem)
 \quicksand~lags behind the SSS-MC approaches due to additional start-up overhead from its tenfold parallelism.
@@
 In \mxtest, in which we do not trust the lock implementation's correctness,
-the control experiment found dramatically fewer bugs (just 1)\footnote{
+% TODO CAMREADY: Fix, SSSMC-ShM finds 2 mx bugs.
+the control experiments
+found dramatically fewer bugs (just 1)\footnote{
 	The one bug SSS-MC found was in a fully-assembler lock implementation.
@@
 
-% TODO: fix this table.
 \begin{table*}[t]
@@
 	\begin{tabular}{r|c||c|c|c|c|c|c||c|c||c|c}
-		% TODO: Update SSSMC column to have more favourable ICB numbers.
 		& {\bf Num.} & {\bf DPOR} & {\bf ICB} & {\bf QS} & {\bf DR} & {\bf Nondet.} & {\bf Malloc} & {\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
@@
 	\caption{Summary of bugs and data races found by each test program.
-		% TODO: Add Shared-Mem
+		% TODO CAMREADY: Add Shared-Mem
 		DPOR/ICB are the controls; QS is \quicksand.
@@
 the sheer number of accesses during each new execution
-% TODO
-%(often hundreds)
+(often thousands)
 added significant performance overhead to the MC when computing DPOR and backtracking.
@@
 {\bf Full verification.}
-% TODO: Discuss how ICB sucks even more.
 For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
-In Figure~\ref{fig:totalverif} we plot the cumulative distribution of \quicksand's total verifications
-%before a given elapsed CPU consumption.
-% TODO TODO TODO decide whether to include dis or not
-\revision{against those provided by SSS-MC-Shared-Mem\footnote{
-	% TODO TODO TODO if not plot this curve, move this discussion into main body
-	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
+In Figure~\ref{fig:totalverif} we plot the cumulative distribution of
+%total
+verifications provided by each approach.
+Though SSS-MC-Shared-Mem appears to outperform \quicksand, its safety guarantee is unsound,
+as many tests break its assumption of stack-allocated memory being private.
 For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
-	SSS-MC-Shared-Mem's verifications are only sound in the absence of such sharing!
-	\quicksand~detects these dynamically, without needing to discriminate between stack and heap.}
-	%Expanding the statically-coded PPs to include stack accesses would make SSS-MC-Shared-Mem's performance even worse.
-	% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
+While SSS-MC-Shared-Mem must avoid excessive PPs on every private stack access for performance,
+\quicksand~detects these dynamically, without needing to discriminate between stack and heap.
+In total, SSS-MC-Shared-Mem claimed to verify 21 tests
+(across 7 of the 9 test cases)
+in which QS-DR ultimately found bugs,
+which further calls into question the rest of its completions.
+
+%\quicksand's total verifications
+%%before a given elapsed CPU consumption.
+%\revision{against those provided by SSS-MC-Shared-Mem\footnote{
+%	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
+%	For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
+%	SSS-MC-Shared-Mem's verifications are only sound in the absence of such sharing!
+%	\quicksand~detects these dynamically, without needing to discriminate between stack and heap.}
+%	%Expanding the statically-coded PPs to include stack accesses would make SSS-MC-Shared-Mem's performance even worse.
+%	% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization.
 Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
-}.
-\quicksand~outperforms SSS-MC, being able to verify safety for
-% TODO
-{\bf 9999x}
-as many tests in 10 CPU-hours.
-We attribute this again to the overhead of the sheer number of statically-coded PPs that SSS-MC-Shared-Mem must use,
-as well as to the work which ICB necessarily repeats when increasing its preemption bound.}
+%}.
+%\quicksand~outperforms SSS-MC, being able to verify safety for
+%{\bf 9999x}
+%as many tests in 10 CPU-hours.
+%We attribute this again to the overhead of the sheer number of statically-coded PPs that SSS-MC-Shared-Mem must use,
+%as well as to the work which ICB necessarily repeats when increasing its preemption bound.}
 
-Among these verified tests, 36 contained no data-race candidates whatsoever,
+Among QS-DR's verified tests, 36 contained no data-race candidates whatsoever,
 so the same verification could be provided
@@
 
-% TODO
 % Actually, while SSS-MC-Shared-Mem proved competitive with \quicksand~at finding data-race-related bugs,
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{totalverifs-squashed.pdf}
+	\includegraphics[width=0.48\textwidth]{totalverifs-squashed-with-unsound-shm.pdf}
 	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
-	Some tests had no data-race candidates, and hence could also be verified by SSS-MC.}
+	%SSS-MC-Shared-Mem completed 1.37x as many tests, but 
+	SSS-MC-Shared-Mem is unsound: 21 tests it completed hid bugs previously found in Figure~\ref{fig:dowefindbugsfaster}.
+	%previously found (Figure~\ref{fig:dowefindbugsfaster}).
+	%Some tests had no data-race candidates, and hence could also be verified by .}
+	36 data-race-free tests were also soundly verified by QS-Sync-Only, SSS-MC-DPOR, and SSS-MC-ICB.}
 	\label{fig:totalverif}
@@
 Likewise, replay-based tools \cite{portend} are dependent upon the data-race detector to provide an execution trace leading to each candidate.
-This experiment suggests that
-such tools could benefit from a similar feedback loop as used in Iterative Deepening.
+This result suggests that
+such tools could benefit from a similar feedback loop as is used in Iterative Deepening.
 %i.e., discovering transitively-reachable data races while testing initial ones.

### 2016-03-24 oopsla-16/evaluation.tex
Improve the prose.

@@
 For all tests, we used {\tt without\_function} to blacklist PPs on {\tt malloc}'s internal lock.
-In total, our evaluation comprises 629 unique tests (pairs of a test program and a Pintos or P2),
+In total, our evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
 % at least 157 of which could expose bugs.
-at least 165 of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
-% TODO: update this number with any new bugs found by PREEMPT_EVERYWHERE
+at least 173 of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
 
@@
 	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
-		Using ICB+BPOR, configured to insert PPs on any shared memory access
-		(determined at runtime, excluding threads' accesses to their own stacks),
+		Using ICB+BPOR, configured to preempt on any shared memory access \cite{inspect}
+		(decided at runtime, excluding threads' accesses to their own stacks),
 		which in principle includes all possible data-race PPs.
@@
 %To match this,
-\quicksand~reports the CPU-time spent in addition to the wall-clock time for a direct comparison
+\quicksand~reports the CPU-time spent in addition to the wall-clock time for a resource-fair comparison
 (although, with the growing importance of multicore for performance, \quicksand's inherent parallelism is a convenient benefit).
@@
 However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
-In Figure~\ref{fig:dowefindbugsfaster}(b), we give \quicksand~full credit for its inherent parallelism,
-and with a 10 core allocation, it outperforms SSS-MC for any fixed budget of wall-clock time.}
+Figure~\ref{fig:dowefindbugsfaster}(b) gives \quicksand~full credit for its inherent parallelism:
+with a 10 core allocation, it outperforms SSS-MC for any fixed budget of wall-clock time.}
 
@@
 %We found the area between the two QS-DR curves to be
-%% TODO
+%% TODO CAMREADY
 %{\bf XX\%}
 %of the area between QS-DR and SSS-MC-ICB (the next closest),
-%% TODO cite
+%% TODO cite CAMREADY
 %which we consider an insignificant amount of variance.
@@
 %Beyond using the state space estimator to guess the percent coverage,
-We know of no approach to quantify the probability
-that a bug would be exposed by an untested interleaving,
-but \quicksand~at least reports which subsets of PPs resulted in state spaces that did complete in time.
+%We know of no approach to quantify the probability
+%that a bug would be exposed by an untested interleaving,
+While recent work \cite{randomized-scheduler} attempts to quantify the probability
+that an untested interleaving would expose a bug,
+\quicksand~instead reports which subsets of PPs resulted in state spaces that did complete in time.
 %\quicksand~does not attempt any such quantitative guarantee,
 %but can at least report
-On 129 tests, the SSS-MC control experiment timed out after 10 hours with no bugs found.
+% TODO: Fix to be ICB not DPOR.
+On 129 tests, SSS-MC-DPOR timed out after 10 hours with no bugs found.
 Among these tests, \quicksand~found bugs in 36.
@@
 which further calls into question the rest of its completions.
+
+% NB. TODO CAMREADY. If a reviewer asks, "How come SSSMC-ShM completed so many state spaces,
+% even though last section you argued that so many PPs slows it down terribly?"
+% The answer is, "Actually, it's not that SSSMC-ShM magically got fast again for the verification experiment;
+% it's that SSSMC-ShM and Quicksand are *BOTH* slow to achieve full completion.
+% Quicksand can get bogged down in a combinatorial explosion of different subsets of PPs,
+% having a zillion jobs on its hands. Future work should improve this with https://github.com/bblum/landslide/issues/216 ."
 

### 2016-03-24 oopsla-16/intro.tex
Improve the prose.

@@
 \revision{To achieve the same level of verification, prior model checkers must decide}~in advance to preempt on every single memory access \cite{spin}, \revision{which is computationally prohibitive for
-%even moderately-sized
-larger tests.
+even moderately-sized
+%larger
+tests.
 Our approach provides the best of both worlds:
@@
 %We show many advantages over both conventional stateless model checking and single-pass data-race analysis.
-We find many of the same bugs found by the conventional approach faster using subsets of preemption points,
-and that data-race preemption points quickly expose many new bugs that prior model checkers could not find at all.
+%We find many of the same bugs found by the conventional approach faster using subsets of preemption points,
+We find that data-race preemption points quickly expose many new bugs that prior model checkers could not find at all,
+and that they enable full verification of many more tests than before.
+%and that they enable full verification of many racy, but non-failing, tests.
 

### 2016-03-24 oopsla-16/paper.tex
some tiny things
present sssmc-shmem results, and clean up some todos
Improve the prose.

@@
 
-% TODO: remove all changebars before submission
-%\setcounter{changebargrey}{0}
 %\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
@@
 Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
-It avoids concerns of state space size, but is prone to false positives:
-spurious reports of access pairs that cannot be reordered to produce a bug.
+It avoids concerns of state space size,
+%but is prone to false positives:
+%spurious reports of access pairs that cannot be reordered to produce a bug.
+%but data races do not always correspond to observable failures.
+but may report races that do not lead to observable failures,
+which jeopardizes a user's willingness to use the analysis.
 
@@
 finding new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
+Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.
+Between data-race bugs and verifications, \quicksand~is shown to provide the best of both worlds.
 By using \numstudence~student operating system implementation projects as our test suite, we also show the potential benefit as a debugging platform in educational settings.
-Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.
 
@@
 
-% TODO: proofs->appeendix
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
+%\appendix
+%\section{Appendix Title}
+%
+%This is the text of the appendix, if you need one.
 
@@
 
-Acknowledgments, if needed.
+%Acknowledgments, if needed.
+{\em omitted for double-blind review}
 

### 2016-03-24 oopsla-16/related.tex
present sssmc-shmem results, and clean up some todos
Improve the prose.

@@
 %
-Inspect \cite{inspect} instruments source code by inserting wrapper calls around all accesses to potentially-shared data.
+Inspect \cite{inspect} instruments source code by
+%inserting wrapper calls around
+instrumenting
+all accesses to potentially-shared data.
 It identifies such instructions in advance with an over-approximating static alias analysis,
@@
 
-% TODO: Expand on MCR and SATcheck.
-{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter},
+{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Optimal DPOR \cite{optimal-dpor}, Dynamic Interface Reduction \cite{demeter},
 %Maximal Causality Reduction \cite{mcr},
@@
 \revision{cannot estimate remaining time to total verification,
-and cannot incorporate data-race PPs not statically coded in advance}.
+and can incorporate data-race PPs only when statically coded in advance}.
 %and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

### 2016-05-20 oopsla-16/response.txt
add prelim author response

We agree with reviewer A's concern about possible bias in the set of
benchmarks. To address this, we will add some analysis of the types of bugs
found. For example, in the Iterative Context Bounding control experiment, 82
bugs were found with preemption bound 1, 15 with bound 2, and 2 with bound 3.
This is consistent with the distribution in [1] and shows our bugs are not
biased towards requiring high preemption bounds.

Our focus on maximizing completed state spaces is motivated by [2], in which
developers prefer compositional testing (manually restricting a test's scope
to complete faster), which suggests that full completions of partial state
spaces are easier for humans to interpret than partial completions of full
spaces. Our evaluation also shows that subset state spaces generally uncover
bugs faster (QS-Sync-Only vs SSS-MC-DPOR). However, your suggestion to
sometimes complete larger jobs first, thereby pruning any subsumed smaller
jobs, is an insightful optimization which we could implement in future work.

If the race detector produces a candidate where only one of two orders is
feasible, the MC will subsequently classify it as a false positive.

To address reviewer B, we choose new preemption points to add for each
data-race candidate, one for each racing instruction. This produces several
new jobs, each with a different combination of the new ones with existing
ones. For the sake of soundness, which assumes an infinite CPU budget, it
would suffice to randomly choose among new jobs until all are exhausted, or
even to always choose the biggest (maximal) job and ignore subset jobs.
Section 5's proofs assume this simple approach (call it Algorithm 0).
Algorithm 1 in section 3 optimizes algorithm 0 for finite CPU budgets to
prioritize finding bugs faster. Sorry for the confusion and we will clarify
this, including Algorithm 0 explicitly.

Regarding the suggestion of Maximal Causality Reduction [3], we were not able
to do a direct comparison because that paper has no open-source
implementation. We agree it is not necessarily orthogonal to our approach, but
could be combined in future work when an implementation becomes available.

We will clarify in our introduction that we assume sequentially-consistent
hardware, although in principle our work could be extended to weak memory
models using the techniques in [4].

Regarding Figure 3, the contrived program is just an example to explain the
background. We agree other approaches exist which can solve this case.

Reviewer C insightfully observes that Figure 7's count of bugs excludes
data-race candidates from the control experiments. This is the mentioned
philosophical difference about data races versus concrete failures; to address
this, we will add a table listing how many data races the control experiments
reported, and how many of those Quicksand refuted as false positives.

Reviewer D mentions that both HB and Limited HB will find all data races when
combined with model checking. We suspected this was true, but were only able
to prove the case for Limited HB. If you could provide a citation which proves
the case of regular HB combined with model checking, we are interested to read
it and would gladly include it for cam-ready. However, one further advantage
of Limited HB is that it reports candidates sooner, allowing Quicksand to test
them immediately, at the expense of more false positives. We will add a
discussion of this trade-off.

Regarding our malloc-recycle solution, we tried your suggestion to never
reallocate memory, but this unacceptably impacted the performance of some
malloc-heavy tests. Our proof allows eliminating these data-race candidates
without hacking the allocator.

Thanks for the clarification about FastTrack and CHESS; we will improve our
discussion of those tools.

Finally, reviewers A and D both criticized the formalism for unclear
definitions and notation. We agree and will work to improve the foundation and
notation definitions for cam-ready. For example, in the definition of MHB, we
define "cannot be reordered" in terms of the ability for threads to be
scheduled at prior preemption points, with no intermediate blocking operation
by the latter thread (this coincides with DPOR's approach to identify
alternate interleavings to explore).

[1] Musuvathi and Qadeer, Iterative Context Bounding, PLDI07
[2] Ball et al, Preemption Sealing, TACAS10
[3] Huang, Maximal Causality Reduction, PLDI15 ([24] in the paper/reviews)
[4] Zhang et al, DPOR for relaxed memory, PLDI15

### 2016-05-20 oopsla-16/response.txt
response 600 words exacto

@@
-We agree with reviewer A's concern about possible bias in the set of
-benchmarks. To address this, we will add some analysis of the types of bugs
-found. For example, in the Iterative Context Bounding control experiment, 82
-bugs were found with preemption bound 1, 15 with bound 2, and 2 with bound 3.
-This is consistent with the distribution in [1] and shows our bugs are not
-biased towards requiring high preemption bounds.
+We agree with reviewer A about possible bias in the benchmarks. To address
+this, we will add some analysis of the types of bugs found. For example, in
+the Iterative Context Bounding control experiment, 82 bugs were found with
+preemption bound 1, 15 with bound 2, and 2 with bound 3, consistent
+with the distribution in [Musuvathi PLDI07]. This shows no bias towards
+requiring high preemption bounds.
 
-Our focus on maximizing completed state spaces is motivated by [2], in which
+Maximizing completed state spaces is motivated by [Ball TACAS10], in which
 developers prefer compositional testing (manually restricting a test's scope
@@
 spaces are easier for humans to interpret than partial completions of full
-spaces. Our evaluation also shows that subset state spaces generally uncover
+spaces. We also showed subset state spaces generally uncover
 bugs faster (QS-Sync-Only vs SSS-MC-DPOR). However, your suggestion to
-sometimes complete larger jobs first, thereby pruning any subsumed smaller
-jobs, is an insightful optimization which we could implement in future work.
+sometimes complete larger jobs first, to prune subsumed smaller jobs, is an
+insightful optimization for future work.
 
-If the race detector produces a candidate where only one of two orders is
-feasible, the MC will subsequently classify it as a false positive.
-
-To address reviewer B, we choose new preemption points to add for each
+Reviewer B: We choose new preemption points to add for each
 data-race candidate, one for each racing instruction. This produces several
@@
 Section 5's proofs assume this simple approach (call it Algorithm 0).
-Algorithm 1 in section 3 optimizes algorithm 0 for finite CPU budgets to
-prioritize finding bugs faster. Sorry for the confusion and we will clarify
-this, including Algorithm 0 explicitly.
+Algorithm 1 in section 3 optimizes algorithm 0 for finite CPU budgets,
+choosing the state space with smallest ETA. We will clarify this, including
+Algorithm 0 explicitly.
 
-Regarding the suggestion of Maximal Causality Reduction [3], we were not able
-to do a direct comparison because that paper has no open-source
-implementation. We agree it is not necessarily orthogonal to our approach, but
-could be combined in future work when an implementation becomes available.
+Regarding the suggestion of Maximal Causality Reduction [Huang PLDI15], we
+were not able to do a direct comparison because that paper has no open-source
+implementation. This is a compelling direction for future work.
 
 We will clarify in our introduction that we assume sequentially-consistent
-hardware, although in principle our work could be extended to weak memory
-models using the techniques in [4].
+hardware. Future work could extend our implementation and proofs using the
+techniques in [Zhang PLDI15].
 
-Regarding Figure 3, the contrived program is just an example to explain the
-background. We agree other approaches exist which can solve this case.
+Figure 2 is meant to show the state of an arbitrary in-progress search,
+although we agree it needs work. Regarding Figure 3, the contrived program is
+just an example to explain the background. We agree other approaches exist
+which can solve this case. Thanks for the suggestion about graph search; we
+will look into it.
 
@@
 Reviewer D mentions that both HB and Limited HB will find all data races when
-combined with model checking. We suspected this was true, but were only able
-to prove the case for Limited HB. If you could provide a citation which proves
-the case of regular HB combined with model checking, we are interested to read
-it and would gladly include it for cam-ready. However, one further advantage
-of Limited HB is that it reports candidates sooner, allowing Quicksand to test
-them immediately, at the expense of more false positives. We will add a
-discussion of this trade-off.
+combined with model checking. We suspected this was true, but have only proven
+the case for Limited HB. Could you provide a citation for us to include during
+revisions? One further advantage of Limited HB is reporting candidates sooner,
+allowing Quicksand to test them immediately, at the expense of more false
+positives. We will clarify this trade-off.
 
-Regarding our malloc-recycle solution, we tried your suggestion to never
-reallocate memory, but this unacceptably impacted the performance of some
-malloc-heavy tests. Our proof allows eliminating these data-race candidates
-without hacking the allocator.
+Regarding our malloc-recycle solution, we tried your never reallocating
+memory, but this unacceptably impacted malloc-heavy tests. Our proof allows
+eliminating these data-race candidates without hacking the allocator.
 
@@
 
-Finally, reviewers A and D both criticized the formalism for unclear
-definitions and notation. We agree and will work to improve the foundation and
-notation definitions for cam-ready. For example, in the definition of MHB, we
-define "cannot be reordered" in terms of the ability for threads to be
-scheduled at prior preemption points, with no intermediate blocking operation
-by the latter thread (this coincides with DPOR's approach to identify
-alternate interleavings to explore).
-
-[1] Musuvathi and Qadeer, Iterative Context Bounding, PLDI07
-[2] Ball et al, Preemption Sealing, TACAS10
-[3] Huang, Maximal Causality Reduction, PLDI15 ([24] in the paper/reviews)
-[4] Zhang et al, DPOR for relaxed memory, PLDI15
+Finally we agree with reviewers A and D and will work to improve the
+formalism's notation and definitions. For example, in the definition of MHB,
+we define "cannot be reordered" in terms of the latter thread's ability to be
+scheduled at all intervening preemption points, with no intermediate blocking
+operation (this coincides with DPOR's approach to identify alternate
+interleavings to explore).

### 2016-07-14 oopsla-16/design.tex
refactor design section intro and write algorithm0

@@
 In this context, the depth is the number of PPs used.
-Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs.
-We refer to each unique set of PPs as a {\em job},
-and prioritize them based on number of PPs, ETA, and whether they include data-race PPs,
+Hence,
+%our tool
+\quicksand~schedules multiple MC instances in parallel to
+%simultaneously
+test many different subsets of the available PPs,
+We refer to each unique set of PPs as a {\em job}.
+
+Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
+A MC tool is still used to test each state space, and other reduction techniques are still applicable.
+Moreover, because Iterative Deepening treats the set of preemption points as mutable,
+it can add new preemption points reactively based on any runtime analysis.
+We focus on run-time data-race detection~\cite{hybriddatarace,tsan,ifrit} as the mechanism for finding new preemption candidates.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\cbstart
+\subsection{Changing State Spaces}
+\label{sec:algorithm0}
+
+To introduce the Iterative Deepening algorithm,
+we first show a simple approach for handling new PPs in the absence of any CPU budget restriction.
+%show how we prioritize smaller state spaces based on ETAs.
+
+{\bf Na\"ive approach.}
+Given unlimited CPU time for testing,
+we would always switch to the new maximal state space whenever adding a new PP.
+The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
+so considering any incomplete subset of the known PPs will be duplicate work.
+Algorithm~\ref{alg:algorithm0} demonstrates this na\"ive approach.
+It would be seeded with the set of all statically-known synchronization API PPs,
+and invoked whenever a new data race PP is added.
+%
+Our proofs in \sect{\ref{sec:totalverif}},
+being concerned with the verification guarantee provided when \quicksand~completes within the CPU budget,
+are based on this simple version of Iterative Deepening.
+%
+
+\newcommand\AllPPs{\ensuremath{\mathcal{A}}}
+\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
+\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
+\newcommand\GetETA[1]{ETA(#1)}
+\newcommand\GetPPSet[1]{PPSet(#1)}
+
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
+	\Input{$j$, the currently-running job}
+	%\Input{$eta$, $j_0$'s predicted completion time}
+	\Input{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
+	%\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
+	%\Input{$T$, the remaining time in the CPU budget}
+	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
+		return NewJob(\AllPPs) // New maximal state space
+	}{
+		return $j$ // $j$ is still maximal
+	}
+	\caption{Na\"ive Iterative Deepening method}
+	\label{alg:algorithm0}
+\end{algorithm}
+
+{\bf Prioritizing smaller jobs.}
+However, in tests where full verification is not feasible,
+focusing on the maximal state space alone is likely to be fruitless.
+%
+Hence, we prioritize all subset jobs
+\cbend
+based on number of PPs, ETA, and whether they include data-race PPs,
 We rely on state-space estimation \cite{estimation}
@@
 as each one serves as a guarantee that all interleavings possible with its PPs were tested.
-
-Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
-A MC tool is still used to test each state space, and other reduction techniques are still applicable.
-Moreover, because Iterative Deepening treats the set of preemption points as mutable,
-it can add new preemption points reactively based on any runtime analysis.
-We focus on run-time data-race detection~\cite{hybriddatarace,tsan,ifrit} as the mechanism for finding new preemption candidates.
+\cbstart
+The next three subsections will show how we schedule these smaller jobs
+based on their PP sets and ETAs.
+\cbend
 
@@
 
-\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
-\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
-\newcommand\GetETA[1]{ETA(#1)}
-\newcommand\GetPPSet[1]{PPSet(#1)}
 With a limited CPU budget, we must avoid running tests that are likely to be fruitless.

### 2016-07-14 oopsla-16/design.tex
reverse order of data-race-pps and choosing-best-job subsecs
forgot stupid changebar package breaks 2col tables

@@
 
-\cbstart
+\revisionx{
 \subsection{Changing State Spaces}
@@
 %
-Hence, we prioritize all subset jobs
-\cbend
+Hence, we prioritize all subset jobs}
 based on number of PPs, ETA, and whether they include data-race PPs,
@@
 as each one serves as a guarantee that all interleavings possible with its PPs were tested.
-\cbstart
+\revisionx{
 The next three subsections will show how we schedule these smaller jobs
 based on their PP sets and ETAs.
-\cbend
+}
 
@@
 which could potentially be improved in future work (\sect{\ref{sec:future}}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Choosing the best job}
-
-With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
-Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
-and a set of {\em pending} jobs (untested ones with unknown ETAs).
-When the MC reports an ETA too high for some job,
-we compare with other pending and suspended jobs to find another one more likely to complete in time.
-%
-Our method,
-%for doing so,
-listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
-%\footnote{
-%Though its worst-case performance is $O(mn)$ in the
-%%number of pending and suspended jobs,
-%sizes of $\mathcal{P}$ and $\mathcal{S}$,
-%in practice the non-constant portion beyond line 4 runs very infrequently
-%and is negligible compared to the exponentially-sized state spaces.}.
-Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
-and $j_1$ is suspended,
-then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
-Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
-%reveals that it might finish in time after all.
-Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.
-
-\begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
-	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
-	\Input{$j_0$, the currently-running job}
-	%\Input{$eta$, $j_0$'s predicted completion time}
-	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
-	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
-	\Input{$T$, the remaining time in the CPU budget}
-	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
-		return $j_0$ // Common case: job is expected to finish.
-	}
-	\ForEach{job $j_P \in$ \PendingJobs}{
-		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
-		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
-			return $j_P$
-		}
-	}
-	%// no pending jobs; maybe resume a suspended job \\
-	\ForEach{job $j_S \in$ \SuspendedJobs}{
-		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
-			$\land$
-			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
-			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
-			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
-				return $j_S$
-			}
-		}
-	}
-	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
-	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
-	\label{alg:shouldworkblock}
-\end{algorithm}
-
-%
-We also account for the inherent inaccuracy of ETA estimates.
-Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
-in case their ETAs are actually overestimated.
-Lines 12-14 account for the
-%bizarre
-possibility that among two suspended jobs,
-%given two jobs,
-%%$j_1,j_2$,
-\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
-but
-\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
-This can arise because estimates tend to get more accurate over time,
-and $j_1$ perhaps ran much longer before suspending.
-% In such scenarios,
-We heuristically assume the smaller job's ETA is more accurate
-to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
-(it lets us avoid thrashing in \quicksand).
 
@@
 
+\subsection{Choosing the best job}
+
+With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
+Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
+and a set of {\em pending} jobs (untested ones with unknown ETAs).
+When the MC reports an ETA too high for some job,
+we compare with other pending and suspended jobs to find another one more likely to complete in time.
+%
+Our method,
+%for doing so,
+listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
+%\footnote{
+%Though its worst-case performance is $O(mn)$ in the
+%%number of pending and suspended jobs,
+%sizes of $\mathcal{P}$ and $\mathcal{S}$,
+%in practice the non-constant portion beyond line 4 runs very infrequently
+%and is negligible compared to the exponentially-sized state spaces.}.
+Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+and $j_1$ is suspended,
+then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
+Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
+%reveals that it might finish in time after all.
+Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.
+
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
+	\Input{$j_0$, the currently-running job}
+	%\Input{$eta$, $j_0$'s predicted completion time}
+	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
+	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
+	\Input{$T$, the remaining time in the CPU budget}
+	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
+		return $j_0$ // Common case: job is expected to finish.
+	}
+	\ForEach{job $j_P \in$ \PendingJobs}{
+		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
+		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
+			return $j_P$
+		}
+	}
+	%// no pending jobs; maybe resume a suspended job \\
+	\ForEach{job $j_S \in$ \SuspendedJobs}{
+		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
+			$\land$
+			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
+			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
+			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
+				return $j_S$
+			}
+		}
+	}
+	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
+	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
+	\label{alg:shouldworkblock}
+\end{algorithm}
+
+%
+We also account for the inherent inaccuracy of ETA estimates.
+Line 1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
+in case their ETAs are actually overestimated.
+Lines 12-14 account for the
+%bizarre
+possibility that among two suspended jobs,
+%given two jobs,
+%%$j_1,j_2$,
+\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
+but
+\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
+This can arise because estimates tend to get more accurate over time,
+and $j_1$ perhaps ran much longer before suspending.
+% In such scenarios,
+We heuristically assume the smaller job's ETA is more accurate
+to avoid repeatedly resuming larger jobs briefly while their ETAs only become worse
+(it lets us avoid thrashing in \quicksand).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Heuristics}

### 2016-07-15 oopsla-16/implementation.tex
Improve the prose.
mando revisions regarding chess, fasttrack, djit, and seqcst

@@
 \landslide~implements DPOR \cite{dpor},
-state space estimation \cite{estimation}, and a hybrid
+state space estimation \cite{estimation},
+\revisionx{the DJIT+ HB data-race analysis \cite{djit,fasttrack},}
+and a hybrid
 lockset/Limited HB
@@
 We suspend jobs simply by making the MC wait on a message-passing reply.
-Should \quicksand~later re-schedule a suspended job, we send a message to continue,
-resuming it where it left off.
+Should \quicksand~later re-schedule a suspended job, it sends a message to continue,
+resuming the job where it left off.
 %otherwise, we resume it only after time runs out, causing it to exit.

### 2016-07-15 oopsla-16/intro.tex
mando revisions regarding chess, fasttrack, djit, and seqcst

@@
 However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
-For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
+For example, \textsc{CHESS} \cite{chess} \revisionx{by default} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
 %
-On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
-are able to preempt threads before any shared memory access.
+On the other hand, SPIN \cite{spin} %and Inspect \cite{inspect}
+%are able to preempt
+preempts threads before any shared memory access.
 Such fine granularity would automatically check each data race for the possibility of failure, but risks timing out before the state space can be completed.
 %makes full state space completion intractable for even modestly-sized tests.
-Choosing preemption points is a tradeoff between schedule coverage and feasibility of completion,
-but even with state-of-the-art reduction techniques,
+\revisionx{Some tools, such as CHESS and Inspect \cite{inspect} can strike a middle ground using compiler instrumentation to statically add preemption points on memory accesses.
+Nevertheless,}
+choosing preemption points is a tradeoff between schedule coverage and feasibility of completion:
+%but
+even with state-of-the-art reduction techniques,
 \revision{fixing the degree of coverage}~in advance necessarily leaves some tests unaffordably large \cite{parrot,mcr}.
@@
 	\item {\em Iterative Deepening}, a new \revision{algorithm}~for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
-	\item A proof of convergence, showing that should it be possible in the given CPU budget,
+	\item A proof of convergence \revisionx{for sequentially-consistent memory models}, showing that should it be possible in the given CPU budget,
 		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;

### 2016-07-15 oopsla-16/overview.tex
Last of garth's commence
mando revisions regarding chess, fasttrack, djit, and seqcst

@@
 \subsection{Stateless Model Checking}
+\label{sec:overview-sssmc}
 
@@
 
-\revision{
 	{\bf Variants of Happens-Before.}
-	Most prior work focuses on {\em Happens-Before} (HB) \cite{lamport-clocks} as the order relation between accesses.
+	Most prior work focuses on {\em Happens-Before} (HB) \cite{lamport-clocks,djit,fasttrack} as the order relation between accesses.
 \cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
@@
 %% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
-Hence, we employ Limited HB in this paper},
-accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
+%Hence, we employ Limited HB in this paper},
+%accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
+\revisionx{Hence, we can accept some overhead from Limited HB's false positives for the sake of finding data-race candidates more quickly.
+In \sect{\ref{sec:eval}} we will evaluate how HB and Limited HB each influence \quicksand's bug-finding and verification speed.
+}
 

### 2016-07-15 oopsla-16/related.tex
mando revisions regarding chess, fasttrack, djit, and seqcst

@@
 This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
-Nevertheless, Iterative Deepening is applicable to these tools.
+%Nevertheless, Iterative Deepening is applicable to these tools.
 
@@
 
+\revisionx{{\bf Happens-Before.}}
 %Too many related projects to list have made contributions to the
@@
 \revision{introduce Limited Happens-Before}, %(\sect{\ref{sec:overview-dr}})},
-which we employ,
-\revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.
+which
+\revisionx{\quicksand~uses to achieve its best bug-finding result.
+Other tools such as DJIT+ \cite{djit} and FastTrack \cite{fasttrack} opt for the precise Happens-Before analysis,
+which produces no false positives but is prone to more false negatives as shown in \sect{\ref{sec:overview-dr}}.
+FastTrack, the most recent development in this approach,
+optimizes the representation of variable write vector clocks for performance,
+which guarantees to detect at least the first race on each variable.
+However, as we are interested in classifying data race candidates as {\em benign} or {\em buggy},
+this optimization would be unsound for total verification if the first race on a variable were benign.
+Hence, we implement precise HB as defined by DJIT+.
+Finally, \cite{predictive-dr} recently introduced the Causally-Precedes relation for a precise analysis which covers some common false negative cases
+such as our example in Figure~\ref{fig:hb-example}(a).
+}
+%\revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.
 
+\revisionx{{\bf Other domains.}}
 \revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
@@
 %although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
-% No, IDGAF about pure happens before.
 %FastTrack \cite{fasttrack} optimizes the performance of pure happens-
 
+\revisionx{{\bf Replay analysis.}}
 Closer to our work, replay analysis \cite{recordreplaydrs} also suppresses false positives by testing multiple thread interleavings.

### 2016-07-15 oopsla-16/soundness.tex
Improve the prose.
mando revisions regarding chess, fasttrack, djit, and seqcst

@@
 discuss our assumptions explicitly and include more formal definitions and structure.
+
+\revisionx{These proofs are built on a DPOR algorithm definition which assumes sequentially-consistent memory hardware.
+%
+% TODO: Aaargh, can't think of how to actually justify this.
+%We believe this assumption is reasonable for the scope of this paper because the main property we require of DPOR is that 
+A recent paper extended DPOR to support weaker memory models \cite{tsopso}, so our proofs could likewise be extended with that version of DPOR in future work.
+We also assume the Limited HB definition for the data-race analysis, leaving the case for precise HB to future work.
+}
 
@@
 %our previous proof's
-Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over Pure HB.
+Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over full HB.
 % TODO CAMREADY: Cite the tech report proofs, which I think call out the use of limited HB in the proof more explicitly.

### 2016-07-15 oopsla-16/design.tex
minor touch up

@@
 The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
-so considering any incomplete subset of the known PPs will be duplicate work.
+so considering any incomplete subset of the known PPs would be duplicate work.
 Algorithm~\ref{alg:algorithm0} demonstrates this na\"ive approach.

### 2016-07-15 oopsla-16/overview.tex
address the rest of reviewer D non-mandos

@@
 We call such unreported access pairs {\em false negatives}.
+However, consider the similar program in Figure~\ref{fig:hb-example}(b),
+in which the access pair ceases to exist in the alternate interleaving.
 \cite{hybriddatarace} introduces the {\em Limited HB} relation,
@@
 by considering only blocking operations like {\tt cond\_wait} to enforce the order.
-However, consider the similar program in Figure~\ref{fig:hb-example}(b),
-in which the access pair ceases to exist in the alternate interleaving.
-Limited HB will report all potential races, avoiding false negatives \cite{tsan},
+Limited HB will report all potential races, avoiding \revisionx{many} false negatives \cite{tsan},
 but at the cost of necessarily reporting some such {\em false positives}.
@@
 
-Stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx}.
+\revisionx{Being dynamic analyses, both HB and Limited HB may suffer false negatives when a racy access pair is not executed at all in a specific interleaving.
+Limited HB offers the advantage of identifying a potential race as long as the access pair is observed under any interleaving,
+rather than requiring the accesses to be adjacent in time, as HB would.}
+While stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
 %However,
-This work incorporates data-race analysis in an internal feedback loop,
+we incorporate data-race analysis in an internal feedback loop,
 using model checking to automatically test each potential race

### 2016-07-15 oopsla-16/related.tex
minor relwork touchup

@@
 \revisionx{\quicksand~uses to achieve its best bug-finding result.
-Other tools such as DJIT+ \cite{djit} and FastTrack \cite{fasttrack} opt for the precise Happens-Before analysis,
+Other tools such as DJIT+ \cite{djit} and FastTrack \cite{fasttrack} opt for the precise Happens-Before relation first defined by Lamport \cite{lamport-clocks},
 which produces no false positives but is prone to more false negatives as shown in \sect{\ref{sec:overview-dr}}.
-FastTrack, the most recent development in this approach,
+FastTrack
+%, the most recent development in this approach,
 optimizes the representation of variable write vector clocks for performance,
@@
 this optimization would be unsound for total verification if the first race on a variable were benign.
-Hence, we implement precise HB as defined by DJIT+.
+Hence, we implement precise HB with write vector clocks as defined by DJIT+.
 Finally, \cite{predictive-dr} recently introduced the Causally-Precedes relation for a precise analysis which covers some common false negative cases

### 2016-07-16 oopsla-16/related.tex
finish related work related revisions

@@
 is a randomized algorithm that can quantify the probability of uncovering bugs.
-PCT is optimized for tests with impossibly large state spaces,
-while for more approachable tests,
-its incompatibility with DPOR would significantly delay completion.
+PCT \revisionx{targets} tests with impossibly large state spaces,
+\revisionx{eschewing DPOR's depth-first search model to
+%ensure it can
+quickly sample a broad cross-section of large state spaces.
+However, it proposes no alternate reduction algorithm to make up for its incompatibility with DPOR,
+so is unsuitable for verification of medium-sized tests.}
+%while for more approachable tests,
+%its incompatibility with DPOR would significantly delay completion.
 Future work could use ETAs to heuristically switch between DPOR and PCT}.
+%
 Finally, ICB \cite{chess-icb} is most similar to our work,
@@
 when full completion is intractable (\sect{\ref{sec:future}}).
-However, ICB is limited to a fixed set of PPs,
+However, ICB
+%is limited to a fixed set of PPs,
 \revision{cannot estimate remaining time to total verification,
 and can incorporate data-race PPs only when statically coded in advance}.
+\revisionx{Our results in \sect{\ref{sec:eval-sssmc}} outperform both such configurations of ICB.}
 %and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

### 2016-07-16 oopsla-16/evaluation.tex
show icb bound distribution stats

@@
 
+\revisionx{
+To ensure that our corpus of P2 and Pintos bugs gives an unbiased comparison between \quicksand~and ICB,
+we also counted the preemption bounds necessary for ICB to find each of its bugs.
+Table~\ref{tab:icb-bounds} shows the distribution of these bounds,
+which is consistent with the results of \cite{chess-icb}.
+%justifies that our portrayal of ICB is fair.
+%and shows no bias towards requiring high preemption bounds.
+}
+
+\begin{table}[h]
+	\begin{center}
+		\revisionx{
+		\small
+	\begin{tabular}{r|c|c}
+		{\bf Bound} & {\bf SSS-MC-ICB} & {\bf SSS-MC-Shared-Mem} \\
+		\hline
+		0	& 2	& 1	\\
+		1	& 82	& 86	\\
+		2	& 15	& 32	\\
+		3	& 2	& 3	\\
+		4+	& 0	& 0	\\
+	\end{tabular}
+		}
+	\end{center}
+	\label{tab:icb-bounds}
+	\caption{\revisionx{Distribution of preemption bounds among bugs found by ICB \cite{chess-icb} control experiments. (Bound 0 means the bug was found by switching threads only on {\tt yield()}s.)}}
+\end{table}
+
 {\bf Partial verification.}

### 2016-07-16 oopsla-16/warpzone.tex
Last of garth's commence
bit extra future work

@@
 to prune some or all such repeated work.
+\revisionx{Prioritizing the maximal state space in particular could also improve completion times:
+whenever the maximal job finishes with no new data-races, future implementations could immediately prune all subset jobs and declare a total verification.}
 
@@
 by configuring a static predicate to preempt on any non-stack memory access.
-This introduced up to thousands of new PPs on each new test execution, with a prohibitive performance impact.
+This introduced hundreds of new PPs on each new test execution, with a prohibitive performance impact.
 While this performance could be improved by

### 2016-07-16 oopsla-16/evaluation.tex
fix some camready todo layout issues now that we are swimming in free space

@@
 	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
-		% TODO CAMREADY "as a function of elapsed"
-		(a) Bugs found by elapsed CPU time. Overall,
-		% TODO CAMREADY "a more resource-fair"
-		a more direct comparison than (b),
+		(a) Bugs found as a function of elapsed CPU time. Overall,
+		a more resource-fair comparison than (b),
 		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\

### 2016-07-19 oopsla-16/paper.tex
refactor design section intro and write algorithm0
forgot stupid changebar package breaks 2col tables
deanonymize

@@
 \documentclass[10pt,times,numbers]{sigplanconf}
-\usepackage[1stsubmission]{oopsla2016}
+\usepackage[2ndsubmission]{oopsla2016}
 
@@
 \usepackage{graphicx}
+\usepackage{xspace}
 \usepackage{amsthm}
@@
 
-%\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
-%\newcommand\revision[1]{\cbstart#1\cbend}
 \newcommand\revision[1]{#1}
 \definecolor{purple}{RGB}{192,0,192}
+\newcommand\revisionx[1]{\hilight{purple}{#1}\xspace}
 %\newcommand\revision[1]{\hilight{purple}{#1}}
@@
 
+\url{https://github.com/bblum/oopsla-quicksand}
 % TODO CAMREADY: Make data set, at least, public.
+\url{https://github.com/bblum/oopsla-dataset}
 
-{\em github links removed for double-blind review}
-%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put QS in its own repository
-%
-%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put final logs somewhere
 % TODO CAMREADY: Don't forget to anonymize the names of all your grupos.
@@
 %\section{Acknowledgements}
-
-% TODO CAMREADY
-%{\em removed for double-blind review}
-%Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of chicago's, and CMU's OS classes respectively.
-%Thanks to Wind River for the use of their simulator \simics.
-%Thanks to
-%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
-%the anonymous PLDI reviewers for their helpful comments.
-%This work was supported in part by
-%the U.S. Army Research Office under grant number W911NF0910273.
 
@@
 
-%Acknowledgments, if needed.
-{\em omitted for double-blind review}
+Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of Chicago's, and CMU's OS classes respectively.
+Thanks to Wind River for the use of their simulator \simics.
+Thanks to
+Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
+the anonymous OOPSLA reviewers for their helpful comments.
+This work was supported in part by
+the U.S. Army Research Office under grant number W911NF0910273.
 

### 2016-07-19 oopsla-16/evaluation.tex
Incorporate pure HB graphs and discussion in evaluation section; tons of prose changes omg

@@
 			}
+			\revisionx{
+			\item How does the choice between HB and Limited-HB affect bug-finding and verification performance?
+			}
 		\end{enumerate}
@@
 		\begin{enumerate}
-			\item Does \quicksand~avoid false positives compared to a single-execution data race analysis?
+		\item Does \quicksand~avoid false positives compared to a single-execution \revisionx{Limited-HB} data race analysis?
 				% Explain later as:
@@
 
-To evaluate the benefits of data-race PPs and Iterative Deepening separately, we ran the test suite under \quicksand~in two different configurations, each of which was given a 1-hour budget and 10 CPUs for each test.
+To evaluate the benefits of data-race PPs and Iterative Deepening separately, we ran the test suite under \quicksand~in \revisionx{three} different configurations, each of which was given a 1-hour budget and 10 CPUs for each test.
 \begin{itemize}
-	\item {\bf QS-DR}: \quicksand~with data-race PPs enabled, and
+		\revisionx{
+	\item {\bf QS-Limited-HB}: \quicksand~using Limited-HB to add data-race PPs,
+	\item {\bf QS-Pure-HB}: \quicksand~using pure HB instead, and
+		}
 	\item {\bf QS-Sync-Only}: \quicksand~with PPs still seeded as described in \sect{\ref{sec:initial-pp}}, but never adding new PPs from reported data-race candidates.
@@
 	\begin{tabular}{p{0.48\textwidth}}
-	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-squashed.pdf} \\
-		(a) Bugs found as a function of elapsed CPU time. Overall,
-		a more resource-fair comparison than (b),
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-v2.pdf} \\
+		(a) Bugs found \revisionx{as a function of} elapsed CPU time. Overall,
+		a more \revisionx{resource-fair} comparison than (b),
 		although \quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
 		\\
-	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-wallclock.pdf} \\
+	\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-wallclock-v2.pdf} \\
 		(b) Bugs found by elapsed wall-clock time.
@@
 	by several configurations of \quicksand~and the SSS-MC control.
-	\quicksand~finds \revision{151}\% as many bugs with data-race PPs.}
+	\quicksand~finds \revisionx{125}\% as many bugs with data-race PPs.}
 	\label{fig:dowefindbugsfaster}
@@
 %The break-even point is at a negligible 90 seconds.
-\revision{Compared to the best SSS-MC series, \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds,
-and ultimately concludes 10 CPU-hours with \revision{151}\% as many bugs in total.
+\revision{Compared to \revisionx{SSS-MC-ICB (the fastest among the control experiments),} \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds.
+\revisionx{Compared to SSS-MC-Shared-Mem (the best SSS-MC approach in the long term),} we ultimately concludes 10 CPU-hours with \revisionx{125\%} as many bugs in total \revisionx{when using Limited HB}.
 \revision{Before the break-even point at 200 seconds,
-\quicksand~lags behind the SSS-MC approaches due to additional start-up overhead from its tenfold parallelism.
+\quicksand~lags behind \revisionx{SSS-MC-ICB} due to additional start-up overhead from its tenfold parallelism.
 However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
 Figure~\ref{fig:dowefindbugsfaster}(b) gives \quicksand~full credit for its inherent parallelism:
-with a 10 core allocation, it outperforms SSS-MC for any fixed budget of wall-clock time.}
+with a 10 core allocation, it outperforms SSS-MC
+for any fixed budget of wall-clock time.}
 
@@
 %Because \quicksand~can be nondeterministic in its scheduling of state spaces,
-%we ran the QS-DR tests a second time to quantify the variance.
+%we ran the QS-Limited-HB and QS-Pure-HB tests a second time to quantify the variance.
 %% TODO CAMREADY
-%% We found the maximum area between any two QS-DR curves to be
-%We found the area between the two QS-DR curves to be
+%% We found the maximum area between any two curves to be
+%We found the area between the two curves to be
 %% TODO CAMREADY
 %{\bf XX\%}
-%of the area between QS-DR and SSS-MC-ICB (the next closest),
+%of the area between QS-TODO(?)-HB and SSS-MC-Shared-Mem (the next closest),
 %% TODO cite CAMREADY
@@
 % TODO CAMREADY: Fix, SSSMC-ShM finds 2 mx bugs.
-the control experiments
+\revisionx{SSS-MC-ICB and SSS-MC-DPOR}
 found dramatically fewer bugs (just 1)\footnote{
@@
 %These results show explicitly that for arbitrary CPU-time budgets
-%\footnote{The initial perfect overlap between QS-DRs and QS-no-DR-bugs indicates how long it takes before the first data-race bug is found.}
+%\footnote{The initial perfect overlap between QS-*-HB and QS-Sync-Only indicates how long it takes before the first data-race bug is found.}
 %even after the extra overhead of verifying them, \quicksand~still slightly edges out SSS-MC
@@
 On the other hand,
-comparing QS-DR to SSS-MC-Shared-Mem shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
+comparing \revisionx{QS-Limited-HB} to SSS-MC-Shared-Mem shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
 %We attribute this to the fact that
@@
 		4+	& 0	& 0	\\
+		\hline
+		Total	& 101	& 122	\\
 	\end{tabular}
@@
 %but can at least report
-% TODO: Fix to be ICB not DPOR.
-On 129 tests, SSS-MC-DPOR timed out after 10 hours with no bugs found.
-Among these tests, \quicksand~found bugs in 36.
-For the other 93, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
+On \revisionx{310} tests, \revisionx{SSS-MC-ICB} timed out after 10 hours with no bugs found.
+Among these tests, \quicksand~found bugs in \revisionx{47}.
+For the other \revisionx{263}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
 %and the number of unique PPs among them.
@@
 {\bf Full verification.}
-For 153 of our 629 tests, \quicksand~was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}}.
+For 153 of our 629 tests, \revisionx{QS-Limited-HB} was able to provide the total verification guarantee described in \sect{\ref{sec:totalverif}},
+\revisionx{and QS-Pure-HB completed a verification for 167 tests.}
 In Figure~\ref{fig:totalverif} we plot the cumulative distribution of
@@
 verifications provided by each approach.
-Though SSS-MC-Shared-Mem appears to outperform \quicksand, its safety guarantee is unsound,
-as many tests break its assumption of stack-allocated memory being private.
-For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
-While SSS-MC-Shared-Mem must avoid excessive PPs on every private stack access for performance,
-\quicksand~detects these dynamically, without needing to discriminate between stack and heap.
-In total, SSS-MC-Shared-Mem claimed to verify 21 tests
-(across 7 of the 9 test cases)
-in which QS-DR ultimately found bugs,
-which further calls into question the rest of its completions.
+\revisionx{The next best approach for verifications was SSS-MC-Shared-Mem, which completed its search in only 39 tests.}
+
+%Though SSS-MC-Shared-Mem appears to outperform \quicksand, its safety guarantee is unsound,
+%as many tests break its assumption of stack-allocated memory being private.
+%For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
+%While SSS-MC-Shared-Mem must avoid excessive PPs on every private stack access for performance,
+%\quicksand~detects these dynamically, without needing to discriminate between stack and heap.
+%In total, SSS-MC-Shared-Mem claimed to verify 21 tests
+%(across 7 of the 9 test cases)
+%in which QS-DR ultimately found bugs,
+%which further calls into question the rest of its completions.
 
@@
 
-Among QS-DR's verified tests, 36 contained no data-race candidates whatsoever,
+% TODO: Update for pure HB (need a re-run of a subset of completed sssmc tests to confirm they still have no pure hb drs, even if they had limited hb drs)
+Among QS-Limited-HB's verified tests, 36 contained no data-race candidates whatsoever,
 so the same verification could be provided
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{totalverifs-squashed-with-unsound-shm.pdf}
+	\includegraphics[width=0.48\textwidth]{totalverifs-v2.pdf}
 	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
@@
 \end{figure}
+
+\revisionx{
+{\bf Happens-Before comparison.}
+Ultimately, using Limited HB for finding data-race candidates allowed \quicksand~to find more bugs, while
+the ``pure'' Happens-Before analysis improved \quicksand's performance on verifications.
+This trade-off is attributable to the fact that QS-Limited-HB need not wait to test many alternate thread interleavings before a potential data-race candidate is confirmed; rather, it can add new jobs to start testing potential races immediately.
+On the other hand, QS-Limited-HB can get overwhelmed by too many false positives, needing to refute such candidates by testing new state spaces,
+while QS-Pure-HB can refute false positives {\em en passant} by testing alternate interleavings in its original state spaces.
+%In both cases, the performance differences seem to become significant only after about 10 minutes of CPU time,
+%which suggests % TODO
+}
 

### 2016-07-19 oopsla-16/evaluation.tex
testing philosophy

@@
 % TODO: Update for pure HB (need a re-run of a subset of completed sssmc tests to confirm they still have no pure hb drs, even if they had limited hb drs)
-Among QS-Limited-HB's verified tests, 36 contained no data-race candidates whatsoever,
+Among \revisionx{QS-Limited-HB's} verified tests, 36 contained no data-race candidates whatsoever,
 so the same verification could be provided
@@
 among these, the otherwise more antiquated SSS-MC-DPOR performs best,
-while the other two lag behind
-}
+while the other two lag behind}
 %\quicksand~is slower than SSS-MC-DPOR
@@
 %which suggests % TODO
+This suggests that
+%each approach has merit, and that
+MCs which feature data-race analysis should implement both modes and offer the user to choose based on their testing philosophy.
 }

### 2016-07-19 oopsla-16/evaluation.tex
buncha minor prose tweaks

@@
 \revision{Compared to \revisionx{SSS-MC-ICB (the fastest among the control experiments),} \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds.
-\revisionx{Compared to SSS-MC-Shared-Mem (the best SSS-MC approach in the long term),} we ultimately concludes 10 CPU-hours with \revisionx{125\%} as many bugs in total \revisionx{when using Limited HB}.
+\revisionx{Compared to SSS-MC-Shared-Mem (the best SSS-MC approach in the long term), \quicksand's Limited HB version} ultimately concludes 10 CPU-hours with \revisionx{125\%} as many bugs in total.
 \revision{Before the break-even point at 200 seconds,
@@
 for any fixed budget of wall-clock time.}
+\revisionx{After 1 hour of wall-clock time, tenfold \quicksand~performs 158\% as well as SSS-MC-ICB.}
 
@@
 Table~\ref{tab:icb-bounds} shows the distribution of these bounds,
-which is consistent with the results of \cite{chess-icb}.
+which is consistent with the results of \cite{chess-icb}, showing no bias towards bugs that would be harder for ICB to find.
 %justifies that our portrayal of ICB is fair.
@@
 	\end{center}
+	\caption{\revisionx{Distribution of preemption bounds among bugs found by ICB control experiments. (Bound 0 means the bug was found by switching threads only on {\tt yield()}s.)}}
 	\label{tab:icb-bounds}
-	\caption{\revisionx{Distribution of preemption bounds among bugs found by ICB \cite{chess-icb} control experiments. (Bound 0 means the bug was found by switching threads only on {\tt yield()}s.)}}
 \end{table}
@@
 	\includegraphics[width=0.48\textwidth]{totalverifs-v2.pdf}
-	\caption{Cumulative distribution of tests \quicksand~fully verified (\sect{\ref{sec:totalverif}}).
+	\caption{Cumulative distribution of tests fully verified
+	\revisionx{by QS-Limited-HB, QS-Pure-HB, and SSS-MC-Shared-Mem}
+	(\sect{\ref{sec:totalverif}}).
 	%SSS-MC-Shared-Mem completed 1.37x as many tests, but 
-	SSS-MC-Shared-Mem is unsound: 21 tests it completed hid bugs previously found in Figure~\ref{fig:dowefindbugsfaster}.
+	%SSS-MC-Shared-Mem is unsound: 21 tests it completed hid bugs previously found in Figure~\ref{fig:dowefindbugsfaster}.
 	%previously found (Figure~\ref{fig:dowefindbugsfaster}).

### 2016-07-19 oopsla-16/related.tex
small parallel structure fix
buncha minor prose tweaks

@@
 %ensure it can
-quickly sample a broad cross-section of large state spaces.
+instead sample broad cross-sections of large state spaces.
 However, it proposes no alternate reduction algorithm to make up for its incompatibility with DPOR,
@@
 %, the most recent development in this approach,
-optimizes the representation of variable write vector clocks for performance,
+optimizes the representation of variable write clocks for performance,
 which guarantees to detect at least the first race on each variable.

### 2016-07-19 oopsla-16/soundness.tex
address the rest of reviewer D non-mandos
minor touch ups
buncha minor prose tweaks

@@
 
-We identify a particular class of false positive data-race candidates in which the associated memory was recycled by re-allocation between the two accesses.
+We identify a particular class of false positive data-race candidates \revisionx{under Limited HB} in which the associated memory was recycled by re-allocation between the two accesses.
 Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
@@
 Here, reordering lines 6-7 before line 3 will allow {\tt x} and {\tt x2} to race.
+\revisionx{Such collisions could be avoided with a hacked allocator which never recycles memory, but this could unacceptably impact performance in {\tt malloc}-heavy tests.}
 

### 2016-07-19 oopsla-16/evaluation.tex
split the giant table by making an auxiliary table for dr statistics; add purehb/shmem to table; add discussion of pure hb

@@
 \subsection{Experimental Setup}
+\label{sec:eval-setup}
 
@@
 		\small
-	\begin{tabular}{r|c||c|c|c|c|c|c||c|c||c|c}
-		& {\bf Num.} & {\bf DPOR} & {\bf ICB} & {\bf QS} & {\bf DR} & {\bf Nondet.} & {\bf Malloc} & {\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
-		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf DRs} & {\bf timeouts} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
+		\revisionx{
+	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+		& {\bf Num.} & {\bf QS-LHB} & {\bf QS-PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem} & {\bf LHB DR} & {\bf Nondet.} & {\bf Mutual} & {\bf Avg. tested} \\
+		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf timeouts} & {\bf subset SSes} \\
 		\hline
-		% FIXME CAMREADY: update FRMs
-					% TODO: update to be ShMem not DPOR
-											% TODO: recount with ICB
-				% #test  sssmc   icb      qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
-			% TODO: If possible, add a column for how many false negatives (incorrect verif claims) ShMem had.
-		{\tt bcast}	& 79	& 7	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
-		{\tt join} 	& 79	& 14 	& 14	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
-		{\tt mx} 	& 79	& 1	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
-		{\tt sem} 	& 79	& 10	& 11	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
-		{\tt signal} 	& 79	& 4	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
-		{\tt rwlock} 	& 79	& 26	& 27	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
+				% XXX FIXME: Need rerun nondet dr expt!
+				% #test  qs-lhb  qs-phb   icb     dpor    everyw  dronly  nondets timeout comp.SSes
+		{\tt bcast}	& 79	& 8	& 8	& 5	& 6	& 7	& 2	& 4	& 7	& 112.3	\\
+		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& 11	& 75.9	\\
+		{\tt mx} 	& 79	& 10	& 9	& 1	& 1	& 12	& 9	& 1	& 0	& -	\\
+		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& 36	& 93.4	\\
+		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& 59.0	\\
+		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& 43	& 88.3	\\
 		\hline
-		{\tt sched} 	& 59	& 1	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
-		{\tt alarm} 	& 44	& 5	& 16	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
-		{\tt wait} 	& 52	& 23	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
+		{\tt sched} 	& 59	& 7	& 7	& 1	& 1	& 8	& 6	& 4	& 2	& 13.0	\\
+		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& 16	& 8.1	\\
+		{\tt wait} 	& 52	& 30	& 27	& 23	& 23	& 1	& 7	& 1	& 13	& 32.8	\\
 		\hline
-		{\bf Total}	& 629	& 91	& 104	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
+		{\bf Total}	& 629	& 153	& 133	& 101	& 88	& 122	& 69	& 32	& 173	& 69.5 	\\
 	\end{tabular}
+		}
 	\end{center}
-	\caption{Summary of bugs and data races found by each test program.
-		% TODO CAMREADY: Add Shared-Mem
-		DPOR/ICB are the controls; QS is \quicksand.
-		``DR bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+	\caption{Summary of bugs
+	%and data races
+	found by each test program.
+		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
+		``LHB DR bugs'' counts among \revisionx{QS-Limited-HB's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
 	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
-	``Malloc DRs'' counts how many malloc-recycle false positives we suppressed (\sect{\ref{sec:eval-dr}}).
-	% TODO CAMREADY: count timeouts for ICB instead?
-	``Mutual timeouts'' counts how often both SSS-MC-DPOR and \quicksand~timed out with no bug found;
+	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
 	% TODO CAMREADY: update avg tested subsets to be from the ICB timeouts
 	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
-	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
-	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
 		}
@@
 
-\begin{table}[h]
+\begin{table}[t]
 	\begin{center}
@@
 using data-race PPs increases our \revision{verification capacity}~by 4.25x.
+\revisionx{Finally, assuming sequentially-consistent hardware,
+QS-Pure-HB classified many % TODO: how many
+true data races as benign, %(non-failing),
+while the SSS-MC-ICB approach could at best report such races to the user.
+We count these cases in Table~\ref{tab:drstatistix}.
+}
 
@@
 the ``pure'' Happens-Before analysis improved \quicksand's performance on verifications.
-This trade-off is attributable to the fact that QS-Limited-HB need not wait to test many alternate thread interleavings before a potential data-race candidate is confirmed; rather, it can add new jobs to start testing potential races immediately.
+This trade-off is attributable to the fact that QS-Limited-HB need not wait to test many alternate thread interleavings before a potential data-race candidate is confirmed; rather, it can add new jobs to start testing potential races immediately\footnote{
+	\revisionx{Table~\ref{tab:drbugs} corroborates: the difference is most dramatic in {\tt alarm}, the test where \quicksand~struggled most to finish even small subset jobs.}
+	}.
 On the other hand, QS-Limited-HB can get overwhelmed by too many false positives, needing to refute such candidates by testing new state spaces,
@@
 %In both cases, the performance differences seem to become significant only after about 10 minutes of CPU time,
-%which suggests % TODO
+%which suggests... I don't actually know!
 This suggests that
@@
 }
+
+\begin{table}[t]
+	\begin{center}
+		\small
+	\begin{tabular}{r||c|c|c||c}
+		& {\bf Total} & {\bf \revisionx{Benign}} & {\bf Untested} & {\bf Malloc} \\
+		{\bf Test} & {\bf DR PPs} & {\bf \revisionx{DRs}} & {\bf DR PPs} & {\bf DRs} \\
+		\hline
+		% FIXME CAMREADY: update FRMs
+				% DRPPs   benign                  untested FRMs
+		% TODO: recount dr pps for pure hb
+		% TODO: count dr pps in totalverifs for benign
+		% TODO: recount untesteds for pure hb
+		{\tt bcast}	& 912	& \revisionx{1337}	& 107	& 51	\\
+		{\tt join} 	& 781	& \revisionx{1337}	& 292	& 333	\\
+		{\tt mx} 	& 829	& \revisionx{1337}	& 1	& 7	\\
+		{\tt sem} 	& 753	& \revisionx{1337}	& 279	& 140	\\
+		{\tt signal} 	& 1118	& \revisionx{1337}	& 391	& 180	\\
+		{\tt rwlock} 	& 915	& \revisionx{1337}	& 634	& 125	\\
+		\hline
+		{\tt sched} 	& 144	& \revisionx{1337}	& 3	& 0	\\
+		{\tt alarm} 	& 115	& \revisionx{1337}	& 89	& 35	\\
+		{\tt wait} 	& 142	& \revisionx{1337}	& 31	& 31	\\
+		\hline
+		{\bf Total}	& 5709	& \revisionx{1337}	& 1827	& 902	\\
+	\end{tabular}
+	\end{center}
+	\caption{
+		\revisionx{Additional data race statistics.}
+		% TODO: update this for pure-hb and add a column,
+		``Total DR PPs'' counts how many unique data-racing instructions \revisionx{QS-Pure-HB identified}.
+		Among those,
+		\revisionx{``Benign DRs'' counts how many we refuted as non-failing (\sect{\ref{sec:eval-sssmc}}), while}
+		``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
+		``Malloc DRs'' counts how many \revisionx{false positive PPs QS-Limited-HB} suppressed (\sect{\ref{sec:eval-dr}}).
+		}
+	\label{tab:drstatistix}
+\end{table}
 
@@
 In \sect{\ref{sec:recycle}} we showed the soundness of suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
-In Table~\ref{tab:drbugs}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
+In Table~\ref{tab:drstatistix}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
 In total, 902 data-races fit the malloc-recycle pattern across all tests,

### 2016-07-19 oopsla-16/evaluation.tex
cite simsa thesis

@@
 %Though for many applications of MC, assuming a correct lock implementation is sufficient,
-% TODO CAMREADY cite simsa thesis
-Though it often suffices to assume correctly-implemented locks,
+Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
 we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.

### 2016-07-19 oopsla-16/overview.tex
fix some camready todo layout issues now that we are swimming in free space
Philosophy of bugs.

@@
 \revision{\normalsize (a) True potential data race.}
-% TODO CAMREADY: put another line break here
+\\
 \\
@@
 
+\revisionx{
+{\bf Philosophy of bugs.}
+% two camps: all races are bugs; citations:
+% and, races must be classified; citations: replay analysis, portend,
+While there is a vast body of work on how to detect data races to begin with,
+judging data races once found is a matter of philosophical debate unto itself.
+Some recent tools classify races depending on how they impact program behaviour \cite{recordreplaydrs,portend}, overlooking {\em benign} races in search of more dangerous ones.
+%In particular,
+\cite{racerx} acknowledges that a program may have too many races for a user to worry about,
+so bug reports must be prioritized by severity of effects. %how likely they may cause failures or security vulnerabilities.
+%
+However, other prior work argues that data races are always bugs \cite{miscompile-benign,data-races-are-evil},
+largely due to the possibility of compiler or hardware reordering of racy accesses.
+%
+We take the former camp:
+we consider a data race a bug only when it results in a visible failure state (e.g., crash or deadlock).
+We bypass concerns of compiler reordering by checking programs at the executable level;
+for a study of hardware reordering in the context of DPOR, we refer the reader to \cite{tsopso}.
+}
+
 \subsection{Terminology}

### 2016-07-19 oopsla-16/evaluation.tex
a small touchup, and move the table to fix latex layout
sully revisions

@@
 		\revisionx{
-	\item {\bf QS-Limited-HB}: \quicksand~using Limited-HB to add data-race PPs,
+	\item {\bf QS-Limited-HB}: \quicksand~using Limited-HB for its data-race analysis,
 	\item {\bf QS-Pure-HB}: \quicksand~using pure HB instead, and
@@
 	by several configurations of \quicksand~and the SSS-MC control.
-	\quicksand~finds \revisionx{125}\% as many bugs with data-race PPs.}
+	\quicksand~finds \revisionx{125\%} as many bugs with data-race PPs.}
 	\label{fig:dowefindbugsfaster}
@@
 %}
-
-% TODO: Fix up this paragraph after preempt-everywhere results come in.
-The left half of Table~\ref{tab:drbugs}
-breaks down the number and types of bugs found by each test program.
-In \mxtest, in which we do not trust the lock implementation's correctness,
-% TODO CAMREADY: Fix, SSSMC-ShM finds 2 mx bugs.
-\revisionx{SSS-MC-ICB and SSS-MC-DPOR}
-found dramatically fewer bugs (just 1)\footnote{
-	The one bug SSS-MC found was in a fully-assembler lock implementation.
-{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
-Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
-%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
-%Though for many applications of MC, assuming a correct lock implementation is sufficient,
-Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
-we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
-%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
-% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
-% (as well as on the mutex API boundaries).
-% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
-
-%% RIP this.
-%Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
-%which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
-%Intuitively, this line shows that for programs with only benign data races,
-%\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
 
@@
 \end{table*}
+
+% TODO: Fix up this paragraph after preempt-everywhere results come in.
+The left half of Table~\ref{tab:drbugs}
+breaks down the number and types of bugs found by each test program.
+In \mxtest, in which we do not trust the lock implementation's correctness,
+% TODO CAMREADY: Fix, SSSMC-ShM finds 2 mx bugs.
+\revisionx{SSS-MC-ICB and SSS-MC-DPOR}
+found dramatically fewer bugs (just 1)\footnote{
+	The one bug SSS-MC found was in a fully-assembler lock implementation.
+{\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions.
+Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
+%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
+%Though for many applications of MC, assuming a correct lock implementation is sufficient,
+Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
+we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
+%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
+% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
+% (as well as on the mutex API boundaries).
+% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
+
+%% RIP this.
+%Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
+%which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
+%Intuitively, this line shows that for programs with only benign data races,
+%\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
 
@@
 verifications provided by each approach.
-\revisionx{The next best approach for verifications was SSS-MC-Shared-Mem, which completed its search in only 39 tests.}
+\revisionx{The next best approach for verifications was SSS-MC-Shared-Mem, which completed its search in only 39 cases.}
 

### 2016-07-19 oopsla-16/intro.tex
sully revisions

@@
 %makes full state space completion intractable for even modestly-sized tests.
-\revisionx{Some tools, such as CHESS and Inspect \cite{inspect} can strike a middle ground using compiler instrumentation to statically add preemption points on memory accesses.
+\revisionx{Some tools, such as CHESS and Inspect \cite{inspect} can strike a middle ground by using compiler instrumentation to statically add preemption points on memory accesses.
 Nevertheless,}
@@
 	\item {\em Iterative Deepening}, a new \revision{algorithm}~for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
-	\item A proof of convergence \revisionx{for sequentially-consistent memory models}, showing that should it be possible in the given CPU budget,
-		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
+	\item A proof of convergence, showing that should it be possible in the given CPU budget,
+		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules \revisionx{(assuming a sequentially-consistent memory model)};
 	\item A new tactic for eliminating one class of false-positive data race candidates,

### 2016-07-19 oopsla-16/overview.tex
sully revisions

@@
 %
+C++ declares any race between two non-atomic locations to be undefined behaviour \cite{cpp-foundation}.
 We take the former camp:
 we consider a data race a bug only when it results in a visible failure state (e.g., crash or deadlock).
+(From a C++ perspective, we assume all concurrent accesses are implemented by {\tt std::atomic}.)
 We bypass concerns of compiler reordering by checking programs at the executable level;

### 2016-07-19 oopsla-16/coverletter/letter.tex
add cover letter skeleton

\documentclass{article}
\special{papersize=8.5in,11in}

%\usepackage[charter]{mathdesign}
\usepackage{fullpage}

\title{OOPSLA Round 2 Cover Letter \\ {\large Paper \#177: Stateless Model Checking with Data-Race Preemption Points}}
\author{Ben Blum (\textsf{bblum@cs.cmu.edu}) \and Garth Gibson (\textsf{garth@cs.cmu.edu})}
\date{July 19, 2016}

\begin{document}

\maketitle
\thispagestyle{empty}

\noindent To the OOPSLA Program Committee,
\\

Thank you for your helpful reviews. Following is a list of revisions we have made, sorted by which of the mandatory revisions they address.

\begin{enumerate}
	\item {\bf Acknowledge sequentially-consistent memory assumption.}
	\begin{itemize}
		\item ...
	\end{itemize}
	\item {\bf Implement and compare with precise Happens-Before.}
	\begin{itemize}
		\item ...
	\end{itemize}
	\item {\bf Explain how we add new preemption points.}
	\begin{itemize}
		\item ...
	\end{itemize}
	\item {\bf Elaborate on related work (MCR and FastTrack).}
	\begin{itemize}
		\item ...
	\end{itemize}
	\item {\bf Philosophical difference between data races and failures.}
	\begin{itemize}
		\item ...
	\end{itemize}
\end{enumerate}

We have also made some changes that were not included in the mandatory revisions, to address other reviewer comments and in one case to fix some bugs in our experiments, as follows:

\begin{itemize}
	\item ...
\end{itemize}

\noindent Regards,
\\

\noindent Ben Blum \\
Garth Gibson

\end{document}

### 2016-07-20 oopsla-16/overview.tex
garth revisions

@@
 and decide on-the-fly whether to pursue each test, or to defer it in favour of others.
+
+\revisionx{
+{\bf Recent advances.}
+Like many prior checkers \cite{chess,dbug-ssv,inspect,portend}, ours implements DPOR for sequentially-consistent hardware.
+In the worst case, these tools may all suffer false negatives as they miss weak-memory-only bugs.
+Recently, Zhang et al. \cite{tsopso} introduced a new formalism with which DPOR can control weak memory nondeterminism,
+such as reordering store buffers.
+Iterative Deepening could use this new technique,
+%We expect this technique would apply directly with Iterative Deepening,
+provided a model checker which implements such reorderings (not yet available to us).
+%We believe this assumption is reasonable for the scope of this paper because the main property we require of DPOR is that
+%A recent paper extended DPOR to support weaker memory models \cite{tsopso}, so our proofs could likewise be extended with that version of DPOR in future work.
+
+Maximal Causality Reduction (MCR) \cite{mcr}, a reduction algorithm which may replace DPOR,
+has also recently shown promising performance improvements over prior model checkers.
+We expect Iterative Deepening to be compatible with MCR,
+and look forward to evaluating the combination when an MCR implementation becomes widely available.
+}
 
@@
 \cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
-it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
-as shown in the example program in Figure~\ref{fig:hb-example}(a).
+it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving.
+%as shown in the example program in Figure~\ref{fig:hb-example}(a).
+\revisionx{Figure~\ref{fig:hb-example}(a) shows a contrived example program in which HB masks the potential race.}
 We call such unreported access pairs {\em false negatives}.
@@
 in which the access pair ceases to exist in the alternate interleaving.
-\cite{hybriddatarace} introduces the {\em Limited HB} relation,
+O'Callahan et al. \cite{hybriddatarace} introduced the {\em Limited HB} relation,
 which will report such potential races
@@
 %
-C++ declares any race between two non-atomic locations to be undefined behaviour \cite{cpp-foundation}.
+(C++ declares any race between two non-atomic locations to be undefined behaviour \cite{cpp-foundation}.)
 We take the former camp:

### 2016-07-20 oopsla-16/soundness.tex
garth revisions

@@
 
-\revisionx{These proofs are built on a DPOR algorithm definition which assumes sequentially-consistent memory hardware.
-%
-% TODO: Aaargh, can't think of how to actually justify this.
-%We believe this assumption is reasonable for the scope of this paper because the main property we require of DPOR is that 
-A recent paper extended DPOR to support weaker memory models \cite{tsopso}, so our proofs could likewise be extended with that version of DPOR in future work.
-We also assume the Limited HB definition for the data-race analysis, leaving the case for precise HB to future work.
+\revisionx{These proofs are built on a DPOR algorithm definition which assumes sequentially-consistent memory hardware,
+as discussed in \sect{\ref{sec:overview-sssmc}}.
+We also assume the Limited HB definition
+for the data-race analysis, as discussed in
+\sect{\ref{sec:overview-dr}}.
+%leaving the case for precise HB to future work.
 }

### 2016-07-20 oopsla-16/coverletter/letter.tex
change letter skeleton a bit
write the cover letter

@@
 
-%\usepackage[charter]{mathdesign}
-\usepackage{fullpage}
+\usepackage[charter]{mathdesign}
+%\usepackage{fullpage}
 
-\title{OOPSLA Round 2 Cover Letter \\ {\large Paper \#177: Stateless Model Checking with Data-Race Preemption Points}}
+\title{OOPSLA Round 2 Cover Letter \\ {\normalsize Paper \#177: Stateless Model Checking with Data-Race Preemption Points}}
 \author{Ben Blum (\textsf{bblum@cs.cmu.edu}) \and Garth Gibson (\textsf{garth@cs.cmu.edu})}
@@
 \begin{enumerate}
-	\item {\bf Acknowledge sequentially-consistent memory assumption.}
+	\item {\bf Acknowledge sequentially-consistent (SC) memory assumption.}
 	\begin{itemize}
-		\item ...
+		\item Added paragraph in section 2.1 (page 3), discussing SC assumption, and dicsussing potential future work to incorporate [Zhang 2015].
+The assumption is justified by the fact that most tools from prior work make the same assumption, and that the assumption's worst case compromises verification at worst, but not bug-finding potential.
+		\item Add mention of SC assumption in the introduction (page 2),
+		when introducing the proofs in section 5 (page 7),
+		and when discussing verification results in section 6.3 (page 11).
 	\end{itemize}
-	\item {\bf Implement and compare with precise Happens-Before.}
+	\item {\bf Implement and compare with precise Happens-Before (HB).}
 	\begin{itemize}
-		\item ...
+		\item Feature implemented and experiments performed. Results are presented as a new data series in Figures 7(a), 7(b), and 8, and as a new column in Table 1 (pages 10 and 12).
+		\item Mentioned implementation of new data-race analysis mode in section 4.1 (page 6).
+		\item Added new evaluation question in the introduction to section 6 (page 9).
+		\item Defined ``QS-Pure-HB'' as a new experimental setup in section 6.2 (page 10).
+Also clarified between QS-Limited-HB and QS-Pure-HB when discussing Quicksand's results on data-race bugs, where before we just said ``Quicksand'' (pages 9-13).
+		\item Added paragraph discussing evaluation results comparing the two HB approaches in section 6.3 (page 12).
 	\end{itemize}
@@
 	\begin{itemize}
-		\item ...
+		\item Added section 3.1 to introduce the design of Iterative Deepening in more detail (page 4).
+		\item Added Algorithm 1 to show explicitly the ``na\"ive'' approach of adding more preemption points (page 4), to use as a starting point to explain algorithms 2 and 3 more clearly.
+		\item Swapped the order of sections 3.3 and 3.4 (pages 5-6) to improve narrative flow.
 	\end{itemize}
-	\item {\bf Elaborate on related work (MCR and FastTrack).}
+	\item {\bf Elaborate on related work (CHESS, PCT, MCR, and FastTrack).}
 	\begin{itemize}
-		\item ...
+		\item Clarified the capabilities of CHESS in the introduction (page 1).
+		\item Added a paragraph comparing MCR to DPOR in section 2.1 (page 3).
+		\item Expanded explanation of PCT in section 8.1 to clarify incompatibility with DPOR (page 14).
+		\item Expanded section 8.2's introductory paragraph to discuss FastTrack and DJIT+ (pages 14-15).
 	\end{itemize}
@@
 	\begin{itemize}
-		\item ...
+		\item Added paragraph in section 2.2 to discuss the philosophical difference explicitly (page 4).
+		\item Added Table 3 in section 6 to count the number of total and verified-benign races during the QS-Pure-HB experiment (page 13).
+(``Benign DRs'' displays new data; ``Total DR PPs'' and ``Untested DR PPs'' are updated to reflect QS-Pure-HB rather than the old numbers from QS-Limited-HB, and the latter two plus ``Malloc DRs'' have been relocated from Table 1 to here.)
 	\end{itemize}
@@
 \begin{itemize}
-	\item ...
+	\item Updated several of the experiments in our evaluation.
+		\begin{itemize}
+			\item Most importantly, we discovered a bug in our implementation of SSS-MC-Shared-Mem which caused it to ignore many shared memory PPs, leading it to find far too few bugs and erroneously declare too many verifications.
+		We fixed the bug and re-ran this experiment, leading to the updated data series shown in Figures 7 and 8 and Table 1;
+		the new results show increased bug-finding capacity (still not as good as Quicksand) and much slower verification speed.
+		We removed the discussion of SSS-MC-Shared-Mem's potential unsoundness from section 6.3 (no longer necessary to justify its position on Figure 8),
+		and updated the overall bug-finding ratio between QS-Limited-HB and SSS-MC from 1.5x to 1.25x (pages 1 and 10).
+			\item We fixed several other minor bugs in our test cases, leading to slightly changed statistics in Table 1 but no significant difference to the layout of our performance graphs.
 		\end{itemize}
+	\item Clarified in section 2.2 that Figure 3 is a contrived example (page 3).
+	\item Clarified in section 2.2 the advantage offered by Limited HB and the possibility of false negatives (page 3).
+	\item Mentioned in section 5.2 the possibility, and downside, of avoiding malloc-recycle false positives by using a hacked allocator (page 8).
+	\item Added analysis of ICB preemption bound distribution in section 6.3 (page 11) and Table 2 (page 12) to defuse concerns about possible bias in our suite of bugs.
+	\item Mentioned the possibility for future work to occasionally prioritize bigger jobs over small ones in section 7 (page 13).
+	\item Reduced clutter among citations, removing ISBNs, DOIs, ``In Proceedings of the Xth annual...'', and so on.
+\end{itemize}
+
+We are still working to improve the formalism, as suggested by Reviewer D,
+although we prioritized the mandatory revisions (especially implementing and evaluating precise HB) for this deadline.
+An improved version of the formalism will be ready for the camera-ready date.
+
+We feel our paper has been much improved since implementing your suggestions, and hope you agree.
+\\
 

### 2016-07-20 oopsla-16/coverletter/letter.tex
tweak letter

@@
 \usepackage[charter]{mathdesign}
-%\usepackage{fullpage}
+\usepackage{fullpage}
 
-\title{OOPSLA Round 2 Cover Letter \\ {\normalsize Paper \#177: Stateless Model Checking with Data-Race Preemption Points}}
+\title{OOPSLA Round 2 Cover Letter \\ {\large Paper \#177: Stateless Model Checking with Data-Race Preemption Points}}
 \author{Ben Blum (\textsf{bblum@cs.cmu.edu}) \and Garth Gibson (\textsf{garth@cs.cmu.edu})}
@@
 	\item Mentioned the possibility for future work to occasionally prioritize bigger jobs over small ones in section 7 (page 13).
-	\item Reduced clutter among citations, removing ISBNs, DOIs, ``In Proceedings of the Xth annual...'', and so on.
+	\item Reduced clutter among citations, removing ISBNs, DOIs, ``In Proceedings of the Xth Annual...'', and so on.
 \end{itemize}
@@
 Garth Gibson
+\thispagestyle{empty}
 

### 2016-07-20 oopsla-16/evaluation.tex
garth revisions
don't hyphenate Limited-HB (unless QS-Limited-HB)

@@
 			\revisionx{
-			\item How does the choice between HB and Limited-HB affect bug-finding and verification performance?
+			\item How does the choice between HB and Limited HB affect bug-finding and verification performance?
 			}
@@
 		\begin{enumerate}
-		\item Does \quicksand~avoid false positives compared to a single-execution \revisionx{Limited-HB} data race analysis?
+		\item Does \quicksand~avoid false positives compared to a single-execution \revisionx{Limited HB} data race analysis?
 				% Explain later as:
@@
 		\revisionx{
-	\item {\bf QS-Limited-HB}: \quicksand~using Limited-HB for its data-race analysis,
+	\item {\bf QS-Limited-HB}: \quicksand~using Limited HB for its data-race analysis,
 	\item {\bf QS-Pure-HB}: \quicksand~using pure HB instead, and
@@
 	by several configurations of \quicksand~and the SSS-MC control.
-	\quicksand~finds \revisionx{125\%} as many bugs with data-race PPs.}
+	\quicksand~finds \revisionx{125\%} as many bugs \revisionx{at the 10-hour mark compared to the best SSS-MC approach}.}
+	%with data-race PPs.}
 	\label{fig:dowefindbugsfaster}

### 2016-07-20 oopsla-16/overview.tex
footnoteify discussion of c++ spec

@@
 %
-(C++ declares any race between two non-atomic locations to be undefined behaviour \cite{cpp-foundation}.)
 We take the former camp:
-we consider a data race a bug only when it results in a visible failure state (e.g., crash or deadlock).
-(From a C++ perspective, we assume all concurrent accesses are implemented by {\tt std::atomic}.)
+we consider a data race a bug only when it results in a visible failure state (e.g., crash or deadlock)\footnote{
+C++ declares any race between two non-atomic locations to be undefined behaviour \cite{cpp-foundation}.
+From a C++ perspective, we assume all concurrent accesses are implemented by {\tt std::atomic} loads and stores.}.
 We bypass concerns of compiler reordering by checking programs at the executable level;

### 2016-07-20 oopsla-16/evaluation.tex
update table 3, drstatistix, w/ real numbers!

@@
 		\small
+		\revisionx{
 	\begin{tabular}{r||c|c|c||c}
@@
 		\hline
-		% FIXME CAMREADY: update FRMs
 				% DRPPs   benign untested FRMs
-		% TODO: recount dr pps for pure hb
-		% TODO: count dr pps in totalverifs for benign
-		% TODO: recount untesteds for pure hb
-		{\tt bcast}	& 912	& \revisionx{1337}	& 107	& 51	\\
-		{\tt join} 	& 781	& \revisionx{1337}	& 292	& 333	\\
-		{\tt mx} 	& 829	& \revisionx{1337}	& 1	& 7	\\
-		{\tt sem} 	& 753	& \revisionx{1337}	& 279	& 140	\\
-		{\tt signal} 	& 1118	& \revisionx{1337}	& 391	& 180	\\
-		{\tt rwlock} 	& 915	& \revisionx{1337}	& 634	& 125	\\
+		{\tt bcast}	& 655	& 97	& 150	& 52	\\
+		{\tt join} 	& 566	& 68	& 249	& 338	\\
+		{\tt mx} 	& 911	& 127	& 44	& 7	\\
+		{\tt sem} 	& 783	& 2	& 414	& 166	\\
+		{\tt signal} 	& 936	& 9	& 510	& 180	\\
+		{\tt rwlock} 	& 543	& 1	& 310	& 156	\\
 		\hline
-		{\tt sched} 	& 144	& \revisionx{1337}	& 3	& 0	\\
-		{\tt alarm} 	& 115	& \revisionx{1337}	& 89	& 35	\\
-		{\tt wait} 	& 142	& \revisionx{1337}	& 31	& 31	\\
+		{\tt sched} 	& 65	& 51	& 3	& 0	\\
+		{\tt alarm} 	& 35	& 0	& 29	& 35	\\
+		{\tt wait} 	& 71	& 1	& 28	& 31	\\
 		\hline
-		{\bf Total}	& 5709	& \revisionx{1337}	& 1827	& 902	\\
+		{\bf Total}	& 4565	& 356	& 1737	& 965	\\
 	\end{tabular}
+		}
 	\end{center}
@@
 		\revisionx{Additional data race statistics.}
-		% TODO: update this for pure-hb and add a column,
-		``Total DR PPs'' counts how many unique data-racing instructions \revisionx{QS-Pure-HB identified}.
+		``Total DR PPs'' counts how many unique data-racing instructions \revisionx{QS-Pure-HB identified} among tests
+		where it found no bugs.
 		Among those,

### 2016-07-20 oopsla-16/coverletter/letter.tex
move the table again; and fix corresponding page numbers in cover letter

@@
 		when introducing the proofs in section 5 (page 7),
-		and when discussing verification results in section 6.3 (page 11).
+		and when discussing verification results in section 6.3 (page 12).
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Feature implemented and experiments performed. Results are presented as a new data series in Figures 7(a), 7(b), and 8, and as a new column in Table 1 (pages 10 and 12).
+		\item Feature implemented and experiments performed. Results are presented as a new data series in Figures 7(a), 7(b), and 8, and as a new column in Table 1 (pages 10-12).
 		\item Mentioned implementation of new data-race analysis mode in section 4.1 (page 6).
@@
 		\item Added paragraph in section 2.2 to discuss the philosophical difference explicitly (page 4).
-		\item Added Table 3 in section 6 to count the number of total and verified-benign races during the QS-Pure-HB experiment (page 13).
+		\item Added Table 3 in section 6 to count the number of total and verified-benign races during the QS-Pure-HB experiment (page 12).
 (``Benign DRs'' displays new data; ``Total DR PPs'' and ``Untested DR PPs'' are updated to reflect QS-Pure-HB rather than the old numbers from QS-Limited-HB, and the latter two plus ``Malloc DRs'' have been relocated from Table 1 to here.)
@@
 	\item Mentioned in section 5.2 the possibility, and downside, of avoiding malloc-recycle false positives by using a hacked allocator (page 8).
-	\item Added analysis of ICB preemption bound distribution in section 6.3 (page 11) and Table 2 (page 12) to defuse concerns about possible bias in our suite of bugs.
+	\item Added analysis of ICB preemption bound distribution in section 6.3 and Table 2 (page 11) to defuse concerns about possible bias in our suite of bugs.
 	\item Mentioned the possibility for future work to occasionally prioritize bigger jobs over small ones in section 7 (page 13).

### 2016-08-04 oopsla-16/evaluation.tex
all mandos done!
move the table again; and fix corresponding page numbers in cover letter
update graphs and tables for updated bug numbers caused by fixing sched/icb assertions

@@
 
+\begin{table*}[t]
+	\begin{center}
+		\small
+		\revisionx{
+	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
+		& {\bf Num.} & {\bf QS-LHB} & {\bf QS-PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem} & {\bf LHB DR} & {\bf Nondet.} & {\bf Mutual} & {\bf Avg. tested} \\
+		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf timeouts} & {\bf subset SSes} \\
+		\hline
+				% XXX FIXME: Need rerun nondet dr expt!
+				% #test  qs-lhb  qs-phb   icb     dpor    everyw  dronly  nondets timeout comp.SSes
+		{\tt bcast}	& 79	& 8	& 8	& 5	& 6	& 7	& 2	& 4	& 7	& 112.3	\\
+		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& 11	& 75.9	\\
+		{\tt mx} 	& 79	& 10	& 9	& 1	& 1	& 12	& 9	& 1	& 0	& -	\\
+		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& 36	& 93.4	\\
+		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& 59.0	\\
+		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& 43	& 88.3	\\
+		\hline
+		{\tt sched} 	& 59	& 7	& 7	& 1	& 1	& 8	& 6	& 4	& 2	& 13.0	\\
+		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& 16	& 8.1	\\
+		{\tt wait} 	& 52	& 30	& 27	& \revisionz{24}
+								& 23	& 1	& 7	& 1	& 13	& 32.8	\\
+		\hline
+		{\bf Total}	& 629	& 153	& 133	& \revisionz{102}
+								& 88	& 122	& 69	& 32	& 173	& 69.5 	\\
+	\end{tabular}
+		}
+	\end{center}
+	\caption{Summary of bugs
+	%and data races
+	found by each test program.
+		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
+		``LHB DR bugs'' counts among \revisionx{QS-Limited-HB's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
+	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
+	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
+		}
+	\label{tab:drbugs}
+\end{table*}
+
 {\bf Finding new data-race bugs.}
@@
 %}
-
-\begin{table*}[t]
-	\begin{center}
-		\small
-		\revisionx{
-	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
-		& {\bf Num.} & {\bf QS-LHB} & {\bf QS-PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem} & {\bf LHB DR} & {\bf Nondet.} & {\bf Mutual} & {\bf Avg. tested} \\
-		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf timeouts} & {\bf subset SSes} \\
-		\hline
-				% XXX FIXME: Need rerun nondet dr expt!
-				% #test  qs-lhb  qs-phb   icb     dpor    everyw  dronly  nondets timeout comp.SSes
-		{\tt bcast}	& 79	& 8	& 8	& 5	& 6	& 7	& 2	& 4	& 7	& 112.3	\\
-		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& 11	& 75.9	\\
-		{\tt mx} 	& 79	& 10	& 9	& 1	& 1	& 12	& 9	& 1	& 0	& -	\\
-		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& 36	& 93.4	\\
-		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& 59.0	\\
-		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& 43	& 88.3	\\
-		\hline
-		{\tt sched} 	& 59	& 7	& 7	& 1	& 1	& 8	& 6	& 4	& 2	& 13.0	\\
-		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& 16	& 8.1	\\
-		{\tt wait} 	& 52	& 30	& 27	& 23	& 23	& 1	& 7	& 1	& 13	& 32.8	\\
-		\hline
-		{\bf Total}	& 629	& 153	& 133	& 101	& 88	& 122	& 69	& 32	& 173	& 69.5 	\\
-	\end{tabular}
-		}
-	\end{center}
-	\caption{Summary of bugs
-	%and data races
-	found by each test program.
-		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
-		``LHB DR bugs'' counts among \revisionx{QS-Limited-HB's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
-	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
-	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
-	% TODO CAMREADY: update avg tested subsets to be from the ICB timeouts
-	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
-		}
-	\label{tab:drbugs}
-\end{table*}
 
@@
 
-\begin{table}[t]
+\begin{table}[h]
 	\begin{center}
@@
 		1	& 82	& 86	\\
-		2	& 15	& 32	\\
+		2	& \revisionz{16}	& 32	\\
 		3	& 2	& 3	\\
@@
 		\hline
-		Total	& 101	& 122	\\
+		Total	& \revisionz{102}	& 122	\\
 	\end{tabular}
@@
 		{\bf Test} & {\bf DR PPs} & {\bf \revisionx{DRs}} & {\bf DR PPs} & {\bf DRs} \\
+		% note: if you wanted to calculate total DR PPs with FAB tests included, here's the nums instead:
+		% alarm 66
+		% bct 849
+		% mx 1193
+		% paradise/sem 1201
+		% paraguay/signal 1088
+		% sema 101
+		% dgr 846
+		% tej 1029
+		% wait 196
 		\hline

### 2016-08-08 oopsla-16/evaluation.tex
change formatting according to camready author package

@@
 
-\subsection{\revision{Comparison to state-of-the-art MC}}%Comparing Iterative Deepening to SSS-MC
+\subsection{\revision{Comparison to State-of-the-Art MC}}%Comparing Iterative Deepening to SSS-MC
 \label{sec:eval-sssmc}
@@
 
-\subsection{Comparing to single-pass data-race analysis}
+\subsection{Comparing to Single-Pass Data-Race Analysis}
 \label{sec:eval-dr}

### 2016-08-08 oopsla-16/paper.tex
Incorporate pure HB graphs and discussion in evaluation section; tons of prose changes omg
fix line break in conclusion
prepare for cam ready revisions
change formatting according to camready author package

@@
 \documentclass[10pt,times,numbers]{sigplanconf}
-\usepackage[2ndsubmission]{oopsla2016}
+\usepackage[final]{oopsla2016}
 
@@
 \usepackage{amsthm}
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}
+\usepackage{flushend}
+\usepackage{color}
+\usepackage[colorlinks=false,breaklinks,draft=false]{hyperref}
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
@@
 
+% presubmit revisions
 \newcommand\revision[1]{#1}
 \definecolor{purple}{RGB}{192,0,192}
-\newcommand\revisionx[1]{\hilight{purple}{#1}\xspace}
 %\newcommand\revision[1]{\hilight{purple}{#1}}
+% 2nd round revisions
+%\newcommand\revisionx[1]{\hilight{purple}{#1}\xspace}
+\newcommand\revisionx[1]{#1}
+% cam ready revisions
+\newcommand\revisionz[1]{\hilight{purple}{#1}\xspace}
 
@@
 
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
-\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
+\authorinfo{Ben Blum}{Carnegie Mellon University, USA}{bblum@cs.cmu.edu}
+\authorinfo{Garth Gibson}{Carnegie Mellon University, USA}{garth@cs.cmu.edu}
 %\authorinfo{\vspace{24pt}}{}{}
@@
 %with data-race preemption points
-1.5x as many bugs and verified 4.3x as many tests compared to \revision{prior model checking approaches}. %stateless model checking without data-race preemption points.
+\revisionx{1.25x} as many bugs and verified 4.3x as many tests compared to \revision{prior model checking approaches}. %stateless model checking without data-race preemption points.
 %Our evaluation shows this technique is
@@
 \revision{\category{D.2.4}{Software Engineering}{Software/Program Verification}}
+\keywords{model checking, data races, verification}
 %\category{D.2.5}{Software Engineering}{Testing and Debugging}
@@
 \url{https://github.com/bblum/oopsla-quicksand}
+
 % TODO CAMREADY: Make data set, at least, public.

### 2016-08-09 oopsla-16/evaluation.tex
update timeouts/subsets

@@
 		{\tt bcast}	& 79	& 8	& 8	& 5	& 6	& 7	& 2	& 4	& 7	& 112.3	\\
-		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& 11	& 75.9	\\
+		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& \revisionz{12}
+													& \revisionz{69.7}	\\
 		{\tt mx} 	& 79	& 10	& 9	& 1	& 1	& 12	& 9	& 1	& 0	& -	\\
-		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& 36	& 93.4	\\
-		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& 59.0	\\
-		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& 43	& 88.3	\\
+		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& \revisionz{50}
+													& \revisionz{77.4}	\\
+		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& \revisionz{59.6}	\\
+		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& \revisionz{44}
+													& \revisionz{86.3}	\\
 		\hline
 		{\tt sched} 	& 59	& 7	& 7	& 1	& 1	& 8	& 6	& 4	& 2	& 13.0	\\
-		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& 16	& 8.1	\\
+		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& \revisionz{17}
+													& \revisionz{7.8}	\\
 		{\tt wait} 	& 52	& 30	& 27	& \revisionz{24}
-								& 23	& 1	& 7	& 1	& 13	& 32.8	\\
+								& 23	& 1	& 7	& 1	& \revisionz{15}
+													& \revisionz{33.8}	\\
 		\hline
 		{\bf Total}	& 629	& 153	& 133	& \revisionz{102}
-								& 88	& 122	& 69	& 32	& 173	& 69.5 	\\
+								& 88	& 122	& 69	& 32	& \revisionz{192}
+													& \revisionz{65.8} 	\\
 	\end{tabular}

### 2016-08-12 oopsla-16/overview.tex
minor tweak

@@
 We also identify the {\em minimal} and {\em maximal state space} for each test.
-The {\em minimal state space} includes only thread switches arising from no preemptions (Figure~\ref{fig:id}, A).
-The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}, D).
+The {\em minimal state space} includes only thread switches arising from no preemptions (Figure~\ref{fig:id}(a)).
+The {\em maximal state space} is the one tested by SSS-MC: all statically-available PPs are enabled (Figure~\ref{fig:id}(d)).
 %However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.

### 2016-08-13 oopsla-16/evaluation.tex
Rerun nondet expt and update results in table/graph

--- 2016-08-08 oopsla-16/evaluation.tex
@@
 
+\newcommand\revz[1]{\revisionz{#1}}
+
 \begin{table*}[t]
@@
 		\revisionx{
-	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
-		& {\bf Num.} & {\bf QS-LHB} & {\bf QS-PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem} & {\bf LHB DR} & {\bf Nondet.} & {\bf Mutual} & {\bf Avg. tested} \\
-		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf timeouts} & {\bf subset SSes} \\
+	\begin{tabular}{r|c||c|c|c|c|c||c|c|c|c||c|c}
+		\multicolumn{2}{c||}{} & \multicolumn{5}{c||}{\bf \revz{Total bugs}} & \multicolumn{4}{c||}{\bf \revz{Data-race bugs}} \\
+		% ---
+		& {\bf Num.} & \multicolumn{2}{c|}{\bf \revz{\quicksand}} & \multicolumn{3}{c||}{\bf \revz{Control (SSS-MC)}}
+		& \multicolumn{2}{c|}{\bf \revz{Limited HB}} & \multicolumn{2}{c||}{\bf \revz{Pure HB}} &
+		%& {\bf LHB DR} & {\bf Nondet.} & & &
+		{\bf Mutual} & {\bf Avg. tested} \\
+		% ---
+		{\bf Test} & {\bf tested} & {\bf LHB} & {\bf PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem}
+		& {\bf All} & {\bf Nondet.} & {\bf \revz{All}} & {\bf \revz{Nondet.}}
+		& {\bf timeouts} & {\bf subset SSes} \\
+		% ---
 		\hline
-				% XXX FIXME: Need rerun nondet dr expt!
-				% #test  qs-lhb  qs-phb   icb     dpor    everyw  dronly  nondets timeout comp.SSes
-		{\tt bcast}	& 79	& 8	& 8	& 5	& 6	& 7	& 2	& 4	& 7	& 112.3	\\
-		{\tt join} 	& 79	& 23	& 20	& 13	& 13 	& 14	& 11	& 6	& 11	& 75.9	\\
-		{\tt mx} 	& 79	& 10	& 9	& 1	& 1	& 12	& 9	& 1	& 0	& -	\\
-		{\tt sem} 	& 79	& 17	& 16	& 12	& 11	& 12	& 7	& 5	& 36	& 93.4	\\
-		{\tt signal} 	& 79	& 10	& 8	& 5	& 5	& 11	& 6	& 6	& 45	& 59.0	\\
-		{\tt rwlock} 	& 79	& 27	& 26	& 25	& 23	& 28	& 4	& 2	& 43	& 88.3	\\
+				% #test  qs-lhb  qs-phb           icb             dpor    everyw  dronly  nondets(LHB)    dronly(PHB)    nondets(LHB) mutualtimeout    comp.SSes
+		{\tt bcast}	& 79	& 8	& 8		& 5		& 6	& 7	& 2	& \revz{1}	& \revz{2}	& \revz{1}	& 7		& 112.3	\\
+		{\tt join} 	& 79	& 23	& 20		& 13		& 13 	& 14	& 11	& \revz{4}	& \revz{7}	& \revz{3}	& \revz{12}	& \revz{69.7}	\\
+		{\tt mx} 	& 79	& 10	& 9		& 1		& 1	& 12	& 9	& \revz{1}	& \revz{8}	& \revz{1}	& 0		& -	\\
+		{\tt sem} 	& 79	& 17	& 16		& 12		& 11	& 12	& 7	& \revz{3}	& \revz{6}	& \revz{2}	& \revz{50}	& \revz{77.4}	\\
+		{\tt signal} 	& 79	& 10	& 8		& 5		& 5	& 11	& 6	& \revz{1}	& \revz{3}	& \revz{2}	& 45		& \revz{59.6}	\\
+		{\tt rwlock} 	& 79	& 27	& 26		& 25		& 23	& 28	& 4	& \revz{1}	& \revz{3}	& \revz{0}	& \revz{44}	& \revz{86.3}	\\
 		\hline
-		{\tt sched} 	& 59	& 7	& 7	& 1	& 1	& 8	& 6	& 4	& 2	& 13.0	\\
-		{\tt alarm} 	& 44	& 21	& 12	& 16	& 5	& 29	& 17	& 3	& 16	& 8.1	\\
-		{\tt wait} 	& 52	& 30	& 27	& \revisionz{24}
-								& 23	& 1	& 7	& 1	& 13	& 32.8	\\
+		{\tt sched} 	& 59	& 7	& 7		& 1		& 1	& 8	& 6	& \revz{4}	& \revz{6}	& \revz{6}	& 2		& 13.0	\\
+		{\tt alarm} 	& 44	& 21	& 12		& 16		& 5	& 29	& 17	& \revz{1}	& \revz{7}	& \revz{6}	& \revz{17}	& \revz{7.8}	\\
+		{\tt wait} 	& 52	& 30	& \revz{26}	& \revz{24}	& 23	& 1	& 7	& \revz{2}	& \revz{2}	& \revz{0}	& \revz{15}	& \revz{33.8}	\\
 		\hline
-		{\bf Total}	& 629	& 153	& 133	& \revisionz{102}
-								& 88	& 122	& 69	& 32	& 173	& 69.5 	\\
+		{\bf Total}	& 629	& 153	& \revz{132}	& \revz{102}	& 88	& 122	& 69	& \revz{15}	& \revz{44}	& \revz{21}	& \revz{192}	& \revz{65.8} 	\\
 	\end{tabular}
@@
 		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
-		``LHB DR bugs'' counts among \revisionx{QS-Limited-HB's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
-	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
+		``\revisionz{Data-race} bugs'' counts among \revisionz{\quicksand's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
+	among those, \revisionz{the ``Nondeterministic'' columns show} how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
 	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{nondet-drs-1-v2.pdf}
+	\includegraphics[width=0.48\textwidth]{nondets.pdf}
 	\caption{Some data-race candidates may not be identified during a single program execution.
 		Using nondeterministic data races as PPs,
-		\quicksand~found 189\% as many data-race bugs compared to using single-pass candidates alone.
+		% lhb: 127.77%
+		% phb: 191.30%
+		\quicksand~found \revisionz{128\% (Limited HB) to 191\% (Pure HB)} as many data-race bugs compared to using single-pass candidates alone.
 	}
@@
 which includes the former type as well as new bugs involving nondeterministic data-races.
-\quicksand~found 68 data-race bugs in total, only 36 of which could be found with single-pass data-race candidates alone.
+\revisionz{QS-LHB found 69 data-race bugs in total, 54} of which could be found with single-pass data-race candidates alone.
+\revisionz{QS-PHB is even more dependent on nondeterministic data-race PPs, requiring nondeterministic data-race PPs in 21 cases among its 44 total data-race bugs.}
 %In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.

### 2016-08-13 oopsla-16/evaluation.tex
Update number of frms-for-realsies

@@
 In Table~\ref{tab:drstatistix}, the column ``Malloc-recycle DRs'' shows the total number of such data-race candidates for each test program.
-In total, 902 data-races fit the malloc-recycle pattern across all tests,
-only 69 of which were observed to avoid the re-allocation in an alternate interleaving.
-Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all 833 other state spaces.
+In total, \revisionz{965} data-races fit the malloc-recycle pattern across all tests,
+% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
+only \revisionz{64} of which were observed to avoid the re-allocation in an alternate interleaving.
+Our proof in \sect{\ref{sec:recycle}} guarantees the safety of pruning all \revisionz{901} other state spaces.
 
-Among those 69 true data-races, %which initially fit the malloc-recycle pattern,
+Among those \revisionz{64} true data-races, %which initially fit the malloc-recycle pattern,
 none exposed a new bug when used as a PP.
@@
 even if unsound without Iterative Deepening.
-However, \quicksand~was able to correctly identify the 69 violations of that heuristic (among 30 distinct tests),
+However, \quicksand~was able to correctly identify the \revisionz{64} violations of that heuristic (among \revisionz{26} distinct tests),
 and fall back to classifying them with DPOR.

### 2016-08-16 sigcse-17/intro.tex
sigcse: write abstrax + intro

\section{Introduction}
\label{sec:intro}

% TODO: lol.
Building real-world multithreaded concurrent programming systems is well-known to be notoriously difficult to design, implement, test, debug, hard to get right, challenging, and error-prone arguably because of the difficulties of finding, reproducing, and fixing concurrency errors, attributable to the astronomically large number of thread interleavings, which arise at random during normal program execution.

In this paper we describe our experience teaching concurrent programming at the operating-system level to advanced undergraduate students at \carnegiemellon (\cmu).
In \fourten, the Operating Systems Design and Implementation class at \cmu,
students complete aggressively low-level thread library and kernel projects,
which run on real hardware,
in which they implement threads and processes.
Presented with a kernel specification called \pebbles%
% TODO camready
%\cite{kspec}
, students build a threading and synchronization library that runs in user space on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
% Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing (SMP).

In comparison with other educational kernel projects,
\pebbles focuses more on ground-up design than on teaching a broad feature set.
At Stanford, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
In both courses, student kernels support multiprocessor execution;
however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
In \fourten, we provide starter code for the machine boot sequence and for a minimal C library (\x{printf()}, \x{malloc()}), leaving students to design threading and scheduling primitives from scratch.

Incorporating SMP into the \fourten curriculum would expand the challenges students face in a single semester.
However, we wish to avoid compromising the complexity
%, hardware fidelity,
and learning experience of the project set.
One solution would be to ease the burden of debugging concurrency errors.
Hence, we have built \landslide%
% TODO camready
%\cite{landslide}
, a tool that uses stateless model checking \cite{verisoft,dpor,chess-icb}
as an alternative to the conventional stress tests.
%to force kernel threads to take execution paths that expose concurrency bugs.
\landslide can test student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
We present our experience sharing \landslide with volunteering students who used it to find bugs in their own code,
and discuss the possibility for integrating it as a main component of operating systems curricula.

### 2016-08-16 sigcse-17/paper.tex
sigcse: write abstrax + intro

\documentclass{sig-alternate-05-2015}

\usepackage{xspace}
\usepackage{color}
\usepackage{graphicx}
\usepackage{listings}

\lstset{language=C,
        frame=none,%single,
        basicstyle=\ttfamily,
        keywordstyle=\bfseries,
        commentstyle=\em,
        tabsize=2,
        showstringspaces=false,
        flexiblecolumns=false,
        mathescape=true,
        escapeinside={\#}{\#},
        % numbers=left, numberstyle=\tiny, numbersep=5pt %, stepnumber=2, 
        numbersep=5pt,
        % literate={->}{{$\rightarrow$}}1 {<->}{{$\leftarrow$}}1
}
\newcommand\x[1]{\lstinline{#1}}

\begin{document}

\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% TODO: ???
%\doi{10.475/123_4}
%\isbn{123-4567-24-567/08/06}
%\acmPrice{\$15.00}
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.

\conferenceinfo{SIGCSE '17}{March 8th - 11th, Seattle, Washington, USA}
\CopyrightYear{2017}

% TODO camready
\newcommand\carnegiemellon{{\em Double-Blind University}\xspace}
\newcommand\cmu{DBU\xspace}
\newcommand\pebbles{\cmu-OS\xspace}
\newcommand\landslide{Landslide\xspace}
\newcommand\fourten{CS999}
%\newcommand\carnegiemellon{{\em Carnegie Mellon University}\xspace}
%\newcommand\cmu{CMU\xspace}
%\newcommand\pebbles{Pebbles\xspace}
%\newcommand\landslide{Landslide\xspace}
%\newcommand\fourten{15-410}

\title{Teaching Low-Level Concurrency with \pebbles and \landslide}

\numberofauthors{3}

% TODO camready
\author{
	\em double-blind submission
%	\alignauthor
%	Ben Blum \\
%	\affaddr{Carnegie Mellon University} \\
%	\affaddr{5000 Forbes Avenue} \\
%	\affaddr{Pittsburgh, PA, USA} \\
%	\email{bblum@cs.cmu.edu}
%	\alignauthor
%	David A. Eckhardt \\
%	\affaddr{Carnegie Mellon University} \\
%	\affaddr{5000 Forbes Avenue} \\
%	\affaddr{Pittsburgh, PA, USA} \\
%	\email{davide@cs.cmu.edu}
%	\alignauthor
%	Garth Gibson \\
%	\affaddr{Carnegie Mellon University} \\
%	\affaddr{5000 Forbes Avenue} \\
%	\affaddr{Pittsburgh, PA, USA} \\
%	\email{garth@cs.cmu.edu}
}

\maketitle
\begin{abstract}

In \carnegiemellon's undergraduate operating systems course
students learn how to write correct and principled concurrent code through
a series of low-level programming projects.
The curriculum includes implementing hardware device drivers and a user-level
threading library, and culminates in a six-week long project in which students
build from the ground up a fully-preemptible, UNIX-like kernel that can run
on real hardware.

We aspire to expand the scope of these projects in response to
typical hardware platforms'
increasing complexity,
especially massive multi-core.
To that end, we wish to reduce the time students spend
debugging concurrency bugs.
We present Landslide, a
%systematic testing
stateless model checker which
offers a more principled approach than stress testing.
% edupar version
%We report on our preliminary experience working with students using Landslide to find previously-overlooked bugs in their own code, and discuss future methods for making Landslide more accessible to struggling students.
% sigcse version
We report on our experience giving students Landslide during the thread library
%and kernel projects.
project,
%Although kernel-mode Landslide requires manual instrumentation, testing thread libraries is fully automatic.
% TODO: verify these, obv. ;)
showing that it allows them to find and fix previously-overlooked bugs.
Further, %even struggling students can take advantage of it,
Landslide's users achieve higher concurrency scores on average in subsequent projects,
% FIXME awkward phrasing
suggesting that debugging with a model checker produces better student retention of concurrency principles.

\end{abstract}

% TODO, do dis
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

% End generated code

\printccsdesc

\keywords{concurrency, debugging, model checking, computer science education}

\input{intro}

\section{The {\secit Body} of The Paper}
% TODO
TODO body

\section{Conclusions}
% TODO
TODO conclusion

% TODO camready
% \section{Acknowledgments}
% Many people have contributed to the development of
% 15-410 over the past decade.
% Stephen Muckle, currently at Qualcomm,
% led the switch to x86 projects running in Simics.
% Nathaniel Wesley Filardo,
% currently a Ph.D.\ candidate at Johns Hopkins,
% % increased the sanity of the build infrastructure,
% designed the \x{swexn()} system call and helped critique this paper.
% The current reference kernel, called ``pathos,''
% was written by Michael J. Sullivan,
% currently a CMU CS Ph.D.\ candidate,
% and Elly Fong-Jones, currently at Google.
% The SMP infrastructure code was contributed by
% Ryan Pearl, currently at Dropbox.
% Joshua Wise, currently at NVIDIA,
% led several kernel-extension projects.
% % including the hypervisor project.
% The course has benefitted
% greatly from free educational Simics licenses
% donated by Virtutech/Wind River.
% Code-size numbers were generated using David A.\ Wheeler's
% ``SLOCCount.''
% Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.

\bibliographystyle{abbrv}
\bibliography{citations}

%\balancecolumns
% That's all folks!
\end{document}

### 2016-08-16 sigcse-17/intro.tex
sigcse: some revisions

@@
 %\cite{kspec}
-, students build a threading and synchronization library that runs in user space on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
-During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
+, students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
+%During these projects
+The students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
 % Finally, in another two-week project, students implement an extension to their Pebbles kernel.

### 2016-08-17 oopsla-16/implementation.tex
take care of some TODO CAMREADies

@@
 Assuming that the lock implementation is already correct enables more productive data-race analysis on the rest of the codebase,
-% TODO CAMREADY: Cite simsa thesis
-while the locks themselves can be verified separately.
+while the locks themselves can be verified separately \cite{dbug-thesis}.
 %Otherwise, %with testing limited to one execution,

### 2016-08-17 oopsla-16/soundness.tex
change formatting according to camready author package
take care of some TODO CAMREADies

@@
 
-\subsection{Convergence to total verification}
+\subsection{Convergence to Total Verification}
 \label{sec:totalverif}
@@
 
-\subsection{Suppressing ``malloc-recycle'' false positives}
+\subsection{Suppressing ``Malloc-Recycle'' False Positives}
 \label{sec:recycle}
@@
 \begin{tabular}{rll}
-	% TODO: Can cut struct defns for space.
 	& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
@@
 Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over full HB.
-% TODO CAMREADY: Cite the tech report proofs, which I think call out the use of limited HB in the proof more explicitly.
 }

### 2016-08-17 oopsla-16/evaluation.tex
fix more TODO CAMREADies

@@
 We fixed these by hand, before running tests, to ensure that every bug in our study required meaningful work by the MC.}.
-% TODO: Is this ok? Too arrogant?
+% Is this ok? Too arrogant?
+% I guess it's fine!
 We believe that merely finding a small handful of new real-world bugs is largely anecdotal,
@@
 	\item {\bf SSS-MC-ICB}: With PPs as above, but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster, and
-	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
+	\item {\bf SSS-MC-Shared-Mem}:
 		Using ICB+BPOR, configured to preempt on any shared memory access \cite{inspect}
@@
 
-% TODO CAMREADY
+% Sorry camready, not gonna do this. It'll have to wait for the thesis.
 %\revision{
 %%{\bf Variance.}
-%% TODO CAMREADY: Probably will need at least a 3rd run of dis.
 %Because \quicksand~can be nondeterministic in its scheduling of state spaces,
 %we ran the QS-Limited-HB and QS-Pure-HB tests a second time to quantify the variance.
-%% TODO CAMREADY
 %% We found the maximum area between any two curves to be
 %We found the area between the two curves to be
-%% TODO CAMREADY
 %{\bf XX\%}
-%of the area between QS-TODO(?)-HB and SSS-MC-Shared-Mem (the next closest),
-%% TODO cite CAMREADY
+%of the area between QS-(?)-HB and SSS-MC-Shared-Mem (the next closest),
+%% would need a citation...
 %which we consider an insignificant amount of variance.
@@
 
-% TODO: Fix up this paragraph after preempt-everywhere results come in.
 The left half of Table~\ref{tab:drbugs}
@@
 In \mxtest, in which we do not trust the lock implementation's correctness,
-% TODO CAMREADY: Fix, SSSMC-ShM finds 2 mx bugs.
 \revisionx{SSS-MC-ICB and SSS-MC-DPOR}
@@
 we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
-%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
-% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
-% (as well as on the mutex API boundaries).
-% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
+% TODO CAMREADY: Report on SSS-MC-Shared-Mem's mutex testing results.
 
@@
 
-% TODO CAMREADY: Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
+% Future work (for thesis maybe?) Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
 {\bf Finding the same bugs faster.}
@@
 
-% NB. TODO CAMREADY. If a reviewer asks, "How come SSSMC-ShM completed so many state spaces,
-% even though last section you argued that so many PPs slows it down terribly?"
-% The answer is, "Actually, it's not that SSSMC-ShM magically got fast again for the verification experiment;
-% it's that SSSMC-ShM and Quicksand are *BOTH* slow to achieve full completion.
-% Quicksand can get bogged down in a combinatorial explosion of different subsets of PPs,
-% having a zillion jobs on its hands. Future work should improve this with https://github.com/bblum/landslide/issues/216 ."
-
 %\quicksand's total verifications
@@
 
-% TODO: Update for pure HB (need a re-run of a subset of completed sssmc tests to confirm they still have no pure hb drs, even if they had limited hb drs)
 Among \revisionx{QS-Limited-HB's} verified tests, 36 contained no data-race candidates whatsoever,
@@
 \revisionx{Finally, assuming sequentially-consistent hardware,
-QS-Pure-HB classified many % TODO: how many
+QS-Pure-HB classified many
 true data races as benign, %(non-failing),
@@
 %although that would still not simultaneously be able to provide total verifications.
-% TODO: read gambit paper
 

### 2016-08-17 oopsla-16/evaluation.tex
fix some last numbers in eval regarding expt reruns

@@
 While recent work \cite{randomized-scheduler} attempts to quantify the probability
-that an untested interleaving would expose a bug,
+\revisionz{that a bug remains in some untested interleaving},
+%that an untested interleaving would expose a bug,
 \quicksand~instead reports which subsets of PPs resulted in state spaces that did complete in time.
@@
 %but can at least report
-On \revisionx{310} tests, \revisionx{SSS-MC-ICB} timed out after 10 hours with no bugs found.
-Among these tests, \quicksand~found bugs in \revisionx{47}.
-For the other \revisionx{263}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
+On \revisionz{229} tests, \revisionx{SSS-MC-ICB} timed out after 10 hours with no bugs found.
+Among these tests, \quicksand~found bugs in \revisionz{37}.
+For the other \revisionz{192}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
 %and the number of unique PPs among them.
@@
 which includes the former type as well as new bugs involving nondeterministic data-races.
-\revisionz{QS-LHB found 69 data-race bugs in total, 54} of which could be found with single-pass data-race candidates alone.
+\revisionz{QS-LHB found 69 data-race bugs in total, 15} of which could \revisionz{not} be found with single-pass data-race candidates alone.
 \revisionz{QS-PHB is even more dependent on nondeterministic data-race PPs, requiring nondeterministic data-race PPs in 21 cases among its 44 total data-race bugs.}

### 2016-08-17 sigcse-17/410.tex
sigcse: import text on 410 from edupar, with some violence

\section{\fourten}
\label{sec:410}

\fourten, Operating System Design and Implementation,
is an upper-level undergraduate class
designed for juniors and seniors in Computer Science
and Electrical and Computer Engineering.
For CS students it is one of five courses which fulfill
a ``systems programming'' degree requirement.
%
The prerequisite course is \twothirteen,
Introduction to Computer Systems,
% TODO CAMREADY
%a course which has been adopted at other
%universities in the U.S.\ and worldwide
adapted from Carnegie Mellon University's 15-213
\cite{sigcse01:CSaPP}.
This section discusses the course learning objectives,
outlines the programming projects,
and mentions some current limitations in the scope
of the course material.

The course has many learning objectives,
ranging from acquiring detailed factual knowledge about
hardware features
through practicing advanced cognitive processes
such as open-ended design.
%%
%Students study high-level concepts
%such as
%\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
%file-system internals,
%and log-based storage.
%%
%Emphasis is placed on acquiring information from ``primary sources,''
%including both manufacturer-provided hardware documentation
%and a non-textbook technical-literature reading assignment.
%
The course's distinguishing feature is that
students begin with a ``blank slate'' rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements from multiple sources
and must
choose their own module boundaries and inter-module conventions.
%
%Due to the foundational nature of kernel code,
The assignment design and grading encourage students to
think about corner cases, including resource exhaustion,
instead of being satisfied by
``the right basic idea''
implementations that handle only auspicious situations.
%
Finally, most relevant to this work,
students %gain substantial experience in
analyze and write lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread
%concurrency
and thread/interrupt concurrency.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Projects}

In the course of a semester, students work on five
programming assignments; the first two are individual,
and the remaining three are the products of two-person
teams.

\subsubsection{Stack Crawler}

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces students to our expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project is built using a standard C tool chain
and debugged as students see fit (generally \x{printf()} and GDB).

\subsubsection{Device Drivers and Game}

The second project is a simple game, such as Hangman,
which runs without an underlying
operating system.
We provide students a small C library %with selected C library routines
(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()}),
who then write a device-driver library consisting of
console output (to a memory-mapped text-mode video display),
keyboard input (interrupt-driven),
and a countdown timer.
The game enables students to test their drivers.
%
% and the need to write enough code to get into trouble
%

This project and the remaining ones are written in
C with a bit of x86-32 assembly code,
which is then
compiled and linked into an ELF executable,
stored into a 1.44-megabyte 3.5-inch floppy-disk image,
and booted via GRUB.
If the image is copied to a real floppy or CD,
%or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can be booted on standard PC hardware.
However, students most often use
the Simics system emulator \cite{simics} (described below).

\subsubsection{User-space Thread Library}

The third project is implementing a 1:1 thread library for
user-space programs,
%essentially a stripped-down
%version of POSIX Pthreads.
similar to POSIX Pthreads.
The primary goal is
for students to experiment
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose.
%(there is substantial variation across groups).
They then write other synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.

A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
There is a semantic gap
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and \x{deschedule()} (the system call
which suspends execution of a thread).

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO: Anonymize this citation
in a twelve-page document~\cite{kspec}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

\subsubsection{\pebbles Kernel}

For the fourth project, students %two-student teams
produce a kernel which implements the
same specification as the reference kernel
they previously relied on.
They design and implement some approach to
synchronizing and blocking threads while
they are in kernel space,
%\shortversion{,}{\footnote{
%Student designs for their kernel's synchronization primitives vary widely.
%For example, ``top of the class'' students who went on to become teaching assistants have alternately used condition variables but not semaphores, and semaphores but not condition variables.
%},}
a simple round-robin scheduler,
basic virtual memory,
a program loader,
%various x86 exception handlers,
and code for setting up and tearing down
threads and processes
(re-using their game-project device drivers).
These ingredients are combined to yield
twenty-five system calls,
which we list in Table~\ref{tab:syscalls}.
%(life-cycle, thread management, memory management,
%console I/O, and miscellaneous),
%which we discuss further in Section~\ref{sec:pebbles}.}

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from kbd input. \\
		\x{readline} & Reads the next line from kbd input. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}

%%% Justify by mentioning average lines of code. -- bblum
%For most students in the class, this is the
%largest and most complicated software artifact they
%have produced.
%Because the test suite and the grading criteria
%emphasize robustness and preemptibility of
%kernel code,
%there are many cross-cutting concerns.
%%
%Because the students are responsible for ensuring
%the run-time invariants underlying all compiler-generated
%code in the system (kernel and user-space),
%they gain experience with debugging at both the
%algorithm level and the register/bit-field level.

{\bf Lifecycle.} The process lifecycle is loosely based on the
Unix \x{fork()}/\x{exec()}/\x{wait()} model.
We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
distinction between ``tasks,''
which are resource containers,
and threads,
each of which executes within a single task.
This differs from %partitioning is a different approach from that of
Plan~9's \x{rfork()}~\cite{Pike90plan9} or Linux's \x{clone()},
%both of
which allow a continuum of sharing
%between identical threads and independent processes,
at the cost of implementation complexity. %which we wish to avoid.
%A task's address space can be edited through two system
%calls that add and delete contiguous regions of memory.
%Students are expected to implement zero-fill-on-demand,
%but copy-on-write and demand-loading of executables are
%optional.
%Threads can print text to and read lines of text from
%the simple console that students implemented at
%the start of the semester.

Widely regarded as the most difficult concurrency problem in the project
is that of coordinating a parent and a child task that ``simultaneously''
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

{\bf Threads.}
%Within a task,
Threads are created via
\x{thread_fork}, which closely
mirrors the specification of \x{fork()}:
the new thread begins with identical %an exact copy of the
register values,
%of the old thread, except that
%the return-value register, \x{\%eax}, contains
%zero;
excepting \x{\%eax}, containing zero;
in the old thread, \x{\%eax} holds the thread ID
of the new thread.
This minimal formulation defers all policy decisions
%(e.g., stack size, location, and contents)
to the
user-space thread library.
A thread may suspend its execution via the
\x{deschedule()} system call,
and a blocked thread can be reactivated via
\x{make_runnable()}.

Each thread
can register an exception handler
%has the ability to register a handler
%for hardware exceptions (illegal instruction,
%division by zero, etc.)
via \x{swexn()},
which specifies the addresses of handler code
and a handler stack.
%When a thread encounters an exception,
The kernel reflects exceptions to user space
by
de-registering the handler,
pushing the thread's register state onto the handler stack,
and invoking the handler code.
The handler
%has the ability to
can re-register itself
(or some other handler),
run arbitrary code,
and/or atomically adopt a new set of register
values.
%The \x{swexn()} facility is used to enable
\x{swexn()} enables
the thread library to detect %when a managed
thread crashes
%thread has crashed
and %, more importantly,
to enable user-space automatic stack growth.
%implementation of automatically-growing stack
%regions.
%Removing the auto-stack policy and mechanism from
%its traditional position in Unix-like kernels
%simplfies the kernel and
%front-loads policy questions to an earlier
%part of the course,
%before students are facing the substantial pressure
%of the kernel project.

\subsubsection{\pebbles Extension}

Students who complete the kernel project on time
then work on a kernel-extension project.
Past projects have included
a sound card driver,
a filesystem,
suspend-to-disk,
profiling,
and an in-kernel debugger.
Two recent, more aggressive, projects have been
paravirtualization
%so that their kernels can host guest kernels and
and multi-processor support.
%to their single-processor kernels.

In the multi-processor project, students
%extended their kernels to
scheduled multiple threads across up to eight cores.
We provided base code for determining the CPU count, booting the application processors, and sending inter-processor interrupts (IPIs).
% Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
% to implement moving threads between processors,
% and
% to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
Students must move threads between processors,
synchronize memory unmapping across multiple TLBs,
%and account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
and find a way to prevent concurrent data access apart from disabling local CPU interrupts.
%
%To allow struggling groups to make progress without solving every synchronization challenge,
To help struggling groups,
we allowed
%the use of
a ``big kernel lock'', and rewarded groups according to how little they used it.
Submissions were also graded according to
correctness of the scheduler and spinlocks,
stability on a suite of stress tests,
and efficient use of IPIs. %for efficient processor coordination.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Simics}

The main execution and debugging
platform used in the class is
Wind River Simics~\cite{simics}.
%originally developed by the Swedish Institute of
%Computer Science and
%currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for correct code %that behaves correctly
but also for kernels that accidentally ``abuse'' hardware.
%{but also for code that is ``incorrect within reason.''
%Occasionally student code manages to invoke an unsupported
%function of some hardware device or even to cause the
%simulator to crash, but this is quite rare, and in almost
%every case when the ``bad event'' is characterized it turns
%out that what the software requested from the hardware was
%a mistake.
%Meanwhile, inside its large envelope, Simics supports
%\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86
%hardware-defined
descriptor tables.
%Unlike some symbolic debuggers,
%Simics is scriptable in Python,
%and it is possible to interpose user-written measurement
%frameworks to evaluate proposed hardware
%features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
%\shortversion{}{Performance is very reasonable (Simics can simulate certain
%cases, such as a CPU halted awaiting a clock interrupt,
%much faster than real time).}
%\shortversion{}{A major advantage of using Simics over the QEMU emulator in particular
%is that QEMU only issues timer interrupts at basic-block boundaries,
%which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Limitations}

At present the course accepts the following limitations
or core assumptions.

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86, for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.

\textbf{Memory model.}
At present we allow students
to reason about concurrency based on a
sequentially-consistent view of
% TODO: add a citation?
memory.
%which is consistent with
%existing x86 hardware~\cite{SewellSOZNM:x86tso-cacm10}
%but simplistic compared to most other
%architectures.
Lecture material educates students about
%the general topic of
out-of-order memory systems
and fencing,
but at present the hands-on experience does not
reinforce that material.

\textbf{Reduced breadth of implementation.}
One cost of programming to the complexities of
real hardware is that
within %the scope of
a single semester,
students do not %\violence{not}{not have the opportunity to}
implement common features of many kernels,
such as a file system, interprocess communication,
a real-time scheduler,
or networking.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Grading}

Thread libraries and kernels
are graded through a combination
of three methods:
performance on a test suite,
compliance with ``hurdle criteria,''
and detailed code review by a member of the course staff.

\subsubsection{Automated tests}

%While developing their code,
Students will uncover both bugs
and design misunderstandings by running test code we
provide.
%Some tests are simple, exercising one feature or condition
%(e.g., allocating and freeing a large amount of memory),
%while other tests exercise a mix of kernel features
%simultaneously (creating and destroying processes,
%some of which do I/O while others handle or are killed
%by various exceptions) over an extended period of time.
The test suite is a number of user-space programs
which test
%functionality including
basic thread and process lifecycles, robustness
against corner-case security vulnerabilities, and long-term stability
under large stress tests.
%
%When students submit their thread libraries and kernels,
%we use an automated testing harness to
%run a superset of the public test suite.
%
Of course, due to the nondeterministic nature of many bugs in student code,
these tests can provide only a rough estimate of a submission's correctness,
rather than verifying it with certainty.
(This hit-or-miss nature motivated our work on Landslide.)
%Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.

\subsubsection{Hurdle completion criteria}

Because of the ambitious learning objectives,
we employ a very detailed notion of completeness
for the thread library and kernel projects.
The ``hurdle'' model is based on the following observations:

% TODO: Figure out how to left-justify these bullet poince.
\begin{itemize}
	\item Students learn more from bugs
		that are harder to find.
		In extreme cases, code may do
		almost everything right, but the last bug may
		demonstrate a deep misunderstanding which requires a re-design.

	\item Because concurrency bugs are triggered rarely,
		finding them requires code capable of running
		for a long period without crashing.

	\item Robust code often requires a different structure
		than a ``usually works'' solution.
		It is easy to %write task-exit code which
		implement \x{vanish()} by
		allocating memory to notify the parent task.
		%parent that it has exited.
		However, such a design suffers from ``terminal irony'':
		a program wishing to exit,
		which would free up memory,
		cannot do so if kernel memory has been exhausted.
\end{itemize}

%For these reasons, \shortversion{much}{we believe much} of the learning
%associated with the kernel project necessarily occurs
%at the end of the experience,
%so it is important for students to finish the process.
% [suppressed: bad to build P4 on a shaky P3]
%\shortversion{It}{We believe it} would be particularly
It would be detrimental to
assign partial credit to a large body of non-working
code.
Instead, we require student thread libraries and kernels to pass a
specified percentage of tests
%each segment of the test suite
to be considered complete.
Furthermore,
because the test suite cannot detect certain
structural shortcuts (e.g., ``fixing'' a concurrency
problem by disabling interrupts during \textit{all}
kernel execution),
we require students to certify, in writing, that
their code possesses certain properties.

% TODO: Talk about P2 hurdle

A ``complete'' kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain distasteful x86-specific code structures.

%Students who pass the completeness hurdle submit their kernels
%and can move on to the kernel-extension project.
%\shortversion{Students}{Meanwhile, students} who were unable to meet the
%hurdle criteria on time receive additional time
%to finish up and improve their code quality.
%%---this
%%requirement is based on our belief that important
%%learning experiences are obtained by solving the
%%hard problems,
%%which turn up at the end of the kernel implementation.
%The hurdle criteria constitute 50\% of the kernel project's grade.
%

\subsubsection{Code review and interview}

Finally, both to make up for the tests' incompleteness and to provide personalized human feedback,
each student submission is manually graded by a member of course staff.
The instructors and teaching assistants print each group's submitted code and mark it with red ink, auditing it for style, architectural issues, races, and preemptibility.
Furthermore, after the kernel project, each group meets with their grader for a ``kernel interview'', essentially a debriefing for correcting any misconceptions.
% TODO: and what for p2?
Staff discretionary points constitute 20\% of the kernel project's grade.

### 2016-08-18 sigcse-17/410.tex
sigcse: save some space

@@
 This project and the remaining ones are written in
-C with a bit of x86-32 assembly code,
+C with some x86-32 assembly,
 which is then
 compiled and linked into an ELF executable,
-stored into a 1.44-megabyte 3.5-inch floppy-disk image,
+stored into a %1.44-megabyte 3.5-inch
+floppy-disk image,
 and booted via GRUB.
@@
 %or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
-it can be booted on standard PC hardware.
+it can boot on standard PC hardware.
 However, students most often use
-the Simics system emulator \cite{simics} (described below).
+Simics \cite{simics} (described in \sect{\ref{sec:simics}}).
 
@@
 \subsection{Simics}
+\label{sec:simics}
 
@@
 
-At present the course accepts the following limitations
-or core assumptions.
+%At present the course accepts the following limitations
+%or core assumptions.
 
@@
 
-\subsubsection{Automated tests}
-
+\textbf{Automated tests.}
 %While developing their code,
@@
 
-\subsubsection{Hurdle completion criteria}
-
+\textbf{Hurdle completion criteria.}
 Because of the ambitious learning objectives,
@@
 		that are harder to find.
-		In extreme cases, code may do
-		almost everything right, but the last bug may
+		In extreme cases, %code may do
+		%almost everything right, but the last bug may
+		subtle bugs may
 		demonstrate a deep misunderstanding which requires a re-design.
 
-	\item Because concurrency bugs are triggered rarely,
-		finding them requires code capable of running
+	\item %Because concurrency bugs are triggered rarely,
+		Finding rare concurrency bugs requires code capable of running
 		for a long period without crashing.
@@
 		implement \x{vanish()} by
-		allocating memory to notify the parent task.
+		allocating memory. %to notify the parent task.
 		%parent that it has exited.
@@
 		which would free up memory,
-		cannot do so if kernel memory has been exhausted.
+		cannot do so if memory is exhausted.
 \end{itemize}
@@
 
-\subsubsection{Code review and interview}
-
+\textbf{Code review and interview.}
 Finally, both to make up for the tests' incompleteness and to provide personalized human feedback,

### 2016-08-19 sigcse-17/landslide.tex
sigcse: write about landlside

\section{Landslide}
\label{sec:landslide}

Motivated by the stress tests' inconsistency at finding concurrency bugs,
% TODO camready
we have built Landslide%\cite{landslide}
, a stateless model checker for \pebbles kernels and thread libraries.
Stateless model checking \cite{verisoft} offers an alternative
%concurrency testing approach
to stress testing,
in which the testing framework controls scheduling to force a new thread interleaving to execute in each iteration of a test.
It follows in the footsteps of related tools such as CHESS \cite{chess},
but also provides several important features for use in an educational setting,
such as heuristics to handle unusual student solutions, kernel-level testing, and user-friendly progress reports and HTML debugging traces.

\subsection{Design}

Conceptually,
Landslide is designed as follows.
Through several annotations in the kernel source, Landslide tracks
%the scheduler state to know
which threads are running or runnable.
(For testing thread libraries, these annotations are already provided in the reference kernel binary;
for testing kernels, the student must add them by hand.)
When execution reaches certain interesting {\em preemption points},
Landslide triggers artificial clock interrupts
to force the scheduler to run a different thread.
When a test finishes execution according
to one pattern of thread switches,
Landslide rewinds the simulation state
and resumes the test according to a different interleaving.
After each instruction,
Landslide applies several bug-detection predicates to
the program state to detect
illegal heap accesses,
deadlock, infinite loops, and panics.

{\bf Example.}
Figure \ref{fig:paradise}(a) shows an example concurrency bug which we will use as a running example.
The code implements a producer/consumer program, in which one thread sends data to the other through a shared list,
synchronized with a mutex and condition variable.
However, because the consumer code only checks the list size once before blocking,
multiple simultaneous consumers can race,
causing a segfault when one receives a null \x{data} pointer. %and tries to dereference it

{\bf State space exploration.}
To trigger the interleaving which exposes this bug, threads must be preempted at the boundaries of the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
Landslide automatically considers these boundaries to be {\em preemption points}: candidate sites for timer-driven thread switches.
It can also add preemption points dynamically,
as identified by an integrated data-race analysis \cite{tsan,fasttrack},
to expose shared-memory bugs when synchronization is absent.

The branching possibilities arising from each preemption point produce an execution tree, or state space,
as shown in Figure \ref{fig:paradise}(b).
%As these state spaces are exponentially-sized,
Landslide uses Dynamic Partial Order Reduction \cite{dpor} to identify and prune equivalent interleavings
to mitigate the exponential explosion of these state spaces.
% TODO make sure this gets mentioned later, if not here
%and state space estimation \cite{estimation} to predict which
% TODO CAMREADY: cite oopsla
% In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.

\subsection{Identifying Bugs}

Landslide checks for bugs in several ways.
It can identify assertion failures, use-after-free and other invalid heap accesses, and
deadlocks, and also includes a heuristic check for infinite loops and livelock.
%Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
%In contrast with analyses that directly detect data races~\cite{tsan},
Although Landslide employs a data-race analysis to decide where to preempt,
it does not directly classify suspicious memory accesses as bugs,
but rather seeks concrete, observable failures that can arise from many different types of errors.

% TODO: make fig trace

When Landslide identifies a bug, it outputs an HTML-formatted
{\em preemption trace}.
Figure~\ref{fig:trace} shows a typical output for the example bug in Figure \ref{fig:paradise}.
The trace reports what kind of bug was detected
and each preemption point in the current interleaving:
which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
Hence, students can bypass tediously reconstructing the schedule using \x{printf},
and hopefully better understand the concurrent execution that exposed the bug,

\subsection{Heuristics}

TODO talk about heuristics

### 2016-08-22 oopsla-16/paper.tex
take care of some TODO CAMREADies
oopsla: add sully and carlo to acks

@@
 
-% TODO CAMREADY: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
 \newcommand\landslide{\textsc{Landslide}}
@@
 Thanks to
-Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
+Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Michael J. Sullivan, Carlo Angiuli, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
 the anonymous OOPSLA reviewers for their helpful comments.

### 2016-08-22 oopsla-16/evaluation.tex
Add citations/urls for supplementary material (and fix some outdated numbers)

@@
 The source code of all 9 test cases is available at
-{\em [submitted as supplementary material]}.
-For all tests, we used {\tt without\_function} to blacklist PPs on {\tt malloc}'s internal lock.
+\url{https://github.com/bblum/oopsla-dataset}.
+For all tests,
+%we used \revisionz{preemption sealing (\sect{\ref{sec:landslide}})}
+%{\tt without\_function}
+%to blacklist PPs on {\tt malloc}'s internal lock.
+\revisionz{we blacklisted PPs on {\tt malloc}'s internal lock
+using \landslide's {\tt without\_function} command (\sect{\ref{sec:landslide}}).}
 In total, our evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
 % at least 157 of which could expose bugs.
-at least 173 of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
+at least \revisionz{181} of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
 
@@
 we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.
-% TODO CAMREADY: Report on SSS-MC-Shared-Mem's mutex testing results.
 

### 2016-08-22 oopsla-16/soundness.tex
Add citations/urls for supplementary material (and fix some outdated numbers)

@@
 In this section we present two theorems concerning Iterative Deepening's correctness.
-Our full proofs
-{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
+Our full proofs,
+available at \cite{quicksand-soundness},
+%{\em [submitted as supplementary material; will be cited as a tech report in the final version of the paper]}
 discuss our assumptions explicitly and include more formal definitions and structure.

### 2016-08-23 sigcse-17/intro.tex
sigcse: import text on 410 from edupar, with some violence
sigcse: add some pretty pictures

@@
 
-% TODO: lol.
-Building real-world multithreaded concurrent programming systems is well-known to be notoriously difficult to design, implement, test, debug, hard to get right, challenging, and error-prone arguably because of the difficulties of finding, reproducing, and fixing concurrency errors, attributable to the astronomically large number of thread interleavings, which arise at random during normal program execution.
+%Building real-world multithreaded concurrent programming systems is well-known to be notoriously difficult to design, implement, test, debug, hard to get right, challenging, and error-prone arguably because of the difficulties of finding, reproducing, and fixing concurrency errors, attributable to the astronomically large number of thread interleavings, which arise at random during normal program execution.
 
@@
 %to force kernel threads to take execution paths that expose concurrency bugs.
-\landslide can test student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
+\landslide tests student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
 We present our experience sharing \landslide with volunteering students who used it to find bugs in their own code,

### 2016-08-23 sigcse-17/landslide.tex
sigcse: add some pretty pictures

@@
 to stress testing,
-in which the testing framework controls scheduling to force a new thread interleaving to execute in each iteration of a test.
+in which the testing framework controls scheduling to force execution of a new thread interleaving in each iteration of a test.
 It follows in the footsteps of related tools such as CHESS \cite{chess},
@@
 {\bf Example.}
-Figure \ref{fig:paradise}(a) shows an example concurrency bug which we will use as a running example.
-The code implements a producer/consumer program, in which one thread sends data to the other through a shared list,
+Figure \ref{fig:paradise}(a) shows an example concurrency bug. %which we will use as a running example.
+The code implements a producer/consumer program, in which one thread sends work to the other through a shared list,
 synchronized with a mutex and condition variable.
-However, because the consumer code only checks the list size once before blocking,
-multiple simultaneous consumers can race,
-causing a segfault when one receives a null \x{data} pointer. %and tries to dereference it
+However, because the consumer code only checks once whether work exists before blocking,
+%multiple
+simultaneous consumers suffer
+%as shown in Figure \ref{fig:paradise}(b),
+a segfault when one receives a null \x{work} pointer. %and tries to dereference it
+
+\definecolor{thread1}{RGB}{87,172,255}
+\definecolor{thread2}{RGB}{255,201,102}
+\definecolor{thread3}{RGB}{255,128,160}
+
+\begin{figure}[t]
+	\begin{tabular}{l|l}
+	Consumer thread & Producer thread \\
+	\hline
+\begin{lstlisting}
+mutex_lock(mx);
+if (!work_exists())
+  cond_wait(cvar,mx);
+work = dequeue();
+mutex_unlock(mx);
+access(work->data);
+\end{lstlisting}
+	&
+\begin{lstlisting}
+mutex_lock(mx);
+
+enqueue(work);
+signal(cvar);
+
+mutex_unlock(mx);
+\end{lstlisting}
+	\end{tabular}
+	{\bf (a}) A buggy producer/consumer program, in which the consumer thread only checks \x{work_exists()} once.
+	\vspace{1em}
+	\\
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%	\begin{tabular}{l|l|l}
+%	\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
+%	\hline
+%\begin{lstlisting}
+%lock();
+%if (!work())
+%  wait();
+%\end{lstlisting}
+%& & \\
+%&
+%\begin{lstlisting}
+%lock();
+%enqueue();
+%signal();
+%unlock();
+%\end{lstlisting}
+%& \\
+%& &
+%\begin{lstlisting}
+%lock();
+%dequeue();
+%unlock();
+%\end{lstlisting}
+%\\
+%\begin{lstlisting}
+%dequeue();
+%unlock();
+%// SIGSEGV
+%\end{lstlisting}
+%& &
+%	\begin{center}
+%	{\bf (b)} Interleaving with two consumer threads. The second bypasses \x{cond_wait()}, causing the first to receive \x{NULL}. %in which one receives \x{NULL}.
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\includegraphics[width=0.48\textwidth]{execution-tree.pdf}
+	{\bf (b)} The state space of all interleavings forms an execution tree. %The interleaving in (b) is the rightmost branch.
+	The bug is observed when the second consumer thread bypasses \x{cond_wait()}, causing the first to receive \x{NULL}.
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\caption{Example concurrency bug.}
+	\label{fig:paradise}
+\end{figure}
 
@@
 %As these state spaces are exponentially-sized,
-Landslide uses Dynamic Partial Order Reduction \cite{dpor} to identify and prune equivalent interleavings
+Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
 to mitigate the exponential explosion of these state spaces.
@@
 % In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.
+This reduction requires only tracing the memory accesses of each thread,
+so we are able to perform it behind-the-scenes, without consuming any student attention.
 
@@
 
-% TODO: make fig trace
-
 When Landslide identifies a bug, it outputs an HTML-formatted
@@
 
+\newcommand\hilight[2]{\color{#1}#2\color{black}\xspace}
+\definecolor{orange}{RGB}{192,96,0}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{brickred}{RGB}{192,0,0}
+\definecolor{commentblue}{RGB}{0,0,192}
+
+\begin{figure*}[t]
+	\hilight{brickred}{{\bf TID 4 was killed by a page fault! (Faulting eip: 0x1000294; addr: 0x00000000)}} (53 interleavings tested)
+	\\
+	\begin{tabular}{|l|l|l|}
+		\hline
+		\multicolumn{1}{|c|}{\bf TID 4} &
+		\multicolumn{1}{c|}{\bf TID 5} &
+		\multicolumn{1}{c|}{\bf TID 6} \\
+		\hline
+		\hline
+		0x010003d5 \hilight{olivegreen}{\x{deschedule}} (syscalls.c:7) & & \\
+		0x010011a4 \hilight{olivegreen}{\x{cond_wait}} (cond.c:21) & & \\
+		0x010001e0 \hilight{olivegreen}{\x{consumer}} (main.c:30) & & \\
+		0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) & & \\
+		\hline
+		& 0x010034f5 \hilight{olivegreen}{\x{mutex_unlock}} (mutex.c:19) & \\
+		& 0x01000037 \hilight{olivegreen}{\x{producer}} (main.c:74) & \\
+		& 0x01000100 \hilight{olivegreen}{\x{main}} (main.c:88) & \\
+		\hline
+		& & 0x010034f5 \hilight{olivegreen}{\x{mutex_unlock}} (mutex.c:19) \\
+		& & 0x0100020a \hilight{olivegreen}{\x{consumer}} (main.c:33) \\
+		& & 0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) \\
+		\hline
+		0x01000294 \hilight{olivegreen}{\x{access}} (main.c:17) & & \\
+		0x01000216 \hilight{olivegreen}{\x{consumer}} (main.c:34) & & \\
+		0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) & & \\
+		\hline
+	\end{tabular}
+	\caption{Example preemption trace provided by Landslide as debugging output for the bug in Figure~\ref{fig:paradise}.}
+	\label{fig:trace}
+\end{figure*}
+
+\subsection{Instrumentation}
+
+For the thread library project, Landslide instruments student implementations automatically.
+We provide a custom reference kernel binary equipped with the kernel instrumentations (described below)
+so Landslide can track threads' lifecycles. %the lifecycle of threads.
+The userspace \x{malloc} and \x{assert} are also provided in the basecode, and tracked similarly.
+Landslide's data-race analysis must understand the locking discipline,
+so it searches the student's code for functions named \x{mutex_lock} and \x{mutex_unlock} (as mandated by the assignment handout),
+and assumes they at least provide mutual exclusion\footnote{
+	To relax this assumption, we provide a special test case for mutexes,
+	in which Landslide may preempt and search for data races in the mutex implementation itself.
+}.
+All other thread library components are treated as unknown, and subject to preemption anywhere.
+
+To test \pebbles kernels directly, they must be annotated to inform Landslide of certain important concurrency events.
+Because the kernel project leaves internal code structure as an open-ended design decision,
+Landslide cannot search for specific function names. % like it does for P2s.
+Instead, we provide a set of functions named with the prefix \x{tell_landslide}
+to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
+Users must also list all core scheduler, context switch, and timer handling routines, so Landslide can ignore their memory accesses for the sake of DPOR
+(otherwise, threads would always conflict with each other and DPOR would achieve no reduction).
+
 \subsection{Heuristics}

### 2016-08-23 sigcse-17/landslide.tex
sigcse: talk about heuristix

@@
 
-TODO talk about heuristics
+Student submissions may behave in extraordinary ways which violate the assumptions of many modern model checkers.
+Landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
+but it must support testing any legal, yet distasteful, solution.
+%Hence, we added several heuristic
+For example, students may use open-coded \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
+To avoid getting stuck in a cyclic (infinite) state space, Landslide keeps a ``yield loop'' counter for each thread;
+when a thread \x{yield}s 10 times without otherwise making progress, Landslide marks it blocked.
+%
+In extreme cases, this may erroneously seem like deadlock when all threads \x{yield} many times coincidentally.
+To avoid wasting student time with false alarms, % the ultimate sin
+Landslide will reset all blocked threads' yield-loop counters,
+retrying potentially up to 128 times before reporting a true deadlock.
+
+Students may also break abstraction by open-coding a \x{mutex_unlock} sequence.
+Landslide must track lock operations for accurate data-race analysis,
+so to catch these patterns, it
+%learns the size of the \x{mutex} struct, then
+checks accesses against stored lock addresses blocked on by other threads.
+Similarly, these rogue unlock accesses may unblock a yield-looping thread.

### 2016-08-23 sigcse-17/landslide.tex
sigcse: shorten some things and lengthen others

@@
 It can also add preemption points dynamically,
-as identified by an integrated data-race analysis \cite{tsan,fasttrack},
+as identified by an integrated data-race analysis \cite{hybriddatarace,fasttrack},
 to expose shared-memory bugs when synchronization is absent.
@@
 In extreme cases, this may erroneously seem like deadlock when all threads \x{yield} many times coincidentally.
-To avoid wasting student time with false alarms, % the ultimate sin
+To avoid wasting precious student attention with false alarms, % the ultimate sin
 Landslide will reset all blocked threads' yield-loop counters,
@@
 Similarly, these rogue unlock accesses may unblock a yield-looping thread.
+These heuristics have enabled Landslide to test over 100 student thread libraries from the past 5 semesters fully automatically.

### 2016-08-23 oopsla-16/overview.tex
take care of some TODO CAMREADies
fix more TODO CAMREADies
oopsla: i don't think i'm actually going to fix figure 2

@@
 
+%\begin{figure}[t]
+%	\begin{tabular}{c}
+%	\includegraphics[width=0.42\textwidth]{tree-maximal-only.pdf}
+%	\end{tabular}
+%	\caption{\revisionz{The state space of possible interleavings is represented as a tree,
+%	which stateless model checkers test exhaustively.
+%	This state space arises when preempting on synchronization APIs used by the program in Figure~\ref{fig:example};
+%	other state spaces are possible with more, or fewer, preemption points.}}
+%	\label{fig:tree-maximal}
+%\end{figure}
+
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{trees-v2-squashed.pdf}
+	\includegraphics[width=0.48\textwidth]{trees.pdf}
 	\caption{Iterative Deepening example.
-		The minimal state space (A) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
-		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (B), {\tt mutex\_unlock} alone (C), or both together (D).
+		The minimal state space (a) includes only voluntary thread switches, such as {\tt yield()}. %or {\tt cond\_wait()}.
+		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (b), {\tt mutex\_unlock} alone (c), or both together (d).
 Each option increases the state space size unpredictably, so multiple state spaces should be tested in parallel.
@@
 Figure~\ref{fig:id} shows several possible resulting state spaces.
-The approach of prior work is to enable all preemption points simultaneously, i.e., to test only the state space denoted (D).
+The approach of prior work is to enable all preemption points simultaneously, i.e., to test only the state space denoted (d).
 
@@
 %Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})
-%% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
 %Hence, we employ Limited HB in this paper},

### 2016-08-23 oopsla-16/proof/di.tex
prepare proof for submission and begin writing system model

@@
 \section{Definitions}
+
+\subsection{System Model}
+
+To reason about programs at the executable level, we will define them as a list of assembly-like instructions,
+including certain concurrency primitives which we evaluate atomically, assuming the runtime provides correct implementations for them.
+\begin{eqnarray*}
+	\mathcal{P} &::=& [\mathcal{I}] \\
+	\mathcal{I} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg}(a,v) \\
+	&|& a \leftarrow \mathsf{malloc}(n) \quad | \quad \mathsf{free}(a) \\
+	&|& \mathsf{mutex\_lock}(m) \quad | \quad \mathsf{mutex\_unlock}(m) \\
+	&|& \mathsf{deschedule} \quad | \quad \mathsf{make\_runnable}(t) \\
+	&|& v \leftarrow \mathsf{thread\_fork} \quad | \quad \mathsf{thread\_exit} \\
+	&|& \mathsf{local}
+\end{eqnarray*}
+
+For the sake of brevity, we will not explicitly define execution semantics, though they are straightforward:
+The execution state consists of any number of threads, each represented by a list of instructions to execute ($[\mathcal{I}]$),
+a stack of value maps of thread-local variables ($v$),
+and an indicator of whether the thread is runnable.
+The state also includes a map of global memory, indexed by addresses denoted $a$,
+and a set of mutex objects, denoted $m$.
+To execute one step, arbitrarily choose one runnable thread, and evaluate the first element of its instruction list.
+This gives rise to thread concurrency which can interleave at instruction granularity,
+although it assumes a sequentially-consistent memory model \cite{tsopso}.
+\begin{eqnarray*}
+	\mathcal{E} &::=& [\mathcal{T}] \times [a \rightarrow \mathsf{int}] \times [m \rightarrow \mathsf{bool}] \\
+	\mathcal{T} &::=& [\mathcal{I}] \times [v \rightarrow \mathsf{int}] \times \mathsf{bool}
+\end{eqnarray*}
+
+{\bf Memory.} $\mathsf{read}$, $\mathsf{write}$, and $\mathsf{xchg}$ access global or heap memory shared by all threads, indicated by some address $a$, and copying and/or modifying a thread-local variable $v$.
+% cmpxchg, xadd, etc omitted for brevity
+$\mathsf{malloc}$ and $\mathsf{free}$ provide access to fresh memory accessible by all threads. % free is irrelevant to the system model in terms of expressive power
+
+{\bf Threads.} $\mathsf{mutex\_lock}$ and
+$\mathsf{mutex\_unlock}$ provide mutual exclusion:
+a thread cannot evaluate $\mathsf{mutex\_lock}$ until that lock is released.
+$\mathsf{deschedule}$ and $\mathsf{make\_runnable}$ allow threads to manipulate their runnability,
+and $\mathsf{thread\_fork}$ and $\mathsf{thread\_exit}$ allow creation and destruction of new threads.
+$\mathsf{thread\_fork}$ is defined in the Pebbles manner \cite{kspec};
+we omit higher-level abstractions such as $\mathsf{create}$ or $\mathsf{join}$, which can be implemented using these primitives \cite{thrlib}.
+
+{\bf Local state.}
+$\mathsf{local}$ represents any thread-local instruction, such as flow control, function definition/calling, and modifying local variables.
+For brevity, we omit a detailed list, although note that $\mathsf{call}$ would be implemented by modifying the instruction stream $[\mathcal{I}]$ and creating a new frame on the stack of local variables.
 

### 2016-08-23 oopsla-16/proof/proof.tex
Import proof
prepare proof for submission and begin writing system model

--- 2015-11-20 pldi-16/proof/proof.tex
@@
 %
-\documentclass[pldi]{../sigplanconf-pldi15}
+%\documentclass[pldi]{../sigplanconf-pldi15}
+\documentclass[10pt,times,numbers]{../sigplanconf}
+\usepackage[proof]{../oopsla2016}
 
@@
 In other words, when Iterative Deepening finishes generating new data-race PPs, and completes all associated state spaces,
-it serves as a full verification of {\em all possible} thread schedules of the given test case.
+it serves as a full verification of {\em all possible} schedules of the given test case.
 
@@
 
-This document is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
+	This document is supplemental material to our main paper \cite{quicksand}; we assume the reader is already familiar with our motivation and terminology.
 \end{abstract}
@@
 
-% TODO
+% TODO CAMREADY
 % We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
@@
 
-{\em removed for double blind review}
-%Thanks to Michael J. ``Sully'' Sullivan and Carlo Angiuli, who double-checked the proof and helped with the flow of explanation.
+%{\em removed for double blind review}
+Michael J. ``Sully'' Sullivan, Carlo Angiuli, and Nathaniel Filardo double-checked the proof and provided valuable feedback on notation style and the flow of explanation.
+Thanks as well to the anonymous OOPSLA reviewer \#4, whose suggestions helped improve the formality of our definitions.
+This work was supported in part by the U.S. Army Research Office under grant number W911NF0910273.
 

### 2016-08-24 oopsla-16/proof/di.tex
oopsla proofs: add system model and fix definition of MHB.

@@
 \begin{eqnarray*}
-	\mathcal{P} &::=& [\mathcal{I}] \\
+	\mathcal{P} &::=& [n, \mathcal{I}] \\
 	\mathcal{I} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg}(a,v) \\
@@
 	&|& \mathsf{deschedule} \quad | \quad \mathsf{make\_runnable}(t) \\
-	&|& v \leftarrow \mathsf{thread\_fork} \quad | \quad \mathsf{thread\_exit} \\
-	&|& \mathsf{local}
+	&|& t \leftarrow \mathsf{thread\_fork} \quad | \quad \mathsf{thread\_exit} \quad | \quad \mathsf{yield} \\
+	&|& \textit{local}
 \end{eqnarray*}
 
-For the sake of brevity, we will not explicitly define execution semantics, though they are straightforward:
 The execution state consists of any number of threads, each represented by a list of instructions to execute ($[\mathcal{I}]$),
-a stack of value maps of thread-local variables ($v$),
+a stack of variable maps for thread-local state ($v$),
 and an indicator of whether the thread is runnable.
 The state also includes a map of global memory, indexed by addresses denoted $a$,
-and a set of mutex objects, denoted $m$.
+and a set of mutex objects, denoted $m$. Thus:
+\begin{eqnarray*}
+	\mathcal{E} &::=& [\mathcal{T}] \times [a \rightarrow \mathsf{int}] \times [m \rightarrow \mathsf{bool}] \\
+	\mathcal{T} &::=& \mathcal{P} \times [[v \rightarrow \mathsf{int}]] \times \mathsf{bool}
+\end{eqnarray*}
+For the sake of brevity, we will not explicitly define execution semantics, though they are straightforward:
 To execute one step, arbitrarily choose one runnable thread, and evaluate the first element of its instruction list.
 This gives rise to thread concurrency which can interleave at instruction granularity,
-although it assumes a sequentially-consistent memory model \cite{tsopso}.
-\begin{eqnarray*}
-	\mathcal{E} &::=& [\mathcal{T}] \times [a \rightarrow \mathsf{int}] \times [m \rightarrow \mathsf{bool}] \\
-	\mathcal{T} &::=& [\mathcal{I}] \times [v \rightarrow \mathsf{int}] \times \mathsf{bool}
-\end{eqnarray*}
+although it assumes a sequentially-consistent memory model. For a treatment of DPOR with relaxed memory, we refer the reader to \cite{tsopso}.
 
@@
 a thread cannot evaluate $\mathsf{mutex\_lock}$ until that lock is released.
-$\mathsf{deschedule}$ and $\mathsf{make\_runnable}$ allow threads to manipulate their runnability,
+$\mathsf{deschedule}$ and $\mathsf{make\_runnable}$ allow threads to manipulate their own or another's runnability, respectively,
 and $\mathsf{thread\_fork}$ and $\mathsf{thread\_exit}$ allow creation and destruction of new threads.
@@
 we omit higher-level abstractions such as $\mathsf{create}$ or $\mathsf{join}$, which can be implemented using these primitives \cite{thrlib}.
+$\mathsf{yield}$ has no effect under these execution semantics, but is included for the sake of synchronization API preemption points (see below).
 
 {\bf Local state.}
-$\mathsf{local}$ represents any thread-local instruction, such as flow control, function definition/calling, and modifying local variables.
+$\textit{local}$ represents any thread-local instruction, such as modifying local variables, flow control, function definition/calling, assertions, and side effects.
 For brevity, we omit a detailed list, although note that $\mathsf{call}$ would be implemented by modifying the instruction stream $[\mathcal{I}]$ and creating a new frame on the stack of local variables.
+
+\begin{definition}[Blocked threads]
+	A thread is blocked when either its runnable flag is false, or when its next instruction is a $\mathsf{mutex\_lock}$ on an unavailable mutex.
+\end{definition}
+
+Model checkers often also include heuristics to identify threads using open-coded $\mathsf{yield}$ loops to block,
+but for the sake of our proofs, we assume such patterns are implemented more tastefully with a condition-variable-like primitive built upon $\mathsf{deschedule}$.
+
+\begin{definition}[Interleaving]
+	An interleaving is a sequence of execution steps, beginning with some initial program state and ending when all threads are blocked or have exited, and/or when a bug is observed.
+\end{definition}
 
@@
 \begin{definition}[Preemption point (PP)]
-	A PP is a code location between two instructions at which we may force threads to switch.
+	%A PP is a code location between two instructions at which we may force threads to switch.
+	A PP is a predicate on the execution state which identifies a class of instruction pairs between which we may force threads to switch.
 \end{definition}
@@
 Hence, the ``same'' PP may occur multiple times during an execution; for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
-For the sake of this proof, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
+%For the sake of this proof, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
+In our proofs, we will use the term ``PP'' to indicate specific instruction sites at which a PP predicate holds.
+
+We also use ``synchronization API PPs'' to denote the class of
+%statically-available
+PPs which occur immediately {\em after} any of
+$\mathsf{mutex\_lock}$,
+$\mathsf{mutex\_unlock}$,
+$\mathsf{deschedule}$,
+$\mathsf{make\_runnable}$,
+$\mathsf{thread\_fork}$,
+$\mathsf{thread\_exit}$,
+$\mathsf{yield}$.
+Because no other instruction affects a thread's runnability, it is always possible to execute a program by switching the currently-executing thread only at synchronization API PPs.
+
+All data-race PPs will occur immediately {\em before} a $\mathsf{read}$, $\mathsf{write}$, or $\mathsf{xchg}$.
 
 \begin{definition}[Transition]
-A sequence of instructions executed by the program between two preemption points (PPs).
+A sequence of execution steps from a program's evaluation between two preemption points (PPs).
 \label{def:transition}
+
 \end{definition}
-Our model checker provides the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-We also include a trace of all memory accesses in the representation of transitions.
+We require the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
+The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
+We also assume a trace of all memory accesses is available in the representation of transitions.
+
+\begin{definition}[State space]
+Given a set of PP predicates, a state space is
+a set of interleavings representing all possible execution sequences which switch threads only on those PPs.
+\end{definition}
 
 \begin{definition}[Must-happen-before (MHB)]
-	Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
+%Given two transitions $A$ and $B$, we say $A$ MHB $B$ if $B$ cannot be reordered to occur before $A$.
+Let $t_1$ and $t_2$ be two transitions of an interleaving, and $T1$ and $T2$ be the corresponding thread IDs,
+and let $t_1$ occur before $t_2$.
+Then $t_1$ MHB $t_2$ if
+\begin{enumerate}
+	\item $T2$ is blocked immediately preceding $t_1$ and not blocked immediately afterward,
+		and there does not occur another $t_2'$ by $T2$ between $t_1$ and $t_2$; or
+	\item there occurs some $t_3$ by thread $T3$ such that $t_1$ MHB $t_3$, $t_3$ MHB $t_2$, and $T3 \ne T2$; or
+	\item $T1 = T2$.
+\end{enumerate}
 \end{definition}
-All transitions of the same thread are trivially MHB-related.
+
+Intuitively, MHB expresses the "cannot be reordered with" (or "enables") relation between transitions.
 Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
@@
 However, as prior work has shown, they miss many other bugs involving unprotected variables accessed alternately before and after mutex-protected critical sections.
-Indeed, because most concurrent malloc implementations are protected by a lock,
-our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
+%Indeed, because most concurrent malloc implementations are protected by a lock,
+%our malloc-recycle false positives are indistinguishable from such false negatives under pure-HB.
 }.
@@
 \end{figure}
-Intuitively, $A$ MHB $B$ can be thought to mean ``$B$ cannot be reordered before $A$''.
-Note also that, while lock-protected critical sections can be reordered around each other (i.e., line 1 not MHB lines 8-9),
+
+Note also that although transitions of the same thread are related by MHB,
+MHB is transitive only when the latter two transitions are not by the same thread (condition 2).
+While lock-protected critical sections can be reordered around each other (i.e., line 1 not MHB lines 8-9),
 one cannot be reordered to be in the middle of the other (i.e, lines 3-4 MHB line 6).
-Hence, MHB is not necessarily transitive.
+%Hence, MHB is not necessarily transitive.
 In the latter case, the MHB relation is established by the mutex's blocking mechanism used during contention.
@@
 
-\begin{definition}[Interleaving]
-	An ordered list of transitions and preemption points between them.
-\end{definition}
+% Outdated. See above.
+%\begin{definition}[Interleaving]
+%	An ordered list of transitions and preemption points between them.
+%\end{definition}
 
@@
 
-\begin{definition}[State space]
-	A set of interleavings subject to the constraint that, given the preemption points used, all equivalence classes of possible interleavings are represented by at least one member.
-\end{definition}
+% Outdated. See above.
+%\begin{definition}[State space]
+%	A set of interleavings subject to the constraint that, given the preemption points used, all equivalence classes of possible interleavings are represented by at least one member.
+%\end{definition}
 
 \begin{definition}[Dynamic Partial Order Reduction (DPOR)]
-	A state-space search algorithm for stateless model checkers,
-	guaranteed to reorder transitions of two threads
-	iff they are not independent and are not related by MHB \cite{dpor}.
+	A state-space search algorithm for stateless model checkers;
+	given a state space $\mathcal{S}$, it will test at least one interleaving from each equivalence class in $\mathcal{S}$.
+	%guaranteed to reorder transitions of two threads
+	%iff they are not independent and are not related by MHB \cite{dpor}.
 	\label{def:dpor}
@@
 
-The soundness of DPOR guarantees that if a program behaviour can possibly be exposed by any thread interleaving around the given transitions/PPs,
-that interleaving will eventually be tested by reordering only such conflicting transitions.
-In other words, reordering memory-independent thread transitions cannot possibly affect program behaviour.
+Considering an interleaving $\mathcal{I}$ in $\mathcal{S}$, if two transitions $t_1$ and $t_2$ by different threads are not independent and not related by MHB, let $\mathcal{J}$ be the interleaving which reorders $t_1$ with $t_2$.
+DPOR is then guaranteed to test some interleaving in $\mathcal{S}$ equivalent to $\mathcal{J}$ \cite{dpor}.
+
+Because equivalent interleavings produce identical execution states,
+DPOR guarantees to expose all reachable execution states by testing its subset of interleavings.
+We refer to this property as {\em the soundness of DPOR}.
+
+%The soundness of DPOR guarantees that if a program behaviour can possibly be exposed by any thread interleaving around the given transitions/PPs,
+%that interleaving will eventually be tested by reordering only such conflicting transitions.
+%In other words, reordering memory-independent thread transitions cannot possibly affect program behaviour.
 

### 2016-08-24 oopsla-16/proof/convergence.tex
oopsla proof: address camreadies

--- 2015-11-20 pldi-16/proof/convergence.tex
@@
 The contrapositive statement offers more structure for an easier proof\footnote{
-For the reader who likes to avoid non-constructive proofs, % \cite{vargomax}, % TODO CAMREADY: uncomment
+For the reader who likes to avoid non-constructive proofs \cite{vargomax},
 note that we are using the constructive contrapositive direction:
@@
 until a different data race PP between two other threads is discovered.
-% TODO CAMREADY
 %as shown in Figure~\ref{fig:threethreads}.

### 2016-08-26 sigcse-17/eval.tex
sigcse: write an eval, i guess

\section{Evaluation}
\label{sec:eval}

We made \landslide available to students for the thread library project
during the Spring '15, Fall '15, and Spring '16 semesters of \fourten.
We introduced stateless model checking and gave a demo during a special lecture,
and required students to pass the thread library hurdle (\sect{\ref{sec:grading}}) to use \landslide.
Of 90 total two-student groups in those semesters, 47 tested their thread libraries with \landslide.

Whenever a student ran \landslide,
it recorded which options were used (test program and time limit),
a snapshot of the student's code,
and the result of the test (completed, timed out, bug found, or ctrl-C'ed).
%
We manually analyzed these snapshots to count how many bugs each group found
(not double-counting the same bug found repeatedly)
and how many of those were fixed before submission
(i.e., the students changed their code and re-ran the same test successfully).
%
We also distinguished {\em deterministic} bugs from concurrency bugs
by identifying whether \landslide needed to test multiple interleavings before finding the bug\footnote{
	For example, many students found deterministic use-after-free bugs,
	previously overlooked because the unit tests do not provide heap checking like \landslide does.
}.

\begin{table}[t]
	\begin{tabular}{r|cc|cc}
	%det bugs found	det bugs fixed	races fixed	races found
	& \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
	\# bugs	& found & fixed & found & fixed \\
	\hline
	0	& 27	& 32	& 15	& 23	\\
	1	& 6	& 5	& 7	& 11	\\
	2	& 9	& 6	& 12	& 8	\\
	3	& 2	& 1	& 7	& 2	\\
	4	& 1	& 1	& 3	& 1	\\
	5	& 2	& 2	& 2	& 1	\\
	11	& 	& 	& 1	& 1	\\
	\hline
	%Total groups
	%	& 47	& 47	& 47	& 47	\\
	Total bugs
		& 44	& 34	& 85	& 53	\\
	\end{tabular}
	\caption{Summary of how many bugs were found and/or fixed by how many groups.
	Each column counts how many groups found/fixed the number of bugs in each row;
	for instance, two groups found 5 concurrency bugs, one of which fixed all 5.}
	\label{tab:this-table-sucks-but-it's-the-best-i-got}
\end{table}

Table \ref{tab:this-table-sucks-but-it's-the-best-i-got} shows our results.
In total, \landslide found 44 deterministic bugs and 85 concurrency bugs.
\landslide found at least one concurrency bug for 32 groups (68\%),
24 of whom (51\%) were able to fix at least one,
verifying their update with a succesful re-run of the same test.

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{p2-distribution.pdf}
	\caption{Distribution of thread library grades.}
	\label{fig:p2-distribution}
\end{figure}
We also collected the overall thread library grades to evaluate whether \landslide helps students achieve more during the project.
Figure~\ref{fig:p2-distribution} shows the distribution of these grades.
Compared to students from the same semesters who opted not to use \landslide,
those who did scored on average 3\% higher. % that's rough, buddy.
%
%However, to account for the possible bias for opting-out students being more at the bottom of the class
%on account of not enough free time to volunteer for \landslide,
However, the latter distribution is potentially biased towards struggling students
who lack enough free time to volunteer for \landslide.
To account for this, we also checked grades from prior semesters when \landslide was not available.
This distribution is indistiguishable from the experimental group,
suggesting that \landslide's benefit is largely offset by the time investment required.
%
In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

### 2016-08-26 sigcse-17/410.tex
sigcse: add some pretty pictures
sigcse: talk about heuristix
sigcse: shorten some things and lengthen others
sigcse: fill in a bunch of TODOs and anonymized citations

@@
 A similar deliberate semantic gap exists between
-\x{cond_wait()} and \x{deschedule()} (the system call
-which suspends execution of a thread).
+\x{cond_wait()} and \x{deschedule()}.
+%(the system call which suspends execution of a thread).
 
@@
 The reference kernel's behavior is specified
-% TODO: Anonymize this citation
-in a twelve-page document~\cite{kspec}.
+% TODO CAMREADY: deanonymize
+in a twelve-page document~\cite{kspec-anonymized}.
 In addition to providing a reliable execution
@@
 %which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}
+It also provides instruction and memory access tracing and checkpointing/rewinding, both of which Landslide employs.
 
@@
 \subsection{Grading}
+\label{sec:grading}
 
@@
 
-% TODO: Figure out how to left-justify these bullet poince.
 \begin{itemize}
@@
 
-% TODO: Talk about P2 hurdle
-
-A ``complete'' kernel can boot up and run the shell,
+The thread library hurdle requires students to pass a suite of five test cases,
+which test mutex locking/unlocking and thread creation, exit, and joining in moderately stressful ways.
+%
+A hurdle-passing kernel can boot up and run the shell,
 can pass 80\% of the basic system-call sanity test,

### 2016-08-26 sigcse-17/landslide.tex
sigcse: fill in a bunch of TODOs and anonymized citations

@@
 
+Apart from which test program to run, \landslide's user interface requires just one option:
+students specify how much time they are willing to wait.
+%Considering this CPU budget,
+\landslide then uses state space estimation \cite{estimation} to predict
+% TODO CAMREADY: deanonymize
+which combinations of preemption points will produce appropriately-sized state spaces for the given CPU-time budget \cite{quicksand-anonymized}.
+If no bug is found before time runs out, \landslide displays a progress report summarizing the estimated remaining time for each state space,
+so ambitious students can decide whether to increase the time limit and retry.
+
 \subsection{Identifying Bugs}
@@
 but rather seeks concrete, observable failures that can arise from many different types of errors.
-
-When Landslide identifies a bug, it outputs an HTML-formatted
-{\em preemption trace}.
-Figure~\ref{fig:trace} shows a typical output for the example bug in Figure \ref{fig:paradise}.
-The trace reports what kind of bug was detected
-and each preemption point in the current interleaving:
-which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
-Hence, students can bypass tediously reconstructing the schedule using \x{printf},
-and hopefully better understand the concurrent execution that exposed the bug,
 
@@
 \end{figure*}
+
+When Landslide identifies a bug, it outputs an HTML-formatted
+{\em preemption trace}.
+Figure~\ref{fig:trace} shows a typical output for the example bug in Figure \ref{fig:paradise}.
+The trace reports what kind of bug was detected
+and each preemption point in the current interleaving:
+which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
+Hence, students can bypass tediously reconstructing the schedule using \x{printf},
+and hopefully better understand the concurrent execution that exposed the bug,
 

### 2016-08-26 sigcse-17/paper.tex
sigcse: some revisions
sigcse: dos2unix paper
sigcse: import text on 410 from edupar, with some violence
sigcse: save some space
sigcse: write about landlside
sigcse: add some pretty pictures
sigcse: write future work and fix some todos, fix abstract claims

@@
 \usepackage{listings}
+\usepackage{enumitem}
+%\usepackage{colortbl}
 
@@
 \newcommand\x[1]{\lstinline{#1}}
+
+\newcommand{\sect}[1]{\S #1}
+
+\setlist[itemize]{leftmargin=*}
 
@@
 \newcommand\landslide{Landslide\xspace}
-\newcommand\fourten{CS999}
+\newcommand\fourten{CS999\xspace}
+\newcommand\twothirteen{CS888\xspace}
 %\newcommand\carnegiemellon{{\em Carnegie Mellon University}\xspace}
@@
 %Although kernel-mode Landslide requires manual instrumentation, testing thread libraries is fully automatic.
-% TODO: verify these, obv. ;)
 showing that it allows them to find and fix previously-overlooked bugs.
-Further, %even struggling students can take advantage of it,
-Landslide's users achieve higher concurrency scores on average in subsequent projects,
-% FIXME awkward phrasing
-suggesting that debugging with a model checker produces better student retention of concurrency principles.
+%% no :( :(
+% Further, %even struggling students can take advantage of it,
+% Landslide's users achieve higher concurrency scores on average in subsequent projects,
+% suggesting that debugging with a model checker improves student retention of concurrency principles.
 
@@
 
-% TODO, do dis
 % The code below should be generated by the tool at
@@
 	 <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+	 <concept_id>10011007.10011006.10011073</concept_id>
+	 <concept_desc>Software and its engineering~Software maintenance tools</concept_desc>
 	 <concept_significance>500</concept_significance>
@@
 	 <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+	 <concept_id>10010147.10011777.10011778</concept_id>
+	 <concept_desc>Computing methodologies~Concurrent algorithms</concept_desc>
 	 <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
 	 </concept>
@@
 
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Software and its engineering~Software maintenance tools}
+\ccsdesc[300]{Computing methodologies~Concurrent algorithms}
 
@@
 \input{intro}
+\input{410}
+\input{landslide}
+\input{eval}
 
-\section{The {\secit Body} of The Paper}
-% TODO
-TODO body
+%\section{Conclusions and Future Work}
+\section{Future Work}
 
-\section{Conclusions}
-% TODO
-TODO conclusion
+We are presently extending \landslide to support Pintos kernels \cite{pintos},
+%used at many top CS schools including Stanford, Berkeley, and University of Chica
+another instructional kernel popular at many top CS schools.
+%a more widely-used instructional kernel than \pebbles.
+Because Pintos mandates more internal APIs than \pebbles,
+\landslide can more easily instrument student code without requiring manual annotations.
+In a preliminary experiment, we tested 79 student Pintos implementations with \landslide,
+finding 59 concurrency bugs with 10 CPU-hours allocated to each of 3 test cases \cite{quicksand-anonymized}.
+In future work, we hope to fully automate Pintos instrumentation, and offer \landslide at other schools beyond \cmu.
+
+%\fourten, the undergraduate operating systems course at \cmu, teaches concurrency in a
 

### 2016-08-26 sigcse-17/intro.tex
sigcse: shorten some things and lengthen others
sigcse: fill in a bunch of TODOs and anonymized citations
sigcse: finish off some intro todos; make a graph bigger

@@
 in which they implement threads and processes.
-Presented with a kernel specification called \pebbles%
+Presented with a kernel specification called \pebbles
 % TODO camready
-%\cite{kspec}
+\cite{kspec-anonymized}
 , students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
@@
 \pebbles focuses more on ground-up design than on teaching a broad feature set.
-At Stanford, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
+At Berkeley, Stanford, and the University of Chicago, among others, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
 In both courses, student kernels support multiprocessor execution;
@@
 %\cite{landslide}
-, a tool that uses stateless model checking \cite{verisoft,dpor,chess-icb}
+, a tool that uses stateless model checking \cite{verisoft,dpor,chess}
 as an alternative to the conventional stress tests.
@@
 \landslide tests student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
-We present our experience sharing \landslide with volunteering students who used it to find bugs in their own code,
+It also features an experimental, fully automatic, Pintos testing mode.
+We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
 and discuss the possibility for integrating it as a main component of operating systems curricula.
+
+% TODO: After submission, email the PC "Paper #1337 was submitted with an anonymized citation; here is the deanonymized version and the camera ready of the paper in case you need to confirm our claims about pintos"

### 2016-08-26 sigcse-17/410.tex
sigcse: rename double blind name of pebbles/landslide

@@
 %which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}
-It also provides instruction and memory access tracing and checkpointing/rewinding, both of which Landslide employs.
+It also provides instruction and memory access tracing and checkpointing/rewinding, both of which \landslide employs.
 
@@
 rather than verifying it with certainty.
-(This hit-or-miss nature motivated our work on Landslide.)
+(This hit-or-miss nature motivated our work on \landslide.)
 %Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.

### 2016-08-26 sigcse-17/landslide.tex
sigcse: rename double blind name of pebbles/landslide

@@
-\section{Landslide}
+\section{\landslide}
+% TODO camready
+%\section{Landslide}
 \label{sec:landslide}
@@
 % TODO camready
-we have built Landslide%\cite{landslide}
+we have built \landslide%\cite{landslide}
 , a stateless model checker for \pebbles kernels and thread libraries.
@@
 Conceptually,
-Landslide is designed as follows.
-Through several annotations in the kernel source, Landslide tracks
+\landslide is designed as follows.
+Through several annotations in the kernel source, \landslide tracks
 %the scheduler state to know
@@
 When execution reaches certain interesting {\em preemption points},
-Landslide triggers artificial clock interrupts
+\landslide triggers artificial clock interrupts
 to force the scheduler to run a different thread.
@@
 to one pattern of thread switches,
-Landslide rewinds the simulation state
+\landslide rewinds the simulation state
 and resumes the test according to a different interleaving.
 After each instruction,
-Landslide applies several bug-detection predicates to
+\landslide applies several bug-detection predicates to
 the program state to detect
@@
 To trigger the interleaving which exposes this bug, threads must be preempted at the boundaries of the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
-Landslide automatically considers these boundaries to be {\em preemption points}: candidate sites for timer-driven thread switches.
+\landslide automatically considers these boundaries to be {\em preemption points}: candidate sites for timer-driven thread switches.
 It can also add preemption points dynamically,
@@
 %As these state spaces are exponentially-sized,
-Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
+\landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
 to mitigate the exponential explosion of these state spaces.
@@
 
-Landslide checks for bugs in several ways.
+\landslide checks for bugs in several ways.
 It can identify assertion failures, use-after-free and other invalid heap accesses, and
@@
 %In contrast with analyses that directly detect data races~\cite{tsan},
-Although Landslide employs a data-race analysis to decide where to preempt,
+Although \landslide employs a data-race analysis to decide where to preempt,
 it does not directly classify suspicious memory accesses as bugs,
@@
 	\end{tabular}
-	\caption{Example preemption trace provided by Landslide as debugging output for the bug in Figure~\ref{fig:paradise}.}
+	\caption{Example preemption trace provided by \landslide as debugging output for the bug in Figure~\ref{fig:paradise}.}
 	\label{fig:trace}
@@
 
-When Landslide identifies a bug, it outputs an HTML-formatted
+When \landslide identifies a bug, it outputs an HTML-formatted
 {\em preemption trace}.
@@
 
-For the thread library project, Landslide instruments student implementations automatically.
+For the thread library project, \landslide instruments student implementations automatically.
 We provide a custom reference kernel binary equipped with the kernel instrumentations (described below)
-so Landslide can track threads' lifecycles. %the lifecycle of threads.
+so \landslide can track threads' lifecycles. %the lifecycle of threads.
 The userspace \x{malloc} and \x{assert} are also provided in the basecode, and tracked similarly.
-Landslide's data-race analysis must understand the locking discipline,
+\landslide's data-race analysis must understand the locking discipline,
 so it searches the student's code for functions named \x{mutex_lock} and \x{mutex_unlock} (as mandated by the assignment handout),
@@
 	To relax this assumption, we provide a special test case for mutexes,
-	in which Landslide may preempt and search for data races in the mutex implementation itself.
+	in which \landslide may preempt and search for data races in the mutex implementation itself.
 }.
@@
 
-To test \pebbles kernels directly, they must be annotated to inform Landslide of certain important concurrency events.
+To test \pebbles kernels directly, they must be annotated to inform \landslide of certain important concurrency events.
 Because the kernel project leaves internal code structure as an open-ended design decision,
-Landslide cannot search for specific function names. % like it does for P2s.
-Instead, we provide a set of functions named with the prefix \x{tell_landslide}
+\landslide cannot search for specific function names. % like it does for P2s.
+Instead, we provide a set of functions named with the prefix
+% TODO camready
+\texttt{tell\_\landslide}
+%\x{tell_landslide}
 to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
-Users must also list all core scheduler, context switch, and timer handling routines, so Landslide can ignore their memory accesses for the sake of DPOR
+Users must also list all core scheduler, context switch, and timer handling routines, so \landslide can ignore their memory accesses for the sake of DPOR
 (otherwise, threads would always conflict with each other and DPOR would achieve no reduction).
@@
 Student submissions may behave in extraordinary ways which violate the assumptions of many modern model checkers.
-Landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
+\landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
 but it must support testing any legal, yet distasteful, solution.
@@
 For example, students may use open-coded \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
-To avoid getting stuck in a cyclic (infinite) state space, Landslide keeps a ``yield loop'' counter for each thread;
-when a thread \x{yield}s 10 times without otherwise making progress, Landslide marks it blocked.
+To avoid getting stuck in a cyclic (infinite) state space, \landslide keeps a ``yield loop'' counter for each thread;
+when a thread \x{yield}s 10 times without otherwise making progress, \landslide marks it blocked.
 %
@@
 To avoid wasting precious student attention with false alarms, % the ultimate sin
-Landslide will reset all blocked threads' yield-loop counters,
+\landslide will reset all blocked threads' yield-loop counters,
 retrying potentially up to 128 times before reporting a true deadlock.
@@
 Students may also break abstraction by open-coding a \x{mutex_unlock} sequence.
-Landslide must track lock operations for accurate data-race analysis,
+\landslide must track lock operations for accurate data-race analysis,
 so to catch these patterns, it
@@
 Similarly, these rogue unlock accesses may unblock a yield-looping thread.
-These heuristics have enabled Landslide to test over 100 student thread libraries from the past 5 semesters fully automatically.
+These heuristics have enabled \landslide to test over 100 student thread libraries from the past 5 semesters fully automatically.

### 2016-08-26 sigcse-17/eval.tex
sigcse: finish off some intro todos; make a graph bigger
sigcse: refine eval description

@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{p2-distribution.pdf}
+	\includegraphics[width=0.48\textwidth]{p2-distribution-large.pdf}
 	\caption{Distribution of thread library grades.}
@@
 To account for this, we also checked grades from prior semesters when \landslide was not available.
-This distribution is indistiguishable from the experimental group,
+%This distribution is indistiguishable from the experimental group,
+Although the minimum grade among the experimental group is 17\% higher than the minimum from prior semesters,
+the middle 50\% of the distributions are quite similar,
 suggesting that \landslide's benefit is largely offset by the time investment required.

### 2016-08-27 sigcse-17/410.tex
sigcse: garth's revisions

@@
 designed for juniors and seniors in Computer Science
-and Electrical and Computer Engineering.
+% garth removed this, i guess
+%and Electrical
+and Computer Engineering.
 For CS students it is one of five courses which fulfill
@@
 %it
-introduces students to our expectations about
+introduces students to course expectations about
 design, analysis, and making choices;
@@
 This project is built using a standard C tool chain
+% FIXME garth wants to cite gdb
 and debugged as students see fit (generally \x{printf()} and GDB).
@@
 This project and the remaining ones are written in
-C with some x86-32 assembly,
+C with a small amount of x86-32 assembly code,
 which is then
@@
 floppy-disk image,
+% FIXME garth wants to cite grub
 and booted via GRUB.
@@
 However, students most often use
+the x86 hardware simulator
 Simics \cite{simics} (described in \sect{\ref{sec:simics}}).
@@
 %version of POSIX Pthreads.
+% FIXME garth wants to cite pthreads
 similar to POSIX Pthreads.
@@
 student code using a variety of
-interleaving policies.
+interleaving policies,
+with the ad-hoc goal of exposing concurrency bugs.
 %\shortversion{}{In order to aid debugging of concurrency problems,
@@
 twenty-five system calls,
-which we list in Table~\ref{tab:syscalls}.
+listed in Table~\ref{tab:syscalls}.
 %(life-cycle, thread management, memory management,
@@
 		\hline
-		\x{getchar}* & Reads one character from kbd input. \\
-		\x{readline} & Reads the next line from kbd input. \\
+		\x{getchar}* & Reads one character from keyboard. \\
+		\x{readline} & Reads the next line from keyboard. \\
 		\x{print} & Prints a memory buffer to the console. \\
@@
 	\end{tabular}
-	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
+	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement calls marked with an asterisk (*), though the reference kernel provides them. }
 	\label{tab:syscalls}
@@
 
-Widely regarded as the most difficult concurrency problem in the project
+%Widely regarded as
+The most difficult concurrency problem in the project
 is that of coordinating a parent and a child task that ``simultaneously''
@@
 and an in-kernel debugger.
-Two recent, more aggressive, projects have been
+Two more aggressive, projects have been
 paravirtualization
@@
 %While developing their code,
-Students will uncover both bugs
-and design misunderstandings by running test code we
-provide.
+Students uncover both bugs
+and design misunderstandings by running course-provided test code.
 %Some tests are simple, exercising one feature or condition
@@
 against corner-case security vulnerabilities, and long-term stability
-under large stress tests.
+under long-running stress tests.
 %
@@
 Because of the ambitious learning objectives,
-we employ a very detailed notion of completeness
+we employ a detailed notion of completeness
 for the thread library and kernel projects.
@@
 to handle resource exhaustion,
-and to avoid certain distasteful x86-specific code structures.
+and to avoid certain prohibited x86-specific code structures.
 

### 2016-08-27 sigcse-17/eval.tex
sigcse: garth's revisions

@@
 during the Spring '15, Fall '15, and Spring '16 semesters of \fourten.
-We introduced stateless model checking and gave a demo during a special lecture,
+We introduced stateless model checking in class, %and gave a demo during a special lecture,
 and required students to pass the thread library hurdle (\sect{\ref{sec:grading}}) to use \landslide.
-Of 90 total two-student groups in those semesters, 47 tested their thread libraries with \landslide.
+Of 90 two-student groups in those semesters, 47 tested their thread libraries with \landslide.
 
@@
 	%det bugs found	det bugs fixed	races fixed	races found
-	& \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
-	\# bugs	& found & fixed & found & fixed \\
+	& \multicolumn{4}{c}{Number of groups} \\
+	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
+	of bugs	& found & fixed & found & fixed \\
 	\hline
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{p2-distribution-large.pdf}
+	\includegraphics[width=0.48\textwidth]{p2-distribution-logos.pdf}
 	\caption{Distribution of thread library grades.}

### 2016-08-27 sigcse-17/intro.tex
sigcse: rename double blind name of pebbles/landslide
sigcse: garth's revisions

@@
 In \fourten, the Operating Systems Design and Implementation class at \cmu,
-students complete aggressively low-level thread library and kernel projects,
+students complete low-level thread library and kernel projects,
 which run on real hardware,
@@
 % TODO camready
-\cite{kspec-anonymized}
-, students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
+\cite{kspec-anonymized}, students build a user-space threading and synchronization library on top of a \pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, implement their own multi-threaded kernel.
 %During these projects
@@
 In comparison with other educational kernel projects,
-\pebbles focuses more on ground-up design than on teaching a broad feature set.
+with \pebbles we focus more on ground-up design than on teaching a broad feature set.
 At Berkeley, Stanford, and the University of Chicago, among others, students build kernels following the Pintos architecture \cite{pintos}, and at MIT, using Xv6/JOS \cite{xv6}.
@@
 
-Incorporating SMP into the \fourten curriculum would expand the challenges students face in a single semester.
+Incorporating symmetric multiprocessing (SMP) into the \fourten curriculum would expand the challenges students face in a single semester.
 However, we wish to avoid compromising the complexity
@@
 We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
-and discuss the possibility for integrating it as a main component of operating systems curricula.
+and discuss the possibility of integrating it as a main component of the operating systems curriculum.
 

### 2016-08-27 sigcse-17/landslide.tex
sigcse: garth's revisions

@@
-\section{\landslide}
+\section{Logos}
 % TODO camready
@@
 \landslide is designed as follows.
-Through several annotations in the kernel source, \landslide tracks
+Using several annotations in the kernel source, \landslide tracks
 %the scheduler state to know
@@
 (For testing thread libraries, these annotations are already provided in the reference kernel binary;
-for testing kernels, the student must add them by hand.)
+for testing kernels, students must add them by hand.)
 When execution reaches certain interesting {\em preemption points},
@@
 \landslide rewinds the simulation state
-and resumes the test according to a different interleaving.
+and resumes the test according to a different thread interleaving.
 After each instruction,
@@
 {\bf State space exploration.}
-To trigger the interleaving which exposes this bug, threads must be preempted at the boundaries of the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
-\landslide automatically considers these boundaries to be {\em preemption points}: candidate sites for timer-driven thread switches.
+To trigger the interleaving which exposes this bug, threads must be preempted when invoking the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
+\landslide automatically considers these boundaries to be {\em preemption points}:
+candidate sites for artificial timer-driven thread switches.
 It can also add preemption points dynamically,
@@
 to mitigate the exponential explosion of these state spaces.
-% TODO make sure this gets mentioned later, if not here
-%and state space estimation \cite{estimation} to predict which
-% TODO CAMREADY: cite oopsla
 % In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.
@@
 Apart from which test program to run, \landslide's user interface requires just one option:
-students specify how much time they are willing to wait.
+students specify how long they are willing to wait.
 %Considering this CPU budget,
-\landslide then uses state space estimation \cite{estimation} to predict
+% TODO CAMREADY: deanonymize
+\landslide then uses state space estimation \cite{estimation-anonymized} to predict
 % TODO CAMREADY: deanonymize
@@
 The trace reports what kind of bug was detected
-and each preemption point in the current interleaving:
+and each preemption point encountered in the current interleaving:
 which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
@@
 	To relax this assumption, we provide a special test case for mutexes,
-	in which \landslide may preempt and search for data races in the mutex implementation itself.
+	in which \landslide will preempt and search for data races in the mutex implementation itself.
 }.
@@
 to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
-Users must also list all core scheduler, context switch, and timer handling routines, so \landslide can ignore their memory accesses for the sake of DPOR
+Users must also list all scheduler, context switch, and timer handling routines, so \landslide can ignore their memory accesses for the sake of DPOR
 (otherwise, threads would always conflict with each other and DPOR would achieve no reduction).
@@
 
-Student submissions may behave in extraordinary ways which violate the assumptions of many modern model checkers.
+Student submissions may behave in extraordinary ways, violating the assumptions of many modern model checkers.
 \landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
-but it must support testing any legal, yet distasteful, solution.
+but it must support testing any legal, if distasteful, solution.
 %Hence, we added several heuristic
-For example, students may use open-coded \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
+For example, students may use potentially infinite \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
 To avoid getting stuck in a cyclic (infinite) state space, \landslide keeps a ``yield loop'' counter for each thread;
@@
 Similarly, these rogue unlock accesses may unblock a yield-looping thread.
-These heuristics have enabled \landslide to test over 100 student thread libraries from the past 5 semesters fully automatically.
+
+%So far,
+These heuristics have enabled \landslide to test over 100 student thread libraries
+%from the past 5 semesters
+fully automatically
+(47 in progress in our evaluation, plus
+79 submitted over the past 5 semesters).

### 2016-08-27 sigcse-17/paper.tex
sigcse: rename double blind name of pebbles/landslide
sigcse: garth's revisions

@@
 \newcommand\cmu{DBU\xspace}
-\newcommand\pebbles{\cmu-OS\xspace}
-\newcommand\landslide{Landslide\xspace}
+\newcommand\pebbles{EthOS\xspace}
+\newcommand\landslide{Logos\xspace}
 \newcommand\fourten{CS999\xspace}
@@
 threading library, and culminates in a six-week long project in which students
-build from the ground up a fully-preemptible, UNIX-like kernel that can run
+build from the ground up a fully-preemptible, streamlined UNIX-like kernel that can run
 on real hardware.
@@
 debugging concurrency bugs.
-We present Landslide, a
+We present \landslide, a
 %systematic testing
@@
 % edupar version
-%We report on our preliminary experience working with students using Landslide to find previously-overlooked bugs in their own code, and discuss future methods for making Landslide more accessible to struggling students.
+%We report on our preliminary experience working with students using \landslide to find previously-overlooked bugs in their own code, and discuss future methods for making \landslide more accessible to struggling students.
 % sigcse version
-We report on our experience giving students Landslide during the thread library
+We report on our experience giving students \landslide during the thread library
 %and kernel projects.
 project,
-%Although kernel-mode Landslide requires manual instrumentation, testing thread libraries is fully automatic.
+%Although kernel-mode \landslide requires manual instrumentation, testing thread libraries is fully automatic.
 showing that it allows them to find and fix previously-overlooked bugs.
@@
 % Further, %even struggling students can take advantage of it,
-% Landslide's users achieve higher concurrency scores on average in subsequent projects,
+% \landslide's users achieve higher concurrency scores on average in subsequent projects,
 % suggesting that debugging with a model checker improves student retention of concurrency principles.
@@
 
-%\section{Conclusions and Future Work}
-\section{Future Work}
+\section{Conclusions and Future Work}
 
@@
 % ``SLOCCount.''
-% Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.
+% Work on \landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.
 

### 2016-08-27 sigcse-17/eval.tex
sigcse: save some frames^Wlines and add a conclusion paragraph

@@
 \begin{table}[t]
+	\begin{center}
 	\begin{tabular}{r|cc|cc}
@@
 	& \multicolumn{4}{c}{Number of groups} \\
-	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Nondeterministic bugs} \\
+	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Concurrency bugs} \\
 	of bugs	& found & fixed & found & fixed \\
@@
 	\end{tabular}
+	\end{center}
 	\caption{Summary of how many bugs were found and/or fixed by how many groups.
@@
 who lack enough free time to volunteer for \landslide.
-To account for this, we also checked grades from prior semesters when \landslide was not available.
+%To account for this,
+Thus, we also checked grades from prior semesters when \landslide was not available.
 %This distribution is indistiguishable from the experimental group,
@@
 the middle 50\% of the distributions are quite similar,
-suggesting that \landslide's benefit is largely offset by the time investment required.
+suggesting that \landslide's measurable benefit is largely offset by the time investment required.
 %
-In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.
+%In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

### 2016-08-27 sigcse-17/paper.tex
sigcse: save some frames^Wlines and add a conclusion paragraph

@@
 %Although kernel-mode \landslide requires manual instrumentation, testing thread libraries is fully automatic.
-showing that it allows them to find and fix previously-overlooked bugs.
+showing that it allows them to find and fix many previously-overlooked concurrency bugs.
 %% no :( :(
@@
 \ccsdesc[500]{Software and its engineering~Software maintenance tools}
-\ccsdesc[300]{Computing methodologies~Concurrent algorithms}
+%\ccsdesc[300]{Computing methodologies~Concurrent algorithms}
 
@@
 
+\fourten, the undergraduate operating systems course at \cmu,
+features low-level thread library and kernel projects
+in which students tackle challenging concurrency problems.
+We have introduced \landslide, a stateless model checker,
+as an alternative debugger to conventional stress testing.
+In our evaluation over three semesters,
+\landslide helped more than half of volunteering students find and fix previously-overlooked concurrency bugs.
+So far, however, this benefit is not reflected in the grades students ultimately achieved.
+In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.
+%helping students achieve more overall.
+
 We are presently extending \landslide to support Pintos kernels \cite{pintos},
@@
 Because Pintos mandates more internal APIs than \pebbles,
-\landslide can more easily instrument student code without requiring manual annotations.
+\landslide can easily instrument student code without requiring manual annotations.
 In a preliminary experiment, we tested 79 student Pintos implementations with \landslide,
@@
 In future work, we hope to fully automate Pintos instrumentation, and offer \landslide at other schools beyond \cmu.
-
-%\fourten, the undergraduate operating systems course at \cmu, teaches concurrency in a
 

### 2016-08-27 sigcse-17/410.tex
sigcse: save some frames^Wlines and add a conclusion paragraph
sigcse: tom's revisions

@@
 %
-The prerequisite course is \twothirteen,
+The prerequisite is \twothirteen,
 Introduction to Computer Systems,
@@
 %universities in the U.S.\ and worldwide
-adapted from Carnegie Mellon University's 15-213
+a course adapted from Carnegie Mellon University %'s 15-213
 \cite{sigcse01:CSaPP}.
@@
 
-Students who complete the kernel project on time
+Students who complete the kernel on time
 then work on a kernel-extension project.
@@
 
+\textbf{Time.}
+Being a high-level systems elective, \fourten is exceptionally time-consuming,
+demanding 20 or more hours weekly during the hardest projects. %thread library and kernel projects.
+A more accessible version of the course could be implemented
+by extending the time available for the first three projects
+and concluding with a stripped-down subset of the kernel,
+or by offering the kernel project in an optional second semester.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-08-27 sigcse-17/intro.tex
sigcse: tom's revisions

@@
 It also features an experimental, fully automatic, Pintos testing mode.
-We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
-and discuss the possibility of integrating it as a main component of the operating systems curriculum.
+%We present our experience sharing \landslide with volunteering \fourten students who used it to find bugs in their own code,
+%and discuss the possibility of integrating it as a main component of the operating systems curriculum.
+
+We describe the \fourten curriculum in detail (\sect{\ref{sec:410}}),
+highlight \landslide's design and functionality (\sect{\ref{sec:landslide}}),
+present our experience sharing \landslide with volunteering students who used it to find concurrency bugs (\sect{\ref{sec:eval}}),
+%and discuss the possibility of integrating it as a main component of the operating systems curriculum (\sect{\ref{sec:warpzone}}).
+and discuss ongoing work to incorporate it into the curriculum at \cmu and other schools (\sect{\ref{sec:warpzone}}).
 

### 2016-08-27 sigcse-17/410.tex
sigcse: SAVE THE ANIMALS!^Wlines.

@@
 
-The course has many learning objectives,
-ranging from acquiring detailed factual knowledge about
+The course's learning objectives
+range from acquiring detailed knowledge about
 hardware features
-through practicing advanced cognitive processes
+to practicing advanced programming skills %cognitive processes
 such as open-ended design.
@@
 %
-The course's distinguishing feature is that
-students begin with a ``blank slate'' rather than a
+%The course's distinguishing feature is that
+Students begin with a blank slate rather than a
 kernel-source template or an existing operating system,
-so they must synthesize design requirements from multiple sources
-and must
-choose their own module boundaries and inter-module conventions.
+so they must synthesize design requirements %from multiple sources
+and %must
+choose their own module boundaries and conventions.
 %
@@
 %hardware platform.}
-We target x86 PCs because machines of this type are
-available from many manufacturers at many price points.
-This means that it is easy for students to obtain access
+We target x86 PCs %because machines of this type are
+%available from many manufacturers at many price points.
+%This means that it is easy for
+so students can easily obtain access
 to a machine (virtual or real) which will let them

### 2016-08-27 sigcse-17/eval.tex
sigcse: SAVE THE ANIMALS!^Wlines.

@@
 We also distinguished {\em deterministic} bugs from concurrency bugs
-by identifying whether \landslide needed to test multiple interleavings before finding the bug\footnote{
-	For example, many students found deterministic use-after-free bugs,
-	previously overlooked because the unit tests do not provide heap checking like \landslide does.
-}.
+by identifying whether \landslide needed to test multiple interleavings before finding the bug.
+(For example, many students found deterministic use-after-free bugs,
+previously overlooked because the unit tests do not provide heap checking like \landslide does.)
 
@@
 \begin{figure}[t]
-	\includegraphics[width=0.48\textwidth]{p2-distribution-logos.pdf}
+	\includegraphics[width=0.48\textwidth]{p2-distribution-logos-squashed.pdf}
 	\caption{Distribution of thread library grades.}

### 2016-08-27 sigcse-17/landslide.tex
sigcse: SAVE THE ANIMALS!^Wlines.

@@
 	{\bf (a}) A buggy producer/consumer program, in which the consumer thread only checks \x{work_exists()} once.
-	\vspace{1em}
-	\\
 	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\landslide checks for bugs in several ways.
-It can identify assertion failures, use-after-free and other invalid heap accesses, and
-deadlocks, and also includes a heuristic check for infinite loops and livelock.
-%Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
-%In contrast with analyses that directly detect data races~\cite{tsan},
-Although \landslide employs a data-race analysis to decide where to preempt,
-it does not directly classify suspicious memory accesses as bugs,
-but rather seeks concrete, observable failures that can arise from many different types of errors.
-
 \newcommand\hilight[2]{\color{#1}#2\color{black}\xspace}
@@
 \end{figure*}
+
+\landslide %checks for bugs in several ways.
+%It can identify
+identifies assertion failures, use-after-free and other invalid heap accesses, and
+deadlocks, and also includes a heuristic check for infinite loops and livelock.
+%Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
+%In contrast with analyses that directly detect data races~\cite{tsan},
+Although \landslide employs a data-race analysis to decide where to preempt,
+it does not directly classify suspicious memory accesses as bugs,
+but rather seeks concrete, observable failures that can arise from many different types of errors.
 

### 2016-08-27 sigcse-17/410.tex
sigcse: some more help from garth

@@
 %it
-introduces students to course expectations about
+introduces course expectations about
 design, analysis, and making choices;
@@
 % ^^ LOL, classic dave.
-This project is built using a standard C tool chain
+This project uses a standard C tool chain
 % FIXME garth wants to cite gdb
-and debugged as students see fit (generally \x{printf()} and GDB).
+and debugged as students see fit (generally GDB, a symbolic debugger, and \x{printf}).
 
@@
 \textbf{Time.}
+%% garth's version
 Being a high-level systems elective, \fourten is exceptionally time-consuming,
-demanding 20 or more hours weekly during the hardest projects. %thread library and kernel projects.
-A more accessible version of the course could be implemented
-by extending the time available for the first three projects
-and concluding with a stripped-down subset of the kernel,
-or by offering the kernel project in an optional second semester.
+especially during the thread library and kernel projects.
+It could be revised to fit another program's available time and effort
+levels by extending the time available for the first three projects,
+ending with the thread library, and optionally offering the kernel
+project in a specialized second semester course.
+
+%% my version
+%Being a high-level systems elective, \fourten is exceptionally time-consuming for students,
+%especially during the thread library and kernel projects.
+%A more accessible version of the course could be implemented
+%by extending the time available for the first three projects
+%and concluding with a stripped-down subset of the kernel,
+%or by offering the kernel project in an optional second semester.
 

### 2016-08-27 sigcse-17/410.tex
sigcse: last minute changes

@@
 % FIXME garth wants to cite gdb
-and debugged as students see fit (generally GDB, a symbolic debugger, and \x{printf}).
+and debugged as students see fit (generally GDB, a symbolic debugger; and \x{printf}).
 
@@
 		\x{set_cursor_pos} & Sets the console cursor location. \\
-		\x{get_cursor_pos} & Retrieves the console cursor location. \\
+		\x{get_cursor_pos} & Retrieves same. \\ %the console cursor location. \\
 		\hline
@@
 We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
-distinction between ``tasks,''
+distinction between {\em tasks},
 which are resource containers,
-and threads,
+and {\em threads},
 each of which executes within a single task.
@@
 The most difficult concurrency problem in the project
-is that of coordinating a parent and a child task that ``simultaneously''
+is that of coordinating a parent and a child task that simultaneously
 exit:
@@
 We target the 32-bit
-subset of x86, for multiple reasons.
+subset of x86 for multiple reasons.
 %%% I'm not sure I believe this is an advantage. Instead of having
@@
 opposed to x86-32 requires more code and more
-debugging time without imparting greater
+debugging time without imparting any greater
 insight.

### 2016-08-27 oopsla-16/design.tex
minor touch ups
fix some camready todo layout issues now that we are swimming in free space
buncha minor prose tweaks
change formatting according to camready author package
take care of some TODO CAMREADies
oopsla: last revisions :O

@@
 Algorithm~\ref{alg:algorithm0} demonstrates this na\"ive approach.
-It would be seeded with the set of all statically-known synchronization API PPs,
-and invoked whenever a new data race PP is added.
+It is seeded with the set of all statically-known synchronization API PPs,
+and invoked whenever a new data-race candidate is found.
 %
-Our proofs in \sect{\ref{sec:totalverif}},
+Our proofs in \sect{\ref{sec:soundness}},
 being concerned with the verification guarantee provided when \quicksand~completes within the CPU budget,
@@
 \begin{algorithm}[t]
+	\revisionx{
 	\SetKwInOut{Input}{Input}
@@
 	\label{alg:algorithm0}
+	}
 \end{algorithm}
@@
 
-\subsection{Initial PP configuration}
+\subsection{Initial PP Configuration}
 \label{sec:initial-pp}
@@
 
-\subsection{Data-race preemption points}
+\subsection{Data-Race Preemption Points}
 \label{sec:classifying}
@@
 the small job risks missing the data race entirely if the original PPs were required to expose it.
-% TODO CAMREADY: Put numbers here.
 In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
@@
 \begin{figure}[t]
-	% TODO CAMREADY: increase back to 0.48
-	\hspace{1em}\includegraphics[width=0.45\textwidth]{dr-jobs-v2.pdf}
+	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
 	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
@@
 %(a) the current thread ID and
-%% FIXME: We don't actually do this.
+%% Oops! We don't actually do this.
 %(b) the most recent {\tt call} instruction.
@@
 
-\subsection{Choosing the best job}
+\subsection{Choosing the Best Job}
 
@@
 in case their ETAs are actually overestimated.
-Lines 12-14 account for the
+Lines 12-15 account for the
 %bizarre

### 2016-08-27 oopsla-16/evaluation.tex
oopsla: last revisions :O

@@
 Figure~\ref{fig:dr-falsenegs} compares the types of data-race candidates necessary to expose each data-race bug in our test suite.
-The first series represents the bugs found using PPs from single-pass data-race candidates,
+The first and third series represent the bugs found using PPs from single-pass data-race candidates,
 % not entirely true, as portend could be given data-race traces from an MC,
@@
 i.e., the state-of-the-art approach used by \cite{racefuzzer,portend}.
-The second series shows all data-race bugs \quicksand~found,
+The second and fourth series show all data-race bugs \quicksand~found,
 which includes the former type as well as new bugs involving nondeterministic data-races.

### 2016-08-27 oopsla-16/intro.tex
oopsla: last revisions :O

@@
 %so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
-Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
+Concurrency bugs are notoriously hard to find and reproduce because they appear only in specific thread interleavings, which arise at random during normal program execution.
 {\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
@@
 %makes full state space completion intractable for even modestly-sized tests.
-\revisionx{Some tools, such as CHESS and Inspect \cite{inspect} can strike a middle ground by using compiler instrumentation to statically add preemption points on memory accesses.
+\revisionx{Some tools, such as CHESS and Inspect \cite{inspect}, can strike a middle ground by using compiler instrumentation to statically add preemption points on memory accesses.
 Nevertheless,}

### 2016-08-27 oopsla-16/overview.tex
oopsla: last revisions :O

@@
 The checker defines the granularity of thread interleavings by the {\em preemption points} it uses to switch threads.
-Most model checkers \cite{chess,dbug-ssv} choose synchronization/thread library API boundaries for these points;
+Most model checkers \cite{chess,dbug-ssv} choose synchronization and thread library API boundaries for these points;
 in our example program, these would be lines 1, 4, and 7.

### 2016-08-27 oopsla-16/soundness.tex
oopsla: last revisions :O

@@
 \begin{definition}[Reachability]
-A data race candidate, and its associated PPs, are reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
+A data race candidate, and its associated PPs, are reachable if it will be identified by a MC configured to preempt only on already-reachable PPs.
 \end{definition}
@@
 To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
-Here, Thread 2's logic usually corresponds to an initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.
+Here, Thread 2's logic usually corresponds to an initialization pattern \cite{eraser}, but for generality we have added an arbitrary {\tt publish} action on line 6.
 

### 2016-08-27 oopsla-16/warpzone.tex
split the giant table by making an auxiliary table for dr statistics; add purehb/shmem to table; add discussion of pure hb
take care of some TODO CAMREADies
oopsla: last revisions :O

@@
 
-% TODO CAMREADY: Measure how much subset overlap there is.
+% Measure how much subset overlap there is.
+% Sorry past-me, mot going to do this.
 {\bf Avoiding redundant work.}
@@
 {\bf Finer-grained PP subsets.}
-\quicksand~was able to partially guarantee safety for some PPs in 89\% of tests with too-large maximal state spaces.
-However, in 4 cases, no more than the minimal state space could be verified,
-and in 6 others, no state spaces were completed at all.
+\quicksand~was able to partially guarantee safety for some PPs in \revisionz{93\%} of tests with too-large maximal state spaces.
+However, in \revisionz{6} cases, no more than the minimal state space could be verified,
+and in \revisionz{18} others, no state spaces were completed at all.
 %Larger state spaces often result from finer-grained locking,
@@
 benign, even though no bug was found.
-In the ``Untested DR PPs'' column of Table~\ref{tab:drbugs}, we show how many such candidates we could not verify (32\%).
+In the ``Untested DR PPs'' column of Table~\ref{tab:drstatistix}, we show how many such candidates we could not verify (\revisionz{38\%}).
 For a more formal treatment of these cases, we refer the reader to the {\em k-witness harmless} metric introduced by \cite{portend},

### 2016-09-01 oopsla-16/proof/proof.tex
oopsla proof: address camreadies
ack pdl consortium in proof

@@
 
-% TODO CAMREADY
-% We suspect that DPOR + pure-HB will always eventually find all data races uncovered by MHB.
-% for the purpose of iterative deepening, MHB is
-% better because it will find the data races much sooner, rather than requiring you
-% to explore the maximal state space before the race can be uncovered
-
 \section{Acknowledgements}
@@
 Thanks as well to the anonymous OOPSLA reviewer \#4, whose suggestions helped improve the formality of our definitions.
-This work was supported in part by the U.S. Army Research Office under grant number W911NF0910273.
+
+This work was supported in part by
+the member companies of the PDL Consortium (Broadcom,
+Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
+Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
+Tintri, Two Sigma, Uber, Veritas and Western Digital)
+and by the U.S. Army Research Office under grant number W911NF0910273.
 

### 2016-09-01 oopsla-16/design.tex
oopsla: clean up/remove outdated comments in main tex

@@
 	\SetKwInOut{Input}{Input}
-	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
 	\Input{$j$, the currently-running job}
-	%\Input{$eta$, $j_0$'s predicted completion time}
 	\Input{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
-	%\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
-	%\Input{$T$, the remaining time in the CPU budget}
 	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
@@
 
-%When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
-%Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
-%(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
-%Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
-%Instead, \landslide~qualifies the PC with
-%(a) the current thread ID and
-%% Oops! We don't actually do this.
-%(b) the most recent {\tt call} instruction.
-%% (a crude approximation of a stack trace)
-%% which are much cheaper, as we carry them around all the time already
-%Note that we do {\em not} qualify data races by the shared memory address,
-%which can change based on different interleavings of previous code
-%(for example, depending on the result of {\tt malloc()}).
-%% especially when malloc is involved.
-%%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Our method,
-%for doing so,
 listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
@@
 	\Input{$j_0$, the currently-running job}
-	%\Input{$eta$, $j_0$'s predicted completion time}
 	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
@@
 We also include a heuristic to
-%ignore ETAs entirely
 never suspend jobs before they pass a certain threshold of interleavings tested,

### 2016-09-01 oopsla-16/evaluation.tex
oopsla: clean up/remove outdated comments in main tex

@@
 				% How many false positives does the free-re-malloc technique suppress?
-				% to do (done) If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
-				% Answer: Just 1. (for p2s at least; as of time of writing, pintos dr-falsenegs not run yet)
 			\item Does \quicksand~find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
@@
 %for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
-%#for i in /tmp/shit/pintoses/daniel-deniz*; do
+%#for i in /tmp/shit/pintoses/*; do
 %	echo -ne "$i";
@@
 We tested P2s with 6 multithreaded programs:
-% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
-% each tailored to exercise a different part of the P2 project
 \mxtest, for locking algorithm correctness,
@@
 In total, our evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
-% at least 157 of which could expose bugs.
 at least \revisionz{181} of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt
-
-%\begin{table}[t]
-%	\begin{tabular}{l|l|l}
-%			& QS bugs & SSS-MC bugs \\
-%		\hline
-%		\mxtest & eg 1000 & eg 0 \\
-%		\bct & & \\
-%		etc... & & \\
-%		\hline
-%		Total & & \\
-%	\end{tabular}
-%	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom 2)}
-%	\label{tab:allbugs}
-%\end{table}
-%
-%\begin{table}[t]
-%	\small
-%	\begin{tabular}{l|l|l||l|l}
-%	& QS bug & \begin{tabular}{c} SSS-MC \\ completed\end{tabular}
-%	& QS bug & \begin{tabular}{c}SSS-MC \\ timeout \end{tabular} \\
-%		\hline
-%		\mxtest & e.g. 5 & 10 & 0 & 0 \\
-%		\bct & & & & \\
-%		etc... & & & & \\
-%		\hline
-%		Total & & & & \\
-%	\end{tabular}
-%	\caption{Bugs requiring data-race PPs to expose, found by \quicksand~but missed by the single-state-space approach.}
-%	\label{tab:drbugs}
-%\end{table}
 
@@
 		& \multicolumn{2}{c|}{\bf \revz{Limited HB}} & \multicolumn{2}{c||}{\bf \revz{Pure HB}} &
-		%& {\bf LHB DR} & {\bf Nondet.} & & &
 		{\bf Mutual} & {\bf Avg. tested} \\
@@
 {\bf Finding new data-race bugs.}
-%We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
-%The break-even point is at a negligible 90 seconds.
 \revision{Compared to \revisionx{SSS-MC-ICB (the fastest among the control experiments),} \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds.
@@
 Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
-%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
-%Though for many applications of MC, assuming a correct lock implementation is sufficient,
 Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
@@
 
-%% RIP this.
-%Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
-%which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
-%Intuitively, this line shows that for programs with only benign data races,
-%\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.
-
 % Future work (for thesis maybe?) Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
 {\bf Finding the same bugs faster.}
-%To test whether Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
 \revision{The QS-Sync-Only experiment tests whether}~Iterative Deepening is effective even for MC domains without data races.
@@
 %Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.
-
-%Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
-%and is faster in the first 60 seconds due to less start-up overhead,
-%QS-no-DR-PPs finds more of the bugs sooner thereafter.
-%Hence, for modest CPU budgets,
-%\quicksand~is likely to find bugs SSS-MC will miss.
-%and for more ambitiously-sized tests,
-%programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
-%We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
-%Iterative Deepening is likely to find bugs SSS-MC will miss.
-%Moreover, it is also easy to imagine scaling up the size of each test case to test ,
-%using more threads or longer sequences of API calls.
-%We hope that is compelling even to users willing to spend many CPU-hours on testing.
-%These results show explicitly that for arbitrary CPU-time budgets
-%\footnote{The initial perfect overlap between QS-*-HB and QS-Sync-Only indicates how long it takes before the first data-race bug is found.}
-%even after the extra overhead of verifying them, \quicksand~still slightly edges out SSS-MC
 
@@
 For the other \revisionz{192}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
-%and the number of unique PPs among them.
-%In 6 cases, \quicksand~also failed to complete anything; beyond these,
-%between 1 and 253 state spaces were completed for each test
-%Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
 These completions guarantee that, if the test program could expose a bug,
-% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
 it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
@@
 }
-
-%Hence, in Table~\ref{tab:drbugs} we count how often \quicksand~uncovered a bug only in state spaces which included data-race PPs, while
-%
-%In Table~\ref{tab:allbugs} ....
 
@@
 \revisionx{The next best approach for verifications was SSS-MC-Shared-Mem, which completed its search in only 39 cases.}
-
-%Though SSS-MC-Shared-Mem appears to outperform \quicksand, its safety guarantee is unsound,
-%as many tests break its assumption of stack-allocated memory being private.
-%For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
-%While SSS-MC-Shared-Mem must avoid excessive PPs on every private stack access for performance,
-%\quicksand~detects these dynamically, without needing to discriminate between stack and heap.
-%In total, SSS-MC-Shared-Mem claimed to verify 21 tests
-%(across 7 of the 9 test cases)
-%in which QS-DR ultimately found bugs,
-%which further calls into question the rest of its completions.
 
@@
 %}.
-%\quicksand~outperforms SSS-MC, being able to verify safety for
-%{\bf 9999x}
-%as many tests in 10 CPU-hours.
-%We attribute this again to the overhead of the sheer number of statically-coded PPs that SSS-MC-Shared-Mem must use,
-%as well as to the work which ICB necessarily repeats when increasing its preemption bound.}
 
@@
 
-% Actually, while SSS-MC-Shared-Mem proved competitive with \quicksand~at finding data-race-related bugs,
-% its heuristic to ignore threads' own stack accesses introduces unsoundness.
-% <Move contents of footnote here.>
-% Indeed, SSS-MC-Shared-Mem claimed a full verification on 999999 tests for which \quicksand~found bugs!
-% % Hence, no SSS-MC strategy competes with quicksand for truly sound verification
-% Hence, only SSS-MC-DPOR and SSS-MC-ICB compete with \quicksand for sound verification, and they are thoroughly outperformed.
-% Between data-race bugs and verifications, \quicksand~is shown to provide the best of both worlds.
-
 \begin{figure}[t]
@@
 	(\sect{\ref{sec:totalverif}}).
-	%SSS-MC-Shared-Mem completed 1.37x as many tests, but 
-	%SSS-MC-Shared-Mem is unsound: 21 tests it completed hid bugs previously found in Figure~\ref{fig:dowefindbugsfaster}.
-	%previously found (Figure~\ref{fig:dowefindbugsfaster}).
 	%Some tests had no data-race candidates, and hence could also be verified by .}
@@
 \revisionz{QS-PHB is even more dependent on nondeterministic data-race PPs, requiring nondeterministic data-race PPs in 21 cases among its 44 total data-race bugs.}
-%In total, we found 189\% as many bugs by using nondeterministic data-race candidates as PPs.
 
@@
 %although that would still not simultaneously be able to provide total verifications.
-
-% Figure out concretely what the data race tricks are that we do, so we can claim them as contributions in the paper. Then ACTUALLY EVALUATE THEM.
-%         - Speculative DR PPs.
-%                 Not a heuristic, rather how to make it work at all to begin with.
-%                 (Cite MS thesis, claim on backwards explorating finding bugs faster)
-%         - Free/re-malloc to eliminate some false positives. See #193.
-%                 Measure how many false positives are eliminated.
-%                 Check, ofc, to make ABSOLUTE SURE, that no bugs missed w/ this trick.
-%                         If there are, it could be because of the implementation
-%                         bug described in #193.
-%         - Using tid/last_call filtering because whole stack traces are too expensive.
-%                 Moderately optional, 1st priority since theoretically interesting:
-%                 Turn on/off and measure how resulting DR bug #s change.
-%         - Optional: Reprioritizing DRs based on "confirmed" / "suspected"
-%                 Shouldn't be hard just make ID wrapper print "s" or "c"!
-%                 Is it helpful for ID to put priorities on DR PPs?
-%                         Test by inverting the priority and see if fewer buges are found.
-%         // Super optional to talk about. Probably not worth the time.
-%         // - "Too suspicious" (during init/destroy)
-%         //      (Cite eraser, section 2.2)

### 2016-09-01 oopsla-16/overview.tex
oopsla: clean up/remove outdated comments in main tex

@@
 {\bf Philosophy of bugs.}
-% two camps: all races are bugs; citations:
-% and, races must be classified; citations: replay analysis, portend,
 While there is a vast body of work on how to detect data races to begin with,

### 2016-09-01 oopsla-16/paper.tex
oopsla: ack PDL consortium, make graph a bit bigger
oopsla: last revisions :O
oopsla: clean up/remove outdated comments in main tex

@@
 \newcommand\numthrlibs{79}
-\newcommand\numpintoses{78} % 'di' is basecode. durr
+\newcommand\numpintoses{78}
 \newcommand\numstudence{157} % total pintoses plus p2s
@@
 % cam ready revisions
-\newcommand\revisionz[1]{\hilight{purple}{#1}\xspace}
+%\newcommand\revisionz[1]{\hilight{purple}{#1}\xspace}
+\newcommand\revisionz[1]{#1\xspace}
 
-\title{Stateless Model Checking with Data-Race Preemption Points}
+% obfuscate title just a little bit to avoid the billion inevitable google searches
+% i don't mind if really dedicated people find their way here, but not everyone please
+\newcommand\titlec{l Check}
+\newcommand\titleg{ion Points}
+\newcommand\titleb{less Mode}
+\newcommand\titlee{h Data-R}
+\newcommand\titled{ing wit}
+\newcommand\titlef{ace Preempt}
+\newcommand\titlea{State}
+
+\title{\titlea\titleb\titlec\titled\titlee\titlef\titleg}
 
@@
 \authorinfo{Garth Gibson}{Carnegie Mellon University, USA}{garth@cs.cmu.edu}
-%\authorinfo{\vspace{24pt}}{}{}
 
@@
 We have presented Iterative Deepening and \quicksand, a new technique and tool for automating the choice of preemption points (PPs) during stateless model checking.
-% , and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
 \quicksand~incorporates data-race analysis to create new PPs tailored specifically to the program under test,
 and automatically finds state spaces that are appropriately sized to complete in a given CPU budget.
-%and manages multiple model checker instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.
 
 We achieve better bug-finding results than either single-pass data-race detection or single-state-space model checking alone,
-%We dramatically reduce false-positive data race reports,
-%%by verifying the absence of bugs in their associated state spaces
-%and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
-%Compared to prior work,
-%, we find bugs faster in smaller subset state spaces, %% not true, or at least, not by very much
-%when the maximal state space is too large to explore
 finding new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
@@
 
-%\section{Acknowledgements}
-
-%\appendix
-%\section{Appendix Title}
-%
-%This is the text of the appendix, if you need one.
-
 \acks
@@
 the anonymous OOPSLA reviewers for their helpful comments.
+%
+We also thank the member companies of the PDL Consortium (Broadcom,
+Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
+Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
+Tintri, Two Sigma, Uber, Veritas and Western Digital).
+%
 This work was supported in part by

### 2016-09-01 oopsla-16/related.tex
oopsla: clean up/remove outdated comments in main tex

@@
 to Verisoft, the original C model checker \cite{verisoft}.
-%, and Eraser, the original data race detector \cite{eraser}.
 %Our model checker
@@
 \revision{\cite{chess} discusses the ability of CHESS to add PPs using a single-pass data-race analysis,
-% FIXME... idk, "discusses (without explicit evaluation)..., but, as our evaluation shows..."?
 but does not evaluate either the increase in bug-finding capacity or the soundness properties.
@@
 %Nevertheless, Iterative Deepening is applicable to these tools.
+% IDK if i should mention it, but OOPSLA 2015, protocol based verification of MPI concurrency paper.
+Different verification approach entirely; doesn't suffer exponential explosion but limited to programs with no shared state and MPI communication only
 
@@
 %and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
+% Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.
 
@@
 %although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
-%FastTrack \cite{fasttrack} optimizes the performance of pure happens-
 
@@
 Future work could combine the two approaches, using MC to produce new data-race traces for Portend to classify, or using Portend's analysis to inform \quicksand's heuristic priorities.
-
-% \subsection{Other Concurrency Testing Approaches}
-%
-% blah blah pldi'15 symbiosis DSP
-
-% Note that BPOR paper claims that ICB(3+) repeats LOADS of work, and that makes it ok for landslide-ID to repeat work.
-
-% IDK if i should mention it, but OOPSLA 2015, protocol based verification of MPI concurrency paper.
-Different verification approach entirely; doesn't suffer exponential explosion but limited to programs with no shared state and MPI communication only
-
-% Probably NOT worth a mention: OOPSLA 2015, stateless model checking of event driven applications.
-Turning timer-driven model on its head and checking single-threaded, but asynch-event-driven programs (i.e.
-device-like signal handlers)

### 2016-09-01 oopsla-16/proof/di.tex
oopsla proof: address camreadies
oopsla proofs: fix system model to reason about execution traces, leaving semantics opaque

@@
 
-To reason about programs at the executable level, we will define them as a list of assembly-like instructions,
-including certain concurrency primitives which we evaluate atomically, assuming the runtime provides correct implementations for them.
+To avoid relying on any particular programming language features,
+we leave the program syntax and execution semantics opaque,
+reasoning instead about execution traces.
+We require that a program's evaluation produce a trace of instructions of the following form:
 \begin{eqnarray*}
-	\mathcal{P} &::=& [n, \mathcal{I}] \\
-	\mathcal{I} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg}(a,v) \\
+	\mathcal{A} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg(a,v)} \\
 		&|& a \leftarrow \mathsf{malloc}(n) \quad | \quad \mathsf{free}(a) \\
-	&|& \mathsf{mutex\_lock}(m) \quad | \quad \mathsf{mutex\_unlock}(m) \\
-	&|& \mathsf{deschedule} \quad | \quad \mathsf{make\_runnable}(t) \\
-	&|& t \leftarrow \mathsf{thread\_fork} \quad | \quad \mathsf{thread\_exit} \quad | \quad \mathsf{yield} \\
-	&|& \textit{local}
+		&|& \textit{local} \\
+		&|& \mathcal{A}_{sync}
 \end{eqnarray*}
+The execution steps take the following forms:
 
-The execution state consists of any number of threads, each represented by a list of instructions to execute ($[\mathcal{I}]$),
-a stack of variable maps for thread-local state ($v$),
-and an indicator of whether the thread is runnable.
-The state also includes a map of global memory, indexed by addresses denoted $a$,
-and a set of mutex objects, denoted $m$. Thus:
-\begin{eqnarray*}
-	\mathcal{E} &::=& [\mathcal{T}] \times [a \rightarrow \mathsf{int}] \times [m \rightarrow \mathsf{bool}] \\
-	\mathcal{T} &::=& \mathcal{P} \times [[v \rightarrow \mathsf{int}]] \times \mathsf{bool}
-\end{eqnarray*}
-For the sake of brevity, we will not explicitly define execution semantics, though they are straightforward:
-To execute one step, arbitrarily choose one runnable thread, and evaluate the first element of its instruction list.
-This gives rise to thread concurrency which can interleave at instruction granularity,
-although it assumes a sequentially-consistent memory model. For a treatment of DPOR with relaxed memory, we refer the reader to \cite{tsopso}.
-
-{\bf Memory.} $\mathsf{read}$, $\mathsf{write}$, and $\mathsf{xchg}$ access global or heap memory shared by all threads, indicated by some address $a$, and copying and/or modifying a thread-local variable $v$.
-% cmpxchg, xadd, etc omitted for brevity
+{\bf Memory.}
+$\mathsf{read}$, $\mathsf{write}$, and $\mathsf{xchg}$ access global or heap memory shared by all threads, indicated by some address $a$.
+(Other atomic swap operataions are omitted for brevity.)
 $\mathsf{malloc}$ and $\mathsf{free}$ provide access to fresh memory accessible by all threads. % free is irrelevant to the system model in terms of expressive power
 
-{\bf Threads.} $\mathsf{mutex\_lock}$ and
+{\bf Local state.}
+$\textit{local}$ represents any thread-local instruction, such as modifying local variables, flow control, function calling, and assertions. %, and side effects.
+We omit a detailed list for brevity, as we do not need to reason about their nature in these proofs.
+%For brevity, we omit a detailed list, although note that $\mathsf{call}$ would be implemented by modifying the instruction stream $[\mathcal{I}]$ and creating a new frame on the stack of local variables.
+
+{\bf Threads.}
+$\mathcal{A}_{sync}$ denotes the subclass of evaluation steps which implement inter-thread synchronization; i.e., the {\em synchronization API}:
+\begin{eqnarray*}
+	\mathcal{A}_{sync} &::=& \mathsf{mutex\_lock}(m) \quad | \quad \mathsf{mutex\_unlock}(m) \\
+		&|& \mathsf{deschedule} \quad | \quad \mathsf{make\_runnable}(t) \\
+		&|& t \leftarrow \mathsf{thread\_fork} \quad | \quad \mathsf{thread\_exit} \quad | \quad \mathsf{yield}
+\end{eqnarray*}
+$\mathsf{mutex\_lock}$ and
 $\mathsf{mutex\_unlock}$ provide mutual exclusion:
-a thread cannot evaluate $\mathsf{mutex\_lock}$ until that lock is released.
-$\mathsf{deschedule}$ and $\mathsf{make\_runnable}$ allow threads to manipulate their own or another's runnability, respectively,
+a thread which evaluates $\mathsf{mutex\_lock}$ on some lock $m$ becomes blocked until no other thread holds $m$.
+% the second formulation here excludes the possibility of m_r waking a thread blocked on mutex_lock (which would be horrible)
+%$\mathsf{deschedule}$ and $\mathsf{make\_runnable}$ allow threads to manipulate their own or another's runnability, respectively,
+$\mathsf{deschedule}$ alows a thread to block itself until another thread wakes it with $\mathsf{make\_runnable}$,
+%
 and $\mathsf{thread\_fork}$ and $\mathsf{thread\_exit}$ allow creation and destruction of new threads.
 $\mathsf{thread\_fork}$ is defined in the Pebbles manner \cite{kspec};
-we omit higher-level abstractions such as $\mathsf{create}$ or $\mathsf{join}$, which can be implemented using these primitives \cite{thrlib}.
+we omit higher-level abstractions such as $\mathsf{cond\_wait}$, $\mathsf{create}$, or $\mathsf{join}$, which can be implemented using these primitives \cite{thrlib}.
 $\mathsf{yield}$ has no effect under these execution semantics, but is included for the sake of synchronization API preemption points (see below).
 
-{\bf Local state.}
-$\textit{local}$ represents any thread-local instruction, such as modifying local variables, flow control, function definition/calling, assertions, and side effects.
-For brevity, we omit a detailed list, although note that $\mathsf{call}$ would be implemented by modifying the instruction stream $[\mathcal{I}]$ and creating a new frame on the stack of local variables.
-
-\begin{definition}[Blocked threads]
-	A thread is blocked when either its runnable flag is false, or when its next instruction is a $\mathsf{mutex\_lock}$ on an unavailable mutex.
+\begin{definition}[Interleaving]
+An interleaving (or execution trace) is a list of these instructions, annotated to indicate the currently-running thread as well as the runnability of all existing threads:
+\begin{eqnarray*}
+	\mathcal{I} &::=& [\mathcal{A}, t, (t \rightarrow \mathsf{bool})]
+\end{eqnarray*}
 \end{definition}
 
-Model checkers often also include heuristics to identify threads using open-coded $\mathsf{yield}$ loops to block,
-but for the sake of our proofs, we assume such patterns are implemented more tastefully with a condition-variable-like primitive built upon $\mathsf{deschedule}$.
+We say a {\em thread switch} occurs when adjacent elements in $\mathcal{I}$ have different thread IDs $t$.
+%
+We say a thread is {\em blocked} when its value in the runnability map is false.
 
-\begin{definition}[Interleaving]
-	An interleaving is a sequence of execution steps, beginning with some initial program state and ending when all threads are blocked or have exited, and/or when a bug is observed.
-\end{definition}
+{\bf Interleaving invariants.}
+We require the evaluation semantics to produce interleavings which fulfil several invariants,
+apart from the obvious ones ensuring correct synchronization and $\mathsf{malloc}$ discipline.
+
+Model checkers often include heuristics to identify blocking via open-coded $\mathsf{yield}$ loops,
+but we assume here that such patterns are implemented more tastefully with a condition-variable-like primitive built upon $\mathsf{deschedule}$.
+
+Most importantly, we assume that threads switch only at instructions identified by the set of {\em preemption-point predicates}, defined below.
 
@@
 
+\begin{definition}[Preemption point (PP) predicate]
+	A predicate on the execution state which identifies a class of instruction pairs between which we may force threads to switch.
+\end{definition}
+
+We use {\em synchronization API PPs} to denote the set of predicates
+%statically-available
+which occur immediately after any of $\mathcal{A}_{sync}$.
+%$\mathsf{mutex\_lock}$,
+%$\mathsf{mutex\_unlock}$,
+%$\mathsf{deschedule}$,
+%$\mathsf{make\_runnable}$,
+%$\mathsf{thread\_fork}$,
+%$\mathsf{thread\_exit}$,
+%$\mathsf{yield}$.
+Because no other instruction affects a thread's runnability, it is always possible to execute a program by switching the currently-executing thread only at synchronization API PPs.
+
+All data-race PP predicates will occur immediately {\em before} a $\mathsf{read}$, $\mathsf{write}$, or atomic swap.
+Other predicates are possible, though we will show they are irrelevant.
+
+When generating execution traces, the evaluation semantics is parameterized by a set of active PP predicates.
+As long as the set contains the synchronization API PPs,
+the execution shall switch threads only when a PP predicate holds.
+
 \begin{definition}[Preemption point (PP)]
 	%A PP is a code location between two instructions at which we may force threads to switch.
-	A PP is a predicate on the execution state which identifies a class of instruction pairs between which we may force threads to switch.
+	%A PP is a predicate on the execution state which identifies a class of instruction pairs between which we may force threads to switch.
+	A PP is any site in an interleaving at which a PP predicate evaluated to true.
 \end{definition}
 
-In the main paper, we identified PPs by predicates on the instruction pointer and current thread ID.
-Hence, the ``same'' PP may occur multiple times during an execution; for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
-%For the sake of this proof, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
-In our proofs, we will use the term ``PP'' to indicate specific instruction sites at which a PP predicate holds.
+As discussed above, all thread switches occur at PPs.
+However, an interleaving may also contain PPs at which the same thread continued running.
 
-We also use ``synchronization API PPs'' to denote the class of
-%statically-available
-PPs which occur immediately {\em after} any of
-$\mathsf{mutex\_lock}$,
-$\mathsf{mutex\_unlock}$,
-$\mathsf{deschedule}$,
-$\mathsf{make\_runnable}$,
-$\mathsf{thread\_fork}$,
-$\mathsf{thread\_exit}$,
-$\mathsf{yield}$.
-Because no other instruction affects a thread's runnability, it is always possible to execute a program by switching the currently-executing thread only at synchronization API PPs.
-
-All data-race PPs will occur immediately {\em before} a $\mathsf{read}$, $\mathsf{write}$, or $\mathsf{xchg}$.
+In the main paper, we used ``PP'' largely to denote what we now call ``PP predicates''.
+Hence, the ``same'' PP could occur multiple times during an execution.
+%for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
+In these proofs, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
 
 \begin{definition}[Transition]
-A sequence of execution steps from a program's evaluation between two preemption points (PPs).
+A sequence of execution steps from a program's evaluation between two PPs.
 \label{def:transition}
@@
 The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
-We also assume a trace of all memory accesses is available in the representation of transitions.
+%We also assume a trace of all memory accesses is available in the representation of transitions. %% It is, now.
 
 \begin{definition}[State space]
-Given a set of PP predicates, a state space is
-a set of interleavings representing all possible execution sequences which switch threads only on those PPs.
+	A state space $\mathcal{S}$ is a set of interleavings representing all execution sequences legal under a given set of PP predicates.
 \end{definition}
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 	\small
@@
 
-In our main paper, we refer to this relation as Limited HB.
+In our main paper, we refer to this relation (in conjunction with a lock-set analysis) as Limited HB.
 

### 2016-09-01 oopsla-16/proof/assumptions.tex
Import proof
oopsla proofs: fix discussion of which PPs happen before and/or after associated instructions

--- 2015-11-10 pldi-16/proof/assumptions.tex
@@
 
-{\bf No forcibly blocking other threads.}
-We assume the sync API behaves much like pthreads and/or message-passing:
-that is, a thread can cause itself to block on some condition,
-and other threads may wake it up,
-but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
+%{\bf No forcibly blocking other threads.}
+%We assume the sync API behaves much like pthreads and/or message-passing:
+%that is, a thread can cause itself to block on some condition,
+%and other threads may wake it up,
+%but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
+%
+%We require this because of the asymmetry between synchronization PPs (which occur {\em after}) synchronization operations)
+%and data-race PPs (which occur {\em before} their associated read/write).
+%For example, if a primitive $\mathsf{deschedule\_other}$ existed,
+%a transition such as $\mathsf{write}(a,v); \mathsf{deschedule\_other}(t_2)$ would lack a PP between the two instructions,
+%which could be necessary to expose some...
 
@@
 i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.
-% , such as for a program trying to find a counterexample for Sally's Conjecture.
-% TODO: cite some unproved mathematical conjecture by a woman, as an example of an undecidable program?
 

### 2016-09-01 oopsla-16/proof/di.tex
oopsla proofs: fix discussion of which PPs happen before and/or after associated instructions

@@
 %statically-available
-which occur immediately after any of $\mathcal{A}_{sync}$.
+which occur immediately before or after any of $\mathcal{A}_{sync}$.
 %$\mathsf{mutex\_lock}$,
@@
 
-All data-race PP predicates will occur immediately {\em before} a $\mathsf{read}$, $\mathsf{write}$, or atomic swap.
+All data-race PP predicates will occur immediately before a $\mathsf{read}$, $\mathsf{write}$, or atomic swap.
 Other predicates are possible, though we will show they are irrelevant.
@@
 
-Intuitively, MHB expresses the "cannot be reordered with" (or "enables") relation between transitions.
+Intuitively, MHB expresses the "cannot be reordered with" relation between transitions
+("enables" in DPOR terminology \cite{dpor}).
 Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
-Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.
+Such synchronization events include $\mathsf{thread\_fork}$, $\mathsf{make\_runnable}$,
+and sometimes but not always $\mathsf{mutex\_unlock}$.
 
@@
 so only one interleaving from each equivalence class need be tested to ensure total schedule coverage \cite{mazurkiewicz}.
+Equivalence is, of course, transitive.
 
@@
 {\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
-that set represents every single code location where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
-Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
+In summary, we are proving that when Iterative Deepening saturates the set of data-race PPs,
+that set represents every single instruction where a preemption could possibly affect the program's behaviour,
+and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings by preempting anywhere.
+A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
 but the technique's iterative nature will eventually find it.
 On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any PPs beyond the ones we consider are irrelevant.
+so any such PPs are irrelevant.
+
+Note that while we defined synchronization API PPs to occur both before and after any $\mathcal{A}_{sync}$ instruction,
+we will only add data-race PPs {\em before} their associated $\mathsf{read}$ or $\mathsf{write}$.
+Our proof requires preempting only before each racing instruction, not after,
+in order to fully saturate the PP set with all possible data races.
+Then, once saturated, preempting after each racing instruction would be extraneous,
+because all subsequent instructions before the next PP must be local operations or non-communicating reads/writes%.
+% Adding a PP both before and after could potentially expose some bugs sooner,
+% if the 'after' PP were redundant with another data-race PP but that other race would take forever to find.
+% So, I think it's an optimization to do it this way (although I haven't done a comparison experiment).
+%
+\footnote{Our implementation also optimizes the synchronization API PPs,
+generally preempting only {\em after} each synchronization instruction.
+$\mathsf{mutex\_lock}$ is the exception, as it can cause other threads to become blocked.
+All the others can only make blocked threads runnable, establishing MHB,
+which also provides MHB for the preceding transition of the invoking thread.
+Hence, coalescing those transitions does not exclude any possible interleavings.
+% If a third thread could "query" whether the 2nd thread is still blocked (or nonexistent),
+% that would break the soundness of this optimization, and you'd need to preempt before mx_unlock and thr_create after all.
+% XXX: In the implementation, such queries are actually possible with m_r or yield;
+% while they do not interact with mutex_unlock, I think an adversary could use them to interleave with thr_create/exit.
+}.
 

### 2016-09-01 oopsla-16/proof/assumptions.tex
oopsla proofs: brevity and clarity in intro material

@@
 
-This section documents the assumptions we make about the concurrency model, language model, and test environment,
-and discusses some limitations that may arise from these assumptions.
+This section documents our assumptions about the concurrency model, language model, and test environment,
+and discusses some limitations that may arise therefrom.
 
@@
 {\bf Maximal state space.}
-We assume the model checker has all static/hard-coded PPs enabled during this state space exploration,
+We assume the model checker has all
+%static/hard-coded
+synchronization API
+PPs enabled during this state space exploration,
 and that we are not limited by a pressing CPU budget.
-Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
+%Further, we assume that the static PPs include all lock/unlock/trylock operations on mutexes (or whatever other low-level locks are used) and also all higher-level sync primitives which can cause HB (either directly, or because they are built on top of mutexes).
+We assume that all synchronization primitives not listed in our system model are built upon them,
+or otherwise annotated for the MC to add additional PPs for them.
 
@@
 However, Iterative Deepening aims to test the most important interleavings with the time available,
-so in the case of not enough time, our point here is that continuing the current state space fits that goal best.
+so in the case of not enough time,
+%our point here is that
+continuing the current state space fits that goal best.
 %An alternate approach would be to accept the malloc-recycle PPs, giving them the lowest possible priority, just in case any CPUs would otherwise be idle with not enough jobs to run.
@@
 Both DPOR and data-race detection rely on this assumption,
-as any other way for threads to affect each other's behaviour would invisibly reduce independence and break soundness.
+as any other way for threads to affect each other's behaviour could invisibly reduce independence and break soundness.
 For brevity in these proofs, we refer to all thread communication as shared memory,
@@
 {\bf Locks are correct.}
-Because hybrid data-race detection uses lockset analysis to conclude that many access pairs could not possibly race,
-we assume that no preemption during a lock-protected critical section could cause a contending thread to make meaningful progress.
-This extends to use of disabling/enabling interrupts during kernel-space testing, which we model as a single global lock,
+Because hybrid data-race detection uses lockset analysis to exclude many candidate access pairs,
+%conclude that many access pairs could not possibly race,
+we assume that no preemption during a lock-protected critical section could cause a contending thread to make progress.
+This extends to use of disabling/enabling interrupts in kernel code, which we model as a single global lock,
 although we do not model RCU \cite{rcu}.
-Our data-race analysis also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically), although this is tangential to the proof.
 
@@
 which would expensively test all the main program's required locking properties in tandem with the program itself.
+
+We also assume that any unusual locking discipline, such as recursive mutexes or lock hand-off,
+is correctly annotated to the data-race analysis.
+Our implementation also models the behaviour of r/w locks and 1-initialized semaphores (the latter heuristically),
+although this is tangential to the proof.
 
@@
 {\bf Halting.}
-From a certain perspective, proving convergence is the same as proving completeness,
+From a certain perspective, convergence is the same as completeness,
 which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
@@
 we are already accepting that many tests will time out.
-We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
-Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
-If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
+We are concerned with the limited case of when \quicksand~{\em does} terminate with all state spaces completed.
+%, the verification is sound.
+
+Inversely, when we prove that Iterative Deepening will eventually find an arbitrary buggy interleaving,
+we assume that no intermediate state spaces contain nontermination bugs.
+If they do,
+%for the sake of testing/verification,
+we are satisfied with finding that bug instead.
 In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
-i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.
+i.e., it will never get stuck forever in an infinite loop without identifying a bug (necessarily accepting some false positives).
 
@@
 {\bf Malloc is a magic black box.}
-We assume the malloc implementation is correct (e.g., it won't double-allocate blocks), although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
+We assume the malloc implementation is correct (e.g., it won't double-allocate),
+although we don't assume any implementation details such as a tendency to reuse blocks or allocate adjacent ones.
 %particular allocation pattern regarding adjacency/coalescing/reuse.
-In fact, in our experiments we instruct our MC to ignore all potential PPs which might be inserted on malloc's internal mutex;
+In fact,
+%in our experiments we instruct our MC to
+our implementation ignores all potential PPs arising from malloc's internal mutex;
 in essence treating it as a ``magic primitive'', because we are not interested in verifying its implementation.
-Furthermore, we configured DPOR to ignore the internal heap metadata accesses
-when tracking shared memory conflicts to achieve greater state space reduction
-(i.e., if the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
+Furthermore, we configured DPOR to ignore any shared memory conflicts arising from internal heap metadata
+%when tracking shared memory conflicts
+to achieve greater state space reduction.
+(If the only consequence of reordering two transitions is malloc returning different addresses, we consider them independent).
 %for the purpose of avoiding malloc-only shared memory conflicts.
@@
 {\bf Sharing heap addresses.}
-Finally, we assume that the only way the program can obtain heap addresses is through the return value of malloc().
+Finally, we assume that the only way the program can obtain heap addresses is through the return value of $\mathsf{malloc}$.
 Because we are testing C programs, any bizarre violations of this assumption are technically possible,
@@
 
-For Section \ref{sec:proof}, we further assume a malloced block's address cannot be obtained through arithmetic on the address of a different block; in Section \ref{sec:owned} we show how to account for this case by relaxing the previous ``black box'' assumption.
+In Section \ref{sec:proof}, we further assume a $\mathsf{malloc}$ed block's address cannot be obtained through arithmetic on the address of a different block;
+in Section \ref{sec:owned} we account for this case by relaxing the previous ``black box'' assumption.

### 2016-09-01 oopsla-16/proof/di.tex
oopsla proofs: brevity and clarity in intro material

@@
 As long as the set contains the synchronization API PPs,
-the execution shall switch threads only when a PP predicate holds.
+%the execution shall switch threads only when a PP predicate holds.
+the thread-switch invariant discussed above will hold.
 
@@
 
-In the main paper, we used ``PP'' largely to denote what we now call ``PP predicates''.
+In the main paper, ``PP'' referred to what we now call ``PP predicates''.
 Hence, the ``same'' PP could occur multiple times during an execution.
 %for example, {\tt mutex\_lock()} may be called from {\tt foo()} and later again from {\tt bar()}.
-In these proofs, we separate such cases into multiple unique PPs: each PP is simply a label denoting the end of a single transition.
+In these proofs, we separate such cases into multiple unique PPs:
+each PP is simply a label denoting the boundary between two transitions.
 
@@
 \end{definition}
-We require the invariant that each transition's instructions are associated with exactly one thread. (That is, the set of PPs always includes all thread switches.)
-The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
+The thread-switch invariant guarantees that each transition's instructions are associated with exactly one thread.
+%The set of synchronization API PPs provides this invariant, and all other PP sets in these proofs will be supersets of those.
 %We also assume a trace of all memory accesses is available in the representation of transitions. %% It is, now.
@@
 
-Intuitively, MHB expresses the "cannot be reordered with" relation between transitions
-("enables" in DPOR terminology \cite{dpor}).
+Intuitively, MHB expresses when two transitions cannot be reordered
+(the ``enables'' relation in DPOR terminology \cite{dpor}).
 Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
@@
 
-In our main paper, we refer to this relation (in conjunction with a lock-set analysis) as Limited HB.
+Our main paper refers to this relation (in conjunction with a lock-set analysis) as Limited HB.
 
@@
 \begin{itemize}
-	\item The intersection of both threads' locksets is empty (i.e., the threads do not hold the same lock during each access), and
-	\item The threads' transitions are not related by MHB.
+	\item The intersection of both threads' locksets is empty (i.e., the same lock does not protect both accesses), and
+	\item The containing transitions are not related by MHB.
 \end{itemize}
@@
 For brevity, we now use ``data race'' to refer both to true races and to potential data-race candidates identified by MHB.
-In this proof we are concerned solely with candidates, and whether they can be observed to race or are false positives.
-It is up to the MC to decide whether true data races are benign or buggy.
+%In this proof we are concerned solely with candidates, and whether they can be observed to race or are false positives.
+%It is up to the MC to decide whether true data races are benign or buggy.
 
 \begin{definition}[False positive data race]
-	An apparent data race that cannot be observed in the opposite order from what was actually executed.
+	An apparent data race that cannot be executed in the opposite order from what was observed.
 \end{definition}
 
-False positives are caused when some data dependency based on some other shared state, invisible to the data-race analysis,
+False positives are caused when some data dependency based on some other shared state %, invisible to the data-race analysis,
 changes some variable values when the threads are reordered, such that the memory addresses no longer collide.
@@
 
-Recent work \cite{sparc-ssm} has proposed hardware techniques for detecting many classes of stale heap pointer accesses, including the one shown in Figure \ref{fig:recycle-bug}.
-This approach could be combined with stateless model checking in future work to identify such bugs immediately,
+Recent work \cite{sparc-ssm} proposed hardware techniques for detecting many classes of stale heap pointer accesses, including the one shown in Figure \ref{fig:recycle-bug}.
+Future work could combine this approach with MC to identify such bugs immediately,
 rather than requiring Iterative Deepening to explore new state spaces corresponding to the data race.
-However, if the {\tt malloc} call were in thread 1 instead of thread 2, the bug would still be nondeterministic, requiring stateless model checking to expose.
+However, if the {\tt malloc} call were in thread 1 instead of thread 2, the bug would still be nondeterministic, requiring MC to expose.
 
@@
 
-In this section we provide (hopefully) intuitive summaries of both our proof goals.
+This section provides (hopefully) intuitive summaries of our proof goals.
 %for readers not interested in double-checking the proofs' internal structure.
@@
 {\bf Intuition for Iterative Deepening convergence.}
-In summary, we are proving that when Iterative Deepening saturates the set of data-race PPs,
-that set represents every single instruction where a preemption could possibly affect the program's behaviour,
-and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings by preempting anywhere.
+%In summary, we are proving
+We will prove
+that when Iterative Deepening saturates the set of data-race PPs,
+that set represents every instruction where a preemption could possibly affect the program's behaviour
+Hence, completing the associated state spaces is as strong a verification as testing all possible thread interleavings by preempting anywhere.
 A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
@@
 {\bf Intuition for malloc-recycle soundness.}
-In summary, we are proving that if a malloc-recycle-pattern data race is a true race, rather than a false positive,
-then DPOR is guaranteed to ``reorder away the free and re-malloc''.
-In other words, DPOR's exploration will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
+%In summary, we are proving
+We prove that if a malloc-recycle-pattern data race is not a false positive, %a true race, rather than a false positive,
+then DPOR %is guaranteed to ``reorder away the free and re-malloc''.
+%In other words, DPOR's exploration
+will eventually interleave threads in such a way that the malloc-recycle pattern will disappear,
 while the access pair remains for the data-race detector to find, as shown in Figure~\ref{fig:recycle-goal}.
 Hence, in the same state space where the malloc-recycle data race was found, if it's a true race, the same race will also appear without the recycle pattern.
-So if that race hides a failure bug (assertion, segfault, etc.), Iterative Deepening will still be led to the necessary preemption point to find that bug.
+So if that race can lead to a failure, Iterative Deepening will still be led to the necessary preemption point to find it.

### 2016-09-02 oopsla-16/proof/di.tex
oopsla proofs: fix some oopses in system model

@@
 \begin{eqnarray*}
-	\mathcal{A} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg(a,v)} \\
+	\mathcal{A} &::=& v \leftarrow \mathsf{read}(a) \quad | \quad \mathsf{write}(a,v) \quad | \quad \mathsf{xchg}(a,v) \\
 		&|& a \leftarrow \mathsf{malloc}(n) \quad | \quad \mathsf{free}(a) \\
-		&|& \textit{local} \\
-		&|& \mathcal{A}_{sync}
+		&|& \mathcal{A}_{local} \quad | \quad \mathcal{A}_{sync}
 \end{eqnarray*}
@@
 $\mathsf{read}$, $\mathsf{write}$, and $\mathsf{xchg}$ access global or heap memory shared by all threads, indicated by some address $a$.
-(Other atomic swap operataions are omitted for brevity.)
+(Other atomic swap operations are omitted for brevity.)
 $\mathsf{malloc}$ and $\mathsf{free}$ provide access to fresh memory accessible by all threads. % free is irrelevant to the system model in terms of expressive power
@@
 {\bf Local state.}
-$\textit{local}$ represents any thread-local instruction, such as modifying local variables, flow control, function calling, and assertions. %, and side effects.
-We omit a detailed list for brevity, as we do not need to reason about their nature in these proofs.
+$\mathcal{A}_{local}$ represents any thread-local instruction, such as modifying local variables, flow control, function calls, and assertions. %, and side effects.
+We omit a detailed list for brevity, as we do not need to reason about them in these proofs.
 %For brevity, we omit a detailed list, although note that $\mathsf{call}$ would be implemented by modifying the instruction stream $[\mathcal{I}]$ and creating a new frame on the stack of local variables.

### 2016-09-02 oopsla-16/proof/recycle.tex
oopsla proof: address camreadies
oopsla proofs: revise recycle proof for prosaic issues

@@
 
-We seek to prove that ignoring malloc-recycle data races cannot cause DPOR + Iterative Deepening to miss a bug that could be found by using the malloc-recycle race as a PP.
-Hence, our soundness statement is as follows:
+We seek to prove that ignoring malloc-recycle data race candidates cannot cause DPOR + Iterative Deepening to miss a bug
+that could be found by using the race as a PP predicate. Our soundness statement is as follows:
 
@@
 The keen reader might ask here, what if T1 contains some extra spurious malloc calls (not related to $a_1$) that affect what T2's malloc returns after being reordered?
-However, these could at best either cause {\tt x}'s memory to be recycled slightly differently (not affecting the proof), or not at all (which simply changes some cases to be immediate use-after-frees).
+These could at best either cause {\tt x}'s memory to be recycled differently (not affecting the proof), or not at all (which simply causes immediate use-after-free).
 %If such a malloc could affect $a_2$, the data race would be a false positive after all.
-In general, adding spurious mallocs that could affect $a_1$ or $a_2$ could only convert the program back into a false-positive scenario;
-and adding more spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
-So, WLOG, we say that the only relevant events are the ones we mention explicitly.
+In general, extra spurious mallocs that could affect $a_1$ or $a_2$ could only convert the program back into a false-positive scenario;
+and extra spurious synchronization events could only make it more easy to find PPs we need to trigger the reordering.
+So we can safely assume the only relevant events are the ones we mention explicitly.
 
-By our last assumption, there must also be an ``original malloc'' which allocated the heap block to begin with.
+By our last assumption, there must also be an ``original malloc'' which allocated the block to begin with.
 We must ask, which thread did the malloc which returned $a_1$'s address in the first place?
-Because of our last assumption, we know that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
+Our last assumption provides that the other thread must obtain that address through some communication mechanism (which we'll reason about later).
 
@@
 
-We ask: was there a synchronization event (unlock or message-pass) between the publish action and $a_1$?
-If so, then by the maximal state space assumption, the event will be a PP, and we are done.
-If not, then  T1's lockset during publish will be the same as the lockset during $a_1$ (and the fact that there is no MHB cannot change).
+If there was a synchronization event between the publish action and $a_1$,
+then the maximal state space assumption provides the necessary PP, and we are done.
+%We ask: was there a synchronization event (unlock or message-pass) between the publish action and $a_1$?
+%If so, then by the maximal state space assumption, the event will be a PP, and we are done.
+%If not,
+Otherwise, T1's lockset will be the same during publish and during $a_1$ (and the MHB-ness cannot change).
 For T1's publish to reach T2, they must access the same memory ({\em outside} of the block containing $a_1$; T2 doesn't have that yet),
 which we'll call $p$.
-By the previous two statements, $p$ must be a data race of its own.
+Hence, $p$ must be a data race of its own.
 %(whether by writing {\tt x}'s address directly into the shared memory, or by sharing a pointer to a data structure containing {\tt x} -- doesn't matter which)
@@
 
-Because it's possible for $p$ to {\em also} be a malloc-recycle data race,
+Because $p$ may also be a malloc-recycle data race,
 as shown in Figure~\ref{fig:induction},
-we do not necessarily get the PP for free here.
+we do not necessarily get the PP for free.
 In this case we need to prove that DPOR will likewise reorder any intermediate malloc-recycle pattern to generate the PP we need\footnote{
@@
 		%DPOR will test some interleaving in which T1's
-		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access.
-		Hence by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
+		Hence a data race PP will be generated on the \publish{n-1}\texttt{->ptr} access,
+		and by Definition~\ref{def:dpor} and Lemma~\ref{lem:reorder},
 		DPOR will reorder T1's and T2's subsequent accesses to \publish{n} sufficiently to place a PP on them.
@@
 Hence, even if the accesses by which T1 shares {\tt x} with T2 appear in a different malloc-recycle pattern,
-a PP will be identified on the publish location $p$.
-Hence by Definition~\ref{def:dpor} DPOR will reorder T2's execution to just after the publish action.
+a PP will be identified on the publish location $p$,
+%Hence by Definition~\ref{def:dpor}
+and DPOR will reorder T2's execution to just after the publish action.
 % XXX: Landslide does not do this! It only puts a PP immediately *before* the access. That's incompatible with this, which needs you to put one both before and after.
 As long as T2's execution occurs after the publish, it will receive the same value for its $a_2$, so the location of the data race does not change.
-Lemma~\ref{lem:reorder} finishes.
+Lemma~\ref{lem:reorder} concludes.
 %Hence for any type of publish action, there will be a PP between it and $a_1$, hence DPOR will reorder $a_2$ to before $a_1$ (without changing the address at which $a_2$ occurs), and Lemma~\ref{lem:reorder} finishes.
@@
 If we relax the ``sharing heap addresses'' assumption, there is another way to share the allocation's address without T1 and T2 communicating outside of $a_1$/$a_2$.
-One thread can overrun a {\em different} heap block which happened to be adjacent to the one containing $a_1$/$a_2$ (call them the ``neighbour block'' and ``real block'' respectively).
+One thread can overrun a {\em different} heap block adjacent to the one containing $a_1$/$a_2$ (call them the ``neighbour block'' and ``real block'' respectively).
 Figure~\ref{fig:overrun} shows an example.
@@
 
-In the evaluation of our main paper, we used the full ``sharing heap addresses'' assumption to heuristically reduce state space size,
-and not test reorderings of threads which could only change the addresses allocated by {\tt malloc}.
-This restricted the scope of our tests to exclude testing for heap-overflow bugs.
-We consider this justified because new techniques \cite{sparc-ssm} have recently emerged for finding such bugs quickly without the need for data-race PPs.
-However, for users wishing to test for this class of bug simultaneously with other concurrency bugs, we show in this section how to strengthen the configuration of DPOR to cover the weakened assumption.
+In our evaluation, we used the full ``sharing heap addresses'' assumption to heuristically reduce state space size,
+skipping reorderings which could only change the addresses allocated by $\mathsf{malloc}$.
+This restricted our tests' scope to exclude such heap-overflow bugs.
+We consider this justified because recent techniques \cite{sparc-ssm} can find such bugs quickly without the need for data-race PPs.
+However, for users wishing to test for this class of bug and concurrency bugs simultaneously,
+we show now how to strengthen the configuration of DPOR to cover the weakened assumption.
 
@@
 a DPOR which ignores malloc's internal metadata accesses would still be unsound with respect to these bugs.
-We illustrate this in Figure~\ref{fig:overrun-notenough}: even if there were PPs in arbitrary places, the two threads' transitions do not collide in this interleaving on anything but {\tt malloc}'s internal metadata, and hence DPOR would not attempt to reorder them.
-Hence, what we have already proved so far can be viewed as showing that suppressing malloc-recycle candidates is sound with respect only to the classes of bugs which DPOR is already sound to.
+We illustrate this in Figure~\ref{fig:overrun-notenough}:
+even with PPs in arbitrary places, the two threads' transitions conflict only on $\mathsf{malloc}$'s internal metadata,
+and hence DPOR would not attempt to reorder them.
+Hence, our proofs so far show that suppressing malloc-recycle candidates is sound with respect to the classes of bugs which DPOR is already sound to.
 
@@
 \caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-Because the mallocs have been reordered, they may no longer collide, which wrongly appears to be a false positive.}
+The mallocs have been reordered and may no longer collide, which wrongly appears to be a false positive.}
 \label{fig:overrun-notenough}
@@
 \end{tabular}
-\caption{Goal interleaving of Figure~\ref{fig:overrun}.
-To ensure collision, the sequence of malloc calls which produce $a_1$ and $a_2$ must not be disrupted, compared to the trace which originally exposed the malloc-recycle race.}
+	\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls producing $a_1$ and $a_2$ must not be disrupted compared to the original interleaving.}
+	%, compared to the trace which originally exposed the malloc-recycle race.}
 \label{fig:overrun-goal}
@@
 we must strengthen our configuration of DPOR as follows:
-accesses from {\tt malloc}'s internal implementation will not be ignored when computing the shared memory conflict relation,
-and we will automatically insert a synchronization API PP at the start and end of each {\tt malloc} and {\tt free} call.
-% TODO CAMREADY: uncomment footnote (it's commented because it seems like it's advertising landslide too much)
+accesses from $\mathsf{malloc}$'s internal implementation are not ignored when computing shared memory conflicts,
+and the synchronization API PP predicates are extended to include the start and end of each $\mathsf{malloc}$ and $\mathsf{free}$ call.
 %\footnote{
@@
 \begin{proof}
-WLOG, let T1's access in the original malloc-recycle race have come first.
+WLOG, let T1's access in the original malloc-recycle race occur first.
 We require that our strengthened DPOR will reorder T2's racing access to before T1's, such that both still occur on the same address.
-There will be a PP in T1's execution between its {\tt malloc} call and the subsequent racing access.
+There will be a PP in T1's execution between its $\mathsf{malloc}$ call and the subsequent racing access.
 If there are no other memory conflicts between T1 and T2, then by the soundness of DPOR, this PP suffices to reorder without changing the address.
@@
 
-% TODO CAMREADY cite landslide
-%%\cite{landslide} % uncomment post-double-blind review
-%
 %This spells QED so we are done.

### 2016-09-02 oopsla-16/proof/convergence.tex
oopsla proofs: revise convergence proofs for prosaic issues

@@
 
+\newcommand\naive{na\"{i}ve\xspace}
 We seek to prove that, given enough time, Iterative Deepening offers the same coverage of thread interleavings that could be achieved by preempting at every single instruction.
-We'll call the latter the {\em na\"{i}ve state space}, and call the condition of testing all its interleavings {\em convergence}.
-(It could also be called soundness, from the perspective of viewing Iterative Deepening as a search re-ordering heuristic that doesn't miss any interleavings.)
+We'll call the latter the {\em \naive state space}, and call the condition of testing all its interleavings {\em convergence}%
+\footnote{
+It could also be called soundness, taking the perspective that Iterative Deepening is a search ordering heuristic that doesn't miss any interleavings.}.
 Hence, our convergence statement is as follows:
@@
 \begin{theorem}[Total verification]
-	If Iterative Deepening fully saturates its set of data-race PPs and completes all associated state spaces,
+	If Iterative Deepening fully saturates its data-race PP predicates and completes all associated state spaces,
 	it serves as a verification of all possible thread schedules of the given test program.
@@
 For the reader who likes to avoid non-constructive proofs \cite{vargomax},
-note that we are using the constructive contrapositive direction:
+note that we use the constructive contrapositive direction:
 Theorem~\ref{thm:convergence} is $A \rightarrow B$ and Theorem~\ref{thm:totalverif} is $\neg B \rightarrow \neg A$.}:
@@
 \begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
+	If a bug is exposed by an interleaving in the \naive state space, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
 	\label{thm:convergence}
@@
 \]
-be an interleaving trace, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$ at its end.
+be an interleaving, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$ at its end.
 We will use the following notation:
 \begin{itemize}
-	\item $\ppinstr{\mathcal{I}}{\alpha}{i}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$ (the {\em next} transition), i.e., the instruction of $\alpha$ that \pai preempted just before.
+	\item $\ppinstr{\mathcal{I}}{\alpha}{i}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$ (the {\em next} transition),
+		i.e., the instruction of $\alpha$ that $\pai$ preempted just before.
 	\item $\ppothers{\mathcal{I}}{\alpha}{i}$ indicates the set of all transitions by other threads between $\tai$ and $t_{\alpha{}(i+1)}$.
-	\item $\conflicts{t_\alpha}{t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
+	\item $\conflicts{t_\alpha}{t_\beta}$ indicates the set of shared memory conflict pairs between $t_\alpha$ and $t_\beta$.
+		%memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair,
+		%one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write;
+		%i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
 \end{itemize}
@@
 
-Iterative Deepening will not, of course, preempt on exactly the same instructions that an arbitrary naive interleaving would,
+Iterative Deepening will not, of course, preempt on exactly the same instructions that an arbitrary \naive interleaving would,
 as it is only capable of preempting on data races and sync API boundaries.
-Thus, our first task is to show that all naive interleavings have an equivalent interleaving with only data-race and sync API PPs.
+Thus, our first task is to show that all \naive interleavings have an equivalent interleaving which includes only data-race and sync API PPs.
 
 \begin{definition}[Relevant preemption point]
-	We say a PP $\pai$ is {\em relevant} if either $\ppinstr{\mathcal{I}}{\alpha}{i}$ is a sync API boundary (including a voluntary yield),
+	We say a PP $\pai$ is {\em relevant} if either
+	%$\ppinstr{\mathcal{I}}{\alpha}{i}$ is a sync API boundary (including a voluntary yield),
+	$\pai$ is a synchronization API PP,
 	or if
@@
 \begin{proof}
-	Let $\pai$ be the first irrelevant PP of a naive interleaving $\mathcal{I}$.
+	Let $\pai$ be the first irrelevant PP of a \naive interleaving $\mathcal{I}$.
 	We ask, did some thread among $\ppothers{\mathcal{I}}{\alpha}{i}$ conflict with any instruction from $t_{\alpha{}(i+1)}$, even though there was no conflict with $\ppinstr{\mathcal{I}}{\alpha}{i}$?
@@
 			% chi for conflict.
-		\item If $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} \ne \emptyset$, or if $\tai$ contains a sync API boundary instruction that's not already a PP,
+		\item If $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} \ne \emptyset$, or if $\tai$ contains a sync API instruction that's not already a PP,
 			then let $\chi$ be either the first instruction by $t_{\alpha{}(i+1)}$ among the conflicts or the first sync API instruction, whichever comes first.
@@
 
-			In other words, we have simply reordered $\ppothers{\mathcal{I}}{\alpha}{i}$ to between $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ and $\chi$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\chi$.
-			We know it must be possible to reorder $\ppothers{\mathcal{I}}{\alpha}{i}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{t_{\alpha{}(i+1)}}{\chi}$, hence the other threads' runnability cannot be affected.
+			In other words, we reorder $\ppothers{\mathcal{I}}{\alpha}{i}$ to between $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ and $\chi$,
+			removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\chi$.
+			It must be possible to reorder $\ppothers{\mathcal{I}}{\alpha}{i}$ to after $\alpha$'s execution because no synchronization occurs during $\pppfx{t_{\alpha{}(i+1)}}{\chi}$, hence the other threads' runnability cannot be affected.
 			Likewise $\pppfx{t_{\alpha{}(i+1)}}{\chi}$ is independent with $\ppothers{\mathcal{I}}{\alpha}{i}$, so by the soundness of DPOR, $\mathcal{I}' \equiv \mathcal{I}$.
-		\item Otherwise, $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} = \emptyset$, and no instructions among $t_{\alpha{}(i+1)}$ are a sync API boundary.
+		\item Otherwise, $\conflicts{\ppothers{\mathcal{I}}{\alpha}{i}}{t_{\alpha{}(i+1)}} = \emptyset$,
+			and no instructions among $t_{\alpha{}(i+1)}$ are synchronization.
 			Then we output the new interleaving:
@@
 			\]
-			In other words, we have reordered {\ppothers{\mathcal{I}}{\alpha}{i}} with the entire next transition by thread $\alpha$.
-			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
+			In other words, we reorder {\ppothers{\mathcal{I}}{\alpha}{i}} with the entire next transition by thread $\alpha$.
+			This must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
 			%&$(t_{\alpha{}(i+1)}) \setminus \pppfx{t_{\alpha{}(i+1)}}{\chi}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
@@
 Now we must show that, starting from a sync-PP-only state space,
-Iterative Deepening will eventually interleave threads sufficiently to expose all data races which are used as PPs in any buggy fully-relevant interleaving.
-The challenge of this proof is that some data-race candidates are nondeterministic to find;
-i.e., they may be hidden in an obscure control flow path which requires a prior preemption on a different data race to expose,
+Iterative Deepening will eventually
+%interleave threads sufficiently to expose
+encounter all data races which are used as PPs in any buggy fully-relevant interleaving.
+The challenge
+%of this proof
+is that some data-race candidates are nondeterministic to find;
+i.e., they may be hidden in
+%an obscure control flow path
+control flow which requires a prior preemption on a different data race to expose,
 as shown in Figure~\ref{fig:nondet-dr}.
 Hence, the maximal state space of statically-available sync API PPs will not necessarily uncover all possible data-race PPs;
-Iterative Deepening may need to iterate through some data-race state spaces before finding certain nondeterministic races.
+%Iterative Deepening
+we may need to iterate through some data-race state spaces before finding certain nondeterministic races.
 
-\begin{figure}[t]
+\begin{figure}[h]
 	\small
@@
 	\end{tabular}
-	\caption{Not all data races will immediately be uncovered by interleaving threads around sync API PPs alone.
-Here, preempting during the race on line 1 is necessary to force the program into the control flow which can race on line 5.}
+	\caption{Not all data races will immediately be uncovered by
+	%interleaving threads around
+	sync API PPs alone.
+	Here, preempting during line 1's race is necessary to force the program
+	%into the control flow which can race on line 5.
+	to execute the racy line 5.}
 	\label{fig:nondet-dr}
@@
 \begin{definition}[Reachable preemption point]
-	A PP $\pai$ such that either $\ppinstr{\mathcal{I}}{\alpha}{i}$ is part of a reachable data race, or a sync API boundary.
+	A PP $\pai$ such that either $\ppinstr{\mathcal{I}}{\alpha}{i}$ is part of a reachable data race, or a synchronization instruction.
 \end{definition}
@@
 \begin{proof}
-Let $\mathcal{I} = [(t_{\alpha{}1}, p_{\alpha{}1}), ...]$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
-We proceed by induction on the PPs according to the order of their preemptions.
-(Note that this is not necessarily the same as the order of the racing instructions $\ppinstr{\mathcal{I}}{\alpha}{i}$, which occur in $t_{\alpha{}(i+1)}$, not in \tai.)
+Let $\mathcal{I} = [(t_{\alpha{}1}, p_{\alpha{}1}), ...]$ be the fully-relevant interleaving and PP sequence in the premise which exposed a bug.
+We proceed by induction on the PPs according to their order in $\mathcal{I}$%the order of their preemptions.
+\footnote{Note that this is not necessarily the same as the order of the racing instructions
+	$\ppinstr{\mathcal{I}}{\alpha}{i}$, which occur in $t_{\alpha{}(i+1)}$, not in \tai.
+}.
 For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbh$ with $\tbh \prec \tai \in \mathcal{I}$, $\pbh$ is reachable.
@@
 {\bf Finding the next reachable data-race PP.}
-Now, we show by induction that a data race on $\ppinstr{\mathcal{I}}{\alpha}{i}$ will be reachable.
+Now, we show by induction that a data race on $\ppinstr{\mathcal{I}}{\alpha}{i}$ is reachable.
 We assume that a state space $\mathcal{S}$ containing $\mathcal{K}$,
@@
 will be reachable.
-(In the base case, the number of new data-race PPs is 0, and $\mathcal{K}$'s reachability is justified above.)
+(In the base case, $n=0$, and $\mathcal{K}$'s reachability is justified above.)
 We must show that in this state space, if the $\ppinstr{\mathcal{I}}{\alpha}{i}$ data race is not reachable,
-a new unique data-race PP will instead be reachable.
+a new unique data-race PP will be reachable instead.
 
-By the definition of fully-relevant interleaving, $\mathcal{I}$ guarantees that some other thread $\omega$ can execute a data-racing instruction with $\ppinstr{\mathcal{I}}{\alpha}{i}$.
+By the definition of relevance, %fully-relevant interleaving,
+$\mathcal{I}$ guarantees that some other thread $\omega$ can execute a data-racing instruction with $\ppinstr{\mathcal{I}}{\alpha}{i}$.
 %
-By the soundness of DPOR, if a program behaviour is possible by interleaving threads at the boundaries of
+By the soundness of DPOR, if a program behaviour is possible by interleaving at the boundaries of
 the given transitions, that interleaving will be tested.
 By contrapositive,
-because $\omega$'s data-racing instruction was not tested in $\mathcal{S}$, one or more necessary sites of interleaving must be in the middle of some transition, rather than at its boundaries.
+because $\omega$'s data-racing instruction was not tested in $\mathcal{S}$,
+one or more necessary interleaving sites must be in the middle of some transition, rather than all at boundaries.
 
@@
 
-We show this by contradiction:
+We show this by contradiction.
 % Sully points out that it's constructive because the set of interleavings / data races is finite.
@@
 Then, the above assumption also applies to $\mathcal{S}'$,
-which shows that for any {\em pair} of transitions such as $\tbk$, adding two new PPs cannot expose new program behaviour.
-Hence, inductively, no set of new PPs of any size would expose new program behaviour not already exposed in $\mathcal{S}$.
+showing that for any {\em pair} of transitions such as $\tbk$, adding two new PPs cannot expose new program behaviour.
+Inductively, no set of new PPs of any size would expose new behaviour not already exposed in $\mathcal{S}$.
 However, the instruction by $\omega$ which conflicts with $\ppinstr{\mathcal{I}}{\alpha}{i}$ was not observed in $\mathcal{S}$,
@@
 {\bf Reaching $\mathbf{p_{\alpha{}i}}$.}
-Hence either a data-race with $\ppinstr{\mathcal{I}}{\alpha}{i}$ will be identified,
+Hence either a data-race will be identified including $\ppinstr{\mathcal{I}}{\alpha}{i}$,
 or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
@@
 \begin{theorem}[Convergence of Iterative Deepening]
-	If a bug is exposed by an interleaving in the na\"{i}ve state space, Iterative Deepening will eventually test an interleaving which exposes the same bug.
+	% restatement of same theorem from beginning
+	If a bug is exposed by an interleaving in the \naive state space, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
 \end{theorem}

### 2016-09-02 oopsla-16/proof/convergence.tex
oopsla proofs: some more strats to save the frames

@@
 \\
-		& Thread 1 (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
+		& {\bf Thread 1} (Thread 2 similar, with {\tt t1} and {\tt t2} vars swapped) \\
 	1 & \texttt{\hilight{brickred}{x = x + 1;}} \\
@@
 %the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
-By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption\footnote{
-		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
-		Because we are concerned with verifying the absence of bugs, this also suffices for convergence, although we have left it implicit for brevity.}.
+By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
+%\footnote{
+%	In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces,
+%	anticipating the same bug.
+%	%Because we are concerned with verifying the absence of bugs, this also suffices for convergence, although we have left it implicit for brevity.
+%	This also suffices for the sake of verification.}.
 Then DPOR will find the buggy interleaving within this state space.

### 2016-09-02 oopsla-16/proof/di.tex
oopsla proofs: some more strats to save the frames

@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & \texttt{\hilight{brickred}{my\_x->foo = ...;}} & \\
@@
 	\\
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & \texttt{publish(x1);} & \\
@@
 \end{tabular}
-\caption{Goal interleaving, which reorders the adversarial program's threads away from the pattern, while the data race remains.}
+\caption{Goal interleaving, reordering the adversarial threads away from the pattern, while the data race remains.}
 \label{fig:recycle-goal}

### 2016-09-02 oopsla-16/proof/recycle.tex
oopsla proofs: some more strats to save the frames

@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 
@@
 %Hence for any type of publish action, there will be a PP between it and $a_1$, hence DPOR will reorder $a_2$ to before $a_1$ (without changing the address at which $a_2$ occurs), and Lemma~\ref{lem:reorder} finishes.
-
 \end{proof}
@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & & \texttt{z = malloc(42);} \\
@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{x1 = malloc(...);} & \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+	%2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
+	2 & & \texttt{x2 = \&z[50];} \\
+	3 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	4 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	5 & \texttt{x1 = malloc(...);} & \\
+	6 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	7 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 \end{tabular}
@@
 \begin{tabular}{rll}
-	& Thread 1 & Thread 2 \\
+	& {\bf Thread 1} & {\bf Thread 2} \\
 	1 & & \texttt{z = malloc(42);} \\
-	2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
-	3 & & \texttt{x2 = \&z[50];} \\
-	4 & \texttt{x1 = malloc(...);} & \\
-	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+	%2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
+	2 & & \texttt{x2 = \&z[50];} \\
+	3 & \texttt{x1 = malloc(...);} & \\
+	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	6 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	7 & \texttt{\hilight{olivegreen}{free}(x1);} \\
 \end{tabular}

### 2016-09-03 pldi-16/testcases/testcases.tex
fix testcases document for public release

@@
 \usepackage{graphicx}
+\usepackage{flushend}
 %\usepackage{setspace}
@@
 implemented by students in Berkeley's and the University of Chicago's undergrad OS classes,
-CS162 and CS230 respectively \cite{pintos}.
+CS162 and CMSC 23000 respectively \cite{pintos}.
 They use the internal kernel API provided by the basecode, available online at \cite{pintos-github}.

### 2016-09-06 oopsla-16/proof/proof.tex
oopsla proofs: revise convergence proofs for prosaic issues
Finish oopsla proofs for TRification

@@
 \usepackage{graphicx}
+\usepackage{xspace}
+\usepackage{cmu-techreport}
+\usepackage{flushend}
 %\usepackage{setspace}
@@
 %\title{Soundness Proof for Eliminating Malloc-Recycle Data Races in Stateless Model Checking}
-\title{Soundness Proofs for Iterative Deepening}
-\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
+\title{\bf Soundness Proofs for Iterative Deepening}
+%\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
+\author{Ben Blum}
+
+% tech report stuff
+\trnumber{CMU-PDL-16-103}
+\keywords{concurrency, stateless model checking, data race analysis, verification}
+
+\abstract{
+	\renewcommand{\baselinestretch}{1.0}
+	\small
+	\em
+	%We have developed the
+	The Iterative Deepening algorithm allows stateless model checkers to adjust preemption points on-the-fly.
+	%incorporate dynamic data-race analysis during testing.
+	%The algorithm
+	It uses dynamic data-race detection to avoid necessarily preempting on every shared memory access,
+	and ignores false-positive data race candidates arising from certain heap allocation patterns.
+	%When an Iterative Deepening test reaches completion,
+	An Iterative Deepening test that reaches completion soundly verifies all possible thread interleavings of that test.
+}
+
+\support{\small{
+	{\bf Acknowledgements:}
+	Michael J. ``Sully'' Sullivan, Carlo Angiuli, and Nathaniel Filardo double-checked the proof and provided valuable feedback on notation style and the flow of explanation.
+Thanks as well to the anonymous OOPSLA reviewer \#4, whose suggestions helped improve the formality of our definitions.
+%This work was supported in part by
+We thank
+the member companies of the PDL Consortium (Broadcom,
+Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
+Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
+Tintri, Two Sigma, Uber, Veritas and Western Digital)
+and by the U.S. Army Research Office under grant number W911NF0910273.
+}}
 
 \maketitle
-\begin{abstract}
-In our paper we show the effectiveness of combining dynamic data-race detection \cite{eraser,hybriddatarace} with stateless model checking \cite{verisoft,dpor}.
-Our approach involves adding new state spaces to explore each time a new data-race candidate is found.
-This document presents two proofs concerning the technique.
+
+\section{Introduction}
+In our OOPSLA paper \cite{quicksand} we introduced Iterative Deepening,
+an algorithm which combines
+stateless model checking \cite{verisoft,dpor}
+and
+data-race analysis \cite{eraser,hybriddatarace}.
+It dynamically adds new state spaces to test each time a data-race candidate is found.
+We present here two proofs concerning the technique.
 
@@
 
-Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic we discuss in the paper.
-Despite many powerful reduction techniques, the state spaces are still exponentially sized,
-so any way to avoid exploring some in advance is obviously beneficial.
-We prove that for the special case of {\em malloc-recycle} false positives, it is safe to eliminate these immediately upon discovering them, without bothering to explore their associated state spaces.
+Second, we prove the soundness of the {\em malloc-recycle} false-positive-elimination tactic discussed in the paper.
+Despite powerful reduction techniques, each state space is still exponentially sized,
+%so any way to avoid exploring some in advance is obviously beneficial.
+so we wish to avoid exploring any we could prove to be redundant.
+We prove that it is safe to eliminate malloc-recycle data race candidates immediately, %upon discovering them,
+without bothering to explore their associated state spaces.
 
-	This document is supplemental material to our main paper \cite{quicksand}; we assume the reader is already familiar with our motivation and terminology.
-\end{abstract}
+This tech report is supplemental material to our main paper; we assume the reader is already familiar with our motivation and terminology.
 
@@
 
-\section{Acknowledgements}
-
-%{\em removed for double blind review}
-Michael J. ``Sully'' Sullivan, Carlo Angiuli, and Nathaniel Filardo double-checked the proof and provided valuable feedback on notation style and the flow of explanation.
-Thanks as well to the anonymous OOPSLA reviewer \#4, whose suggestions helped improve the formality of our definitions.
-
-This work was supported in part by
-the member companies of the PDL Consortium (Broadcom,
-Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
-Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
-Tintri, Two Sigma, Uber, Veritas and Western Digital)
-and by the U.S. Army Research Office under grant number W911NF0910273.
-
 \bibliographystyle{abbrvnat}

### 2016-09-06 oopsla-16/proof/recycle.tex
Finish oopsla proofs for TRification

@@
 
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& {\bf Thread 1} & {\bf Thread 2} \\
-	1 & & \texttt{z = malloc(42);} \\
-	%2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
-	2 & & \texttt{x2 = \&z[50];} \\
-	3 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	4 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	5 & \texttt{x1 = malloc(...);} & \\
-	6 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	7 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
-The mallocs have been reordered and may no longer collide, which wrongly appears to be a false positive.}
-\label{fig:overrun-notenough}
-\end{figure}
-
-\begin{figure}[t]
-	\small
-\begin{tabular}{rll}
-	& {\bf Thread 1} & {\bf Thread 2} \\
-	1 & & \texttt{z = malloc(42);} \\
-	%2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
-	2 & & \texttt{x2 = \&z[50];} \\
-	3 & \texttt{x1 = malloc(...);} & \\
-	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
-	6 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
-	7 & \texttt{\hilight{olivegreen}{free}(x1);} \\
-\end{tabular}
-	\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls producing $a_1$ and $a_2$ must not be disrupted compared to the original interleaving.}
-	%, compared to the trace which originally exposed the malloc-recycle race.}
-\label{fig:overrun-goal}
-\end{figure}
-
 % For shown it is in complexity class, not possible to complete,
@@
 %This spells QED so we are done.
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & & \texttt{z = malloc(42);} \\
+	2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
+	3 & & \texttt{x2 = \&z[50];} \\
+	4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	5 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	6 & \texttt{x1 = malloc(...);} & \\
+	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+\end{tabular}
+\caption{Without a PP between lines 4 and 5 of Figure~\ref{fig:overrun}, this is the only alternate interleaving DPOR would explore.
+The mallocs have been reordered and may no longer collide, which wrongly appears to be a false positive.}
+\label{fig:overrun-notenough}
+\end{figure}
+
+\begin{figure}[t]
+	\small
+\begin{tabular}{rll}
+	& {\bf Thread 1} & {\bf Thread 2} \\
+	1 & & \texttt{z = malloc(42);} \\
+	2 & & \texttt{\hilight{commentblue}{// TODO bounds check??}} \\
+	3 & & \texttt{x2 = \&z[50];} \\
+	4 & \texttt{x1 = malloc(...);} & \\
+	5 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+	6 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	7 & \texttt{\hilight{brickred}{x1->foo = ...;}} & \\
+	8 & \texttt{\hilight{olivegreen}{free}(x1);} \\
+\end{tabular}
+	\caption{Goal interleaving of Figure~\ref{fig:overrun}. To ensure collision, the sequence of malloc calls producing $a_1$ and $a_2$ must not be disrupted compared to the original interleaving.}
+	%, compared to the trace which originally exposed the malloc-recycle race.}
+\label{fig:overrun-goal}
+\end{figure}

### 2016-09-06 oopsla-16/proof/di.tex
Finish oopsla proofs for TRification
oopsla proof  - correct some stale text about yield

@@
 we omit higher-level abstractions such as $\mathsf{cond\_wait}$, $\mathsf{create}$, or $\mathsf{join}$, which can be implemented using these primitives \cite{thrlib}.
-$\mathsf{yield}$ has no effect under these execution semantics, but is included for the sake of synchronization API preemption points (see below).
+%$\mathsf{yield}$ has no effect under these execution semantics, but is included for the sake of synchronization API preemption points (see below).
+$\mathsf{yield}$
+allows the execution semantics to switch threads %without blocking the invoker,
+while respecting the preemption-point-switching invariant discussed below.
+%would have no effect under an execution semantics which schedules threads arbitrarily
+%but is included for the sake of synchronization API preemption points (see below).
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Intuition}
+
+This section provides (hopefully) intuitive summaries of our proof goals.
+%for readers not interested in double-checking the proofs' internal structure.
+
+{\bf Intuition for Iterative Deepening convergence.}
+%In summary, we are proving
+We will prove
+that when Iterative Deepening saturates the set of data-race PPs,
+that set represents every instruction where a preemption could possibly affect the program's behaviour
+Hence, completing the associated state spaces is as strong a verification as testing all possible thread interleavings by preempting anywhere.
+A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
+but the technique's iterative nature will eventually find it.
+On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
+so any such PPs are irrelevant.
+
 \begin{figure}[t]
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Intuition}
-
-This section provides (hopefully) intuitive summaries of our proof goals.
-%for readers not interested in double-checking the proofs' internal structure.
-
-{\bf Intuition for Iterative Deepening convergence.}
-%In summary, we are proving
-We will prove
-that when Iterative Deepening saturates the set of data-race PPs,
-that set represents every instruction where a preemption could possibly affect the program's behaviour
-Hence, completing the associated state spaces is as strong a verification as testing all possible thread interleavings by preempting anywhere.
-A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
-but the technique's iterative nature will eventually find it.
-On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
-so any such PPs are irrelevant.
-
 Note that while we defined synchronization API PPs to occur both before and after any $\mathcal{A}_{sync}$ instruction,
